

\begin{exercises}

\item Given $\Pr{A}=0.3$, $\Pr{B}=0.5$ and $\Pr{A\cap B}=0.24$,
find
\begin{enumerate}
\item $\Pr{A\cup B}$
\item $\Pr{\overline{A}\cap B}$
\item $\Pr{\overline{A} \cup \overline{B}}$
\item $\Pr{A\mid B}$
\item Are events $A$ and $B$ independent?
\end{enumerate}

\begin{answer}
You can use Venn diagrams to make the formulae clearer.
\begin{enumerate}
\item $\Pr{A\cup B}=\Pr{A} + \Pr{B} - \Pr{A\cap B} = 0.56$.
\item $\Pr{\overline{A}\cap B} = \Pr{B} - \Pr{A\cap B} = 0.26$.
\item $\Pr{\overline{A}\cup\overline{B}} = \Pr{\overline{A}} +
\Pr{\overline{B}} - \Pr{\overline{A}\cap\overline{B}} = 0.76$ or
$1-\Pr{A\cap B} = 0.76$.
\item $\Pr{A\mid B} = \Pr{A \cap B}/\Pr{B} =
0.24/0.50 = 0.48$.
\item Since $\Pr{A}\times \Pr{B} = 0.3\times
0.5 = 0.15\ne \Pr{A\cap B}$, $A$ and $B$ are not independent
events.
\end{enumerate}
\end{answer}

\item A computer programmer is running a subroutine to solve the
general quadratic equation $y=ax^2 + bx + c=0$. This is an
`experiment' of choosing values for the constants $a$, $b$ and $c$
satisfying the quadratic equation.
\begin{enumerate}
\item Define the sample space $S$ for the experiment.
\item Define
the the event $A$: `the equation has two equal real roots'.
\end{enumerate}

\begin{answer}
\begin{enumerate}
\item All real numbers are possible values for $a$, $b$ and $c$;
hence $S=\{ (a,b,c)\mid -\infty<a<\infty, -\infty<b<\infty,
-\infty<c<\infty\}$.
\item The solutions to a quadratic equation
are given by
\[
   x = \frac{-b \pm \sqrt{b^2 -4ac} }{2a}.
\]
For two equal roots, $b^2-4ac=0$. So event $A$ is defined as $A=\{
(a,b,c)\mid b^2-4ac=0\}$.
\end{enumerate}
\end{answer}

\item
Senie~\etal~\cite{BIB:Senie:breast} studied the relationship
between breast self-examination and the age of diagnosis of breast
cancer; the data are shown in Table~\ref{TB:prob:breast}.
\begin{enumerate}
   \item
   Compute
   $\Pr{\mbox{age$<$45}}$.
   \item
   Compute
   $\Pr{\mbox{conduct self-examinations monthly}}$.
   \item
   Compute
   $\Pr{\mbox{age under 45}\cap \mbox{never use self-examination}}$.
   \item
   Compute
   $\Pr{\mbox{age between 45 and 59}\cup \mbox{never use self-examination}}$.
   \item
   Compute
   $\Pr{\mbox{age between 45 and 59} \mid  \mbox{never use self-examination}}$.
   \item
   Is the frequency of breast self-examination
   independent of age at diagnosis?
   Justify.
\end{enumerate}


\begin{table}[hb]
\begin{center}
\begin{tabular}{cccc}
& \multicolumn{3}{c}{Frequency of self-examination}
\botnewrule\\
\cline{2-4} \topnewrule Age & Monthly & Occasionally & Never
\botnewrule \\
\hline \topnewrule
$<$ 45 & 91 & 90 & 51 \\
$45$--$59$ & 150 & 200 & 155 \\
$60+{}$ & 109 & 198 & 172
\end{tabular}
\caption{The frequency of breast self-examination according to age
at diagnosis of breast cancer.} \label{TB:prob:breast}
\end{center}
\end{table}


\begin{answer}
$232/1216$; $350/1216$; $51/1216$;
$(150+200+155+51+172)/1216=728/1216\approx 0.599$; $155/378$; The
variables are not independent (eg $\Pr{\mbox{age between 45 and
59} \mid  \mbox{never use self-examination}}=0.599$
but \\
$\Pr{ \mbox{age between 45 and 59}} = 505/1216 \approx 0.415$.
\end{answer}

\item A courier company is interested in the length of time a
certain set of traffic lights is on green. The lights have been
set so that the longest possible time between green lights in any
one direction is 2 minutes, and the shortest possible time is 15
seconds. An employee is instructed to observe the lights and
record the length of time between consecutive green lights.
\begin{enumerate}
   \item What is the variable of interest?
   \item What is the sample space?
   \item Can the classical approach to probability be used to
         determine \\
        $\Pr{\mbox{less than 60 seconds between green lights}}$?
         Why or why not?
    \item What about the relative frequency approach?
    If so, how?  If not, why not?
\end{enumerate}

\begin{answer}
\begin{enumerate}
\item The length of time (in seconds) between green lights at the
intersection, say G
\item $S=\{G\mid 15\le G\le 120\}$
\item No---no equally
likely events are defined
\item The probability can be *approximated*---observe the lights a large number of times
and count how often there is less than 60 seconds between green
lights.
\end{enumerate}
\end{answer}

\item On October~13, the American television programme *Nightline* interviewed Dr~Richard Andrews, director of the
California Office of Emergency Services. They were discussing
various natural disasters that were being predicted as a result of
the El Ni\~no. This is part of their interview:

> *Ted Koppel*:  Dr~Andrews, I'm sure you have heard such cautionary advice before, so on what basis is the assumption being made that this is the one that's going to have the kind of impact on southern California\dots that's being predicted?
> *Richard Andrews*: Well, in the business that I'm in and that local government and state government is in, which is to protect lives and property, we have to take these forecasts very seriously... I listen to earth scientists talk about earthquake probabilities a lot and in my mind every probability is 50--50, either it will happen or it won't happen\dots

Explain *why* Dr Andrews is incorrect. Give an example to
show why he *must* be incorrect. (Chance News~6.12.)

\begin{answer}
He is erroneously assuming that the events `an earthquake occurs'
and `an earthquake does not occur' are equally likely events.
\end{answer}

\item
The data is Table~\ref{TB:prob:pilotdeaths} was part of the data
collected by Dr Peter Ruscoe in an investigation into aviation
deaths of private pilots in Australia.

\begin{enumerate}
   \item
   What is the probability that a randomly chosen death in 1997
   was of a pilot $50$ or older?
   \item
   What proportion of deaths from 1997 to 1999
   were of pilots aged under 30?
   \item
   What other information may be useful
   in studying the effect of age on pilot deaths?
\end{enumerate}

\begin{table}
\begin{center}
\begin{tabular}{cccc}
    & \multicolumn{3}{c}{Deaths in the year:}
    \botnewrule\\
    \cline{2-4}
    \topnewrule
Age &   1997 & 1998 & 1999
\botnewrule\\
\hline \topnewrule
$<30$    &   2  &  1  &  3\\
$30-49$ &   5  &  12  &  5\\
$\ge 50$ &   9  &  11  &  9\\
\end{tabular}
\caption{Aviation deaths of private pilots in Australia from 1997
to 1999 according to the pilot's age.} \label{TB:prob:pilotdeaths}
\end{center}
\end{table}

\begin{answer}
$9/16$; $6/57$; pilot experience, type of flight (solo, etc.),
type of aircraft, etc.
\end{answer}

\item
A bag contains 21 balls, which are either blue or red. Two balls
are drawn, without replacement, from the bag.

\begin{enumerate}
\item Define the sample space for the experiment. \item If it is
known that the probability of a blue ball being drawn
      followed by a red ball is $7/30$,
      determine the number of red balls in the bag.
\item What would happen if the answer to the previous part was
determined
      assuming that the balls were drawn *with* replacement?
\end{enumerate}

\begin{answer}
$S=\{ (B,B); (B,R); (R,B), (R,R)\}$ where $B$ refers to drawing a
blue ball and $R$ to drawing a red ball; $7$ or $14$ red balls;
$13.2111$ or $7.7889$, neither of which are possible.
\end{answer}

\item
International practice is for electronic mail (or e-mail)
addresses to conclude with a two letter code indicating the
country of origin.  E-mail addresses in Australia, for example,
end with \texttt{au}, and so e-mail addressed to Queensland
government departments will often have the form
\url{persons-name@department-name.qld.gov.au}.
\begin{enumerate}
\item How many countries can be accommodated in such a scheme?
\item Do you think this is a sufficient number?
\item To avoid potential confusion between similar country codes, suppose it was
decided that countries could not use the same two letters in their
code as any other country.  (That is, no country would be allowed
to use the suffix \texttt{ua}, as it uses the same two letters as
Australia's code.) How many countries can be accommodated in this
scheme?
\end{enumerate}

\begin{answer}
$26^2=676$; There are under 200 sovereign countries, so should be
sufficient (but takes no account of whether there are enough
*sensible* codes for each country); $C^{26}_2=26\times 25/2
=325$.
\end{answer}

\item
At the Cricket World Cups, each cricket squad consist of
fifteen players from which a team of eleven must be chosen for
each game. The squad consists of five bowlers, seven batsmen, two
all-rounders and and one wicketkeeper.

\begin{enumerate}
    \item
    Find the number of teams possible if the team consists of
    five batters,
    four bowlers,
    one all-rounder
    and one wicketkeeper.

    \item
    After a game concludes,
    each member of one team shakes hands with
    each member of the opposing team,
    and each member of both teams shakes hands
    with the two umpires.
    How many handshakes are there in total at the conclusion of a game?

\end{enumerate}

\begin{answer}
$C^7_5 \times C5_4 \times C^2_1 \times C^1_1$; $11^2 + (2\times
22) =1 65$.
\end{answer}

\item
Suppose I am dealt five cards from a pack of cards (as in
poker).  Explain why the number of possible hands is ${52\choose
5}$ and not $52\times51\times50\times49\times48 = {}^{50}P_5$.

\begin{answer}
Because the order is not important (after getting the cards, they
can be arranged in any order in your hand).
\end{answer}




\item\label{QN:gameshow}
(Computer exercise) Use \R\ to simulate the gameshow described
in Section~\ref{SC:gameshow}.

\begin{answer}
{\em Step 1}: Generate a random sequence of length 100 consisting of the integers
1,2 and 3, putting the sequence in column C1 to represent the position of the car.\\
{\em Step 2}: Generate another such sequence (and put in column C2) to represent
the contestants first choice on each of the 100 nights.\\
{\em Step 3}: Count the number of times the numbers in the two column agree representing
the number of times the contestant will win {\em if the contestant doesn't change doors}.
If the numbers in the two columns don't agree then the contestant will win
*only if the contestant decides to change doors*.

```{r}
car.door <- sample(1:3, 
                   size = 100,
                   replace = TRUE)
first.choice <- sample(1:3,
                       size = 100,
                       replace = TRUE)
confusion.table <- table(car.door,
                         first.choice)
agree<-sum(diag(confusion.table))
```

One simulation produced the confusion table
\begin{center}
\begin{tabular}{rrrr}
 & \multicolumn{3}{c}{first.choice}  \\
car.door & 1 & 2 & 3 \\
     1   & 14 & 12 & 13 \\
     2 &   13 &  9 &  8 \\
     3   & 14 & 10 & 7 \\
\end{tabular}
\end{center}

Notice contestants would have won the car
30 times if they retained their first choice which means they
would have won 70 times if they had changed.

The apparent paradox is explained by the fact that the contestant did not know about the second chance at the start.
\end{answer}

\item
(Computer exercise) Use \R\ to simulate the experiment described in Section~\ref{SC:CAO}.

\begin{answer}

```{r}
sim.results<-numeric(0) # names a NULL vector ready to collect results
for ( i in 1:1000){
   Ncond1 <- sum(sample(0:1,
                        size = 21,
                        replace = TRUE)) # number of 1's in sample of 21 (0,1)
sim.results <- c(sim.results,
                 Ncond1) } # accumulate incidences of 1's
print(table(sim.results))  # frequency of counts out of 21
```
\end{answer}

\end{exercises}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\section{Exercises}

\begin{exercises}

\item
For any two events $A$ and $B$ in the sample space $S$ prove (a),
(b) and (c), stating condition(s) for the equality sign to hold.
\begin{enumerate}
\item $ \Pr{A \cup B} \le \Pr{A} + \Pr{B} $
\item $ \Pr{A \cap B} \le \Pr{A} $
\item $ \Pr{A \cup B} \ge \Pr{A} $
\item If $ \Pr{A}=2/3$ and $\Pr{A \cup B}=3/4$ what are the upper and lower bounds
on the value of $\Pr{B}$.
\end{enumerate}
%hint
\begin{answer}
\begin{enumerate}
\item Use Theorem~\ref{TM:probaddition} noting $\Pr{A \cap B} \ge 0$.
(What is the condition on $A$ and $B$ for equality?)
\item First write $A$ as union of two mutually exclusive events; that is,
 $A=(A \cap B) \cup (A \cap \overline{B})$.
\item Note $A \cup B = A \cup (B \cap \overline{A})$.
\item $1/12 \le \Pr{B} \le 3/4$.
\end{enumerate}
\end{answer}


\item
Given an experiment such that $\Pr{A}=\frac{2}{3}$, $\Pr{B}=\frac{1}{4}$,
$\Pr{A \cup B}=\frac{5}{6}$, evaluate the following algebraically
using set identities and
the rules of probability.\\[3mm]
$\begin{array}{l l}
(\mbox{a}) \ \Pr{\overline{B}} & (\mbox{b}) \ \Pr{A \cap B}\\
(\mbox{c}) \ \Pr{A \cap \overline{B}}&
(\mbox{d}) \ \Pr{\overline{A} \cap \overline{B}}\\
(\mbox{e}) \ \Pr{\overline{A} \cup \overline{B}}&
(\mbox{f}) \ \Pr{A \cup \overline{B}}
\end{array}$.

\begin{answer}
Note using Venn diagrams alone is not acceptable.\\
(a) 3/4 \ (b) 1/12  \ (c) 7/12 \ (d) 1/6 \ (e) 11/12 \ (f) 5/6
\end{answer}


\item If a box contains 100 tickets numbered from 1 to 100 inclusive and four
tickets are drawn from the box one at a time (without replacement),
find the probability that
\begin{enumerate}
\item all four numbers drawn are even,
\item exactly two even numbers are drawn,
\item at least two odd numbers are drawn before drawing the first even number,
(at least one even number to be drawn).
\item the sum of the numbers drawn is odd.
\end{enumerate}

\begin{answer}
\begin{enumerate}
\item $\displaystyle{\frac{188}{3201}}$
\item $\displaystyle{\frac{1225}{3201}}$
\item $\Pr{\mbox{2 odd before 1st even}}$\\
$= \Pr{\mbox{odd,odd,even,odd or even}} + \Pr{\mbox{odd,odd,odd,even}}
=\displaystyle{\frac{3625}{19206}}$
\item $\Pr{\mbox{sum odd}} = \Pr{\mbox{1 odd number OR 3 odd numbers drawn}}
  = \displaystyle{\frac{1600}{3201}}$
\end{enumerate}
\end{answer}


\item Prove that if $A$ and $B$ are events with non-zero probabilities then
if $A$ and $B$ are independent they are not mutually exclusive.

\begin{answer}
If $A$ and $B$ are independent then $\Pr{A \cap B} = \Pr{A}\Pr{B} > 0$. Hence
show condition for $A$ and $B$ mutually exclusive not satisfied.
\end{answer}


\item An ordinary die is rolled twice and the number of of spots on the
upper face is observed after each roll. The events $A$,$B$ and $C$ are defined
as\\
$A$: The number of spots showing on the first roll is 3.\\
$B$: The sum of the number of spots in the two rolls is 6.\\
$C$: The sum of the number of spots  in the two rolls is 7.\\
Show that $A$ and $B$ are not independent but that $A$ and $C$ are independent.
Are $B$ and $C$ independent?

\begin{answer}
Apply Definition~\ref{DF:independence}.
\end{answer}


\item In a two person game a fair die is thrown in turn by each player. The
player who throws a six first wins.
\begin{enumerate}
\item Find the probability that the first player to throw the die wins.
\item Show that if the player to throw first is selected by the toss of a fair coin
then each player has an equal chance of winning.
\end{enumerate}

\begin{answer}
$\Pr{\mbox{Player throwing first wins}}
= \Pr{\mbox{First six on throw 1 or 3 or 5 or \dots}}
= \Pr{\mbox{First six on throw 1}} + \Pr{\mbox{First six on throw 3}} + \cdots$
and sum the geometric progression obtained.\\
% (See {\em Important Preliminaries}, page (vi) of study guide.)
Use Theorem~\ref{TM:totalprob} for the second part. Define the events
$A=\{\mbox{Player 1 wins}\}$, $B_1 = \{\mbox{Player 1 throws first}\}$,
$B_2 = \{\mbox{Player 1 throws second}\}$.
\end{answer}


\item \label{QN:1:randresp}
In order to get honest answers
to sensitive questions the *randomised response technique* is sometimes used.
For example, suppose the aim is
to discover the proportion of students who have used illegal
drugs in the past twelve months. To do this we prepare $N$ cards. On $m$
of these is the statement `I have used an illegal drug in the past twelve
months'. On the remaining $N-m$ cards is the statement
`I have not used an illegal drug
in the past twelve months'.\\
A number of students are now chosen at random from all the students available.
Each student so chosen then selects one card
at random from the prepared pile of $N$ cards and answers yes or no to the question
`Is the statement on the card you have selected true?'
without divulging which statement is actually on the card. If $\Pr{Y}$ is the
probability that a student answers yes, and $\Pr{D}$ is the probability that
a student chosen at random has used an illegal drug and
assuming that each student answers truthfully,
\begin{enumerate}
\item find $\Pr{Y}$ in terms of $\Pr{D}$, $m$ and $N$ stating any conditions
required on $m$ and $N$.
\item If in a sample of 400 students, 175 answer yes so that the estimate of $\Pr{Y}$
is $\frac{175}{400}$, estimate $\Pr{D}$ from the expression found above, given that
$N=100$ and $m=25$.
\end{enumerate}

\begin{answer}
\begin{enumerate}
\item Use Theorem~\ref{TM:totalprob} to find $\Pr{Y}$.\\
$Y$ = \{Yes\}, $D$ = \{Student has used illegal drugs in last 12 months\},
$\overline{D}$ = \{student hasn't used illegal drugs in last 12 months\}.
\item $\displaystyle{ \Pr{D}= \frac{N\Pr{Y}-[N-m]}{2m-N}}$
\end{enumerate}
\end{answer}


\item\label{QN:1:multchoice}
A multiple choice question contains $m$
possible choices. There is
a probability of $p$ that a candidate chosen at random will know the
correct answer (and probability $q = 1 - p$ that this candidate will
not know the answer). If a candidate does not know the answer, the candidate
guesses and is equally likely to select any of the $m$ choices. \\
If a candidate is
randomly selected what is the probability of the question being answered correctly?

\begin{answer}
Use Theorem~\ref{TM:totalprob} to find $\Pr{C}$ where $C =$ \{correct answer\},\\
 $K =$ \{student knows answer\}. Ans: $\Pr{C}=\displaystyle {\frac{mp+q}{m}}$
\end{answer}




\item
Bateson~\cite{BIB:Bateson:mendel} repeated many of the famous
experiments of Mendel (see Mendel~\cite{BIB:Mendel:hybrids}). One
of these experiments involved crossing a pure strain of peas with
purple flowers and long pollen, with a pure strain of peas with
red flowers and round pollen. In the second generation, the data
in Table~\ref{TB:prob:bateson} were collected.

\begin{table}
\begin{center}
\begin{tabular}{lccc}
& Purple & Red & Total
\botnewrule \\
\hline \topnewrule
Long pollen & 1528 & 117 & 1645 \\
Round pollen & 106 & 381 & 487%
\botnewrule\\
\hline \topnewrule Total & 1634 & 498 & 2132
\end{tabular}
\caption{The results in the second generation of Bateson's
experiment crossing a pure strain of peas with purple flowers and
long pollen, with a pure strain of peas with red flowers and round
pollen.} \label{TB:prob:bateson}
\end{center}
\end{table}

\begin{enumerate}
   \item
   Determine $\Pr{\mbox{red flower}}$ from the data.
   \item
   Determine $\Pr{\mbox{round pollen}}$ from the data.
   \item
   Do the data suggest the colour of the pollen
   is independent of the shape?  Explain.
   \item
   Determine $\Pr{\mbox{red}\cap\mbox{long}}$.
   \item
   Determine $\Pr{\mbox{red}\cup\mbox{long}}$.
\end{enumerate}



\item

A computer system requires all user to have passwords, which must
be made up entirely of the letters \texttt{a} to \texttt{z}.
Capital letters are considered to be different than lower case
letters.  (For example, the passwords \texttt{aaaaaa},
\texttt{AAAAAA}, \texttt{aaaAAA} and \texttt{AaaAaA} are all
considered different passwords.)

\begin{enumerate}
\item Determine the total number of passwords that are possible if
every password must contain exactly six letters.


\item Determine the total number of passwords possible if every
password must be at least four letters long, and at most six
letters long.


\item Determine the total number of passwords if no letter can be
repeated in any one password.  (The passwords must still be at
least four and at most six letters long, and lower case and upper
case letters are still considered different.)

\end{enumerate}


\item\label{QN:CHD}
Table~\ref{TB:CHD} displays data from a cross-classification of coronary heart
disease (CHD) by coffee drinking and smoking habits in a
case-control study of 66 CHD cases and 85 unmatched controls. The
results are all for men aged 40 to 55. Heavy smokers are defined
as those individuals who smoke one pack per day or more. The data
are from Paul et al.~\cite{BIB:Paul:CHD}.

\begin{table}
\begin{center}
\begin{tabular}{cccccc}
                & \multicolumn{2}{c}{\textit{Heavy Smokers}} && \multicolumn{2}{c}{\textit{Others}}
                \botnewrule \\
                \cline{2-3} \cline{5-6}
                \topnewrule
                & \multicolumn{2}{c}{\textit{Coffee use}} && \multicolumn{2}{c}{\textit{Coffee use}} \\
                & \textit{Heavy} & \textit{Other} && \textit{Heavy} & \textit{Other}
                \botnewrule \\
                \hline
                \topnewrule
\textit{CHD cases}       &   25  &   11  &&   15  &   15\\
\textit{Non-CHD cases}   &   14  &   8   &&   21  &   42
\end{tabular}
\caption{Data for Exercise~\ref{QN:CHD}}\label{TB:CHD}
\end{center}
\end{table}

\begin{enumerate}

   \item
   What is the probability that a randomly selected individual is a heavy coffee drinker?

   \item
   What is the probability that a randomly chosen individual
   is a heavy coffee drinker and a heavy smoker?

   \item
   Suppose you meet one of the men in the study whom you know does not smoke.
   What is the probability this man will have had a CHD problem?

   \item
   Could the data be used to conclude that coffee use causes CHD?

   \item
    Determine if coffee use and CHD cases appear to be independent,
    justifying your answers with appropriate probability calculations.


\end{enumerate}


\item Consider the following argument.

\begin{quote}
   When I toss a coin twice, there are only three outcomes:
   a Head and a Head; a Tail and a Tail; or one of each.
   So the probability of obtaining two Tails must be one-third.
\end{quote}

Explain *why* the reasoning is incorrect, and then determine the correct answer.


\item Consider the following experiment: A die is loaded so that
the chance of obtaining a \epsdice{1} is three times the chance of
obtaining a \epsdice{2}. The chance of obtaining a \epsdice{2} is
the same as obtaining a \epsdice{3}, the same as obtaining a
\epsdice{4}, the same as obtaining a \epsdice{5}, the same as
obtaining a \epsdice{6}. The experiment consists of rolling the
die once and observing the uppermost face.

\begin{enumerate}
\item List the sample space for the experiment. \item Give the
definition for the classical approach to probability.
      Is it possible to use the classical approach to probability
      to determine $\Pr{\mbox{a \epsdice{1} is rolled}}$?
      If so, determine the probability.  If not, why not?
\item Is it possible to use the relative frequency approach to
probability
      to determine
       $\Pr{\mbox{a \epsdice{1} is rolled}}$.
      If so, how?  If not, why not?
\item Is it possible to use the three axioms of probability
      to determine $\Pr{\mbox{a \epsdice{1} is rolled}}$.
      If so, how?  If not, why not?
\item A class of students is given the die without being told it
is loaded
      and is asked to determine $\Pr{\mbox{a \epsdice{2} is rolled}}$ by
      repeated rolls of the die.
      One group of students repeats the experiment 100 times and records
      a \epsdice{2} on 15 occasions.  Do you think they have evidence that
      the die is loaded?  Why or why not?
%\item Another group of student tosses the die 500 times and records a
%      \textbf{2} 61 times.  What is the probability of obtaining 61 \textbf{2}
%      (or something more extreme) from 500 tosses if the die was
%      fair\footnote{This hasn't strictly been covered
%      yet.}?  What should the students conclude?
\end{enumerate}

\item Prove, using only the three axioms of probability, that
$\Pr{A}\le 1$ for any event $A$.


\item Partridge~\cite[p~21]{BIB:Partridge:willitrain} gives the
data in Table~\ref{TB:prob:SOI} showing the relationship between
the southern oscillation index (SOI) and rainfall in eastern
Australia. The proportions can be considered as historical
probabilities.

\begin{table}
\begin{center}
\begin{tabular}{cccc}
Average     & \multicolumn{3}{c}{Proportion of Eastern Australia} \\
Annual SOI  & \multicolumn{3}{c}{receiving annual rainfall:}
\botnewrule\\
\cline{2-4} \topnewrule
        & below average & average   & above average
\botnewrule\\
\hline \topnewrule
below $-10$ & 0.63 & 0.31 & 0.06 \\
$-10$ to $10$ & 0.25 & 0.48 & 0.27\\
above $10$ & 0.03 & 0.26 & 0.71
\end{tabular}
\caption{The effect of the southern oscillation index (SOI) on
rainfall in eastern Australia. The table shows the proportion of
years that historically give significantly above and below average
rainfall in years with the given average SOI.}\label{TB:prob:SOI}
\end{center}
\end{table}


\begin{enumerate}
   \item
   What method has been used to compute the probabilities given in the table?

   \item
   Can the data be used to make predictions about the future?
   Explain.

   \item
   Are the SOI and rainfall in eastern Australia independent?
   Justify your answer.
   What does this mean in this context?

   \item
   Construct a tree diagram that summarises the data.

   \item
   The text in Partridge~\cite[p~20]{BIB:Partridge:willitrain}
   implies that for $8$ years, the average annual SOI was below $-10$,
   and that for $8$ years the average annual SOI was above $10$;
   for the remaining $53$ years the average annual SOI was between $-10$ and $10$.

Use this information to determine the following historical probabilities.
   \begin{enumerate}
      \item
      $\Pr{\mbox{SOI below $-10$} \cap \mbox{below average rainfall in eastern Australia}}$;
      \item
      $\Pr{\mbox{SOI above $-10$} \mid  \mbox{below average rainfall in eastern Australia}}$;
      \item
      $\Pr{\mbox{receiving below average rainfall in eastern Australia}}$.
   \end{enumerate}

\end{enumerate}


\item
The Brisbane newspaper, \emph{The Sunday Mail}, gave details about
the numbers of people who, after strip-searching, were
subsequently charged. The data are from the twelve months from
April 1998 and May 1999 for various districts in Queensland. The
following data were deduced from the article for the Cairns
region: the probability that a person was charged in Cairns after
being strip-searched is $0.077$ (which seems frighteningly low);
the probability that the person was male and also charged after
being strip-searched is $0.074$; the probability that a person was
male is $0.903$.

Construct a Venn diagram to answer the following questions.
\begin{enumerate}
   \item
   Determine the probability that a person who was strip-searched was female.

   \item
   Determine the probability that a female who was strip-searched was then charged.

   \item
   Determine the probability of selecting a male who was charged.

   \item
   Determine the probability of either a female \emph{or}
   someone who was not charged.

\end{enumerate}

Repeat the questions using a table rather than a Venn diagram.

%     Y   N
% M  26  327  353
% F   4   34   38
% ---------------
% T  30  361   391
%

\item

Here is an amazing true story about the Arizona Lottery (Chance
News, 27 June 1998 to 8 August 1998 edition). The Arizona Lottery
started a new game ``Pick~3'' on May 3, 1998. In Pick~3, three
numbers are chosen from zero to nine (repetitions are allowed).
Three \emph{ordered} winning numbers are picked and prizes are
awarded depending on how the numbers match the winning numbers.
New winning numbers are picked every day of the week except
Sunday.

After 32 games, no nines had been drawn in any of the 96 winning
numbers. One woman, who always chose 9-0-7, called to complain
that there must be a problem. It was later discovered that the
company supplying the random number generator provided a program
that did not select nines.

About 1.2 million tickets with nines had been sold during this
period. To compensate, lottery officials ran the same game for the
period July~15 to July~31 but with all the prizes doubled. They
also went back to the old-fashioned method of using numbered balls
to determine the winning numbers.

\begin{enumerate}
\item Determine the probability that 96 numbers are selected from
zero to nine
      without selecting a nine.
      When do you think the error could reasonably have been detected?

\item When you play the game for \$1, three of the options
available to play are
      Exact Order, Front Pair and Back Pair. Three
      ordered winning numbers are then drawn. If you choose ``Exact Order''
      and your numbers agree with the winning
      numbers in their chosen order, you win \$500. If
      you chose ``Front Pair'' and your first two numbers match the first
      two numbers of the winning numbers in their chosen order,
      or you chose ``Back Pair'' and your last two numbers match the last
      two of the winning numbers in their chosen order order, you win \$50.
      Find the expected winnings under each choice. Were any of the
      choices favourable games when the prizes were doubled?
\item How much do you think the doubling of prize money
      affected the number of people who played the game?
\end{enumerate}


\item\label{QN:Goodman}
The data in Table~\ref{TB:Goodman} were collected by L.A.
Goodman~\cite{BIB:Goodman}. They concern eye and hair colour of
5387 children in the Scottish town of Caithness.

\begin{table}
\begin{center}
\begin{tabular}{lccccc}
&\multicolumn{5}{c}{\textit{Hair Colour}}
\botnewrule\\
\cline{2-6} \topnewrule \textit{Eye Colour} &
\textit{Fair}&\textit{Red}&\textit{Medium}&
                      \textit{Dark}&\textit{Black} \\
 \botnewrule        & (HF) & (HR) & (HM) & (HD) & (HB) \\
\hline
\topnewrule\textit{Blue} (EB)& 326 & 38 & 241 & 110 & 3 \\
\textit{Light} (EL) & 688 & 116 & 584 & 188 & 4 \\
\textit{Medium} (EM) & 343 & 84 & 909 & 412 & 26 \\
\botnewrule\textit{Dark} (ED) & 98 & 48 & 403 & 681 & 85 \\
\hline
\end{tabular}
\caption{Data for Exercise~\ref{QN:Goodman}}\label{TB:Goodman}
\end{center}
\end{table}
%%   HAND 188

\begin{enumerate}
\item Determine the following probabilities.
\begin{enumerate}
\item $\Pr{\mbox{Blue eyes}}$; \item $\Pr{\mbox{Dark or black
hair}}$; \item $\Pr{\mbox{HF}\cap\mbox{ED}}$; \item
$\Pr{\mbox{HM}\mid \mbox{EB}}$.
\end{enumerate}

\item Explain the \emph{meaning} of each of the following in words
and in the context of the question (you do \emph{not} have to
determine probabilities).
\begin{enumerate}
\item $\mbox{HF}\cap\mbox{EB}$; \item $\mbox{EM}\cup\mbox{EL}$;
\item $(\mbox{HF}\cup\mbox{HR})\mid \mbox{ED}$.
\end{enumerate}

\item Does it appear that hair colour and eye colour are
independent? Justify your answer.
\end{enumerate}


\item
A rugby league team of 13 consists of five backs, six forwards and
two halves.
\begin{enumerate}
   \item
   How many teams are possible from a squad of 17,
   consisting of six backs,
   eight forwards and three halves?

   \item
   How many teams are possible from a squad of 17,
   consisting of six backs,
   eight forwards and two halves,
   and one player who could play as a back or a half?

\end{enumerate}




\item \label{QN:payoff}
In \emph{Gold Lotto}, players pick any six numbers from  the
numbers one through to 45 (inclusive). Six winning numbers are
then drawn (without replacement), and players who have selected
the same six numbers win the first prize. (Two ``supplementary
numbers'' are also drawn, but we will not be concerned with them.)
To play, players need to play a minimum of four games, costing
\$1.75 (at the time of writing). Suppose the McMillan family play
the minimum four games every week.
\begin{enumerate}
\item What is the probability of the McMillans correctly choosing
the six winning numbers?
%1.2277 x 10^{-7}, or 1/( 45-C-6 )
\item The expected payoff is the amount one can expect to win or
lose for each game played in the long term.  It is calculated as
\begin{align*}
\hbox{Expected Payoff}=\,&\Pr{\hbox{win}}\times\hbox{(Net amount won)} +\\
                       & \quad\Pr{\hbox{lose}} \times \hbox{({Net amount lost})}.
\end{align*}
Suppose the first prize is one million dollars (\$1\,000\,000).
(This may not be true---the first prize changes weekly.) What is
the expected payoff from playing Gold Lotto?
%It is possible for there to be more than one winner.  In this case, the
%winners divide the first prize evenly between them.  Determine the
%expected payoff for the McMillans if there are two winners.
\end{enumerate}




\item
In the game of \emph{Yahtzee}, players roll five six-sided dice
and are awarded points for various combinations that appear.
Assuming the five dice are fair, determine the following.
\begin{enumerate}
\item The probability that all five dice show the same number (for
example,
       \epsdice{2}, \epsdice{2}, \epsdice{2}, \epsdice{2}, \epsdice{2}).
\item The probability of obtaining three dice with the same
number, and
      the remaining two dice as the same number, but different
      to what is on the other three (a ``full house''; for example,
      \epsdice{3}, \epsdice{3}, \epsdice{3}, \epsdice{6}, \epsdice{6}.)
\item The probability of obtaining one pair.  (That is, two dice
      have the same number, but the other three are all different than
      each other, and also different than the pair; for example
      \epsdice{1}, \epsdice{1}, \epsdice{3}, \epsdice{5}, \epsdice{4}).
\item The probability of obtaining a ``large (or long) straight''.
      (That is, five consecutive numbers, like \epsdice{1}, \epsdice{2},
      \epsdice{3}, \epsdice{4}, \epsdice{5}.)
\end{enumerate}


\item
In Orleans Parish, Louisiana, the law states juries must be
selected by jury commissioners at random from a list compiled from
the general population. In 1958, the list for one case consisted
of 30~black people and 40~white people. Twelve people were chosen
for the jury, none of whom were black.

\begin{enumerate}

   \item
   The commissioners claimed that the jury was selected
   according to the law
   (that is, at random).
   Given that this is true,
   what is the probability that such a jury was selected?

   \item
   Do you believe the jury really was selected at random?
   (Justify your answer.)

\end{enumerate}

% Brockett and Levine, p 101 question 46.



\item
In drug testing, there are two keys aspects of the test:

\begin{itemize}
\item
the \emph{sensitivity} of the test, defined as:\\
$\Pr{\mbox{Positive test result}\mid \mbox{person actually has the
disease}}$; and \item
the \emph{specificity} of the test, defined as:\\
$\Pr{\mbox{Negative test result}\mid \mbox{person actually doesn't
have the disease}}$.
\end{itemize}
In testing for HIV, the standard test is the Wellcome
Elisa test, commonly called the Elisa test.  For the Elisa test,
the sensitivity is approximately 0.993 and the specificity
approximately 0.9999, according to the US Department of Food and
Drug Administration. In a research paper (Johnson and
Gastwirth~\cite{BIB:Johnson}), the incidence of HIV-positive in
the general population is estimated to be about 0.000025.


\begin{enumerate}
\item Explain \emph{why} the specificity is important.  What does
this tell you about the Elisa test?

\item Explain \emph{why} the sensitivity is important.  What does
this tell you about the Elisa test? \item Explain \emph{how} the
probabilities for specificity and sensitivity have most likely
been found.

\item Explain \emph{why} $\Pr{\mbox{a person has HIV}\mid \mbox{the
test is positive}}$ is important.

\item Explain \emph{why} $\Pr{\mbox{a person has HIV}\mid \mbox{the
test is negative}}$ is important.

\item Determine $\Pr{\mbox{a person has HIV}\mid \mbox{the test is
positive}}$.

\item Determine $\Pr{\mbox{a person has HIV}\mid \mbox{the test is
negative}}$.

\item Discuss the practical consequences of your answers to
parts~(f) and~(g) for someone who has been given the Elisa test.

\item Suppose you work as a statistician for a local hospital.
The administrator is considering compulsory HIV testing for all
new employees, by using the Elisa test.  The administrator,
Dr~Tam, has asked you to write a \emph{one page} report on the
proposal from a \emph{statistical} point-of-view. Your report
should be easily understood by Dr~Tam (who is a medical expert,
but not an expert on statistics), and should focus only on your
conclusion from a statistical point-of-view.  (That is, she will
get others' opinions on costs, medical aspects, privacy concerns,
and so on; your contribution is the statistical component only.)
You \emph{must} write this report like a real report:  addressed
to Dr~Tam, typed wherever possible, using proper language and
grammar.  This question will be marked as if you were submitting
in a real job.  While you are asked to submit a one-page report,
you can include one extra page of calculations if you wish.

Before answering the question, consider what issues are important
to Dr~Tau and to the individual staff members who have been given
the test, and what probabilities help address these issues. These
issues should be listed somewhere in the report.

\end{enumerate}



\item Consider the following events concerning the type of
computer a customer purchases: $A$: orders a 17'' monitor; $B$:
orders a computer with a ergonomic keyboard; $C$: orders a
computer with a modem.  Write in words what is \emph{meant} by the
following (don't determine probabilities).
\begin{enumerate}
   \setlength{\parsep}{0pt}
   \item $A\cap B$
   \item $A'\cup C$
   \item $A'\cap B \cap C$
   \item $C\mid A$
   \item $(B\cap C)\cup A'$
\end{enumerate}

\item Given two independent events $C$ and $D$, prove that
$\overline D$ and $C$ are \emph{not} mutually exclusive.


\item An experiment involves tossing a coin and rolling a standard
die.
\begin{enumerate}
   \item List the sample space.
   \item Determine the probability of tossing a Head and rolling
         a \epsdice{2}.
   \item Determine the probability of rolling a \epsdice{5} given
         that a Head has been tossed.  What is the \emph{reduced
         sample space} in this situation?
\end{enumerate}



\item The following problem appeared in an American newspaper.

\noindent A woman and a man (unrelated) each have two children.
At least one of the women's children is a boy, and the man's older
child is a boy.  Do the chances that woman has two boys equal the
chances that the man has two boys?

One newspaper reader wrote in

\begin{quote}
I will send \$1000 to your favourite charity if you can prove me
wrong.  The chances of both the woman and the man having two boys
are equal.
\end{quote}

What \emph{is} the answer to this problem?  If the reply is
incorrect, what was the misunderstanding? (From Chance News 6.12,
October~10, 1997 to November~9, 1997.)


\item The following item is taken from
Shen~\cite{BIB:Shen:probproof}. It is known that oceans cover more
than one half of the earth's surface. The following probabilistic
argument has been used to prove that there are at least two
symmetric points covered by water. (`Two symmetric points' are two
points on directly opposite sides of the globe.)

\begin{enumerate}
   \item
   Choose a point $X$ on the surface of the earth at random.
   Let $X'$ the point antipodal to $X$ (on the opposite side of the globe).
   Determine $\Pr{\mbox{$X$ is in the ocean}}$
   and $\Pr{\mbox{$X'$ is in the ocean}}$.

   \item
   Write an addition formula for computing $\Pr{\mbox{$X$ \emph{or} $X'$ is in the ocean}}$.

   \item
   Use this formula to deduce that
   $\Pr{\mbox{$X$ and $X'$}}>0$ for the addition formula to make sense.
\end{enumerate}



\item In a given situation, the sample space is given as
$S=\lbrace 1, 2, 3, 4, 5\rbrace$.   The probabilities of obtaining
these numbers from the sample space are only known to be:
\begin{eqnarray*}
   \Pr{1} = a, & \quad  \Pr{2} = 0.2-a, & \\
   \Pr{3} = b, & \quad \Pr{4} = 2b, \quad& \Pr{5}= 0.8-3b,
\end{eqnarray*}
where $a>0$ and $b>0$. It is also known that the events
``obtaining 1, 2 or 3'' and ``obtaining a 3 or 4'' are
independent.
\begin{enumerate}
\item Confirm that the sample space consists of the set $S$ as
      given above.
\item Use the information above to solve for $b$. \item  If it is
also known that the events ``obtain a 1 or 5'' and
      ``obtain a 1 or 3'' are also independent, show that
      $a$ can take two possible values.
\end{enumerate}

\item Sleigh~\etal~\cite{BIB:Sleigh:bell} compared two methods for
detecting \emph{Schistosoma mansoni} eggs in faeces: Bell's method
and the Kato--Katz method. Let $B$ refer to the event ``Bell's
test returned positive'' and $K$ refer to the event ``the
Kato--Katz test returned positive''. Then, the following results
were obtained from the study.
$$
   \Pr{B} = 0.756;\quad
   \Pr{K} = 0.629;\quad
   \Pr{B \cap K} = 0.584.
$$

\begin{enumerate}
   \item
   Explain in simple terms what the probability
   $\Pr{B\cap K}$ actually \emph{means}.

   \item
   Determine the probability that neither test was positive.

   \item
   Compute $\Pr{B\mid K}$.

   \item
   Compute $\Pr{B\cup K}$.

   \item
   Would you expect the two test to be independent?
   Explain.
   Perform some calculations to confirm.
\end{enumerate}




\item
The following information is taken from the abstract of Baron et
al.~\cite{BIB:Baron:colorectal}.

\begin{quotation}
   \noindent
   We performed a randomized, double-blind trial of aspirin as a
   chemopreventive agent against colorectal adenomas.
   We randomly assigned 1121 patients with a recent history of histologically documented
   adenomas to receive placebo (372 patients),
   81~mg of aspirin (377 patients),
   or 325~mg of aspirin (372 patients) daily
   \dots
   Follow-up colonoscopy was performed at least one year
   after randomization in 1084 patients (97 percent).
   The incidence of one or more adenomas was 47 percent in the placebo group,
   38 percent in the group given 81~mg of aspirin per day,
   and 45 percent in the group given 325~mg of aspirin per day.

\end{quotation}


Of the original 1121 patients, 1084 remain in the study. Assume
that the $1121-1084=37$ people who pulled out left 360 in the
placebo group, 364 in the 81~mg aspirin group, and 360 in the
325~mg aspirin group (that is, left the study randomly).

\begin{enumerate}
   \item
   Determine the proportion of people in the study
   who were found to have an adenoma.

   \item
   Given that a person has an adenoma,
   what is the probability that the person was in the placebo group?

   \item
   What is the probability that a person had no adenomas
   and was in the 325~mg of aspirin group.
   \item
   Do you think the aspirin had any effect on the
   occurrence of adenomas?
   Why or why not?
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\item(Computer exercise)  Simulate the situation in Exercise~\ref{QN:1:randresp}
when the proportion of students
taking an illegal drug is 30\%. (\Note: The advantage of {\em simulation} is that
we {\em know} what the answer should be.)

\begin{enumerate}
\item The \R\ script is listed below. The first few lines allocate numerical values
to parameters prior to generating the random samples.

```{r}
# lines like this are comments - no need to type them
#  set the parameter values
p <- 0.3           # assumed prob of a person using illegal drugs
sample.size <- 400
Ncards <- 100      # size of card deck
m <- 25            # no of cards which state that ``I have used ..''
# generate the card deck
card.deck <- c(rep(0, (Ncards - m)),
               rep(1, m))

# Draw a sample of students
incidence.sample <-  sample(c(0, 1),
                            size = sample.size,
                            replace = TRUE,
                            prob = c( (1 - p), p) )
# Draw a sample from the card deck
answer.sample <- sample( card.deck,
                         size = sample.size,
                         replace = TRUE)
# match the students with their answers
confusion.tab <- table(incidence.sample,
                       answer.sample)

print(confusion.tab)
```

\item Estimate the proportion of students using drugs by using
the result of~\ref{QN:1:randresp}.
Is the result reasonable?

\item Try the simulation again but vary the proportion of students taking drugs and
both $N$ and $m$.
\end{enumerate}


\item (Computer exercise) Simulate the situation in Exercise~\ref{QN:1:multchoice}
when 60\% of students know the correct answer.
\begin{enumerate}
\item First generate a sample of 100 students (1 = student knows answer,
0=student doesn't know answer) randomly chosen from a population in which
60\% of students know the correct answer.
(Use the {\sf sample()} function as in the previous question to generate
a sample from a discrete distribution. \newline
Count the number of one's in your sample. (These know the answer and will
answer correctly.) Use the number of zeros as the sample size for generating
the random sample in the next step.
\item Now generate a random sample for the students who did not know the
answer. Since these students select the answers 1 to 4 at random they have
a probability of 25\% of getting the correct answer.
That is to sample 1 and 0 with the probabilities 0.25 and 0.75 respectively.\newline
Count the number of 1's in the random sample (the number of correct guesses)
\item Find the total number of correct answers obtained (`Know' plus `guessed'!)
to get your estimate.
\item How did it compare with the theoretical answer?
\end{enumerate}



\end{exercises}

\makeanswers

\endinput


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%unused qns


\item
Efron's dice were invented by Bradley Efron. The dice consist of
four cubes as shown in Figure~\ref{FG:efronsdice}.

\begin{enumerate}
   \item
   Compute
   $\Pr{A\mbox{\ beats\ }B}$,
   $\Pr{B\mbox{\ beats\ }C}$ and
   $\Pr{C\mbox{\ beats\ }D}$.
   How do you think $A$ and $C$ would compare?
   ((\Hint: Fill out a table similar to ****************************?

   \item
   Compute
   $\Pr{D\mbox{\ beats\ }A}$
   and compare to your previous answer.
   Table~\ref{TB:efrondicetable}.
\end{enumerate}


\begin{figure}[ht]
\begin{center}
   \setlength{\unitlength}{0.7mm}
   \begin{picture}(150,50)(0,-10)
      % Dice faces, unrolled
      \thicklines
      \multiput(0,20)(40,0){4}{\line(1,0){30}}
      \multiput(0,30)(40,0){4}{\line(1,0){30}}
      \multiput(10,0)(40,0){4}{\line(0,1){40}}
      \multiput(20,0)(40,0){4}{\line(0,1){40}}
      \multiput(10,0)(40,0){4}{\line(1,0){10}}
      \multiput(10,40)(40,0){4}{\line(1,0){10}}
      \multiput(0,20)(40,0){4}{\line(0,1){10}}
      \multiput(30,20)(40,0){4}{\line(0,1){10}}
      \multiput(10,10)(40,0){4}{\line(1,0){10}}
      % Text--numbers of the dice
      % Die A
      \put(5,25){\makebox(0,0){0}}
      \multiput(15,5)(0,10){4}{\makebox(0,0){4}}
      \put(25,25){\makebox(0,0)0}
      % Die B
      \multiput(55,5)(0,10){4}{\makebox(0,0){3}}
      \multiput(45,25)(20,0){2}{\makebox(0,0){3}}
      % Die C
      \multiput(95,5)(0,10){4}{\makebox(0,0)2}
      \multiput(85,25)(20,0){2}{\makebox(0,0)6}
      % Die D
      \multiput(125,25)(10,0){3}{\makebox(0,0){1}}
      \multiput(135,5)(0,10){2}{\makebox(0,0){5}}
      \put(135,35){\makebox(0,0)5}
      % Labels under dice
      \put(15,-10){\makebox(0,0){Die~$A$}}
      \put(55,-10){\makebox(0,0){Die~$B$}}
      \put(95,-10){\makebox(0,0){Die~$C$}}
      \put(135,-10){\makebox(0,0){Die~$D$}}
   \end{picture}
   \caption{Efron's Dice.  Imagine folding the outlines into a cube.}
   \label{FG:efronsdice}
   \end{center}
\end{figure}


\begin{table}
   \begin{center}
      \newcommand{\win}{$\surd$}
      \newcommand{\loss}{$\times$}
      \begin{tabular}{cc|cccccc}
         &\textbf{Die A}&\multicolumn{6}{c}{\textbf{Die B}} \\
         & & 3 & 3 & 3 & 3 & 3 & 3
         \botnewrule \\
         \hline
         \topnewrule
         &4 & \win & \win & \win & \win & \win & \win \\
         &4 & \win & \win & \win & \win & \win & \win \\
         &4 & \win & \win & \win & \win & \win & \win \\
         &4 & \win & \win & \win & \win & \win & \win \\
         &0 & \loss & \loss & \loss & \loss & \loss & \loss \\
         &0 & \loss & \loss & \loss & \loss & \loss & \loss
      \end{tabular}
      \caption{Determining probabilities of winning for Efron's dice:
      When does Die~A beat Die~B?}
      \label{TB:efrondicetable}
   \end{center}
\end{table}

\begin{answer}
$A$ beats $B$; $B$ beats $C$; $C$ beats $D$ and (surprisingly?)
$D$ beats $A$ all with probability $1/3$.
\end{answer}


\item
While the probabilities for cubic dice are relatively
straightforward, it is not clear how to compute probabilities when
the dice have the dimension $1\times 1\times r$; see
Figure~\ref{FG:prism}. For various values of~$r$, numerous rolls
were conducted and the results shown in Table~\ref{TB:diceresults}
(see Dunn~\cite{BIB:Dunn:dice}).

\begin{enumerate}
   \item
   Can you use the classical approach to probability to
   determine $\Pr{\epsdice{6}}$ for various~$r$?
   Why or why not?

   \item
   Estimate $\Pr{\epsdice{6}}$ for each value of~$r$
   using the empirical approach.

   \item
   Plot a graph of~$r$ against $\Pr{\epsdice{6}}$.

   \item
   What value do you expect for $\Pr{\epsdice{6}}$ when $r=1$.
   Why?

\end{enumerate}

\begin{answer}
No (in general)---no equally-likely events are defined; (no
answers); (no answers); expect $\Pr{\epsdice{6}}=1/3$.
\end{answer}


\begin{figure}[ht]
   \begin{center}
   \setlength{\unitlength}{0.3mm}
      \begin{picture}(100,100)(25,-15)
         \thicklines
         % Face
         \put(0,0){\line(1,0){70}}
         \put(70,0){\line(0,1){70}}
         \put(70,70){\line(-1,0){70}}
         \put(0,70){\line(0,-1){70}}
         % Six marking
         \put(15,20){\circle*{15}}
         \put(35,20){\circle*{15}}
         \put(55,20){\circle*{15}}
         \put(15,50){\circle*{15}}
         \put(35,50){\circle*{15}}
         \put(55,50){\circle*{15}}
         % 3rd dimension
         \put(70,0){\line(3,1){60}}
         \put(0,70){\line(3,1){60}}
         \put(70,70){\line(3,1){60}}
         \put(60,90){\line(1,0){70}}
         \put(130,90){\line(0,-1){70}}
         % Lines for arrows
         \thinlines
         \put(0,-5){\line(0,-1){10}}
         \put(70,-5){\line(0,-1){10}}
         \put(-5,0){\line(-1,0){10}}
         \put(-5,70){\line(-1,0){10}}
         \put(130,15){\line(0,-1){10}}
         % Arrows
         \put(30,-10){\vector(-1,0){30}}
         \put(40,-10){\vector(1,0){30}}
         \put(-10,25){\vector(0,-1){25}}
         \put(-10,45){\vector(0,1){25}}
         \put(95,-1){\vector(-3,-1){25}}
         \put(105,1){\vector(3,1){25}}
         % Text
         \put(35,-10){\makebox(0,0){$1$}}
         \put(-10,35){\makebox(0,0){$1$}}
         \put(100,0){\makebox(0,0){$r$}}
         % Back end
         \put(160,65){\vector(-3,-1){30}}
         \put(160,65){\makebox(0,0)[l]{Six also}}
      \end{picture}
      \caption{A non-cubic die.
         Notice that the {{6}} is on \emph{both} $1\times1$ faces on the die.}
      \label{FG:prism}
   \end{center}
\end{figure}


\begin{table}
\begin{center}
\begin{tabular}{ccccc}
\textbf{Ratio}, $r$ & \textbf{Sixes} && \textbf{Ratio}, $r$ &
\textbf{Sixes}
\botnewrule\\
\hline \topnewrule
$   0.25    $ & $   744 $ && $ 1.10 $ & $190$ \\
$   0.50    $ & $   609 $ && $ 1.15 $ & $180 $ \\
$   0.75    $ & $   467 $ && $ 1.25 $ & $137 $ \\
$   0.85    $ & $   412 $ && $ 1.50 $ & $47$ \\
$   0.90    $ & $   355 $ && $ 1.75 $ & $26 $ \\
$   1.00    $ & $   253 $ && $ 2.00 $ & $14 $ \\
\end{tabular}
\caption{The results of throwing $1\times1\times r$ dice. Both
$1\times1$ faces were marked with a~$6$. Each die was rolled $760$
times.} \label{TB:diceresults}
\end{center}
\end{table}

\item
A common way of gathering information is to use door-to-door
surveys. However, because such a procedure uses face to face
questioning, people are often less inclined to answer some
questions truthfully; for example ``Have you ever lied on your tax
return forms?''.  One method of dealing with such a problem is
described in this question.

To obtain information, the surveyor is given a pile of $N$ cards,
of which $M$ have the statement ``I have lied in my tax return
forms''. Half of the remaining $N-M$ cards contain a simple
statement, such as ``I am a male'', and the other half contain the
statement ``I am female''.  The surveyor then shows the cards to
the person being interviewed so that they can see there are three
different questions that could be chosen.  The person then
randomly selects a card so that the surveyor does not see which
card the person has chosen.  The person being interviewed then
answers the question with a `Yes' or a `No'.  Since the
interviewer does not know what question has been answered, people
are more likely to answer truthfully.

\begin{enumerate}
\item Suppose an interviewer has a pile of 100 cards, and 40
contain the statement ``I have lied in my tax return forms''.
Half of the remaining cards contain the statement ``I am male'',
and the other half contain the statement ``I am female''.  The
interviewer interviews 500 people, 203 of whom are male, and
receives a total of 158 `Yes' responses.
\begin{enumerate}
\item Draw a tree diagram to illustrate the experiment that will
be useful
      in determining the proportion of people who lie on their tax return
      forms, $p$.
\item Estimate the proportion of people who have lied in their
      tax return forms.
\end{enumerate}

\item Construct a formula for estimating the proportion of people
who have lied on their tax return forms from a sample of size $n$
of whom proportion $y$ respond `Yes'.  Assume that a fraction $s$
of the cards contain the statement ``I have lied in my tax return
forms'', and that a proportion $m$ of the sample were male.

\item Discuss the \emph{practical} (not statistical) consequences
if 98 of the 100 cards contained the statement ``I have lied in my
tax return forms''.

\item Discuss the \emph{practical} (not statistical) consequences
if $M$ cards contained the statement ``I have lied in my tax
return forms'' and the remaining $N-M$ contain the statement ``I
am male''.

\end{enumerate}

\item
In less than ten sentences, discuss the following quote from
Monique Lloyd:
\begin{quote}
   The lottery:  A tax on people who flunked math.
\end{quote}
(\Note: ``flunked math'' means ``failed mathematics''.)

\item
The success of mammograms has been well documented.  Even though
mammograms are generally conducted on women over 40, breast cancer
is not unknown in women under 40. In this question, we examine
mammograms and breast cancer for women under 40. We take the data
(United States data only) from the articles ``Early Breast Cancer
Detection With Screening Mammography'' by Ella H. Wong, M.D. from
the Department of Radiology,  Sansum Medical Clinic and ``Mayo
doctors downplay mammogram false alarms'' by Gordon Slovut,
Minneapolis--St Paul Star Tribune (April 21, 1998).

Suppose breast cancer in women under 40 has a probability of 1\%
of occurring.  If the mammogram detects a cancer, there is a
chance of a false positive.  This probability is estimated to be
about 9\% by Mary Barton, a Harvard researcher. In addition, the
chance of a false negative is much smaller, possibly about 0.1\%.




\item In the game of Scrabble\copyright, players make words out of
a pool of 100 letter tiles containing the~$26$ letters from the
English alphabet (plus two blank tiles). The distribution of the
letters (in alphabetical order going down columns) is given in
Table~\ref{TB:prob:EG:scrabble}.

\begin{table}
\begin{center}
   \begin{tabular}{cc|cc|cc}
Letter & Number  & Letter & Number & Letter & Number \\
\botnewrule       & of tiles & & of tiles & & of tiles \\
\hline \topnewrule
      A & 9 & J & 1 & S & 4 \\
      B & 2 & K & 1 & T & 6  \\
      C & 2 & L & 4 & U & 4 \\
      D & 4 & M & 2 & V & 2 \\
      E & 12& N & 6 & W & 2 \\
      F & 2 & O & 8 & X & 1 \\
      G & 3 & P & 2 & Y & 2 \\
      H & 2 & Q & 1 & Z & 1 \\
      I & 9 & R & 6 & Blanks & 2 \\
   \end{tabular}
   \caption{The distribution of the letters in a game of Scrabble\copyright.}
\label{TB:prob:EG:scrabble}
\end{center}
\end{table}

\noindent (The blank tiles can be used in general play to
represent any letter.) Two players---Khan and Brown---agree to
play a game. To decide who starts, both players draw a tile from
the pool of letters. The player with the letter closest to the
start of the alphabet begins the game. If a player draws a blank
tile, the blank is returned to the pool of letters, the tiles are
mixed, and then another tile is drawn. Khan draws the first tile.
After Khan has drawn her (non-blank) tile, then Brown selects his
(non-blank) tile.

\begin{enumerate}
   \item
   Determine the probability that Khan will eventually draw the letter \framebox{E}.
   (Don't forget to allow for the possibility that Khan may
   draw blank tiles.)

   \item
   If Khan draws an \framebox{E}, determine the probability that Brown will
   go first in the actual game;
   that is,
   eventually draw an \framebox{A}\,, \framebox{B}\,, \framebox{C} or \framebox{D}.
   (Assume that Khan keeps his tile.
   Don't forget to allow for the possibility that Brown may
   draw blank tiles.)

\end{enumerate}


\item Give three examples of pairs of events that are mutually
exclusive.

\item Give three examples of pairs of events that are independent.


\item A government report into housing obtained (among other
things) results about education level and type of housing. The
results are tabulated below.

\begin{center}
\begin{tabular}{lcccr}
   & \multicolumn{3}{c}{\textbf{Highest Level of Education}} & \botnewrule\\
   \cline{2-4}
   \topnewrule & Less than & Year 12 & Tertiary & \\
    \textbf{Housing}   & Year 12 $(A_1)$ & $(A_2)$ & $(A_3)$ & \botnewrule \\
   \hline
   \topnewrule House ($B_1$) & 3 & 10 & 7 & 20 \\
   Flat/Unit ($B_2$) & 5 & 5 & 0 & 10 \\
   Living with Family ($B_3$) & 3 & 6 & 4 & 13 \\
   Caravan Parks/Other ($B_4$) & 1 & 4 & 2 & 7 \botnewrule\\
   \hline
   \topnewrule & 12 & 25 & 13 & 50 \\
\end{tabular}
\end{center}
\begin{enumerate}
   \item Calculate the following probabilities.
   \begin{enumerate}
      \item $\Pr{A_3\cap B_1}$
      \item $\Pr{A_2 \cup B_3}$
      \item $\Pr{A_3\cup A_1}$
      \item $\Pr{A_3 \cap B_4'}$
      \item $\Pr{\{A_1\cap A_2\} \cup B_1}$
      \item $\Pr{A_1 \mid  B_2}$
   \end{enumerate}
   \item Explain the meaning of each of the probabilities
   found above. (That is, what do the symbols \emph{mean}?)
   \item  In an examination, the question is asked: ``Are the
   two events independent?''.
   \begin{enumerate}
      \item  What are the two events?
      \item  Here is the solution given by one student.
      \begin{quote}
         Two events are independent if $\Pr{A}\times\Pr{B} = \Pr{A\cap B}$.
         I'll consider the events $A_2$ and $B_1$.
         First, $\Pr{A_2} = 25/50 = 1/2$.
         Then, $\Pr{B_1} = 20/50 = 2/5$.
         So,
         $\Pr{A_2}\times\Pr{B_1} = 1/2 \times 2/5 = 1/5$.
         Then,
         $\Pr{A_2\cap B_1} = 10/50 = 1/5$.
         Since
         $\Pr{A_2}\times\Pr{B_1} = \Pr{A_2\cap B_1}$,
         the events
         ``Highest Level of Education'' and ``Housing Status''
         are independent.
      \end{quote}
      \begin{enumerate}
         \item Check the student's proof, and find the errors (if any).
         \item Is the student's conclusion correct?  If not,
         state why not.  If so, what does the conclusion mean?
      \end{enumerate}

   \end{enumerate}

\end{enumerate}



\item Explain, in simple, non-statistical words, why is it
sensible to claim that two events $A$ and $B$ are independent if
$\Pr{A\mid B} = \Pr{A}$.


\item
Consider an experiment that involves rolling a die and tossing a
coin. Define $E$ as the event ``the number \epsdice{3} is observed
on the die'' and the event $F$ refer to the event ``a head is
observed on the coin''. Write down in words the \emph{meaning} of
each of the following.  \emph{Do not} list the events in the
sample space or determine probabilities.
   \begin{enumerate}
      \item $E\cap \overline F$.
      \item $F\mid E$.
   \end{enumerate}




%\item
%
%In Figure~\ref{FG:garden}, there is a picture of a backyard showing the
%position of a garden.   There is no fence around the backyard, but
%there is a wall to the west.
%
%\begin{enumerate}
%\item
%Assuming that rain falls straight down (that is, $\alpha=90^\circ$),
%determine the probability that
%a rain drop that lands in the yard will land in the garden.
%\item-
%Rather than assume that the rain falls straight down, instead assume
%that the rain falls at an angle $\alpha$.  Rain is said to come from
%angle $\alpha$ if it is coming from the west at an angle of
%$\alpha$ degrees to the horizontal (see Figure~\ref{FG:garden}).
%Assume that the rain comes perfectly aligned in the east-west direction.
%Determine the probability that
%a rain drop that lands in the yard will land in the garden.
%
%\item
%Draw a graph of the probability that a rain drop that lands
%in the yard will land in the garden versus the angle $\alpha$, over the
%range $30^\circ<\alpha<150^\circ$.
%\item
%How does your answer to part~(b) change if the rain comes at an
%angle to the east-west direction as well?
%\end{enumerate}
%
%\begin{figure}[hb]
%   \begin{center}
%      \includegraphics[scale=0.7]{../pics/prob/garden.eps}
%      \caption{A garden in a backyard:
%               (a)~View of backyard from the top;
%               (b)~Cross-section view of backyard}
%      \label{FG:garden}
%   \end{center}
%\end{figure}
%
%\clearpage
%

%\item Given the sample space $\lbrace 1, 2, 3, 4, 5, 6, 7, 8\rbrace$
%where the probability of selecting each item is equally likely,
%the following events are defined:
%
%\begin{tabular}{llll}
%$A$:\quad& $\lbrace 1, 2, 3, 4\rbrace$;\quad
%$B$: & $\lbrace 1, 2, 5, 6\rbrace$; \\
%$C$: & $\lbrace 1, 2, 7, 8\rbrace$;\quad
%$D$: & $\lbrace 1, 3, 5, 7\rbrace$. \\
%\end{tabular}
%\begin{enumerate}
%   \item Determine if $A$, $B$ and $C$ are mutually independent or not.
%   \item Determine if $A$, $B$ and $D$ are mutually independent or not.
%\end{enumerate}
%


