# (PART) Theoretical foundations {-}

# Essentials of set theory {#ChapterSetTheory}


::: {.objectivesBox .objectives data-latex="{iconmonstr-target-4-240.png}"}
UpOn completion of this chapter you should be able to:

* define sets, and define elements of sets.
* define the cardinality of sets.
* perform basic set operations.
:::


## Introduction to sets {#Sets}

The basis of probability is working with *sets*, which we first define.
Informally, a *set* is a collection of 'things'.


::: {.definition #Sets name="Set"}
A *set*\index{Set} is a well-defined, unordered collection of distinct *elements*.\index{Elements of sets}
:::

In this definition:

* 'distinct' means that an element is either present or absent in the set; there is no concept of repeated elements in a set.
* 'unordered' means that the order of the elements in the set is irrelevant.
* 'well-defined' means that it must be clear whether any object belongs to the set or does not belong to the set.

Sets are usually denoted using a capital letter, such as $A$.
Sets can be finite, countably infinite, or uncountable; the distinctions are discussed further in Sect.\ \@ref(FiniteInfiniteSets).

For some sets, listing the individual elements is possible:
$$
   A = \{ a_1, a_2, a_3, a_4\},
$$
when we can write $a_2\in A$ to indicate that\ $a_2$ is an element of the set\ $A$.
However, listing the elements like this is not always possible (e.g., for uncountably infinite sets).,

The *elements*\index{Elements of sets} can be interpreted broadly, and can refer to letters, digits, names, suits in a pack of cards, numbers, words, animals, people, machine components, plants... or even other sets.


::: {.example #DefineSets name="Sets"}
Consider these two sets:
\begin{align*}
  A &= \{\text{Brisbane}, \text{Sydney}, \text{Hobart}, \text{Melbourne}, \text{Adelaide}, \text{Perth}\};\\
  B &= \{\text{Hobart}, \text{Melbourne}, \text{Perth, }\text{Brisbane}, \text{Sydney}, \text{Adelaide}\}.
\end{align*}
These two sets are *equal*, since the order of the elements does not matter: both sets contain six distinct elements ('Australian state capital cities').
We can write $A = B$.

In contrast, 
$$
  C = \{\text{Brisbane}, \text{Sydney}, \text{Hobart}, \text{Sydney}, \text{Hobart}, \text{Hobart} \}
$$
contains redundancy; the elements 'Sydney' and 'Hobart' are repeated.
By definition, sets do not contain repeated elements so\ $C$ is equivalent to writing
$$
  C = \{\text{Brisbane}, \text{Sydney}, \text{Hobart} \}.
$$.
:::


::: {.example #ExamplesOfSets name="Examples of sets"}
Define the set\ $M$ as:
$$
  M = \{\text{Echidnas}, \text{Platypuses}\},
$$
$M$ is the set of all known monotremes. 
Set\ $M$ has two elements.

We could define the set\ $E$ as
$$
   E = \{1, 3, 5, 7, 9\},
$$
the set of all odd digits. 
Set\ $E$ has five elements.

Then, we could define the set\ $W$ as:
$$
  W = \{ M, E\} = \{ (\text{Echidnas}, \text{Platypuses}), (1, 3, 5, 7, 9)\}.
$$
The set\ $W$ contains two elements: the sets\ $M$ and\ $E$.
:::


The symbol '$\in$' is used to denote that an element is a member of a set; the symbol '$\notin$' is used to denote that an element is *not* a member of a set.
If $a$ is an element of the set $A$, we write $a \in A$, and we say '$a$ is an element of set\ $A$'.


::: {.example #DefineSetsIn name="Elements of sets"}
For set\ $A$ in Example\ \@ref(exm:DefineSets), we can write
$$
  \text{Hobart} \in A
$$
to indicate that 'Hobart' is a member of set\ $A$.
Similarly, we can write 
$$
  \text{Canberra} \notin A
$$
to indicate that 'Canberra' is *not* a member of\ $A$.
:::


## Defining the elements of sets {#DescribingElementsOfSets}

Elements\index{Elements of sets!defining} of a set can be defined in several ways:

* *Enumerating*: listing all the individual elements.
* *Description*: specifying a property that all elements share.
* *Rule*: defining a precise condition that determines set membership.

Regardless of the method used, any given object must be clearly a member of the set or not a member of the set.


::: {.example #DefineSetsElements name="Elements in a set"}
Set\ $A$ in Example\ \@ref(exm:DefineSets) lists its elements explicitly.
Alternatively, $A$ could be defined by a *description*, by stating 'the set of capital cities in Australian states'.

'Canberra' is *not* an element of\ $A$, since Canberra is not an Australia state capital.
:::


::: {.example #DefiningSets2 name="Defining elements of a set"}
Define the set $L$ as 'all books at the local library with more than two authors'.
Listing all elements is impractical, but the rule clearly determines membership: any book either satisfies the condition (and belongs to\ $L$) or does not.
:::


## Specific sets {#SpecialSets}

Some sets are standard in mathematics and probability, and have special names or symbols.


:::{.definition #EmptySet name="Empty set"}
The *null set*, or the *empty set*,\index{Empty set} has no elements, and is denoted by $\varnothing$.
:::


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
The *set* itself is denoted $\varnothing$, which contains zero elements.
:::


:::{.definition #UniversalSet name="Universal set"}
The *universal set*\index{Universal set} is the set of all elements under consideration in a specific context.
It is often denoted by $S$, $U$ or $\Omega$.
:::

In practice, a universal is almost always defined.
Theoretically, unrestricted universal sets lead to paradoxes such as *Russell's paradox*.\index{Russell's paradox}


::: {.tipBox .tip data-latex="{iconmonstr-info-6-240.png}"}
Russell's paradox,\index{Russell's paradox} discovered by Bertrand Russell in\ 1901, arises from considering the set\ $R$ of all sets that are *not* members of themselves. 
Then:

* If $R\in R$, then (by definition) $R\notin R$, which is a contradiction.
* If $R\notin R$, then (by definition) $R\in R$, which is a contradiction.

Both cases lead to a contradiction.
Modern set theory (such as ZFC)\index{ZFC} restricts sets formation to avoid such paradoxes.
:::


:::{.example #NullSet name="Null set"}
Consider rolling a standard six-sided die
The *universal set* is 
$$
  U = \{1, 2, 3, 4, 5, 6\}.
$$
The set of rolls greater than\ $4$ is
$$
  B = \{5, 6\}.
$$
The set of rolls greater than\ $10$ contains *no* elements, and so
$$
  C = \varnothing.
$$
We do not write $C = \{\varnothing\}$; the symbol $\varnothing$ itself represents a set (with no elements).
If we write 
$$
  D = \{\varnothing\},
$$
then set\ $D$ has one element, and that element is the empty set.
:::


Some commonly used sets of numbers include:

* *Natural* (or counting) numbers\index{Number sets!natural} $\mathbb{N}$: $\{1, 2, 3, \dots\}$.
* *Integers*\index{Number sets!integers} $\mathbb{Z}$: $\{\ldots, -2, 1, 0, 1, 2, 3, \dots\}$.
* *Rational*\index{Number sets!rationals} numbers $\mathbb{Q}$.
* *Real*\index{Number sets!reals} numbers $\mathbb{R}$.
* *Positive real*\index{Number sets!positive reals} numbers $\mathbb{R}_{+}$.
* *Negative real*\index{Number sets!negative reals} numbers $\mathbb{R}_{-}$.
* *Complex*\index{Number sets!complex} numbers $\mathbb{C}$.

(Some authors include\ $0$ in the set of natural numbers.)


## Types of sets and cardinality {#FiniteInfiniteSets}

Sets can be:\index{Cardinality}

* *finite*, where elements can be counted; the number of elements\ $n$ is finite.
* *countable infinite*: elements could be listed in a sequence, but there are infinitely many.
* *uncountably infinite*: elements cannot all be listed in a sequence; there are 'more than infinitely many' elements.

For finite or countably infinite sets, each element is called an *element* or a *sample point*.


### Finite sets {#FiniteSets}

If\index{Finite sets} the number of elements in a set can be counted, the set is called a *finite* set.
A finite set has exactly\ $n$ distinct elements for some non-negative integer\ $n$.

The two sets\ $A$ and\ $B$ in Example\ \@ref(exm:DefineSets) are *finite* sets, as they contain a finite number of elements ($n = 6$).


::: {.example #DefiningSets3 name="Defining elements of a set"} 
Define set $A$ as
$$
   A = \left\{ \text{Countries whose names begin and end with the letter A}\right\}.
$$
The elements of $A$ include, for example, 'Austria', 'Antigua and Barbuda' and 'Albania'.
$A$ contains a *finite* number of elements.
:::


::: {.example #DefiningSets name="Defining elements of a finite sets"}
Define the set $S$ as the 'set of suits in a standard pack of cards'.
Then, $S = \{\clubsuit, \spadesuit, \heartsuit, \diamondsuit\}$.
We can write, for example, that $\spadesuit \in S$.

The elements of the set are listed; the set has four elements.
:::



::: {.example #DefiningSets4 name="Defining elements of a finite sets"}
Define the set\ $O$ as the 'set of odd numbers between\ $0$ and\ $100$'.
Alternatively, the elements of the set can defined by giving the pattern:
$$
  O = \{1, 3, 5, 7, 9, \dots 95, 97, 99\}
$$
Set\ $O$ has $50$\ elements.
:::


::: {.example #FiniteSets name="Finite sets"}
Define set $B$ as 'the even numbers after rolling a single die'; then,
`r if (knitr::is_html_output()) '<!--'`
 $$
  B = \big\{
   \largedice{2},
   \largedice{4},
   \largedice{6} \big\}.
$$
`r if (knitr::is_html_output()) '-->'`
`r if (knitr::is_latex_output()) '<!--'`
<em>B</em> = {<span class="larger-die">&#9857;</span>, <span class="larger-die">&#9859;</span>, <span class="larger-die">&#9861;</span>}
`r if (knitr::is_latex_output()) '-->'`

The set\ $B$ is *finite*, with three elements.
:::


### Countably infinite sets {#CountablyInfiniteSets}

::: {.definition #CountablyInfiniteSets name="Countably infinite set"}
A set is called *countably infinite* if its elements can be put into a one-to-one correspondence with the natural numbers 
$$
  N = \{1, 2, 3, \dots\}.
$$
Each element appears exactly once in the sequence.
:::


::: {.example #DefiningSets5 name="Countably infinite sets"}
Define $S$ as the number of rolls of a die needed to roll a
`r if (knitr::is_latex_output()) {
   '\\largedice{6}'
} else {
   '<span class="larger-die">&#9861;</span>'
}`.
The number of rolls needed could be $1$ or $2$ or $3$... in theory, no upper limit exists.
That is, the set\ $S$ is
$$
  S = \{1, 2, 3, 4, \dots\}
$$
Set\ $S$ is *countably infinite*.
:::


::: {.example #DefiningSets6 name="Countably infinite sets"}
Define the set $O$ as the 'set of odd numbers'.
The elements of the set can be described (odd numbers'), or by giving the pattern:
$$
  O = \{1, 3, 5, 7, 9, \dots\}
$$
Set\ $O$ is *countably infinite*.
:::


### Uncountably infinite sets {#InfiniteSets}

:::{.definition #UncountablyInfiniteSets name="Uncountably infinite sets"}
A set is *uncountably infinite* if its elements cannot be listed in a sequence without missing some elements.
:::


::: {.example #UncountablyInfiniteSets name="Uncountably infinite sets"}
The set $R$ refers to the positive real numbers, which is an *uncountably infinite* set.
The elements of $R$ cannot be listed without missing some elements.
For instance, an infinite number of elements exists between the elements\ $0$ and\ $0.0001$.

Set\ $R$ is *uncountably infinite*.
:::

An example of a uncountably infinite set is the positive real numbers $R = {x \in \mathbb{R} : x > 0}$.


:::{.example #HeightsOfPeople name="Uncountably infinite set"}
Consider the heights of people.
Between the two heights $173\cms$ and $174\cms$, an infinite number of heights exist:
$173.1\cms$, $173.01\cms$, $173.001\cms$, $173.0001\cms$, and so on.

The set of heights in uncountably infinite.

Nonetheless, for convenience heights are rounded to (say) the nearest whole number of centimetres, or perhaps to one decimal place of a centimetre.
:::


Since the elements of uncountably infinite sets cannot be listed---even with a pattern---the elements of uncountably infinite sets are often concisely defined using *set-builder notation*.\index{Set-builder notation}
(Set-builder notation can be used for any sets, but are especially useful for uncountably infinite sets.)

Set-builder notation usually has the format:
\begin{align*}
  A &= \{ x \mid \text{some condition for $x$ to belong to the set}\}, \text{or}\\
  A &= \{ x :    \text{some condition for $x$ to belong to the set}\}.
\end{align*}
In this notation, $\{\dots\}$ means that\ $A$ is a set; $x$ is a variable representing the elements of the set\ $A$; the colon or vertical bar is read as 'such that', and is followed by the condition for $x$ to belong to the set\ $A$.
Often, the notation to the left of the 'such that' defines the universal set for the context.


:::{.example #SetBuilderNotation1 name="Set-builder notation"}
Consider the set defined by
$$
   A = \{x \in \mathbb{Z} \mid x > 9\}.
$$
This is read as '$A$ is the set of all integers\ $x$, such that the elements\ $x$ are larger than\ $9$'.
:::
:::{.example #SetBuilderNotation name="Set-builder notation"}
Consider the set defined by
\begin{align*}
   F = &\{s \in \text{Names of students in a certain course} \mid{}\\
       &\qquad \text{students whose mark is less than $50\%$}\}.
\end{align*}
This is read as '$F$ is the set of all student names in a certain course\ $s$, such that the elements\ $s$ are the names of all students in that course with a mark less than $50$%'.
That is, $F$ represents the set of all students in that specific course who scored less than\ $50$%.
:::


:::{.example #SetBuilderNotation2 name="Set-builder notation"}
Consider the set defined by
$$
   B = \{x \in \text{Animal species}: \text{$x$ is an animal species with two legs}\}
$$
This is read as '$B$ is the set of all animals species\ $x$, such that the elements\ $x$ are all animal species with two legs'.
We could also have written
$$
   B = \{x : \text{$x$ is an animal species with two legs}\}
$$
:::


:::{.example #SetBuilderNotation3 name="Set-builder notation"}
Consider the set defined by
$$
    V = \{x \in \text{letters in the alphabet} \mid \text{$x$ is a vowel}\}.
$$
This is read as '$V$ is the set of all letters \ $x$, such that the elements\ $x$ are all vowels'.
:::


:::{.example #SetBuilderNotation4 name="Set-builder notation"}
Consider the set defined by
$$
   P = \{y \in \mathbb{Z} : \text{$y$ is a two-digit integer}\}.
$$
This is read as '$P$ is the set of integers\ $y$, such that the elements\ $y$ are all two-digit integers'.
:::




:::{.example #CricketBallDefine name="Throwing a cricket ball"}
Consider throwing a cricket ball.
Let\ $D$ be the set of all possible distances (in metres) that the ball could be thrown; then
$$
   D = \{ d \in \mathbb{R} \mid d \ge 0 \}.
$$
The cardinality of $D$ is *uncountably infinite*.
:::


:::{.example #UncountablyInfiniteSets2 name="Uncountably infinite sets"}
Define $C$ as the set of 'all possible distances travelled by cars in Europe over one year'.
Set\ $C$ is *uncountably infinite*, and can be defined as
$$
  C = \{c \in \mathbb{R} \mid c \ge 0 \}.
$$
:::


### Cardinality {#Cardinality}

*Cardinality*\index{Cardinality} denotes the number of elements in a set.
Cardinality of a set\ $A$ is denoted\ $|A|$.

Cardinality of a finite set is the number of elements in the set.
Countably infinite sets and uncountably infinite sets both have *infinite* cardinality, but different infinite cardinalities.
Countably infinite sets have cardinality $\aleph_0$, where $\aleph_0$ ('aleph-null') is the smallest infinite cardinality.
Uncountably infinite sets have cardinality $\mathfrak{c}$ where $\mathfrak{c}$ is the 'cardinality of the continuum'.
Cantor showed that $\mathfrak{c} = 2^{\aleph_{0}} > \aleph_0$.





::: {.example #SizeOfSets name="Cardinality of finite sets"}
The set $S$, defined in Example\ \@ref(exm:DefiningSets), is
$$
  S = \{\clubsuit, \spadesuit, \heartsuit, \diamondsuit\}.
$$
The elements of the set are listed; the set has four elements.
That is, $|S| = 4$.
:::



## Operations on sets {#RelationshipsBetweenSets}

Sets can be combined in various ways, which sometimes makes defining sets easier too.
Various operations are defined for sets.\index{Set operations}


:::{.definition #IntersectionUnionSubset name="Set operations"}
Consider two sets\ $A$ and\ $B$, and a universal set\ $S$.
The following operations are defined.

* **Intersection**:\index{Set operations!intersection}  
  The *intersection* of of sets $A$ and $B$, written as $A\cap B$, is the set of elements in *both* $A$ and $B$: $A\cap B = \{x \mid x\in A \text{ and } x \in B\}$.
  We often say $A$ **and** $B$.
* **Union**:\index{Set operations!union}  
  The *union* of sets $A$ and $B$, written as $A\cup B$, is the combined set of all the elements in either $A$ or $B$, or in both: $A\cup B = \{x \mid x\in A \text{ or } x \in B\}$.
  We often say $A$ **or** $B$.
* **Complement**:\index{Set operations!complement}  
  The *complement* of set\ $A$, written\ $A^c$, is all the elements of\ $S$ that are *not* in set\ $A$: $A^c = \{x \mid x\notin A\}$.
* **Difference**, **set difference** or **set subtraction**:\index{Set operations!set difference}  
  The *difference* between sets\ $A$ and set\ $B$, written $A\setminus B$ or $A - B$, is all the elements in\ $A$ that are *not* in set\ $B$: $A\setminus B = \{x \mid x\in A \text{ and } x \notin B\}$.
* **Cartesian product**:\index{Set operations!Cartesian product}  
  The Cartesian product of\ $A$ and\ $B$, written $A\times B$, is the combination of all elements of\ $A$ with all elements of\ $B$.
  For example,  if $A = \{ 1, 2\}$ and $B = \{10, 20\}$, then $A\times B = \{(1, 10), (1, 20), (2, 10), (2, 20) \}$.
* **Subset**:\index{Set operations!subset}    
  $A$ is a *subset* of\ $B$, written $A\subseteq B$, if every elements of\ $A$ is also an element of $B$. 
  (Set $B$ may also have elements that are not in\ $A$.)
  This definition requires that $A$ is equal to or smaller than\ $B$.
* **Proper subset**:\index{Set operations!proper subset}    
   $A$ is a *proper subset* of\ $B$, written $A\subset B$, if every elements of\ $A$ is also an element of $B$, but $B$ has at least one element that is *not* in\ $A$.
:::

Two sets whose intersection is the empty set (i.e., if $A\cap B = \varnothing$ for two sets\ $A$ and\ $B$) the sets are called *disjoint sets*.\index{Sets!disjoint}

The visual representation in Fig.\ \@ref(fig:IntersectionUnionSubset) may help clarify some of these definitions.
These diagrams are called *Venn diagrams*.\index{Venn diagrams}


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
Be aware of variations in notation!\index{Set operations!complement}
Other notation for the complement of an event $A$ (everything *not* in event $A$), which we denote as\ $A^c$, includes\ $\overline{A}$ or\ $A'$.
:::


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
Be careful using **and** and **or**.

In general conversation, **and** can mean the *intersection* or the *union*, depending on the context.
For example:

* 'Can the people who are female and aged under $20$ please stand?'
* 'Can the engineers and doctors please stand?' 

In the first, the context clearly means 'intersection': being 'female' (say, $F$) and being 'aged under\ $20$' (say, $T$) both apply to the same individual.
The sentence refers to $F\cap T$.

In the second, the context clearly means 'union': people who are engineers (say, $E$), or people who are doctors (say, $D$), or both.
The sentence refers to $E \cup D$.
In this sentence, **and** is combining two groups of people, not two *conditions* on the same people.

Similarly, in everyday speech **or** can be inclusive (one, the other, or both) or exclusive (one, or the other, but *not* both).
For example, 

* 'Do you have milk or sugar in your coffee?' uses an inclusive **or**: you may have sugar, you may have milk, but you may also have both.
* 'Answer this quiz using true or false' uses an exclusive **or**; it means to answer with true, or with false, but *not* both.

Context matters!
In the context of sets in mathematics and statistics, $A$ **and** $B$ always means 'in both sets', and $A$ **or** $B$ means 'in either, or both, sets'.
:::




```{r IntersectionUnionSubset, echo=FALSE, out.width='80%', fig.height = 8, fig.width=5.5, fig.align="center", fig.cap="Venn diagrams showing intersection, union, proper subset, complement and set differences. The rectangle represents the universal set, $S$."}
#par( mar = c(0.05, 0.05, 0.05, 0.05))
par(mfrow = c(3, 2) )

colourA <- rgb(0, 0, 255, 
               max = 255, 
               alpha = 125)
colourB <- rgb(0, 255, 0, 
               max = 255, 
               alpha = 125)
colourS <- rgb(100, 100, 100, 
               max = 255, 
               alpha = 125)


### INTERSECTION
plot( x = c(0, 1),
      y = c(0, 1),
      type = "n",
      ylab = "",
      xlab = "",
      main = expression( atop(
        paste("The ", bold("intersection"), " of sets ", italic(A), " and ", italic(B)),
        italic(A)~intersect(italic(B)) ) ), 
      axes = FALSE)


polygon( x = c(0, 0, 1, 1),
         y = c(0, 1, 1, 0))
plotrix::draw.ellipse(x = 0.4, 
                     y = 0.5,
                     a = 0.22,
                     b = 0.16,
                     col = colourA)
plotrix::draw.circle(x = 0.6,
                     y = 0.5,
                     radius = 0.19,
                     col = colourB)
text(x = 0.4,
     y = 0.65,
     labels = expression(italic(A)), 
     pos = 3)
text(x = 0.65,
     y = 0.75,
     labels = expression(italic(B)), 
     pos = 3)
arrows( x0 = 0.5, 
        y0 = 0.15,
        x1 = 0.5,
        y1 = 0.5,
        angle = 15,
        length = 0.15,
        lwd = 1)
text(x = 0.5,
     y = 0.18,
     pos = 1,
     labels = expression(italic(A) ~~ intersect() ~~ italic(B)))
mtext(text = expression(italic(S)),
      side = 1,
      adj = 0.25)


### UNION
colour1 <- plotColour1

plot( x = c(0, 1),
      y = c(0, 1),
      type = "n",
      xlab= "",
      ylab= "",
      main = expression(atop(
        paste("The ", bold(union), " of sets ", italic(A), " and ", italic(B)),
        italic(A)~union(italic(B)) ) ),
      axes = FALSE)


polygon( x = c(0, 0, 1, 1),
         y = c(0, 1, 1, 0))
plotrix::draw.ellipse(x = 0.4, 
                     y = 0.5,
                     a = 0.22,
                     b = 0.16,
                     col = colour1,
                     border = colour1)
plotrix::draw.circle(x = 0.6,
                     y = 0.5,
                     radius = 0.2,
                     col = colour1,
                     border = colour1)
text(x = 0.5,
     y = 0.5,
     labels = expression(italic(A) ~~ union() ~~ italic(B)))


text(x = 0.4,
     y = 0.75,
     labels = expression(italic(A)), 
     pos = 3)
text(x = 0.65,
     y = 0.75,
     labels = expression(italic(B)), 
     pos = 3)
mtext(text = expression(italic(S)),
      side = 1,
      adj = 0.25)



### SUBSET
plot( x = c(0, 1),
      y = c(0, 1),
      type = "n",
      xlab= "",
      ylab= "",
      main = expression(atop(
        paste( italic(A), " is a ", bold(proper~subset), " of ", italic(B)),
        italic(A)%subset%italic(B) ) ),
      axes = FALSE)


polygon( x = c(0, 0, 1, 1),
         y = c(0, 1, 1, 0))
plotrix::draw.ellipse(x = 0.52, 
                     y = 0.55,
                     a = 0.1,
                     b = 0.07,
                     col = colourA)
plotrix::draw.circle(x = 0.5,
                     y = 0.5,
                     radius = 0.2,
                     col = colourB)
text(x = 0.52,
     y = 0.65,
     labels = expression(italic(A)),
     pos = 1)
text(x = 0.6,
     y = 0.75,
     labels = expression(italic(B)),
     pos = 3)
mtext(text = expression(italic(S)),
      side = 1,
      adj = 0.25)




### COMPLEMENT
plot( x = c(0, 1),
      y = c(0, 1),
      type = "n",
      xlab= "",
      ylab= "",
      main = expression(atop(
        paste("The ", bold(complement), " of ", italic(A) ),
              italic(A)^italic(c)) ),
      axes = FALSE)


polygon( x = c(0, 0, 1, 1),
         y = c(0, 1, 1, 0),
         col = plotColour)
plotrix::draw.ellipse(x = 0.4, 
                     y = 0.5,
                     a = 0.22,
                     b = 0.16,
                     col = "white")
text(x = 0.4,
     y = 0.4,
     labels = expression(italic(A)))
text(x = 0.75,
     y = 0.55,
     labels = expression(italic(A)^c))
mtext(text = expression(italic(S)),
      side = 1,
      adj = 0.25)




### SET DIFFERENCE A - B
plot( x = c(0, 1),
      y = c(0, 1),
      type = "n",
      ylab = "",
      xlab = "",
      main = expression( atop(
        paste("The ", bold("difference"), " between two sets ", italic(A), " and ", italic(B)),
        italic(A)~"\u2216"~italic(B) ) ), 
      axes = FALSE)


polygon( x = c(0, 0, 1, 1),
         y = c(0, 1, 1, 0))
plotrix::draw.ellipse(x = 0.4, 
                     y = 0.5,
                     a = 0.22,
                     b = 0.16,
                     col = colourA)
plotrix::draw.circle(x = 0.6,
                     y = 0.5,
                     radius = 0.19,
                     col = "white")
# Now redraw left circle
plotrix::draw.ellipse(x = 0.4, 
                     y = 0.5,
                     a = 0.22,
                     b = 0.16,
                     col = NA)
text(x = 0.4,
     y = 0.65,
     labels = expression(italic(A)), 
     pos = 3)
text(x = 0.7,
     y = 0.75,
     labels = expression(italic(B)), 
     pos = 3)
arrows( x0 = 0.3, 
        y0 = 0.15,
        x1 = 0.3,
        y1 = 0.5,
        angle = 15,
        length = 0.10,
        lwd = 1)
text(x = 0.27,
     y = 0.16,
     pos = 1,
     labels = expression(italic(A)~"\u2216"~italic(B)))
mtext(text = expression(italic(S)),
      side = 1,
      adj = 0.25)




### SET DIFFERENCE: B - A
plot( x = c(0, 1),
      y = c(0, 1),
      type = "n",
      ylab = "",
      xlab = "",
      main = expression( atop(
        paste("The ", bold("difference"), " between two sets ", italic(B), " and ", italic(A)),
        italic(B)~"\u2216"~italic(A) ) ), 
      axes = FALSE)


polygon( x = c(0, 0, 1, 1),
         y = c(0, 1, 1, 0))
plotrix::draw.circle(x = 0.6,
                     y = 0.5,
                     radius = 0.19,
                     col = colourB)
plotrix::draw.ellipse(x = 0.4, 
                     y = 0.5,
                     a = 0.22,
                     b = 0.16,
                     col = "white")
# Now redraw right circle
plotrix::draw.circle(x = 0.6,
                     y = 0.5,
                     radius = 0.19,
                     col = NA)
text(x = 0.4,
     y = 0.65,
     labels = expression(italic(A)), 
     pos = 3)
text(x = 0.7,
     y = 0.75,
     labels = expression(italic(B)), 
     pos = 3)
arrows( x0 = 0.7, 
        y0 = 0.15,
        x1 = 0.7,
        y1 = 0.5,
        angle = 15,
        length = 0.10,
        lwd = 1)
text(x = 0.68,
     y = 0.16,
     pos = 1,
     labels = expression(italic(B)~"\u2216"~italic(A) ) )
mtext(text = expression(italic(S)),
      side = 1,
      adj = 0.25)



```




::: {.example #RelationshipBetweenSets5 name="Relationships between sets"}
Suppose we define these sets:
\begin{align*}
   C &= \{ 0, 1, 2, 3, 4, 5\};\\
   D &= \{ 4, 5, 6, 7\};\\
   E &= \{ 6 \}.
\end{align*}
Then:
\begin{align*}
   C \cap D &= \{ 4, 5\};                   &          C \cup D &= \{ 0, 1, 2, 3, 4, 5, 6, 7\};\\ 
   E &\in D;                                &          E &\notin C;\\ 
   C \cap E &= \varnothing;                   &          D \cap E &= D;\\
   C\setminus D &= \{0, 1, 2, 3\};          &          D\setminus C &= \{6, 7\}.
\end{align*}
:::



::: {.example #RelationshipBetweenSets6 name="Relationships between sets"}
Suppose we define these sets:
\begin{align*}
   U &= \{\text{All the digits}\} = \{ 0, 1, 2, 3, 4, 5, 6, 7, 8, 9\};\\
   B &= \{ 6, 7, 8, 9\};\\
   E &= \{ 2, 4, 6, 8\}.
\end{align*}
Here, $U$ is the *universal set*, the set of all digits,
Then, for example:
\begin{align*}
   U \cap B &= \{ 6, 7, 8, 9 \} = B;        &          B \cup E &= \{2, 4, 6, 7, 8, 9\};\\ 
   B &\in U;                                &          U &\notin B;\\ 
   B \cap E &= \{6, 8\};                    &          U \cup E &= U;\\
   B \setminus E &= \{7, 8, 9\};            &          E\setminus U &= \varnothing.
\end{align*}
:::


::: {.example #RelationshipBetweenSets4 name="Relationships between sets"}
Consider the set\ $A$ defined using set-builder notation as:\index{Set-builder notation}
$$
  A = \{ x  \in \mathbb{R} \mid \sqrt{x^2 - 4} \in \mathbb{R}\}.
$$
This means that the set\ $A$ comprises the real elements\ $x$ such that the values of $\sqrt{x^2 - 4}$ are real numbers; that is, such that $x^2 - 4 \ge 0$.
The values of\ $x$ can be defined as
$$
  A = \{ x \in\mathbb{R} \mid (-\infty, -2] \cup [2, \infty)\};
$$
set\ $A$ contains real numbers less than or equal to\ $-2$, and real numbers greater than or equal to\ $2$.

Notice the use of the brackets: using the square brackets\ '$[$' and\ '$]$' is standard for showing that the indicated value *is included* in the interval, where as the round brackets\ '$($' and\ '$)$' is standard for showing that the indicated value *is not included* in the interval.
:::


:::{.example #QuickUseOfOperations name="Set operations"}
We can define $\mathbb{N}_0 = \{ 0 \} \cup \mathbb{N}$.

We could write $\mathbb{Q} = \{\frac{m}{n}\mid m\in\mathbb{Z}, n \in \mathbb{Z}, n\ne 0\}$.

We can also write $\mathbb{Z} \in \mathbb{R}$, but $\mathbb{R} \notin \mathbb{Z}$.
:::


:::{.example #TwoDSets name="Sets can be two-dimensional"}
Sets do not have to be one-dimensional.\index{Sets!two-dimensional}
We could define the set $P$ as
$$
  P = \{(x, y)\in \mathbb{R}\times\mathbb{R}  \mid (0 < x < 1)\cap(0 < y < 1) \},
$$
where $\mathbb{R}\times\mathbb{R}$ means 'the ordered pairs $(x, y)$ where both\ $x$ and\ $y$ are real numbers'.
$P$ defines a unit square on the Cartesian axes.
:::




## Set algebra {#SetAlgebra}

Set algebra has many rules; we only provide some.\index{Set}
For sets\ $A$, $B$ and\ $C$ defined on the universal set\ $S$ these rules are defined.

:::::: {.cols data-latex=""}
::: {.col data-latex="{0.49\textwidth}"}


The *commutative* laws are:\index{Set algebra!commutative laws}

* $A\cup B = B \cup A$ 
* $A\cap B = B \cap A$.

The *associative laws* are:\index{Set algebra!associative laws}

* $A\cup(B\cup C) = (A\cup B)\cup C$.
* $A\cap(B\cap C) = (A\cap B)\cap C$.

The *distributive laws* are:\index{Set algebra!distributive laws}

* $A\cap (B\cup C) = (A\cap B)\cup (A \cap C)$.
* $A\cup (B\cap C) = (A\cup B)\cap (A \cup C)$.

*de Morgan's* law are:\index{Set algebra!de Morgan's laws} 

* $(A \cap B)^c = A^c \cup B^c$.
* $(A \cup B)^c = A^c \cap B^c$.

:::

::: {.col data-latex="{0.025\textwidth}"}
\ 
<!-- an empty Div (with a white space), serving as
a column separator -->
:::

::: {.col data-latex="{0.49\textwidth}"}


The *absorptions laws* are:\index{Set algebra!absorption laws}

* $A\cup (A\cap B) = A$.
* $A\cap (A\cup B) = A$.

The *dominance laws*:

* $A\cup S = S$
* $A\cap\varnothing = \varnothing$.

The *idempotent laws*:\index{Set algebra!idempotent laws}

* $A\cup A = A$.
* $A\cap A = A$.

:::

::::::

<!-- These are two separate lines, since a second level of indent begins with a dash, and look like a minus sign, which is confusing -->

:::{.proof}
We only give one example proof, for one of the commutative laws: prove $A\cup B = B \cup A$ (which may seem 'obvious').

To show this proposition, we need to show that 
\begin{align}
   (A \cup B) &\subseteq B\cup A\qquad\text{and also that}\\
   (B \cup A) &\subseteq A\cup B.
\end{align}
Consider the first statement, and consider an element\ $x$ such that $x\in (A\cup B)$.
This means that either:

* $x\in A$, in which case $x\in (B\cup A)$; or
* $x\in B$, in which case $x\in (B\cup A)$.

In either case, $x \in (B\cup A)$ and so $(A\cup B)\subseteq B\cup A$.

The proof of the second statement is similar, and hence $A\cup B = B \cup A$.
:::


These rules can be used to prove numerous other properties of sets.


:::{.example #UsingSetAlgebra name="Using set algebra"}
Consider the statement $A\cap (A^c \cup B)$.
To simplify, set algebra rules can be used:
\begin{align*}
     A\cap (A^c \cup B)
     &= (A\cap A^c) \cup (A\cap B)\qquad\text{(distributive law)}\\
     &= \varnothing \cup (A\cap B)\qquad\text{(by definition of $A^c$)}\\
     &= A\cap B.\qquad\text{(dominance law)}
\end{align*}
Thus, $A\cap (A^c \cup B) = A\cap B$.
:::


## Statistical computing with R {#StatisticalComputingWithR}


### Using R {#StatisticalComputingIntro}

Computers and computer packages are essential tools in the application of statistics to real problems. 
In this book, the statistical package **R** is used, and will be used to illustrate various concepts to help you understand the theory (Sect.\ \@ref(SimulationTheoryDemo)).

One way in which **R** can be used is to easily compute probabilities for specific distributions.
Also, **R** can be used to verify (not *prove*) theoretical results obtained.
To do this, a technique called *computer simulation* can be used. 

Simulation can also be used to solve problems for which it may be difficult (or impossible) to obtain a theoretical result. 
Sometimes these numerical solutions to intractable analytical problems is termed *Monte Carlo simulation*.

We introduce **R** by showing how to operate with sets; more general information about using **R** is given in Sect.\ \@ref(ShortRIntro).
Greater functionality in working with sets is provided by installing and using the **R** package `sets`.


### Simple example {#SetsInRExample1}

Consider rolling a standard, fair six-sided die.
Define\ $A$ as the set of numbers that can be rolled on the die, $B$ as the set of *even* numbers that can be rolled, and\ $C$ as the of numbers that can be rolled that are divisible by three.

This is a simple example, but **R** can be used to determine the elements in these sets, and the union and intersection of the sets.

First, we define the sets:

```{r}
DieRolls <- 1:6 # Set A
RollsEven <- seq(2, 6, by = 2)  # Set B
RollsDivisibleBy3 <- c(3, 6)    # Set C
RollsEven                       # Set B
RollsDivisibleBy3               # Set C
```

Then use **R** to find the intersection and union:

```{r}
intersect(RollsEven, RollsDivisibleBy3)
union(RollsEven, RollsDivisibleBy3)
sort( union(RollsEven, RollsDivisibleBy3) )
```

The function `setdiff()` find the difference between two sets:

```{r}
setdiff(RollsEven, RollsDivisibleBy3)  # B \ C
setdiff(RollsDivisibleBy3, RollsEven)  # C \ B
```

The function `is.element()` determines if a given element is contained in a given set:

```{r}
is.element(RollsDivisibleBy3, RollsEven)
is.element(RollsEven, RollsDivisibleBy3)
```




### More involved example {#SetsInRExample2}

The above examples was simple enough to complete manually.
However, the power of using computers is more obvious for more involved examples.

Consider the two sets:
\begin{align*}
  A &= \{\text{Countries whose names have `a' as the \emph{second} letter}\};\\
  B &= \{\text{Countries whose names end with `a'}\}.
\end{align*}

In **R**, the names of the countries can be found by loading the `countries` package (assuming it is installed):

```{r}
library("countries")
```

The list of country names are in the **R** vector `list_countries()`:

```{r}
head( list_countries() ) # Show the first five values only
```

First, we find the countries whose names have an `a` as the second letter (assuming it is always lower case); that is, we identify the elements in set\ $A$:

```{r}
# Find where, in the list of countries, 
# a country name has an `a` in position 2:
# (Assumes the letter is lower case)
Locate_a_Second <-  which( substr(list_countries(),
                                  start = 2, 
                                  stop = 2)
                           == "a")

# Now find the countries listed at those locations:
a_Second <- list_countries()[Locate_a_Second]

# How many countries are there whose names have an `a` as second letter?
length(a_Second)
```


Then, we find the countries whose names end with the letter `a` (assuming it is always lower case);  that is, we identify the elements in set\ $B$.
To do this, we first need to know how long the name of each country is:

```{r}
# Find the length of the name of each country in the list:
Length_Country_Names <- nchar(list_countries()) # Length of each country name
```

Then we can proceed similar to having `a` as the second letter of the name:

```{r}
# Find where, in the list of countries, a country name has an A in last position:
locate_Ends_With_a <- which( substr(list_countries(),
                                    start = Length_Country_Names,
                                    stop = Length_Country_Names)
                            == "a")

# Now find the countries listed at those locations:
Ends_With_a <- list_countries()[locate_Ends_With_a]

# How many countries are there whose names end with an `a`?
length(Ends_With_a)

```

Now we can find the countries whose names have an `a` as the second and as the last letters by finding the intersection between the two sets:

```{r}
a_Second_And_Last <- intersect(a_Second, 
                               Ends_With_a)

# How many countries are there whose names have
# `a` as second and last letters?
length(a_Second_And_Last)

# What are these countries?
a_Second_And_Last
```



```{r echo=FALSE, eval=FALSE}
### This is the R help from `?sets` in R:
(x <- c(sort(sample(1:20, 9)), NA))
(y <- c(sort(sample(3:23, 7)), NA))
union(x, y)
intersect(x, y)
setdiff(x, y)
setdiff(y, x)
setequal(x, y)

## True for all possible x & y :
setequal( union(x, y),
          c(setdiff(x, y), intersect(x, y), setdiff(y, x)))

is.element(x, y) # length 10
is.element(y, x) # length  8

## Factors:
x <- as.factor(c("A", "B", "A"))
y <- as.factor(c("B", "b"))
union(x, y)
intersect(x, y)
setdiff(x, y)
setdiff(y, x)
## (Note that union() and intersect() merge the levels.)
```



## Exercises {#SetsExercises}

Selected answers appear in Sect.\ \@ref(AnswersChapSetTheory).



:::{.exercise #AlgebraProof1}
Prove one of the idempotent laws in Sect.\ \@ref(SetAlgebra).
:::


:::{.exercise #AlgebraProof2}
Prove one of the dominance laws in Sect.\ \@ref(SetAlgebra).
:::


:::{.exercise #ImaginaryNumbers}
Consider the set of complex numbers $\mathbb{C}$ written in the form $a + bi$.

1. Use set-builder notation to define the set of complex numbers in terms of the set of real numbers\ $\mathbb{R}$.
2. Adapt your notation to define the set of purely imaginary numbers\ $\mathbb{I}$.
:::



:::{.exercise #SetsQuadratic}
Consider the real solutions to the general quadratic equation $y = ax^2 + bx + c = 0$ (where $a \ne 0$, since then the equation does not define a quadratic). 

Using set-builder notation, define: 

1. the universal set\ $U$, which defines all quadratics.
2. the set\ $R$ that contains all quadratics with two equal real roots.
3. the set\ $Z$ that contains all quadratics with no real roots.
4. the set\ $S$, in terms of\ $U$, $R$ and\ $z$, that contains all quadratics with exactly one real root.
:::


:::{.exercise #SetsLogFunctions}
Consider the equation $y = \log(x^2 - 1)$ for real values of\ $y$.
Using set-builder notation, define: 

1. the set of values of\ $x$ for which\ $y$ has real values.
2. the set of values of\ $x$ for which $y > 1$.
3. the set of values of\ $x$ for which\ $y$ has the value zero.
:::



:::{.exercise #SetsCards}
Define these sets:
\begin{align*}
  S &= \{\clubsuit, \spadesuit\};\\
  P &= \{ \text{Jack}, \text{Queen}, \text{King}, \text{Ace}\}\quad \text{(i.e., the picture cards)},
\end{align*}
where\ $U$ is the universal set of all cards in a pack.
Using set-builder notation, define:

1. the set\ $A$ that contains the black picture cards.
1. the set\ $B$ that contains the red picture cards.
1. the set\ $C$ that contains the red non-picture cards.
:::


:::{.exercise #Rbasics1}
Define a vector\ $A$ as the even numbers from\ $2$ to\ $20$, and vector\ $B$ as all integers from\ $1$ to\ $10$. 
Write **R** code to find:

1. the intersection of\ $A$ and\ $B$.
2. the elements in\ $A$ that are *not* in\ $B$.
:::


:::{.exercise #Rbasics2}
Define a vector\ $N$ as the odd numbers from\ $1$ to\ $51$, and vector\ $M$ as all triangular numbers between\ $1$ and\ $51$. (Triangular numbers take the form $n(n + 1)/2$ for integer\ $n > 0$.)
Write **R** code to find:

1. all the triangular numbers between\ $1$ and\ $51$ inclusive, and hence the elements of set\ $M$.
2. the intersection of\ $N$ and\ $M$.
3. the elements in\ $N - M$, and explain what this means.
4. the elements in\ $M - N$, and explain what this means.
:::


:::{.exercise #USStates}
In **R**, the names of all US states are contains in the built-in dataset `state.name`.
Using this dataset, use **R** to:

1. find the set of US states who names begin with\ `W`.
1. find the set of US states who names begin with `North`.
1. find the set of US states who names end in `a`.
1. find the set of US states who names end in `a`, **or** start with\ `W`.
1. find the set of US states who names end in `a`, **and** start with\ `W`.
1. find the set of US states who names end in `a`, **and** do not start with\ `W`.
:::

:::{.exercise #USStatesData}
In **R**, information about each US states in (mostly) 1977 is contained in the built-in dataset `state.x77`.
Using this dataset, use **R** to:

1. find the set of US states whose average life expectancy exceeds\ $70$.
1. find the set of US states whose area is less than\ $500\,000$ square miles.
1. find the set of US states whose illiteracy rate is greater than\ $2$%.
1. find the set of US states whose percent of high-school graduates is less than\ $50$%.
1. find the set of US states whose illiteracy rate is greater than\ $2$%, **and** whose percent of high-school graduates is less than\ $50$%.
1. find the set of US states whose illiteracy rate is greater than\ $2$%, **or** whose percent of high-school graduates is less than\ $50$%.

(To view the help for this data, type `?state.x77` in **R**.)
:::


:::{.exercise #SetBuilder1}
Using set-builder notation, define set\ $T$ as the values of\ $x$ for which $y = 1/x$ is a real value.
:::


:::{.exercise #SetBuilder2}
Using set-builder notation, define set\ $L$ as the values of\ $x$ for which $y = \tan x$ is a real value.
:::


:::{.exercise #SetBuilderReading1}
Explain what values are contained in set\ $G$, if it is defined as
$$
   G = \{ x\in\mathbb{R} \mid x^2 < 4\}
$$
What is the cardinality of set\ $G$?
:::


:::{.exercise #SetBuilderReading2}
Explain what values are contained in set\ $B$, if it is defined as
$$
   B = \{x\in\mathbb{R} \mid \lfloor x \rfloor = x\}.
$$
In this expression, $\lfloor x \rfloor$ is the *floor* function, that rounds the value of\ $x$ to the next lowest integer (for example, $\lfloor \pi \rfloor = 3$).

What is the cardinality of set\ $B$?
:::


:::{.exercise #SetAlgebra1}
Use set algebra to show that $A\setminus(A\cap B) = A\cap B^c$.
:::


:::{.exercise #SetAlgebra2}
Use set algebra to show that $(A\setminus B)^c = A^c\cup B^c$.
:::


:::{.exercise #SetAlgebra3}
Use set algebra to show that $(A\cup B) \cap (A\cup B^c) \cap B = A\cap B$.
:::

:::{.exercise #SetAlgebra4}
Use set algebra to show that $(A\cap B) \cup (A\cap B^c) = A$.
:::


:::{.exercise #TwoCoinsTosses}
Suppose the set $C$ is defined as
`r if (knitr::is_latex_output()) {
   '$C = \\{ \\Heads, \\Tails\\}$,'
} else {
   '<em>C</em> = {<span class="HTcircle">H</span>, <span class="HTcircle">T</span>}'
}`
the set of results from tossing a coin.
Define the set\ $D$ that shows the results of *two* tosses of a coin, as an ordered pair.
(See Example\ \@ref(exm:TwoDSets).)
:::



:::{.exercise #CardsInPack}
Define the following two sets:
$$
   S = \{\spadesuit, \clubsuit, \diamondsuit, \heartsuit\}
   \qquad\text{and}\qquad
   D = \{2, 3, 4, \dots, \text{Jack}, \text{Queen}, \text{King}, \text{Ace}\}.
$$
Use these two sets to define Set\ $C$ of 'all cards in a standard pack', as an ordered pair.
(See Example\ \@ref(exm:TwoDSets).)
:::


:::{.exercise #ThreeViruses}
A hospital tests patients for three respiratory viruses: Influence (the flu) ($F$), COVID ($C$), and RSV ($R$).
A patient is classified as 'High Risk' if they have at least two of these viruses simultaneously. 

Write the set expression for 'High Risk' patient.
:::


:::{.exercise #ParticleDetector}
Imagine a circular particle detector. 
Let\ $S$ be the entire area of the detector. 
Let\ $D_r$ be the set of points $(x, y)$ such that $x^2 + y^2 \leq r^2$ (i.e., a disk of radius $r$). 

If a particle is detected in the 'outer ring', between radius\ $5$ and radius\ $10$, express this region using $D_5$ and $D_{10}$ using set notation.
:::