# (PART) Theoretical foundations {-}

# Essentials of set theory {#ChapterSetTheory}


::: {.objectivesBox .objectives data-latex="{iconmonstr-target-4-240.png}"}
Upon completion of this chapter you should be able to:

* define sets, and define elements of sets.
* define the cardinality of sets.
* perform basic set operations.
:::


## Introduction to sets {#Sets}

The basis of probability is working with *sets*.
Informally, a *set* is simply a collection of objects.


::: {.definition #Sets name="Set"}
A *set*\index{Set} is a well-defined, unordered collection of distinct *elements*.\index{Elements of sets}
:::

In this definition, 'distinct' means that there is no concept of repeated elements in a set.
Also, notice that a set is an 'unordered' collection: the order of the elements in the set is irrelevant.
'Well-defined' means any object must clearly belong to the set, or not belong to the set.

Sets are usually denoted using a capital letter, such as $A$.
Sets can be finite, countably infinite, or uncountable; the distinctions are discussed further in Sect.\ \@ref(FiniteInfiniteSets).


::: {.example #DefineSets name="Sets"}
Consider these two sets:
\begin{align*}
  A &= \{\text{Brisbane}, \text{Sydney}, \text{Hobart}, \text{Melbourne}, \text{Adelaide}, \text{Perth}\};\\
  B &= \{\text{Hobart}, \text{Melbourne}, \text{Perth, }\text{Brisbane}, \text{Sydney}, \text{Adelaide}\}.
\end{align*}
These two sets are *equal*, since the order of the elements does not matter: both sets contain six distinct elements ('Australian state capital cities').
We can write $A = B$.

In contrast, 
$$
  C = \{\text{Brisbane}, \text{Sydney}, \text{Hobart}, \text{Sydney}, \text{Hobart}, \text{Hobart} \}
$$
contains redundancy; the elements 'Sydney' and 'Hobart' are repeated.
By definition, repeated elements in sets are redundant, so\ $C$ is equivalent to writing
$$
  C = \{\text{Brisbane}, \text{Sydney}, \text{Hobart} \}.
$$
:::


For some sets, listing the individual elements is possible, when the individual elements are denoted with lower-case letters:
$$
   A = \{ a_1, a_2, a_3, a_4\}.
$$

The symbol '$\in$' denotes that an element is a member of a set; the symbol '$\notin$' is used to denote that an element is *not* a member of a set.
If $a$ is an element of the set $A$, we write $a \in A$, and we say '$a$ is an element of set\ $A$'.
However, listing the elements like this is not always possible (e.g., for uncountably infinite sets).


::: {.example #DefineSetsIn name="Elements of sets"}
For set\ $A$ in Example\ \@ref(exm:DefineSets), we can write
$$
  \text{Hobart} \in A
$$
to indicate that 'Hobart' is a member of set\ $A$.
Similarly, we can write 
$$
  \text{Canberra} \notin A
$$
to indicate that 'Canberra' is *not* a member of\ $A$.
:::


Set *elements*\index{Elements of sets} can be interpreted broadly, and can refer to letters, digits, names, suits in a pack of cards, numbers, words, animals, people, machine components, plants... or even other sets.


::: {.example #ExamplesOfSets name="Examples of sets"}
Define the set\ $M$ as:
$$
  M = \{\text{Echidnas}, \text{Platypuses}\};
$$
$M$ is the set of all known monotremes. 
Set\ $M$ has two elements and, for example, $\text{Echindna}\in M$.

Also, define the set\ $D$ as
$$
   D = \{1, 3, 5, 7, 9\},
$$
the set of all odd digits. 
Set\ $D$ has five elements.

Then, we could define the set\ $W$ as:
$$
  W = \{ M, D\} = \{ (\text{Echidnas}, \text{Platypuses}), (1, 3, 5, 7, 9)\}.
$$
The set\ $W$ contains two elements: set\ $M$ and set\ $D$.
:::


## Defining the elements of sets {#DescribingElementsOfSets}

Elements\index{Elements of sets!defining} of a set can be defined in several ways:

* *Enumerating*: listing all the individual elements.
* *Description*: specifying a property that all elements share.
* *Using a rule*: defining a precise condition that determines set membership.

Regardless of the method used, any given object must be clearly a member of the set or not a member of the set.


::: {.example #DefineSetsElements name="Elements in a set"}
Set\ $A$ in Example\ \@ref(exm:DefineSets) lists its elements explicitly.
Alternatively, $A$ could be defined by a *description*, by stating 'the set of capital cities in Australian states'.

'Canberra' is *not* an element of\ $A$, since Canberra is not an Australia state capital.
:::


::: {.example #DefiningSets2 name="Defining elements of a set"}
Define the set $L$ as 'all books at the local library with more than two authors'.
Listing all elements is impractical, but the rule clearly determines membership: any book either satisfies the condition (and belongs to\ $L$) or does not.
:::


## Specific sets {#SpecialSets}

Some sets are standard in mathematics and probability, and have special names or symbols.


:::{.definition #EmptySet name="Empty set"}
The *null set*, or the *empty set*,\index{Empty set} has no elements, and is denoted by $\varnothing$.
:::


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
The *set* itself is denoted $\varnothing$, which contains zero elements.
:::


:::{.definition #UniversalSet name="Universal set"}
The *universal set*\index{Universal set} is the set of all elements under consideration in a specific context.
It is often denoted by $S$, $U$ or $\Omega$.
:::


:::{.example #universalSet name="Universal sets"}
The universal set is the set of all objects currently under discussion, so it depends on context.

If we are talking about people in a room, $U$ is the set of  all people in the room.
If we are talking about integers, $U$ is the set of all integers.
If we are talking about dealing cards, $U$ is the set of all cards in the pack. 

:::

In practice, a universal set is almost always defined.
Theoretically, unrestricted universal sets lead to paradoxes such as *Russell's paradox*.\index{Russell's paradox}


::: {.tipBox .tip data-latex="{iconmonstr-info-6-240.png}"}
Russell's paradox,\index{Russell's paradox} discovered by Bertrand Russell in\ 1901, arises from considering this set:
$$
   R = \{ \text{all sets that are not members of themselves} \}.
$$
Then:

* If $R\in R$, then (by definition) $R\notin R$, which is a contradiction.
* If $R\notin R$, then (by definition) $R\in R$, which is a contradiction.

Both cases lead to a contradiction.
Modern set theory (such as ZFC)\index{ZFC} restricts sets formation to avoid such paradoxes.
:::


:::{.example #NullSet name="Null set"}
Consider rolling a standard six-sided die and observing the uppermost face.
The *universal set* is 
$$
  U = \{1, 2, 3, 4, 5, 6\}.
$$
The set of rolls greater than\ $4$ is
$$
  B = \{5, 6\}.
$$
The set of rolls greater than\ $10$ contains *no* elements, and so
$$
  C = \varnothing.
$$
We do not write $C = \{\varnothing\}$; the symbol $\varnothing$ itself represents a set (with no elements).
Writing 
$$
  C = \{\varnothing\},
$$
defines\ $C$ as a set with one element, and that element is the empty set.
:::


Some commonly used sets of numbers include:

* *Natural* (or counting) numbers\index{Number sets!natural} $\mathbb{N}$: $\{1, 2, 3, \dots\}$.
* *Integers*\index{Number sets!integers} $\mathbb{Z}$: $\{\ldots, -2, 1, 0, 1, 2, 3, \dots\}$.
* *Rational*\index{Number sets!rationals} numbers $\mathbb{Q}$, including numbers like $1/2 = 0.5$.
* *Real*\index{Number sets!reals} numbers $\mathbb{R}$, including numbers like $\pi$.
* *Positive real*\index{Number sets!positive reals} numbers $\mathbb{R}_{+}$.
* *Negative real*\index{Number sets!negative reals} numbers $\mathbb{R}_{-}$.
* *Complex*\index{Number sets!complex} numbers $\mathbb{C}$.

(Some authors include\ $0$ in the set of natural numbers.)


## Cardinality of sets {#FiniteInfiniteSets}

Sets can be:\index{Cardinality}

* *finite*, where elements can be counted; the number of elements\ $n$ is finite.
* *countable infinite*: elements could be listed in a sequence, but there are infinitely many.
* *uncountably infinite*: elements cannot all be listed in a sequence.

For finite or countably infinite sets, each element is called an *element* or a *sample point*.


### Finite sets {#FiniteSets}

If\index{Finite sets}\index{Cardinality!finite sets} the number of elements in a set can be counted, the set is called a *finite* set.
A finite set has exactly\ $n$ distinct elements for some non-negative integer\ $n$.


::: {.example #DefiningSets name="Defining elements of a finite set"}
Define the set $S$ as the 'set of suits in a standard pack of cards'.
Then, $S = \{\clubsuit, \spadesuit, \heartsuit, \diamondsuit\}$.
We can write, for example, that $\spadesuit \in S$.

The elements of the set are listed; the set has $n = 4$ elements.
:::


::: {.example #DefiningSets4 name="Defining elements of a finite set"}
Define the set\ $O$ as the 'set of odd numbers between\ $0$ and\ $100$'.
Alternatively, the elements of the set can defined by giving the pattern:
$$
  O = \{1, 3, 5, 7, 9, \dots 95, 97, 99\}
$$
Set\ $O$ has $50$\ elements.
:::


::: {.example #FiniteSets name="Finite sets"}
Define set $B$ as 'the even numbers after rolling a single die'; then,
`r if (knitr::is_html_output()) '<!--'`
 $$
  B = \big\{
   \largedice{2},
   \largedice{4},
   \largedice{6} \big\}.
$$
`r if (knitr::is_html_output()) '-->'`
`r if (knitr::is_latex_output()) '<!--'`
<em>B</em> = {<span class="larger-die">&#9857;</span>, <span class="larger-die">&#9859;</span>, <span class="larger-die">&#9861;</span>}
`r if (knitr::is_latex_output()) '-->'`

The set\ $B$ is *finite*, with three elements.
:::


### Countably infinite sets {#CountablyInfiniteSets}

If\index{Countably infinite sets}\index{Cardinality!countably infinite} the elements in a set could (in principle) be listed without omitting any element, but an infinite number of them exist, the set is called a *countably infinite* set.


::: {.definition #CountablyInfiniteSets name="Countably infinite set"}
A set\ $A$ is called *countably infinite* if a *bijection*\index{Bijection} exists between the elements of\ $A$ and the natural numbers\ $\mathbb{N}$; that is, the elements of\ $A$ can be listed in a sequence $a_1, a_2, a_3,\dots$ where every element of\ $A$ appears exactly once.
:::


::: {.example #DefiningSets5 name="Countably infinite sets"}
Define $S$ as the number of rolls of a die needed to roll a
`r if (knitr::is_latex_output()) {
   '\\largedice{6}'
} else {
   '<span class="larger-die">&#9861;</span>'
}`.
The number of rolls needed could be $1$ or $2$ or $3$..., and no upper limit exists (though some numbers are very unlikely).
That is, the set\ $S$ is
$$
  S = \{1, 2, 3, 4, \dots\}.
$$
Set\ $S$ is *countably infinite*.
:::


::: {.example #DefiningSets6 name="Countably infinite sets"}
Define the set $O$ as the 'set of odd numbers'.
The elements of the set can be described ('odd numbers'), or by giving the pattern:
$$
  O = \{1, 3, 5, 7, 9, \dots\}
$$
Set\ $O$ is *countably infinite*.
:::


### Uncountably infinite sets {#InfiniteSets}

If\index{Uncountably infinite sets}\index{Cardinality!uncountably infinite sets} the elements cannot be listed without omitting elements, the set is called a *uncountably infinite* set.


:::{.definition #UncountablyInfiniteSets name="Uncountably infinite sets"}
A set\ $A$ is called *uncountably infinite* if it is infinite but *cannot* be listed in a sequence $a_1, a_2, a_3, \cdots$; that is, no matter how clever the listing, some elements will always be omitted. 
Such a set is too large to be paired with $\mathbb{N}$.
:::


::: {.example #UncountablyInfiniteSets name="Uncountably infinite sets"}
The set $\mathbb{R}$ refers to the positive real numbers, which is an *uncountably infinite* set.
The elements of $\mathbb{R}$ cannot be listed without missing some elements.
For instance, an infinite number of elements exists between the elements\ $0$ and\ $0.0001$.

Set\ $\mathbb{R}$ is *uncountably infinite*.
:::


:::{.example #HeightsOfPeople name="Uncountably infinite set"}
Consider the heights of people.
Between the two heights $173\cms$ and $174\cms$, an infinite number of heights exist:
$173.1\cms$, $173.01\cms$, $173.001\cms$, $173.0001\cms$, and so on.

The set of heights is uncountably infinite.

For convenience heights are rounded to (say) the nearest whole number of centimetres, or perhaps to one decimal place of a centimetre.
:::


Since the elements of uncountably infinite sets cannot be listed---even with a pattern---the elements of uncountably infinite sets are often concisely defined using *set-builder notation*.\index{Set-builder notation}
(Set-builder notation can be used for any sets, but are especially useful for uncountably infinite sets.)


### Set-builder notation {#SetBuilderNotation}

\index{Set-builder notation|(}
Set-builder notation usually has the format:
\begin{align*}
  A &= \{ x \mid \text{some condition for $x$ to belong to the set}\}, \text{or}\\
  A &= \{ x :    \text{some condition for $x$ to belong to the set}\}.
\end{align*}
In this notation, $\{\dots\}$ means that\ $A$ is a set; $x$ is a variable representing the elements of the set\ $A$; the colon or vertical bar is read as 'such that', and is followed by the condition for $x$ to belong to the set\ $A$.
Often, the notation to the left of the 'such that' defines the universal set for the context.


:::{.example #SetBuilderNotation1 name="Set-builder notation"}
Consider the set defined by
$$
   A = \{x \in \mathbb{Z} \mid x > 9\}.
$$
This is read as '$A$ is the set of all integers\ $x$, such that the elements\ $x$ are larger than\ $9$'.
:::


:::{.example #SetBuilderNotation name="Set-builder notation"}
Consider the set defined by
\begin{align*}
   F = &\{s \in \text{Names of students in a certain course} \mid{}\\
       &\phantom{\{x \in {}} \text{students whose mark is less than $50\%$}\}.
\end{align*}
This is read as '$F$ is the set of all\ $s$ drawn from the names of students in a certain course, such that\ $s$ is a student whose mark less than $50$%'.
That is, $F$ represents the set of all students in that specific course who scored less than\ $50$%.
:::


:::{.example #SetBuilderNotation2 name="Set-builder notation: finite sets"}
Consider the set defined by
$$
   B = \{x \in \text{Animal species}: \text{$x$ has two legs}\}
$$
This is read as '$B$ is the set of all animals species\ $x$, such that the elements\ $x$ have two legs'.
We could also have written
$$
   B = \{x : \text{$x$ is an animal species with two legs}\}
$$
:::


:::{.example #SetBuilderNotation3 name="Set-builder notation: finite sets"}
Consider the set defined by
$$
    V = \{x \in \text{letters in the alphabet} \mid \text{$x$ is a vowel}\}.
$$
This is read as '$V$ is the set of all letters \ $x$, such that the elements\ $x$ are all vowels'.

Consider the set defined by
$$
   P = \{y \in \mathbb{Z} : \text{$y$ is a two-digit integer}\}.
$$
This is read as '$P$ is the set of integers\ $y$, such that the elements\ $y$ are all two-digit integers'.
:::


:::{.example #CricketBallDefine name="Set-builder notation: uncountably infinite sets"}
Consider throwing a cricket ball.
Let\ $D$ be the set of all possible distances (in metres) that the ball could be thrown; then
$$
   D = \{ d \in \mathbb{R} \mid d \ge 0 \}.
$$
Set $D$ is *uncountably infinite*.

Define $C$ as the set of 'all possible distances travelled by cars in Europe over one year'.
Set\ $C$ is *uncountably infinite*, and can be defined as
$$
  C = \{c \in \mathbb{R} \mid c \ge 0 \}.
$$
Set $C$ is *uncountably infinite*.
:::
\index{Set-builder notation|)}


### Cardinality {#Cardinality}

*Cardinality*\index{Cardinality} denotes the number of elements in a set.
Cardinality of a set\ $A$ is denoted\ $|A|$.

Cardinality of a finite set is the number of elements in the set.
Countably infinite sets and uncountably infinite sets both have *infinite* cardinality, but different infinite cardinalities.
Countably infinite sets have cardinality $\aleph_0$, where $\aleph_0$ ('aleph-null') is the smallest infinite cardinality.
Uncountably infinite sets have cardinality $\mathfrak{c}$ where $\mathfrak{c}$ is the 'cardinality of the continuum'.
Cantor showed that $\mathfrak{c} = 2^{\aleph_{0}} > \aleph_0$.


::: {.example #SizeOfSetCardinality name="Cardinality: finite sets"}
Set $S$ defined in Example\ \@ref(exm:DefiningSets) is
$$
  S = \{\clubsuit, \spadesuit, \heartsuit, \diamondsuit\}.
$$
The elements of the set are listed; the set has four elements.
That is, $|S| = 4$.
:::


:::{.example #SizeOfCardinality1 name="Cardinality: uncountably infinite sets"}
Set\ $A$ in Example\ \@ref(exm:SetBuilderNotation1) is
$$
   A = \{x \in \mathbb{Z} \mid x > 9\}.
$$
The cardinality of\ $A$ is  $\aleph_0$: $|A| = \aleph_0$.
:::


:::{.example #CricketBallCardinality name="Cardinality: uncountably infinite sets"}
Set\ $D$ in Example\ \@ref(exm:CricketBallDefine) is
$$
   D = \{ d \in \mathbb{R} \mid d \ge 0 \}.
$$
The cardinality of set $D$ is $\mathfrak{c}$:  $|D| = \mathfrak{c}$.
:::




## Operations on sets {#RelationshipsBetweenSets}

Sets can be combined in various ways, which sometimes makes defining sets easier too.
Many operations are defined for sets.\index{Set operations}


:::{.definition #IntersectionUnionSubset name="Set operations"}
Consider two sets\ $A$ and\ $B$, and a universal set\ $S$.
The following operations are defined.

* **Intersection**:\index{Set operations!intersection}  
  The *intersection* of sets $A$ and $B$, written as $A\cap B$, is the set of elements in *both* $A$ and $B$: $A\cap B = \{x \mid x\in A \text{ and } x \in B\}$.
  We often say $A$ **and** $B$.
* **Union**:\index{Set operations!union}  
  The *union* of sets $A$ and $B$, written as $A\cup B$, is the combined set of all the elements in either $A$ or $B$, or in both: $A\cup B = \{x \mid x\in A \text{ or } x \in B\}$.
  We often say $A$ **or** $B$.
* **Complement**:\index{Set operations!complement}  
  The *complement* of set\ $A$, written\ $A^c$, is all the elements of\ $S$ that are *not* in set\ $A$: $A^c = \{x \mid x\notin A\}$.
* **Cartesian product**:\index{Set operations!Cartesian product}  
  The Cartesian product of\ $A$ and\ $B$, written $A\times B$, is the combination of all elements of\ $A$ with all elements of\ $B$.
  For example,  if $A = \{ 1, 2\}$ and $B = \{10, 20\}$, then $A\times B = \{(1, 10), (1, 20), (2, 10), (2, 20) \}$.
* **Difference**, **set difference** or **set subtraction**:\index{Set operations!set difference}  
  The *difference* between sets\ $A$ and set\ $B$, written $A\setminus B$ or $A - B$, is all the elements in\ $A$ that are *not* in set\ $B$: $A\setminus B = \{x \mid x\in A \text{ and } x \notin B\}$.
  In other words, $A\setminus B = A\cap B^c$.
  Note that $A\setminus B$ is not the same as $B\setminus A$, in general.
* **Subset**:\index{Set operations!subset}    
  $A$ is a *subset* of\ $B$, written $A\subseteq B$, if every element of\ $A$ is also an element of $B$.
  Set $B$ may have elements that are not in\ $A$.
* **Proper subset**:\index{Set operations!proper subset}    
  $A$ is a *proper subset* of\ $B$, written $A\subset B$, if every element of\ $A$ is also an element of $B$, and set $B$ has elements that are not in\ $A$.
* **Symmetric difference**:\index{Set operations!symmetric difference} 
  The *symmetric difference* of sets $A$ and $B$ is written as $A\Delta B$, and is the combined set of all the elements in either $A$ or $B$, *but not* in both.
  By definition, $A\Delta B = (A\setminus B) \cup (B\setminus A)$.
  This operation is also called *exclusive or*.\index{Set operations!exclusive or}
:::

Two sets whose intersection is the empty set (i.e., $A\cap B = \varnothing$ for two sets\ $A$ and\ $B$) are called *disjoint sets*.\index{Sets!disjoint}

::: {.notationBox .notation data-latex="{iconmonstr-infinity-8-240.png}"}
Other common notations for the complement of a set\ $A$ include\ $\overline{A}$ or\ $A'$.
:::



::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
Be careful using **and** and **or**.
In general conversation, **and** can mean the *intersection* or the *union*, depending on the context.
For example:

* 'People who are female and aged under $20$: please stand.'

The context means 'intersection': being 'female' (say, $F$) and being 'aged under\ $20$' (say, $T$) both apply to the same individual.
The sentence refers to $F\cap T$.

* 'Engineers and doctors: please stand.' 

The context means 'union': people who are engineers (say, $E$), or people who are doctors (say, $D$), or both.
The sentence refers to $E \cup D$.
In this sentence, **and** is combining two groups of people, not two *conditions* on the same people.

Similarly, in everyday speech **or** can be inclusive (one, the other, or both) or exclusive (one, or the other, but *not* both).
For example, 

* 'Do you have milk or sugar in your coffee?'

The context means inclusive **or**, as defined above.
Taking sugar (say, $S$) and taking milk (say, $M$) are combined as $S\cup M$: you may take milk, you may take sugar, but you may also take both.

* 'Answer this quiz using true or false'.

The context here means **exclusive or**.
Answering true (say, $T$) and answering false (say, $F$) are combined as $T \Delta F$: it means to answer with true, or with false, but *not* both.

Context matters!
In the context of sets in mathematics and statistics, $A$ **and** $B$ always means $A \cap B$ (intersection), and $A$ **or** $B$ always means $A\cup B$ (union).
:::


::: {.example #RelationshipBetweenSets5 name="Relationships between sets"}
Suppose we define these sets:
\begin{align*}
   C &= \{ 0, 1, 2, 3, 4, 5\};\\
   D &= \{ 4, 5, 6, 7\};\\
   E &= \{ 6 \}.
\end{align*}
Then:
\begin{align*}
   C \cap D &= \{ 4, 5\};                   &          C \cup D &= \{ 0, 1, 2, 3, 4, 5, 6, 7\};\\ 
   E &\in D;                                &          E &\notin C;\\ 
   C \cap E &= \varnothing;                   &          D \cap E &= D;\\
   C\setminus D &= \{0, 1, 2, 3\};          &          D\setminus C &= \{6, 7\};\\
   C\Delta D &= \{0, 1, 2, 3, 6, 7\}.
\end{align*}
:::



::: {.example #RelationshipBetweenSets6 name="Relationships between sets"}
Suppose we define these sets:
\begin{align*}
   U &= \{\text{All the digits}\} = \{ 0, 1, 2, 3, 4, 5, 6, 7, 8, 9\};\\
   B &= \{ 6, 7, 8, 9\};\\
   E &= \{ 2, 4, 6, 8\}.
\end{align*}
Here, $U$ is the *universal set*, the set of all digits,
Then, for example:
\begin{align*}
   U \cap B &= \{ 6, 7, 8, 9 \} = B;        &          B \cup E &= \{2, 4, 6, 7, 8, 9\};\\ 
   B &\in U;                                &          U &\notin B;\\ 
   B \cap E &= \{6, 8\};                    &          U \cup E &= U;\\
   B \setminus E &= \{7, 8, 9\};            &          E\setminus U &= \varnothing;\\
   B\Delta E &= \{2, 4, 7, 9\};             &          E^c &= \{0,1, 3, 5, 7, 9\}.
\end{align*}
:::


::: {.example #RelationshipBetweenSets4 name="Relationships between sets"}
Consider the set\ $A$ defined using set-builder notation as:\index{Set-builder notation}
$$
  A = \{ x  \in \mathbb{R} \mid \sqrt{x^2 - 4} \in \mathbb{R}\}.
$$
This means that the set\ $A$ comprises the real elements\ $x$ such that the values of $\sqrt{x^2 - 4}$ are real numbers; that is, such that $x^2 - 4 \ge 0$.
Alternatively
$$
  A = \{ x \in\mathbb{R} \mid (-\infty, -2] \cup [2, \infty)\}.
$$
:::



::: {.notationBox .notation data-latex="{iconmonstr-infinity-8-240.png}"}
The square brackets\ '$[$' and\ '$]$' indicate the value *is included* in the interval, where as round brackets\ '$($' and\ '$)$' indicated the value *is not included* in the interval.

When an interval extends to infinity, it must always use round brackets; for example,
$$
  (-\infty, -2] \quad\text{and}\quad [2, \infty)
$$
as used above.
This is because\ $\infty$ and\ $-\infty$ are not real numbers, but symbols representing unboundedness. 
Since they are not elements of\ $\mathbf{R}$, they cannot be included in any set, and square brackets are never used with infinity.
:::


:::{.example #QuickUseOfOperations name="Set operations"}
We can define $\mathbb{N}_0 = \{ 0 \} \cup \mathbb{N}$.

We could write $\mathbb{Q} = \{\frac{m}{n}\mid m\in\mathbb{Z}, n \in \mathbb{Z}, n\ne 0\}$.

We can also write $\mathbb{Z} \in \mathbb{R}$, but $\mathbb{R} \notin \mathbb{Z}$.
:::


:::{.example #TwoDSets name="Sets can be two-dimensional"}
Sets do not have to be one-dimensional.\index{Sets!two-dimensional}
We could define the set $P$ as
$$
  P = \{(x, y)\in \mathbb{R}\times\mathbb{R}  \mid (0 < x < 1)\cap(0 < y < 1) \},
$$
where $\mathbb{R}\times\mathbb{R}$ means 'the ordered pairs $(x, y)$ where both\ $x$ and\ $y$ are real numbers'.
$P$ defines a unit square on the Cartesian axes.
:::



## Venn diagrams {#VennDiagramsSets}

*Venn diagrams*\index{Venn diagrams} provide a visual way to represent sets and their relationships.
A rectangle represents the universal set\ $S$, and closed curves inside it (such as circle or ellipses) represent subsets.
Shaded regions indicate the elements belonging to a particular set or operation.

The visual representations of many of the basic set operations from Sect.\ \@ref(RelationshipsBetweenSets) (Def.\ \@ref(def:IntersectionUnionSubset)) are shown in Fig.\ \@ref(fig:IntersectionUnionSubset).


```{r IntersectionUnionSubset, echo=FALSE, out.width='95%', fig.height = 8, fig.width=5.5, fig.align="center", fig.cap="Venn diagrams showing intersection, union, complement, proper subset, set differences and symmetric difference. The rectangle represents the universal set, $S$.", message=FALSE}
#par( mar = c(0.05, 0.05, 0.05, 0.05))
par(mfrow = c(4, 2), 
    mar = c(2.5, 1, 3.5, 1) + 0.1 )

colourA <- rgb(0, 0, 255, 
               max = 255, 
               alpha = 125)
colourB <- rgb(0, 255, 0, 
               max = 255, 
               alpha = 125)
colourS <- rgb(100, 100, 100, 
               max = 255, 
               alpha = 125)
colourXOR <- "white"

make_ellipse <- function(cx, cy, a, b, n = 200) {
  theta <- seq(0, 2 * pi, length.out = n)
  coords <- cbind(
    cx + a * cos(theta),
    cy + b * sin(theta)
  )
  coords <- rbind(coords, coords[1, ])  # CLOSE the polygon
  st_polygon(list(coords))
}


# Create universal set as rectangle
U <- st_sfc(st_polygon(list(
  matrix(c(0, 0,
           1, 0,
           1, 1,
           0, 1,
           0, 0), 
         ncol = 2, 
         byrow = TRUE)
)))


# Sets A and B
A <- st_sfc( make_ellipse(0.4, 0.5, 0.22, 0.16) )
B <- st_sfc( make_ellipse(0.6, 0.5, 0.20, 0.30) )

# Universal set S
S <- st_sfc(
  st_polygon(list(rbind(
    c(0, 0), 
    c(1, 0), 
    c(1, 1), 
    c(0, 1), 
    c(0, 0)
  )))
)

A_only <- st_difference(A, B)         # A minus B
B_only <- st_difference(B, A)         # B minus A
A_subset <- st_sfc( make_ellipse(0.45, 0.5, 0.07, 0.12) )
INTERSECT <- st_intersection(A, B)
UNION <- st_union(A, B)
XOR <- st_sym_difference(A, B)        # symmetric difference
COMPLEMENT <- st_difference(S, A)

colour1 <- plotColour1


####################

### INTERSECTION
# Universal set
plot(0, 0, 
     type = "n", 
     xlim = c(0, 1), 
     ylim = c(0, 1),
     xlab = "", 
     ylab = "", 
     axes = FALSE,
     main = expression( 
        paste(bold("Intersection"), " of ", italic(A)~"and"~italic(B)) ) )

# Universal set frame
polygon(x = c(0, 0, 1, 1), 
        y = c(0, 1, 1 , 0), 
        border = "black")

# Plot regions in correct order
plot(A_only, 
     col = "white", 
     border = NA, 
     add = TRUE)
plot(B_only, 
     col = "white", 
     border = NA, 
     add = TRUE)
plot(INTERSECT, 
     col = colour1, 
     border = NA, 
     add = TRUE)

# Draw outlines
plot(st_geometry(A), 
     add = TRUE, 
     border = "black", 
     lwd = 1)
plot(st_geometry(B), 
     add = TRUE, 
     border = "black", 
     lwd = 1)

# Labels
text(0.4, 0.75, 
     labels = expression(italic(A)))
text(0.7, 0.85, 
     labels = expression(italic(B)))
mtext(expression(italic(S)), 
      side = 1, 
      adj = 0.25)


arrows( x0 = 0.5, 
        y0 = 0.18,
        x1 = 0.5,
        y1 = 0.5,
        angle = 15,
        length = 0.10,
        lwd = 1)
text(0.5, 0.18,
     pos = 1,
     label = expression(italic(A) ~~ intersect() ~~ italic(B) ) )






### UNION
# Universal set
plot(0, 0, 
     type = "n", 
     xlim = c(0, 1), 
     ylim = c(0, 1),
     xlab = "", 
     ylab = "", 
     axes = FALSE,
     main = expression(
        paste(bold(Union), " of ", italic(A)~"and"~italic(B)) ) )

# Universal set frame
polygon(x = c(0, 0, 1, 1), 
        y = c(0, 1, 1 , 0), 
        border = "black")

# Plot regions in correct order
plot(A_only, 
     col = "white", 
     border = "black",
     lty = 3,
     add = TRUE)
plot(B_only, 
     col = "white", 
     border = "black", 
     lty = 3,
     add = TRUE)
plot(UNION, 
     col = colour1, 
     border = "black", 
     add = TRUE)


# Labels
text(0.4, 0.75, 
     labels = expression(italic(A)))
text(0.7, 0.85, 
     labels = expression(italic(B)))
mtext(expression(italic(S)), 
      side = 1, 
      adj = 0.25)


text(0.5, 0.18,
     pos = 1,
     label = expression(italic(A) ~~ union() ~~ italic(B) ) )






### COMPLEMENT
# Complement
plot(0, 0, 
     type = "n", 
     xlim = c(0, 1), 
     ylim = c(0, 1),
     xlab = "", 
     ylab = "", 
     axes = FALSE,
     main = expression(
        paste(bold(Complement), " of ", italic(A)) ) )

# Universal set frame
polygon(x = c(0, 0, 1, 1), 
        y = c(0, 1, 1 , 0), 
        border = "black")

# Plot regions in correct order
plot(A_only, 
     col = "white", 
     border = NA, 
     add = TRUE)
plot(COMPLEMENT, 
     col = colour1, 
     border = "black", 
     add = TRUE)


# Labels
text(0.4, 0.5, 
     labels = expression(italic(A)))
mtext(expression(italic(S)), 
      side = 1, 
      adj = 0.25)


text(0.5, 0.18,
     pos = 1,
     label = expression(italic(A) ^ italic(c) ) )








### SUBSET
# Universal set
plot(0, 0, 
     type = "n", 
     xlim = c(0, 1), 
     ylim = c(0, 1),
     xlab = "", 
     ylab = "", 
     axes = FALSE,
     main = expression( 
        paste(bold("Proper subset"), " of ", italic(A)) ) )

# Universal set frame
polygon(x = c(0, 0, 1, 1), 
        y = c(0, 1, 1 , 0), 
        border = "black")

# Plot regions in correct order
plot(A_only, 
     col = "white", 
     border = NA, 
     add = TRUE)
plot(A_subset, 
     col = colour1, 
     border = "black", 
     add = TRUE)

# Draw outlines
plot(st_geometry(A), 
     add = TRUE, 
     border = "black", 
     lwd = 1)

# Labels
text(0.4, 0.75, 
     labels = expression(italic(A)))
text(0.45, 0.5, 
     labels = expression(italic(B)))
mtext(expression(italic(S)), 
      side = 1, 
      adj = 0.25)

text(0.5, 0.18,
     pos = 1,
     label = expression(italic(B)%subset%italic(A) ) )






### SET DIFFERENCE A - B
# Universal set
plot(0, 0, 
     type = "n", 
     xlim = c(0, 1), 
     ylim = c(0, 1),
     xlab = "", 
     ylab = "", 
     axes = FALSE,
     main = expression( 
        paste(bold("Difference "), italic(A)~~"\u2216"~italic(B) ) ))

# Universal set frame
polygon(x = c(0, 0, 1, 1), 
        y = c(0, 1, 1 , 0), 
        border = "black")

# Plot regions in correct order
plot(A_only, 
     col = colour1, 
     border = NA, 
     add = TRUE)
plot(B_only, 
     col = "white", 
     border = NA, 
     add = TRUE)
plot(INTERSECT, 
     col = "white", 
     border = NA, 
     add = TRUE)

# Draw outlines
plot(st_geometry(A), 
     add = TRUE, 
     border = "black", 
     lwd = 1)
plot(st_geometry(B), 
     add = TRUE, 
     border = "black", 
     lwd = 1)

# Labels
text(0.4, 0.75, 
     labels = expression(italic(A)))
text(0.7, 0.85, 
     labels = expression(italic(B)))
mtext(expression(italic(S)), 
      side = 1, 
      adj = 0.25)


arrows( x0 = 0.35, 
        y0 = 0.18,
        x1 = 0.35,
        y1 = 0.5,
        angle = 15,
        length = 0.10,
        lwd = 1)
text(0.35, 0.18,
     pos = 1,
     label = expression(italic(A) ~~"\u2216"~~ italic(B) ) )







### SET DIFFERENCE: B - A
# Universal set
plot(0, 0, 
     type = "n", 
     xlim = c(0, 1), 
     ylim = c(0, 1),
     xlab = "", 
     ylab = "", 
     axes = FALSE,
     main = expression( 
        paste(bold("Difference "), italic(B)~~"\u2216"~italic(A) ) ))

# Universal set frame
polygon(x = c(0, 0, 1, 1), 
        y = c(0, 1, 1 , 0), 
        border = "black")

# Plot regions in correct order
plot(A_only, 
     col = "white", 
     border = NA, 
     add = TRUE)
plot(B_only, 
     col = colour1, 
     border = NA, 
     add = TRUE)
plot(INTERSECT, 
     col = "white", 
     border = NA, 
     add = TRUE)

# Draw outlines
plot(st_geometry(A), 
     add = TRUE, 
     border = "black", 
     lwd = 1)
plot(st_geometry(B), 
     add = TRUE, 
     border = "black", 
     lwd = 1)

# Labels
text(0.4, 0.75, 
     labels = expression(italic(A)))
text(0.7, 0.85, 
     labels = expression(italic(B)))
mtext(expression(italic(S)), 
      side = 1, 
      adj = 0.25)


arrows( x0 = 0.70, 
        y0 = 0.18,
        x1 = 0.70,
        y1 = 0.5,
        angle = 15,
        length = 0.10,
        lwd = 1)
text(0.70, 0.18,
     pos = 1,
     label = expression(italic(B) ~~"\u2216"~~ italic(A) ) )








### EXCLUSIVE OR
make_ellipse <- function(cx, cy, a, b, n = 200) {
  theta <- seq(0, 2 * pi, length.out = n)
  coords <- cbind(
    cx + a * cos(theta),
    cy + b * sin(theta)
  )
  coords <- rbind(coords, coords[1, ])  # CLOSE the polygon
  st_polygon(list(coords))
}


# Create universal set as rectangle
U <- st_sfc(st_polygon(list(
  matrix(c(0,0,
           1,0,
           1,1,
           0,1,
           0,0), 
         ncol = 2, 
         byrow = TRUE)
)))



# Compute the regions
A_only <- st_difference(A, B)         # A minus B
B_only <- st_difference(B, A)         # B minus A
XOR <- st_sym_difference(A, B)        # symmetric difference
overlap <- st_intersection(A, B)      # intersection


# Universal set
plot(0, 0, 
     type = "n", 
     xlim = c(0, 1), 
     ylim = c(0, 1),
     xlab = "", 
     ylab = "", 
     axes = FALSE,
     main = expression( atop(
        paste(bold("Symmetric difference"), " of ", italic(A)~"and"~italic(B)),
              italic(A)*Delta*italic(B) ) ) )

# Universal set frame
polygon(x = c(0, 0, 1, 1), 
        y = c(0, 1, 1 , 0), 
        border = "black")

# Plot regions in correct order
plot(A_only, 
     col = colour1, 
     border = NA, 
     add = TRUE)
plot(B_only, 
     col = colour1, 
     border = NA, 
     add = TRUE)
plot(overlap, 
     col = "white", 
     border = NA, 
     add = TRUE)

# Draw outlines
plot(st_geometry(A), 
     add = TRUE, 
     border = "black", 
     lwd = 1)
plot(st_geometry(B), 
     add = TRUE, 
     border = "black", 
     lwd = 1)

# Labels
text(0.4, 0.75, 
     labels = expression(italic(A)))
text(0.75, 0.85, 
     labels = expression(italic(B)))
mtext(expression(italic(S)), 
      side = 1, 
      adj = 0.25)


arrows( x0 = 0.5, 
        y0 = 0.15,
        x1 = 0.3,
        y1 = 0.5,
        angle = 15,
        length = 0.10,
        lwd = 1)
arrows( x0 = 0.5, 
        y0 = 0.15,
        x1 = 0.6,
        y1 = 0.3,
        angle = 15,
        length = 0.10,
        lwd = 1)
text(0.5, 0.15,
     pos = 1,
     label = expression(italic(A)~Delta~italic(B) ) )
```


## Set algebra {#SetAlgebra}

Set algebra has many laws; we only provide some.\index{Set}
For sets\ $A$, $B$ and\ $C$ defined on the universal set\ $S$ these laws are defined.

:::::: {.cols data-latex=""}
::: {.col data-latex="{0.49\textwidth}"}


The *commutative* laws are:\index{Set algebra!commutative laws}

* $A\cup B = B \cup A$.
* $A\cap B = B \cap A$.
* $A\Delta B = B\Delta A$.
* $A\setminus B \ne B\setminus A$ in general.

The *associative laws* are:\index{Set algebra!associative laws}

* $A\cup(B\cup C) = (A\cup B)\cup C$.
* $A\cap(B\cap C) = (A\cap B)\cap C$.
* $(A\Delta B)\Delta C = A\Delta (B\Delta C)$.

The *distributive laws* are:\index{Set algebra!distributive laws}

* $A\cap (B\cup C) = (A\cap B)\cup (A \cap C)$.
* $A\cup (B\cap C) = (A\cup B)\cap (A \cup C)$.
* $A\setminus(B \cup C) = (A\setminus B)\cap (A\setminus C)$.
* $A\setminus(B \cap C) = (A\setminus B)\cup (A\setminus C)$.
* $(A \cup B)\setminus C = (A\setminus C)\cup(B\setminus C)$.
* $(A \cap B)\setminus C = (A\setminus C)\cap(B\setminus C)$.

:::

::: {.col data-latex="{0.025\textwidth}"}
\ 
<!-- an empty Div (with a white space), serving as
a column separator -->
:::

::: {.col data-latex="{0.49\textwidth}"}


The *absorption laws* are:\index{Set algebra!absorption laws}

* $A\cup (A\cap B) = A$.
* $A\cap (A\cup B) = A$.
* $A\setminus \varnothing = A$.
* $A\setminus A = \varnothing$.
* $A\Delta\varnothing = A$.
* $A\Delta A = \varnothing$.

The *dominance laws*:

* $A\cup S = S$.
* $A\cap\varnothing = \varnothing$.

The *idempotent laws*:\index{Set algebra!idempotent laws}

* $A\cup A = A$.
* $A\cap A = A$.

*De Morgan's* law are:\index{Set algebra!de Morgan's laws} 

* $(A \cap B)^c = A^c \cup B^c$.
* $(A \cup B)^c = A^c \cap B^c$.

:::

::::::


<!-- These are two separate lines, since a second level of indent begins with a dash, and look like a minus sign, which is confusing -->

Two proofs are given below to show how these laws can be used.

:::{.example #ProofsWithSetAlgebra name="Using set algebra"}
Prove $A\cup B = B \cup A$ (which may seem 'obvious').

To show this proposition, we need to show that 
\begin{align}
   (A \cup B) &\subseteq B\cup A\qquad\text{and also that}\\
   (B \cup A) &\subseteq A\cup B.
\end{align}
Consider the first statement, and consider an element\ $x$ such that $x\in (A\cup B)$.
This means that either:

* $x\in A$, in which case $x\in (B\cup A)$; or
* $x\in B$, in which case $x\in (B\cup A)$.

In either case, $x \in (B\cup A)$ and so $(A\cup B)\subseteq B\cup A$.

The proof of the second statement is similar, and hence $A\cup B = B \cup A$.
:::


These set-algebra rules can be used to prove numerous other properties of sets too.


:::{.example #UsingSetAlgebra name="Using set algebra"}
Consider the statement $A\cap (A^c \cup B)$.
Set algebra rules can be used to simplify:
\begin{align*}
     A\cap (A^c \cup B)
     &= (A\cap A^c) \cup (A\cap B)\qquad\text{(distributive law)}\\
     &= \varnothing \cup (A\cap B)\qquad\text{(by definition of $A^c$)}\\
     &= A\cap B.\qquad\text{(dominance law)}
\end{align*}
Thus, $A\cap (A^c \cup B) = A\cap B$.
:::

:::{.example #UsingSetAlgebra2 name="Using set algebra"}
Prove $B\setminus (A\setminus B) = B\setminus A$.

By definition of the set difference:
\begin{align*}
  B\setminus (A\setminus B) 
  &= B\setminus(A\cap B^c)\\
  &= (B\setminus A)\cup (B\setminus B^c)\quad\text{(distributive law)}\\
  &= (B\cap A^c) \cup B.
\end{align*}
Applying a distributive law again:
\begin{align*}
  B\setminus (A\setminus B) 
  &= (B\cup B) \cap (A^c \cap B),
\end{align*}
and since $B\cup B = B$,
\begin{align*}
  (B\cup B) \cap (A^c \cap B)  
  &= B \cap (A^c \cap B)\\
  &= B \cap (B \cap A^c)\\
  &= (B \cap B) \cap A^c\\
  &= B\cap A^c,
\end{align*}
which is the definition of $B\setminus A$.
:::


## Statistical computing with R {#StatisticalComputingWithR}

We introduce **R** by showing how to operate with sets; more general information about using **R** is given in Sect.\ \@ref(ShortRIntro).
Greater functionality in working with sets is provided by installing and using the **R** package **sets**.


### Simple example {#SetsInRExample1}

Consider rolling a standard, fair six-sided die.
Define\ $A$ as the set of numbers that can be rolled on the die, $B$ as the set of *even* numbers that can be rolled, and\ $C$ as the numbers that can be rolled that are divisible by three.

**R** can be used to determine the elements in these sets, and the union and intersection of the sets.

First, we define the sets:

```{r}
Die_Rolls <- 1:6 # Set A
Rolls_Even <- seq(2, 6, by = 2)    # Set B
Rolls_Divisible_By_3 <- c(3, 6)    # Set C
Rolls_Even                         # Set B
Rolls_Divisible_By_3               # Set C
```

Then use **R** to find the intersection and union:

```{r}
intersect(Rolls_Even, Rolls_Divisible_By_3)
union(Rolls_Even, Rolls_Divisible_By_3)
sort( union(Rolls_Even, Rolls_Divisible_By_3) )
```

The function `setdiff()` finds the difference between two sets:

```{r}
setdiff(Rolls_Even, Rolls_Divisible_By_3)  # B \ C
setdiff(Rolls_Divisible_By_3, Rolls_Even)  # C \ B: not symmetric!
```

`is.element(A, B)` determines if the elements in\ $A$ are contained in set\ $B$:

```{r}
is.element(Rolls_Divisible_By_3, Rolls_Even)  # The two elements of A
is.element(Rolls_Even, Rolls_Divisible_By_3)  # The three elements of B
```



### More involved example {#SetsInRExample2}

The above example could be completed manually without the need to use **R**.
However, the power of using computers is more obvious for more involved examples.

Consider the two sets:
\begin{align*}
  A &= \{\text{Countries whose names have `a' as the \emph{second} letter}\};\\
  B &= \{\text{Countries whose names end with `a'}\}.
\end{align*}

In **R**, the names of the countries can be found by loading the `countries` package (assuming it is installed):

```{r}
library("countries")
```

The list of country names are in the **R** vector `list_countries()`:

```{r}
head( list_countries() ) # Show the first five values only
```

First, we find the countries whose names have an `a` as the second letter (assuming it is always lower case); that is, we identify the elements in set\ $A$:

```{r}
# Find where, in the list of countries, 
# a country name has an `a` in position 2:
# (Assumes the letter is lower case)
Locate_a_Second <-  which( substr(list_countries(),  # substr()  means 'sub-string'
                                  start = 2,         # Start at letter 2 
                                  stop = 2)          # End at letter 2
                           == "a")
Locate_a_Second    # List of TRUE and FALSE for each listed country

# Now find the countries listed at those locations:
a_Second <- list_countries()[Locate_a_Second]

# How many countries are there whose names have an `a` as second letter?
length(a_Second)
```


Then, we find the countries whose names end with the letter `a` (assuming it is always lower case);  that is, we identify the elements in set\ $B$.
To do this, we first need to know how long the name of each country is:

```{r}
# Find the length of the name of each country in the list:
# nchar() finds the length of each country name
Length_Country_Names <- nchar(list_countries()) 
```

Then we can proceed similar to having `a` as the second letter of the name:

```{r}
# Find where, in the list of countries, a country name has an A in last position:
locate_Ends_With_a <- which( substr(list_countries(),
                                    start = Length_Country_Names,
                                    stop = Length_Country_Names)
                            == "a")

# Now find the countries listed at those locations:
Ends_With_a <- list_countries()[locate_Ends_With_a]

# How many countries are there whose names end with an `a`?
length(Ends_With_a)

```

Now we can find the countries whose names have an `a` as the second and as the last letters by finding the intersection between the two sets:

```{r}
a_Second_And_Last <- intersect(a_Second, 
                               Ends_With_a)

# How many countries are there whose names have
# `a` as second and last letters?
length(a_Second_And_Last)

# What are these countries?
a_Second_And_Last
```



```{r echo=FALSE, eval=FALSE}
### This is the R help from `?sets` in R:
(x <- c(sort(sample(1:20, 9)), NA))
(y <- c(sort(sample(3:23, 7)), NA))
union(x, y)
intersect(x, y)
setdiff(x, y)
setdiff(y, x)
setequal(x, y)

## True for all possible x & y :
setequal( union(x, y),
          c(setdiff(x, y), intersect(x, y), setdiff(y, x)))

is.element(x, y) # length 10
is.element(y, x) # length  8

## Factors:
x <- as.factor(c("A", "B", "A"))
y <- as.factor(c("B", "b"))
union(x, y)
intersect(x, y)
setdiff(x, y)
setdiff(y, x)
## (Note that union() and intersect() merge the levels.)
```



## Exercises {#SetsExercises}

Selected answers appear in Sect.\ \@ref(AnswersChapSetTheory).



:::{.exercise #AlgebraProof1}
Prove one of the idempotent laws in Sect.\ \@ref(SetAlgebra).
:::


:::{.exercise #AlgebraProof2}
Prove one of the dominance laws in Sect.\ \@ref(SetAlgebra).
:::


:::{.exercise #ImaginaryNumbers}
Consider the set of complex numbers $\mathbb{C}$ written in the form $a + bi$.

1. Use set-builder notation to define the set of complex numbers in terms of the set of real numbers\ $\mathbb{R}$.
2. Adapt your notation to define the set of purely imaginary numbers\ $\mathbb{I}$.
3. How would you define the complex numbers, when written in polar form: $r \exp(i\theta)$?
:::



:::{.exercise #SetsQuadratic}
Consider the general quadratic equation $ax^2 + bx + c = 0$ for real values of \ $a\ne 0$, $b$ and\ $c$. 
Using set-builder notation, define: 

1. the universal set\ $U$ in terms of\ $a$, $b$ and\ $c$, which defines all quadratics.
2. the set\ $R$ in terms of\ $a$, $b$ and\ $c$ that contains all quadratics with two equal real roots.
3. the set\ $Z$ in terms of\ $a$, $b$ and\ $c$ that contains all quadratics with no real roots.
4. the set\ $S$, in terms of\ $U$, $R$ and\ $Z$, that contains all quadratics with exactly one real root.
:::


:::{.exercise #SetsLogFunctions}
Consider the equation $y = \log(x^2 - 1)$ for real values of\ $y$.
Using set-builder notation, define: 

1. the set of values of\ $x$ for which\ $y$ has real values.
2. the set of values of\ $x$ for which $y > 1$.
3. the set of values of\ $x$ for which\ $y$ has the value zero.
:::


:::{.exercise #SetsCards}
Define these sets:
\begin{align*}
  S &= \{\clubsuit, \spadesuit\};\\
  P &= \{ \text{Jack}, \text{Queen}, \text{King}, \text{Ace}\}\quad \text{(i.e., the picture cards)},
\end{align*}
where\ $U$ is the universal set of all cards in a pack.
Using set-builder notation, define:

1. the set\ $B$ that contains the black picture cards.
1. the set\ $R$ that contains the red picture cards.
1. the set\ $N$ that contains the red non-picture cards.
:::


:::{.exercise #Rbasics1}
Define a vector\ $A$ as the even numbers from\ $2$ to\ $20$, and vector\ $B$ as all integers from\ $1$ to\ $10$. 
Write **R** code to find:

1. the intersection of\ $A$ and\ $B$.
2. the elements in\ $A$ that are *not* in\ $B$.
:::


:::{.exercise #Rbasics2}
Define a vector\ $N$ as the odd numbers from\ $1$ to\ $51$, and vector\ $M$ as all triangular numbers between\ $1$ and\ $51$. (Triangular numbers take the form $n(n + 1)/2$ for integer\ $n > 0$.)
Write **R** code to find:

1. all the triangular numbers between\ $1$ and\ $51$ inclusive, and hence the elements of set\ $M$.
2. the intersection of\ $N$ and\ $M$.
3. the elements in\ $N - M$, and explain what this means.
4. the elements in\ $M - N$, and explain what this means.
:::

:::{.exercise #SetBuilder1}
Using set-builder notation, define set\ $T$ as the values of\ $x$ for which $y = 1/x$ is a real value.
:::


:::{.exercise #SetBuilder2}
Using set-builder notation, define set\ $L$ as the values of\ $x$ for which $y = \tan x$ is a real value.
:::


:::{.exercise #USStates}
In **R**, the names of all US states are contains in the built-in dataset `state.name`.
Using this dataset, use **R** to:

1. find the set of US states who names begin with\ `W`.
1. find the set of US states who names begin with `North`.
1. find the set of US states who names end in `a`.
1. find the set of US states who names end in `a`, **or** start with\ `W`.
1. find the set of US states who names end in `a`, **and** start with\ `W`.
1. find the set of US states who names end in `a`, **and** do not start with\ `W`.
:::

:::{.exercise #USStatesData}
In **R**, information about each US states in (mostly) 1977 is contained in the built-in dataset `state.x77`.
Using this dataset, use **R** to:

1. find the set of US states whose average life expectancy exceeds\ $70$.
1. find the set of US states whose area is less than\ $500\,000$ square miles.
1. find the set of US states whose illiteracy rate is greater than\ $2$%.
1. find the set of US states whose percent of high-school graduates is less than\ $50$%.
1. find the set of US states whose illiteracy rate is greater than\ $2$%, **and** whose percent of high-school graduates is less than\ $50$%.
1. find the set of US states whose illiteracy rate is greater than\ $2$%, **or** whose percent of high-school graduates is less than\ $50$%.

(To view the help for this data, type `?state.x77` in **R**.)
:::


:::{.exercise #SetBuilderReading1}
Explain what values are contained in set\ $G$, if it is defined as
$$
   G = \{ x\in\mathbb{R} \mid x^2 < 4\}
$$
What is the cardinality of set\ $G$?
:::


:::{.exercise #SetBuilderReading2}
Explain what values are contained in set\ $B$, if it is defined as
$$
   B = \{x\in\mathbb{R} \mid \lfloor x \rfloor = x\}.
$$
In this expression, $\lfloor x \rfloor$ is the *floor* function\index{Floor function}, that rounds the value of\ $x$ to the next lowest integer towards $\-infty$ (for example, $\lfloor \pi \rfloor = 3$, $\lfloor e\rfloor = 2$, $\lfloor -\pi \rfloor = -4$ and $\lfloor -e\rfloor = -3$).

What is the cardinality of set\ $B$?
:::


:::{.exercise #SetAlgebra1}
Show that $A\setminus(A\cap B) = A\cap B^c$.
:::


:::{.exercise #SetAlgebra2}
Show that $(A\setminus B)^c = A^c\cup B^c$.
:::


:::{.exercise #SetAlgebra3}
Show that $(A\cup B) \cap (A\cup B^c) \cap B = A\cap B$.
:::

:::{.exercise #SetAlgebra4}
Show that $(A\cap B) \cup (A\cap B^c) = A$.
:::


:::{.exercise #SetAlgebra5}
Show that $A\Delta B = A^c\Delta B^c$.
:::

:::{.exercise #SetAlgebra6}
Show that $(A\Delta B) = (A\cup B) \setminus (A\cap B)$.
:::


:::{.exercise #TwoCoinsTosses}
Suppose the set $C$ is defined as
`r if (knitr::is_latex_output()) {
   '$C = \\{ \\Heads, \\Tails\\}$,'
} else {
   '<em>C</em> = {<span class="HTcircle">H</span>, <span class="HTcircle">T</span>}'
}`
the set of results from tossing a coin.
Define the set\ $D$ that shows the results of *two* tosses of a coin, as an ordered pair.
(See Example\ \@ref(exm:TwoDSets).)
:::



:::{.exercise #CardsInPack}
Define the following two sets:
$$
   S = \{\spadesuit, \clubsuit, \diamondsuit, \heartsuit\}
   \qquad\text{and}\qquad
   D = \{2, 3, 4, \dots, \text{Jack}, \text{Queen}, \text{King}, \text{Ace}\}.
$$
Use these two sets to define Set\ $C$ of 'all cards in a standard pack', as an ordered pair.
(See Example\ \@ref(exm:TwoDSets).)
:::


:::{.exercise #ThreeViruses}
A hospital tests patients for three respiratory viruses: Influence (the flu) ($F$), COVID ($C$), and RSV ($R$).
A patient is classified as 'High Risk' if they have at least two of these viruses simultaneously. 

Write the set expression for 'High Risk' patient.
:::


:::{.exercise #ParticleDetector}
Imagine a circular particle detector. 
Let\ $S$ be the entire area of the detector. 
Let\ $D_r$ be the set of points $(x, y)$ such that $x^2 + y^2 \leq r^2$ (i.e., a disk of radius $r$). 

If a particle is detected in the 'outer ring', between radius\ $5$ and radius\ $10$, express this region using $D_5$ and $D_{10}$ using set notation.
:::

