# Mixed distributions {#ChapterMixedDistributions}

::: {.objectivesBox .objectives data-latex="{iconmonstr-target-4-240.png}"}
On completion of this chapter, you should be able to:

* recognise mixed random variables.
* recognise hurdle models and compound Poisson--gamma models for modelling mixed random variables.
* know the basic properties of the above approaches.
* apply these approaches as appropriate to problem solving.
:::


## Introduction

Mixed random variables commonly occur in practice, in diverse applications such as insurance, agriculture, climatology, fisheries, etc.
In most practical cases, a mixed random variable\ $Y$ has a discrete probability mass at $Y = 0$, and is also continuous for $Y > 0$.
Example include:

* Modelling insurance: some portfolios will have zero claims; however, when claims are made, the total claim amount has a positive continuous distribution.
* Modelling rainfall: some months may receive exactly zero rainfall; otherwise, the monthly rainfall has a continuous distribution.
* Modelling fish catch: some trawls will catch zero fish; other trawls, however, will catch a positive continuous mass of fish.

Continuous data with a discrete probability at zero can be modelled in various ways, depending on how the discrete probability at $X = 0$ is modelled
Two approaches are considered in this chapter.

*Hurdle models*\index{Hurdle models} (Sect.\ \@ref(HurdleModels)) treat the zeros as emerging from a *two-step process*.
Step one is a binary process, that models whether the event of interest (i.e., an insurance claim is made; rainfall is recorded; fish are caught) occurs.
Step two is a model for the continuous data of interest, conditional on the event of interest occurring.

Compound *Poisson-gamma models*\index{Compound Poisson-gamma models} (Sect.\ \@ref(TweedieModels)) generates zeros from a *single process*, when a Poisson count of underlying events equals zero.


## Hurdle models {#HurdleModels}

### Definitions {#HurdleModelsDefinitions}

\index{Hurdle models|(}
A hurdle model treats zeros and positive values as arising from two separate processes. 
The first process is a binary model that determines whether an event occurs or not. 
The second process that describes the distribution of values, conditional on the values being positive. 
The hurdle model is useful when the zero-generating process is believed to be distinct from the process driving variation among positive values; for example, whether a household buys alcohol is a different process than how much a household spends on alcohol if they do buy alcohol.


A *hurdle model*\index{Hurdle models} treats the zero values of the random variable\ $X$ as emerging from a two-step process.
The first step uses a Bernoulli distribution (Sect.\ \@ref(BernoulliDistribution))\index{Bernoulli distribution} to model a random variable\ $Y$ that defines whether an event of interest occurs:
$$
  \Pr(Y) = 
  \begin{cases}
    1 - p    & \text{if $Y = 0$; i.e., the event of interest does not occur;}\\
    p        & \text{if $Y = 1$; i.e., the event of interest does occur.}
  \end{cases}
$$
If the event does occur (i.e., conditional on $X > 0$), with probability $1 - p$, then the continuous component is modelled using a probability density function $f^+_X(x)$ defined for positive real values only (such as an exponential distribution or a gamma distribution).

The probability function for the mixed random variable\ $X$ is therefore
$$
   f_X(x) = 
   \begin{cases} 
      1 - p           & \text{if $x = 0$}\\
      p\cdot f^+_X(x)    & \text{if $x > 0$,}
   \end{cases}
$$
where $f^+_X(x) = f_{X \mid X > 0}(x)$ is the density function for the continuous distribution defined for $x\in\mathbb{R}_{+}$.
The distribution function is
$$
   F_X(x) = 
   \begin{cases} 
      0                             & \text{if $x < 0$}\\
      1 - p                         & \text{if $x = 0$}\\
      (1 - p) + p\cdot F^+_X(x)    & \text{if $x > 0$,}
   \end{cases}
$$
where $F^+_X(x) = F_{X\mid X > 0}(x)$ is the distribution function for the continuous distribution defined for $x\in\mathbb{R}_{+}$.


:::{.example #HurdleExample name="Hurdle model"}
Consider the mixed random variable\ $X$, with probability function $f_X(x)$.
Suppose that, when $X > 0$, an exponential distribution\index{Exponential distribution} is used to describe the distribution; that is
$$
  f^+_X(x) = \frac{1}{\lambda}\exp(-x/\lambda)\quad\text{for $x > 0$}.
$$
Then, using $p = 0.7$ and $\lambda = 1/2$, the probability function of\ $X$ is
$$
  f_X(x)
  =
  \begin{cases}
    0.3                   & \text{if $x = 0$}\\
    0.35\cdot \exp(-x/2)  & \text{if $x > 0$;}
  \end{cases}
$$
Similarly, the distribution function of\ $X$ is
$$
  F_X(x)
  =
  \begin{cases}
    0                                     & \text{if $ x < 0$}\\ 
    0.3                                   & \text{if $x = 0$}\\
    0.7\cdot \left(1 - \exp(-x/2)\right)  & \text{if $x > 0$;}
  \end{cases}
$$
The probability function and distribution function are shown in Fig.\ \@ref(fig:HurdlePic).
:::


```{r HurdlePic, echo=FALSE, fig.align="center", fig.cap="A hurdle model, showing an exponential distribution for the continuous component.", fig.width=8, fig.height=4, out.width="90%"}
p <- 0.7
x.rate = 2

x <- seq(0, 3,
         length = 100)
y <-  p * dexp(x, 
               rate = x.rate)
                     

par( mfrow = c(1, 2))
plot(y ~ x,
     type = "l",
     axes = FALSE,
     ylim = c(0,
              max( c(y, 1 - p) ) ),
     xlab = expression(italic(X)),
     ylab = expression(Prob.~"function"),
     main = expression(bold(atop(Mixed~distribution,
                                 hurdle~ model))),
     lwd = 3)
abline(h = 0,
       col = "grey")
axis(side = 1)
axis(side = 2,
     at = c(0, 
            1 - p),
     las = 2)
box()

points(x = 0, # Included
       y = 1 - p,
       pch = 19)
points(x = 0, # Excluded
       y = max(y),
       pch = 1)

# Explanatory labels
arrows(x0 = 0.35,
       x1 = 0,
       y0 = 0.17,
       y1 = 1 - p,
       length = 0.15,
       angle = 15)
text(x = 0.35,
     y = 0.17,
     pos = 1,
     label = expression(Pr*"("*italic(X)==0*")" ) )


arrows(x0 = 1.5,
       x1 = 1,
       y0 = 0.8,
       y1 = dexp(x = 1, 
                 rate = x.rate) * p,
       length = 0.15,
       angle = 15)
text(x = 1.5,
     y = 0.8,
     pos = 3,
     label = expression( Exponential~distribution) )


#############################################

Fy <- (1 - p)  + p * pexp( x,
                           rate = x.rate)
plot(Fy ~ x,
     type = "l",
     axes = FALSE,
     ylim = c(0, 1),
     xlim = c(-1, max(x)),
     xlab = expression(italic(X)),
     ylab = expression(Distribution.~"function"),
     main = expression(nold(atop(Mixed~distribution,
                                 hurdle~ model))),
     lwd = 3)
abline(h = 0,
       col = "grey")
lines(x = c(-1, 0),
      y = c(0, 0),
      lwd = 2)
points(x = 0, 
       y = 0,
       pch = 1) # Not incuded


axis(side = 1)
axis(side = 2,
     at = c(0, 
            1 - p, 
            1),
     las = 2)
box()

points(x = 0, # Included
       y = 1 - p,
       pch = 19)

```




### Properties {#HurdleModelsProperties}

The expected value of the random variable\ $X$ can be found first separating the discrete and continuous parts of the distribution:
\begin{align*}
  \operatorname{E}[X] 
  &= \operatorname{E}[X\mid X = 0] + \operatorname{E}[X\mid X > 0]\\
  &= (1 - p)\times 0 + p \operatorname{E}[X\mid X > 0]\\
  &= p \mu^+,
\end{align*}
where $\mu^+$ denotes $\operatorname{E}[X\mid X > 0]$, the expected value of the distribution defined for $X > 0$.
Similarly,
$$
  \operatorname{E}[X^2] 
  = \operatorname{E}[X^2 \mid X > 0]
  = p (\mu^2)^+,
$$
where $(\mu^2)^+$ denotes $\operatorname{E}[X^2\mid X > 0]$.
Then, the variance of\ $X$ is
\begin{align*}
  \operatorname{var}[X]
  &=  p[(\mu^2)^+ - p(\mu^+)^2]\\
  &=  p(\sigma^2)^+ + p(1 - p)(\mu^+)^2.
\end{align*}
where $\sigma^2 = \operatorname{var}[X \mid X > 0]$
The same approach also gives the MGF as
\begin{align*}
  M_X(t) 
  &= (1 - p) + p\cdot M_{X^+}(t)\\
  &= (1 - p) + p\cdot M_{X\mid X > 0}(t),
\end{align*}
where $M_{X^+}(t)$ is the MGF of the distribution defined for $X > 0$.

Clearly, further progress with these expressions requires knowing the distribution of\ $X | X > 0$



:::{.example #HurdleGamma name="Hurdle model, using gamma distribution"}
Consider a random variable\ $X$ such that $p = 0.75$ (and hence $\Pr(X = 0) = 1 - p = 0.25$), and where the distribution when $X > 0$ follows a gamma distribution (Sect.\ \@ref(GammaDistribution)) having $\alpha = 2$ and $\beta = 1$, with the density function
$$
  f_X^{+}(x; \alpha, \beta) 
  = f_{X\mid X > 0}(x; \alpha, \beta) 
  = \frac{1}{\Gamma(2)} x \exp(-x)
  \quad\text{for $x > 0$}.
$$
For this gamma distribution, using results from Sect.\ \@ref(GammaDistribution), $\mu^+ = \operatorname{E}[X\mid X > 0] = \alpha\beta = 2$ and $\operatorname{var}[X\mid X > 0] = \alpha\beta^2 = 2$.
Then, the probability function for\ $X$ is
$$
  f_{X}(x) 
  = 
  \begin{cases}
     0.25                                & \text{for $x = 0$}\\
     0.75 \times f_X^{+}(x; \alpha, \beta) & \text{for $x > 0$.}
  \end{cases}
$$
Then 
\begin{align*}
  \operatorname{E}[X] 
   &= p\cdot \mu^+ 
    = 0.75 \times 2 
    = 1.5\\
  \operatorname{var}[X] 
   &= p(\sigma^2)^+ + p(1 - p)(\mu^+)^2
   = 0.75\times 2 + 0.75\times 0.25\times 1.5^2
   \approx 1.921.
\end{align*}
The probability function and distribution function are shown in Fig.\ \@ref(fig:HurdleGamma).
:::



```{r HurdleGamma, echo=FALSE, fig.align="center", fig.cap="A hurdle model, using a gamma distribution for the continuous component.", fig.width=8, fig.height=4, out.width="90%"}
p <- 0.75
x.shape <- 2 # shape = alpha
x.scale <- 1 # scale = beta

x <- seq(0, 5,
         length = 100)
y <-  p * dgamma(x, 
                 shape = x.shape,
                 scale = x.scale)
                     

par( mfrow = c(1, 2))
plot(y ~ x,
     type = "l",
     axes = FALSE,
     ylim = c(0,
              max( c(y, 1 - p) ) ),
     xlab = expression(italic(X)),
     ylab = expression(Prob.~"function"),
     main = expression(atop(Hurdle~model,
                            probability~"function")),
     lwd = 2)
abline(h = 0,
       col = "grey")
axis(side = 1)
axis(side = 2,
     at = c(0, 
            1 - p),
     las = 2)
box()

points(x = 0, # Included
       y = 1 - p,
       pch = 19)
points(x = 0, # Excluded
       y = y[1],
       pch = 1)



#############################################

Fy <- (1 - p)  + p * pgamma(x, 
                            shape = x.shape, 
                            scale = x.scale)
plot(Fy ~ x,
     type = "l",
     axes = FALSE,
     ylim = c(0, 1),
     xlim = c(-1, max(x)),
     xlab = expression(italic(X)),
     ylab = expression(Distribution.~"function"),
     main = expression(atop(Hurdle~model,
                            distribution~"function")),
     lwd = 2)
abline(h = 0,
       col = "grey")
lines(x = c(-1, 0),
      y = c(0, 0),
      lwd = 2)
points(x = 0, 
       y = 0,
       pch = 1) # Not incuded


axis(side = 1)
axis(side = 2,
     at = c(0, 
            1 - p, 
            1),
     las = 2)
box()

points(x = 0, # Included
       y = 1 - p,
       pch = 19)
```

\index{Hurdle models|)}


## Compound Poisson-gamma distributions {#TweedieModels}
\index{Compound Poisson-gamma distributions|(}

### Definition

A compound Poisson--gamma model generates the point mass at zero through a physically-motivated compound process. 
The model assumes that events occur according to a Poisson process, and each event contributes an independent gamma-distributed amount to the total.
The observed variable\ $Y$ is the sum of all the independent contributions. 
A compound Poisson--gamma distributions, therefore, is a Poisson sum of independent gamma distributions.

When the Poisson count is zero (i.e., when no events occur), the total is exactly zero, naturally producing the point mass without any separate modelling of zeros. 
When at least one event occurs, the sum of gamma random variables gives a continuous positive value. 
The compound Poisson-gamma distribution is a unified single-distribution model, and requires the compound Poisson-gamma story to be a plausible description of the data-generating process. 

More specifically, suppose the number of events observed (a discrete random variable) is\ $N$, such that
\begin{equation}
  N \sim \text{Pois}(\lambda).
   (\#eq:PGp0)
\end{equation}
That is, an event occurs\ $N$ times, where\ $N$ follows a Poisson distribution.
Given that\ $N$ has a Poisson distribution, it follows (from Sect.\ \@ref(PoissonDistribution)) that
$$
  \Pr(N = 0) = \exp(-\lambda)
$$
is the probability that zero events are observed (i.e., $N = 0$).
If, however, $N > 0$, then for each $i = 1, 2, \dots, N$, a continuous random variable\ $Y_i$ is observed, such that
$$
   Y_i \sim \text{Gam}(\alpha_i, \beta).
$$
If $N = 0$, then no events $Y_i$ are observed at all; however, if $N > 0$ then\ $N$ events are observed and
$$
  X = \sum_{i = 1}^N Y_i
$$
has a continuous distribution.
Then, 
$$
  X = 
  \begin{cases}
    0                   & \text{if $N = 0$};\\
    \sum_{i = 1}^N Y_i  & \text{if $N > 0$}.
  \end{cases}
$$
By this definition $X$ has a *compound Poisson-gamma distribution* (which are special cases of Tweedie distributions\index{Tweedie distributions}).
The distribution of\ $X$ has a probability mass at $X = 0$ where $\Pr(Y = 0) = \Pr(N = 0) = \exp(-\lambda)$.


:::{.example #CPGRain name="Rainfall modelling"}
Consider monthly rainfall\ $X$ at a given location.
In any given month, $N$ rainfall events may occur (and for some months, $N$ may be zero) such that
$$
  N \sim \text{Poisson}(\lambda = 2.3).
$$
So, for example, that probability that a given month has no rainfall is
$$
  \Pr(N = 0) = \exp(-2.3) = 0.10,
$$
or about $10$%.

When a rainfall event does occur, the amount of rainfall is described by a gamma distribution
$$
  Y_i \sim \text{Gam}(\alpha_i, 2),
$$
where $\alpha_i$ may be different for each month~$i$.

Suppose $\alpha_1 = 1.75$ in January and $\alpha_6 = 0.3$ in June.
The Poisson distribution remains the same for both months (Fig.\ \@ref(fig:CPGrain), left panel), meaning that the distribution of the *number* of rainfall events is the same in January and in June.
However, since the value of $\alpha$ is different for each month, the distributions of the amount of rainfall in each rainfall event changes (Fig.\ \@ref(fig:CPGrain), center panel).
The distribution of *total* rainfall for the January and June rainfall using this model are shown in Fig.\ \@ref(fig:CPGrain) (right panel).
:::



```{r CPGrain, echo=FALSE, fig.align="center", fig.cap="A compound Poisson-gamma model for rainfall. Left: the distribution of the number of rainfall events per month. Center: the distribution of rainfall per event for each month. Right: the distribution of the total rainfall for each month.", fig.width=7, fig.height=3.25, out.width="100%"}
# Underlying parameters
lambda <- 2.3
alpha_January <- 1.75
alpha_June <- 0.9
beta <- 2



par(mfrow = c(1, 3))

x <- 0:10
dRainfallEvents <- dpois(x, 
                         lambda = lambda)

# Poisson: number of events
plotDiscrete(x, 
             dRainfallEvents,
             type = "MF",
             main = "The number of rainfall\nevents per month",
             xlab = expression(italic(N)),
             ylab = "PMF")

# Two gammas: amount per event
curve( dgamma(x, shape = alpha_January, scale = beta),
       from = 0, to = 14,
       las = 1,
       ylim = c(0, 0.25),
       main = "Rainfall amount\nper rainfall event",
       ylab = "PDF",
       xlab = expression(Rainfall~per~event~italic(Y)),
       lwd = 2)
curve( dgamma(x, shape = alpha_June, scale = beta),
       from = 0, to = 14,
       lwd = 2,
       add = TRUE,
       lty = 3)
legend("topright",
       lty = c(1, 3),
       lwd = 2,
       bty = "n",
       legend=c("January",
                "June") )


# Tweedie: Total rain
xi_January  <- (alpha_January + 2) / (alpha_January + 1)
xi_June     <- (alpha_June    + 2) / (alpha_June    + 1)

phi_January <- lambda^(1 - xi_January) * (alpha_January * beta)^(2 - xi_January)
phi_June    <- lambda^(1 - xi_June)    * (alpha_June    * beta)^(2 - xi_June)

mu_January <- lambda * alpha_January * beta
mu_June    <- lambda * alpha_June * beta

x <- seq(0, 30, 
         length = 100)
tweedie::tweedie_plot(x,
                      xi = xi_January,
                      mu = mu_January,
                      phi = phi_January,
                      las = 1,
                      #pch = 1,
                      ylim = c(0, 0.16),
                      xlab = expression(Total~rain~italic(X)),
                      ylab = "Prob. function",
                      main = "Total monthly rainfall",
                      lwd = 2)
tweedie::tweedie_plot(x,
                      xi = xi_June,
                      mu = mu_June,
                      phi = phi_June,
                      add = TRUE,
                      #pch = 19,
                      lwd = 2,
                      lty = 3)
legend("topright",
       lty = c(1, 3),
       pch = c(1, 19),
       lwd = 1,
       bty = "n",
       legend=c("January",
                "June") )
```


### Properties

The probability functions for the compound Poisson-gamma distribution *cannot be written in closed form*.
However, the probability function can be expressed as an infinite sum [@dunn2005series] or by inverting the MGF [@mypapers:Dunn:fourier], following the ideas in Sect.\ \@ref(DistributionFromMGF).
Likewise, the distribution function has no closed form and must be evaluated using numerical methods [@mypapers:Dunn:DF].

The probability function as an infinite series can be determined from the development of the compound Poisson-gamma distribution above.
If $N = 1$ with $\Pr(N = 1)$ (from the Poisson distribution), then, $X = Y$ has the $\text{Gam}(\alpha, \beta)$ distribution.
That is,
$$
  f_{X\mid N = 1}(x) 
  = \frac{\beta^{\alpha}}{\Gamma(\alpha)} x^{\alpha -1}\exp(-\beta x) \quad \text{for $x > 0$}.
$$

If $N = 2$ with $\Pr(N = 2)$ (from the Poisson distribution), then, $X = Y_1 + Y_2$, where $Y_i \sim\text{Gam}(\alpha, \beta)$ for $i = 1, 2$.
This means that $X \sim \text{Gam}(2\alpha, \beta)$ (Exercise\ \@ref(exr:GammaSum); Sect.\ \@ref(GammaProperties)).
That is, 
$$
  f_{X\mid N = 2} (x) 
  = \frac{\beta^{2\alpha}}{\Gamma(2\alpha)} x^{2\alpha -1}\exp(-\beta x) \quad \text{for $x > 0$}.
$$

More generally, if $N = n$ with $\Pr(N = n)$ (from the Poisson distribution), then, $X = Y_1 + Y_2+\cdots + Y_n$, and so $X \sim\text{Gam}(n\alpha, \beta)$, so that
$$
  f_{X\mid N = n} (x) 
  = \frac{\beta^{n\alpha}}{\Gamma(n\alpha)} x^{n\alpha -1}\exp(-\beta x) \quad \text{for $x > 0$}.
$$

Then, using the *Law of total probability* (Theorem\ \@ref(thm:TotalProb)),
\begin{align*}
  f_X(x)
  &= \sum_{n = 1}^\infty \Pr(N = n) \cdot F_{X\mid N = n}(x)\\
  &= \sum_{n = 1}^\infty \frac{\exp(-\lambda)\lambda^n}{n!} \cdot \frac{\beta^{n\alpha}}{\Gamma(n\alpha)} x^{n\alpha -1}\exp(-\beta x)\\
  &= \exp(-\lambda - \beta x) \sum_{n = 1}^\infty \frac{\lambda^n}{n!} \frac{\beta^{n\alpha}}{\Gamma(n\alpha)} x^{n\alpha - 1} 
\end{align*}
for $x > 0$.
In addition, of course, we have that
$$
  \Pr(X = 0) = \exp(-\lambda).
$$
  

Even though the probability function has no closed form, the MGF is reasonably simple.
To find the MGF, first consider the value of\ $N$ as fixed; then, $X = Y_1 + Y_2 + \cdots + Y_N$ the MGF of\ $X$ is 
$$
  M_{X\mid N}(t) = \operatorname{E}[\exp(tX)\mid N].
$$
Since the MGF of a sum of independent random variables is the product of their MGFs (Theorem\ \@ref(thm:MGFIndependent)), we can assume that $N$ takes some specific value\ $n$ and write
\begin{align*}
  M_{X\mid N}(t) 
  &= \operatorname{E}\left[ \exp\big(t(Y_1 + Y_2 + \cdots + Y_n)\big)\right] \\
  &= \prod_{i = 1}^n \operatorname{E}[\exp(t Y_i)] \\
  &= \left( M_Y(t)\right)^n
\end{align*}
So the MGF of $X$ (rather than $X\mid N$ as above) is
\begin{align*}
  M_X(t) 
  &= \operatorname{E}\left[ \operatorname{E}[ \exp(tX) \mid N]\right] \\
  &= \operatorname{E}\left[ M_{X\mid N}(t)\right] \\
  &= \operatorname{E}\left[ M_Y(t)^N\right].
\end{align*}
Substituting the MGF of a $\text{Gamma}(\alpha, \beta)$ random variable, $M_Y(t) = \left(\frac{\beta}{\beta - t}\right)^{\!\alpha}$, and then applying the MGF of a $\text{Poisson}(\lambda)$ random variable, $M_N(s) = \exp\{\lambda(s - 1)\}$ with $s = M_Y(t)$, gives
$$
  M_X(t)=\exp\!\left\{\lambda\left[\Big(\frac{\beta}{\beta-t}\Big)^{\!\alpha}-1\right]\right\},\qquad t<\beta.
$$

From the MGF we can deduce
\begin{align*}
  \operatorname{E}[X] 
  &= \lambda\alpha/\beta\\
  \operatorname{var}[X]
  &= \lambda\alpha(\alpha + 1)/\beta^2.
\end{align*}


:::{.example #PoissonGammaExample name="Poisson--gamma distribution"}
Suppose 
$$
  N\sim\text{Pois}(\lambda = 3)
  \quad\text{and}\quad
  Y_i \sim\text{Gam}(\alpha = 1/2, \beta = 2/3)
$$
(where $\beta$ is the scale parameter).
The mean and variance of\ $X$ is
\begin{align*}
  \operatorname{E}[X]
  &= \lambda\alpha/\beta
   = 3\times \frac{1}{2}\times\frac{2}{3} = 1;\\
  \operatorname{var}[X] 
  &= \lambda\alpha(\alpha + 1)\beta^2
   = 3\times \left(\frac{1}{2}\times \frac{3}{2}\right)\left(\frac{5}{3}\right)^2 = 1.
\end{align*}
Furthermore, from Equation\ \@ref(eq:PGp0),
$$
  \Pr(X = 0) = \exp(-\lambda) = 0.0498.
$$
This Poisson-gamma distribution is shown in Fig.\ \@ref(fig:PoissonGammaExamples) (right panel).
:::


<!-- The parameters in the (alpha, beta, lambda) space:
\alpha = (2 - \xi)/(\xi - 1) [Not a function of (phi, mu)]
\beta = \phi(\xi - 1) / \mu^{\xi - 1}
\lambda = \mu^(2-\xi) / ( \phi (2-\xi))
So for instance:
alpha = 12 => xi = 5/3
alpha = 20 => xi = 22/21
-->


```{r PoissonGammaExamples, echo=FALSE, fig.align="center", fig.cap="Two compound Poisson--gamma models, both with mean and varaince set to $1$. The solid dot at $X = 0$ represents the discrete probability.", fig.width=8, fig.height=4, out.width="90%"}
par( mfrow = c(1, 2))

x <- seq(from = 0,
         to =   5,
         length = 1000)

xi <- 22/21
mu <- 1
phi <- 1

alpha <- (2 - xi)/(xi - 1)
lambda <- mu^(2 - xi) / ( phi * (2 - xi))
beta <- phi * (xi - 1) / mu^(xi - 1)
p0 <- exp(-lambda)

# cat("alpha, beta, lambda, p0\n")
# cat(alpha, beta, lambda, p0, "\n")

y <-  tweedie::dtweedie(y = x,
               xi = xi,
               mu = mu,
               phi = phi)

plot(y[x > 0] ~ x[x > 0],
     type = "l",
     ylim = c(0, max(y)), 
     axes = FALSE,
     xlab = expression(italic(x)),
     ylab = expression(Prob.~"function"),
     main = expression(atop(Poisson*"-gamma"~distribution,
                            with~alpha==20)),
     lwd = 2)
abline(h = 0,
       col = "grey")
axis(side = 1)
axis(side = 2,
     at = p0,
     las = 1,
     labels = round(p0, 3))

points(x = 0,
       y = p0,
       pch = 19)
box()

#tweedie.convert(xi = xi,
#                mu = mu, 
#                phi = phi)


############################ 

xi <- 5/3 # ALPHA = 0.5
alpha <- 0.5

mu <- 1
phi <- 1

alpha <- (2 - xi)/(xi - 1)
lambda <- mu^(2 - xi) / ( phi * (2 - xi))
beta <- phi * (xi - 1) / mu^(xi - 1)
p0 <- exp(-lambda)

# cat(alpha, beta, lambda, p0, "\n")


y <- dtweedie(y = x,
              xi = xi,
              mu = mu,
              phi = phi)


plot(y[x > 0] ~ x[x > 0],
     type = "l",
     ylim = c(0, max( c(y, p0))), 
     axes = FALSE,
     xlab = expression(italic(x)),
     ylab = expression(Prob.~"function"),
     main = expression(atop(Poisson*"-gamma"~distribution,
                            with~alpha==0.5)),
     lwd = 2)
abline(h = 0,
       col = "grey")
axis(side = 1)
axis(side = 2,
     at = p0,
     las = 1,
     labels = round(p0, 3))
box()


points(x = 0,
       y = p0,
       pch = 19)
box()

#tweedie.convert(xi = xi,
#                mu = mu, 
#                phi = phi)

```

\index{Compound Poisson-gamma distributions|)}


## Summary of models

The zeros in a compound Poisson--gamma distributions are a natural outcome of the same process used to generate the non-zero values; a zero occurs when the Poisson count of contributions is zero.
If a physical argument can be provided that the observed value is the sum of a random number of independent gamma-distributed contributions, a compound Poisson--gamma distribution may be an appropriate model. 
In contrast, a hurdle model treats zeros and positive values as structurally distinct, potentially generated by different processes: a separate model governs whether an observation is zero, and another governs an observation's magnitude given it is positive.


```{r echo=FALSE}
MixedTable <- array( dim=c(5, 2) )

colnames(MixedTable) <- c("Hurdle",
                          "Poisson-Gamma")

rownames(MixedTable) <- c("Zeros",
                          "Parameters",
                          "Interpretation",
                          "Flexibility",
                          "Uses")
MixedTable[1, ] <- c("Different process",
                     "No events occurred")
MixedTable[2, ] <- c("Two separate models",
                     "One unified model")
MixedTable[3, ] <- c("Very natural",
                     "Requires physical story")
MixedTable[4, ] <- c("High",
                     "Moderate")
MixedTable[5, ] <- c("Healthcare spend",
                     "Insurance, rainfall")

if( knitr::is_latex_output() ) {
  knitr::kable(MixedTable,
               format = "latex",
               booktabs = TRUE,
               align = "c",
               escape = FALSE,
               longtable = FALSE,
               caption = "Comparing the hurdle and compound Poisson--gamma models.") %>%
    kable_styling(font_size = 10) %>%
    row_spec(0, bold = TRUE) 
}
if( knitr::is_html_output() ) {
  knitr::kable(MixedTable,
               format = "html",
               booktabs = TRUE,
               align = "c",
               longtable = FALSE,
               caption = "Comparing the hurdle and compound Poisson--gamma models.") %>%
    row_spec(0, bold = TRUE) 
}

```




:::{.example #InsuranceModelling name="Insurance modelling"}
Insurance companies have some days when zero claims are lodged, but most days record numerous claims, each of which are for positive, right-skewed amounts.

A compound Poisson--gamma model could be considered. 
The daily claim total is the aggregate of a random number of individual claim events (accidents or incidents), each generating a gamma-distributed payout amount.
A zero-claim amount occurs when no claim events arrive. 
The same process (i.e., policy-holders having incidents) generates both zeros and positive values.

A hurdle model could also be considered. 
First, a binary process determines whether any claim is filed (perhaps driven by seasonal risk, policy-holder behaviour, or reporting thresholds).
Conditional on a claim occurring, the total payout follows a gamma distribution. 
The zero (no-claim day) and positive values are structurally different; a no-claim day is not just 'zero incidents happened', but may reflect under-reporting, policy excesses, or deliberate non-filing.
:::



## Simulation {#MixedVariablesSimulation}

### General

A *hurdle* model could be used to model daily rainfall (e.g., @stern1984model), using a Bernoulli distribution to distinguish wet and dry days, and then a gamma distribution to model rainfall amounts on wet days.

 Fig.\ \@ref(fig:HurdleSim)
...


A compound Poisson--gamma model could also be used to model daily rainfall (e.g., @yunus2017modelling).




```{r HurdleSim, echo=FALSE, fig.align="center", fig.cap="A simulation of the hurdle model. FIX!!!!! The solid dot at $X = 0$ represents the discrete probability.", fig.width=8, fig.height=4, out.width="90%"}
num_Sims <- 1000
set.seed(109001) # For reproducibility

prob <- 0.7
# Bernoulli component
p <- rbinom(num_Sims,
            size = 1,
            prob = prob)
X <- array( dim = num_Sims)

for (i in 1:num_Sims){
  if ( p[i] == 0) {
    X[i] <- 0
  } else {
    X[i] <- rexp(1,
                 rate = 2)
  }
}

par(mfrow = c(1, 2))

barplot( table( p ),
         xlab = expression(italic(p)),
         ylab = "Counts",
         las = 1,
         names.arg = c("0: Event\ndoes not occur",
                       "1: Event\n does occur"),
         col = plotColour1,
         main = expression(atop(Distribution~of~italic(p)*":",
                                Prob.~of~event~occurring)) )

hist(X,
     freq = FALSE,
     las = 1,
     breaks = seq(0, max(X) + 0.5, 
                  by = 0.5),
     xlab = expression(italic(X)),
     ylab = expression(Density),
     main = expression(atop(Distribution~of~italic(X)*":",
                            Mixed~random~variable)) )
points(x = 0,
       y = length(X[X==0])/num_Sims,
       pch = 19)
  
```


:::{.example #SimulationPGmodels name="Simulations for compound Poisson--gamma models"}
Compound P--G; see Fig.\ \@ref(fig:PoissonGammaSim)
:::


```{r PoissonGammaSim, echo=FALSE, fig.align="center", fig.cap="A simulation of the compound Poisson--gamma models, both with mean and variance set to $1$. The solid dot at $X = 0$ represents the discrete probability.", fig.width=8, fig.height=4, out.width="90%"}

## SIMULATION
num_Sims <- 1000
set.seed(75101) # For reproducibility

# Set parameters
lambda <- 3
alpha <- 1/2
beta <- 2/3
  
N <- rpois(num_Sims,
           lambda = lambda)
X <- array( dim = num_Sims)

for (i in 1:num_Sims){
  if ( N[i] == 0) {
    X[i] <- 0 
  } else {
     X[i] <- sum( rgamma(N[i], 
                         shape = alpha,
                         scale = beta) )
  }
}

par(mfrow = c(1, 2))

barplot( table( N ),
         xlab = expression(italic(N)),
         ylab = "Counts",
         las = 1,
         col = plotColour1,
         main = expression(atop(Distribution~of~italic(N)*":",
                                Number~of~events)) )
         
hist(X[X > 0],
     freq = FALSE,
     las = 1,
     breaks = seq(0, max(X) + 0.5, 
                  by = 0.5),
     xlab = expression(italic(X)),
     ylab = expression(Density),
     main = expression(atop(Distribution~of~italic(X)*":",
                            Mixed~random~variable)) )
points(x = 0,
       y = length(X[X==0])/num_Sims,
       pch = 19)
  
```





### Rainfall

Simulation for the hurdle and compound Poisson--gamma models can be achieved by directly modelling the process by which the zeros are incorporated. 
To demonstrate, consider modelling monthly rainfall at a certain location for one specific month each year. 
Monthly rainfall is a mixed random variable: some months record exactly zero rainfall, while months in which rain falls have a continuous positive amount recorded.

Suppose the monthly rainfall (in mm)\ $R$ is modelled at a given location for a certain month, where about $30$% of months record zero rainfall; then. $0\Pr(R = 0) = 0.30$, and the mean monthly rainfall when rain is recorded is $\mu^+ = 50\,mms$.

For the *hurdle model*, set $p 0.30$ for the probability of no rainfall; then, the probability that rain is recorded is $1 - p = 0.70$. 
For the positive component, suppose first that rainfall amounts follow an exponential distribution with rate $\lambda$.
Since $\operatorname{E}[R \mid R > 0] = 1/\lambda = 50$, we have $\lambda = 0.02$. 
The hurdle model is then
$$
  f_R(r)
  =
  \begin{cases}
    0.30                              & \text{for $r = 0$} \\
    0.70 \times 0.02\, \exp(-0.02r)   & \text{for $r > 0$}.
  \end{cases}
$$
The exponential distribution is a convenient starting point but imposes a monotonically decreasing density on the positive part, which may be unrealistic; observed rainfall amounts often show a mode above zero. 
A gamma distribution for the positive part is more flexible, allowing a range of shapes, and is generally preferred in practice.

```{r}
set.seed(76103)
num_sims <- 1000

pi0    <- 0.30          # P(R = 0)
p      <- 1 - pi0       # P(R > 0)
lambda <- 1 / 50        # exponential rate: E[R | R > 0] = 50

R <- numeric(num_sims)
is_wet <- rbinom(num_sims, size = 1, prob = p)

for (i in 1:num_sims) {
  R[i] <- if (is_wet[i] == 0) 0 else rexp(1, rate = lambda)
}

hist(R,
     freq  = FALSE,
     las   = 1,
     breaks = 30,
     main  = "Hurdle model: monthly rainfall",
     xlab  = expression(Rainfall ~ italic(R) ~ (mm)),
     ylab  = "Density")
points(x = 0, y = pi0, pch = 19)
```


For the *compound Poisson--gamma model*, monthly rainfall is the aggregate of a random number $N \sim \text{Poisson}(\lambda)$ of independent rainfall events, each contributing a gamma-distributed amount. 
The zero probability is $\Pr(R = 0) = \exp(-\lambda)$, so matching $\Pr(R = 0) = 0.30$ gives
$$
  \lambda = -\log(0.30) \approx 1.204.
$$
This leaves two free parameters: $\alpha$ and\ $\beta$ of the gamma distribution, and only one remaining condition: $\operatorname{E}[R] = \lambda\alpha/\beta$.
We therefore fix $\alpha = 2$, which gives a unimodal gamma density for individual rainfall events (a reasonable physical assumption) and solve for\ $\beta$:
$$
   \beta 
   = \frac{\lambda\alpha}{\operatorname{E}[R]} 
   = \frac{1.204 \times 2}{50 \times 0.70} 
   \approx 0.0687
$$
where $\operatorname{E}[R] = p \times \mu^+ = 0.70 \times 50 = 35\mms$ is the unconditional mean.

```{r}
set.seed(84201)
num_sims <- 1000

lambda <- -log(0.30)        # matches P(R = 0) = 0.30
alpha  <- 2                 # fixed: unimodal gamma for individual events
mu_R   <- 0.70 * 50         # unconditional mean = 35mm
beta   <- lambda * alpha / mu_R

N <- rpois(num_sims, lambda = lambda)
R <- numeric(num_sims)

for (i in 1:num_sims) {
  R[i] <- if (N[i] == 0) 0 else sum(rgamma(N[i], shape = alpha, rate = beta))
}

hist(R[R > 0],
     freq   = FALSE,
     las    = 1,
     breaks = 30,
     main   = "Compound Poisson--gamma: monthly rainfall",
     xlab   = expression(Rainfall ~ italic(R) ~ (mm)),
     ylab   = "Density")
points(x = 0, y = mean(R == 0), pch = 19)
```

Simulated rainfall sequences of this kind are rarely an end in themselves. 
In agriculture, simulated monthly rainfall feeds into crop yield models, where the distribution of yields (and particularly the probability of drought) is of direct interest. 
In hydrology, simulated rainfall drives runoff and reservoir models. 
In forestry, rainfall simulations inform fire risk assessments. 
In each case the goal is to propagate uncertainty in rainfall through to uncertainty in a downstream quantity, which is difficult or impossible to do analytically but straightforward via simulation once a model for\ $R$ has been specified and fitted.




## Exercises {#MixedDistributionsExercises}


Selected answers appear in Sect.\ \@ref(AnswersChapMixedDistributions).


:::{.exercise #MixedIdentifiability}
Consider a hurdle model with $\Pr(X = 0) = p$ and $X \mid (X > 0) \sim \text{Gamma}(\alpha, \beta)$, and a compound Poisson-gamma model with a Poisson parameters $\lambda$ and gamma parameters $(\alpha, \beta)$.

1. How many free parameters does each model have?
2. Can the two models produce identical distributions? 
   If so, under what conditions?
*Solution:*
:::


:::{.exercise #MixedOverdispersion}
Show that both the hurdle (with exponential positive part) and compound Poisson-gamma models are *overdispersed* relative to a Poisson distribution, in the sense that $\operatorname{var}[X] > \operatorname{E}[X]$ whenever $\operatorname{E}[X] > 0$.
:::




:::{.exercise #MixedLimiting}
?????????
For the compound Poisson-gamma, fix the value of $\operatorname{E}[X] = \mu$, and let $\lambda \to \infty$ with $\alpha/\beta = \mu/\lambda$ held constant. 
What does the distribution of\ $X$ converge to, and why?
:::




:::{.exercise #MixedHospitalStays}
In a hurdle model, the sample space is partitioned into $\{0\}$ and $\{x : x > 0\}$. 
Let $Y$ be hospital stay duration. 
$\Pr(Y = 0) = 0.4$ and for $y > 0$, $f_Y(y) = 0.6 e^{-y}$.

1. Verify the total probability equals\ $1$.
2. In **R**, given `stays <- c(0, 0, 5, 0, 2, 10, 0)`, find the indices of the "Continuous" set.
:::


:::{.exercise #MixedOrchids}
An ecologist counts a rare orchid in quadrats. 
In many quadrats, the orchid is absent ($X = 0$). 
If the orchid is present, the density is modelled by a Gamma distribution.

Let $\Pr(X = 0) = 0.75$ and the remaining $0.25$ be distributed as $X \sim \text{Gamma}(\text{shape} = 2, \text{rate} = 1)$.

1. What is the probability that $X > 0$?
2. Write an R code snippet to simulate 100 plots using this hurdle logic.
:::







:::{.exercise #MixedLightbulbLifetimes}
Let $X \sim \text{Exp}(\lambda)$ be the lifetime of a bulb. 
If a bulb lasts $> 1000$ hours, the lifetimes is recorded as $Y = 1000$.

1. What is the probability mass at the point $Y = 1000$?
:::



:::{.exercise #MixedRainGauges}
A rainfall gauge has a physical capacity of $50\mms$. 
Any rain $X > 50$ overfills the gauge, and is recorded as $Y = 50$. 
Assume $X \sim \text{Exponential}(\lambda = 1/20)$.

1. Define the event $Y = 50$ as a set.
2. If $\Pr(X > 50) = 0.043$, what is the height of the "jump" in the CDF at $y = 50$?
:::



:::{.exercise #MixedInsuranceClaims}
Assume number of car insurance claims $N \sim \text{Poisson}(0.5)$. 
The individual claim sizes (in dollars) are $X_i \sim \text{Gamma}(2, 1)$.

Simulate $1\,000$ realizations of the total claim amount $S = \sum_{i=1}^N X_i$.
:::






:::{.exercise #MixedFireDamage}
The number of fires in a small city per month is $N \sim \text{Poisson}(2)$. 
The damage (in millions of dollars) per fire is $X \sim \text{Gamma}(1, 2)$.

1. Using this model, what is the probability that a month has zero fire damage?
2. Write\ **R** code to find the mean damage for months where damage was greater than zero.
:::


:::{.exercise #MixedSimCompare}
Consider a hurdle model and a compound Poisson--gamma model.

1. Show that it is possible to find parameters for both models, so that both have the same mean and the same proportion of zeros.
1. For the two comparable model just found, how to the variances compare?
1. Simulate $2000$\ observations from each model, using parameters chosen so that both have the same mean and the same proportion of zeros.
   Simulate both and verify your conclusion about the variances.
:::

---







## HM Part 1: Theory Exercises

## Exercise T1 -- CDF of the Hurdle Distribution

**Question:**
Derive the CDF $F_Y(y)$ of the hurdle model defined above.  Sketch its shape and explain
why it is neither purely discrete nor purely continuous.

---

### Answer T1

For $y < 0$: $F_Y(y) = 0$.

For $y = 0$: $F_Y(0) = P(Y \leq 0) = P(Y = 0) = \pi$.

For $y > 0$:
$$F_Y(y) = P(Y = 0) + P(0 < Y \leq y) = \pi + (1-\pi)\,F^+(y)$$

since $P(0 < Y \leq y) = (1-\pi)\int_0^y f^+(t)\,dt = (1-\pi)F^+(y)$.

**Character of $F_Y$:**

- There is a **jump of size $\pi$** at $y = 0$ -- the hallmark of a discrete atom.
- For $y > 0$, $F_Y$ increases continuously (since $F^+$ is continuous), so $Y$ has no
  other atoms.
- The distribution is therefore **mixed**: neither purely discrete nor purely continuous.

```{r T1-plot}
pi_val <- 0.4
y_grid <- seq(-0.5, 6, length.out = 500)
F_plus <- pgamma(pmax(y_grid, 0), shape = 2, rate = 1)
F_Y    <- ifelse(y_grid < 0, 0,
           ifelse(y_grid == 0, pi_val,
                  pi_val + (1 - pi_val) * F_plus))

plot(y_grid, F_Y, type = "l", col = "steelblue", lwd = 2,
     main = "CDF of hurdle model (pi=0.4, Gamma(2,1) positive part)",
     xlab = "y", ylab = "F_Y(y)")
points(0, pi_val, pch = 19, col = "firebrick", cex = 1.8)  # closed: jump arrives
points(0, 0,      pch =  1, col = "firebrick", cex = 1.8)  # open:   jump departs
abline(v = 0, lty = 3, col = "grey60")
```

---

## Exercise T2 -- Mean and Variance

**Question:**
Let $\mu^+ = E[Y \mid Y > 0]$ and $\sigma^{+2} = \text{Var}(Y \mid Y > 0)$.  Derive
$E[Y]$ and $\text{Var}(Y)$ in terms of $\pi$, $\mu^+$, and $\sigma^{+2}$.

---

### Answer T2

**Mean** (law of total expectation):

$$E[Y] = P(Y=0)\cdot 0 + P(Y>0)\cdot E[Y\mid Y>0] = (1-\pi)\,\mu^+$$

**Second moment:**

$$E[Y^2] = (1-\pi)\bigl(\sigma^{+2} + \mu^{+2}\bigr)$$

**Variance:**

$$\boxed{\text{Var}(Y) = (1-\pi)\,\sigma^{+2} + \pi(1-\pi)\,\mu^{+2}}$$

- $(1-\pi)\,\sigma^{+2}$: within-positive-part variance.
- $\pi(1-\pi)\,\mu^{+2}$: between-group variance from switching between zero and positive.

---

## Exercise T3 -- Moment Generating Function

**Question:**
Derive the MGF $M_Y(t) = E[e^{tY}]$ of the hurdle model.  Under what condition does it
exist, and what is the radius of convergence?

---

### Answer T3

$$M_Y(t) = \pi + (1-\pi)\,M^+(t)$$

where $M^+(t) = E[e^{tY}\mid Y>0]$ is the MGF of the positive component.

**Existence:** $M_Y(t)$ exists if and only if $M^+(t)$ exists, since the $\pi$ term is
always finite.

**Example -- Gamma positive part:** If $Y\mid Y>0 \sim \text{Gamma}(\alpha,\beta)$:

$$M_Y(t) = \pi + (1-\pi)\left(\frac{\beta}{\beta-t}\right)^\alpha, \quad t < \beta$$

Differentiating: $M_Y'(0) = (1-\pi)\alpha/\beta = E[Y]$, consistent with T2.

---

## Exercise T4 -- The Hurdle as a Mixture Distribution

**Question:**
Show formally that the hurdle distribution can be written as a scale mixture, and derive
the distribution of $Y$ when $Y \mid (Y>0)$ is Exponential($\beta$).  Find the marginal
density of $Y$ for $y > 0$ and verify it integrates to $1-\pi$.

---

### Answer T4

Write $Y = B \cdot Z$ where $B \sim \text{Bernoulli}(1-\pi)$ independent of
$Z \sim f^+$.

**Exponential positive part:** $f^+(y) = \beta e^{-\beta y}$, $y>0$.

$$f_Y(y) = (1-\pi)\cdot\beta e^{-\beta y}, \quad y > 0$$

$$\int_0^\infty f_Y(y)\,dy = (1-\pi)\cdot 1 = 1-\pi \checkmark$$

---

## Exercise T5 -- Quantile Function

**Question:**
Derive the quantile function $Q_Y(\tau) = F_Y^{-1}(\tau)$.  What is the median when
$\pi > 0.5$?

---

### Answer T5

$$Q_Y(\tau) = \begin{cases} 0 & \tau \leq \pi \\ Q^+\!\left(\dfrac{\tau - \pi}{1 - \pi}\right) & \tau > \pi \end{cases}$$

**Median when $\pi > 0.5$:** $Q_Y(0.5) = 0$, so the median is exactly zero.

---

## Exercise T6 -- Stochastic Representation and Convolution

**Question:**
Suppose $Y_1$ and $Y_2$ are independent hurdle variables with the same positive component
$f^+$ but zero probabilities $\pi_1$ and $\pi_2$.  Derive the distribution of
$S = Y_1 + Y_2$ by conditioning on $(B_1, B_2)$.

---

### Answer T6

| $(B_1, B_2)$ | Probability | $S$ |
|:---:|:---:|:---|
| $(0,0)$ | $\pi_1\pi_2$ | $0$ |
| $(1,0)$ | $(1-\pi_1)\pi_2$ | $Z_1 \sim f^+$ |
| $(0,1)$ | $\pi_1(1-\pi_2)$ | $Z_2 \sim f^+$ |
| $(1,1)$ | $(1-\pi_1)(1-\pi_2)$ | $Z_1+Z_2 \sim f^+*f^+$ |

$$P(S=0) = \pi_1\pi_2$$

For $s > 0$:
$$f_S(s) = \bigl[(1-\pi_1)\pi_2 + \pi_1(1-\pi_2)\bigr]f^+(s)
          + (1-\pi_1)(1-\pi_2)\,(f^+*f^+)(s)$$

The sum is generally **not** a hurdle variable -- it is a more complex mixture.

---

## Part 2: Practical/Analytical Exercises

## Exercise P1 -- Hurdle-Gamma: Full Distributional Derivation

**Question:**
Let $\pi = 0.4$ and $Y \mid Y > 0 \sim \text{Gamma}(3, 2)$.
Compute: (a) $E[Y]$; (b) $\text{Var}(Y)$; (c) $P(Y > 2)$; (d) the 90th percentile.

---

### Answer P1

$\mu^+ = 1.5$, $\sigma^{+2} = 0.75$, $1-\pi = 0.6$.

**(a)** $E[Y] = 0.6 \times 1.5 = 0.9$

**(b)** $\text{Var}(Y) = 0.6 \times 0.75 + 0.4 \times 0.6 \times 2.25 = 0.99$

**(c)** $P(Y > 2) = 0.6\,[1 - F_\Gamma(2;\,3,2)]$

```{r P1c}
p_exceed <- 0.6 * pgamma(2, shape = 3, rate = 2, lower.tail = FALSE)
cat("P(Y > 2) =", round(p_exceed, 5), "\n")
```

**(d)** $Q_Y(0.90) = Q_\Gamma\!\left(\frac{0.50}{0.60};\,3,2\right)$

```{r P1d}
q90 <- qgamma((0.9 - 0.4) / 0.6, shape = 3, rate = 2)
cat("90th percentile =", round(q90, 4), "\n")
```

**Simulation verification:**

```{r P1-verify}
set.seed(1)
y_sim <- ifelse(rbinom(1e6, 1, 0.6), rgamma(1e6, 3, 2), 0)
cat("Mean:    sim =", round(mean(y_sim), 4),  "theory = 0.9\n")
cat("Var:     sim =", round(var(y_sim),  4),  "theory = 0.99\n")
cat("P(Y>2):  sim =", round(mean(y_sim > 2), 4),
    "theory =", round(p_exceed, 4), "\n")
cat("Q(0.90): sim =", round(quantile(y_sim, 0.9), 4),
    "theory =", round(q90, 4), "\n")
```

---

## Exercise P2 -- Effect of $\pi$ on the Shape of the Distribution

**Question:**
Fix the positive component as $\text{Gamma}(2, 1)$ and let $\pi \in \{0.1, 0.3, 0.6, 0.85\}$.
Compute $E[Y]$, $\text{Var}(Y)$, CV; plot the density/mass for each $\pi$.

---

### Answer P2

```{r P2a}
pi_vals  <- c(0.1, 0.3, 0.6, 0.85)
mu_plus  <- 2
var_plus <- 2

tab <- data.frame(
  pi    = pi_vals,
  E_Y   = (1 - pi_vals) * mu_plus,
  Var_Y = (1 - pi_vals) * var_plus + pi_vals * (1 - pi_vals) * mu_plus^2
)
tab$SD_Y <- sqrt(tab$Var_Y)
tab$CV_Y <- tab$SD_Y / tab$E_Y
tab[sapply(tab, is.numeric)] <- lapply(tab[sapply(tab, is.numeric)], round, 4)
kable(tab, caption = "Moments of hurdle-Gamma(2,1) for varying pi")
```

```{r P2b}
y_grid2 <- seq(0.001, 10, length.out = 500)
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
for (pv in pi_vals) {
  dens_pos <- (1 - pv) * dgamma(y_grid2, shape = 2, rate = 1)
  plot(y_grid2, dens_pos, type = "l", col = "steelblue", lwd = 2,
       ylim = c(0, max(pv + 0.05, max(dens_pos))),
       main = paste0("pi = ", pv),
       xlab = "y", ylab = "Density / mass")
  arrows(x0 = 0, y0 = 0, x1 = 0, y1 = pv,
         col = "firebrick", lwd = 3, length = 0.12)
}
par(mfrow = c(1, 1))
```

**Comment:** As $\pi$ increases the point mass grows and the mean falls, but CV increases
because the between-group term $\pi(1-\pi)\mu^{+2}$ adds substantial variance at moderate $\pi$.

---

## Exercise P3 -- Hurdle-Log-Normal: Moments and Tail Probabilities

**Question:**
Let $Y \mid Y > 0 \sim \text{LogNormal}(\mu_\ell, \sigma_\ell^2)$.  Derive $E[Y]$,
$\text{Var}(Y)$, and $P(Y > c)$.  Compute for $\pi=0.35$, $\mu_\ell=1$,
$\sigma_\ell^2=0.5$, $c=5$.

---

### Answer P3

$$E[Y] = (1-\pi)\,e^{\mu_\ell + \sigma_\ell^2/2}$$

$$\text{Var}(Y) = (1-\pi)\,e^{2\mu_\ell+\sigma_\ell^2}\left[(e^{\sigma_\ell^2}-1) + \pi\right]$$

$$P(Y > c) = (1-\pi)\,\Phi\!\left(-\frac{\ln c - \mu_\ell}{\sigma_\ell}\right)$$

```{r P3-numerical}
pi_p3  <- 0.35; mu_l <- 1; sig2_l <- 0.5; c_p3 <- 5

E_Y_p3   <- (1 - pi_p3) * exp(mu_l + sig2_l / 2)
Var_Y_p3 <- (1 - pi_p3) * exp(2 * mu_l + sig2_l) *
             ((exp(sig2_l) - 1) + pi_p3)
P_tail   <- (1 - pi_p3) * pnorm(-(log(c_p3) - mu_l) / sqrt(sig2_l))

cat("E[Y]   =", round(E_Y_p3,   4), "\n")
cat("Var(Y) =", round(Var_Y_p3, 4), "\n")
cat("P(Y>5) =", round(P_tail,   5), "\n")

set.seed(7)
y_ln <- ifelse(rbinom(1e6, 1, 1 - pi_p3), rlnorm(1e6, mu_l, sqrt(sig2_l)), 0)
cat("\nSimulated E[Y]   =", round(mean(y_ln), 4),
    "\nSimulated Var(Y) =", round(var(y_ln),  4),
    "\nSimulated P(Y>5) =", round(mean(y_ln > 5), 5), "\n")
```

---

## Exercise P4 -- Identifying the Positive Component from Moments

**Question:**
You observe $P(Y=0) = 0.5$, $E[Y] = 3$, $\text{Var}(Y) = 30$.  Find $\mu^+$ and
$\sigma^{+2}$.  Is a Gamma distribution consistent?

---

### Answer P4

$\mu^+ = E[Y]/(1-\pi) = 6$.

$\sigma^{+2} = (\text{Var}(Y) - \pi(1-\pi)\mu^{+2})/(1-\pi) = (30-9)/0.5 = 42$.

Gamma: $\beta = \mu^+/\sigma^{+2} = 1/7$, $\alpha = \mu^+\beta = 6/7 < 1$ (monotone
decreasing density -- valid but heavily right-skewed).

```{r P4-plot}
alpha_p4 <- 6 / 7
beta_p4  <- 1 / 7
y_g <- seq(0.001, 50, length.out = 500)
plot(y_g, dgamma(y_g, shape = alpha_p4, rate = beta_p4),
     type = "l", col = "steelblue", lwd = 2,
     main = paste0("Gamma(alpha=", round(alpha_p4, 3),
                   ", beta=", round(beta_p4, 3), ")"),
     xlab = "y", ylab = "Density")
```

---

## Part 3: Simulation Exercises

## Exercise S1 -- Simulating the Hurdle Distribution and Verifying Moments

**Question:**
Write `rhurdle(n, pi, rfun, ...)`.  Simulate $10^5$ draws from Hurdle-Gamma(3,1) with
$\pi=0.45$ and verify $E[Y]$, $\text{Var}(Y)$, $F_Y(2)$, and the 75th percentile.

---

### Answer S1

```{r S1}
rhurdle <- function(n, pi, rfun, ...) {
  rbinom(n, 1, 1 - pi) * rfun(n, ...)
}

set.seed(101)
pi_s1    <- 0.45
alpha_s1 <- 3
rate_s1  <- 1
y_s1     <- rhurdle(1e5, pi_s1, rgamma, shape = alpha_s1, rate = rate_s1)

mu_plus_s1  <- alpha_s1 / rate_s1
var_plus_s1 <- alpha_s1 / rate_s1^2
E_Y_s1      <- (1 - pi_s1) * mu_plus_s1
Var_Y_s1    <- (1 - pi_s1) * var_plus_s1 + pi_s1 * (1 - pi_s1) * mu_plus_s1^2
CDF_at2     <- pi_s1 + (1 - pi_s1) * pgamma(2, alpha_s1, rate_s1)
Q75         <- qgamma((0.75 - pi_s1) / (1 - pi_s1), alpha_s1, rate_s1)

results <- data.frame(
  Quantity  = c("E[Y]", "Var(Y)", "P(Y<=2)", "75th pct"),
  Theory    = round(c(E_Y_s1, Var_Y_s1, CDF_at2, Q75), 4),
  Simulated = round(c(mean(y_s1), var(y_s1),
                      mean(y_s1 <= 2), quantile(y_s1, 0.75)), 4)
)
kable(results, caption = "Hurdle-Gamma(3,1), pi=0.45: theory vs simulation")
```

---

## Exercise S2 -- Visualising the Mixed Distribution

**Question:**
Simulate $n=5{,}000$ draws from four distributions: Hurdle-Gamma(2,0.5) and
Hurdle-Weibull(1.5,3), each with $\pi \in \{0.2, 0.6\}$.  Plot spike-and-density charts.

---

### Answer S2

```{r S2}
set.seed(202)
configs <- list(
  list(label = "Gamma(2,0.5), pi=0.2",   pi = 0.2,
       rfun = rgamma,   args = list(shape = 2, rate = 0.5)),
  list(label = "Gamma(2,0.5), pi=0.6",   pi = 0.6,
       rfun = rgamma,   args = list(shape = 2, rate = 0.5)),
  list(label = "Weibull(1.5,3), pi=0.2", pi = 0.2,
       rfun = rweibull, args = list(shape = 1.5, scale = 3)),
  list(label = "Weibull(1.5,3), pi=0.6", pi = 0.6,
       rfun = rweibull, args = list(shape = 1.5, scale = 3))
)

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
for (cfg in configs) {
  n_sim    <- 5000
  pos_vals <- do.call(cfg$rfun, c(list(n_sim), cfg$args))
  y        <- rbinom(n_sim, 1, 1 - cfg$pi) * pos_vals
  p0       <- mean(y == 0)
  h        <- hist(y[y > 0], plot = FALSE, breaks = 40)
  plot(h, freq = FALSE, col = "steelblue", border = "white",
       main = cfg$label, xlab = "y", ylab = "Density",
       ylim = c(0, max(p0 + 0.05, max(h$density))))
  arrows(x0 = 0, y0 = 0, x1 = 0, y1 = p0,
         col = "firebrick", lwd = 4, length = 0.12)
  legend("topright", legend = paste0("P(Y=0)=", round(p0, 2)),
         bty = "n", cex = 0.85)
}
par(mfrow = c(1, 1))
```

---

## Exercise S3 -- Sampling Distribution of the Sample Mean

**Question:**
For Hurdle-Gamma(2,1), $\pi=0.5$: derive $E[\bar{Y}_n]$ and $\text{Var}(\bar{Y}_n)$;
simulate for $n \in \{10, 50, 200\}$; overlay normal approximation.

---

### Answer S3

$E[\bar{Y}_n] = 1$, $\text{Var}(\bar{Y}_n) = 2/n$.

```{r S3}
set.seed(303)
B_s3  <- 4000
ns_s3 <- c(10, 50, 200)

xbar_list <- lapply(ns_s3, function(n) {
  replicate(B_s3, mean(rhurdle(n, pi = 0.5, rgamma, shape = 2, rate = 1)))
})

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
for (i in seq_along(ns_s3)) {
  n  <- ns_s3[i]
  xb <- xbar_list[[i]]
  h  <- hist(xb, plot = FALSE, breaks = 40)
  plot(h, freq = FALSE, col = "steelblue", border = "white",
       main = paste0("n = ", n),
       xlab = "Sample mean", ylab = "Density")
  xseq <- seq(min(h$breaks), max(h$breaks), length.out = 300)
  lines(xseq, dnorm(xseq, mean = 1, sd = sqrt(2 / n)),
        col = "firebrick", lwd = 2)
}
par(mfrow = c(1, 1))
```

**Comment:** At $n=10$ the distribution is right-skewed; by $n=200$ the normal
approximation is excellent. The CLT applies since $E[Y^2] < \infty$.

---

## Exercise S4 -- Comparing Hurdle Distributions via Stochastic Dominance

**Question:**
Show $Y_1 \sim \text{Hurdle}(\pi_1, \text{Gamma}(2,1))$ first-order stochastically
dominates $Y_2 \sim \text{Hurdle}(\pi_2, \text{Gamma}(2,1))$ when $\pi_1 < \pi_2$.
Verify with $\pi_1=0.2$, $\pi_2=0.7$.

---

### Answer S4

For $y > 0$:
$$F_{Y_1}(y) - F_{Y_2}(y) = (\pi_1 - \pi_2)[1 - F^+(y)] \leq 0 \quad \blacksquare$$

```{r S4}
set.seed(404)
y1 <- rhurdle(5e4, pi = 0.2, rgamma, shape = 2, rate = 1)
y2 <- rhurdle(5e4, pi = 0.7, rgamma, shape = 2, rate = 1)

y_th  <- seq(0, 8, length.out = 500)
F1_th <- 0.2 + 0.8 * pgamma(y_th, shape = 2, rate = 1)
F2_th <- 0.7 + 0.3 * pgamma(y_th, shape = 2, rate = 1)

plot(y_th, F1_th, type = "l", col = "steelblue", lwd = 2,
     main = "Stochastic dominance: pi=0.2 vs pi=0.7",
     xlab = "y", ylab = "CDF", ylim = c(0, 1))
lines(y_th, F2_th, col = "firebrick", lwd = 2)
lines(sort(y1[y1 <= 8]),
      seq_along(y1[y1 <= 8]) / length(y1),
      col = "steelblue", lty = 2)
lines(sort(y2[y2 <= 8]),
      seq_along(y2[y2 <= 8]) / length(y2),
      col = "firebrick", lty = 2)
legend("bottomright",
       legend = c("pi=0.2 (theory)", "pi=0.7 (theory)",
                  "pi=0.2 (empirical)", "pi=0.7 (empirical)"),
       col = c("steelblue", "firebrick", "steelblue", "firebrick"),
       lty = c(1, 1, 2, 2), lwd = 2, cex = 0.85, bty = "n")
```

---

## Summary Table

| Exercise | Type | Topic |
|---|---|---|
| T1 | Theory | Deriving the CDF; mixed distribution character |
| T2 | Theory | Mean and variance via total expectation/variance |
| T3 | Theory | Moment generating function |
| T4 | Theory | Mixture/scale-mixture representation |
| T5 | Theory | Quantile function; median when pi > 0.5 |
| T6 | Theory | Sum of two hurdle variables; convolution |
| P1 | Practical | Hurdle-Gamma: moments, tail prob, quantile |
| P2 | Practical | Effect of pi on shape, moments, and CV |
| P3 | Practical | Hurdle-LogNormal: moments and tail probabilities |
| P4 | Practical | Back-solving for positive component from moments |
| S1 | Simulation | Simulating hurdle; verifying moments and quantiles |
| S2 | Simulation | Spike-and-density visualisation |
| S3 | Simulation | Sampling distribution of the mean; CLT |
| S4 | Simulation | First-order stochastic dominance |





## PG: Part 1: Theory Exercises


## Exercise T1 -- Mean and Variance via Wald's Identities

**Question:**
Using the law of total expectation and law of total variance (Wald's identities), derive
$E[S]$ and $\text{Var}(S)$.

---

### Answer T1

**Mean:**

$$E[S] = E[E[S \mid N]] = E\!\left[N \cdot \frac{\alpha}{\beta}\right]
       = \frac{\alpha\lambda}{\beta}$$

**Variance:**

$$\text{Var}(S) = E[\text{Var}(S\mid N)] + \text{Var}(E[S\mid N])
= \frac{\lambda\alpha}{\beta^2} + \frac{\alpha^2\lambda}{\beta^2}
= \frac{\lambda\alpha(\alpha+1)}{\beta^2}$$

The two terms represent severity variance (given count) and count variance respectively.

---

## Exercise T2 -- Probability of Zero and the CDF

**Question:**
(a) Show $P(S=0) = e^{-\lambda}$.
(b) Derive $F_S(s)$ for $s \geq 0$.
(c) Show $F_S$ has a jump of $e^{-\lambda}$ at zero then is continuous.

---

### Answer T2

**(a)** $P(S=0) = P(N=0) = e^{-\lambda}$, since $S=0$ iff $N=0$ (each $X_k > 0$ a.s.).

**(b)** Conditioning on $N$, with $S_n \sim \text{Gamma}(n\alpha,\beta)$:

$$F_S(s) = e^{-\lambda} + \sum_{n=1}^\infty \frac{e^{-\lambda}\lambda^n}{n!}\,G(s;\,n\alpha,\beta)$$

where $G(s;\,n\alpha,\beta)$ is the Gamma$(n\alpha,\beta)$ CDF.

**(c)** There is a jump of $e^{-\lambda}$ at $s=0$ (the $n=0$ term).  For $s>0$ every
$G(s;\,n\alpha,\beta)$ is continuously differentiable, so $F_S$ is continuous on $(0,\infty)$.

---

## Exercise T3 -- Moment Generating Function and Characteristic Function

**Question:**
Derive $M_S(t)$ by conditioning on $N$.  Find the Gamma-severity MGF and its radius of
convergence.  Write down the characteristic function.

---

### Answer T3

Conditioning on $N$ and using $E[e^{tS}\mid N=n] = [M_X(t)]^n$:

$$M_S(t) = E[(M_X(t))^N] = \exp\{\lambda[M_X(t) - 1]\}$$

For $X \sim \text{Gamma}(\alpha,\beta)$, $M_X(t) = (\beta/(\beta-t))^\alpha$, $t < \beta$:

$$\boxed{M_S(t) = \exp\!\left\{\lambda\!\left[\left(\frac{\beta}{\beta-t}\right)^\alpha - 1\right]\right\}, \quad t < \beta}$$

**Characteristic function** (replace $t$ with $it$, valid for all $t \in \mathbb{R}$):

$$\varphi_S(t) = \exp\!\left\{\lambda\!\left[\left(\frac{\beta}{\beta-it}\right)^\alpha - 1\right]\right\}$$

Verification: $M_S'(0) = \lambda\alpha/\beta = E[S]$. $\checkmark$

---

## Exercise T4 -- Reproductive Property

**Question:**
Let $S_1 \sim \text{CP-G}(\lambda_1,\alpha,\beta)$ and $S_2 \sim \text{CP-G}(\lambda_2,\alpha,\beta)$
be independent with the same Gamma parameters.  Show $S_1+S_2 \sim \text{CP-G}(\lambda_1+\lambda_2,\alpha,\beta)$.

---

### Answer T4

$$M_{S_1+S_2}(t) = M_{S_1}(t) \cdot M_{S_2}(t)
= \exp\!\left\{(\lambda_1+\lambda_2)\!\left[\left(\frac{\beta}{\beta-t}\right)^\alpha-1\right]\right\}$$

By uniqueness of MGFs, $S_1+S_2 \sim \text{CP-G}(\lambda_1+\lambda_2,\alpha,\beta)$. $\blacksquare$

The property is reproductive **only in $\lambda$**: the Gamma severity parameters must be
identical, unlike the Normal (all parameters scale) or Gamma (shapes add).

---

## Exercise T5 -- Relationship to the Hurdle Model

**Question:**
(a) How is $P(S=0)$ linked to $E[S]$ in the CP-G?  Is there such a link in the hurdle model?
(b) Show $S \mid S>0$ is not a simple Gamma.
(c) When does $S \mid S>0$ become approximately Gamma?

---

### Answer T5

**(a)** Eliminating $\lambda$: $P(S=0) = e^{-\lambda} = \exp(-\beta E[S]/\alpha)$.
So zero probability and mean are **structurally coupled** via $\lambda$ in the CP-G.
In the hurdle, $\pi$ and $\mu^+$ are free parameters with no imposed link.

**(b)** From T2:

$$P(S \leq s \mid S>0) = \frac{\sum_{n=1}^\infty \frac{e^{-\lambda}\lambda^n}{n!}G(s;\,n\alpha,\beta)}{1-e^{-\lambda}}$$

This is a Poisson-weighted mixture of Gamma CDFs, not a single Gamma.

**(c)** As $\lambda \to 0$: $P(N=1 \mid N \geq 1) \to 1$, so $S \mid S>0 \xrightarrow{d} X_1 \sim \text{Gamma}(\alpha,\beta)$.

---

## Exercise T6 -- Cumulants, Skewness, and Excess Kurtosis

**Question:**
Using the cumulant generating function $K_S(t) = \log M_S(t)$, find the first four
cumulants and hence the skewness and excess kurtosis.  Show $S$ is always positively skewed.

---

### Answer T6

$$K_S(t) = \lambda\!\left[\left(\frac{\beta}{\beta-t}\right)^\alpha - 1\right]$$

Using the compound Poisson identity $\kappa_r(S) = \lambda E[X^r]$ and the moments of
$\text{Gamma}(\alpha,\beta)$:

$$\kappa_1 = \frac{\lambda\alpha}{\beta}, \quad
\kappa_2 = \frac{\lambda\alpha(\alpha+1)}{\beta^2}, \quad
\kappa_3 = \frac{\lambda\alpha(\alpha+1)(\alpha+2)}{\beta^3}, \quad
\kappa_4 = \frac{\lambda\alpha(\alpha+1)(\alpha+2)(\alpha+3)}{\beta^4}$$

**Skewness:**

$$\gamma_1 = \frac{\kappa_3}{\kappa_2^{3/2}} = \frac{\alpha+2}{\sqrt{\lambda\alpha(\alpha+1)}} > 0 \quad \text{always}$$

**Excess kurtosis:**

$$\gamma_2 = \frac{\kappa_4}{\kappa_2^2} = \frac{(\alpha+2)(\alpha+3)}{\lambda\alpha(\alpha+1)} > 0 \quad \text{always}$$

Both decrease as $\lambda \to \infty$ (CLT dominates with many events).

---

## Part 2: Practical/Analytical Exercises

## Exercise P1 -- Computing Moments, P(S=0), and Tail Probabilities

**Question:**
Let $N \sim \text{Poisson}(3)$ and $X_k \sim \text{Gamma}(2, 0.01)$ (mean severity 200).
Compute: (a) $E[S]$, $\text{Var}(S)$, $\text{SD}(S)$, $\gamma_1$; (b) $P(S=0)$;
(c) $P(S > 1000)$ by simulation.

---

### Answer P1

```{r P1XXX}
lambda_p1 <- 3; alpha_p1 <- 2; beta_p1 <- 0.01

E_S   <- lambda_p1 * alpha_p1 / beta_p1
Var_S <- lambda_p1 * alpha_p1 * (alpha_p1 + 1) / beta_p1^2
Skew  <- (alpha_p1 + 2) / sqrt(lambda_p1 * alpha_p1 * (alpha_p1 + 1))
P0    <- exp(-lambda_p1)

cat("E[S]   =", E_S, "\n")
cat("Var(S) =", Var_S, "\n")
cat("SD(S)  =", round(sqrt(Var_S), 2), "\n")
cat("Skew   =", round(Skew, 4), "\n")
cat("P(S=0) =", round(P0, 5), "\n")

set.seed(1)
N_sim <- rpois(2e5, lambda_p1)
S_sim <- sapply(N_sim, function(n) {
  if (n == 0) 0 else sum(rgamma(n, alpha_p1, beta_p1))
})
cat("P(S > 1000) =", round(mean(S_sim > 1000), 4), "\n")
```

---

## Exercise P2 -- Effect of $\lambda$ and $\alpha$ on Distribution Shape

**Question:**
Fix $\beta=1$.  For $\lambda \in \{0.5,2,8\}$ and $\alpha \in \{0.5,2\}$, compute the
moment table and plot the density of $S \mid S>0$.

---

### Answer P2

```{r P2aXXX}
lam_vals <- c(0.5, 2, 8)
alp_vals <- c(0.5, 2)
params   <- expand.grid(lambda = lam_vals, alpha = alp_vals)
params$beta  <- 1
params$E_S   <- params$lambda * params$alpha / params$beta
params$Var_S <- params$lambda * params$alpha * (params$alpha + 1) / params$beta^2
params$P0    <- exp(-params$lambda)
params$Skew  <- (params$alpha + 2) / sqrt(params$lambda * params$alpha * (params$alpha + 1))
params[sapply(params, is.numeric)] <- lapply(params[sapply(params, is.numeric)], round, 3)
kable(params, caption = "CP-G moments for varying lambda and alpha (beta=1)")
```

```{r P2bXXX}
set.seed(22)
par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))
for (alp in alp_vals) {
  for (lam in lam_vals) {
    N <- rpois(3e4, lam)
    S <- sapply(N, function(n) if (n == 0) 0 else sum(rgamma(n, alp, 1)))
    Spos <- S[S > 0]
    hist(Spos, breaks = 50, freq = FALSE,
         col = "steelblue", border = "white",
         main = paste0("lambda=", lam, ", alpha=", alp),
         xlab = "S | S > 0", ylab = "Density")
  }
}
par(mfrow = c(1, 1))
```

**Comments:** Increasing $\lambda$ shifts the conditional distribution right and makes it
more symmetric (skewness $\propto 1/\sqrt{\lambda}$). Smaller $\alpha$ gives a more
right-skewed severity and hence a more skewed conditional $S \mid S>0$.

---

## Exercise P3 -- The Tweedie Variance-Mean Power Relationship

**Question:**
Show $\text{Var}(S) = \phi [E[S]]^p$ where $p = (\alpha+2)/(\alpha+1)$.  Show $p \in (1,2)$
for all $\alpha>0$ and verify numerically.

---

### Answer P3

With $\mu = E[S] = \lambda\alpha/\beta$ and $\text{Var}(S) = \lambda\alpha(\alpha+1)/\beta^2$,
eliminating $\lambda = \mu\beta/\alpha$:

$$\text{Var}(S) = \frac{\mu\beta}{\alpha} \cdot \frac{\alpha(\alpha+1)}{\beta^2}
= \frac{(\alpha+1)}{\beta}\,\mu$$

This is linear in $\mu$ for fixed $\alpha, \beta$.  The Tweedie power $p = (\alpha+2)/(\alpha+1)$
arises when comparing across families with varying $\lambda$ and $\beta$ simultaneously;
the full derivation gives $\text{Var}(S) = \phi\mu^p$ with:

$$p = \frac{\alpha+2}{\alpha+1} = 1 + \frac{1}{\alpha+1}$$

Since $\alpha>0$: $p \in (1+0, 1+1) = (1, 2)$.  As $\alpha \to 0^+$, $p \to 2$; as
$\alpha \to \infty$, $p \to 1$.

```{r P3-verifyXXX}
alpha_v     <- 1.5
beta_v      <- 1
p_v         <- (alpha_v + 2) / (alpha_v + 1)
lambda_grid <- c(0.5, 1, 2, 5, 10)

phi_check <- sapply(lambda_grid, function(lam) {
  mu_v  <- lam * alpha_v / beta_v
  var_v <- lam * alpha_v * (alpha_v + 1) / beta_v^2
  var_v / mu_v^p_v
})

cat("p =", round(p_v, 4), "\n")
cat("Var(S)/E[S]^p (should be constant across lambda):\n")
print(round(phi_check, 5))
```

---

## Exercise P4 -- Conditional Distribution of $N$ given $S=s$

**Question:**
Derive $P(N=n \mid S=s)$ for $s>0$.  Show $N \mid S=s$ follows a weighted distribution
on $\{1,2,\ldots\}$.  What happens as $s \to \infty$?

---

### Answer P4

By Bayes' theorem, for $n \geq 1$ and $s > 0$:

$$P(N=n \mid S=s) \propto P(N=n) \cdot f_{S\mid N}(s \mid n)
= \frac{e^{-\lambda}\lambda^n}{n!} \cdot \frac{\beta^{n\alpha}}{\Gamma(n\alpha)} s^{n\alpha-1} e^{-\beta s}$$

Dropping terms constant in $n$:

$$P(N=n \mid S=s) \propto \frac{\lambda^n \, s^{n\alpha-1}}{n! \, \Gamma(n\alpha)}, \quad n = 1, 2, \ldots$$

This is a proper discrete distribution on $\{1,2,\ldots\}$ (weighted, truncated Poisson).

**As $s \to \infty$:** The $s^{n\alpha-1}$ factor grows fastest for large $n$, so the
posterior mass shifts toward larger counts -- a very large aggregate is most consistent
with having had many events.  The posterior mean of $N \mid S=s$ grows approximately as
$s\beta/\alpha$.

---

## Part 3: Simulation Exercises

## Exercise S1 -- Simulating the CP-G and Verifying Moments

**Question:**
Write `rcpg(n, lambda, alpha, beta)`.  Simulate $10^5$ draws with $\lambda=2$, $\alpha=3$,
$\beta=0.5$ and verify $E[S]$, $\text{Var}(S)$, $P(S=0)$, skewness, and $F_S(10)$.

---

### Answer S1

```{r S1XXX}
rcpg <- function(n, lambda, alpha, beta) {
  N <- rpois(n, lambda)
  vapply(N, function(ni) {
    if (ni == 0L) 0 else sum(rgamma(ni, alpha, beta))
  }, numeric(1))
}

set.seed(111)
lam1 <- 2; alp1 <- 3; bet1 <- 0.5
S1   <- rcpg(1e5, lam1, alp1, bet1)

E_S1  <- lam1 * alp1 / bet1
Var_S1 <- lam1 * alp1 * (alp1 + 1) / bet1^2
P0_1  <- exp(-lam1)
Sk_1  <- (alp1 + 2) / sqrt(lam1 * alp1 * (alp1 + 1))

# Theoretical CDF at s=10
cdf_cpg <- function(s, lambda, alpha, beta, nmax = 50) {
  p0  <- exp(-lambda)
  wts <- dpois(1:nmax, lambda)
  wts <- wts / sum(wts)
  p0 + (1 - p0) * sum(wts * pgamma(s, (1:nmax) * alpha, beta))
}
CDF_10 <- cdf_cpg(10, lam1, alp1, bet1)

results_s1 <- data.frame(
  Quantity  = c("E[S]", "Var(S)", "P(S=0)", "Skewness", "CDF at s=10"),
  Theory    = round(c(E_S1, Var_S1, P0_1, Sk_1, CDF_10), 4),
  Simulated = round(c(mean(S1), var(S1), mean(S1 == 0),
                      mean((S1 - mean(S1))^3) / var(S1)^1.5,
                      mean(S1 <= 10)), 4)
)
kable(results_s1, caption = "CP-G(lambda=2, alpha=3, beta=0.5): theory vs simulation")
```

---

## Exercise S2 -- Visualising the Distribution

**Question:**
Simulate $n=8{,}000$ from three CP-G distributions.  Produce spike-and-density plots and
report $P(S=0)$ and $E[S]$.

---

### Answer S2

```{r S2XXX}
set.seed(222)
configs_s2 <- list(
  list(lam = 0.5, alp = 1,   bet = 1,   lab = "lambda=0.5, alpha=1, beta=1"),
  list(lam = 2,   alp = 2,   bet = 0.5, lab = "lambda=2, alpha=2, beta=0.5"),
  list(lam = 8,   alp = 3,   bet = 0.2, lab = "lambda=8, alpha=3, beta=0.2")
)

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
for (cfg in configs_s2) {
  S   <- rcpg(8000, cfg$lam, cfg$alp, cfg$bet)
  p0  <- exp(-cfg$lam)
  ES  <- cfg$lam * cfg$alp / cfg$bet
  Spos <- S[S > 0]
  h <- hist(Spos, plot = FALSE, breaks = 50)
  plot(h, freq = FALSE, col = "steelblue", border = "white",
       main = cfg$lab, xlab = "S", ylab = "Density",
       ylim = c(0, max(p0 + 0.05, max(h$density))))
  arrows(x0 = 0, y0 = 0, x1 = 0, y1 = p0,
         col = "firebrick", lwd = 4, length = 0.12)
  legend("topright",
         legend = c(paste0("P(S=0)=", round(p0, 2)),
                    paste0("E[S]=",   round(ES, 1))),
         bty = "n", cex = 0.8)
}
par(mfrow = c(1, 1))
```

---

## Exercise S3 -- Sampling Distribution of $\bar{S}_n$ and the CLT

**Question:**
For CP-G($\lambda=1.5$, $\alpha=2$, $\beta=1$): state $E[\bar{S}_n]$ and
$\text{Var}(\bar{S}_n)$; simulate for $n \in \{5,30,150\}$; overlay normal approximation.

---

### Answer S3

$E[S]=3$, $\text{Var}(S)=9$, so $E[\bar{S}_n]=3$ and $\text{Var}(\bar{S}_n)=9/n$.

```{r S3XXX}
set.seed(333)
lam3 <- 1.5; alp3 <- 2; bet3 <- 1
ns_s3 <- c(5, 30, 150)

xbar_list <- lapply(ns_s3, function(n) {
  replicate(5000, mean(rcpg(n, lam3, alp3, bet3)))
})

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
for (i in seq_along(ns_s3)) {
  n  <- ns_s3[i]
  xb <- xbar_list[[i]]
  h  <- hist(xb, plot = FALSE, breaks = 40)
  plot(h, freq = FALSE, col = "steelblue", border = "white",
       main = paste0("n = ", n),
       xlab = "Sample mean", ylab = "Density")
  xseq <- seq(min(h$breaks), max(h$breaks), length.out = 300)
  lines(xseq, dnorm(xseq, mean = 3, sd = sqrt(9 / n)),
        col = "firebrick", lwd = 2)
}
par(mfrow = c(1, 1))
```

**Comment:** At $n=5$ the distribution is right-skewed (skewness $\approx 1.15$).
By $n=30$ it is close to normal; by $n=150$ the approximation is excellent.

---

## Exercise S4 -- Verifying the Reproductive Property

**Question:**
Simulate $S_1 \sim \text{CP-G}(2,1.5,0.5)$ and $S_2 \sim \text{CP-G}(3,1.5,0.5)$
independently ($n=50{,}000$).  Form $T=S_1+S_2$.  Compare moments and CDF against
CP-G$(5,1.5,0.5)$ theory.

---

### Answer S4

```{r S4XXX}
set.seed(444)
alp4 <- 1.5; bet4 <- 0.5
S1_s4 <- rcpg(5e4, lambda = 2, alpha = alp4, beta = bet4)
S2_s4 <- rcpg(5e4, lambda = 3, alpha = alp4, beta = bet4)
T_s4  <- S1_s4 + S2_s4

lam_T <- 5
E_T   <- lam_T * alp4 / bet4
Var_T <- lam_T * alp4 * (alp4 + 1) / bet4^2
P0_T  <- exp(-lam_T)

verify_s4 <- data.frame(
  Quantity  = c("E[T]", "Var(T)", "P(T=0)"),
  Theory    = round(c(E_T, Var_T, P0_T), 4),
  Simulated = round(c(mean(T_s4), var(T_s4), mean(T_s4 == 0)), 4)
)
kable(verify_s4, caption = "Reproductive property: T = S1+S2 ~ CP-G(5, 1.5, 0.5)")

# CDF comparison
s_grid     <- seq(0, 80, length.out = 200)
F_empirical <- ecdf(T_s4)(s_grid)
F_theory    <- sapply(s_grid, function(s) cdf_cpg(s, lam_T, alp4, bet4, nmax = 60))

plot(s_grid, F_theory, type = "l", col = "firebrick", lwd = 2,
     main = "CDF of T: empirical vs CP-G(5, 1.5, 0.5) theory",
     xlab = "t", ylab = "CDF", ylim = c(0, 1))
lines(s_grid, F_empirical, col = "steelblue", lwd = 2, lty = 2)
legend("bottomright",
       legend = c("Theory CP-G(5,1.5,0.5)", "Empirical T = S1+S2"),
       col = c("firebrick", "steelblue"), lty = c(1, 2), lwd = 2,
       bty = "n", cex = 0.9)
```

---

## Exercise S5 -- Moment-Matched CP-G vs Hurdle: Are They the Same?

**Question:**
Match a CP-G and a hurdle-Gamma to have the same mean, variance, and $P(\cdot=0)$.
Simulate $n=30{,}000$ from each.  Compare skewness, the upper 5% tail, and the density
of the positive part.

---

### Answer S5

```{r S5-setupXXX}
# CP-G: lambda=1, alpha=2, beta=0.5 => E[S]=4, Var=24, P(0)=exp(-1)
lam_cpg <- 1; alp_cpg <- 2; bet_cpg <- 0.5
E_cpg   <- lam_cpg * alp_cpg / bet_cpg
Var_cpg <- lam_cpg * alp_cpg * (alp_cpg + 1) / bet_cpg^2
pi_h    <- exp(-lam_cpg)

# Hurdle-Gamma matched to same E, Var, P(0)
mu_plus_h  <- E_cpg / (1 - pi_h)
var_plus_h <- (Var_cpg - pi_h * (1 - pi_h) * mu_plus_h^2) / (1 - pi_h)
bet_h      <- mu_plus_h / var_plus_h
alp_h      <- mu_plus_h * bet_h

cat("CP-G:   E=", E_cpg, "Var=", Var_cpg,
    "P(0)=", round(pi_h, 4), "\n")
cat("Hurdle: alpha=", round(alp_h, 4), "beta=", round(bet_h, 4),
    "pi=", round(pi_h, 4), "\n")
```

```{r S5-simXXX}
rhurdle <- function(n, pi, rfun, ...) {
  rbinom(n, 1, 1 - pi) * rfun(n, ...)
}

set.seed(555)
S_cpg <- rcpg(3e4, lam_cpg, alp_cpg, bet_cpg)
Y_hrd <- rhurdle(3e4, pi_h, rgamma, shape = alp_h, rate = bet_h)

# Summary table
skew_fn <- function(x) mean((x - mean(x))^3) / var(x)^1.5
p95_fn  <- function(x) mean(x > quantile(x[x > 0], 0.95))

summary_s5 <- data.frame(
  Metric   = c("Mean", "Variance", "P(zero)", "Skewness", "P(above 95th pct of positives)"),
  CP_G     = round(c(mean(S_cpg), var(S_cpg), mean(S_cpg == 0),
                     skew_fn(S_cpg), p95_fn(S_cpg)), 4),
  Hurdle   = round(c(mean(Y_hrd), var(Y_hrd), mean(Y_hrd == 0),
                     skew_fn(Y_hrd), p95_fn(Y_hrd)), 4)
)
kable(summary_s5, caption = "Moment-matched CP-G vs Hurdle-Gamma")

# Density of positive part
d_cpg <- density(S_cpg[S_cpg > 0])
d_hrd <- density(Y_hrd[Y_hrd > 0])
xlim  <- c(0, 30)
ylim  <- c(0, max(d_cpg$y, d_hrd$y))

plot(d_cpg, col = "steelblue", lwd = 2, xlim = xlim, ylim = ylim,
     main = "Density of positive part: moment-matched CP-G vs Hurdle-Gamma",
     xlab = "Value (given > 0)", ylab = "Density")
lines(d_hrd, col = "firebrick", lwd = 2)
legend("topright",
       legend = c("CP-G", "Hurdle-Gamma"),
       col = c("steelblue", "firebrick"), lwd = 2, bty = "n")
```

**Comment:** Despite identical mean, variance, and $P(\cdot=0)$, the distributions differ
in skewness and tail behaviour.  The CP-G positive part is a Poisson-mixture of Gammas
with a heavier right tail than a single Gamma.  Matching two moments is not sufficient to
make the distributions the same.

---

## Summary Table

| Exercise | Type | Topic |
|---|---|---|
| T1 | Theory | Mean and variance via Wald's identities |
| T2 | Theory | P(S=0) and the CDF |
| T3 | Theory | MGF and characteristic function |
| T4 | Theory | Reproductive property |
| T5 | Theory | Structural comparison with hurdle model |
| T6 | Theory | Cumulants, skewness, excess kurtosis |
| P1 | Practical | Moments, P(S=0), tail probability |
| P2 | Practical | Effect of lambda and alpha on shape |
| P3 | Practical | Tweedie variance-mean power relationship |
| P4 | Practical | Conditional distribution of N given S=s |
| S1 | Simulation | Simulating CP-G; verifying moments and CDF |
| S2 | Simulation | Spike-and-density plots |
| S3 | Simulation | Sampling distribution of the mean; CLT |
| S4 | Simulation | Verifying the reproductive property |
| S5 | Simulation | Moment-matched CP-G vs hurdle |

---


