
# Transformations of random variables {#ChapterTransformations}

::: {.objectivesBox .objectives data-latex="{iconmonstr-target-4-240.png}"}
On completion of this chapter, you should be able to:

* derive the distribution of a transformed variable, given the distribution of the original variable, using the distribution function method, the change of variable method, and the moment generating function method as appropriate.
* find the joint distribution of two transformed variables in a bivariate situation.
:::


## Introduction

\index{Transformations of random variables|(}
In this chapter, we consider the distribution of a random variable $Y = u(X)$, given a random variable\ $X$ with known distribution, and a function $u(\cdot)$.
Among several available techniques, three are considered:

1. the change of variable method (Sect.\ \@ref(ChangeOfVariable));
2. the distribution function method for continuous random variable only (Sect.\ \@ref(DistributonFunctionMethod));
3. the moment generating function method (Sect.\ \@ref(TransformationMoments)).

An important concept in this context is a *one-to-one transformation*.


:::{.definition #OneOneTransformation name="One-to-one transformation"}
Given random variables\ $X$ and\ $Y$ with ranges\ $\mathcal{R}_X$ and\ $\mathcal{R}_Y$ respectively, the function\ $u(\cdot)$ is a *one-to-one transformation* (or mapping) if, for each $y\in \mathcal{R}_Y$, there corresponds exactly one $x\in \mathcal{R}_X$.
:::


When $Y = u(X)$ is a one-to-one transformation, the inverse function is uniquely defined; that is, $X$ can be written uniquely in terms of\ $Y$. 
This is important when considering the distribution of\ $Y$ when the distribution of\ $X$ is known.


:::{.example #OneToOneTransformation}
The transformation $Y = (X - 1)^2$ is not a one-to-one transformation for $\mathcal{R}_X = \mathbb{R}$; that is, if $X$ is defined on $(-\infty, +\infty)$.
For example, the inverse transformation is $X = 1 \pm \sqrt{Y}$, and two values of\ $X$ exists for any given value of $Y > 0$ (Fig.\ \@ref(fig:TwoTransforms), left panel).

However, if the random variable\ $X$ is only defined for $X > 2$, then the transformation is a one-to-one function  (Fig.\ \@ref(fig:TwoTransforms), right panel).
:::


```{r TwoTransforms, echo=FALSE, fig.align="center", fig.cap="Two transformations: a non-one-to-one transformation (left panel), and a one-to-one transformation (right panel) where the random variable is defined only on $X > 2$.", fig.height=4, out.width='80%', fig.width=7}
par(mfrow = c(1, 2))

x <- seq(-2, 4, 
         length = 500)
y <- (x - 1)^2

x1 <- seq(2, max(x) + 1,
          length = 500)
y1 <- (x1 - 1)^2

plot( y ~ x,
      type = "l",
      las = 1,
      lwd = 3,
      col = plotColour1,
      xlab = expression(italic(X)),
      ylab = expression(italic(Y)),
      main = "A transformation that is\nnot one-to-one")

plot( y ~ x,
      type = "l",
      las = 1,
      col = "grey",
      lwd = 3,
      lty = 2,
      xlab = expression(italic(X)),
      ylab = expression(italic(Y)),
      main = "A transformation that is\none-to-one")
lines( y1 ~ x1,
       col = plotColour1,
       lwd = 3)
lines(x = c(min(x1), min(x1)),
      y = c(0, min(y1)),
      lty = 3)
lines(x = c(max(x1), max(x1)),
      y = c(0, max(y1)),
      lty = 3)
```




## The change of variable method {#ChangeOfVariable}

\index{Transformations of random variables!change of variable method|(}
The method is relatively straightforward for one-to-one transformations (such as $Y = 1 - X$ or $Y = \exp(X)$).
Considerable care needs to be exercised if the transformation is not one-to-one; examples are given below.
The discrete and continuous cases are considered separately. 


### Discrete random variables {#ChangeOfVarDiscrete}

Let\ $X$ be a discrete random variable with probability function $p_X(x)$, and let\ $\mathcal{R}_X$ denote the set of discrete points for which $p_X(x) > 0$.
Let $Y = u(X)$ define a *one-to-one transformation* that maps\ $\mathcal{R}_X$ onto\ $\mathcal{R}_Y$, the set of discrete points for which the transformed variable\ $Y$ has a non-zero probability. 
If we solve $Y = u(X)$ for\ $X$ in terms of $Y$, say $X = w(Y) = u^{-1}(Y)$, then for each $y \in \mathcal{R}_Y$, we have $x = w(y)\in \mathcal{R}_X$.


:::{.example #Transform1 name="One-to-one transformation"}
Suppose
$$
   p_X(x) = 
   \begin{cases}
      x/15 & \text{for $x = 1, 2, 3, 4, 5$};\\
      0    & \text{elsewhere}.
   \end{cases}
$$
To find the probability function of\ $Y$ where $Y = 2X + 1$ (i.e., $u(x) = 2x + 1$), first see that $\mathcal{R}_X = \{1, 2, 3, 4, 5\}$.
Hence $\mathcal{R}_Y = \{3, 5, 7, 9, 11\}$ and the mapping $y = 2x + 1 = u(x)$ is one-to-one. 
Also, $w(y) = u^{-1}(y) = (y - 1)/2$.
Hence,
$$
   \Pr(Y = y)
   = \Pr(2X + 1 = y)
   = \Pr\left(X = \frac{y - 1}{2}\right)
   = \left(\frac{y - 1}{2}\right) \times\frac{1}{15}
   = \frac{y - 1}{30}. 
$$
So the probability function of\ $Y$ is
$$
   \Pr(Y = y)
   = \begin{cases}
      (y - 1)/30 & \text{for $y = 3, 5, 7, 9, 11$};\\
      0          & \text{elsewhere}.
    \end{cases}
$$
(Note: The probabilities in this probability function add to\ $1$.)
:::


The above procedure when $Y = u(X)$ is a one-to-one mapping can be stated generally as
\begin{align*}
   \Pr(Y = y) 
   &= \Pr\big(u(X) = y\big)\\
   &= \Pr\big(X = u^{-1} (y)\big)\\
   &= p_X\big(u^{-1}(y)\big), \quad\text{for $y\in \mathcal{R}_Y$}.
\end{align*}


:::{.example #Transform2 name="Transformation (1:1)"}
Let $X$ have a binomial distribution with probability function
$$
   p_X(x) = \begin{cases}
               \binom{3}{x}0.2^x (0.8)^{3 - x} & \text{for $x = 0, 1, 2, 3$};\\
               0 & \text{otherwise}.
            \end{cases}
$$
To find the probability function of $Y = X^2$, first note that $Y = X^2$ is *not* a one-to-one transformation in general, but is one-to-one over the range of\ $X$ (i.e., for $x = 0, 1, 2, 3$).

The transformation $y = u(x) = x^2$, $\mathcal{R}_X = \{ x \mid x = 0, 1, 2, 3 \}$ maps onto $\mathcal{R}_Y = \{y \mid y = 0, 1, 4, 9\}$.
The inverse function is $x = w(y) = \sqrt{y}$, and hence the probability function of $Y$ is
$$
   p_Y(y) = p_X(\sqrt{y})
   = \begin{cases}
               \binom{3}{\sqrt{y}}0.2^{\sqrt{y}} (0.8)^{3 - \sqrt{y}} & \text{for $y = 0, 1, 4, 9$};\\
               0 & \text{otherwise}.
     \end{cases}
$$
:::


When the transformation \ $u(\cdot)$ is *not* 1:1, more care is needed.


:::{.example #TransformNot11 name="Transformation not 1:1"}
Suppose $\Pr(X = x)$ is defined as in Example\ \@ref(exm:Transform1), and define $Y = |X - 3|$.
Since $\mathcal{R}_Y = \{0, 1, 2\}$ the mapping is *not* one-to-one: the event $Y = 0$ occurs if $X = 3$, the event $Y = 1$ occurs if $X = 2$ or $X = 4$, and the event $Y = 2$ occurs if $X = 1$ or $X = 5$.
Hence, $\mathcal{R}_Y  \{ 0, 1, 2\}$.

To find the probability distribution of\ $Y$:
\begin{align*}
   \Pr(Y = 0) 
   &= \Pr(X = 3) = 3/15 = \frac{1}{5};\\
   \Pr(Y = 1) 
   &= \Pr(X = 2 \text{ or } 4) = \frac{2}{15} + \frac{4}{15} = \frac{2}{5};\\
   \Pr(Y = 2) 
   &= \Pr(X = 1 \text{ or } 5) = \frac{1}{15} + \frac{5}{15} = \frac{2}{5}.
\end{align*}
The probability function of $Y$ is
$$
   p_Y(y) = 
   \begin{cases}
       1/5 & \text{for $y = 0$};\\
       2/5 & \text{for $y = 1$};\\
       2/5 & \text{for $y = 2$};\\
       0   & \text{elsewhere}.
   \end{cases}
$$
:::


```{r CoVnotOneToOne, echo=FALSE, fig.align="center", fig.cap="The PMF for $X$ and for the transformed variable $Y$.", fig.width=8, out.width="70%", fig.height=4}
x <- 1:5
px <- x/15

y <- 0:2
py <- c(1, 2, 2)/5

par(mfrow = c(1, 2))

plotDiscrete(x, px,
             type = "MF",
             ylim = c(0, 0.40),
             main = expression(bold(The~PMF~"for"~italic(X))),
             xlab = expression(italic(X)),
             ylab = "PMF")


plotDiscrete(y, py,
             type = "MF",
             ylim = c(0, 0.45),
             main = expression(bold(The~PMF~"for"~italic(Y))),
             xlab = expression(italic(Y)),
             ylab = "PMF")

par(mfrow = c(1,1))
```


### Continuous random variables {#ChangeOfVarContinuous}

The change of variable method is similar for continuous random variables, but requires the use of a *Jacobian*.\index{Jacobian}
For a discrete random variable, probability exists at distinct values of the random variable, but probability only exists over *intervals* for continuous random variables.
Suppose the original variable\ $X$ has a given probability defined over (say) $0 < X < 1$; the equivalent probability in the transformed variable\ $Y$ may be defined over (say) $-4 < Y  < 4$.
The Jacobian is needed to ensure the height of the corresponding density function accommodates this *local* change in the length of the interval (which may not be linear) over which the equivalent probability is defined.


:::{.theorem #ChangeOfVarCont name="Change of variable (continuous rv)"}
If\ $X$ has PDF $f_X(x)$ for $x\in \mathcal{R}_X$ and\ $u(\cdot)$ is a one-to-one function for $x\in \mathcal{R}_X$, then the random variable $Y = u(X)$ has PDF
$$
   f_Y(y) = f_X(x) \left|\frac{dx}{dy}\right|
$$
where the right-hand side is expressed as a function of\ $y$.
The term $\left|dx/dy\right|$ is called the *Jacobian of the transformation*.\index{Jacobian}

::::{.proof}
Let the inverse function be $X = w(Y)$ so that $w(y) = u^{-1}(x)$.

**Case\ 1:** $y = u(x)$ is a strictly *increasing* function (Fig.\ \@ref(fig:Transformation), left panel).
If $a < y < b$ then $w(a) < x < w(b)$ and  $\Pr(a < Y < b) = \Pr\big(w(a) < X <w(b)\big)$, so
$$
   {\int^b_a f_Y(y)\,dy
   =\int^{w(b)}_{w(a)}f_X(x)\,dx
   =\int^b_af\big( w(y)\big)\frac{dx}{dy}\,\,dy}.
$$
Therefore, $\displaystyle {f_Y(y) = f_X\big( w(y) \big)\frac{dx}{dy}}$, where $w(y) = u^{-1}(x)$.


**Case\ 2:** $y = u(x)$ is a strictly *decreasing* function of $x$ (Fig.\ \@ref(fig:Transformation), right panel).
If $a < y < b$ then $w(b) < x < w(a)$ and $\Pr(a < Y < b) = \Pr\big(w(b) < X < w(a)\big)$, so that
\begin{align*}
     \int^b_a f_Y(y)\,dy & = \int^{w(a)}_{w(b)}f_X(x)\,dx\\
     & = \int^a_bf_X(x)\frac{dx}{dy}\,\,dy\\
     & = - \int ^b_a f_X(x)\frac{dx}{dy}\,dy.
\end{align*}
Therefore $f_Y(y) = -f_X\left( w(y) \right)\displaystyle{\frac{dx}{dy}}$.
But $dx/dy$ is negative in the case of a decreasing function, so in general
$$
   f_Y(y) = f_X(x)\left|\frac{dx}{dy} \right|. 
$$
::::
:::



```{r Transformation, echo=FALSE, fig.align="center", fig.cap="A strictly increasing transformation function (left panel) and strictly decreasing function (right panel).", fig.width=9, fig.height=4, out.width='100%'}
par( mfrow = c(1, 2))

logit <- function(x) { 
  log(x / (1 - x)) 
  }

y <- seq(0.01, 1,
         length = 100)
x <- logit(y)

y0 <- 0.4
y1 <- 0.8

x0 <- logit( y0 )
x1 <- logit( y1 )

plot( x = x,
      y = y,
      lwd = 3,
      col = plotColour1,
      axes = FALSE,
      type = "l",
      main = "Increasing function",
      xlab = expression(italic(x)),
      ylab = expression(italic(y)==italic(u)(italic(x))))
axis(side = 2,
     las = 1,
     at = c(y0, y1),
     labels = c(expression(italic(a)), 
                expression(italic(b))) ) 
axis(side = 1,
     at = c(x0, x1 ),
     labels = c(expression(italic(w)*paste("(",italic(a),")")),
                expression(italic(w)*paste("(",italic(b),")")) ) )
box()

lines( x = c( min(x), x0, x0),
       y = c(y0, y0, 0),
       lty = 2)
lines( x = c( min(x), x1, x1),
       y = c(y1, y1, 0),
       lty = 2)

text(x = 3.6,
     y = 0.85,
     label = expression(italic(y)==italic(u)(italic(x))))



logit2 <- function(x) { 
  -log(x / (1 - x)) 
  }

y <- seq(0.01, 0.99,
         length = 100)
x <- logit2(y)

y0 <- 0.4
y1 <- 0.8

x0 <- logit2( y0 )
x1 <- logit2( y1 )

plot( x = x,
      y = y,
      lwd = 3,
      col = plotColour1,
      axes = FALSE,
      type = "l",
      main = "Decreasing function",
      xlab = expression(italic(x)),
      ylab = expression(italic(y)==italic(u)(italic(x))))
axis(side = 2,
     las = 1,
     at = c(y0, y1),
     labels = c(expression(italic(a)), 
                expression(italic(b))) ) 
axis(side = 1,
     at = c(x0, x1),
     labels = c(expression(italic(w)*paste("(",italic(a),")")),
                expression(italic(w)*paste("(",italic(b),")")) ) )
box()

lines( x = c( min(x), x0, x0),
       y = c(y0, y0, 0),
       lty = 2)
lines( x = c( min(x), x1, x1),
       y = c(y1, y1, 0),
       lty = 2)

text(x = 2,
     y = 0.7,
     label = expression(italic(y)==italic(u)(italic(x))))
```


:::{.example #Transform3 name="Transformation"}
Let the PDF of\ $X$ be given by
$$
   f_X(x) = 
   \begin{cases}
      1 & \text{for $0 < x < 1$};\\
      0 & \text{elsewhere}.
   \end{cases}
$$
Consider the transformation $Y = u(X) = -2\log X$ (where $\log$ refers to logarithms to base\ $e$, or *natural logarithms*).
The transformation is one-to-one, and the inverse transformation is
$$
   X = \exp( -Y/2) = u^{-1}(Y) = w(Y).
$$
The space $\mathcal{R}_X = \{x \mid 0 < x < 1\}$ is mapped to $\mathcal{R}_Y = \{y \mid 0 < y < \infty\}$.
Then,
$$
   w'(y) = \frac{d}{dy} \exp(-y/2) = -\frac{1}{2}\exp(-y/2),
$$
and so the *Jacobian* of the transformation $|w'(y)| = \exp(-y/2)/2$.\index{Jacobian}
The PDF of $Y = -2\log X$ is
\begin{align*}
   f_Y(y)
   &= f_X\big(w(y)\big) |w'(y)| \\
   &= f_X\big(\exp(-y/2)\big) \exp(-y/2)/2 \\
   &= \frac{1}{2}\exp(-y/2)\quad\text{for $y > 0$}.
\end{align*}
That is, $Y$ has an exponential distribution with $\beta = 2$:\index{Beta distribution} $Y \sim \text{Exp}(2)$ (Def.\ \@ref(def:ExponentialDistribution)).
:::


:::{.example #TransformSquRt name="Square root transformation"}
Consider the random variable\ $X$ with PDF $f_X(x) = \exp(-x)$ for $x \geq 0$.
To find the PDF of $Y = \sqrt{X}$, first see that $y = \sqrt{x}$ is a strictly increasing function for $x \geq 0$ (Fig.\ \@ref(fig:SqrtRt)).

The inverse relation is $X = Y^2$, and $dx/dy = |2y| = 2y$ for $x \ge 0$.
The PDF of\ $Y$ is
$$
    f_Y(y)
    = f_X(x)\left|\frac{dx}{dy}\right|
    = 2y \exp(-y^2)\quad \text{for $y\geq0$}.
$$
:::




```{r SqrtRt, echo=FALSE, fig.align="center", fig.cap="The square-root transformation (left panel); the PDF of $X$ (centre panel) and the PDF of $Y$ (right panel).", fig.width=7, fig.height=2.5, out.width='100%'}
par(mfrow = c(1, 3))

x <- seq(0, 20,
         length = 100)
y <- sqrt(x)

plot(x = x,
     y = y,
     lwd = 3,
     col = plotColour1,
     las = 1,
     main = expression(bold(The~transformation~italic(y) == sqrt(italic(x)))),
     xlab = expression(italic(x)),
     ylab = expression( italic(y) == sqrt(italic(x))),
     type = "l")




x <- seq(0, 5,
         length = 500)
fx <- exp(-x)

plot(fx ~ x,
     xlab = expression(italic(X)),
     ylab = "Prob. fn.",
     main = expression(bold(PF~"for"~italic(X))),
     type="l",
     las = 1,
     lwd = 3,
     col = plotColour1)






y <- seq(0, 5,
         length = 500)
fy <- 2 * y * exp(-y^2)

plot(fy ~ y,
     xlab = expression(italic(Y)),
     ylab = "Prob. fn.",
     main = expression(bold(PF~"for"~italic(Y))),
     type="l",
     las = 1,
     lwd = 3,
     col = plotColour1)
```


::: {.example #TransformationTan name="Tan transformation"}
Let random variable\ $X$ be uniformly distributed on $[-\pi/2, \pi/2]$. 
Suppose we seek the distribution of $Y = \tan X$ (Fig.\ \@ref(fig:TanXform)).

For the mapping $Y = \tan X$, we see that $\mathcal{R}_Y = \{ y\mid -\infty <y<\infty\}$.
The mapping is one-to-one, and so $X = \tan^{-1}Y$, and $dx/dy = 1/(1 + y^2)$.
Hence
$$
   f_Y(y)
   = f_X(x)\left|\frac{dx}{dy}\right|
   = \frac{1}{\pi(1 + y^2)}.
$$
This is the *Cauchy distribution*.\index{Cauchy distribution}
:::
<!-- CONNECT TO KICKING A FOOTBALL -->


```{r TanXform, echo=FALSE, fig.align="center", fig.cap="The tan transformation (left panel); the PDF of $X$ (centre panel) and the PDF of $Y$ (right panel).", fig.width=7, fig.height=2.5, out.width='100%'}
par(mfrow = c(1, 3))

x <- seq(-pi/2 + 0.1, 
         pi/2 - 0.1,
         length = 100)
y <- tan(x)

plot(x = x,
     y = y,
     lwd = 3,
     col = plotColour1,
     las = 1,
     axes = FALSE,
     main = expression(bold(The~transformation~italic(y) == plain(tan)(italic(x)))),
     xlab = expression(italic(x)),
     ylab = expression( italic(y) == plain(tan)(italic(x))),
     type = "l")
axis(side = 2,
     las = 1)
axis(side = 1,
     at = c(-pi/2, 0, pi/2),
     labels = c(expression(-pi/2),
                "0",
                expression(pi/2)) )
abline(h = 0, 
       lty = 1,
       col = "grey")
abline(v = 0, 
       lty = 1,
       col = "grey")
box()




x <- seq(-pi/2, pi/2,
         length = 500)
fx <- rep(1/pi, length(x) )

plot(fx ~ x,
     xlab = expression(italic(X)),
     ylab = "Prob. fn.",
     main = expression(bold(PF~"for"~italic(X))),
     type="l",
     ylim = c(0, 0.35),
     xlim = c(-2, 2),
     las = 1,
     lwd = 3,
     col = plotColour1)
lines( x = c(-pi/2, -pi/2),
       y = c(0, 1/pi),
       lty = 3,
       col = "grey")
lines( x = c(pi/2, pi/2),
       y = c(0, 1/pi),
       lty = 3,
       col = "grey")


#######################

y <- seq(-3, 3,
         length = 500)
fy <- 1 / (1 + y^2)

plot(fy ~ y,
     xlab = expression(italic(Y)),
     ylab = "Prob. fn.",
     main = expression(bold(PF~"for"~italic(Y))),
     type="l",
     las = 1,
     lwd = 3,
     col = plotColour1)
```


A case where the function\ $u(\cdot)$ is *not* a one-to-one transformation is considered using an example, using a modification of Theorem\ \@ref(thm:ChangeOfVarCont).


:::{.example #TransformationNon11 name="Transformation (not 1:1)"}
Given a random variable\ $Z$ which follows a $N(0, 1)$ distribution, suppose we seek the probability distribution of $Y = \frac{1}{2} Z^2$.

The relationship $Y = u(Z) = \frac{1}{2}z^2$ is not strictly increasing or decreasing in $(-\infty, \infty )$ so Theorem\ \@ref(thm:ChangeOfVarCont) cannot be applied directly.
Instead, subdivide the range of $Z$ and $Y$ so that in each portion the relationship *is* monotonic.
Then:
$$
   f_Z(z) =
   \frac{1}{\sqrt{2\pi}}\,\exp\left( -\frac{1}{2} z^2\right)\quad\text{for $-\infty < z < \infty$}.
$$
The inverse relation, $Z = u^{-1}(Y)$ is $Z = \pm \sqrt{2Y}$. 
For a given value of $Y$, two values of $Z$ are possible. 
However, in the range $-\infty < z < 0$, $Y$ and $Z$ have a monotonic relationship. 
Similarly, for $0 < z <\infty$, $Y$ and $Z$ have a monotonic relationship. 
Thus (see Fig.\ \@ref(fig:Non11Xform)),
$$
   \Pr(a < Y <b) = \Pr(-\sqrt{2b} < Z < -\sqrt{2a}\,) + \Pr(\sqrt{2a} < Z < \sqrt{2b}\,).
$$
The two terms on the right are equal because the distribution of\ $Z$ is symmetrical about\ $z = 0$.
Thus $\Pr(a < Y < b) = 2\Pr(\sqrt{2a} < Z < \sqrt{2b}\,)$, and
\begin{align*}
     f_Y(y)
     &= 2f_Z(z)\left| \frac{dz}{dy}\right|\\
     &= 2\frac{1}{\sqrt{2\pi}}\exp(-y)\frac{1}{\sqrt{2y}};
\end{align*}
that is,
$$
   f_Y(y)
   = e^{-y}y^{-\frac{1}{2}} / \sqrt{\pi}\quad\text{for $0 < y < \infty$}.
$$
This PDF is a [gamma distribution](GammaDistribution) with parameters $\alpha = 1/2$ and $\beta = 1$.
It follows that if $X\sim N(\mu,\sigma^2)$, then the PDF of $Y = \frac{1}{2} (X - \mu )^2 / \sigma^2$ is $\text{Gamma}(\alpha = 1/2,\beta = 1)$ since then $(X - \mu)/\sigma$ is distributed as $N(0, 1)$.
:::



```{r Non11Xform, echo=FALSE, fig.align="center", fig.cap="A transformation not 1:1.", fig.height=3.5, fig.width=4.5, out.width='50%'}
x <- seq(-5, 5,
         length = 100)
square <- function(x) { x^2 }

y <- square(x)

x1 <- -4
x2 <- -2
x3 <- 2
x4 <- 4

y1 <- square(x1)
y2 <- square(x2)
y3 <- square(x3)
y4 <- square(x4)

plot(x = x,
     y = y,
     lwd = 3,
     col = plotColour1,
     las = 1,
     axes = FALSE,
     main = "A non-1:1 transformation",
     xlab = expression(italic(z)),
     ylab = expression(italic(y) == italic(z)^2),
     type = "l")
axis(side = 1,
     at = c(x1, x2, x3, x4),
     labels = c( expression(-sqrt(2)),
                 expression(-sqrt(-2*italic(a))),
                 expression(sqrt(2*italic(a))),
                 expression(sqrt(2)) ) )
axis(side = 2,
     las = 1,
     at = c(y1, y2),
     label = c(expression(italic(b)),
               expression(italic(a))))
box()


lines( x = c(x1, x1, x4, x4),
       y = c(0, y1, y4, 0),
       lty = 2)
lines( x = c(x2, x2, x3, x3),
       y = c(0, y2, y3, 0),
       lty = 2)

abline(h = 0,
       col = "grey")
abline(v = 0,
       col = "grey")

```


Note that the probability can only be doubled as in Example\ \@ref(exm:TransformationNon11) if both $Y = u(Z)$ and the PDF of\ $Z$ are symmetrical about the same point.
\index{Transformations of random variables!change of variable method|)}


### Bivariate discrete case {#ChangeOfVarDiscreteBivariate}

The bivariate discrete case is similar to the univariate discrete case.
Consider a joint probability function $p_{X_1, X_2}(x_1, x_2)$ of two discrete random variables\ $X_1$ and\ $X_2$ defined on the two-dimensional set of points\ $R^2_X$ for which $p(x_1, x_2) > 0$.
Two *one-to-one transformations* need to be defined:
$$
   y_1 = u_1( x_1, x_2)\qquad\text{and}\qquad y_2 = u_2( x_1, x_2)
$$
that map\ $R^2_X$ onto\ $R^2_Y$ (the two-dimensional set of points for which $p(y_1, y_2) > 0$).
The two inverse functions are
$$
   x_1 = w_1( y_1, y_2)\qquad\text{and}\qquad x_2 = w_2( y_1, y_2).
$$
Then the joint probability function of the new (transformed) random variables is
$$
   p_{Y_1, Y_2}(y_1, y_2) =
   \begin{cases}
      p_{X_1, X_2}\big( w_1(y_1, y_2), w_2(y_1, y_2)\big) & \text{where $(y_1, y_2)\in R^2_Y$};\\
      0 & \text{elsewhere}.
   \end{cases}
$$


:::{.example #TransformBivariate name="Transformation discrete (bivariate)"}
Let the two discrete random variables\ $X_1$ and\ $X_2$ have the joint probability function shown in Table\ \@ref(tab:Bivar4).
Consider the two one-to-one transformations
$$
   Y_1 = X_1 + X_2 \qquad\text{and}\qquad Y_2 = 2 X_1.
$$
The joint probability function of\ $Y_1$ and\ $Y_2$ can be found by noting where the $(x_1, x_2)$ pairs are mapped to in the $y_1, y_2$ space:

| $(x_1, x_2)$ | $\mapsto$ | $(y_1,y_2)$
|:-----------:|:----------:|:----------------:
| $(-1, 0)$   | $\mapsto$  | $(-1, -2)$ 
| $(-1, 1)$   | $\mapsto$  | $(0, -2)$
| $(-1, 2)$   | $\mapsto$  | $(1, -2)$
| $(1, 0)$    | $\mapsto$  | $(1, 2)$ 
| $(1, 1)$    | $\mapsto$  | $(2, 2)$ 
| $(1, 2)$    | $\mapsto$  | $(3, 2)$ 


The joint probability function can then be constructed as shown in Table\ \@ref(tab:Bivar5).
:::




```{r Bivar4, echo=FALSE}
Bivar4Table <- array( dim = c(2, 4))

Bivar4Table[1, ] <- c("$x_1 = -1$", "$0.3$", "$0.1$", "$0.1$")
Bivar4Table[2, ] <- c("$x_1 = +1$", "$0.2$", "$0.2$", "$0.1$")


Bivar4Table.caption <- "A bivariate probability function for $X_1$ and $X_2$."
if( knitr::is_latex_output() ) {
  knitr::kable(Bivar4Table,
               format = "latex",
               booktabs = TRUE,
               col.names = c("",
                             "$x_2 = 0$",
                             "$x_2 = 1$",
                             "$x_2 = 2$"),
               align = "r",
               longtable = FALSE,                
               escape = FALSE,
               caption = Bivar4Table.caption) %>%
    kable_styling(font_size = 10) %>%
    row_spec(0, bold = TRUE) 
}
if( knitr::is_html_output() ) {
  knitr::kable(Bivar4Table,
               escape = TRUE,
               format = "html",
               booktabs = TRUE,
               align = "r",
               longtable = FALSE,                
               col.names = c("",
                             "$x_2 = 0$",
                             "$x_2 = 1$",
                             "$x_2 = 2$"),
               caption = Bivar4Table.caption) %>%
    row_spec(0, bold = TRUE) 
}
```





```{r Bivar5, echo=FALSE}
Bivar5Table <- array( dim = c(2, 6))

Bivar5Table[1, ] <- c("$y_2 = -2$", "$0.3$", "$0.1$", "$0.1$", "$0.0$", "$0.0$")
Bivar5Table[2, ] <- c("$y_2 = +2$", "$0.0$", "$0.0$", "$0.2$", "$0.2$", "$0.1$")


Bivar5Table.caption <- "The joint probability function for $Y_1$ and $Y_2$"
if( knitr::is_latex_output() ) {
  knitr::kable(Bivar5Table,
               format = "latex",
               booktabs = TRUE,
               col.names = c("",
                             "$y_1 = -1$",
                             "$y_2 = 0$",
                             "$y_3 = 1$",
                             "$y_4 = 2$",
                             "$y_5 = 3$"),
               align = "r",
               longtable = FALSE,                
               escape = FALSE,
               caption = Bivar5Table.caption) %>%
    kable_styling(font_size = 10) %>%
    row_spec(0, bold = TRUE) 
}
if( knitr::is_html_output() ) {
  knitr::kable(Bivar5Table,
               escape = TRUE,
               format = "html",
               booktabs = TRUE,
               align = "r",
               longtable = FALSE,                
               col.names = c("",
                             "$y_1 = -1$",
                             "$y_1 = 0$",
                             "$y_1 = 1$",
                             "$y_1 = 2$",
                             "$y_1 = 3$"),
               caption = Bivar5Table.caption) %>%
    row_spec(0, bold = TRUE) 
}
```


Sometimes, a joint probability function of two random variables is given, but only *one* new random variable is required.
In this case, a second (dummy) transformation is used, usually a very simple transformation.


:::{.example #TransformBivariate2 name="Transformation (bivariate)"}
Let\ $X_1$ and\ $X_2$ be two independent random variables with the joint probability function
$$
   p_{X_1, X_2}(x_1, x_2) =
         \frac{\mu_1^{x_1} \mu_x^{x_2} \exp( -\mu_1 - \mu_2 )}{x_1!\, x_2!}
         \quad\text{for $x_1$ and $x_2 = 0, 1, 2, \dots$}
$$
This is the joint probability function of two independent Poisson random variables.
Suppose we wish to find the probability function of $Y_1 = X_1 + X_2$.

Consider the two *one-to-one transformations*, where $Y_2 = X_2$ is just a dummy transformation:
\begin{align}
   y_1 &= x_1 + x_2             = u_1(x_1, x_2)\\
   y_2 &= x_2\phantom{{} + x_2} = u_2(x_1, x_2)
\end{align}
which map the points in\ $R^2_X$ onto
$$
   R^2_Y = \left\{ (y_1, y_2)\mid y_1 = 0, 1, 2, \dots; y_2 = 0, 1, 2, \dots, y_1\right\}.
$$
$Y_2$ is a dummy transform, and is chosen to be very simple; any second transform could be chosen (as it is not of direct interest).
The inverse functions are
\begin{align*}
   x_1 &= y_1 - y_2              = w_1(y_1, y_2)\\
   x_2 &= y_2 \phantom{{} - y_2} = w_2(y_2)
\end{align*}
by rearranging the original transformations.
Then the *joint* probability function of\ $Y_1$ and\ $Y_2$ is
\begin{align*}
   p_{Y_1, Y_2}(y_1, y_2)
   &= p_{X_1, X_2}\big( w_1(y_1, y_2), w_2(y_1, y_2)\big) \\
   &= \frac{\mu_1^{y_1 - y_2}\mu_2^{y_2} \exp(-\mu_1 - \mu_2)}{(y_1 - y_2)! y_2!}\quad
   \text{for $(y_1, y_2)\in R^2_Y$}.
\end{align*}
Recall that we seek the probability function of just\ $Y_1$, so we need to find the marginal probability function of $p_{Y_1, Y_2}(y_1, y_2)$.
The marginal probability function of\ $Y_1$ is
$$
   p_{Y_1}(y_1) = \sum_{y_2 = 0}^{y_1} p_{Y_1, Y_2}(y_1, y_2)
   = \sum_{y_2 = 0}^{y_1} \frac{\mu_1^{y_1 - y_2}\mu_2^{y_2} \exp(-\mu_1 - \mu_2)}{(y_1 - y_2)!\, y_2!},
$$
which is equivalent to
$$
   p_{Y_1}(y_1) =
   \begin{cases}
      \displaystyle{\frac{(\mu_1 + \mu_2)^{y_1}\exp\big[-(\mu_1 + \mu_2)\big]}{y_1!}} & \text{for $y_1 = 0, 1, 2, \dots$}\\
      0 & \text{otherwise}.
   \end{cases}
$$
You should recognise this as the probability function of a Poisson random variable (Def.\ \@ref(def:PoissonDistribution)) with mean $\mu_1 + \mu_2$.
Thus $Y_1 \sim \text{Poisson}(\lambda = \mu_1 + \mu_2)$.
:::


### Bivariate continuous case {#ChangeOfVarContinuousBivariate}

With two continuous random variables $(X_1, X_2)$ with a joint PDF $f_{X_1, X_2}(x_1, x_2)$, two transformations are needed to transform them into two new random variables $(Y_1, Y_2)$:
$$
  Y_1 = u_1(X_1, X_2) \quad \text{and} \quad Y_2 = u_2(X_1, X_2).
$$
The joint PDF of the new variables is
$$
  f_{Y_1, Y_2}(y_1, y_2) = f_{X_1, X_2}(x_1, x_2) \cdot |J|
$$
where $|J|$ is the *absolute value* of the *Jacobian*\index{Jacobian} determinant:
$$
  J = \det 
        \begin{bmatrix} 
          {\partial x_1}/{\partial y_1} &  {\partial x_1}/{\partial y_2} \\ 
          {\partial x_2}/{\partial y_1} &  {\partial x_2}/{\partial y_2} 
        \end{bmatrix}.
$$
The Jacobian plays an equivalent role as $\left|dx/dy\right|$ in the univariate case.


:::{.example #BivarContExample name="Transformation continuous (bivariate)"}
Let\ $X_1$ and\ $X_2$ be independent exponential random variables with $\lambda = 1$, such that both PDFs are
$$
  f_X(x) = \exp(-x)\quad\text{for $x > 0$}.
$$
Consider the two transformations $Y_1 = X_1 + X_2$ (the sum) and $Y_2 = X_1 / X_2$ (the ratio).
Then, $y_1 > 0$ and $y_2 > 0$.

Since the random variable\ $X_1$ and\ $X_2$ are independent, the joint PDF is
$$
  f_{X_1, X_2}(x_1, x_2) = \exp(-x_1) \cdot \exp(-x_2) = \exp\left(-(x_1 + x_2)\right) \quad \text{for $x_1, x_2 > 0$}.
$$
Inverting the transformations gives $y_1 = x_1 + x_2$ and $y_2 = x_1 / x_2$ (and so $x_1 = y_2 x_2$).
These give
$$
  x_1 = \frac{y_1 y_2}{1 + y_2}
  \quad\text{and}\quad
  x_2 = \frac{y_1}{1 + y_2}.
$$
The Jacobian requires the four partial derivatives:
$$
\begin{array}{ll}
  \displaystyle \frac{\partial x_1}{\partial y_1} = \frac{y_2}{1+y_2};\quad & \quad\displaystyle \frac{\partial x_1}{\partial y_2} = \frac{y_1(1+y_2) - y_1 y_2}{(1+y_2)^2} = \frac{y_1}{(1+y_2)^2};\\[6pt]
  \displaystyle \frac{\partial x_2}{\partial y_1} = \frac{1}{1+y_2};\quad & \quad\displaystyle \frac{\partial x_2}{\partial y_2} = \frac{-y_1}{(1+y_2)^2}.
\end{array}
$$
So the determinant of the Jacobian is 
\begin{align*}
  \det J 
  &= \left( \frac{y_2}{1+y_2} \cdot \frac{-y_1}{(1+y_2)^2} \right) - \left( \frac{1}{1+y_2} \cdot \frac{y_1}{(1+y_2)^2} \right)\\
  &= \frac{-y_1 y_2 - y_1}{(1+y_2)^3}\\
  &= \frac{-y_1(y_2 + 1)}{(1+y_2)^3}.
\end{align*}
Finally, substitute\ $x_1$ and\ $x_2$ and\ $|J|$ into the formula:
$$
  f_{Y_1, Y_2}(y_1, y_2) = \exp(-(x_1 + x_2)) \cdot \left| \frac{-y_1}{(1 + y_2)^2} \right|.
$$
Since $x_1 + x_2 = y_1$, write
$$
  f_{Y_1, Y_2}(y_1, y_2) = \frac{y_1 \exp(-y_1)}{(1+y_2)^2}\quad\text{for $y_1 > 0$ and $y_2 > 0$}.
$$
:::


### Summary

The change of variable method requires find the pre-image, then accounting for the how transformation redistributes probability (through counting in the discrete case; through the Jacobian in the continuous case); see Table\ \@ref(tab:CoVSummary).


```{r CoVSummary, echo=FALSE}
CoVsummary <- array( dim = c(2, 3))

CoVsummary[1, ] <- c("Univariate",
                     "Sum $p_X$ over every $x$ that transforms into $y$",
                     "Substitute inverse; multiply by $\\|J\\|$")
CoVsummary[2, ] <- c("Bivariate",
                     "Sum $p_{(X_1, X_2)}$ over every $(x_1, x_2)$ that transforms into $(y_1, y_2)$",
                     "Substitute inverses; multiply by $\\|J\\|$ (det of matrix)" )

colnames(CoVsummary) <- c("Setting",
                          "Discrete",
                          "Continuous")

if( knitr::is_latex_output() ) {
  knitr::kable(CoVsummary,
               format = "latex",
               booktabs = TRUE,
               align = c("r", "c", "c"),
               longtable = FALSE,                
               escape = FALSE,
               caption = "Summarising the change of variable method.") %>%
    kable_styling(font_size = 10) %>%
    column_spec(2, width = "5cm") %>%
    column_spec(3, width = "5cm") %>%
    row_spec(0, bold = TRUE) %>% 
    column_spec(1, bold = TRUE) 
}
if( knitr::is_html_output() ) {
  knitr::kable(CoVsummary,
               escape = TRUE,
               format = "html",
               booktabs = TRUE,
               align = "r",
               longtable = FALSE,                
               caption = "Summarising the change of variable method.") %>%
    column_spec(2, width = "20%") %>%
    column_spec(2, width = "40%") %>%
    column_spec(3, width = "40%") %>%
    row_spec(0, bold = TRUE) %>% 
    column_spec(1, bold = TRUE)
}

```




## The distribution function method {#DistributonFunctionMethod}

The distribution function method is the most broadly applicable technique for finding the distribution of a transformed random variable. 
The strategy is to:

1. Write $F_Y(y) = \Pr(Y \leq y) = \Pr(g(X) \leq y)$.
2. Re-express the event $\{g(X) \leq y\}$ as an equivalent event in terms of\ $X$.
3. Evaluate using the known distribution of\ $X$.
4. Differentiate (continuous case) or take differences (discrete case) to obtain the PDF: $f_Y(y) = F_Y'(y)$.

The method operates with probabilities rather than densities, so it works with non-monotone transformations, awkward supports, and many-to-one mappings without requiring the special conditions needed with the change of variables method.


### Continuous random variables

\index{Transformations of random variables!distribution function method|(}

The general procedure above is best demonstrated using an example.


:::{.example #DFMethod name="Distribution function method"}
Consider the random variable\ $X$ with PDF
$$
   f_X(x) = \begin{cases}
               x/4 & \text{for $1 < x < 3$};\\
               0 & \text{elsewhere}.
            \end{cases}
$$
To find the PDF of the random variable\ $Y$ where $Y = X^2$, first see that $1 < y < 9$ and the transformation is monotonic over this region.
The distribution function for\ $Y$ is
\begin{align*}
   F_Y(y)
   &= \Pr(Y\le y) \qquad\text{(by definition)}\\
   &= \Pr(X^2 \le y) \qquad\text{(since $Y = X^2$)}\\
   &= \Pr(X\le \sqrt{y}\,).
\end{align*}
This last step is not trivial, but is critical. 
Sometimes, more care is needed (see Example\ \@ref(exm:TransformA)). 
In this case, there is a one-to-one relationship between\ $X$ and\ $Y$ over the region of which\ $X$ is defined (i.e., has a positive probability); see Fig.\ \@ref(fig:SquaringXform).

Then continue as follows:
\begin{align*}
   F_Y(y)
    =\Pr( X\le \sqrt{y}\,)
   &= F_X\big(\sqrt{y}\,\big) \qquad\text{(definition of $F_X(x)$)} \\
   &= \int_1^{\sqrt{y}} (x/4) \,dx
    = (y - 1)/8
\end{align*}
for $1 < y < 9$, and is zero elsewhere. 
This is the *distribution function* of\ $Y$; to find the PDF:
$$
   f_Y(y)
   = \frac{d}{dy} (y - 1)/8
   = \begin{cases}
        1/8 & \text{for $1 < y < 9$};\\
        0 & \text{elsewhere}.
     \end{cases}
$$
Note the range for which\ $Y$ is defined; since $1 < x < 3$, then $1 < y < 9$.
:::


```{r SquaringXform, echo=FALSE, warning=FALSE, fig.align="center", fig.cap="The transformation $Y = X^2$ when $X$ is defined from $1$ to $3$. The thicker line corresponds to the region where the transformation applies. Note that if $Y < y$, then $2 - \\sqrt{y - 1} < X < 2 + \\sqrt{y - 1}$.", fig.width=8, fig.height=4, out.width='100%'}
par( mfrow = c(1, 3))

x <- seq(-1, 4,
         length = 100)
y <- x^2

plot(x = x,
     y = y,
     type = "l",
     col = plotColour1,
     lwd = 1,
     axes = FALSE,
     xlab = expression(italic(x)),
     ylab = expression(italic(y) == italic(x)^2),
     main = "The squaring\ntransformation",
     las = 1)

Thicker <- ( x >= 1 ) & ( x <= 3 )
lines(x = x[ Thicker],
      y = y[Thicker],
      lwd = 5)
lines( x = c( min(x), sqrt(3), sqrt(3)),
       y = c( 3, 3, 0),
       lty = 2)
axis(side = 2,
     at = c(0, 3, 6, 8, 10, 12, 14, 16),
     las = 1,
     labels = c(0, "y = 3", 6, 8, 10, 12, 14, 16))
axis(side = 1,
     at = c(-1, 0, 1, sqrt(3), 3, 4),
     labels = c(-1, 0, 1, 
                expression(italic(x) == sqrt(3) ),
                3, 4))
box()




funx <- function(x){
  y <- array(0, dim = length(x))
  y <- ifelse( (x > 1) & (x < 3), x/4, 0)
}

x <- seq(0, 4,
         length = 500)
fx <- funx(x)

plot(fx ~ x,
     xlab = expression(italic(X)),
     ylab = "Prob. fn.",
     main = expression(bold(PF~"for"~italic(X))),
     type="l",
     las = 1,
     lwd = 2)





funy <- function(y){
   fy <- ifelse( (y > 1) & (y < 2),
                 1/(2 * sqrt(y - 1)),
                 0)  
}

y <- seq(0, 3,
         length = 500)
fy <- funy(y)

plot(fy ~ y,
     xlab = expression(italic(Y)),
     ylab = "Prob. fn.",
     main = expression(bold(PF~"for"~italic(Y))),
     type="l",
     las = 1,
     lwd = 2)
```



:::{.example #TransformA name="Transformation"}
Consider the same random variable\ $X$ as in the previous example, but the transformation $Y = (X - 2)^2 + 1$ (Fig.\ \@ref(fig:SquaringXform2)).

In this case, the transformation is **not** a one-to-one transform.
Proceed as before to find the distribution function of\ $Y$:
\begin{align*}
   F_Y(y)
   &= \Pr(Y\le y) \qquad\text{(by definition)}\\
   &= \Pr\big( (X - 2)^2 + 1  \le y\big)
\end{align*}
since $Y = (X - 2)^2 + 1$.
From Fig.\ \@ref(fig:SquaringXform2), whenever $(X - 2)^2 + 1 < y$ for some value\ $y$, then\ $X$ must be in the range $2 - \sqrt{y - 1}$ to $2 + \sqrt{y - 1}$.
So:
\begin{align*}
   F_Y(y)
   &= \Pr\big( (X - 2)^2 + 1 \le y\big) \\
   &= \Pr\left( 2 - \sqrt{y - 1} < X < 2 + \sqrt{y - 1} \right)\\
   &= \int_{2-\sqrt{y - 1}}^{2 + \sqrt{y - 1}} x/4\,dx \\
   &=  \sqrt{y - 1}.
\end{align*}
Again, this is the distribution function; so differentiating:
$$
   f_Y(y) = \begin{cases}
               \frac{1}{2\sqrt{y - 1}} & \text{for $1 < y < 2$};\\
               0 & \text{elsewhere}.
            \end{cases}
$$
:::





```{r SquaringXform2, echo=FALSE, fig.align="center", fig.cap="The transformation $Y = (X - 2)^2 + 1$ when $X$ is defined from $1$ to $3$. The thicker line corresponds to the region where the transformation applies. Note that if $Y < y$, then $2 - \\sqrt{y - 1} < X < 2 + \\sqrt{y - 1}$.", fig.width=5, fig.height=4}

x <- seq(0, 4,
         length = 100)
xf <- function(x) { (x - 2)^2 + 1}

y <- xf(x)

plot(x = x,
     y = y,
     type = "l",
     lwd = 1,
     col = plotColour1,
     ylim = c(0, 2.25),
     axes = FALSE,
     xlab = expression(italic(x)),
     ylab = expression(italic(y)==group("(", italic(x)-2,")")^2+1),
     main = "Another transformation",
     las = 1)

x0 <- 1.25
x1 <- 2.75
y0 <- xf( x0 )
y1 <- xf( x1 )


Thicker <- ( x >= 1 ) & ( x <= 3 )
lines(x = x[ Thicker],
      y = y[Thicker],
      lwd = 5)

lines( x = c( min(x), x0, x0),
       y = c(y0, y0, 0),
       lty = 2)
lines( x = c( min(x), x1, x1),
       y = c(y1, y1, 0),
       lty = 2)
axis(side = 2,
     at = c(0, 1, y0, 2),
     las = 1,
     labels = c(0, 1, expression(italic(y)), 2))
axis(side = 1,
     at = c(x0, 2, x1),
     labels = c(expression(2-sqrt(italic(y)-1)),
                2, 
                expression(2+sqrt(italic(y)-1)) ))
box()
```


:::{.example #TransformB name="Transformation"}
Example\ \@ref(exm:TransformationNon11) is repeated here using the distribution function method.
Given\ $Z$ is distributed $N(0, 1)$ we seek the probability distribution of $Y = \frac{1}{2} Z^2$.
First,
$$
   f_Z(z) 
   = (2\pi )^{-\frac 12}\,e^{-z^2/2}\quad\text{for $z\in (-\infty ,\,\infty )$}.
$$
Let\ $Y$ have PDF $f_Y(y)$ and df $F_Y(y)$. 
Then
\begin{align*}
     F_Y(y) 
      = \Pr(Y\leq y) 
     &= \Pr\left(\frac{1}{2}Z^2\leq y\right)\\
     &= \Pr(Z^2\leq 2y)\\
     & = \Pr(-\sqrt{2y}\leq Z\leq \sqrt{2y}\,)\\
     & = F_Z(\sqrt{2y}\,) - F_Z(-\sqrt{2y}\,)
\end{align*}
where\ $F_Z$ is the distribution function of\ $Z$.
Hence
\begin{align*}
     f_Y(y) 
       = F_Y'(y)
     &= F_Z'(\sqrt{2y}\,)-F_Z'(-\sqrt{2y}\,)\\
     &= \frac{\sqrt{2}}{2\sqrt{y}}f_Z(\sqrt{2y}\,) - \frac{\sqrt{2}}{-
2\sqrt{y}}f_Z(-\sqrt{2y}\,)\\[2mm]
     &= \frac{1}{\sqrt{2y}}[f_Z(\sqrt{2y}\,) + f_Z(-\sqrt{2y}\,)]\\
     &= \frac{1}{\sqrt{2y}} \left[ \frac{1}{\sqrt{2\pi}}\,\exp(-y) + \frac{1}{\sqrt{2\pi}}\,\exp(-y)\right]\\
     &= \frac{\exp(-y)y^{-\frac{1}{2}}}{\sqrt{\pi}}
\end{align*}
as before.
:::

Care is needed to ensure the steps are followed logically. 
Diagrams like Fig.\ \@ref(fig:SquaringXform) and\ \@ref(fig:SquaringXform2) are encouraged. 


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
The functions that are produced should be PDFs; check that this is the case.
:::


:::{.example #LOTUSProof name="Proof of Theorem\ \@ref(thm:ExpectationFunction)"}
Theorem\ \@ref(thm:ExpectationFunction) gave an important theorem involving the expectation of function of a random variable.
Here, we prove this theorem on the discrete and continuous cases.

For the *discrete* case , we need to show that $\sum_y y \Pr(g(X) = y) = \sum_x g(x) \Pr(X = x)$.

Let $Y = g(X)$. 
By definition, $\operatorname{E}[Y] = \sum_y y \Pr(Y = y)$.
Recall that the event $\{Y = y\}$ is actually the set of all\ $x$ such that $g(x) = y$, so substituting gives $\Pr(Y = y) = \sum_{x: g(x) = y} \Pr(X = x)$.
Substituting this into the expectation:
$$
  \operatorname{E}[g(X)] = \sum_y y \left( \sum_{x: g(x) = y} \Pr(X = x) \right).
$$
Since $y = g(x)$ for all\ $x$ in that inner sum, it can be moved inside the inner summation:
\begin{align*}
  \operatorname{E}[g(X)] 
  &= \sum_y \sum_{x: g(x) = y} y \Pr(X = x)\\
  &= \sum_y \sum_{x: g(x) = y} g(x) \Pr(X = x).
\end{align*}
Summing over all $y$, and then all $x$ that map to those $y$, is just the same as summing over all possible $x$.
In other word, $\operatorname{E}[g(X)] = \sum_x g(x) \Pr(X = x)$.

For the *continuous* case, we need to show that  $\int y f(g(X)=y) = \int_x g(x) f(x)$.

Assume $g(x)$ is a strictly increasing, differentiable function (this makes the proof simpler, but the result applies more generally). 
Let $Y = g(X)$.
Find the PDF of $Y$ using the transformation method:
$$
  f_Y(y) = f_X(x) \left| \frac{dx}{dy} \right| \quad \text{where } x = g^{-1}(y).
$$
Then, using the definition of $\operatorname{E}[Y]$:
$$
  \operatorname{E}[Y] = \int_{-\infty}^{\infty} y f_Y(y) \, dy.
$$
Substitute $y = g(x)$ and $f_Y(y) dy = f_X(x) dx$:
$$
  \operatorname{E}[g(X)] = \int_{-\infty}^{\infty} g(x) f_X(x) \, dx.
$$
:::


### Discrete univariate case

Let\ $X$ be a discrete random variable with PMF $p_X(x)$ and CDF $F_X(x) = \Pr(X \leq x)$.
For $Y = g(X)$, the CDF of $Y$ is
$$
  F_Y(y) = \Pr(Y \leq y) = \Pr(g(X) \leq y) = \sum_{\{x:\, g(x) \leq y\}} p_X(x).
$$
The PMF of\ $Y$ is then recovered as
$$
  p_Y(y) = F_Y(y) - F_Y(y^-) = \Pr(Y \leq y) - \Pr(Y \leq y - 1)
$$
where\ $y^-$ denotes the largest value in the support of\ $Y$ strictly less than\ $y$.

In the discrete case, the CDF method and the change of variables method are closely related---the difference is mainly one of presentation. 
The CDF framing is especially natural when\ $g()$ is monotone non-decreasing, since the event $\{g(X) \leq y\}$ maps cleanly to $\{X \leq g^{-1}(y)\}$.


:::{.example name="Shifting the geometric distribution"}
Let $X \sim \text{Geometric}(p)$, so that\ $X$ represents the number of *failures* until the first success, with support $\{0, 1, 2, \ldots\}$ (as in Sect.\ \@ref(GeometricDistribution)).
Then,
$$
  \Pr(X \leq x) = 1 - (1-p)^{\lfloor x\rfloor + 1}, \quad x = 0, 1, 2, \ldots
$$
Define $Y = X + 1$, so that\ $Y$ is the number of *trials* before the first success, with support $\{1, 2, 3, \ldots\}$. 
For $y \in \{1, 2, 3, \ldots\}$:
$$
  F_Y(y) = \Pr(Y \leq y) = \Pr(X + 1 \leq y) = \Pr(X \leq y - 1) = 1 - (1 - p)^{(y - 1) + 1} = 1 - (1 - p)^{y},
$$
using the DF for\ $X$ above.
The PMF is the difference between the DF at the values of\ $Y$ with positive probabilities (see Sect.\ \@ref(PFfromDF)):
$$
  p_Y(y) = F_Y(y) - F_Y(y - 1) = (1 - p)^{y - 1} p, \quad y = 1, 2, 3, \ldots
$$
This is the $\text{Geometric}(p)$ distribution with a different parameterisation.
A simple shift of support corresponds to a reparameterisation within the Geometric family of distributions.
:::


```{r echo=FALSE, fig.align="center", fig.cap="The PMF and DF of $X$ following a geometric distribution (left panels), and the PMF and DF of the transformation $Y = X + 1$.", fig.width=7, fig.height=7, out.width="80%"}
prob <- 0.3
x <- 0 : 12
px <- dgeom(x, 
            prob = prob)

y <- 0:12
py <- dgeom( (x - 1),
            prob = prob) 

par( mfrow = c(2, 2))
plotDiscrete(x, px, 
             type = "MF",
             ylab = "PMF",
             xlab = expression(italic(X)),
             main = expression(bold(Prob.~mass~"function"~of~italic(X))))
plotDiscrete(y, py, 
             type = "MF",
             ylab = "PMF",
             xlab = expression(italic(Y)),
             main = expression(bold(Prob.~mass~"function"~of~italic(Y))))

plotDiscrete(x, px, 
             type = "DF",
             ylab = "DF",
             xlab = expression(italic(X)),
             main = expression(bold(Distribution~"function"~of~italic(X))))
plotDiscrete(y, py, 
             type = "DF",
             ylab = "DF",
             xlab = expression(italic(Y)),
             main = expression(bold(Distribution~"function"~of~italic(Y))))

par( mfrow = c(1, 1))
```




### Continuous univariate case

Let\ $X$ be a continuous random variable with PDF\ $f_X$ and CDF $F_X$. 
Then, for $Y = g(X)$
$$
  F_Y(y) = \Pr(g(X) \leq y).
$$
Re-express $\{g(X) \leq y\}$ as an event in\ $X$, evaluate using $F_X$, then differentiate:
$$
  f_Y(y) = \frac{d}{dy} F_Y(y).
$$
When $g()$ is monotone increasing: $F_Y(y) = F_X(g^{-1}(y))$, and differentiation immediately recovers the change of variables formula. 
The CDF method is more flexible because it handles non-monotone transformations $g()$ without partitioning.


:::{.example name="Distribution of squared normal rv"}
Let $X \sim N(0, 1)$ and $Y = X^2$. 
Since $g(x) = x^2$ is not monotone on $\mathbb{R}$, the change of variable method requires partitioning; the distribution function method is simpler.

First, see that the support for $X$ is over\ $\mathbb{R}$, but the support for $Y = X^2$ must be over the non-negative reals.
Furthermore, since individual points have zero probability for continuous random variables, $X = 0$ is not technically possible, so the support is best given as $(0, \infty)$ or $[0, \infty)$ (see Fig.\ \@ref(fig:DFcontinuousUnivariate)).

So, for $y > 0$:
$$
  F_Y(y) = \Pr(X^2 \leq y) = \Pr(-\sqrt{y} \leq X \leq \sqrt{y})
  = 
  F_X(\sqrt{y}) - F_X(-\sqrt{y}) = 2F_X(\sqrt{y}) - 1.
$$

Differentiating:
$$
  f_Y(y) = 2 f_X(\sqrt{y}) \cdot \frac{1}{2\sqrt{y}}
  = \frac{1}{\sqrt{y}} \cdot \frac{1}{\sqrt{2\pi}} \exp(-y/2)
  = \frac{y^{1/2 - 1} \exp(-y/2)}{2^{1/2}\,\Gamma(1/2)}, \quad y > 0.
$$
Notice that $Y > 0$; if $Y = 0$ the PMF is undefined.
The final expressions identifies this as the $\chi^2(1)$ distribution\index{Chi-squared distribution} (i.e., $\text{Gamma}(1/2,\, 1/2)$).
:::


```{r DFcontinuousUnivariate, echo=FALSE, fig.align="center", out.width="100%", fig.width=8.25, fig.height=3.5, fig.cap="A normal distribution (left panel), the transformation (centre panel), and the resulting chi-squared distribution (right panel)."}
x <- seq(-4, 4,
         length = 100)
fX <- dnorm(x)

y <- seq(0, 16,
         length = 100)
fY <- dchisq(y, 
            df = 1)

par( mfrow = c(1, 3))

plot( fX ~ x,
      type = "l",
      las = 1,
      lwd = 3,
      main = expression(bold(PDF~of~italic(X))),
      xlab = expression(italic(X)),
      ylab = "PDF")

X2 <- x ^2
plot(X2 ~ x, 
     type = "l", 
     lwd = 3,
     las = 1,
     xlab = expression(italic(X)),
     ylab = expression(italic(Y)==italic(X)^2),
     main = "The transformation")
abline(h = 4,
       lwd = 2, 
       lty = 3)
text(x = 0,
     y = 4,
     pos = 3,
     label = expression(sqrt(italic(Y))))
segments(x0 = 2,
         x1 = 2,
         y0 = 0,
         y1 = 4,
         lty = 3,
         lwd = 2)
segments(x0 = -2,
         x1 = -2,
         y0 = 0,
         y1 = 4,
         lty = 3,
         lwd = 2)


plot( fY ~ y,
      type = "l",
      lwd = 3,
      las = 1,
      main = expression(bold(PDF~of~italic(Y)==italic(X)^2)),
      xlab = expression(italic(Y)),
      ylab = "PDF")
```



### Bivariate continuous case

Let $(X_1, X_2)$ have joint PDF $f_{\mathbf{X}}(x_1, x_2)$. 
To find the distribution of $Y = g(X_1, X_2)$ (a scalar function of\ $X_1$ and\ $X_2$), compute
$$
  F_Y(y) = \Pr(g(X_1, X_2) \leq y) 
  = 
  \iint_{\{(x_1,x_2):\, g(x_1,x_2) \leq y\}} f_{X_1, X_2}(x_1, x_2)\, dx_1\, dx_2,
$$
then differentiate: $f_Y(y) = F_Y'(y)$.

The key step is characterising the region of integration $\{g(x_1,x_2) \leq y\}$ carefully as a function of $y$, paying attention to how its boundaries move as $y$ changes. 
This is often where the difficulty lies, and a sketch of the support is strongly recommended.



:::{.example name="Distribution of $M = \max(X_1, X_2)$ for independent uniforms distributions"}
Let $X_1, X_2 \overset{\text{iid}}{\sim} \text{Uniform}(0,1)$ with joint PDF $f_{X_1, X_2}(x_1, x_2) = 1$ on the unit square. 
Define $M = \max(X_1, X_2)$.

The event $\{M \leq m\}$ is equivalent to $\{X_1 \leq m \text{ and } X_2 \leq m\}$,
so for $m \in [0,1]$:
$$
  F_M(m) = \Pr(X_1 \leq m,\, X_2 \leq m) = m \cdot m = m^2.
$$
Differentiating,
$$
  f_M(m) = 2m, \quad m \in [0, 1].
$$
This is the $\text{Beta}(2, 1)$ distribution. 
More generally, for $n$ iid $\text{Uniform}(0,1)$ random variables, $M = \max(X_1,\ldots,X_n)$ has CDF $m^n$ and PDF $nm^{n-1}$, a $\text{Beta}(n, 1)$ distribution. 
This extends naturally to order statistics (Chap.\ \@ref(OrderStatisticsChapter)).
:::


```{r bivariate-continuous, echo=FALSE, fig.width = 9, fig.height = 4.5}
set.seed(42)
N <- 80000
X1 <- runif(N); X2 <- runif(N)
M  <- pmax(X1, X2)

m_seq <- seq(0, 1, length.out = 300)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Show the integration region
plot(c(0,1), 
     c(0,1), 
     type = "n",
     xlab = expression(x[1]), 
     ylab = expression(x[2]),
     main = expression("Region {max(X"[1]*", X"[2]*") " <= " m}"),
     las = 1, asp = 1)
m_ex <- 0.65
rect(0, 0, m_ex, m_ex, col = adjustcolor("#2c7bb6", 0.25), border = NA)
rect(0, 0, 1,    1,    border = "grey40", lwd = 1.5)
abline(v = m_ex, lty = 2, col = "#2c7bb6", lwd = 1.5)
abline(h = m_ex, lty = 2, col = "#2c7bb6", lwd = 1.5)
text(m_ex / 2, m_ex / 2,
     expression(paste("Area = ", m^2)), cex = 1, col = "#2c7bb6")
text(m_ex + 0.03, 0.05, expression(m), col = "#2c7bb6", cex = 0.9)
grid(col = "grey85", lty = 1)

# Derived pdf vs simulation
hist(M, breaks = 60, freq = FALSE,
     col = "grey85", border = "white",
     main = expression("pdf of M = max(X"[1]*", X"[2]*") ~ Beta(2,1)"),
     xlab = "m", las = 1, xlim = c(0, 1))
lines(m_seq, 2 * m_seq, col = "#d7191c", lwd = 2.5)
lines(m_seq, dbeta(m_seq, 2, 1), col = "#1a9641", lwd = 1.5, lty = 2)
legend("topleft",
       legend = c("f(m) = 2m  (derived)", "Beta(2,1) pdf"),
       col = c("#d7191c", "#1a9641"), lwd = c(2.5, 1.5),
       lty = c(1, 2), bty = "n")
grid(col = "grey85", lty = 1)

par(mfrow = c(1, 1))
```



### Bivariate discrete case

Let $(X_1, X_2)$ have joint PMF $p_{\mathbf{X}}(x_1, x_2)$. 
For $Y = g(X_1, X_2)$, the CDF of\ $Y$ is
$$
  F_Y(y) = \Pr(Y \leq y) = \sum_{\{(x_1,x_2):\, g(x_1,x_2) \leq y\}} p_{\mathbf{X}}(x_1, x_2),
$$
and the PMF is recovered as $p_Y(y) = F_Y(y) - F_Y(y^-)$ where $y^-$ denotes the largest value in the support of\ $Y$ strictly less than\ $y$.

The bivariate discrete CDF method is often the most natural approach when\ $g()$ is a many-to-one function (such as a maximum or minimum) where enumerating pre-images directly would be cumbersome, but the event $\{g(X_1, X_2) \leq y\}$ has a simple geometric or combinatorial description.


:::{.example name="Maximum of Two Independent Binomials"}
Let $X_1 \sim \text{Binomial}(n_1, p)$ and $X_2 \sim \text{Binomial}(n_2, p)$ independently. 
Define $M = \max(X_1, X_2)$.

Using the same device as in the continuous case, for $m \in \{0, 1, \ldots, \max(n_1, n_2)\}$:
$$
  F_M(m) = \Pr(X_1 \leq m,\, X_2 \leq m) = F_{X_1}(m) \cdot F_{X_2}(m)
$$
by independence. 
The pmf is then
$$
  p_M(m) = F_M(m) - F_M(m - 1)
  = 
  F_{X_1}(m)F_{X_2}(m) - F_{X_1}(m-1)F_{X_2}(m-1).
$$
This does not simplify to a standard named distribution in general, illustrating that the CDF method can yield the distribution even when it has no closed-form family name.
:::


```{r bivariate-discrete, echo=FALSE, fig.width = 9, fig.height = 5, out.width="100%"}
n1 <- 10; n2 <- 8; p <- 0.4
m_vals <- 0:max(n1, n2)

# CDF of M via independence
F_X1 <- pbinom(m_vals, n1, p)
F_X2 <- pbinom(m_vals, n2, p)
F_M  <- F_X1 * F_X2

# PMF of M
p_M  <- diff(c(0, F_M))

# Simulation to verify
set.seed(7)
N   <- 100000
sim <- pmax(rbinom(N, n1, p), rbinom(N, n2, p))
sim_pmf <- table(sim) / N

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# CDF plot
plot(m_vals, F_M, type = "s", col = "#2c7bb6", lwd = 2,
     xlab = "m", ylab = "F(m)",
     main = expression("CDF of M = max(X"[1]*", X"[2]*")"),
     ylim = c(0, 1), las = 1)
points(m_vals, F_M, pch = 16, col = "#2c7bb6")
lines(m_vals, F_X1, lty = 2, col = "#d7191c",  lwd = 1.5)
lines(m_vals, F_X2, lty = 3, col = "#1a9641",  lwd = 1.5)
legend("topleft",
       legend = c(expression("F"[M]*"(m)"),
                  expression("F"[X[1]]*"(m)  Bin(10,0.4)"),
                  expression("F"[X[2]]*"(m)  Bin(8,0.4)")),
       col = c("#2c7bb6","#d7191c","#1a9641"),
       lty = c(1,2,3), lwd = 2, bty = "n", cex = 0.85)
grid(col = "grey85", lty = 1)

# PMF: derived vs simulated
plot(m_vals, p_M, type = "h", col = "#2c7bb6", lwd = 4,
     xlab = "m", ylab = "Probability",
     main = expression("PMF of M = max(X"[1]*", X"[2]*")"),
     ylim = c(0, max(p_M) * 1.15), las = 1)
points(as.integer(names(sim_pmf)), as.numeric(sim_pmf),
       pch = 4, col = "#d7191c", cex = 1.5, lwd = 2)
legend("topleft",
       legend = c("Derived pmf", "Simulated"),
       col = c("#2c7bb6", "#d7191c"),
       lty = c(1, NA), pch = c(NA, 4),
       lwd = c(4, 2), bty = "n")
grid(col = "grey85", lty = 1)

par(mfrow = c(1, 1))
```


### Summary

The distribution function method is preferred over the change of variables method when 

1. the transformation is many-to-one (e.g.\ maximum, minimum, absolute value); 
2. the result is not a standard named distribution; or 
3.  only a marginal distribution is needed from a multivariate problem.

The approach is summarised in Table\ \@ref(tab:DFSummary).



```{r DFSummary, echo=FALSE}
DFsummary <- array( dim = c(4, 3))

DFsummary[1, ] <- c("Discrete univariate",
                     "$F_Y(y) = \\sum_{\\{x:\\,g(x)\\leq y\\}} p_X(x)$; pmf = $\\Delta F_Y$ ",
                     "Difference the CDF")
DFsummary[2, ] <- c("Continuous univariate",
                     "$F_Y(y) = \\Pr(g(X)\\leq y)$; invert inequality; use $F_X$",
                     "Differentiate the CDF")
DFsummary[3, ] <- c("Discrete bivariate",
                     "$F_Y(y) = \\iint_{\\{g\\leq y\\}} f_{X_1, X_2}$; characterise region",
                     "Differentiate the CDF")
DFsummary[4, ] <- c("Continuous bivariate",
                     "$F_Y(y) = \\sum\\sum_{\\{g\\leq y\\}} p_{X_1, X_2}$; pmf = $\\Delta F_Y$",
                     "Difference the CDF" )


colnames(DFsummary) <- c("Setting",
                         "Strategy",
                         "PDF recovery")

if( knitr::is_latex_output() ) {
  knitr::kable(DFsummary,
               format = "latex",
               booktabs = TRUE,
               align = c("r", "c", "c"),
               longtable = FALSE,                
               escape = FALSE,
               caption = "Summarising the distribution function method.") %>%
    kable_styling(font_size = 10) %>%
    column_spec(2, width = "60mm") %>%
    column_spec(3, width = "45mm") %>%
    row_spec(0, bold = TRUE) %>% 
    column_spec(1, bold = TRUE) 
}
if( knitr::is_html_output() ) {
  knitr::kable(DFsummary,
               escape = TRUE,
               format = "html",
               booktabs = TRUE,
               align = "r",
               longtable = FALSE,                
               caption = "Summarising the distribution function method.") %>%
    column_spec(2, width = "20%") %>%
    column_spec(2, width = "50%") %>%
    column_spec(3, width = "40%") %>%
    row_spec(0, bold = TRUE) %>% 
    column_spec(1, bold = TRUE)
}

```





| Setting | Strategy | PDF recovery |
|---------|----------|--------------|
| **Discrete univariate** | $F_Y(y) = \sum_{\{x:\,g(x)\leq y\}} p_X(x)$; pmf = $\Delta F_Y$ | Difference the CDF |
| **Continuous univariate** | $F_Y(y) = \Pr(g(X)\leq y)$; invert inequality; use $F_X$ | Differentiate the CDF|
| **Bivariate continuous** | $F_Y(y) = \iint_{\{g\leq y\}} f_\mathbf{X}$; characterise region | Differentiate the CDF |
| **Bivariate discrete** | $F_Y(y) = \sum\sum_{\{g\leq y\}} p_\mathbf{X}$; pmf = $\Delta F_Y$ | Difference the CDF |

\index{Transformations of random variables!distribution function method|(}




## The moment generating function method {#TransformationMoments}

\index{Transformations of random variables!moment generating function method|(}
The moment generating function (MGF) method is useful for finding the distribution of a linear combination of $n$ independent random variables.
The method essentially involves the computation of the MGF of the transformed variable $Y = u(X_1, X_2, \dots, X_n)$ when the joint distribution of independent $X_1, X_2, \dots, X_n$ is given.\index{Moment generating function}

The MGF method relies on this observation: since the MGF of a random variable (if it exists) completely specifies the distribution of the random variable, then if two random variables have the same MGF they must have identical distributions.
Below, the transformation $Y = X_1 + X_2 + \cdots X_n$ is demonstrated, but the same principles can be applied for other linear combinations also.

Consider\ $n$ independent random variables $X_1, X_2, \dots, X_n$ with MGFs $M_{X_1}(t)$, $M_{X_2}(t)$, $\dots$, $M_{X_n}(t)$, and consider the transformation $Y = X_1 + X_2 + \cdots X_n$.
Since the\ $X_i$ are independent, $f_{X_1,X_2\dots X_n}(x_1, x_2, \dots, x_n) = f_{X_1}(x_1).f_{X_2}(x_2)\dots f_{X_n}(x_n)$.
So, by definition of the MGF,
\begin{align*}
   M_Y(t)
   &= \operatorname{E}(\exp(tY)) \\
   &= \operatorname{E}(\exp[t(X_1 + X_2 + \cdots X_n)]) \\
   &= \int\!\!\!\int\!\!\!\cdots\!\!\!\int \exp[t(x_1 + x_2 + \cdots x_n)] f(x_1, x_2, \dots x_n)\,dx_n\dots dx_2\, dx_1 \\
   &= \int\!\!\!\int\!\!\!\cdots\!\!\!\int \exp(tx_1) f(x_1) \exp(t{x_2}) f(x_2)\dots \exp(t{x_n})f(x_n) \,dx_n\dots dx_2\, dx_1 \\
   &= \int \exp(t x_1) f(x_1)\,dx_1 \int \exp(t{x_2}) f(x_2)\,dx_2 \dots \int \exp(t{x_n})f(x_n)\,dx_n \\
   &= M_{X_1}(t) M_{X_2}(t)\dots M_{X_n}(t) \\
   &= \prod_{i = 1}^n M_{X_i}(t).
\end{align*}
($\prod$ is the symbol for a *product* of terms, in the same way that $\sum$ is the symbol for a *summation* of terms.)
The above result also holds for discrete variables, where summations replace integrations.

This result follows: if $X_1, X_2, \dots, X_n$ are independent random variables and $Y  =  X_1 + X_2 + \dots + X_n$, then the MGF of\ $Y$ is
$$
   M_Y(t)  =  \prod_{i = 1}^n M_{X_i}(t)
$$
where $M_{X_i}(t)$ is the MGF of\ $X_i$ at $t$ for $i = 1, 2, \dots, n$.


:::{.example #MGFLinearX name="MGF method for transformations"}
Suppose that $X_i \sim \text{Poisson}(\lambda_i)$ for $i  =  1, 2, \dots, n$, and we wish to find the distribution of $Y  =  X_1  +  X_2  + \dots  +  X_n$.

Since\ $X_i$ has a Poisson distribution with parameter $\lambda_i$ for $i, 2, \dots n$, the MGF of\ $X_i$ is
$$
   M_{X_i}(t) = \exp[ \lambda_i(e^t - 1)].
$$
The MGF of $Y  = X_1 + X_2 + \cdots X_n$ is
\begin{align*}
   M_Y(t)
   &= \prod_{i = 1}^n \exp[ \lambda_i(e^t - 1)] \\
   &= \exp[ \lambda_1(e^t - 1)] \exp[ \lambda_2(e^t - 1)] \dots \exp[ \lambda_n(e^t - 1)] \\
   &= \exp\left[ (e^t - 1)\sum_{i = 1}^n \lambda_i\right].
\end{align*}
Using $\Lambda = \sum_{i = 1}^n \lambda_i$, the MGF of\ $Y$ is
$$
   M_Y(t) = \exp\left[ (e^t - 1)\Lambda \right],
$$
which is the MGF of a Poisson distribution with mean $\Lambda = \sum_{i = 1}^n \lambda_i$.
This means that the sum of\ $n$ independent Poisson distribution is also a Poisson distribution, whose mean is the sum of the individual Poisson means.
:::
\index{Transformations of random variables!moment generating function method|)}



## Choosing a method

The *change of variables method*\index{Transformations of random variables!change of variable method} is used when the transformation is smooth and monotone (or piecewise monotone), and the full (joint) distribution is required. 
This method is particularly useful in the bivariate and multivariate continuous case: in the Gamma/Beta example (CITE), for example, the joint PDF is obtained directly, and so independence is i=easily identified from factorisation. 
In the discrete case, the change of variable method is effectively careful bookkeeping rather than a distinct technique. 
The change of variable method is best avoided when the Jacobian computation is intractable or difficult, or when the full PDF is not needed and only moments are of interest.

The *distribution function method*\index{Transformations of random variables!distribution function method}
 is the most general method: it works for any transformation, discrete or continuous.
It is also intuitive: compute $\Pr(Y le y)$ and differentiate or find differences.
The distribution function method, however, can become algebraically tedious, and can become complicated with multivariate or multi-step transformations.

The *MGF method*\index{Transformations of random variables!moment generating function method} is only useful in very specific situations, but in these situations it excels. 
It is most applicable for sums of random variables; for example,  proving that a sum (or linear combination) of independent random variables follows a particular named distribution, since the MGFs are multiplied, and the resulting MGF matched to MGFs of known distributions. 
The MGF method cnbe used, for example, to show that the sum of independent normal distributions is a normal distributions, that the sum of independent Poisson distributions sum to Poisson distributions, and so on. 
However, the MGF may not exist, and the method reveals nothing about the joint distribution of a transformed vector; it is useless when the transformation is not a sum or linear combination.
\index{Transformations of random variables|)}



## Exercises

Selected answers appear in Sect.\ \@ref(AnswersChapterTransformations).


:::{.exercise #X2}
Suppose the PDF of\ $X$ is given by
$$
   f_X(x) = \begin{cases}
                x/2 & \text{$0 < x < 2$};\\
                0 & \text{otherwise}.
             \end{cases}
$$

1. Find the PDF of $Y = X^3$ using the change of variable method.
1. Find the PDF of $Y = X^3$ using the distribution function method. 
:::


:::{.exercise #JointDiscrete}
The discrete bivariate random vector $(X_1, X_2)$ has the joint probability function
$$
   f_{X_1, X_2}(x_1, x_2) = 
   \begin{cases}
      (2x_1+ x _2)/6 & \text{for $x_1 = 0, 1$ and $x_2 = 0, 1$};\\
      0               & \text{elsewhere}.
   \end{cases}
$$
Consider the transformations
\begin{align*}
   Y_1 &= X_1 + X_2 \\
   Y_2 &= \phantom{X_1+{}} X_2
\end{align*}

1. Determine the joint probability function of $(Y_1, Y_2)$.
1. Deduce the distribution of\ $Y_1$.
:::



:::{.exercise #GammaSum}
Consider\ $n$ random variables $X_i$ such that $X_i \sim \text{Gamma}(\alpha_i, \beta)$.
Determine the distribution of $Y = \sum_{i = 1}^n X_i$ using MGFs.
:::


:::{.exercise #CauchySquared}
The random variable\ $X$ has PDF
$$
   f_X(x) = \frac{1}{\pi(1 + x^2)}
$$
for $-\infty < x < \infty$.
Find the PDF of\ $Y$ where $Y = X^2$.
:::

:::{.exercise #ContTransform}
A random variable\ $X$ has distribution function
$$
   F_X(x) = 
   \begin{cases}
      0                & \text{for $x \le -0.5$};\\
      \frac{2x + 1}{2} & \text{for $-0.5 < x < 0.5$};\\
      1                & \text{for $x \ge 0.5$}.
   \end{cases}
$$

1. Find, and plot, the PDF of\ $X$.
1. Find the distribution function, $F_Y(y)$, of the random variable $Y = 4 - X^2$.
1. Hence find, and plot, the PDF of $Y$, $f_Y(y)$.
:::


:::{.exercise #Projectile}
Suppose a projectile is fired at an angle\ $\theta$ from the horizontal with velocity\ $v$. 
The horizontal distance that the projectile travels\ $D$ is
$$
   D = \frac{v^2}{g} \sin 2\theta,
$$
where\ $g$ is the acceleration due to gravity ($g\approx 9.8$ m.s^$-2$^).

1. If\ $\theta$ is uniformly distributed over the range $(0, \pi/4)$, find the probability density function of\ $D$.
1. Sketch the PDF of\ $D$ over a suitable range for $v = 12$ and using $g\approx 9.8$m.s^$-2$^.
:::


:::{.exercise #RandomExp}
Most computers have facilities to generate continuous uniform (pseudo-)random numbers between zero and one, say\ $X$.
When needed, exponential random numbers are obtained from\ $X$ using the transformation $Y = -\alpha\ln X$.

1. Show that\ $Y$ has an exponential distribution and determine its parameters.
1. Deduce the mean and variance of\ $Y$.
:::


:::{.exercise #RandomW}
Consider a random variable\ $W$ for which $\Pr(W = 2) = 1/6$, $\Pr(W = -2) = 1/3$ and $\Pr(W = 0) = 1/2$.

1. Plot the probability function of\ $W$.
1. Find the mean and variance of\ $W$.
1. Determine the distribution of $V = W^2$.
1. Find the distribution function of\ $W$.
:::


:::{.exercise #NormalMixture}
In a study to model the load on bridges [@lu2019evaluating], the researchers modelled the Gross Vehicle Weight (GVM, in kilonewtons) weight of smaller trucks\ $S$ using $S\sim N(390, 740$, and the weight of bigger trucks\ $B$ using $L\sim N(865, 142)$.
The total load distribution\ $L$ was then modelled as $L = 0.24S + 0.76B$, reflecting the expected proportion if smaller and bigger trucks using the bridge.

1. Plot the distribution of\ $L$.
1. Compute the mean and standard deviation of\ $L$.
:::


:::{.exercise #LogNormal}
Suppose the random variable\ $X$ has a normal distribution with mean $\mu$ and variance\ $\sigma^2$.
The random variable $Y = \exp X$ is said to have a *log-normal distribution*.

1. Determine the distribution function of\ $Y$ in terms of the function $\Phi(\cdot)$ (see Def.\ \@ref(def:StandardNormalDF)). 
1. Differentiate to find the PDF of $Y$.
1. Plot the log-normal distribution for various parameter values.
1. Determine $\Pr(Y > 2 | Y < 1)$ when $\mu = 2$ and $\sigma^2 = 2$.

(**Hint**: Use the `dlnorm()` and `plnorm()` functions in **R**, where $\mu = {}$`meanlog` and $\sigma = {}$`sdlog`.)
:::


:::{.exercise #BinomXform}
If\ $X$ is a random variable with probability function
$$
   \Pr(X = x) = \binom{4}{x} (0.2)^x (0.8)^{4 - x} \quad \text{for $x = 0, 1, 2, 3, 4$}, 
$$
find the probability function of the random variable defined by $Y = \sqrt{X}$.
:::


:::{.exercise #Xminus3}
Given the random variable\ $X$ with probability function
$$
   \Pr(X = x) = \frac{x^2}{30} \quad \text{for $x = 1, 2, 3, 4$},
$$
find the probability function of $Y= (X - 3)^2$.
:::


:::{.exercise #GiveDF2}
A random variable\ $X$ has distribution function
$$
   F_X(x) = 
   \begin{cases}
      0 & \text{for $x < -0.5$};\\
      \frac{2x + 1}{2}, & \text{for $-0.5 < x < 0.5$}; \\
      1 & \text{for $x > 0.5$}.
   \end{cases}
$$

1. Find the distribution function, $F_Y(y)$, of the random variable $Y = 4 - X^2$.
1. Hence find the PDF of\ $Y$.
:::


:::{.exercise}
If the random variable\ $X$ has an exponential distributed with mean\ $1$, show that the distribution of $-\log(X)$ has a Gumbel distribution (Eq.\ \@ref(eq:Gumbel))\index{Gumbel distribution} with $\mu = 0$ and $\sigma = 1$.
:::



:::{.exercise}
Let\ $X$ have a gamma distribution with parameters $\alpha > 2$ and $\beta > 0$.

1. Prove that the mean of\ $1/X$ is $\beta/(\alpha - 1)$.
2. Prove that the variance of $1/X$ is $\beta^2/[(\alpha - 1)^2(\alpha - 2)]$.
:::



:::{.exercise #BetaHospital2}
In a study modelling waiting times at a hospital [@khadem2008evaluating], patients are classified into one of three categories: Red (critically ill or injured patients), Yellow (moderately ill or injured patients), or Green (injured or uninjured patients).

For 'Green' patients, the service time\ $S$ was modelled as $S = 4.5 + 11V$, where $V \sim \text{Beta}(0.287, 0.926)$.

1. Produce well-labelled plots of the PDF and DF of\ $S$, showing important features. 
1. What proportion of patients have a service time exceeding\ $15\mins$?
1. The quickest\ $20$% of patients are serviced within what time?
:::



:::{.exercise #ExpHospital2}
In a study modelling waiting times at a hospital [@khadem2008evaluating], patients are classified into one of three categories:

* Red: Critically ill or injured patients.
* Yellow: Moderately ill or
injured patients.
* Green: Minimally injured or
uninjured patients.

The time (in minutes) spent in the reception are for 'Yellow' patients, say\ $T$, is modelled as $T = 0.5 + W$, where $W\sim \text{Exp}(16.5)$.

1. Plot the PDF and df of\ $T$.
1. What proportion of patients wait more than $20\mins$, if they have already been waiting for $10\mins$?
1. How long to the slowest\ $10$% of patients need to wait?
:::



:::{.exercise #Not1to1}
Suppose the random variable\ $Z$ has the PDF
$$
  f_Z(z) = 
  \begin{cases}
    \frac{1}{3} & \text{for $-1 < z < 2$};\\
    0 & \text{elsewhere}.
  \end{cases}
$$

1. Find the *probability density function* of\ $Y$, where $Y = Z^2$, using the **distribution function method**.
1. Confirm that your final PDF of\ $Y$ is a valid PDF.
1. Produce a well-labelled plot of the PDF of\ $Y$.
   Ensure all important features and points are clearly labelled.
:::


:::{.exercise #ChiSquaredGamma}
Show that the chi-squared distribution is a special case of the gamma distribution, with $\alpha = \nu/2$ and $\beta = 2$.
:::


:::{.exercise #TrickyXform1}
Suppose the random variable\ $X$ is defined as shown in Fig.\ \@ref(fig:TrickyX).

1. Determine the distribution function for\ $X$.
2. Find the probability density function for the random variable\ $Y$, where $Y = 6 - 2X$.
3. Confirm that your probability density function for\ $Y$ is a valid pdf.
4. Plot the probability density function of $Y$.
:::

```{r TrickyX, echo=FALSE, out.width="90%", fig.width=8, fig.height=4, fig.align="center", fig.cap="The probability density function for the random variable\\ $X$."}
x <- seq(-0.5, 3.5,
         by = 0.5)
fy <- function(x){ ifelse( (x > 0) & (x < 1),
                           2 * x / 3,
                           ifelse( (x >= 1) & (x < 3),
                                   (3 - x)/3,
                                   0) ) }

plot( fy(x) ~ x,
      type = "l",
      lwd = 3,
      col = plotColour1,
      las = 1,
      axes = FALSE,
      main = expression(bold(The~PDF~"for"~the~rv~italic(X))),
      xlab = expression(italic(X)),
      ylab = expression(italic(f)[italic(X)(italic(x))]))
axis(side = 1)
axis(side = 2,
     at = c(0, 1, 2)/3,
     label = c("0", "1/3", "2/3"),
     las = 1)
box()
abline(v = 1,
       lwd = 1, 
       lty = 2,
       col = "grey")
```





:::{.exercise #TrickyXform2}
Suppose the random variable\ $X$ is defined as shown in Fig.\ \@ref(fig:TrickyX).

1. Determine the distribution function for\ $X$ (this was done in Exercise\ \@ref(exr:TrickyXform1)).
2. Find the probability density function for the random variable\ $Z$, where $Z = (X - 2)^2$.
3. Confirm that your probability density function for\ $Z$ is a valid pdf.
4. Plot the probability density function of $Z$.
:::




:::{.exercise #RunnerVelocity}
The time taken to run a distance $D$ (in metres) by a professional athlete, say\ $T$ (in seconds), varies with the distribution shown in Fig.\ \@ref(fig:RunnerVelocityVoltage) (left panel).
The average velocity of the runner, say\ $V$, is related to the time by $V = D/T$.

1. Determine the probability density function for the runner's velocity.
2. Suppose $D = 100$, $\mu = 12$ and $\Delta = 0.25$.
   Plot the probability density function for\ $V$.
:::


```{r RunnerVelocityVoltage, echo=FALSE, out.width="90%", fig.width=8, fig.height=4, fig.align="center", fig.cap="The probability density function for the random variable\\ $T$, the time for the run."}
f <- function(t, mu, a){
  ifelse( ( t >= (mu - a) & (t <= mu)), 
          (a - mu + t)/(a^2),
          (a + mu - t)/(a^2))}
          
          
g <- function(v, mu, a, D){
  ifelse( ( v <= D/(mu - a) & (v >= D/mu)), 
          (D * (v*a - v*mu + D ) )/(a^2 * v^3 ),
          (D * (v*a + v*mu - D ) )/(a^2 * v^3 ) ) }
          


###
par( mfrow = c(1, 2) )

a <- 0.25
mu <- 12
D <- 100

t <- seq(mu - a, mu + a,
         length = 1000)
plot( f(t, mu = mu, a = a) ~ t,
      xlab = expression(Time~italic(t)),
      ylab = "PDF",
      main = "PDF of running time\n(in seconds)",
      type = "l",
      lwd = 3,
      col = plotColour1,
      axes = FALSE,
      las = 1)
axis(side = 1,
     at = c(mu - a,
            mu + a),
     labels = c(expression(mu - Delta),
                expression(mu + Delta) ) )
axis(side = 2,
     at = c(0, 1/a),
     label = c("0",
               expression(1/Delta)),
     las = 1)
box()


###

mu <- 0
sigma <- 1

voltage <- seq(mu -3.5 * sigma,
               mu + 3.5 * sigma,
               length = 1000)

plot( dnorm(voltage, mean = mu, sd = sigma) ~ voltage,
      xlab = expression(Voltage~italic(v)),
      ylab = "PDF",
      xlim = c(-3.25, 3.25),
      main = "PDF of instantaneous voltage\n(in volts)",
      type = "l",
      lwd = 3,
      col = plotColour1,
      las = 1)

box()

```



:::{.exercise #VoltagePower}
Suppose the instantaneous voltage\ $V$ (in volts) in a circuit varies over time such that
$$
  f_V(v) = \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left\{ -\frac{x^2}{2\sigma^2}\right\}.
$$
as shown in Fig.\ \@ref(fig:RunnerVelocityVoltage) (right panel).
(Later, we will identify this as the *normal distribution*.)\index{Normal distribution}

1. Determine the probability density function of the instantaneous *power* in the circuit\ $P$, where $P = V^2/R$ for some circuit resistance\ $R$ (in ohms).
2. Suppose $\sigma = 1$, and $R = 10$.
   Plot the probability density function for\ $P$.
:::



:::{.exercise #LogNormalProduct}
Let $X_1, X_2, \ldots, X_n$ be independent positive random variables, and let
$$
  P = X_1 \cdot X_2 \cdots X_n = \prod_{i=1}^n X_i
$$
be their product.

1. Show that $\log P = \sum_{i=1}^n \log X_i$, where $P = \prod_{i=1}^n X_i$.
2. Hence show that $P \sim \text{LogNormal}\!\left(\sum_{i=1}^n \mu_i,\, \sum_{i=1}^n \sigma_i^2\right)$.
3. Suppose $X_1, X_2, \ldots, X_{10}$ are independent with $)\log X_i \sim N(0.2, 0.09)$ for each\ $i$.
   Find the distribution of $P = \prod_{i=1}^{10} X_i$ and hence find $\Pr(P > \exp 3)$.
:::






