# Mathematical expectation for bivariate distributions {#ExpectationsBivariate}



```{r setup, include=FALSE}
source("R/SetColours.R")
```


::: {.objectivesBox .objectives data-latex="{iconmonstr-target-4-240.png}"}
Upon completion of this chapter you should be able to:

* compute the expectation and variance of linear combinations of random variables.
* interpret and compute the covariance and the coefficient of correlation between two random variables.
* compute the conditional mean and conditional variance of a random variable for some given value of another random variable.
:::


## Expected values of a bivariate function {#ExpectationsBivariateFunctions}

In a manner analogous to the univariate case, the expectation of functions of two random variables can be given.


:::{.definition #BVExpectation name="Expectation for bivariate distributions"}
Let $(X, Y)$ be a $2$-dimensional random variable and let $u(X, Y)$ be a function of\ $X$ and\ $Y$. 

For a *discrete* bivariate distribution with probability mass function $p_{X, Y}(x, y)$ defined over $(x, y) \in R$, the *expectation* or *expected value* of $\operatorname{E}[u(X, Y)]$ is
$$
   \operatorname{E}[u(X, Y)]
   = \sum_{(x, y)\in R} u(x, y)\, p_{X, Y}(x, y).
$$
For a *continuous* bivariate distribution with probability density function $f_{X, Y}(x, y)$ defined over $(x, y) \in R$, the *expectation* or *expected value* of $\operatorname{E}[u(X, Y)]$ is
$$
   \operatorname{E}[u(X, Y)]
   = \iint_{(x, y)\in R} u(x, y)\, f_{X, Y}(x, y).
$$
:::


This definition can be extended to the expectation of a function of any number of random variables.


:::{.example #ExpectationFunctionTwoDiscrete name="Expectation of function of two rvs (discrete)"}
Consider the joint distribution of\ $X$ and\ $Y$ in Example\ \@ref(exm:BVDiscrete3). 
Determine $\operatorname{E}[X + Y]$; i.e., the mean of the number of heads plus the number showing on the die.

From Def.\ \@ref(def:BVExpectation), write $u(X, Y) = X + Y$ and so
\begin{align*}
   \operatorname{E}[X + Y] 
   &= \sum_{x = 0}^2 \sum_{y = 1}^6 (x + y)\, p_{X, Y}(x, y)\\
   &= 1\times(1/24) + 2\times(1/24) + \dots + 6\times(1/24)\\
   & \qquad + 2\times(1/12) + 3\times(1/12) + \dots + 7\times(1/12)\\
   & \qquad + 3\times(1/24) + 4\times(1/24) + \dots + 8\times(1/24)\\
   &= 21/24 + 27/12 + 33/24\\
   &= 4.5.
\end{align*}
The answer is just $\operatorname{E}[X] + \operatorname{E}[Y] = 1 + 3.5 = 4.5$.
This is no coincidence, as we see from Theorem\ \@ref(thm:ExpTwoRV).
:::


:::{.example #ExpectationFunctionTwoContinuous name="Expectation of function of two rvs (continuous)"}
Consider Example\ \@ref(exm:BVDiscreteBank).
To determine $\operatorname{E}[XY]$, write $u(X, Y) = XY$ and proceed:
$$
  \operatorname{E}[XY]
   = \frac{6}{5} \int_0^1\int_0^1 xy(x + y^2)\,dx\,dy
   = \frac7{20}.
$$
Unlike the previous example, an alternative simple calculation based on $\operatorname{E}[X]$ and $\operatorname{E}[Y]$ is not possible, since $\operatorname{E}[XY]\neq\operatorname{E}[X] \operatorname{E}[Y]$ in general.
:::


:::{.theorem #ExpTwoRV name="Expectations of two rvs"}
If\ $X$ and\ $Y$ are any random variables, and\ $a$ and\ $b$ are any constants, then
$$
   \operatorname{E}[aX + bY] = a\operatorname{E}[X] + b\operatorname{E}[Y].
$$

::::{.proof}
The proof given here is for the discrete case.
\begin{align*}
  \operatorname{E}[aX + bY]
  &= \sum_{(x, y) \in R}(ax + by) \, p_{X, Y}(x, y) \quad\text{by definition}\\
  &= \sum_x \sum_y ax\, p_{X, Y}(x, y) + \sum_x \sum_y by\, p_{X, Y}(x, y)\\
  &= a\sum_x x\sum_y p_{X, Y}(x, y) + b\sum_y y\sum_x p_{X, Y}(x, y)\\
  &= a\sum_x x \Pr(X = x) + b\sum_y y \Pr(Y = y)\\
  &= a\operatorname{E}[X] + b\operatorname{E}[Y].
\end{align*}
The continuous case is analogous.
::::
:::


This theorem is no surprise after seeing Theorem\ \@ref(thm:ExpectationLinear), but is powerful and useful. 

This result is true whether or not\ $X$ and\ $Y$ are independent.
Theorem\ \@ref(thm:ExpTwoRV) naturally generalises to the expected value of a *linear combination of random variables* (Theorem\ \@ref(thm:ExpLinear)).


## Moments of a bivariate distribution: covariance

The idea of a moment in the univariate case naturally extends to the bivariate case. 
Hence, define $\mu'_{rs} = \operatorname{E}[X^r Y^s]$ or $\mu_{rs} = \operatorname{E}\big[(X - \mu_X)^r (Y - \mu_Y)^s\big]$ as the raw and central moments for a bivariate distribution.

The most important of these moments is the covariance.

:::{.definition #BVCovariance name="Covariance"}
The *covariance* of\ $X$ and\ $Y$ is defined as 
$$
   \operatorname{Cov}(X, Y) = \operatorname{E}[(X - \mu_X)(Y - \mu_Y)].
$$
When\ $X$ and\ $Y$ are discrete,
$$
   \operatorname{Cov}(X, Y) =
   \sum_{x} \sum_{y} (x - \mu_X)(y - \mu_Y)\, p_{X, Y}(x, y).
$$
When\ $X$ and\ $Y$ are continuous,
$$
   \operatorname{Cov}(X, Y) =
   \int_{-\infty}^\infty\!\int_{-\infty}^\infty (x - \mu_X)(y - \mu_Y)\, f_{X, Y}(x, y)\, dx\, dy.
$$
:::

The covariance is a measure of how\ $X$ and\ $Y$ vary jointly, in the sense that a positive covariance indicates that 'on average'\ $X$ and\ $Y$ increase (or decrease) together whereas a negative covariance indicates that `on average' as\ $X$ increases and\ $Y$ decreases (and vice versa). 
We say that covariance is a measure of *linear dependence*.

Covariance is best evaluated from the computational formula.


:::{.theorem #Covariance name="Covariance"}
For any random variables\ $X$ and\ $Y$,
$$
   \operatorname{Cov}(X, Y)
   =
   \operatorname{E}[XY] - \operatorname{E}[X]\operatorname{E}[Y].
$$

::::{.proof}
The proof uses Theorems\ \@ref(thm:ExpectationLinear) and\ \@ref(thm:ExpTwoRV).
\begin{align*}
   \operatorname{Cov}(X, Y)
   &= \operatorname{E}\big[ (X - \mu_X)(Y-\mu_Y)\big] \\
   &= \operatorname{E}[ XY - \mu_X Y - \mu_Y X + \mu_X\mu_Y] \\
   &= \operatorname{E}[ XY ] - \mu_X\operatorname{E}[Y] - \mu_Y\operatorname{E}[X] +  \mu_X \mu_Y \\
   &= \operatorname{E}[ XY ] - \mu_X\mu_Y - \mu_Y\mu_X +  \mu_X \mu_Y \\
   &= \operatorname{E}[ XY ] - \mu_X \mu_Y.
\end{align*}
::::
:::


Computing the covariance is tedious: $\operatorname{E}[X]$, $\operatorname{E}[Y]$, $\operatorname{E}[XY]$ need to be computed, and so the joint and marginal distributions of\ $X$ and\ $Y$ are needed.

Covariance has units given by the product of the units of\ $X$ and\ $Y$. 
For example, if\ $X$ is measured in metres and\ $Y$ is measured in seconds then $\operatorname{Cov}(XY)$ has the units metre--seconds.
To compare the strength of covariation amongst pairs of random variables, a unitless measure is useful.
Correlation does this by scaling the covariance in terms of the standard deviations of the individual variables.


:::{.definition #Correlation name="Correlation"}
The *correlation coefficient* between the random variables\ $X$ and\ $Y$ is denoted by $\text{Corr}(X, Y)$ or $\rho_{X, Y}$ and is defined as
$$
   \rho_{X, Y} 
   = \frac{\operatorname{Cov}(X, Y)}{\sqrt{ \operatorname{var}[X]\operatorname{var}[Y]}}
   = \frac{\sigma_{X, Y}}{\sigma_X \sigma_Y}.
$$
:::


If there is no confusion over which random variables are involved, we write\ $\rho$ rather than\ $\rho_{XY}$.
It can be shown that $-1 \leq \rho \leq 1$.


:::{.example #CorrCoefDiscrete name="Correlation coefficient (discrete rvs)"}
Consider two discrete random variables\ $X$ and\ $Y$ with the joint pf given in Table\ \@ref(tab:Joint3).
To compute the correlation coefficient, the following steps are required.

* $\text{Corr}(X, Y) = \operatorname{Cov}(X, Y)/\sqrt{ \operatorname{var}[X]\operatorname{var}[Y]}$, so $\operatorname{var}[X]$, $\operatorname{var}[Y]$ must be computed;
* To find $\operatorname{var}[X]$ and $\operatorname{var}[Y]$, $\operatorname{E}[X]$ and $\operatorname{E}[X^2]$, $\operatorname{E}[Y]$ and $\operatorname{E}[Y^2]$ are needed, so the marginal probability functions of\ $X$ and\ $Y$ are needed.

So first, the marginal pfs are
$$
   p_X(x) = \sum_{y = -1, 1} p_{X, Y}(x, y) =
      \begin{cases}
          7/24 & \text{for $x = 0$};\\
          8/24 & \text{for $x = 1$};\\
          9/24 & \text{for $x = 2$}
      \end{cases}
$$
and
$$
   p_Y(y) = \sum_{x = 0}^2 p_{X, Y}(x, y) =
      \begin{cases}
          1/2 & \text{for $y = -1$};\\
          1/2 & \text{for $y = 1$}.
      \end{cases}
$$
Then,
\begin{align*}
   \operatorname{E}[X]   &= (7/24 \times 0) + (8/24 \times 1) + (9/24\times 2) = 26/24;\\
   \operatorname{E}[X^2] &= (7/24 \times 0^2) + (8/24 \times 1^2) + (9/24\times 2^2) = 44/24;\\
   \operatorname{E}[Y]   &= (1/2 \times -1) + (1/2 \times 1) = 0;\\
   \operatorname{E}[Y^2] &= (1/2 \times (-1)^2) + (1/2 \times 1^2) = 1,
\end{align*}
giving $\operatorname{var}[X] = 44/24 - (26/24)^2 = 0.6597222$ and $\operatorname{var}[Y] = 1 - 0^2 = 1$. 
Then,
\begin{align*}
   \operatorname{E}[XY] &= \sum_x\sum_y xy\,p_{X,Y}(x,y) \\
    &= (0\times -1 \times 1/8)  + (0\times 1 \times 1/6) + \cdots + (2\times 1 \times 1/4) \\
    &= 1/12.
\end{align*}
Hence,
$$
   \operatorname{Cov}(X,Y) 
   = \operatorname{E}[XY] - \operatorname{E}[X] \operatorname{E}[Y] 
   = \frac{1}{12} - \left(\frac{26}{24}\times 0\right) = 1/12,
$$
and
$$
   \text{Corr}(X,Y)
   = \frac{ \operatorname{Cov}(X,Y)}{\sqrt{ \operatorname{var}[X]\operatorname{var}[Y] } }
   = \frac{1/12}{\sqrt{0.6597222 \times 1}}
   = 0.1025978,
$$
so the correlation coefficient is about\ $0.10$, and a small positive linear association exists between\ $X$ and\ $Y$.
:::




```{r Joint3, echo=FALSE}
Joint3 <- array( dim = c(3, 5))

Joint3[1, ] <- c("$y = -1$", "$1/8$", "$1/4$", "$1/8$", "$1/2$")
Joint3[2, ] <- c("$y = +1$", "$1/6$", "$1/12$", "$1/4$", "$1/2$")
Joint3[3, ] <- c("Total", "$7/24$", "$1/3$", "$3/8$", "$1$")


Joint3.caption <- "A bivariate discrete probability function."
if( knitr::is_latex_output() ) {
  knitr::kable(Joint3,
               format = "latex",
               booktabs = TRUE,
               align = "r",
               longtable = FALSE,                
               escape = FALSE,
               col.names = c("",
                             "$x = 0$",
                             "$x = 1$" ,
                             "$x = 2$" ,
                             "Total"),
               caption = Joint3.caption) %>%
    kable_styling(font_size = 10) %>%
    row_spec(2, hline_after = TRUE) %>%
    row_spec(0, bold = TRUE)   %>%
    column_spec(1, bold = TRUE)
}
if( knitr::is_html_output() ) {
  knitr::kable(Joint3,
               escape = TRUE,
               col.names = c("",
                             "$x = 0$",
                             "$x = 1$" ,
                             "$x = 2$" ,
                             "Total"),
               caption = Joint3.caption,
               format = "html",
               booktabs = TRUE,
               align = "r",
               longtable = FALSE) %>%                
    row_spec(0, bold = TRUE)  %>%
    column_spec(1, bold = TRUE) 
}
```




## Properties of covariance and correlation

* The correlation has no units.
* The covariance has units; if\ $X$ is measured in kilograms and\ $Y$ in centimetres, then the units of the covariance are kg-cm.
* If the units of measurements change, the numerical value of the covariance changes, but the numerical value of the correlation stays the same.
   (For example, if\ $X$ is changed from kilograms to grams, the numerical value of the correlation will not change in value, but the numerical values of covariance will change.)
* The correlation is a number between\ $-1$ and\ $1$ (inclusive).
  When the correlation coefficient (or covariance) is negative, a *negative linear relationship* is said to exist between the two variables.
  Likewise, when the correlation coefficient (or covariance) is positive, a *positive linear relationship* is said to exist between the two variables.
* When the correlation coefficient (or covariance) is zero, no *linear* dependence is said to exist.


:::{.theorem #CovarianceProperties name="Properties of the covariance"}
For random variables\ $X$, $Y$ and\ $Z$, and constants\ $a$ and\ $b$:

* $\operatorname{Cov}(X, Y) = \operatorname{Cov}(Y, X)$.
* $\operatorname{Cov}(aX,bY) = ab\,\operatorname{Cov}(X, Y)$.
* $\operatorname{var}[aX + bY] = a^2\operatorname{var}[X] + b^2\operatorname{var}[Y] + 2ab\,\operatorname{Cov}(X, Y)$.
* If $X$ and $Y$ are independent, then $\operatorname{E}[XY] = \operatorname{E}[X]\operatorname{E}[Y]$ and hence $\operatorname{Cov}(X,Y) = 0$.
* $\operatorname{Cov}(X, Y) = 0$ does not imply\ $X$ and\ $Y$ are independent, except for the special case of the bivariate normal distribution.
:::

A zero correlation coefficient in an indication of no *linear* dependence only.
A relationship may still exist between\ $X$ and\ $Y$ even if the correlation is zero.


:::{.example #LinearDependence name="Linear dependence and correlation"}
Consider the random variable\ $X$ with the probability function:


| $x$        | $-1$  | $0$   | $1$
|------------|-------|-------|-------
| $p_{X}(x)$ | $1/3$ | $1/3$ | $1/3$


Then, define\ $Y$ to be explicitly related to\ $X$: $Y = X^2$.
A relationship definitely exists between\ $X$ and\ $Y$ (but is non-linear). 
The joint probability function for $(X, Y)$ is shown in Table\ \@ref(tab:JointRship).
Then
\begin{equation*}
   \operatorname{Cov}(X, Y)
   = \operatorname{E}[X, Y] - \operatorname{E}[X]\cdot\operatorname{E}[Y]
   = 0 - 0\times 2/3 = 0
\end{equation*}
so $\text{Corr}(X, Y) = 0$. 
But\ $X$ and\ $Y$ are *certainly related*: $Y$ was explicitly defined as a function of\ $X$.

Since the correlation is a measure of the strength of the *linear* relationship between two random variables, a correlation of zero simply is indication of no *linear* relationship between\ $X$ and\ $Y$. 
(As in this example, a non-linear relationship may exist between the variables, but no linear relationship.)
:::




```{r JointRship, echo=FALSE}
JointRship <- array( dim = c(3, 5))

JointRship[1, ] <- c("$y = 0$", "$0$", "$1/3$", "$0$", "$1/3$")
JointRship[2, ] <- c("$y = 1$", "$1/3$", "$0$", "$1/3$", "$2/3$")
JointRship[3, ] <- c("Total", "$1/3$", "$1/3$", "$1/3$", "$1$")


JointRship.caption <- "A bivariate discrete probability function."
if( knitr::is_latex_output() ) {
  knitr::kable(JointRship,
               format = "latex",
               booktabs = TRUE,
               align = "r",
               longtable = FALSE,                
               escape = FALSE,
               col.names = c("",
                             "$x = -1$", 
                             "$x = 0$", 
                             "$x = 1$", 
                             "Total"),
               caption = JointRship.caption) %>%
    kable_styling(font_size = 10) %>%
    row_spec(2, hline_after = TRUE) %>%
    row_spec(0, bold = TRUE)   %>%
    column_spec(1, bold = TRUE)
}
if( knitr::is_html_output() ) {
  knitr::kable(JointRship,
               escape = TRUE,
               col.names = c("",
                             "$x = -1$", 
                             "$x = 0$", 
                             "$x = 1$", 
                             "Total"),
               caption = JointRship.caption,
               format = "html",
               booktabs = TRUE,
               align = "r",
               longtable = FALSE) %>%                
    row_spec(0, bold = TRUE)  %>%
    column_spec(1, bold = TRUE) 
}
```



## Conditional expectations {#ConditionalExpectation}

Conditional expectations are expectations computed from a conditional distribution.
The conditional mean is the expected value computed from a conditional distribution.


:::{.definition #ConditionalExpectation name="Conditional expectation"}
The *conditional expected value* or *conditional mean* of a random variable\ $X$ for given $Y = y$ is denoted by $\operatorname{E}[X \mid Y = y]$.

If the conditional distribution is discrete with probability mass function $p_{X\mid Y=y}(x)$, then
$$
  \operatorname{E}[X \mid Y = y] =
  \displaystyle \sum_{x} x \cdot p_{X\mid Y=y}(x).
$$
If the conditional distribution is continuous with probability density function $f_{X\mid Y=y}(x)$, then
$$
  \operatorname{E}[X \mid Y = y] =
  \int_{-\infty}^\infty x \cdot f_{X\mid Y=y}(x)\, dx.
$$
:::


$\operatorname{E}[X \mid Y = y]$ is often denoted $\mu_{X \mid Y = y}$.


:::{.example #ConditionalMean name="Conditional mean (continuous)"}
Consider the two random variables\ $X$ and\ $Y$ with joint PDF
$$
   f_{X, Y}(x, y) = \frac{3}{5}(x + xy + y^2) \qquad \text{for $0 < x < 1$ and $-1 < y < 1$}.
$$
To find $f_{Y \mid X = x}(y)$, first find $f_X(x)$:
$$
   f_X(x) = \int_{-1}^1 f_{X,Y}(x,y) dy = \frac{3}{15}(6x + 2)\qquad\text{for $0 < x < 1$}. 
$$

Then,
$$
   f_{Y \mid X = x}(y)
   = \frac{ f_{X, Y}(x, y)}{ f_X(x) }
   = \frac{3(x + xy + y^2)}{6x + 2}
$$
for $-1 < y < 1$ and given $0 < x < 1$. 
The expected value of\ $Y$ given $X = x$ is then
$$
  \operatorname{E}[Y\mid X = x]
   = \frac{x}{3x + 1}
   \qquad\text{for $0 < x < 1$}.
$$
This expression indicates that the conditional expected value of\ $Y$ depends on the given value of\ $X$; for example,
\begin{align*}
   \operatorname{E}[Y\mid X = 0]   &= 0;\\
   \operatorname{E}[Y\mid X = 0.5] &= 0.2;\\
   \operatorname{E}[Y\mid X = 1]   &= 1/4.
\end{align*}
Since $\operatorname{E}[Y\mid X = x]$ depends on the value of\ $X$, then\ $X$ and\ $Y$ are *not* independent.
:::


The conditional variance is the variance computed from a conditional distribution.

:::{.definition #ConditionalVariance name="Conditional variance"}
The *conditional variance* of a random variable\ $X$ for given $Y = y$ is denoted by $\operatorname{var}[X \mid Y = y]$.

If the conditional distribution is discrete with probability mass function $p_{X\mid Y}(x)$, then
$$
  \operatorname{var}[X \mid Y = y] =
  \displaystyle \sum_{x} (x - \mu_{X\mid y})^2\, p_{X\mid Y}(x),
$$
where $\mu_{X \mid y}$ is the *conditional mean* of\ $X$ given $Y = y$.

If the conditional distribution is continuous with probability density function $f_{X\mid Y}(x)$, then
$$
  \operatorname{var}[X \mid Y = y] =
  \int_{-\infty}^\infty (x - \mu_{X\mid y})^2\, f_{X\mid Y}(x)\, dx.
$$
where $\mu_{X \mid y}$ is the *conditional mean* of\ $X$ given $Y = y$.
:::


For brevity, $\operatorname{var}[X \mid Y = y]$ is often denoted $\sigma^2_{X \mid Y = y}$.


:::{.example #ConditionVariance name="Conditional variance (continuous)"}
Refer to Example\ \@ref(exm:ConditionalMean). 
The conditional variance of\ $Y$ given $X = x$ is found by first computing $\operatorname{E}[Y^2\mid X = x]$:
\begin{align*}
   \operatorname{E}[Y^2\mid X = x]
   &= \int_{-1}^1 y^2 f_{Y\mid X = x}(y)\,dy \\
   &= \frac{3}{6x + 2} \int_{-1}^1 y^2 (x + xy + y^2)\, dy \\
   &= \frac{5x + 3}{5(3x + 1)}.
\end{align*}
The conditional variance is
\begin{align*}
   \operatorname{var}[Y\mid X = x]
   &= \operatorname{E}[Y^2\mid X = x] - \left( \operatorname{E}[Y\mid X = x] \right)^2 \\
   &= \frac{5x+3}{5(3x + 1)} - \left( \frac{x}{3x + 1}\right)^2 \\
   &= \frac{10x^2 + 14x + 3}{5(3x + 1)^2}
\end{align*}
for given $0 < x < 1$. 
Hence the variance of\ $Y$ depends on the given value of\ $X$; for example,
\begin{align*}
   \operatorname{var}[Y\mid X = 0]   &= 3/5 = 0.6\\
   \operatorname{var}[Y\mid X = 0.5] &= \frac{10\times (0.5)^2 + (14\times0.5) + 3}{5(3\times0.5 + 1)^2} = 0.4\\
   \operatorname{var}[Y\mid X = 1]   &= 27/80 = 0.3375.
\end{align*}
:::


In general, to compute the conditional variance of $X\mid Y = y$ given a joint probability function, the following steps are required.

* Find the marginal distribution of\ $Y$.
* Use this to compute the conditional probability function $p_{X \mid Y = y}(x) = p_{X, Y}(x, y)/p_{X}(x)$.
* Find the conditional mean $\operatorname{E}[X \mid Y = y]$.
* Find the conditional second raw moment $\operatorname{E}[X^2 \mid Y = y]$.
* Finally, compute $\operatorname{var}[X\mid Y = y] = \operatorname{E}[X^2\mid Y=y] - (\operatorname{E}[X\mid Y=y])^2$.


:::{.example #ConditionalVar name="Conditional variance (discrete)"}
Two discrete random variables\ $U$ and\ $V$ have the joint probability function given in Table\ \@ref(tab:CondVar).
To find the conditional variance of\ $V$ given $U = 11$, use the steps above.

First, find the marginal distribution of\ $U$:
$$
      p_U(u) = \begin{cases}
         4/9 & \text{for $u = 10$};\\
         7/18 & \text{for $u = 11$};\\
         1/6 & \text{for $u = 12$}.
         \end{cases}
$$
Secondly, compute the conditional probability function:
\begin{align*}
  p_{V\mid U = 11}(v \mid u = 11)
  &= p_{U, V}(u,v)/p_{U}(u = 11) \\
  &= \begin{cases}
        \frac{1/18}{7/18} = 1/7 & \text{if $v = 0$};\\
        \frac{1/3}{7/18}  = 6/7 & \text{if $v = 1$}
      \end{cases}
\end{align*}
using $p_U(u = 11) = 7/18$ from the Step\ 1.

Thirdly, find the conditional mean:
$$
  \operatorname{E}[V\mid U = 11] 
  = \sum_v v\, p_{V\mid U=11}(v) 
  = \left(\frac{1}{7}\times 0\right) + \left(\frac{6}{7}\times 1\right)  
  = 6/7.
$$
Fourthly, find the conditional second raw moment:
$$
  \operatorname{E}[V^2\mid U] 
  = \sum_v v^2\, p_{V\mid U=11}(v) 
  = \left(\frac{1}{7}\times 0^2\right) + \left(\frac{6}{7}\times 1^2\right)  
  = 6/7.
$$
Finally, compute:
\begin{align*}
  \operatorname{var}[V\mid U = 11] 
  &= \operatorname{E}[V\mid U = 11] - (\operatorname{E}[V\mid U = 11])^2\\
  &= (6/7) - (6/7)^2\\
  &\approx  0.1224.
\end{align*}
:::


```{r CondVar, echo=FALSE}
CondVar <- array( dim = c(3, 5))

CondVar[1, ] <- c("$v = 0$", "$1/9$", "$1/18$", "$1/6$", "$1/3$")
CondVar[2, ] <- c("$v = 1$", "$1/3$", "$1/3$", "$0$", "$2/3$")
CondVar[3, ] <- c("Total", "$4/9$", "$7/18$", "$1/6$", "$1$")


CondVar.caption <- "A bivariate discrete probability function."
if( knitr::is_latex_output() ) {
  knitr::kable(CondVar,
               format = "latex",
               booktabs = TRUE,
               align = "r",
               longtable = FALSE,                
               escape = FALSE,
               col.names = c("", 
                             "$u = 10$", 
                             "$u = 11$", 
                             "$u = 12$", 
                             "Total"),
               caption = CondVar.caption) %>%
    kable_styling(font_size = 10) %>%
    row_spec(2, hline_after = TRUE) %>%
    row_spec(0, bold = TRUE)   %>%
    column_spec(1, bold = TRUE)
}
if( knitr::is_html_output() ) {
  knitr::kable(CondVar,
               escape = TRUE,
               col.names = c("", 
                             "$u = 10$", 
                             "$u = 11$", 
                             "$u = 12$", 
                             "Total"),
               caption = CondVar.caption,
               format = "html",
               booktabs = TRUE,
               align = "r",
               longtable = FALSE) %>%                
    row_spec(0, bold = TRUE)  %>%
    column_spec(1, bold = TRUE) 
}
```


## Laws of total expectation and variance {#TotalExpectation}

The conditional expectation leads naturally to an important result: the unconditional mean of\ $X$ can be recovered by averaging the conditional mean over all values of\ $Y$.
Intuitively, the law of total expectation says that the overall mean of\ $X$ is a weighted average of the conditional means of\ $X$, weighted by the distribution of\ $Y$.


:::{.theorem #LawTotalExpectation name="Law of total expectation"}
For any two random variables\ $X$ and\ $Y$,
$$
  \operatorname{E}[X] = \operatorname{E}\big[\operatorname{E}[X \mid Y]\big].
$$
For a *discrete* random variable\ $Y$,
$$
  \operatorname{E}[X] = \sum_y \operatorname{E}[X \mid Y = y]\cdot p_Y(y).
$$
For a *continuous* random variable\ $Y$,
$$
  \operatorname{E}[X] = \int_{-\infty}^\infty \operatorname{E}[X \mid Y = y]\cdot f_Y(y)\, dy.
$$

::::{.proof}
The proof is given for the discrete case.
Starting from the definition of conditional expectation:
\begin{align*}
  \operatorname{E}\big[\operatorname{E}[X \mid Y]\big]
  &= \sum_y \operatorname{E}[X \mid Y = y]\cdot p_Y(y)\\
  &= \sum_y \left(\sum_x x\cdot p_{X\mid Y}(x)\right) p_Y(y)\\
  &= \sum_y \sum_x x\cdot \frac{p_{X,Y}(x, y)}{p_Y(y)} \cdot p_Y(y)\\
  &= \sum_x x \sum_y p_{X,Y}(x, y)\\
  &= \sum_x x\cdot p_X(x)\\
  &= \operatorname{E}[X].
\end{align*}
The continuous case is analogous, with sums replaced by integrals.
::::
:::



:::{.example #LawTotalExpectationDiscrete name="Law of total expectation (discrete)"}
Consider the joint distribution of\ $X$ and\ $Y$ in Example\ \@ref(exm:CorrCoefDiscrete), where $\operatorname{E}[X] = 26/24$.
The conditional means are:
\begin{align*}
  \operatorname{E}[X \mid Y = -1] &= \sum_x x\cdot p_{X\mid Y= -2}(x)
  = 0\times\frac{1/8}{1/2} + 1\times\frac{1/4}{1/2} + 2\times\frac{1/8}{1/2} = 1;\\
  \operatorname{E}[X \mid Y = +1] &= \sum_x x\cdot p_{X\mid Y=1}(x)
  = 0\times\frac{1/6}{1/2} + 1\times\frac{1/12}{1/2} + 2\times\frac{1/4}{1/2} = \frac{8}{6}.
\end{align*}
By the law of total expectation:
$$
  \operatorname{E}[X] 
  = \operatorname{E}[X\mid Y = -1]\cdot p_Y(-1) + \operatorname{E}[X\mid Y = +1]\cdot p_Y(+1)
  = 1\times\frac{1}{2} + \frac{8}{6}\times\frac{1}{2} = \frac{26}{24},
$$
which agrees with the value computed directly from the marginal distribution of\ $X$.
:::


:::{.example #LawTotalExpectationCPG name="Law of total expectation"}
Let $X = \sum_{i=1}^N Y_i$ where $N\sim\text{Poisson}(\lambda)$ and $Y_i\sim\text{Gamma}(\alpha, \beta)$ independently.
Conditioning on\ $N$, the conditional mean of\ $X$ given $N = n$ is:
$$
  \operatorname{E}[X\mid N = n] = \sum_{i=1}^n \operatorname{E}[Y_i] = n\alpha\beta.
$$
By the law of total expectation:
$$
  \operatorname{E}[X] 
  = \operatorname{E}\big[\operatorname{E}[X\mid N]\big] 
  = \operatorname{E}[N\alpha\beta] 
  = \alpha\beta\operatorname{E}[N] 
  = \lambda\alpha\beta,
$$
as seen later in Theorem\ \@ref(thm:CPGProperties).
:::





An analogous result holds for the variance.

:::{.theorem #LawTotalVariance name="Law of total variance"}
For any two random variables\ $X$ and\ $Y$,
$$
  \operatorname{var}[X] 
  = \operatorname{E}\big[\operatorname{var}[X \mid Y]\big] + \operatorname{var}\big[\operatorname{E}[X \mid Y]\big].
$$

::::{.proof}
Using the computational formula $\operatorname{var}[X] = \operatorname{E}[X^2] - \operatorname{E}[X]^2$, and applying 
the law of total expectation (Theorem\ \@ref(thm:LawTotalExpectation)) twice:
\begin{align*}
  \operatorname{E}\big[\operatorname{var}[X\mid Y]\big] 
  &= \operatorname{E}\big[\operatorname{E}[X^2\mid Y] - \operatorname{E}[X\mid Y]^2\big]\\
  &= \operatorname{E}[X^2] - \operatorname{E}\big[\operatorname{E}[X\mid Y]^2\big].
\end{align*}
Also,
$$
  \operatorname{var}\big[\operatorname{E}[X\mid Y]\big] 
  = \operatorname{E}\big[\operatorname{E}[X\mid Y]^2\big] - \operatorname{E}\big[\operatorname{E}[X\mid Y]\big]^2
  = \operatorname{E}\big[\operatorname{E}[X\mid Y]^2\big] - \operatorname{E}[X]^2.
$$
Adding these two expressions:
$$
  \operatorname{E}\big[\operatorname{var}[X\mid Y]\big] + \operatorname{var}\big[\operatorname{E}[X\mid Y]\big]
  = \operatorname{E}[X^2] - \operatorname{E}[X]^2 = \operatorname{var}[X].
$$
::::
:::


:::{.example #LawTotalVarianceDiscrete name="Law of total variance (discrete)"}
Suppose a fair coin is tossed: if it lands heads ($Y = 1$, probability $1/2$), 
then\ $X \sim \text{Bin}(4, 0.5)$; if it lands tails ($Y = 0$, probability $1/2$),
then\ $X \sim \text{Bin}(4, 0.9)$.

The conditional means and variances are:
\begin{align*}
  \operatorname{E}[X\mid Y = 1] &= 4\times 0.5 = 2, \quad 
  \operatorname{var}[X\mid Y = 1] = 4\times 0.5\times 0.5 = 1;\\
  \operatorname{E}[X\mid Y = 0] &= 4\times 0.9 = 3.6, \quad 
  \operatorname{var}[X\mid Y = 0] = 4\times 0.9\times 0.1 = 0.36.
\end{align*}
The overall mean is:
$$
  \operatorname{E}[X] = 2\times\frac{1}{2} + 3.6\times\frac{1}{2} = 2.8.
$$
The first term (expected conditional variance):
$$
  \operatorname{E}\big[\operatorname{var}[X\mid Y]\big] 
  = 1\times\frac{1}{2} + 0.36\times\frac{1}{2} = 0.68.
$$
The second term (variance of conditional mean):
$$
  \operatorname{var}\big[\operatorname{E}[X\mid Y]\big]
  = (2 - 2.8)^2\times\frac{1}{2} + (3.6 - 2.8)^2\times\frac{1}{2} = 0.64.
$$
By the law of total variance:
$$
  \operatorname{var}[X] = 0.68 + 0.64 = 1.32.
$$
:::


:::{.example #LawTotalVarianceCPG name="Law of total variance"}
Let $X = \sum_{i=1}^N Y_i$ where $N\sim\text{Poisson}(\lambda)$ and 
$Y_i\sim\text{Gamma}(\alpha, \beta)$ independently (as in Sect.\ \@ref(TweedieModels)).

The conditional mean and variance of\ $X$ given $N = n$ are:
$$
  \operatorname{E}[X\mid N = n] = n\alpha\beta
  \quad\text{and}\quad
  \operatorname{var}[X\mid N = n] = n\alpha\beta^2.
$$
The first term is
$$
  \operatorname{E}\big[\operatorname{var}[X\mid N]\big] 
  = \operatorname{E}[N\alpha\beta^2] 
  = \lambda\alpha\beta^2.
$$
The second term is
$$
  \operatorname{var}\big[\operatorname{E}[X\mid N]\big]
  = \operatorname{var}[N\alpha\beta]
  = \alpha^2\beta^2\operatorname{var}[N]
  = \lambda\alpha^2\beta^2.
$$
Then, by the law of total variance
$$
  \operatorname{var}[X] 
  = \lambda\alpha\beta^2 + \lambda\alpha^2\beta^2
  = \lambda\alpha\beta^2(1 + \alpha).
$$
:::


## Exercises {#ExercisesChapExpectationBivariate}

Selected answers appear in Sect.\ \@ref(AnswersExpectationBivariate).


::::{.exerciseList}
:::{.exercise #AandBsimple}
The random variable\ $A$ has mean\ $13$ and variance\ $5$. 
The random variable\ $B$ has mean\ $4$ and variance\ $2$. 
Assuming\ $A$ and\ $B$ are independent, find:

1. $\operatorname{E}[A + B]$.
1. $\operatorname{var}[A + B]$.
1. $\operatorname{E}[2A - 3B]$.
1. $\operatorname{var}[2A - 3B]$.
:::


:::{.exercise #AandBsimple2}
Repeat Exercise\ \@ref(exr:AandBsimple), but with $\operatorname{Cov}(A, B) = 0.2$.
:::


:::{.exercise #CovLinearCombo}
Suppose $\operatorname{E}[X] = 2$, $\operatorname{E}[Y] = 3$, $\operatorname{var}[X] = 4$, $9\operatorname{var}[Y] = 9$ and $\operatorname{Cov}(X,Y) = 3$.
Define $U = 2X - Y$ and $V = X + 2Y$.

1. Find $\operatorname{E}[U]$ and $\operatorname{E}[V]$.
1. Find $\operatorname{var}[U]$ and $\operatorname{var}[V]$.
1. Find $\operatorname{Cov}(U, V)$.
1. Find $\operatorname{Corr}(U, V)$.
:::


:::{.exercise #CovLinearCombo2}
Suppose $\operatorname{E}[X] = -1$, $\operatorname{E}[Y] = 4$, $\operatorname{var}[X] = 9$,
$\operatorname{var}[Y] = 1$ and $\operatorname{Cov}(X,Y) = -2$.
Define $S = X + 3Y$ and $T = 2X - Y$.

1. Find $\operatorname{E}[S]$ and $\operatorname{E}[T]$.
2. Find $\operatorname{var}[S]$ and $\operatorname{var}[T]$.
3. Find $\operatorname{Cov}(S, T)$.
4. Find $\operatorname{Corr}(S, T)$.
:::
::::

