
# Transformations of bivariate random variables {#ChapterTransformationsBivariate}

```{r setup, include=FALSE}
source("R/SetColours.R")
```


::: {.objectivesBox .objectives data-latex="{iconmonstr-target-4-240.png}"}
On completion of this chapter, you should be able to:

* find the joint distribution of two transformed variables in a bivariate situation, using the change of variable method.
* find the joint distribution of two transformed variables in a bivariate situation, using the distribution function method.
:::


## Bivariate change of variable method {#BivariateChangeOfVariable}

### Bivariate discrete case {#ChangeOfVarDiscreteBivariate}

The bivariate discrete case is similar to the univariate discrete case.
Consider a joint probability function $p_{X_1, X_2}(x_1, x_2)$ of two discrete random variables\ $X_1$ and\ $X_2$ defined on the two-dimensional set of points\ $R^2_X$ for which $p(x_1, x_2) > 0$.
Two *one-to-one transformations* need to be defined:
$$
   y_1 = u_1( x_1, x_2)\qquad\text{and}\qquad y_2 = u_2( x_1, x_2)
$$
that map\ $R^2_X$ onto\ $R^2_Y$ (the two-dimensional set of points for which $p(y_1, y_2) > 0$).
Denote the two inverse functions as
$$
   x_1 = w_1( y_1, y_2)\qquad\text{and}\qquad x_2 = w_2( y_1, y_2).
$$
Then the joint probability function of the new (transformed) random variables is
$$
   p_{Y_1, Y_2}(y_1, y_2) =
      p_{X_1, X_2}\big( w_1(y_1, y_2), w_2(y_1, y_2)\big) \qquad \text{where $(y_1, y_2)\in R^2_Y$}.
$$


:::{.example #TransformBivariate name="Transformation discrete (bivariate)"}
Let the two discrete random variables\ $X_1$ and\ $X_2$ have the joint probability function shown in Table\ \@ref(tab:Bivar4).
Consider the two one-to-one transformations
$$
   Y_1 = |X_1 + X_2| \qquad\text{and}\qquad Y_2 = X_1 + 1.
$$
The joint probability function of\ $Y_1$ and\ $Y_2$ can be found by noting where the $(x_1, x_2)$ pairs are mapped to in the $y_1, y_2$ space:


$\Pr(x_1, x_2)$ | $(x_1, x_2)$ | $\mapsto$ | $(y_1,y_2)$
:--------------:|:-----------:|:----------:|:----------------:
$0.3$           | $(-1, 0)$   | $\mapsto$  | $(1, 1)$ 
$0.1$           | $(-1, 1)$   | $\mapsto$  | $(0, 2)$
$0.1$           | $(-1, 2)$   | $\mapsto$  | $(1, 3)$
$0.2$           | $(1, 0)$    | $\mapsto$  | $(1, 1)$ 
$0.2$           | $(1, 1)$    | $\mapsto$  | $(2, 2)$ 
$0.1$           | $(1, 2)$    | $\mapsto$  | $(3, 3)$ 


The joint probability function can then be constructed as shown in Table\ \@ref(tab:Bivar5).
:::




```{r Bivar4, echo=FALSE}
Bivar4Table <- array( dim = c(2, 4))

Bivar4Table[1, ] <- c("$x_1 = -1$", "$0.3$", "$0.1$", "$0.1$")
Bivar4Table[2, ] <- c("$x_1 = +1$", "$0.2$", "$0.2$", "$0.1$")


Bivar4Table.caption <- "A bivariate probability function for $X_1$ and $X_2$."
if( knitr::is_latex_output() ) {
  knitr::kable(Bivar4Table,
               format = "latex",
               booktabs = TRUE,
               col.names = c("",
                             "$x_2 = 0$",
                             "$x_2 = 1$",
                             "$x_2 = 2$"),
               align = "r",
               longtable = FALSE,                
               escape = FALSE,
               caption = Bivar4Table.caption) %>%
    kable_styling(font_size = 10) %>%
    row_spec(0, bold = TRUE) 
}
if( knitr::is_html_output() ) {
  knitr::kable(Bivar4Table,
               escape = TRUE,
               format = "html",
               booktabs = TRUE,
               align = "r",
               longtable = FALSE,                
               col.names = c("",
                             "$x_2 = 0$",
                             "$x_2 = 1$",
                             "$x_2 = 2$"),
               caption = Bivar4Table.caption) %>%
    row_spec(0, bold = TRUE) 
}
```





```{r Bivar5, echo=FALSE}
Bivar5Table <- array( dim = c(3, 5))

Bivar5Table[1, ] <- c("$y_2 = 1$", "$0.0$", "$0.5$", "$0.0$", "$0.0$")
Bivar5Table[2, ] <- c("$y_2 = 2$", "$0.1$", "$0.0$", "$0.2$", "$0.0$")
Bivar5Table[3, ] <- c("$y_2 = 3$", "$0.0$", "$0.1$", "$0.0$", "$0.1$")


Bivar5Table.caption <- "The joint probability function for $Y_1$ and $Y_2$"
if( knitr::is_latex_output() ) {
  knitr::kable(Bivar5Table,
               format = "latex",
               booktabs = TRUE,
               col.names = c("",
                             "$y_1 = 0$",
                             "$y_1 = 1$",
                             "$y_1 = 2$",
                             "$y_1 = 3$"),
               align = "r",
               longtable = FALSE,                
               escape = FALSE,
               caption = Bivar5Table.caption) %>%
    kable_styling(font_size = 10) %>%
    row_spec(0, bold = TRUE) 
}
if( knitr::is_html_output() ) {
  knitr::kable(Bivar5Table,
               escape = TRUE,
               format = "html",
               booktabs = TRUE,
               align = "r",
               longtable = FALSE,                
               col.names = c("",
                             "$y_1 = 0$",
                             "$y_1 = 1$",
                             "$y_1 = 2$",
                             "$y_1 = 3$"),
               caption = Bivar5Table.caption) %>%
    row_spec(0, bold = TRUE) 
}
```


Sometimes, a joint probability function of two random variables is given, but only *one* new random variable is required.
In this case, a second (dummy) transformation is used, usually a very simple transformation.


:::{.example #TransformBivariate2 name="Transformation (bivariate)"}
Let\ $X_1$ and\ $X_2$ be two independent Poisson random variables:
\begin{align*}
  p_{X_1}(x_1) &= \frac{\lambda_1^{x_1} \exp(-\lambda_1)}{x_1!}\ \text{for $x_1 = 0, 1, 2, \dots$}
  \quad\text{and}\\
  p_{X_2}(x_2) &= \frac{\lambda_2^{x_2} \exp(-\lambda_2)}{x_2!}\ \text{for $x_2 = 0, 1, 2, \dots$}.
\end{align*}
The joint probability function is
$$
   p_{X_1, X_2}(x_1, x_2) =
         \frac{\lambda_1^{x_1} \lambda_2^{x_2} \exp( -\lambda_1 - \lambda_2 )}{x_1!\, x_2!}
         \quad\text{for $x_1$ and $x_2 = 0, 1, 2, \dots$}
$$
Suppose we wish to find the probability function of $Y_1 = X_1 + X_2$.

Consider the two *one-to-one transformations* (where $Y_2 = X_2$ is just a dummy transformation since we need to work with *two* transformations):
\begin{align}
   y_1 &= x_1 + x_2             = u_1(x_1, x_2)\\
   y_2 &= x_2\phantom{{} + x_2} = u_2(x_1, x_2)
\end{align}
which map the points in\ $\mathcal{R}^2_X$ onto
$$
   \mathcal{R}^2_Y = \left\{ (y_1, y_2)\mid y_1 = 0, 1, 2, \dots; y_2 = 0, 1, 2, \dots, y_1\right\}.
$$
$Y_2$ is a dummy transform, and is chosen to be very simple; any second transform could be chosen (since it is not of direct interest).
The inverse functions are
\begin{align*}
   x_1 &= y_1 - y_2              = w_1(y_1, y_2)\\
   x_2 &= y_2 \phantom{{} - y_2} = w_2(y_2)
\end{align*}
by rearranging the original transformations.
Then the *joint* probability function of\ $Y_1$ and\ $Y_2$ is
\begin{align*}
   p_{Y_1, Y_2}(y_1, y_2)
   &= p_{X_1, X_2}\big( w_1(y_1, y_2), w_2(y_1, y_2)\big) \\
   &= \frac{\lambda_1^{y_1 - y_2}\lambda_2^{y_2} \exp(-\lambda_1 - \lambda_2)}{(y_1 - y_2)! y_2!}\quad
   \text{for $(y_1, y_2)\in \mathcal{R}^2_Y$}.
\end{align*}
Recall that we seek the probability function of just\ $Y_1$, so we need to find the marginal probability function of $p_{Y_1, Y_2}(y_1, y_2)$.
The marginal probability function of\ $Y_1$ is
$$
   p_{Y_1}(y_1) 
   = \sum_{y_2 = 0}^{y_1} p_{Y_1, Y_2}(y_1, y_2)
   = \sum_{y_2 = 0}^{y_1} \frac{\lambda_1^{y_1 - y_2}\lambda_2^{y_2} \exp(-\lambda_1 - \lambda_2)}{(y_1 - y_2)!\, y_2!}.
$$
To simplify, first multiply and divide by\ $y_1!$:
\begin{align*}
  p_{Y_1}(y_1)
  &= \frac{\exp(-\lambda_1 - \lambda_2)}{y_1!} 
    \sum_{y_2 = 0}^{y_1} \frac{y_1!}{(y_1 - y_2)!\, y_2!} \lambda_1^{y_1 - y_2}\lambda_2^{y_2}\\
  &= \frac{\exp(-(\lambda_1 + \lambda_2))}{y_1!} 
     \sum_{y_2 = 0}^{y_1} \binom{y_1}{y_2} \lambda_1^{y_1 - y_2}\lambda_2^{y_2}.\\ 
\end{align*}
Then, using the binomial theorem (Sect.\ \@ref(FiniteSeries))\index{Binomial theorem}, this simplifies to
\begin{align*}
  p_{Y_1}(y_1)
  &= \frac{\exp\big(-(\lambda_1 + \lambda_2)\big)}{y_1!} 
     (\lambda_1 + \lambda_2)^{y_1} \\
  &= \frac{\exp(-\Lambda)}{y_1!} \Lambda^{y_1} \quad\text{for $y_1 = 0, 1, 2, \dots$}
\end{align*}
where $\Lambda = \lambda_1 + \lambda_2$.
This is the probability function of a Poisson random variable (Def.\ \@ref(def:PoissonDistribution)) with mean $\Lambda$.
Thus $Y_1 = X_1 + X_2$ has a Poisson distribution with a mean of $\lambda_1 + \lambda_2$: the sum of two independent Poisson random variables is also a Poisson random variable.
:::


### Bivariate continuous case {#ChangeOfVarContinuousBivariate}

With two continuous random variables $(X_1, X_2)$ with a joint PDF $f_{X_1, X_2}(x_1, x_2)$, two transformations are needed to transform them into two new random variables $(Y_1, Y_2)$:
$$
  Y_1 = u_1(X_1, X_2) \quad \text{and} \quad Y_2 = u_2(X_1, X_2).
$$
The joint PDF of the new variables is
$$
  f_{Y_1, Y_2}(y_1, y_2) = f_{X_1, X_2}(x_1, x_2) \cdot |\det J|
$$
where $|\det(J)|$ is the *absolute value* of the *Jacobian determinant*:\index{Jacobian!determinant}
$$
  J = 
  \begin{bmatrix} 
    {\partial x_1}/{\partial y_1} &  {\partial x_1}/{\partial y_2} \\[6pt]
    {\partial x_2}/{\partial y_1} &  {\partial x_2}/{\partial y_2} 
  \end{bmatrix},
$$
so that
$$
  \det(J)
  =
    \frac{\partial x_1}{\partial y_1}
    \frac{\partial x_2}{\partial y_2}
    -
    \frac{\partial x_1}{\partial y_2}
    \frac{\partial x_2}{\partial y_1}.
$$
The Jacobian plays an equivalent role to $\left|dx/dy\right|$ in the univariate case.


:::{.example #BivarContExample name="Transformation continuous (bivariate)"}
Let\ $X_1$ and\ $X_2$ be independent exponential random variables with $\lambda = 1$, such that both PDFs are
$$
  f_{X_1}(x_1) = \exp(-x_1)\quad\text{for $x_1 > 0$}
  \quad\text{and}\quad
  f_{X_2}(x_2) = \exp(-x_2)\quad\text{for $x_2 > 0$}.
$$
That is, $X_1, X_2 \overset{\text{iid}}{\sim} \text{Exponential}(\lambda = 1)$.\index{iid}
Consider the two transformations $Y_1 = X_1 + X_2$ (the sum) and $Y_2 = X_1 / X_2$ (the ratio).
Then, $y_1 > 0$ and $y_2 > 0$.

Since the random variable\ $X_1$ and\ $X_2$ are iid, the joint PDF of\ $X_1$ and\ $X_2$ is
$$
  f_{X_1, X_2}(x_1, x_2) = \exp(-x_1) \cdot \exp(-x_2) = \exp\left(-(x_1 + x_2)\right) \quad \text{for $x_1, x_2 > 0$}.
$$
Inverting the transformations gives
$$
  x_1 = \frac{y_1 y_2}{1 + y_2}
  \quad\text{and}\quad
  x_2 = \frac{y_1}{1 + y_2}.
$$
The Jacobian requires the four partial derivatives:
$$
\begin{array}{ll}
  \displaystyle \frac{\partial x_1}{\partial y_1} = \frac{y_2}{1+y_2};\quad & \quad\displaystyle \frac{\partial x_1}{\partial y_2} = \frac{y_1(1+y_2) - y_1 y_2}{(1+y_2)^2} = \frac{y_1}{(1+y_2)^2};\\[12pt]
  \displaystyle \frac{\partial x_2}{\partial y_1} = \frac{1}{1+y_2};\quad & \quad\displaystyle \frac{\partial x_2}{\partial y_2} = \frac{-y_1}{(1+y_2)^2}.
\end{array}
$$
So the Jacobian determinant is 
\begin{align*}
  \det J 
  &= \left( \frac{y_2}{1+y_2} \cdot \frac{-y_1}{(1+y_2)^2} \right) - \left( \frac{1}{1+y_2} \cdot \frac{y_1}{(1+y_2)^2} \right)\\
  &= \frac{-y_1 y_2 - y_1}{(1+y_2)^3}\\
  &= \frac{-y_1(y_2 + 1)}{(1+y_2)^3}.
\end{align*}
Finally, substitute\ $x_1$ and\ $x_2$ and\ $\det J$ into the formula:
$$
  f_{Y_1, Y_2}(y_1, y_2) = \exp(-(x_1 + x_2)) \cdot \left| \frac{-y_1}{(1 + y_2)^2} \right|.
$$
Since $x_1 + x_2 = y_1$, write
$$
  f_{Y_1, Y_2}(y_1, y_2) = \frac{y_1 \exp(-y_1)}{(1+y_2)^2}\quad\text{for $y_1 > 0$ and $y_2 > 0$}.
$$
:::




:::{.example #SquaredNormalDistribution name="Distribution of squared normal rv"}
Let $X \sim N(0, 1)$ and $Y = X^2$. 
Since $u(x) = x^2$ is not monotone on $\mathbb{R}$, the change of variable method requires partitioning; the distribution function method is simpler.

First, see that the support for $X$ is over\ $\mathbb{R}$, but the support for $Y = X^2$ must be over the non-negative reals.
Furthermore, since individual points have zero probability for continuous random variables, $X = 0$ is not technically possible, so the support is best given as $(0, \infty)$ or $[0, \infty)$ (see Fig.\ \@ref(fig:DFcontinuousUnivariate)).

So, for $y > 0$:
$$
  F_Y(y) = \Pr(X^2 \leq y) = \Pr(-\sqrt{y} \leq X \leq \sqrt{y})
  = 
  F_X(\sqrt{y}) - F_X(-\sqrt{y}) = 2F_X(\sqrt{y}) - 1.
$$

Differentiating:
$$
  f_Y(y) = 2 f_X(\sqrt{y}\,) \cdot \frac{1}{2\sqrt{y}}
  = \frac{1}{\sqrt{y}} \cdot \frac{1}{\sqrt{2\pi}} \exp(-y/2)
  = \frac{y^{1/2 - 1} \exp(-y/2)}{2^{1/2}\,\Gamma(1/2)}, \quad y > 0.
$$
Notice that $Y > 0$; if $Y = 0$ the PMF is undefined.
The final expressions identifies this as the $\chi^2(1)$ distribution\index{Chi-squared distribution} (i.e., $\text{Gamma}(1/2,\, 1/2)$).
:::


```{r DFcontinuousUnivariate, echo=FALSE, fig.align="center", out.width="100%", fig.width=8, fig.height=3.25, fig.cap="A normal distribution (left panel), the transformation (centre panel), and the resulting chi-squared distribution (right panel)."}
x <- seq(-4, 4,
         length = 100)
fX <- dnorm(x)

y <- seq(0, 16,
         length = 100)
fY <- dchisq(y, 
            df = 1)

par( mfrow = c(1, 3))

plot( fX ~ x,
      type = "l",
      las = 1,
      lwd = 3,
      main = expression(bold(PDF~of~italic(X))),
      xlab = expression(italic(x)),
      ylab = "PDF")

X2 <- x ^2
plot(X2 ~ x, 
     type = "l", 
     lwd = 3,
     las = 1,
     xlab = expression(italic(x)),
     ylab = expression(italic(y)==italic(x)^2),
     main = "The transformation")
abline(h = 4,
       lwd = 2, 
       lty = 3)
text(x = 0,
     y = 4,
     pos = 3,
     label = expression(sqrt(italic(y))))
segments(x0 = 2,
         x1 = 2,
         y0 = 0,
         y1 = 4,
         lty = 3,
         lwd = 2)
segments(x0 = -2,
         x1 = -2,
         y0 = 0,
         y1 = 4,
         lty = 3,
         lwd = 2)


plot( fY ~ y,
      type = "l",
      lwd = 3,
      las = 1,
      main = expression(bold(PDF~of~italic(Y)==italic(X)^2)),
      xlab = expression(italic(y)),
      ylab = "PDF")
```





## Bivariate distribution function method {#BivariateDF}

### Bivariate continuous case {#BivariateDFcontinuous}

Let $(X_1, X_2)$ have joint PDF $f_{X_1, X_2}(x_1, x_2)$. 
To find the distribution of $Y = u(X_1, X_2)$ (a scalar function of\ $X_1$ and\ $X_2$), compute
$$
  F_Y(y) = \Pr(u(X_1, X_2) \leq y) 
  = 
  \iint_{\{(x_1,x_2):\, u(x_1,x_2) \leq y\}} f_{X_1, X_2}(x_1, x_2)\, dx_1\, dx_2,
$$
then differentiate: $f_Y(y) = F_Y'(y)$.

The key step is characterising the region of integration $\{u(x_1,x_2) \leq y\}$ carefully as a function of\ $y$, paying attention to how its boundaries move as\ $y$ changes. 
This is often where the difficulty lies, and a sketch of the support is strongly recommended.



:::{.example #JointMaximumUniform name="Distribution of $M = \max(X_1, X_2)$ for independent uniforms distributions"}
Let $X_1, X_2 \overset{\text{iid}}{\sim} \text{Uniform}(0,1)$ with joint PDF $f_{X_1, X_2}(x_1, x_2) = 1$ on the unit square (Fig.\ \@ref(fig:BivariateContinuous), left panel).
Define $M = \max(X_1, X_2)$.

The event $\{M \leq m\}$ is equivalent to $\{X_1 \leq m \text{ and } X_2 \leq m\}$ (see Fig.\ \@ref(fig:BivariateContinuous)),
so for $m \in [0, 1]$:
$$
  F_M(m) = \Pr(X_1 \leq m,\, X_2 \leq m) = m \cdot m = m^2.
$$
Differentiating, the PDF is
$$
  f_M(m) = 2m, \quad m \in [0, 1].
$$
This is the $\text{Beta}(2, 1)$ distribution (Fig.\ \@ref(fig:BivariateContinuous), right panel). 
More generally, for $X_1, X_2, \dots, X_n \overset{\text{iid}}{\sim} \text{Uniform}(0,1)$, we find $M = \max(X_1,\ldots,X_n)$ has CDF $m^n$ and PDF $nm^{n-1}$, a $\text{Beta}(n, 1)$ distribution. 
This extends naturally to order statistics (Chap.\ \@ref(OrderStatisticsChapter)).
:::


```{r BivariateContinuous, echo=FALSE, fig.width = 8, fig.height = 4.5, fig.align="center", fig.cap="The joint distribution of the maximum from two uniform distributions.", out.width='90%'}
set.seed(42)
N <- 80000
X1 <- runif(N)
X2 <- runif(N)

M  <- pmax(X1, X2)

m_seq <- seq(0, 1, 
             length = 300)

par(mfrow = c(1, 2), 
    mar = c(4, 4, 3, 1))

m_ex <- 0.65

# Show the integration region
plot(c(0,1), 
     c(0,1), 
     type = "n",
     xlab = expression(italic(x)[1]), 
     ylab = expression(italic(x)[2]),
     main = expression(bold("Region {max(X"[1]*", X"[2]*") " <= " m}")),
     axes = FALSE,
     las = 1, 
     asp = 1)
axis(side = 1,
     at = c(0, 0.2, 0.4, m_ex, 0.8, 1.0),
     labels = c(0, 0.2, 0.4, expression(italic(m)), 0.8, 1.0))
axis(side = 2,
     las = 1,
     at = c(0, 0.2, 0.4, m_ex, 0.8, 1.0),
     labels = c(0, 0.2, 0.4, expression(italic(m)), 0.8, 1.0))
box()

rect(0,    0, 
     m_ex, m_ex, 
     col = "lightgrey", 
     border = NA)
rect(0, 0, 
     1, 1,
     border = "darkgrey", 
     lwd = 1.5)

abline(v = m_ex, 
       lty = 2, 
       col = "grey", 
       lwd = 1.5)
abline(h = m_ex, 
       lty = 2, 
       col = "grey", 
       lwd = 1.5)

text(x = m_ex / 2, 
     y = m_ex / 2,
     expression(paste("Area = ", italic(m)^2)), 
     cex = 1)
grid(col = "darkgrey", 
     lty = 1)

# Derived pdf vs simulation
plot( x = c(0, 1),
      y = c(0, 2), 
      lwd = 3,
      type = "l",
      las = 1,
      xlab = expression(italic(m)),
      ylab = "PDF",
      main = expression(bold(PDF~of~italic(M))))
grid(col = "grey85", 
     lty = 1)

par(mfrow = c(1, 1))
```



### Bivariate discrete case {#BivariateDFdiscrete}

Let $(X_1, X_2)$ have joint PMF $p_{\mathbf{X}}(x_1, x_2)$. 
For $Y = u(X_1, X_2)$, the CDF of\ $Y$ is
$$
  F_Y(y) = \Pr(Y \leq y) = \sum_{\{(x_1,x_2):\, g(x_1,x_2) \leq y\}} p_{\mathbf{X}}(x_1, x_2),
$$
and the PMF is recovered as $p_Y(y) = F_Y(y) - F_Y(y^-)$ where $y^-$ denotes the largest value in the support of\ $Y$ strictly less than\ $y$.

The bivariate discrete CDF method is often the most natural approach when\ $u(\cdot)$ is a many-to-one function\index{Transformation!many-to-one} (such as a maximum or minimum), where enumerating pre-images\index{Pre-image} directly would be cumbersome, but the event $\{u(X_1, X_2) \leq y\}$ has a simple geometric or combinatorial description.


:::{.example #TwoIndBinomials name="Maximum of two independent binomial distributions"}
Let $X_1 \sim \text{Binomial}(n_1, p)$ and $X_2 \sim \text{Binomial}(n_2, p)$ independently. 
Define $M = \max(X_1, X_2)$.

Using the same ideas as in the continuous case, for $m \in \{0, 1, \ldots, \max(n_1, n_2)\}$:
$$
  F_M(m) = \Pr(X_1 \leq m,\, X_2 \leq m) = F_{X_1}(m) \cdot F_{X_2}(m)
$$
by independence. 
The probability mass function is then
$$
  p_M(m) = F_M(m) - F_M(m - 1)
  = 
  F_{X_1}(m)F_{X_2}(m) - F_{X_1}(m-1)F_{X_2}(m-1).
$$
This does not simplify to a standard named distribution in general; the CDF method can yield the distribution even when it has no closed-form family name.
:::


```{r bivariate-discrete, echo=FALSE, fig.width = 9, fig.height = 5, out.width="100%"}
n1 <- 10; n2 <- 8; p <- 0.4
m_vals <- 0:max(n1, n2)

# CDF of M via independence
F_X1 <- pbinom(m_vals, n1, p)
F_X2 <- pbinom(m_vals, n2, p)
F_M  <- F_X1 * F_X2

# PMF of M
p_M  <- diff(c(0, F_M))

# Simulation to verify
set.seed(7)
N   <- 100000
sim <- pmax(rbinom(N, n1, p), rbinom(N, n2, p))
sim_pmf <- table(sim) / N

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# CDF plot
plot(m_vals, F_M, 
     type = "s", 
     col = ColourSolid,
     lwd = 2,
     xlab = expression(italic(m)), 
     ylab = "F(m)",
     main = expression("CDF of M = max(X"[1]*", X"[2]*")"),
     ylim = c(0, 1), 
     las = 1)
points(m_vals, 
       F_M, 
       pch = 16, 
       col = ColourSolid)
lines(m_vals, 
      F_X1, 
      lty = 2, 
      col = ColourLight,  
      lwd = 1.5)
lines(m_vals, 
      F_X2, 
      lty = 3, 
      col = ColourVeryLight,
      lwd = 1.5)
legend("topleft",
       legend = c(expression("F"[M]*"(m)"),
                  expression("F"[X[1]]*"(m)  Bin(10,0.4)"),
                  expression("F"[X[2]]*"(m)  Bin(8,0.4)")),
       col = c(ColourSolid, ColourLight, ColourVeryLight),
       lty = c(1,2,3), 
       lwd = 2, 
       bty = "n", 
       cex = 0.85)
grid(col = "grey85", 
     lty = 1)

# PMF: derived vs simulated
plot(m_vals, 
     p_M, 
     type = "h", 
     col = ColourLight, 
     lwd = 4,
     xlab = expression(italic(m)), 
     ylab = "Probability",
     main = expression("PMF of M = max(X"[1]*", X"[2]*")"),
     ylim = c(0, max(p_M) * 1.15), 
     las = 1)
points(as.integer(names(sim_pmf)), 
       as.numeric(sim_pmf),
       pch = 4, 
       col = ColourSolid, 
       cex = 1.5, 
       lwd = 2)
legend("topleft",
       legend = c("Derived pmf", "Simulated"),
       col = c(ColourLight, ColourSolid),
       lty = c(1, NA), 
       pch = c(NA, 4),
       lwd = c(4, 2), 
       bty = "n")
grid(col = "grey85", 
     lty = 1)

par(mfrow = c(1, 1))
```



## Exercises {#ExercisesTransformationsBivariate}

Selected answers appear in Sect.\ \@ref(AnswersTransformationsBivariate).


::::{.exerciseList}
:::{.exercise #JointDiscrete}
The discrete bivariate random vector $(X_1, X_2)$ has the joint probability function
$$
   f_{X_1, X_2}(x_1, x_2) = (2x_1+ x _2)/6 \qquad \text{for $x_1 = 0, 1$ and $x_2 = 0, 1$}.
$$
Consider the transformations
\begin{align*}
   Y_1 &= X_1 + X_2 \\
   Y_2 &= \phantom{X_1+{}} X_2
\end{align*}

1. Determine the joint probability function of $(Y_1, Y_2)$.
1. Deduce the distribution of\ $Y_1$.
:::



:::{.exercise #XformSimBetaMaxuniforms}
As in Example\ \@ref(exm:JointMaximumUniform), consider the maximum of two independent continuous uniform distributions $X_1$ and $X_2$ defined on $[0, 1]$.

1. Create an **R** simulation to show that the distribution of $M = \max(X_1, X_2)$ is a $\text{Beta}(2, 1)$ distribution. 2. Extend the simulation to show that the distribution of $M = \max(X_1, \dots, X_n)$ is a $\text{Beta}(n, 1)$ distribution for $n = 3, 5$ and\ $10$, for $X_n \overset{\text{iid}}{\sim} \text{Uniform}(0, 1)$. 
:::
::::




