# (PART) Multivariate random variables and distributions* {-}



```{r setup, include=FALSE}
source("R/SetColours.R")
```



<!-- PERHAPS: a chapter (or more) for bivariate; a chapter (or more) for multivariate with vector/matrix notation. -->


# Multivariate distributions* {#ChapMultivariate}

::: {.objectivesBox .objectives data-latex="{iconmonstr-target-4-240.png}"}
On completion of this chapter you should be able to:

* apply the concept of multivariate random vectors.
* compute joint probability functions and the distribution function for multivariate random vectors.
* find the marginal and conditional probability functions of random vectors in both discrete and continuous cases.
* compute the expectation and variance of linear combinations of random variables.
* interpret and compute the covariance and the coefficient of correlation between two random variables.
* compute the conditional mean and conditional variance of a random variable for some given value of another random variable.
* use the multinomial distribution.
:::


## Introduction {#MultivariateIntroduction}

Not all random processes are sufficiently simple to have the outcome denoted by a single variable\ $X$. 
Many situations require observing two or more numerical characteristics simultaneously.
This chapter mainly discusses the two-variable (bivariate) case, but also discusses the multivariable (more than two variables) case using matrix notation (Sect.\ \@ref(Multivariate)).


## Multivariate random variables and matrix notation {#Multivariate}

So far, we have studied *univariate* random variables (i.e., a single random variable) 
and *bivariate* random variables (two jointly distributed random variables). 
These ideas can be extended to more random variables; the case of *multivariate* random variables, where several random variables are considered simultaneously.  

To do this, using *random vectors* is convenient.\index{Random vector}
A random vector is a column vector of $n$\ random variables:
$$
  \mathbf{X} = [X_1, X_2, \dots, X_n]^T,
$$
where the superscript\ $T$ means 'transpose'.
Each\ $X_i$ is a random variable, and together they form an $n$-dimensional random vector.
In the bivariate case, for example, we could write
$$
   \mathbf{X} 
   = \begin{bmatrix} 
       X_1 \\ 
       X_2 
     \end{bmatrix}.
$$

The $n$\ random variables do not have to have the same distribution, though this is common.
For example, a common scenario is for $n$\ random variables all being drawn from the same population (see, for example, Chap.\ \@ref(ChapDescribingSamples)).



:::{.example #Dice3DRV name="Random vector"}
Consider the random vector $\mathbf{X} = (X_1, X_2, X_3)$ representing the outcome of independently rolling three fair six-sided dice.
Then,
$$
  X_i \sim \text{Discrete uniform}(1, 6)
$$
for $i = 1, 2, 3$.
Each component has the same distribution, but the outcome from the three rolls may be different.
:::


:::{.example #Dice3DRV name="Random vector"}
Consider rolling four fair, six-sided dice independently.
Define the random vector $\mathbf{Y} = (Y_1, Y_2, Y_3, Y_4)$, where $Y_i$ is the roll on which the $i$th occurrence of a six is observed when repeatedly rolling die\ $i$.
(For example, $Y_3$ is the roll on which the third six is observed on Die\ 3.)

Then,
$$
  Y_i \sim \text{Negative binomial}(i, p = 1/6)\quad\text{for $y_i = i, i + 1, i + 2, \dots$}
$$
for $i = 1, 2, 3, 4$, where the *alternative parameterisation* of the negative binomial distribution (Sect.\ \@ref(NegBinAlternative)) has been used.
The components have different negative binomial distributions, and different support:
$$
  S_{Y_i} = \{i, i + 1, i +2, \dots\}.
$$
:::



## Joint probability functions {#MultivariateProbFn}

The *joint probability function* of $\mathbf{X}$ describes the probability distribution for all $n$\ random variables simultaneously.\index{Joint probability function}


:::{.definition #JointPMF name="Joint probability function (discrete)"}
Let $\mathbf{X} = (X_1, \dots, X_n)$ be an $n$-dimensional discrete random variable. 
The random variable $X_j$ (for $j = 1, \dots, n$) takes values in some set $S_j$ (the range).  

Then, the range of the random vector is\index{Range!for random vectors}
$$
  \mathcal{X} \subseteq S_1\times \cdots \times S_n.
$$
The *joint probability mass function* is\index{Joint probability mass function}
$$
  p_{\mathbf{X}}mathbf{x}) 
  = \Pr(\mathbf{X} = \mathbf{x}) \quad \text{for $\mathbf{x} \in \mathcal{X}$},
$$
such that 
$$
  p_{\mathbf{X}}(\mathbf{x}) \geq 0 
  \quad \text{for all $\mathbf{x}) \in \mathcal{X}$},
$$
and
$$
  \sum_{\mathbf{x} \in \mathcal{X}} p_{\mathbf{X}}(\mathbf{x}) = 1.
$$
The support is\index{Support!for random vectors}
$$
  \{\mathbf{x} \in \mathcal{X}: p_{\mathbf{X}}(\mathbf{x}) > 0\}.
$$
:::


:::{.example #Dice3D name="Three dice"}
Consider the random vector $\mathbf{X} = (X_1, X_2, X_3)$ representing the outcome of rolling three fair six-sided dice (Example\ \@ref(exm:Dice3DRV)).  

Since each $X_j \in \{1, 2, 3, 4, 5, 6\}$, the range is
$$
  \mathcal{X} = \{1, 2, 3, 4, 5, 6\} \times \{1, 2, 3, 4, 5, 6\} \times \{1, 2, 3, 4, 5, 6\}.
$$
Since the dice rolls are independent, the joint probability mass function is
\begin{align*}
  p_{X_1, X_2, X_3}(x_1, x_2, x_3) 
  &= 
  \Pr(X_1 = x_1, X_2 = x_2, X_3 = x_3)\\ 
  &=   \Pr(X_1 = x_1)^3 = \frac{1}{6^3} = \frac{1}{216}
\end{align*}
for $(x_1, x_2, x_3) \in \mathcal{X}$.

For instance, the probability that the sum of the three dice equals\ $10$ is
$$
  \Pr(X_1 + X_2 + X_3 = 10) 
  = \sum_{\substack{(x_1, x_2, x_3) \in \mathcal{X}\\ x_1 + x_2 + x_3 = 10}} p(x_1,x_2,x_3).
$$

To obtain the answer in **R**, use:

```{r}
# List all possible outcomes of rolling three dice
dice_Outcomes <- expand.grid(x1 = 1:6, 
                             x2 = 1:6, 
                             x3 = 1:6)

# Show the first few rows (outcomes)
print(head(dice_Outcomes))

# Find where the rows sum to 10
favourable_Outcome <- rowSums(dice_Outcomes) == 10

# Probability
prob <- sum(favourable_Outcome) / nrow(dice_Outcomes)
prob
```
:::


For continuous random vectors, the definition is similar.


:::{.definition #JointPDF name="Joint probability function (continuous)"}
Let $\mathbf{X} = (X_1, \dots, X_n)$ be an $n$-dimensional continuous random vector.
For each $j = 1, \dots, n$, the random variable\ $X_j$ takes values in a subset $S_j \subseteq \mathbb{R}$.  

The range of $\mathbf{X}$ is\index{Range!for random vectors}
$$
  \mathcal{X} \subseteq S_1 \times \cdots \times S_n \subseteq \mathbb{R}^n.
$$
The *joint probability density function* of $\mathbf{X}$ is a function
$$
  f_{X_1, \dots, X_n}(x_1, \dots, x_n), \quad \text{for $(x_1, \dots, x_n) \in \mathcal{X}$},
$$
such that, for every region $A\subseteq \mathcal{X}$,
$$
  \Pr(\mathbf{X}\in A) = \int_A f_{X_1, X_2, \dots, X_n}(x_1, x_2, \dots, x_n)\, dx_1\, dx_2\dots dx_n.
$$
The density satisfies
$$
  f_{X_1, \dots, X_n}(x_1, \dots, x_n) \geq 0 
  \quad \text{for all $(x_1, \dots, x_n) \in \mathcal{X}$},
$$
and
$$
  \int_{\mathcal{X}} f_{X_1, \dots, X_n}(x_1, \dots, x_n)\, dx_1 \cdots dx_n = 1.
$$
The *support* of\ $\mathbf{X}$ is\index{Support!for random vectors}
$$
  \{\mathbf{x}\in\mathcal{X}: f_\mathbf{X}(\mathbf{x}) > 0\}.
$$
:::


:::{.example #UniformCube3D name="Three-dimensional uniform example (continuous)"}
Consider the random vector $\mathbf{X} = (X_1, X_2, X_3)$, uniformly distributed on the unit cube $[0, 1]^3$.  
The range is
$$
  \mathcal{X} = [0, 1] \times [0, 1] \times [0, 1] \subset \mathbb{R}^3.
$$

The joint probability density function is:\index{Joint probability density function}
$$
  f_{X_1, X_2, X_3}(x_1,x_2,x_3) =
  \begin{cases} 
    1 & \text{for $0 \le x_j \le 1$ for all $j = 1, 2, 3$},\\
    0 & \text{otherwise}.
  \end{cases}
$$

The marginal distributions are all uniform on $[0, 1]$; for example:
$$
  f_{X_1}(x_1) = \int_0^1 \int_0^1 f(x_1, x_2, x_3)\, dx_2\, dx_3 = 1, \quad 0\le x_1 \le 1.
$$

For instance, the probability that the sum of the three variables is less than or equal to\ $1$ is
$$
  \Pr(X_1 + X_2 + X_3 \le 1) 
  = \text{volume of tetrahedron } \{(x_1,x_2,x_3)\in\mathcal{X}: x_1 + x_2 + x_3 \le 1\} 
  = \frac{1}{6}.
$$

The answer can be obtained using integration, knowledge of volumes, or using a simulation in **R**:

```{r}
set.seed(30991) # For reproducibility

# Generate 1 000 000 random points in each dimension of the cube
N <- 1e6
X <- matrix(runif(3 * N), 
            ncol = 3)  # N rows of (x1,x2,x3)
print(head(X))

# How many of these random points satisfy: sum less than one?
prob_Est <- mean(rowSums(X) <= 1)
prob_Est
```
:::


## Joint distribution functions {#MultivariateDistFn}

The *multivariate distribution function (CDF)* is\index{Joint distribution function}
$$
  F_{X_1, \dots, X_n}(x_1, \dots, x_n) 
  = \Pr(X_1 \leq x_1, \dots, X_n \leq x_n).
$$


:::{.example #Dice3D_CDF name="Three dice"}
Let $\mathbf{X} = (X_1, X_2, X_3)$ be the random vector representing the outcome of rolling three fair six-sided dice, as in Example\ \@ref(exm:Dice3D) (where the joint probability mass function is given).

The *joint cumulative distribution function* is
$$
  F_{X_1, X_2, X_3}(x_1, x_2, x_3) 
  = \Pr(X_1 \le x_1, X_2 \le x_2, X_3 \le x_3)
  = \sum_{i = 1}^{\lfloor x_1 \rfloor} 
    \sum_{j = 1}^{\lfloor x_2 \rfloor} \sum_{k = 1}^{\lfloor x_3 \rfloor} p(i, j, k),
$$
where $\lfloor x \rfloor$ is the *floor* function (the greatest integer less than or equal to\ $x$).

For instance, the probability that all dice shows less than or equal to\ $3$ is
$$
  F_{X_1, \dots, X_n}(3, 3, 3) 
  = \sum_{i = 1}^{3} \sum_{j = 1}^{3} \sum_{k = 1}^{3} \frac{1}{216} 
  = \frac{27}{216} = \frac{1}{8}.
$$
:::


:::{.example #UniformCube3D_CDF name="Three-dimensional uniform cube: CDF"}
Let $\mathbf{X} = (X_1, X_2, X_3)$ be uniformly distributed on the unit cube $[0, 1]^3$, as in Example\ \@ref(exm:UniformCube3D) (where the joint probability density function is given)

The *joint cumulative distribution function* is
\begin{align*}
  F_{X_1, \dots, X_n}(x_1, x_2, x_3) 
  &= \Pr(X_1 \le x_1, X_2 \le x_2, X_3 \le x_3)\\
  &= \int_0^{x_1}\!\! \int_0^{x_2}\!\! \int_0^{x_3} f(u_1, u_2, u_3)\, du_3\, du_2\, du_1,
\end{align*}
for $0 \le x_1, x_2, x_3 \le 1$.

For instance, the probability that each variable is less than\ $0.5$ is
$$
  Fvs(0.5, 0.5, 0.5) = \int_0^{0.5}\!\! \int_0^{0.5}\!\! \int_0^{0.5} 1 \, du_3\, du_2\, du_1 = 0.5^3 = 0.125.
$$
:::


## Marginal and conditional distributions

Marginal distributions are obtained by summing or integrating out unwanted variables. For example,
$$
  f_{X_1}(x_1) = \int_{\mathbb{R}^{n-1}} f(x_1, x_2, \dots, x_n)\, dx_2 \cdots dx_n.
$$

Conditional distributions are defined in the natural way:
$$
  f_{X_1 \mid X_2, \dots, X_n}(x_1 \mid x_2, \dots, x_n) =
  \frac{f(x_1, x_2, \dots, x_n)}{f_{X_2,\dots,X_n}(x_2, \dots, x_n)}.
$$

EXAMPLES


## Multivariate independence {#MultivariateIndependence}

In the multivariate case, $n$\ random variables $X_1, \dots, X_n$ are *independent* if knowing the value of any subset of them gives no information about the others.
Formally, let $\mathbf{X} = (X_1, \dots, X_n)$ be an $n$-dimensional random vector with joint probability distribution. 

In the *discrete case*, the random variables are independent if and only if the joint probability mass function factors as the product of the marginal probability functions:
$$
  p_{X_1, \dots, X_n}(x_1, \dots, x_n) 
  = \prod_{i = 1}^{n} p_{X_i}(x_i),
  \quad (x_1, \dots, x_n) \in \mathcal{X}.
$$

The *continuous case* is similar; the random variables are independent if and only if the joint probability density function factors as the product of the marginal densities:
$$
  f_{X_1, \dots, X_n}(x_1, \dots, x_n) 
  = \prod_{i = 1}^{n} f_{X_i}(x_i),
  \quad (x_1, \dots, x_n) \in \mathcal{X}.
$$

Independence can also be determined using the distribution functions:
$$
  F_{X_1, \dots, X_n}(x_1, \dots, x_n) = \prod_{i=1}^{n} F_{X_i}(x_i),
$$
where $F$ is the joint cumulative distribution function.  

In practice, independence allows joint probabilities or volumes to be computed by multiplying the corresponding marginal probabilities or integrating products of marginal densities.  


:::{.example #Dice3D_Indep name="Three dice example: independence"}
Consider the outcomes of rolling three fair six-sided dice, represented by the random vector $\mathbf{X} = (X_1, X_2, X_3)$.  
Each $X_j \in \{1,2,3,4,5,6\}$, so $\mathcal{X} = \{1,\dots,6\}^3$.
The dice are independent, so the joint probability mass function can be written as the product of the marginal probability functions:
$$
  p_{X_1, X_2, X_3}(x_1, x_2, x_3) 
  = \Pr(X_1 = x_1, X_2 = x_2, X_3 = x_3) 
  = \prod_{i = 1}^3 \Pr(X_i = x_i) 
  = \frac{1}{6}\cdot   \frac{1}{6} \cdot \frac{1}{6} = \frac{1}{216}.
$$
Equivalently, the joint distribution factors as
$$
  F_{X_1, X_2, X_3}(x_1, x_2, x_3) = \Pr(X_1\le x_1, X_2 \le x_2, X_3 \le x_3) 
  = \prod_{i = 1}^3 F_{X_i}(x_i).
$$
The probability that all dice show a value less than or equal to\ $3$ is
$$
  F_{X_1, \dots, X_n}(3, 3, 3) 
  = \prod_{i = 1}^3 F_{X_i}(3) 
  = \left(\frac{3}{6}\right)^3 
  = \frac{27}{216} = \frac{1}{8}.
$$
:::


:::{.example #UniformCube3D_Indep name="Three-dimensional uniform cube: independence"}
Let $\mathbf{X} = (X_1, X_2, X_3)$ be uniformly distributed on the unit cube $[0, 1]^3$, with joint probability density function
$$
  f_{X_1, X_2, X_3}(x_1, x_2, x_3) = 1, \quad (x_1, x_2, x_3) \in [0, 1]^3.
$$

The variables $X_1, X_2, X_3$ are independent, so the joint density function can be written as
$$
  f(x_1, x_2, x_3) 
  = f_{X_1}(x_1)\cdot f_{X_2}(x_2)\cdot f_{X_3}(x_3) 
  = 1 \cdot 1 \cdot 1.
$$

Equivalently, the joint distribution can be written as
$$
  F_{X_1, X_2, X_3}(x_1, x_2, x_3) 
  = \Pr(X_1 \le x_1, X_2 \le x_2, X_3 \le x_3) 
  = \prod_{i = 1}^3 F_{X_i}(x_i).
$$

For example, the probability that the values of all three variables are less than\ $0.5$ is
$$
  F_{X_1, X_2, X_3}(0.5, 0.5, 0.5) = 0.5^3 = 0.125.
$$
:::


## Multinomial distribution {#MultinomialDistribution}

\index{Multinomial distribution|(}
A specific example of a *discrete* multivariate distribution is the *multinomial distribution*, a generalization of the binomial distribution. 


:::{.definition #MultinomialDistribution name="Multinomial distribution"}
Consider an experiment with the the sample space partitioned as $S = \{B_1, B_2, \ldots, B_k\}$. 
Let $p_i = \Pr(B_i), \ i = 1, 2,\ldots k$ where $\sum_{i = 1}^k p_i = 1$. 
Suppose there are $n$\ repetitions of the experiment in which\ $p_i$ is constant. 
Let the random variable\ $X_i$ be the number of times (in the $n$\ repetitions) that the event\ $B_i$ occurs. 
In this situation, the random vector $(X_1, X_2, \dots, X_k)$ is said to have a *multinomial distribution* with probability function
\begin{equation}
   \Pr(X_1 = x_1, X_2 = x_2, \ldots, X_k = x_k)
   = \frac{n!}{x_1! \, x_2! \ldots x_k!} 
      p_1^{x_1}\, p_2^{x_2} \ldots p_k^{x_k},
   (\#eq:Multinomial)
\end{equation}
where $\mathcal{R}_X = \{(x_1, \ldots x_k) : x_i = 0,1,\ldots,n, \, i = 1, 2, \ldots k, \, \sum_{i = 1}^k x_i = n\}$.
:::


The part of Eq.\ \@ref(eq:Multinomial) involving factorials arises as the number of ways of arranging $n$\ objects, $x_1$\ of which are of the first kind, $x_2$\ of which are of the second kind, etc. 
The above distribution is $(k - 1)$-variate since $x_k = n-\sum_{i = 1}^{k - 1}x_i$.
In particular if $k = 2$, the multinomial distribution reduces to the binomial distribution, which is a univariate distribution.

$X_i$\ is the number of times (out of\ $n$) that the event\ $B_i$, which has probability\ $p_i$, occurs. 
So the random variable\ $X_i$ clearly has a binomial distribution with parameters\ $n$ and\ $p_i$.
Thus, the marginal probability distribution of one of the components of a multinomial distribution is a binomial distribution.

Notice that the distribution in Example\ \@ref(exm:BVDiscreteTwoDice) is an example of a *trinomial distribution*.\index{Trinomial distribution} 
The probabilities shown in Table\ \@ref(tab:BVTwoDice) can be expressed algebraically as
$$
   \Pr(X = x, Y = y)
   = \frac{2!}{x! \, y!\, (2 - x - y)!}
      \left(\frac{1}{6}\right)^x\left(\frac{1}{6}\right)^y\left(\frac{2}{3}\right)^ {2 - x - y}
$$
for $x, y = 0 , 1 , 2$; $x + y \leq 2$.

The following are the basic properties of the multinomial distribution.


:::{.theorem #MultinomialProperties name="Multinomial distribution properties"}
Suppose $(X_1, X_2, \ldots, X_k)$ has the multinomial distribution given in Def.\ \@ref(def:MultinomialDistribution). 
Then for $i = 1, 2, \ldots, k$:

* $\operatorname{E}[X_i] = np_i$.
* $\operatorname{var}[X_i] = n p_i(1 - p_i)$.
* $\operatorname{Cov}(X_i, X_j) = -n p_i p_j$ for $i \neq j$.

::::{.proof}
The first two results follow because $X_i \sim \text{Bin}(n, p_i)$.

We will use\ $x$ for\ $x_1$ and\ $y$ for\ $x_2$ in the third for convenience.
Consider only the case $k = 3$, and note that
$$
   \sum_{(x, y) \in R} 
   \frac{n!}{x! \, y! (n - x - y)!} p_1^x\, p_2^y\, (1 - p_1 - p_2)^{n - x - y} = 1.
$$
Then, putting $p_3 = 1 - p_1 - p_2$,
\begin{align*}
  \operatorname{E}(XY)
  &= \sum_{(x, y)} xy \Pr(X = x, Y = y)\\
  &= \sum_{(x, y)}\frac{n!}{(x - 1)!(y - 1)!(n - x - y)!} p_1^x\,  p_2^y\, p_3^{n - x - y}\\
  &= n(n - 1) p_1 p_2\underbrace{\sum_{(x,y)}\frac{(n - 2)!}{(x - 1)!(y - 1)!(n - x - y)!} p_1^{x - 1}\, p_2^{y - 1}\, p_3^{n - x - y}}_{ = 1}.
\end{align*}
So $\operatorname{Cov}(X, Y) = n^2 p_1 p_2 - n p_1 p_2 - (n p_1)(n p_2) = -n p_1 p_2$.
::::
:::


::: {.softwareBox .software data-latex="{iconmonstr-laptop-4-240.png}"}
The two **R** functions for working with the multinomial distribution functions have the form `[dr]multinom(size, prob)` where `size`${}= n$ and `prob`${} = (p_1, p_2, \dots, p_k)$ for $k$ categories (see App.\ \@ref(UseRDistributions)):

* `dmultinom(x, size, sprob)` computes the probability function at $X = {}$`x`;
* `rmultinom(n, size, prob)` generates `n` random numbers.

The functions `qmultinom()` and `pmultinom()` are not defined.
:::


:::{.example #Multinomial name="Multinomial distribution"}
Suppose the four basic blood groups\ O, A, B and\ AB are known to occur in the proportions $9:8:2:1$. 
Given a random sample of $8$\ individuals, what is the probability that there will be $3$\ each of Types\ O and\ A and $1$\ each of Types\ B and\ AB?

The probabilities are $p_1 = 0.45$, $p_2 = 0.4$, $p_3 = 0.1$, $p_4 = 0.05$, and
\begin{align*}
   \Pr(X_O = 3, X_A = 3, X_B = 1, X_{AB} = 1)
   &= \frac{8!}{3!\,3!\,1!\,1!}(0.45)^3 (0.4)^3 (0.1)(.05)\\
   &= 0.033.
\end{align*}
In **R**:

```{r echo=TRUE}
dmultinom(x = c(3, 3, 1, 1), 
          size = 8, 
          prob = c(0.45, 0.4, 0.1, 0.05))
```
:::
\index{Multinomial distribution|)}

 






## Exercises {#MultivariateExercises}

Selected answers appear in Sect.\ \@ref(AnswersChapMultivariate).



::::{.exerciseList}
:::{.exercise #MultivariateNormal}
Let $X_1, X_2\overset{\text{iid}}{\sim} N(0,1)$. 
Consider the polar coordinates as a bivariate transformation:
$$
  X_1 = R \cos(\Theta) \quad \text{and} \quad X_2 = R \sin(\Theta).
$$
1. Calculate the Jacobian of the transformation.
2. Find the joint PDF in terms of\ $r$ and\ $\theta$.
3. Show that the resulting PDF can be factored into a function of $r$ and a function of $\theta$.
   What does that tell us about the independence of the radius and the angle?
3. Hence show that the bivariate normal distribution is a valid joint PDF.
:::
::::
