# (PART) Sampling distributions {-}

# Describing samples {#ChapDescribingSamples}


```{r setup, include=FALSE}
source("R/SetColours.R")
```


::: {.objectivesBox .objectives data-latex="{iconmonstr-target-4-240.png}"}

* use appropriate techniques to determine the sampling distributions of\ $t$, $F$, and\ $\chi^2$\ distributions.
* explain how the above distributions are related to the normal distribution.
* apply the concepts of the Central Limit Theorem in appropriate circumstances.
* use the Central Limit Theorem to approximate binomial probabilities by normal probabilities in appropriate circumstances.
:::

## Populations and samples {#PopulationsSamples}

This chapter introduces the application of distribution theory to the discipline of statistics.
Distribution theory is about describing random variables in theoretical populations using probability.
Statistics is about collecting data and extracting information from subsets of the population (i.e., samples) to learn about populations.
This chapter explains how to describe those subsets.


### From populations to observations {#Sampling}

\index{Sample|(}
Until now, theoretical probability distributions describing ideal distributions of infinite populations have been our focus.
In practice, of course, ideal distributions and infinite populations do not exist, and finite (usually relatively small) samples from unknown distributions are observed.\index{Sample}


:::{.example #Heights name="Populations and models"}
Suppose our *population* of interest is Australian adult females, and we are interested in their heights.
'Height of Australian adult females' is a continuous random variable, say\ $X$.

We may decide to model this population of heights theoretically using some specified distribution, with the probability density function $f_X(x)$.
Using this model, the mean height of Australian adult females is modelled as (Def.\ 
\@ref(def:Expectation))
$$
  \mu = \operatorname{E}[X] = \int_{-\infty}^\infty x\, f_X(x)\,dx.
$$

If we model their heights using the normal distribution $N(\mu = 173, \sigma^2 = 9)$, the mean is modelled as\ $173\cms$ and the variance as $9\cms^2$.
:::


While height is a continuous random variable, and heights can be modelled using a continuous probability distribution, in practice height measurements of individual Australian females are *not* made with infinite precision; measured heights must be rounded and hence are recorded as discrete values.
Any tool to measure actual heights must record a *rounded* version of the heights, to some minimum unit\ $\delta$.
Heights, for example, may be rounded to the nearest centimetre (i.e., $\delta = 1\cms$) or the nearest millimetre (i.e., $\delta = 0.1\cms$).
So even if it was possible to measure the height of every Australian adult female, we would in practice have a list of discrete values $\{x_1, x_2, \dots , x_N\}$, where\ $N$ is the size of the population.

So, the observed heights are actually discrete (rounded) values, and the population mean of the measured heights would be computed (from Def.\ \@ref(def:Expectation)) using
$$
  \mu 
  = \operatorname{E}[X]
  = \frac{1}{N} \sum_{i = 1}^N x_i.
$$


### Random sampling and statistics {#RandomSamples}

In practice, entire populations are very rarely studied: for example, the population is usually too large (and so studying the whole population is prohibitively expensive and time consuming), and some members of population may be inaccessible.
So rather than studying entire populations, a *sample* of observations from that population is almost always studied.

The task of the *inferential statistics*\index{Inferential statistics}\index{Statistic} is to learn about the entire population, based on what is learnt from studying a small subset of that population: the *sample*. \index{Sample}
Inferential statistics requires an understanding of how samples behave.


:::{.definition #Sample name="Sample"}
A subset of observations from the population is called a sample.\index{Sample}
:::


This definition is very broad: *any* subset from the population is a *sample*.
Importantly, *how* the sample is drawn determines its statistical properties; the most important type of sample for statistical inference is the idea of a *random sample*.\index{Sample!random}


:::{.definition #RandomSample name="Random sample"}
The random variables $X_1, X_2, \dots, X_n$ are a *random sample* from distribution\ $D$ if and only if they are independently and identically distributed (iid)\index{iid} with distribution function $F_X(x)$.
We write $X_1, X_2, \dots, X_n \overset{\text{iid}}{\sim} D$..
:::


The definition uses the distribution function and so applies to discrete, continuous and mixed random variables.
Because the elements of the random sample are assumed to be independent,\index{Independence} results from previous chapters concerning independent random variables therefore apply directly.

Once the sample is obtained, the sample data can be used to compute the value of some *statistic*.
A *statistic*\index{Statistic} is a function of the sample values.


:::{.definition #Statistic name="Statistic"}
A *statistic* is any function $T = g(X_1, X_2, \ldots, X_n)$ that is computable from the sample alone; that is, it does not depend on any unknown population parameters.\index{Statistic}
:::


Recall that a *parameter*\index{Parameter} is a numerical characteristic of a population or its distribution (Def.\ \@ref(def:Parameter)). 
In statistical inference, parameters are typically unknown, and it is the goal of estimation to infer their values from sample data.


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
Since the individuals chosen to be in the sample can vary from sample to sample, the *value* of a statistic can vary from sample to sample.
:::


:::{.example #StatisticsExample name="Statistic"}
For a random sample $X_1, X_2, \dots, X_n$, the statistic\ $R$ can be defined such that
$$
  R = \max(X_1, X_2, \dots, X_n) - \min(X_1, X_2, \dots, X_n).
$$
:::


A special type of statistic is an *estimator*.


:::{.definition #Estimator name="Estimator"}
An estimator $\hat{\theta} = g(X_1, X_2, \dots, X_n)$ is a statistic used to estimate an unknown population parameter\ $\theta$.
As a function of random variables, an estimator is itself a random variable.\index{Estimator}
:::


:::{.definition #Estimate name="Estimate"}
An estimate $\hat{\theta}(x_1, x_2, \dots, x_n)$ is the numerical value taken by an estimator upon observing a realisation $x_1, x_2, \dots, x_n$ of the sample.\index{Estimate} 
:::


The study of estimators and their properties---including point estimation, interval estimation, and hypothesis testing---is called *statistical inference*.\index{Statistical inference}


:::{.example #StatisticsExampleMin name="The sample median"}
Suppose we take a sample of $n = 12$ Australian adult females, and measure their heights to the nearest centimetre.
We could define the *statistic*\index{Statistic}
$$
  M = \text{median}(X_1, X_2, \dots, X_{12}).
$$
$M$ is a *statistic*, as it only depends on the sample data.
Since\ $M$ is likely to be different when computed from another sample of $12$\ Australian adult females, $M$ is a random variable. 

If the purpose of computing the sample median is to estimate the unknown median value in the population, then\ $M$ is an *estimator* of the population median.
While\ $M$ has been proposed as an estimator of the population median, we have not studied whether it is a 'good' estimator of the population median (i.e., has desirable properties).

If the median height in the sample is $169\cms$, then
$$
  m = 169
$$
is an *estimate* of the unknown population median.

Notice that a capital\ $M$ is used for the estimator, and a lowercase\ $m$ for the realised value.
:::
\index{Sample|)}


## Sampling distributions {#SamplingDistributions}

### Sampling distributions: a small example {#SamplingDistributionsSmallExample}

\index{Sampling distribution|(}
Since a statistic is a function of the random variables $X_1, X_2, \ldots, X_n$, *statistics are random variables*, and so the value of a statistic changes from sample to sample. 
This means that *statistics have a distribution* that describes *how* they vary.


:::{.definition #SamplingDistributionStatistic name="Sampling distribution of a statistic"}
The *sampling distribution of a statistic* is the theoretical probability distribution of the values taken by the statistic across all possible samples of a particular size drawn from a particular population.
:::


Typically, a population will be *very* large, and listing all elements of the population is impossible.
However, for demonstration purposes, suppose the population contains only $N = 5$ discrete elements:
$$
   22\qquad 23\qquad 27\qquad 28\qquad 38.
$$
The median value in the population (a *parameter*)\index{Parameter} is\ $27$.
The sample median is an *estimator*\index{Estimator} of this parameter.

For each possible sample of size $n < 5$ (chosen without replacement), we can compute the sample median.
Since the population is (artificially) small in this scenario, the exact sampling distribution can be determined by enumerating every possible sample of any given size\ $n$.

Using **R**:

```{r}
Population <- c(22, 23, 27, 28, 38)      # ALL values in the population
median(Population)                       # Find the median of the population
```

Suppose we consider samples of size $n = 2$; there are $\binom{5}{2} = 10$ such samples:

```{r}
combn(Population, 2)                     # List all samples
```

The median of each sample is easily computed also:

```{r}
median_2 <- combn(Population, 2, FUN = median)  # Medians of every sample
median_2
```

The *sampling distribution* is the distribution of these sample medians, which can be displayed a table (or graph):

```{r}
table(median_2)                          # Tabulate the sample medians
```

Since the population median is\ $27$, the sample medians are often a poor estimate of the population median (they range from\ $22.5$ to\ $33$).
Suppose we looked at larger sample size; say, samples of size $n = 4$:

```{r}
median_S_4 <- combn(Population, 4, FUN = median)
table(median_S_4)
```

So the sample median when $n = 4$ is often a very good estimate of the population median (they range from\ $25$ to\ $27.5$).
The larger sample size produces a better estimate, in general, and the value of the statistic varies less from sample to sample.


### Larger populations {#SamplingDistributionsLargerPopulations}

For larger populations, enumerating every possible is sample implausible or impossible, and obtaining the selected elements is usually impractical or impossible.
Nonetheless, the idea of a sampling distribution of a statistic can be *simulated* using\ **R**.
These sampling distributions, generated from simulated data, are called *empirical distributions*;\index{Empirical distributions} that is, they are based on sample data not theoretical ideals.


:::{.example #SampleEstimating name="Emoirical sampling distributions"}
Consider taking a single random sample (with replacement) of $n = 5$ observations from a $N(\mu = 173, \sigma^2 = 9)$ distribution, and computing the median value in the sample:

```{r echo=FALSE}
set.seed(1592875)
```
```{r}
x_Mn <- 173; x_Sd <- 3        # Population (model) mean and std dev.
sample_Size <- 5              # Sample size is 5
x1 <- rnorm(sample_Size,      # Take a sample of this size...
            mean = x_Mn, sd = x_Sd)     # ... from this distribution
median(x1)                    # Compute the sample median for this sample
```

This process can be repeated numerous times in **R** to simulate selecting many samples of size $n=5$, and computing the median for each (Fig.\ \@ref(fig:SDmedianPlots), left panel):

```{r SDmedian1, fig.show='hide'}
num_Sims <- 100000           # Number of simulations

# Generate random values, and place in a suitably-sized array
Xsample_5 <- array(
  rnorm(num_Sims * sample_Size, mean = x_Mn, sd = x_Sd),
  dim = c(sample_Size, num_Sims)
)

Xmedian_5 <- apply(Xsample_5,     # For this data...
                   MARGIN = 2,    # ... and for each column...
                   FUN = median)  # ... compute the median
hist(Xmedian_5)                   # Plot the histogram of sample medians
```

Suppose we take larger samples, and find the sampling distribution again (Fig.\ \@ref(fig:SDmedianPlots), right panel):

```{r SDmin2, fig.show='hide'}
sample_Size2 <- 100

# Generate random values, and place in a suitably-sized array
Xsample_100 <- array(
  rnorm(num_Sims * sample_Size2, mean = x_Mn, sd = x_Sd),
  dim = c(sample_Size2, num_Sims) 
)

# Find the minimum of each sample (i.e., column)
Xmedian_100 <- apply(Xsample_100,   # For this data...
                  MARGIN = 2,       # ... and for each column...
                  FUN = median)     # ... compute the median
hist(Xmedian_100)                   # Plot the histogram of sample medians
```
:::

```{r SDmedianPlots, echo=FALSE, out.width='90%', fig.width=8, fig.height = 3.25, fig.align="center", fig.cap="Sampling distributions for the sample medians, for samples of size\ $5$ (left) and size\ $100$ (right)."}
par( mfrow = c(1, 2))

set.seed(1592875)

hist(Xmedian_5, 
     xlim = c(165, 185),
     xlab = "Sample median",
     ylab = "",
     breaks = seq(165, 185, by = 1),
     main = "Sampling medians:\nsamples of size 5",
     axes = FALSE)
axis(side = 1)

#####

hist(Xmedian_100, 
     xlim = c(165, 185),
     xlab = "Sample median",
     ylab = "",
     breaks = seq(165, 185, by = 0.5),
     main = "Sampling medians:\nsamples of size 100",
     axes = FALSE)
axis(side = 1)
```

These histograms display the sampling distributions of the sample median for samples of size $n = 5$ and $n = 100$.
Again, the value of the statistic varies less from sample to sample for the larger sample size.


## Estimating population parameters

### Estimating the population mean {#SampligDistributionMean}

One of the most common reasons for taking a sample is to estimate the population mean\ $\mu$.
For a population of\ $N$ values $X_1, X_2, \dots, X_N$, the population mean is $\mu = \operatorname{E}[X] = \frac{1}{N}\sum_i X_i$, as noted earlier.
The natural estimator of\ $\mu$ is the *sample mean* $\overline{X}$, an *estimator*,\index{Estimator} defined as
$$
  \overline{X} = \frac{1}{n}\sum_{i = 1}^n X_i.
$$

We can use\ **R** to simulate the selection of many samples of size $n = 5$, and many sample of size $n = 100$, and compute the empirical mean\index{Empirical mean} for each sample, using the same samples created in Sect.\ \@ref(SamplingDistributionsLargerPopulations) (Fig.\ \@ref(fig:SDmeanPlots)):

```{r SDmean, fig.show='hide'}
# Mean: n = 5
Xmean   <- apply(Xsample_5, 
                 MARGIN = 2, 
                 FUN = mean) 
hist(Xmean)          # Create the histogram of sample minimums

# Mean: n = 100
Xmean_2   <- apply(Xsample_100, 
                   MARGIN = 2, 
                   FUN = mean) 
hist(Xmean_2)          # Create the histogram of sample minimums
```


```{r SDmeanPlots, echo=FALSE, out.width='90%', fig.width=8.5, fig.height = 3.5, fig.align="center", fig.cap="Sampling distributions for the sample mean for sample of size $n = 5$ (left) and $n = 100$ (right). The solid dots represent the population mean."}
par( mfrow = c(1, 2))

hist(Xmean, 
     xlim = c(165, 180),
     xlab = expression(Sample~mean~bar(italic(x))),
     ylab = "",
     breaks = seq(150, 180, by = 1),
     main = "Sampling mean:\nsamples of size 5",
     sub = paste0("(Average sample estimate: ", round(mean(Xmean), 1), ")"),
     axes = FALSE)
axis(side = 1)
points(x = x_Mn, 
       y = 0,
       pch = 19)

#####

hist(Xmean_2, 
     xlim = c(165, 180),
     xlab = expression(Sample~mean~bar(italic(x))),
     ylab = "",
     breaks = seq(150, 180, by = 0.5),
     main = "Sampling mean:\nsamples of size 100",
     sub = paste0("(Average sample estimate: ", round(mean(Xmean_2), 1), ")"),
     axes = FALSE)
axis(side = 1)
points(x = x_Mn, 
       y = 0,
       pch = 19)


```


The sample means vary around the true population mean\ $\mu$ (some are smaller; some are larger), but are centred on\ $\mu$, suggesting that\ $\overline{X}$ is an unbiased estimator. 
Larger samples produce estimates that are more tightly concentrated around\ $\mu$, reflecting the smaller variance of\ $\overline{X}$ for larger values of\ $n$.


### Unbiased estimators {#UnbiasedEstimators}

The sample means are centred on\ $\mu$ (some estimates fall below the true value, some above), but on average the estimator is equal to the parameter it estimates. 
An estimator with this property is called *unbiased*.\index{Estimator!unbiased}


:::{.definition #Unbiased name="Unbiased estimator"}
A statistic\ $\hat{\theta}$ is an *unbiased estimator*\index{Estimator!unbiased} of a parameter\ $\theta$ if
$$
  \operatorname{E}[\hat{\theta}] = \theta.
$$
:::


When we say that\ $\overline{X}$ is an unbiased estimator of\ $\mu$, we mean that $\operatorname{E}[\overline{X}] = \mu$,


:::{.theorem #SampleMeanUnbiased name="An unbiased estimator for $\mu$"}
The sample mean
$$
  \overline{X} = \frac{1}{n} \sum_{i = 1}^n X_i
$$
is an unbiased estimator\index{Estimator!unbiased}\index{Estimator!sample mean} of the population mean\ $\mu$.

::::{.proof}
Using the definition of an unbiased estimator and the properties of expectations:
\begin{align*}
  \operatorname{E}[\overline{X}] 
  &= \operatorname{E}\left[ \frac{1}{n}(X_1 + X_2 + \cdots X_n) \right]\\
  &= \frac{1}{n} \operatorname{E}[ X_1 + X_2 + \cdots X_n ]\\
  &= \frac{1}{n} \left( \operatorname{E}[ X_1 ] + \operatorname{E}[X_2] + \cdots \operatorname{E}[X_n] \right) \\
  &= \frac{1}{n} (n\times \mu) = \mu.
\end{align*}
That is, $\operatorname{E}[\overline{X}] = \mu$, and so\ $\overline{X}$ is an *unbiased estimator* of\ $\mu$.
::::
:::


:::{.example #UnbiasedPoisson name="A unbiased estimate"}
Let $X_1, \ldots, X_n \overset{\text{iid}}{\sim} \operatorname{Poisson}(\lambda)$, and consider the estimate $\overline{X}$.
Because the expectation is a linear operator, and because $\operatorname{E}[X_i] = \lambda$, 
$$
  \operatorname{E}[\overline{X}] 
  = 
  \operatorname{E}\!\left[\frac{1}{n}\sum_{i=1}^n X_i\right] = \frac{1}{n}\sum_{i=1}^n \operatorname{E}[X_i] = \frac{1}{n} \cdot n\lambda = \lambda.
$$
Hence $\overline{X}$ is an unbiased estimator of\ $\lambda$ for a Poisson distribution.
:::

Not all estimators\index{Estimator} are unbiased estimators; even estimators that appear to be natural estimators may not be unbiased estimators.\index{Estimator!unbiased}


:::{.example #BiasedEstimate name="A biased estimate"}
Let $X_1, \ldots, X_n \overset{\text{iid}}{\sim} \operatorname{Exponential}(\lambda)$, where $\lambda > 0$ is the *rate* parameter, so that $\operatorname{E}[X_i] = 1/\lambda$ (Sect.\ \@ref(ExponentialProperties)).
Consider the natural estimator of $\lambda$: that is, $1/\overline{X}$.

Since $X_i \sim \operatorname{Exponential}(\lambda)$, then $\overline{X} = \frac{1}{n}\sum_{i=1}^n X_i$ (see Example\ \@ref(exm:GammaExp)).
Since
$$
  \overline{X} \sim \operatorname{Gamma}(\alpha = n,\, 1/(n\lambda) ),
$$
where we use the shape--rate parameterisation, so $\operatorname{E}[\overline{X}] = 1/\lambda$ and $\operatorname{Var}(\overline{X}) = 1/(n\lambda^2)$.

$\operatorname{E}[1/\overline{X}]$ can be computed directly:
$$
  \operatorname{E}\!\left[\frac{1}{\overline{X}}\right]
  = \int_0^\infty \frac{1}{x} \cdot
    \frac{(n\lambda)^n}{\Gamma(n)}\,x^{n-1} \exp(-n\lambda x)\,dx
  = \frac{(n\lambda)^n}{\Gamma(n)}
    \int_0^\infty x^{n-2} \exp(-n\lambda x)\,dx.
$$
The integral is related to the probability density function of a gamma distribution (Def.\ \@ref(def:GammaDistribution)) with shape $n - 1$ and rate $1/(n\lambda)$, and so
$$
  \int_0^\infty x^{n - 2} \exp(-n\lambda x)\,dx
  = 
  \frac{\Gamma(n - 1)}{(n\lambda)^{n - 1}},
$$
valid for $n > 1$. 
Substituting then,
$$
  \operatorname{E}\!\left[\frac{1}{\overline{X}}\right]
  = \frac{(n\lambda)^n}{\Gamma(n)} \cdot \frac{\Gamma(n - 1)}{(n\lambda)^{n - 1}}
  = n\lambda \cdot \frac{\Gamma(n - 1)}{\Gamma(n)}
  = n\lambda \cdot \frac{1}{n-1}
  = \frac{n\lambda}{n - 1},
$$
using $\Gamma(n) = (n - 1)\,\Gamma(n-1)$. 
Hence
$$
  \operatorname{E}\!\left[\frac{1}{\overline{X}}\right]
  = \frac{n}{n - 1}\,\lambda
  = \lambda + \frac{\lambda}{n - 1},
$$
and so the bias is $\lambda/(n - 1) > 0$: the estimator $1/\overline{X}$ systematically *overestimates* the value of $\lambda$.

Thus, $1/\overline{X}$ is a *biased* estimator of $\lambda$.
Note also that the bias $\lambda/(n - 1) \to 0$ as $n \to \infty$, so $1/\overline{X}$ is *asymptotically unbiased*.\index{Unbiased estimator!asymptotically}

A simulation (Fig.\ \@ref(fig:BiasExponential)) using $\lambda = 2$ shows that, in a sample of size $n = 3$, the mean value of $1/\overline{x}$ is $2.88$, a poor estimate of $\lambda  = 2$. 
The theoretical expected value is
$$
  \frac{n\lambda}{n - 1} = \frac{3\times 2}{2 - 1} = 3,
$$
which is close to the  value in the simulation.
:::


```{r BiasExponential, echo=FALSE, fig.align="center", fig.cap="The sampling distribution of $1/\\overline{X}$ shows that this is a biased estimator of\\ $\\lambda$.", out.width='80%', fig.width=7, fig.height = 3.75}
set.seed(139796)

sample_Size <- 3
num_Sims <- 2000
lambda <- 2

X <- array(
  rexp(sample_Size * num_Sims, rate = lambda),
  dim = c(sample_Size, num_Sims)
)

Xest <- apply(X,
              MARGIN = 2,
              FUN = function(x){ 1/mean(x)} )


h <- hist(Xest,
          plot = FALSE)
ylim_max <- max(h$density) * 1.2  # 10% space for the label

hist(Xest,
     col = ColourLight,
     axes = FALSE,
     ylim = c(0, ylim_max),
     freq = FALSE,
     main = expression(Estimates~of~lambda~using~1/bar(italic(X))),
     ylab = "",
     xlab = expression(1/bar(italic(x))))
axis(side = 1)
abline(v = lambda,
       lty = 2,
       lwd = 2)
text(x = lambda, 
     y = ylim_max * 0.95,
     pos = 4,
     expression(lambda))

mean(Xest)
lambda

```


### Estimating the population variance {#EstimatePopVariance}

\index{Variance!estimation}
In the same way that a population mean can be defined, the population variance for a population of\ $N$ values $X_1, X_2, \dots, X_N$ is
$$
  \operatorname{var}[X] = \sigma^2 = \frac{1}{N}\sum_{i = 1}^N (x_i - \mu)^2.
$$


:::{.theorem #SampleVarianceUnbiased name="An unbiased estimator for $\sigma^2$"}
The sample variance
\begin{equation}
  S^2_\mu = \frac{1}{n} \sum_{i=1}^n (X_i - \mu)^2
   (\#eq:S2WithPopMean)
\end{equation}
is an unbiased estimator\index{Estimator!unbiased} of the population variance\ $\sigma^2$.

::::{.proof}
See Exercise\ \@ref(exr:UnbiasedVarianceMu)
::::
:::


Using the samples found earlier (Sect.\ \@ref(SamplingDistributionsLargerPopulations)), the empirical estimates\ $S^2_\mu$\index{Empirical variances} can be found for samples of size $n = 5$ and $n = 100$ (Fig.\ \@ref(fig:SampleVarMu)), where $\mu = 173$:


```{r, fig.show='hide'}
s2mu <- function(x) { (sum(x - x_Mn)^2) / length(x) }  # Function to find sS2_mu
Xvar_5 <- apply(Xsample_5,                             # Apply fn to each col
                MARGIN = 2,
                FUN = s2mu)
hist(Xvar_5) 
mean(Xvar_5)

Xvar_100 <- apply(Xsample_100,
                  MARGIN = 2,
                  FUN = s2mu)
hist(Xvar_100) 
mean(Xvar_100)
```

```{r SampleVarMu, echo=FALSE, out.width='100%', fig.width=8, fig.height=3.5, fig.cap="Sampling distributions for the sample variance for sample of size $n = 5$ (left) and $n = 100$ (right), using the unbiased estimator $S^2_\\mu$. The solid dots represent the population variance."}
par( mfrow = c(1, 2))

hist(Xvar_5,
     xlab = expression(italic(x)),
     main = "Variance estimates\nn = 5",
     ylab = "",
     axes = FALSE) 
axis(side = 1)

points(x = x_Sd^2,
       y = 0,
       pch = 19)

hist(Xvar_100,
     xlab = expression(italic(x)),
     main = "Variance estimates\nn = 100",
     ylab = "",
     axes = FALSE) 
axis(side = 1)
points(x = x_Sd^2,
       y = 0,
       pch = 19)
```


The average value of the estimates, from the simulations, appear close to\ $\sigma^2 = 9$.
Again the larger sample size produces estimates with less variation compared to the estimates from the smaller sample size.

While $S^2_\mu$ is an unbiased estimator of\ $\sigma^2$, Eq.\ \@ref(eq:S2WithPopMean) has a practical difficulty: its computation requires the population mean\ $\mu$, which is unknown in practice.
This suggests using the estimator
$$
  S^2_n = \frac{1}{n} \sum_{i=1}^n (X_i - \overline{X})^2.
$$
to estimate the population variance.
To determine if this is an unbiased estimator, find $\operatorname{E}[S^2_n]$:
\begin{align*}
    \operatorname{E}[S^2_n] 
  &= \operatorname{E}\left[ \frac{1}{n} \sum_{i=1}^n (X_i - \overline{X})^2 \right]\\
  &= \frac{1}{n}\operatorname{E}\left[ \sum_{i=1}^n \left(X^2_i - 2X_i\overline{X} + \overline{X}^2\right)\right]\\
  &= \frac{1}{n}\operatorname{E}\left[ \sum_{i=1}^n X^2_i - \overline{X}\sum_{i=1}^n 2X_i + \sum_{i=1}^n \overline{X}^2\right].
\end{align*}
Since $\overline{X} = \sum_i X_i/n$, then $\sum_i X_i = n \overline{X}$; hence
\begin{align*}
     \operatorname{E}[S^2_n] 
  &= \frac{1}{n}\operatorname{E}\left[ \sum_{i=1}^n X^2_i - 2n\overline{X}^2 + n \overline{X}^2\right]\\
  &= \frac{1}{n}\operatorname{E}\left[ \sum_{i=1}^n X^2_i - n \overline{X}^2 \right]\\
  &= \frac{1}{n}\left(\operatorname{E}\left[ \sum_{i=1}^n X^2_i\right] - n\operatorname{E}\left[ \overline{X}^2 \right]\right).
\end{align*}
This expression can be simplified by noting that
$$
  \operatorname{var}[\overline{X}]
  = \operatorname{E}[\overline{X}^2] - \operatorname{E}[\overline{X}]^2 
  = \operatorname{E}[\overline{X}^2] - \mu^2, 
$$
and so $\operatorname{E}[\overline{X}^2] = \operatorname{var}[\overline{X}] + \mu^2 = \sigma^2/n + \mu^2$.
Also, 
$$
  \operatorname{var}[X] 
  = \operatorname{E}[X^2] - \operatorname{E}[X]^2 
  = \operatorname{E}[X^2] - \mu^2,
$$
so that $\operatorname{E}[X^2] = \sigma^2 + \mu^2$.
Hence,
\begin{align*}
     \operatorname{E}[S^2_n] 
  &= \frac{1}{n}\left[ \sum_{i=1}^n (\sigma^2 + \mu^2) - n\left(\frac{\sigma^2}{n} + \mu^2\right)\right]\\
  &= \frac{1}{n}\left[ n\sigma^2 + n\mu^2 - \sigma^2 - n\mu^2\right]\\
  &= \frac{n - 1}{n} \sigma^2.
\end{align*}
This shows that $\operatorname{E}[S^2_n] \ne\sigma^2$, and so the estimator is a *biased estimator* of\ $\sigma^2$.
Again, the estimator is asymptotically unbiased,\index{Unbiased estimator!asymptotically} as $\operatorname{E}[S^2_n] \to\sigma^2$ as $n\to\infty$.
Replacing the unknown\ $\mu$ with the estimate\ $\overline{X}$ introduces additional variation, causing $S^2_n$ to systematically underestimate the value of] $\sigma^2$.

Using the sample found earlier, the empirical estimates\ $S^2_n$\index{Empirical variances} can be found for samples of size $n = 5$ and $n = 100$ (Fig.\ \@ref(fig:SDvarPlots), top panels).
The values of $\operatorname{E}[S^2_n]$ for $n = 5$ and $n = 100$ are
$$
  \operatorname{E}[S^2_n] = \frac{5 - 1}{5}\times 3^2 = 7.2
  \quad\text{and}\quad
  \operatorname{E}[S^2_n] = \frac{100 - 1}{100}\times 3^2 = 8.91
$$
respectively, similar to those means computed from the **R** simulation.
These both suggest that the estimator is unbiased.


```{r SDvarPlots, echo=FALSE, out.width='90%', fig.width=8.5, fig.height = 5.5, fig.align="center", fig.cap="Sampling distributions for the sample variance for sample of size $n = 5$ (left panels) and $n = 100$ (right panels), using the biased estimator $S^2_n$ (top panels) and the unbiased estimator $S^2$ (bottom panels). The solid dots represent the population variance."}

s2n <- function(x) { (sum( (x - mean(x))^2) ) / length(x)}

Xvarn   <- apply(Xsample_5, 
                MARGIN = 2, 
                FUN = s2n) 
Xvarn_2   <- apply(Xsample_100, 
                   MARGIN = 2, 
                   FUN = s2n) 


Xvar   <- apply(Xsample_5, 
                MARGIN = 2, 
                FUN = var) 
Xvar_2   <- apply(Xsample_100, 
                  MARGIN = 2, 
                  FUN = var) 

xRange   <- c(0, 50)
xRange_2 <- c(0, 65)

#####

par( mfrow = c(2, 2))


hist(Xvarn, 
     xlim = xRange,
     xlab = "Sample variance",
     ylab = "",
     breaks = seq(min(xRange), max(xRange), by = 2),
     main = "Sample variance:\nsamples of size 5",
     sub = paste0("(Average value: ", round(mean(Xvarn), 3), ")"),
     axes = FALSE)
axis(side = 1)
points(x = x_Sd^2, 
       y = 0,
       pch = 19)



#####

hist(Xvarn_2,
     xlim = xRange,
     xlab = "Sample variance",
     ylab = "",
     breaks = seq(min(xRange), max(xRange), by = 2),
     main = "Sample variance:\nsamples of size 100",
     sub = paste0("(Average value: ", round(mean(Xvarn_2), 3), ")"),
     axes = FALSE)
axis(side = 1)
points(x = x_Sd^2, 
       y = 0,
       pch = 19)


####
####
####


hist(Xvar, 
     xlim = xRange_2,
     xlab = "Sample variance",
     ylab = "",
     breaks = seq(min(xRange_2), max(xRange_2), by = 2),
     main = "Sample variance:\nsamples of size 5",
     sub = paste0("(Average value: ", round(mean(Xvar), 3), ")"),
     axes = FALSE)
axis(side = 1)
points(x = x_Sd^2, 
       y = 0,
       pch = 19)



#####

hist(Xvar_2, 
     xlim = xRange_2,
     xlab = "Sample variance",
     ylab = "",
     breaks = seq(min(xRange_2), max(xRange_2), by = 2),
     main = "Sample variance:\nsamples of size 100",
     sub = paste0("(Average value: ", round(mean(Xvar_2), 3), ")"),
     axes = FALSE)
axis(side = 1)
points(x = x_Sd^2, 
       y = 0,
       pch = 19)

```



To obtain an *unbiased estimator*, we could multiply $S^2_n$ by $(n - 1)/n$ (a constant), which *does* produce an unbiased estimator.


:::{.theorem #SampleVarianceUnbiasedXbar name="An unbiased estimator for $\sigma^2$ ($\mu$ unknown)"}
When the value of\ $\mu$ is unknown, an *unbiased* estimator of\ $\sigma^2$ is 
\begin{equation}
  S^2 = \frac{1}{n - 1} \sum_{i=1}^n (X_i - \overline{X})^2.
   (\#eq:S2WithoutPopMean)
\end{equation}
Notice that the divisor in front of the summation is now\ $n - 1$ rather than\ $n$. 

::::{.proof}
Easily shown by adapting the proof for $S^2_n$ above; see Exercise\ \@ref(exr:UnbiasedVariance).
::::
:::


Using the sample found earlier, the empirical estimates\ $S^2$\index{Empirical variances} can be found for samples of size $n = 5$ and $n = 100$ (Fig.\ \@ref(fig:SDvarPlots), bottom panels).


A result that is useful later is the perhaps-surprising result that the sample mean and the sample variance are independent.

:::{.theorem #MeanVarInd name="Sample mean and variance: Independent"}
Let $X_1, X_2,\dots, X_n$ be a random sample of size\ $n$ from $N(\mu, \sigma^2)$.
Then the sample mean $\overline{X}$ and the sample variance\ $S^2$ are independent.

::::{.proof}
This proof is not given.
::::
:::

This result relies on normality; for non-normal distributions, $\overline{X}$ and $S^2$ are generally dependent.


## The distribution of the sample mean {#SamplingDistributionsMean}

### Introduction {#SamplingDistributionsMeanIntro}

In previous sections, unbiased estimators for\ $\mu$ and\ $\sigma^2$ were introduced and their properties studied.
All estimators are random variables, and so have distributions. 
Previously, we used simulations to display the empirical sampling distribution\index{Sampling distributions!empirical}\index{Empirical distributions} of various statistics.

In this section, theoretical sampling distributions are derived (beginning with the distribution of the sample mean\ $\overline{X}$) which form the foundation for statistical inference.


### The mean and variance of $\overline{X}$ {#SamplingDistributionMean}

Usually the sample mean\ $\overline{X}$ is of greatest interest when sampling.
Figure\ \@ref(fig:SDmeanPlots) suggests that larger sample sizes produce sampling distributions with smaller variances.
The following theorem  quantifies how the variance of\ $\overline{X}$ decreases with sample size.


:::{.theorem #SamplingDistMean name="Sampling distribution of the mean"}
If $X_1, X_2, \dots, X_n$ is a *random sample* of size\ $n$ from a population with mean\ $\operatorname{E}[X] = \mu$ and variance\ $\operatorname{var}[X] = \sigma^2$, then the sample mean\ $\overline{X}$ has a sampling distribution with mean $\mu$ and variance $\sigma^2 / n$.

::::{.proof}
To show that $\operatorname{E}[\overline{X}] = \mu$, apply Theorem\ \@ref(thm:SampleMeanUnbiased).
Similarly, using the properties of the variance, the variance of $\overline{X}$ is:
\begin{align*}
    \operatorname{var}[\overline{X}]
  &= \operatorname{var}[(X_1 + X_2 + \cdots + X_n)/n]\\
  &= \frac{1}{n^2} \operatorname{var}[X_1 + X_2 + \cdots X_n]\\
  &= \frac{1}{n^2} \times n\sigma^2\quad\text{(since the $X_i$ are iid)}\\
  &= \frac{\sigma^2}{n}.
\end{align*}
We sometimes write $\sigma^2_{\overline{X}} = \operatorname{var}[\overline{X}]$.

The quantity $\sigma_{\overline{X}} = \operatorname{sd}[\overline{X}]$ is called the *standard error of the mean*.\index{Standard error of the mean} 

::::
:::


Theorem\ \@ref(thm:SamplingDistMean) applies to random samples as defined in Sect.\ \@ref(RandomSamples), since independence and identical distributions are guaranteed by definition.

Theorem\ \@ref(thm:SamplingDistMean) shows that the expected value of\ $\overline{X}$ does not depend on the size of the sample, but the variance of\ $\overline{X}$ *decreases* as the sample size *increases*: larger samples tend to produce more precise estimates of\ $\mu$.
This makes sense intuitively; after all, one reason for preferring larger samples over smaller samples is that they produce more precise estimates.

Importantly, Theorem\ \@ref(thm:SamplingDistMean) shows *how* the variance of\ $\overline{X}$ decreases as\ $n$ gets larger: inversely proportional to\ $n$.
A sample size twice as large reduces the variance by half or, more intuitively, reduces the standard deviation by a factor of\ $\sqrt{2}$.
In practice this means that to double the precision of\ $\overline{X}$ (i.e., to halve the size of the standard deviation), the sample size needs to be increased four-fold. 

Another important feature of Theorem\ \@ref(thm:SamplingDistMean) is that the results do not rely on knowing the distribution of the population\ $X$.
The result applies for random samples drawn from *any* distribution, provided the $X_i$ are iid.
In Sect.\ \@ref(SamplingDistributionsNormal), we consider the shape of the sampling distribution when the population specifically has a normal distribution; then in Sect.\ \@ref(CentralLimitTheorem) we consider the shape of the sampling distribution when the population has *any* distribution


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
In Theorem\ \@ref(thm:SamplingDistMean), the mean and variance of\ $\overline{X}$ depend only on\ $\mu$ and\ $\sigma^2$, not on the shape of the population distribution.
:::


:::{.example #SamplingMnVarExp name="Sampling mean and variance from an exponential distribution"}
Suppose $X_1, \dots, X_n \overset{\text{iid}}{\sim}\text{Exponential}(\lambda = 2)$.

Hence, $\operatorname{E}[X_i] = 1/2 = 0.5$, and $\operatorname{var}[X_i] = 1/\lambda^2 = 0.25$, so $\operatorname{E}[\overline{X}] = 0.5$, and $\operatorname{var}[\overline{X}] = 0.25/n$.

The empirical distribution of the means of the samples is shown in Fig.\ \@ref(fig:ExpSD), for samples of size $n = 5$, $n = 20$ and $n = 100$.
The empirical estimates of\ $\operatorname{E}[\overline{X}]$ and\ $\operatorname{var}[\overline{X}]$ (shown on the figure) are very close to the theoretical values.
:::


```{r, fig.show='hide'}
set.seed(123)

ns <- c(5, 20, 100)
lambda <- 2
num_Sims <- 10000  # number of simulations

sim_means <- lapply(ns, function(n) {
  replicate(num_Sims, 
            mean(rexp(n, rate = lambda)))
})

hist(sim_means[[1]]) 
hist(sim_means[[2]])
hist(sim_means[[3]])

# The variance when n = 100 should be about twenty times larger
# than the variance for n = 5:
var(sim_means[[1]]) / var(sim_means[[3]])
```



```{r ExpSD, echo=FALSE, fig.align="center", fig.cap="Sampling distributions, for samples drawn from an exponential distribution with $\\lambda = 2$, for various sample sizes. The solid dots represent the population mean $1/\\lambda = 0.5$.", out.width='100%', fig.width=7.5, fig.height=2.5}
par(mfrow = c(1, 3))

hist(sim_means[[1]], 
     axes = FALSE,
     xlim = c(0, 2),
     main = "Sample means:\nn = 5", 
     sub = bquote("(mean;" ~ .(round(mean(sim_means[[1]]), 3)) ~ "and var:" ~ .(round(var(sim_means[[1]]), 4))),
     ylab = "",
     xlab = "Sample means")
axis(side = 1)
points(x = 1/lambda,
       y = 0,
       pch = 19)

hist(sim_means[[2]], 
     axes = FALSE,
     xlim = c(0, 2),
     main = "Sample means:\nn = 15", 
     sub = bquote("(mean;" ~ .(round(mean(sim_means[[2]]), 3)) ~ "and var:" ~ .(round(var(sim_means[[2]]), 4))),
     ylab = "",
     xlab = "Sample mean")
axis(side = 1)
points(x = 1/lambda,
       y = 0,
       pch = 19)

hist(sim_means[[3]], 
     axes = FALSE,
     xlim = c(0, 2),
     breaks = 5,
     main = "Sample means:\nn = 100", 
     sub = bquote("(mean;" ~ .(round(mean(sim_means[[3]]), 3)) ~ "and var:" ~ .(round(var(sim_means[[3]]), 4))),
     ylab = "",
     xlab = "Sample mean")
axis(side = 1)
points(x = 1/lambda,
       y = 0,
       pch = 19)


```


### Sampling distribution of $\overline{X}$: when $X$ has a normal distribution {#SamplingDistributionsNormal}

The normal distribution is used to model many naturally occurring phenomena (Sect.\ \@ref(Normal)). 
Consequently, we begin by studying the sampling distributions of statistics based on data drawn from normal populations.

The following result proves establishes that any linear combination of independent normal random variables is itself normally distributed.


:::{.theorem #LinearComb name="Linear combinations"}
Let $X_i\sim N(\mu_i,\sigma^2_i)$ where $i = 1, 2, \dots n$. 
Define the linear combination\ $Y$ as
$$
   Y = a_1 X_1 + a_2 X_2 + \dots + a_nX_n.
$$
Then $Y \sim N(\sum_i a_i\mu_i, \sum_i a^2_i\sigma^2_i)$.

::::{.proof}
From Theorem\ \@ref(thm:NormalProperties), the MGF of the random variable\ $X_i$ is
$$
   M_{X_i} (t)
   = \exp\left\{\mu_i t + \frac{1}{2} \sigma^2_i t^2\right\}\quad\text{for $i = 1, 2, \dots, n$}.
$$
So, for a constant\ $a_i$ (using Theorem\ \@ref(thm:ExpectationLinearINCORRRECT!) with $\beta = 0$, $\alpha = a_i$):
\begin{align*}
   M_{a_i X_i}(t)
   &= M_{X_i}(a_it)\\
   &= \exp\left\{\mu_i a_i t + \frac{1}{2} \sigma^2_i a^2_i t^2\right\}.
\end{align*}
Since the MGF of a sum of independent random variables is equal to the product of their MGFs,
\begin{align*}
     M_Y(t)   
     &= \prod^n_{i = 1} \exp\left( \mu_i a_i t + \frac{1}{2} \sigma^2_i a^2_i t^2 \right)\\
     &= \exp\left( t\Sigma a_i\mu_i +\frac{1}{2} t^2 \Sigma a^2_i \sigma^2_i \right).
\end{align*}
This is the MGF of a normal random variable with mean $\operatorname{E}[Y] = \Sigma a_i\mu_i$ and variance $\operatorname{var}[Y] = \Sigma a^2_i\sigma^2_i$.
::::
:::


:::{.example #LinearCombExample name="Linear combinations"}
Define
$$
  X_1 \sim N(2, \sigma^2 = 5)
  \quad\text{and}\quad
  X_2 \sim N(25, \sigma^2 = 1)
$$
and $a_1 = -8$ and $a_2 = 10$.
Then, by Theorem\ \@ref(thm:LinearComb), $Y = a_1 X_1  + a_2 X_2$ has a normal distribution, with
\begin{align*}
  \operatorname{E}[Y]    &= a_1\mu_1       + a_2\mu_2         = (-8\times 2) + (10\times 25) = 234;\\
  \text{and}\quad
  \operatorname{var}[Y]  &= a^2_1\sigma^2_1 + a_2^2\sigma^2_2 = ((-8)^2\times 5) + (10^2\times 1) = 420.
\end{align*}
This can be demonstrated in **R** (below); the simulated distribution of\ $Y$ is in Fig.\ \@ref(fig:HistLinearComb).
:::

```{r, fig.show='hide'}
set.seed(8979704) # For reproducibility

a_1 <- -8; a_2 <- 10
num_Sims <- 10000

X_1 <- rnorm(num_Sims,                 # This many simulations from a Normal...
             mean = 2, sd = sqrt(5) )  # ... with these parameters
X_2 <- rnorm(num_Sims,
             mean = 25, sd = sqrt(1) )

Y <- a_1 * X_1 + a_2 * X_2             # The linear combinations
hist(Y)                                # Produce the histogram

mean(Y); var(Y)

```

```{r HistLinearComb, echo=FALSE, out.width='100%', fig.width=8, fig.height=2.5, fig.align="center", fig.cap="Left: The normal distribution of $X_1$. Centre: the normal distribution of $X_2$. Right: the empirical distribution of a linear combination of the two normal random variates is a normal distribution. The solid dot represents the theoretical mean of $Y$, and the solid curve is the theoretical distribution of\\ $Y$."}
par(mfrow = c(1, 3))

curve( dnorm(x, mean=2, sd = sqrt(5)),
       from = -4, to = 9,
       lwd = 2,
       ylab = "",
       xlab = expression(italic(x)),
       main = expression(The~distribution~of~italic(X)[1]),
       axes = FALSE)
axis(side = 1)

curve( dnorm(x, mean=25, sd = sqrt(1)),
       from = 22, to = 28,
       lwd = 2,
       ylab = "",
       xlab = expression(italic(x)),
       main = expression(The~distribution~of~italic(X)[2]),
       axes = FALSE)
axis(side = 1)

hist(Y,
     axes = FALSE,
     freq = FALSE,
     ylab = "",
     main = expression(bold(atop(The~distribution~of~the,
                                 random~variable~italic(Y)))),
     xlab = expression(italic(y)))
axis(side = 1)
points(x = 234,
       y = 0,
       pch = 19)

curve(dnorm(x, mean = 234, sd = sqrt(420)),
      add = TRUE,
      lwd = 2,
      from = 150, 
      to = 350)
```


The sampling distribution of the sum and mean of a random sample from a normal population follows directly from Theorem\ \@ref(thm:LinearComb).


:::{.theorem #SumMean name="Sum and mean of a random sample"}
Let $X_1, X_2, \dots, X_n \overset{\text{iid}}{\sim} N(\mu,\sigma^2)$. 
Define the sum\ $T_n$ and mean\ $\overline{X}$ respectively as
\begin{align*}
   T_n
   &= X_1 + X_2 + \dots + X_n\\
   \overline{X}
   &= (X_1+X_2 + \dots + X_n)/n.
\end{align*}
Then $S\sim N(n\mu, n\sigma^2)$ and $\overline{X}\sim N(\mu, \sigma^2/n)$.

::::{.proof}
See Exercise\ \@ref(exr:SumMean).
::::
:::


Theorem\ \@ref(thm:SumMean) holds for any sample size\ $n$, provided the population has a normal distribution.



::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
For any size sample drawn from a normal distribution, the sample means have a normal distribution.
:::


:::{.example #CLTSugarSachets name="Sums of rvs"}
Sugar sachets (for sweetening hot drinks) typical are manufactured so that the weights of sugar (in grams) in each sachet\ $W$ are iid with a normal distribution having a mean of $4\gs$ and standard deviation of\ $0.1\gs$.

For catering purposes, sachets are sold in cartons of\ $1000$.
What is the probability that a carton contains more than\ $4020\gs$ of sugar? 

Since the weights\ $W$ follow a normal distribution, then the weight of the sugar in each carton is 
$$
   T = W_1 + W_2 + \cdots + W_{1000} \sim N(1000\times 4, \sigma^2=1000 \times 0.1^2).
$$
or $N(4000, 10)$.
Then, we seek $\Pr(T > 4020)$:
$$
  \Pr(T > 4020)
  = 
  \Pr(Z > (4020 - 4000) / 10) 
  = 
  \Pr(Z > 2)
  =
  1 - \phi(2)
$$
In **R**:

```{r}
1 - pnorm(2)
```

The probability is about $2.3$\%.
:::


:::{.example #CLTweights name="CLT"}
The IQs for a large population of $10$\ year-old boys is assumed to be normally distributed, with a mean of\ $110$ and a variance of\ $144$. 
How large a sample is needed to have a probability of\ $0.9$ that the *mean* IQ of the sample would not differ from the expected value\ $110$ by more than\ $5$?

Let\ $X_i$ be the IQ of the $i$th\ boy; then $X_i \sim N(110, 144)$.
Consider a sample of size\ $n$ and let $\overline{X} = \sum^n_{i = 1}X_i/n$; then $\overline{X}\sim N(110, 144/n)$.
The question asks to find the values of\ $n$ such that
$$
   \Pr(|\overline{X} - 110|\leq 5) = 0.90.
$$
That is,
$$
   \Pr\left(\frac{|\overline{X} - 110|}{12/\sqrt{n}} \leq \frac{5}{12/\sqrt{n}}\right) = 0.90
$$
hence
$$
   \Pr(Z \leq 5\sqrt{n} /12) = 0.90. 
$$
Using **R** and being careful (see Fig.\ \@ref(fig:IQFindZ)):
```{r echo=TRUE}
z <- qnorm(0.95); z
```

Thus,
$$
  5\sqrt{n}/12 = 1.644854 \implies n = (12\times 1.644854/5)^2 = 15.58.
$$


The *smallest* size sample, then, would be a sample of $n = 16$.
:::


```{r IQFindZ, echo=FALSE, fig.align="center", out.width='90%', fig.width=8, fig.height=3, fig.cap="Left: the situation in terms of $Z$. Right: finding the required value of $Z$."}
par(mfrow = c(1, 2))

xstar <- 1.6

out <- plotNormal(mu = 0,
                  sd = 1,
                  showX = c(-3, -xstar, 0, xstar, 3),
                  showXlabels = c(-3, expression(-italic(z)), 0, expression(italic(z)), 3))
shadeNormal(x = out$x,
            y = out$y,
            lo = -xstar, 
            hi =  xstar,
            col = "grey")
text(x = 2.5,
     y = 0.3,
     pos = 3,
     expression(Area*":"~0.90))
arrows(x0 = 2.5,
       y0 = 0.3,
       x1 = 0.75,
       y1 = 0.1,
       lwd = 2,
       angle = 15,
       length = 0.1)

## 

out <- plotNormal(mu = 0,
                  sd = 1,
                  showX = c(-3, -2, -1, 0, xstar, 3),
                  showXlabels = c(-3, -2, -1, 0, expression(italic(z)), 3))
shadeNormal(x = out$x,
            y = out$y,
            lo = -Inf, 
            hi =  xstar,
            col = "grey")
text(x = 2.5,
     y = 0.3,
     pos = 3,
     expression(Area*":"~0.95))
arrows(x0 = 2.5,
       y0 = 0.3,
       x1 = 0.75,
       y1 = 0.1,
       lwd = 2,
       angle = 15,
       length = 0.1)
```

:::{.example #CLTPlunger name="Two random variables"}
A design involves a plunger fitting into a cylindrical tube.
The diameter of the plunger can be considered a normal random variable with mean $2.1\cms$ and standard deviation $0.1\cms$.
The inside diameter of the cylindrical tube is a normal random variable with mean $2.3\cms$ and standard deviation $0.05\cms$. 
For a plunger and tube chosen randomly from a day's production run, find the probability that the plunger will *not* fit into the cylinder.

Let\ $X$ and\ $Y$ be the diameter of the plunger and cylinder respectively.
Then $X\sim N\big(2.1, (0.1)^2\big)$ and $Y\sim N\big( 2.3, (0.05)^2\big)$, and we seek $\Pr(Y < X)$.
The distribution of $Y - X$ is $N(2.3 - 2.1, 0.0025 + 0.01)$ so that
\begin{align*}
   \Pr(Y - X < 0) 
   &= \Pr\left(Z <\frac{0 - 0.2}{\sqrt{0.0125}}\right) \quad\text{where $Z\sim N(0,1)$}\\
   &= \Pr(Z < -1.78)\\
          &= 0.0375.
\end{align*}
In **R**:
```{r}
pnorm( -1.78 )
```
:::


### The Central Limit Theorem {#CentralLimitTheorem}

In Sect.\ \@ref(SamplingDistributionMean), general results are given describing the *mean* and *variance* of the sample mean which holds for any population distribution, but do not say anything about the *distribution* of the sample means.
Theorem\ \@ref(thm:SumMean) states that the distribution is normal when the population is normally distributed.

Although these results are important, usually the distribution of the population from which the random sample is drawn is unknown.
Remarkably, even when the population is *not* normally distributed, the sampling distribution still has an approximate normal distribution for large sample sizes.


:::{.theorem #CLT name="Central Limit Theorem (CLT)"}
Let $X_1, X_2, \dots, X_n$ be a random sample from a distribution with mean\ $\mu$ and variance\ $\sigma^2$.
Then the random variable
$$
   Z_n 
   = \frac{\overline{X} - \mu}{\sigma / \sqrt{n}}
$$
converges in distribution to a standard normal variable as $n\to\infty$.
'Convergence in distribution' means that the distribution of $Z_n$ converges to a standard normal as $n\to \infty$. 

::::{.proof}
The proof is long, so is deferred to Appendix\ \@ref(CLTProof).
::::
:::


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
The *Central Limit Theorem* (Theorem\ \@ref(thm:CLT)) is one of the most important theorems in statistics.
:::


The Central Theorem states that when\ $n$ is 'large', $Z_n$ is expected to approximate a standard normal distribution. 
Transforming $Z_n$ back to the sample mean, $\overline{X}$ can be expected to approximate a $N(\mu, \sigma^2/n)$ distribution.

In practice, the distribution of\ $\overline{X}$ is sufficiently close to that of a normal distribution when\ $n$ is larger than about\ $25$ in most situations.
However, if the population distribution is severely skewed, larger samples sizes may be necessary for the approximation to be adequate in practice.


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
In simple terms, the *Central Limit Theorem* states that, for a random variable\ $X$, whatever its distribution, the distribution of the *sample mean* $\overline{X}$ becomes approximately normal as the sample size \(n\) increases.\index{Central Limit Theorem}

In particular, for large\ $n$,

* the distribution of $\overline{X}$ is approximately normal,
* with mean $\operatorname{E}[\overline{X}] = \mu$ where $\mu = \operatorname{E}[X]$, and
* with variance $\operatorname{var}[\overline{X}] = \sigma^2/n$, where $\sigma^2 = \operatorname{var}[X]$.

The approximation generally improves as the sample size\ $n$ increases.

For many distributions that are not highly skewed, the approximation is reasonably good for sample sizes larger than about\ $20$ to $30$.

If the data come from a normal distribution, then the distribution of the sample mean is exactly normal for every sample size $n$ (Sect.\ \@ref(SamplingDistributionsNormal)).
:::


```{r, child = if (knitr::is_html_output())  './children/CLT/CLT_html.Rmd' }
```

```{r, child = if (knitr::is_latex_output()) './children/CLT/CLT_latex.Rmd'}
```


:::{.example #CLTPoisson name="CLT and a Poisson distribution"}
In the **R** code below, data come from a Poisson distribution (Fig.\ \@ref(fig:PoissonCLT), left panel) with $\lambda = 2$.
However, the mean of samples of size $n = 10$ (centre panel) and $n = 100$ (right panel) are approximately distributed as a normal distribution.
The variance of the sample means for the larger sample size is smaller than that for $n = 10$, and the distribution looks more like a normal distribution for the larger sample size.
:::


```{r PoissonCLT, echo=FALSE, fig.cap="The Central Limit Theorem. Left: the distribution of the individual observations that follow a Poisson distribution. Centre: the empirical sampling distribution of the sample mean for samples of size $n = 10$. Right: the empirical sampling distribution of the sample mean for samples of size $n = 100$. The thick dashed lines represent the theoretical asymptotic normal distribution. Note that the scale on the horizontal axis is different for each plot.", fig.align="center", fig.width=7, out.width='100%', fig.height=2.75}
# The distribution of individuals has a uniform distribution:
set.seed(40516) # For reproducibility
sample_Size_1 <- 10
sample_Size_2 <- 100
sample_Means_1 <- NULL
sample_Means_2 <- NULL
lambda <-2

number_Simulations <- 10000

for (i in 1:number_Simulations){
  # Take a sample of size 5
  Xs_1 <- rpois(sample_Size_1, 
                   lambda = lambda)
  # Take a sample of size 20
  Xs_2 <- rpois(sample_Size_2, 
                   lambda = lambda)

  # Find the sample mean
  sample_Means_1 <- c(sample_Means_1, mean(Xs_1) )
  sample_Means_2 <- c(sample_Means_2, mean(Xs_2) )
}

par(mfrow = c(1, 3))

x <- 0:10
plotDiscrete( x = x,
              type = "MF",
              px = dpois(x,
                         lambda = lambda),
              las = 1,
              xlab = expression(italic(x)),
              ylab = "Prob. fn",
              main = "Distribution of the\nindividuals observations")

hist( sample_Means_1,
      las = 1,
      ylim = c(0, 0.9),
      freq = FALSE,
      axes = FALSE,
      xlab = expression(Sample~means~bar(italic(x)) ),
      main = expression( bold( atop(Distribution~of~sample~means,
                               "for"~italic(n)==10) ) ) )
curve(dnorm(x, mean=2, sd = sqrt(2) / sqrt(sample_Size_1) ),
      lwd = 2,
      lty = 2,
      add = TRUE)
axis(side = 1)

hist( sample_Means_2,
      las = 1,
      ylim = c(0, 2.75),
      freq = FALSE,
      axes = FALSE,
      xlab = expression(Sample~means~bar(italic(x)) ),
      main = expression( bold( atop(Distribution~of~sample~means,
                               "for"~italic(n)==100) ) ) )
curve(dnorm(x, mean=2, sd = sqrt(2) / sqrt(sample_Size_2) ),
      lwd = 2,
      lty = 2,
      add = TRUE)
axis(side = 1)
```


:::{.example #CoffeeVending name="CLT"}
When a coffee machine makes a single espresso-style shot, the volume of water in millimetres\ $W$ that is dispensed has a mean of $30\mLs$ with a standard deviation of $2.2\mLs$.

Find the probability that the *average* volume of water is a random sample of size\ $36$ shots is less than $31\mLs$.

The distribution of\ $W$ is not given, but the mean and standard deviation of\ $W$ are given.
The distribution of the sample mean\ $\overline{W}$ can be approximated by the normal distribution with mean of $\mu = 30$ and standard error of $\sigma_{\overline{W}} = \sigma/\sqrt{n} = 2.2/\sqrt{36} = 0.3718\mLs$, according to the CLT.
That is, approximately $\overline{W} \sim N(30, 0.3718^2)$.
Now
$$
   \Pr(\overline{W} < 31) 
   \approx \Pr\left(Z < \frac{31 - 30}{0.3718}\right)
   = \Pr(Z < 2.69).
$$
In **R**:

```{r}
pnorm(2.69)
```

Hence, $\Pr(\overline{W} < 31) \approx 0.0036$, or approximately\ $99.6$%.
:::


:::{.example #FairDiceSum name="Throwing dice"}
Consider throwing a fair die\ $n$ times and observing the sum of the faces showing. 
For $n = 12$ rolls, find the probability that the sum of the faces is at least\ $52$.

Let the random variable\ $X_i$ be the number showing on the $i$th\ throw.
Then define $Y = X_1 + \dots + X_{12}$, so that we seek $\Pr(Y\geq 52)$.

To use Theorem\ \@ref(thm:CLT), see that '$Y\geq 52$' is equivalent to '$\overline{X}\geq 52/12$', where $\overline{X} = Y/12$ is the mean number showing from the $n = 12$\ tosses.

Since the distribution of each\ $X_i$ is a continuous uniform distribution with $\Pr(X_i = x) = 1/6$ (for $x = 1, 2, \dots, 6$), then $\operatorname{E}[X_i] = 7/2$ and $\operatorname{var}[X_i] = 35/12$ (Sect.\ \@ref(DiscreteUniformProperties)).
Hence, $\operatorname{E}[\overline{X}] = 7/2$ and $\operatorname{var}[\overline{X}] = 35/(12^2)$, and so 
$\overline{X} \sim N(7/2, \sigma^2 = 35/(12^2))$. 

Then, from Theorem\ \@ref(thm:CLT),
\begin{align*}
  \Pr(Y\geq 52) 
    & \simeq \Pr(\overline{X}\geq 52/12)\\
    &= \Pr_N \left(Z\geq \frac{52/12 - 7/2}{\sqrt{35/144}}\right)\\
    &= 1 - \Phi(1.690) = 0.0455
\end{align*}
The probability is approximately\ $4.6$%.
:::


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
The Central Limit Theorem also applies to the *sum*
$$
  Y = \sum_{i=1}^n X_i,
$$
not just the sample mean.
If $X_1,\ldots,X_n$ are iid random variables with
$$
  \operatorname{E}[X_i] = \mu
  \qquad
  \text{and}
  \qquad
  \operatorname{var}[X_i] = \sigma^2,
$$
then for large\ $n$,
$$
  Y
  =
  \sum_{i=1}^n X_i
  \approx
  N(n\mu, n\sigma^2).
$$
That is, the sum is approximately normally distributed with

* mean $\operatorname{E}[Y] = n\mu$, and
* variance $\operatorname{var}[Y] = n\sigma^2$.

This result is often useful when modelling totals, such as total sales, total demand, total rainfall, or total waiting time.
:::


:::{.example #NoiseVoltages name="CLT (Voltages)"}
Suppose we have a number of independent noise voltages\ $V_i$ (for $i = 1, 2, \dots, n$). 
Let\ $V$ be the sum of the voltages, and suppose each\ $V_i$ is distributed $U(0, 10)$.
For $n = 20$, find $\Pr(V > 105)$.

This is an example of the CLT written as a sum. 
To find $\Pr(V > 105)$, the distribution of\ $V$ must be known.

Since $\operatorname{E}[V_i] = 5$ and $\operatorname{var}[V_i] = 100/12$, $V$ has an approximate normal distribution with mean $20\times 5 = 100$ and variance $20\times 100/12$.
That is, $\displaystyle{\frac{V - 100}{10\sqrt{5/3}}}$ is distributed $N(0, 1)$ approximately. 
So
$$ 
   \Pr(V > 105) 
   \approx \Pr\left (Z > \frac{105 - 100}{12.91}\right )
   = 1 - \Phi(0.387)
   = 0.352.
$$
The probability is approximately\ $35$%.
:::



### The normal approximation to the binomial {#NormalApproxBinomialCLT}

\index{Binomial distribution!normal approximation}
The normal approximation to the binomial distribution (Sect.\ \@ref(NormalApproxBinomial)) can be seen as an application of the [Central Limit Theorem](#thm:CLT).\index{Central Limit Theorem}
The essential observation is that a sample proportion is a sample mean. 
Consider a sequence of independent [Bernoulli trials](#def:BernoulliTrials) resulting in the random sample $X_1, X_2, \dots, X_n$ where
$$
  X_i
  =
  \begin{cases}
     0  &  \text{if failure}\\
     1  &  \text{if success}
  \end{cases}
$$
denotes whether or not the $i$th\ trial is a success.
Then the sum
$$
   Y = \sum_{i = 1}^n X_i
$$
represents the number of successes in the $n$\ trials and
$$
   \overline{X}
   = 
   \frac{1}{n}\sum_{i = 1}^n X_i
   = 
   \frac{Y}{n}
$$
is a sample mean representing the proportion or fraction of trials which are successful. 
In this context, $\overline{X}$ is usually denoted by the sample proportion\ $\widehat{p}$.

Note that $\operatorname{E}[X_i] = p$ and $\operatorname{var}[X_i] = p(1 - p)$.
Therefore
$$
   \operatorname{E}[Y] = np
   \quad\text{and}\quad
   \operatorname{var}[Y] = np(1 - p)
$$
and
$$
   \operatorname{E}[\overline{X}] = p
   \quad\text{and}\quad
   \operatorname{var}[\overline{X}] = \frac{p(1 - p)}{n}.
$$
Theorem\ \@ref(thm:CLT) is applicable to\ $\overline{X}$ and\ $Y$ respectively. 
Hence
\begin{align*}
   \overline{X} &= \widehat{p}
   \sim N\left(p, \frac{p(1 - p)}{n}\right)\text{ approximately}\\
\text{and}\quad
   Y &= n\widehat{p}
   \sim N(np, np(1 - p))\text{ approximately}.
\end{align*}


## The $\chi^2$-distribution {#ChiSquaredDistribution}

Having discussed the sampling distribution of the sample mean, we now turn to the sampling distribution of the sample variance
\begin{equation}
  S^2 = \frac{1}{n-1} \sum_{i=1}^n (X_i - \overline{X})^2
  (\#eq:S2)
\end{equation}
for a sample $X_1, \dots, X_n$ drawn iid from $N(\mu, \sigma^2)$ (Theorem\ \@ref(thm:SampleVarianceUnbiasedXbar)).
We seek the distribution of $S^2$.


\index{Chi-squared distribution|(}
The formula for the estimator\ $S^2$ in Eq.\ \@ref(eq:S2) involves an\ $\sum X_i^2$ term.
Hence, to find the distribution of\ $S^2$, we start with the distribution of a squared normal variate.
Specifically, consider
$$
  Z = \frac{X - \mu}{\sigma} \sim N(0,1)
$$
and the distribution of $Y = Z^2$.
The distribution of\ $Y$ is easy to determine using the techniques in Chap.\ \@ref(ChapterTransformations) (see Exercise\ \@ref(exr:DistZ2)), and yields
$$
  f_Y(y) = \frac{y^{-1/2}\exp(-y/2)}{\sqrt{2\pi}} \quad\text{for $y > 0$}.
$$
Note that $Y \sim \text{Gamma}(\alpha = 1/2, \beta = 2)$.

Now consider $Z_1, Z_2, \dots Z_n\overset{\text{iid}}{\sim} N(0, 1)$, and the summation $\sum_{i = 1}^n Z_i^2$.
Since each\ $Z_i^2$ has the same gamma distribution, the distribution of the summation can be found, using MGFs (Exercise\ \@ref(exr:SumZ2)), as
$$
  \sum_{i = 1}^n Z_i^2 \sim \text{Gamma}(\alpha = n/2, \beta = 2).
$$
This distribution occurs so often in practice that it has its own name.
A $\text{Gamma}(\alpha = n/2, \beta = 2)$ distribution is defined as a $\chi^2_n$-distribution, and so
\begin{equation}
  \sum_{i = 1}^n Z_i^2 \sim \chi^2_n,
  (\#eq:SumZ2)
\end{equation}
where\ $n$ here is the *degrees of freedom*.\index{Degrees of freedom}


:::{.definition #ChiSquaredDistribution name="Chi-squared distribution"}
A continuous random variable\ $X$ with probability density function\index{Chi-squared distribution}
\begin{equation}
   f_X(x)
   = \frac{x^{(\nu/2) - 1}\exp(-x/2)}{2^{\nu/2}\,\Gamma(\nu/2)}
   \quad\text{for $x > 0$}
\end{equation}
is said to have a *chi-squared distribution* with parameter $\nu > 0$, called the *degrees of freedom*.
We write $X \sim \chi^2_\nu$.
:::


Some plots of $\chi^2$-distributions are shown in Fig.\ \@ref(fig:ChisqPlots).


::: {.linkBox .link data-latex="{iconmonstr-link-1-240.png}"}
The chi-squared distribution is a special case of the gamma distribution, with $\alpha = \nu/2$ and $\beta = 2$.
:::


```{r ChisqPlots, echo=FALSE, out.width='100%', fig.height=2.5, fig.width=6.5, fig.cap="Some $\\chi^2\\!$-distribution probability density functions."}
par( mfrow = c(1, 3))

xx <- seq(0.01, 4, length = 100)
plot( x = xx,
      y = dchisq(xx, df = 1),
      lwd = 2,
      type = "l",
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression( bold(chi^2*"-"*distribution: df==1)))

xx <- seq(0.001, 15, length = 100)
plot( x = xx,
      y = dchisq(xx, df = 5),
      lwd = 2,
      type = "l",
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression( bold(chi^2*"-"*distribution: df==5)))

xx <- seq(0.001, 50, length = 100)
plot( x = xx,
      y = dchisq(xx, df = 20),
      lwd = 2,
      type = "l",
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression( bold(chi^2*"-"*distribution: df==20)))
```


:::{.theorem #ChisqProperties name="Properties of the chi-squared distribution"}
If $X\sim\chi^2_\nu$ then

* $\operatorname{E}(X) = \nu$.
* $\operatorname{var}(X) = 2\nu$.
* $M_X(t) = (1 - 2t)^{-\nu/2}$ for $t < 1/2$.

::::{.proof}
Since the chi-squared distribution is a special case of the gamma distribution, these properties can be obtained directly from those for the gamma distribution (see Theorem\ \@ref(thm:GammaProperties)).
::::
:::
\index{Chi-squared distribution|)}


Having found the distribution of the sum of\ $n$ random variables, we can return to the original task: finding the sampling distribution of the sample variance\ $S^2$ in Eq.\ \@ref(eq:S2).
First, write
$$
  \sum_{i=1}^n (X_i - \overline{X})^2
  = 
  \sum_{i=1}^n (X_i - \mu)^2 - 
  n (\overline{X} - \mu)^2
$$
(see Exercise\ \@ref(exr:VarianceDecomposition)).
Dividing this expression by\ $\sigma^2$ then gives
\begin{align*}
  W
  &=
  \sum_{i=1}^n \frac{(X_i - \overline{X})^2}{\sigma^2} \\
  &=
  \sum_{i=1}^n \left( \frac{X_i - \mu}{\sigma}\right)^2
  -
  \frac{n(\overline{X} - \mu)^2}{\sigma^2}.
\end{align*}
Now, write $W = U - V$, where
\begin{align*}
  U
  &=
  \sum_{i=1}^n \frac{(X_i - \mu)^2}{\sigma^2}
  \sim \chi^2_n\\
  \text{and}\quad
  V
  &=
  \frac{n(\overline{X} - \mu)^2}{\sigma^2}
  \sim \chi^2_1.
\end{align*}
Using the fact that\ $U$ and\ $V$ are independent (a consequence of the independence of sample mean and sample variance under normality), we proceed via MGFs.
Since $U = W + V$ and\ $W$ and\ $V$ are independent, the MGF of\ $U$ factorises as
$$
  M_U(t)
  =
  M_W(t)\cdot M_V(t).
$$
The MGF of a \(\chi^2_k\) distribution is $M(t) = (1- 2 t)^{-k/2}$.
Hence
$$
  (1-2t)^{-n/2}
  =
  M_W(t)\cdot (1-2t)^{-1/2},
$$
which implies
$$
  M_W(t)
  =
  (1 - 2t)^{-(n-1)/2}.
$$
This is the MGF of a $\chi^2_{n-1}$ distribution, and hence
$$
  W
  \sim
  \chi^2_{n-1}.
$$
This means that
$$
  \frac{1}{\sigma^2} \sum_{i=1}^n (X_i - \overline{X})^2 \sim \chi^2_{n-1}
$$
and hence
$$
  \frac{(n - 1)S^2}{\sigma^2} \sim \chi^2_{n-1}
$$
after substituting Eq.\ \@ref(eq:S2) and rearranging.
Thus, \(S^2\) has a scaled \(\chi^2_{n-1}\) distribution.


:::{.theorem #SamplingDistVariance name="Sampling distribution of the variance"}
If $X_1, X_2, \dots, X_n$ is an *iid sample* of size\ $n$ from a $N(\mu, \sigma^2)$ distribution, then 
$$
  \frac{(n - 1)S^2}{\sigma^2} \sim \chi^2_{n - 1}.
$$

::::{.proof}
The proof has been outlined in the above discussion.
::::
:::


::: {.softwareBox .software data-latex="{iconmonstr-laptop-4-240.png}"}
The four **R** functions for working with the $\chi_\nu^2$-distribution have the form `[dpqr]chisq(., df)`, where `df`${} = \nu$ is the degrees of freedom (see Appendix\ \@ref(UseRDistributions)):

* `dchisq(x, df)` computes the PDF at $X = {}$`x`;
* `pchisq(q, df)` computes the CDF at $X = {}$`q`;
* `qchisq(p, df)` computes the quantile for cumulative probability `p`; and
* `rchisq(n, df)` generates `n` random observations.
:::


:::{.example #SamplingDistVariance name="Sampling distribution of the variance"}
Consider taking samples of size $n = 5$ and $n = 20$ from a $N(5, \sigma^2 = 4)$ distribution.
The sampling distribution of
$$
  \frac{n - 1}{4} S^2 \sim \chi^2_{n - 1}.
$$
Figure\ \@ref(fig:ChiS2R) shows the empirical sampling distribution for the scaled sample standard deviation for both sample sizes.
Again the larger sample size produces a sampling distribution with a smaller variance.
:::

```{r fig.show='hide'}
set.seed(42)

n_a <- 5; n_b <- 20
mu <- 5; sigma2 <- 4
num_Sims   <- 10000

# Simulate nsim values of (n-1)S^2 / sigma^2
scaled_S2_a <- replicate(num_Sims, {
  x <- rnorm(n_a, 
             mean = mu, sd = sqrt(sigma2))
  (n_a - 1) * var(x) / sigma2
})
scaled_S2_b <- replicate(num_Sims, {
  x <- rnorm(n_b, 
             mean = mu, sd = sqrt(sigma2))
  (n_b - 1) * var(x) / sigma2
})

# Plot histogram vs true chi^2(n-1) density
hist(scaled_S2_a, freq = FALSE)

curve(dchisq(x, df = n_a - 1), add = TRUE, lwd = 2)

hist(scaled_S2_b, freq = FALSE)

curve(dchisq(x, df = n_b - 1), add = TRUE, lwd = 2)

```


```{r ChiS2R, echo=FALSE, fig.align="center", fig.cap="The sampling distribution of the scaled standard deviation. The scale on the horizontal axis is different for the two plots.", fig.width=8, fig.height=3.25, out.width='90%'}
# Plot histogram vs true chi^2(n-1) density
par(mfrow = c(1, 2))

hist(scaled_S2_a, 
     freq = FALSE,
     ylim = c(0, 0.18),
     axes = FALSE,
     main = "Scaled sample standard\ndeviation: n = 5",
     ylab = "",
     xlab = "Sample standard deviation")
axis(side = 1)

curve(dchisq(x, df = n_a - 1), 
      add = TRUE, 
      lwd = 2)

###

hist(scaled_S2_b, 
     freq = FALSE,
     axes = FALSE,
     ylab = "",
     main = "Scaled sample standard\ndeviation: n = 20",
     xlab = "Sample standard deviation")
axis(side = 1)

curve(dchisq(x, df = n_b - 1), 
      add = TRUE, 
      lwd = 2)


```


:::{.example #Chi2Eng name="Sample standard deviation"}
Quality control engineers routinely need to assess the variability of a production process. 
A process may be centred correctly, yet still produce defective items if the spread is too large.

Suppose a manufacturer produces steel rods whose diameter (in mm) should follow a specified $N(\mu, \sigma^2)$ distribution. The target standard deviation is $\sigma = 0.05\mms$; large variances are outside tolerance.

The production line is monitored by a quality engineer by drawing a random sample of $n = 25$ rods each hour and computing the sample variance $S^2$.
In one sample of $n = 25$ rods, the engineer finds $s^2 = 0.0031\mms^2$.
Find the probability that $S^2 > 0.0031$ when $\sigma^2 = 0.0025$.

Using the chi-squared result, 
$$
  \frac{n-1}{\sigma^2} S^2 
  \sim  \chi^2_{n-1} 
  \quad
  \text{so that}
  \quad  
  \frac{24}{0.0025} S^2 
  =
  9600\, S^2 
  \sim  \chi^2_{24}.
$$

Thus, 
$$
  \Pr(S^2 > 0.0031)
  =
  \Pr\left(9600\,S^2 > \frac{24\times 0.0031}{0.0025}\right).
$$
In\ **R**:

```{r ChiExample}
1 - pchisq(29.76, 
           df = 24)
```
There is an approximate $19$% probability of observing a sample variance this large or larger purely by chance, if the true variance is $\sigma^2 = 0.0025$.
:::


## The $t$-distribution {#TDistribution}

\index{t-distribution@$t$-distribution}
In statistical inference, a common goal is to make statements about a population mean\ $\mu$ using the sample mean\ $\overline{X}$.
For example, Lin et al.\ (@lin2021sleep) studied whether the average overnight sleep time for pre-school children equals the recommended $10\hs$.

If the population variance\ $\sigma^2$ was known, then the Central Limit Theorem implies that
$$
  Z = \frac{\overline{X} - \mu}{\sigma/\sqrt{n}}
$$
is approximately standard normal for large\ $n$, regardless of the distribution of\ $X$.

In practice, however, the population variance\ $\sigma^2$ is almost never known. 
Instead, it is estimated using the sample variance\ $S^2$, leading to the statistic
$$
  T = \frac{\overline{X} - \mu}{S/\sqrt{n}}.
$$
Unlike\ $Z$, this statistic does not have a normal distribution. 
Its exact distribution is the *Student's $t$-distribution*, which accounts for the additional uncertainty introduced by estimating\ $\sigma$.
This extra uncertainty is especially important for small samples.

Suppose that the population distribution of\ $X$ is normal.
Then,
\begin{align*}
  Z &= \frac{\overline{X} - \mu}{\sigma/\sqrt{n}} \sim N(0, 1) \quad\text{and}\\
  V &= \frac{(n - 1) S^2}{\sigma^2}\sim \chi^2_{n-1}.
\end{align*}
In addition, $Z$\ and $V$ are independent.
Hence
$$
  T = \frac{Z}{\sqrt{V/(n-1)}}
$$
This ratio occurs so often in statistical inference, and in practice, that it has its own name.
We say that\ $T$ has a $t$-distribution (or, a *Student's $t$-distribution*)\index{Student's $t$ distribution} with\ $\nu$ degrees of freedom, written $T \sim t_\nu$.
More generally, if \(V \sim \chi^2_\nu\), then the following construction defines the \(t\)-distribution.


:::{.definition #TDistribution name="$t$-distribution"}
Suppose \(Z \sim N(0,1)\) and \(V \sim \chi^2_\nu\). 
Then the random variable
\begin{equation}
  T = \frac{Z}{\sqrt{V/\nu}}
  (\#eq:Tdist)
\end{equation}
has a $t$-distribution with\ $\nu$ degrees of freedom.
:::


For non-normal populations, the statistic\ $T$ is not exactly $t$-distributed. 
However, for large samples, its distribution is often well approximated by a $t$-distribution due to the Central Limit Theorem and the fact that\ $S^2$ converges to\ $\sigma^2$ for large samples


:::{.theorem #TDistributionPDF name="$t$-distribution PDF"}
A continuous random variable\ $X$ with probability density function
\begin{equation}
   f_X(x)
   = \frac{\Gamma\left((\nu + 1)/2\right)}{\sqrt{\pi \nu}\,\Gamma(\nu/2)}
     \left(1 + \frac{x^2}{\nu}\right)^{-(\nu + 1)/2}
     \quad\text{for $x\in\mathbb{R}$}
\end{equation}
is said to have a *$t$-distribution* with parameter $\nu > 0$.
The parameter\ $\nu$ is called the *degrees of freedom*. 
We write $X \sim t_\nu$.

::::{.proof}
This proof is not given.
::::
:::


The probability density function of the $t$-distribution is very similar to the standard normal distribution (Fig.\ \@ref(fig:TPlots)): bell-shaped and symmetric about zero. 
However, the distribution has heavier tails than the standard normal distribution, reflecting its larger variance (Theorem\ \@ref(thm:TDistProperties)).



::: {.softwareBox .software data-latex="{iconmonstr-laptop-4-240.png}"}
The four **R** functions for working with the $t$-distribution have the form `[dpqr]t(., df)`:

* `dt(x, df)` computes the PDF at $X = {}$`x`;
* `pt(q, df)` computes the CDF at $X = {}$`q`;
* `qt(p, df)` computes the quantile for cumulative probability `p`; and
* `rt(n, df)` generates `n` random numbers,

where `df` is the degrees of freedom (see Appendix\ \@ref(UseRDistributions)). 
:::


```{r TPlots, echo=FALSE, out.width='100%', fig.height=2.75, fig.width=6.5, fig.cap="Some $t$-distributions (with normal distributions in lighter-coloured lines), with mean\\ $0$ and variance\\ $1$."}
par( mfrow = c(1, 3))

xx <- seq(-4, 4,
          length = 100)

plot( x = xx,
      y = dt(xx, df = 2),
      ylim = c(0, 0.4),
      lwd = 4,
      type = "l",
      axes = FALSE,
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression( bold(atop(The~italic(t)*"-"*distribution,
                                   with~2~df))) )
axis(side = 1)
lines( x = xx,
       y = dnorm(x = xx),
       lwd = 2,
       col = ColourLight) 

plot( x = xx,
      y = dt(xx, df = 7),
      lwd = 4,
      type = "l",
      axes = FALSE,
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression( bold(atop(The~italic(t)*"-"*distribution,
                                   with~7~df))))
axis(side = 1)
lines( x = xx,
       y = dnorm(x = xx),
       lwd = 2,
       col = ColourLight)

plot( x = xx,
      y = dt(xx, df = 30),
      lwd = 4,
      type = "l",
      axes = FALSE,
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression( bold(atop(The~italic(t)*"-"*distribution,
                                   with~30~df))))
axis(side = 1)
lines( x = xx,
       y = dnorm(x = xx),
       lwd = 2,
       col = ColourLight)
```


:::{.theorem #TDistProperties name="Properties of the $t$-distribution"}
If $X\sim t_\nu$ then

* For $\nu > 1$, $\operatorname{E}[X] = 0$.
* For $\nu > 2$, $\operatorname{var}[X] = \displaystyle{\frac{\nu}{\nu - 2}}$.
* The MGF does not exist. 
  Only moments of order less than\ $\nu$ exist; that is, $\operatorname{E}[|X|^k]$ is finite if and only if $k < \nu$.

::::{.proof}
This proof is not given.
::::
:::


Although we won't prove it, as $\nu\to\infty$ the $t$-distribution converges to the standard normal (which can be seen in Fig.\ \@ref(fig:TPlots)).
In addition, from Theorem\ \@ref(thm:TDistProperties), $\operatorname{var}[X] \to 1$ as $\nu \to \infty$.
Notice that\ $T$ represents a *standardised* version of the sample mean.


:::{.example #CalcaltingT name="Calculating $T$"}
A random sample $\{21, 18, 16, 24, 16\}$ is drawn from a normal population with mean of\ $20$.

To find the value of\ $T$ for this sample, firdst find (from the sample): $\overline{x} = 19.0$ and $s^2 = 12.0$. 
Therefore $t = \frac{19.0 - 20}{\sqrt{12.0/5}} = -0.645$.
(Again, lower-case symbols are used for specific values of statistics, and upper-case symbols for the random variables.)

We can then ask: In random samples from this population, what is the probability that\ $T$ is less than the value found above?

Interest here is in $\Pr(T < -0.645)$ where $T\sim t_4$. 
From **R**, the answer is approximately\ $0.277$:

```{r echo=TRUE}
pt(-0.645, df = 4)
```
:::


:::{.example #Tsleeptime name="$T$-statistic"}
The recommendation for Taiwanese pre-school children is to have at least\ $10\hs$ of sleep per night.
@lin2021sleep sampled\ $39$ girls and found the sample mean was =\ $8.64\hs$ with a sample standard deviation of $s = 0.37\hs$.
Determine the probability that the sample would would be less than\ $8.64\hs$ if the mean in the population was\ $10\hs$, as recommended.

Define the overnight sleep time for each girls as\ $X$.
Since the sample size is moderately large ($n=39$), the Central Limit Theorem implies that the sample mean\ $\overline{X}$ is approximately normally distributed.

Because the population standard deviation\ $\sigma$ is unknown, we use\ $s$.
Then
$$
  T = \frac{\overline{X} - \mu}{S/\sqrt{n}}
$$
has an approximate $t$-distribution, with $\nu = n - 1 = 38$ degrees of freedom.

Using $\mu = 10$, $s = 0.37$ and $\overline{X} = 8.64$, compute
$$
  t = \frac{8.64 - 10}{0.37/\sqrt{39}} \approx -22.97.
$$
Hence $\Pr(\overline{X} < 8.64) = \Pr(T < -22.97)$; using\ **R**:

```{r}
pt(-22.97, df = 38)
```

If the population mean is\ $10\hs$ (as per the recommendation), it is *extremely* unlikely that the sample mean from a sample of\ $39$ girls would be lower than $8.64\hs$. 
:::



## The $F$-distribution {#FDistribution}

\index{F-distribution@$F$-distribution}
Another distribution that arises naturally in statistical inference is the $F$-distribution.
It appears as the distribution of a ratio of two independent chi-squared random variables, and is used for comparing variances and in analysis of variance (ANOVA)\index{ANOVA}\index{Analysis of variance}, a standard method for comparing multiple population means.

Suppose we have two independent random samples from normal populations with variances\ $\sigma_1^2$ and\ $\sigma_2^2$, of sizes\ $n_1$ and\ $n_2$, respectively.
Then, from Theorem\ \@ref(thm:SamplingDistVariance),
$$
 U_1 =  \frac{(n_1 - 1)S_1^2}{\sigma_1^2} \sim \chi^2_{n_1 - 1}
  \quad\text{and}\quad
  U_2 = \frac{(n_2 - 1)S_2^2}{\sigma_2^2} \sim \chi^2_{n_2 - 1}.
$$
The two sample variances can be compared using
$$
  \frac{U_1/(n_1-1)}{U_2/(n_2-1)}.
$$
This ratio occurs so often in practice that is has its own name: an $F$-distribution.


:::{.definition #FDistribution name="$F$-distribution"}
A random variable \(F\) has an $F$-distribution with degrees of freedom\ $\nu_1$ and\ $\nu_2$, written $F \sim F_{\nu_1,\nu_2}$, if it can be expressed as
$$
  F = \frac{U_1/\nu_1}{U_2/\nu_2},
$$
where $U_1 \sim \chi^2_{\nu_1}$, $U_2 \sim \chi^2_{\nu_2}$, and\ $U_1$ and\ $U_2$ are independent.
We write $X \sim F_{\nu_1, \nu_2}$.
:::

This structure makes the $F$-distribution central to inference on variances and mean comparisons.
Some $F$-distributions are shown in Fig.\ \@ref(fig:FPlots).


:::{.theorem #FProperties name="Properties of the $F$-distribution"}
If $X\sim F_{\nu_1, \nu_2}$ then

* For $\nu_2 > 2$,
  $\operatorname{E}[X] = \displaystyle{\frac{\nu_2}{\nu_2-2}}$.
* For $\nu_2 > 4$,
  $\operatorname{var}[X] = \displaystyle{\frac{2\nu_2^2(\nu_1 + \nu_2-2)}
  {\nu_1(\nu_2 - 2)^2(\nu_2 - 4)}}$.
* The MGF does not exist.

::::{.proof}
Not covered.
::::
:::


::: {.softwareBox .software data-latex="{iconmonstr-laptop-4-240.png}"}
The four **R** functions for working with the $F$-distribution have the form `[dpqr]f(., df1, df2)`, where `df1`${} = \nu_1$ and
`df2`${} = \nu_2$ are the numerator and denominator degrees of freedom respectively (see Appendix\ \@ref(UseRDistributions)):

* `df(x, df1, df2)` computes the PDF at $X = {}$`x`;
* `pf(q, df1, df2)` computes the CDF at $X = {}$`q`;
* `qf(p, df1, df2)` computes the quantile for cumulative probability `p`;
  and
* `rf(n, df1, df2)` generates `n` random observations.
:::


```{r FPlots, echo=FALSE, out.width='100%', fig.height=2.5, fig.width=6.5, fig.cap="Some $F$-distributions, for various numerator (df${}_1$) and denominator (df$_2$) degrees of freedom."}
par(mfrow = c(1, 3))

xx2 <- seq(0.01, 5, length = 100)
xx <- seq(0.01, 4, length = 100)

plot( x = xx2,
      y = df(xx2, df1 = 1, df2 = 100),
      ylim = c(0, 0.5),
      lwd = 2,
      type = "l",
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      axes = FALSE,
      main = expression(bold(atop(The~italic(F)*"-"*distribution,
                                  with~df[1]==1))))
axis(side = 1)
box()
curve(df(x, df1 = 1, df2 = 10),
      from = min(xx2),
      to = max(xx2),
      lwd = 2,
      lty = 2,
      add = TRUE)
curve(df(x, df1 = 1, df2 = 1 ),
      from = min(xx2),
      to = max(xx2),
      lty = 3,
      add = TRUE)
legend("topright",
       lty = c(3, 2, 1),
       lwd = 2,
       bty = "n",
       legend = c(expression(df[2] == 1),
                  expression(df[2] == 5),
                  expression(df[2] == 100)) )


plot( x = xx,
      y = df(xx, df1 = 5, df2 = 100),
      lwd = 2,
      type = "l",
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      axes = FALSE,
      main = expression(bold(atop(The~italic(F)*"-"*distribution,
                                  with~df[1]==5))))
axis(side = 1)
box()

curve(df(x, df1 = 5, df2 = 10 ),
      from = min(xx),
      to = max(xx),
      lwd = 2,
      lty = 2,
      add = TRUE)
curve(df(x, df1 = 5, df2 = 1 ),
      from = min(xx),
      to = max(xx),
      lwd = 2,
      lty = 3,
      add = TRUE)
legend("topright",
       lty = c(3, 2, 1),
       lwd = 2,
       bty = "n",
       legend = c(expression(df[2] == 1),
                  expression(df[2] == 5),
                  expression(df[2] == 100)) )




plot( x = xx,
      y = df(xx, df1 = 10, df2 = 100),
      lwd = 2,
      type = "l",
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      axes = FALSE,
      main = expression(bold(atop(The~italic(F)*"-"*distribution,
                                  with~df[1]==20))) )
axis(side = 1)
box()

curve(df(x, df1 = 10, df2 = 10 ),
      from = min(xx),
      to = max(xx),
      lwd = 2,
      lty = 2,
      add = TRUE)
curve(df(x, df1 = 10, df2 = 1 ),
      from = min(xx),
      to = max(xx),
      lwd = 2,
      lty = 3,
      add = TRUE)
legend("topright",
       lty = c(3, 2, 1),
       lwd = 2,
       bty = "n",
       legend = c(expression(df[2] == 1),
                  expression(df[2] == 5),
                  expression(df[2] == 100)) )

```


The main result connecting the $F$-distribution to samples from normal populations is stated in the following Theorem.


:::{.theorem #FGenerate name="Sampling distribution of the ratio of two sample variances"}
Let $X_1, X_2, \dots, X_{n_1}$ be a random sample of size\ $n_1$ from $N(\mu_1, \sigma_1^2)$ and $Y_1, Y_2,\dots, Y_{n_2}$ be an independent random sample of size\ $n_2$ from $N(\mu_2, \sigma_2^2)$.
Then the random variable
$$
   F = \frac{S_1^2/\sigma_1^2}{S_2^2/\sigma_2^2}
$$
follows an $F_{n_1 - 1,\, n_2 - 1}$-distribution.

::::{.proof}
**A partial proof only is given.**
By Theorem\ \@ref(thm:SamplingDistVariance),
$$
   U_1 = \frac{(n_1 - 1)S_1^2}{\sigma_1^2} \sim \chi^2_{n_1 - 1}
   \qquad\text{and}\qquad
   U_2 = \frac{(n_2 - 1)S_2^2}{\sigma_2^2} \sim \chi^2_{n_2 - 1},
$$
and $U_1$, $U_2$ are independent because the two samples are independent.
Setting $\nu_1 = n_1 - 1$ and $\nu_2 = n_2 - 1$, the statistic can be written as
$$
   F = \frac{S_1^2/\sigma_1^2}{S_2^2/\sigma_2^2}
     = \frac{U_1/\nu_1}{U_2/\nu_2}.
$$
That this ratio has an $F_{\nu_1, \nu_2}$-distribution follows directly from Def.\ \@ref(def:FDistribution); the derivation of the PDF is not given here.
::::
:::


::: {.linkBox .link data-latex="{iconmonstr-link-1-240.png}"}
If $T \sim t_n$ then $T^2 \sim F_{1,\, n}$.
:::


::::{.proof}
From Def.\ \@ref(def:TDistribution), $T = Z / \sqrt{V/n}$ where $Z \sim N(0,1)$ and $V \sim \chi^2_n$ are independent.
Then
$$
  T^2 = \frac{Z^2/1}{V/n}.
$$
Since $Z^2 \sim \chi^2_1$ (Sect.\ \@ref(ChiSquaredDistribution)), this is a ratio of the form $\chi^2_1/1$ over $\chi^2_n/n$, which by Def.\ \@ref(def:FDistribution) has an $F_{1,n}$-distribution.
::::



:::{.example #FProbs name="$F$-distribution probabilities"}
Suppose $X\sim F_{2, 10}$.
Find:

* $\Pr(X < 1)$
* $x$ such that $\Pr(X > x) = 0.01$

Using **R**:

```{r echo=TRUE}
pf(1,    df1 = 2, df2 = 10)
qf(0.99, df1 = 2, df2 = 10)
```
:::

:::{.example #FVarianceTest name="Comparing two variances"}
Suppose two independent random samples are taken from normal populations.

* Sample 1: $n_1 = 12$ and $S_1^2 = 18$;
* Sample 2: $n_2 = 10$ and $S_2^2 = 8$.

To compare the population variances, consider the statistic
$$
  F
  =
  \frac{S_1^2}{S_2^2}
  =
  \frac{18}{8}
  =
  2.25.
$$

If we assume that the two population variances are equal (so that $\sigma_1^2 = \sigma_2^2$), then
$$
  F \sim F_{11, 9},
$$
since the numerator and denominator degrees of freedom are
$$
  \nu_1 = n_1 - 1 = 11
  \qquad\text{and}\qquad
  \nu_2 = n_2 - 1 = 9.
$$

Large values of \(F\) provide evidence that the first population variance is larger than the second.
The probability that we would observe a value of the $F$-statistic as large as $F = 2.25$ or larger is:

```{r}
1 - pf(2.25, df1 = 11, df2 = 9)
```

Thus, it is reasonably likely that we would observe an $F$-value as large as $F = 2.25$ or larger, just through random sampling.
:::




## Appendix: Proof of the CLT {#CLTProof}

This Appendix outlines the proof of the Central Limit Theorem (Theorem\ \@ref(thm:CLT)).

Without loss of generality, centre the $X_i$ by letting $Y_i = (X_i - \mu)/\sigma$,
so that $\operatorname{E}[Y_i] = 0$, $\operatorname{var}[Y_i] = 1$, and
$$
  Z_n = \frac{1}{\sqrt{n}} \sum_{i=1}^n Y_i.
$$
Let $\varphi(t) = \operatorname{E}[\exp(itY)]$ denote the common characteristic function (CF) of the\ $Y_i$.
Since the\ $Y_i$ are independent, the CF of\ $Z_n$ is
\begin{align*}
  \varphi_{Z_n}(t)
  &= \operatorname{E}[\exp(itZ_n)]\\
  &= \operatorname{E}\left[\exp\left(\frac{it}{\sqrt{n}}\sum_{i=1}^n Y_i\right)\right]\\
  &= \prod_{i=1}^n \operatorname{E}\left[\exp(it Y_i/\sqrt{n})\right]\\
  &= \left[\varphi\!\left(t/\sqrt{n}\right)\right]^n.
\end{align*}

Since $\operatorname{E}[Y^2] = 1 < \infty$, the CF $\varphi(t)$ is twice differentiable at $t = 0$, with
$$
  \varphi'(t)\big|_{t=0} = i\operatorname{E}[Y] = 0
  \qquad\text{and}\qquad
  \varphi''(t)\big|_{t=0} = i^2 \operatorname{E}[Y^2] = -1.
$$
The Taylor expansion of $\varphi$ about $t = 0$ is therefore
\begin{align*}
  \varphi(t)
  &= 1 + \varphi'(0)\,t + \frac{1}{2}\varphi''(0)\,t^2 + o(t^2)\\
  &= 1 - \frac{1}{2}t^2 + o(t^2)
  \quad \text{as $t \to 0$},
\end{align*}
where $o(t^2)$ means $o(t^2)/t^2 \to 0$ as $t\to 0$.
Substituting to write as a function of $t/\sqrt{n}$ gives
$$
  \varphi\!\left(\frac{t}{\sqrt{n}}\right) 
  = 1 - \frac{t^2}{2n} + o\!\left(\frac{t^2}{n}\right).
$$

Taking logs (which is valid for large enough\ $n$ since $\varphi(t/\sqrt{n}\,) \to 1$) and using
$\log(1 + u) = u + o(u)$ as $u \to 0$:
$$
  \log \varphi\!\left(\frac{t}{\sqrt{n}}\right) 
  = -\frac{t^2}{2n} + o\!\left(\frac{t^2}{n}\right).
$$
Therefore
\begin{align*}
  \log \varphi_{Z_n}(t)
  &= n \log \varphi\!\left(\frac{t}{\sqrt{n}}\right)\\
  &= n\left[-\frac{t^2}{2n} + o\!\left(\frac{t^2}{n}\right)\right]\\
  &= -\frac{t^2}{2} + n\cdot o\left(\frac{t^2}{n}\right)
  \to\, -\frac{t^2}{2}
\end{align*}
as $n \to \infty$, for each fixed $t \in \mathbb{R}$. 
Taking exponents,
$$
  \varphi_{Z_n}(t) \to \exp(-t^2/2) \quad \text{for all $t \in \mathbb{R}$}.
$$

The function $\exp(-t^2/2)$ is the CF of the $N(0, 1)$ distribution.
Since $\varphi_{Z_n}(t) \to \varphi(t)$ pointwise for all\ $t$, and the limit $\varphi(t)$ is continuous at $t = 0$, then $Z_n \overset{d}{\to} Z$, where\ $Z$ has characteristic function $\varphi(t)$, and $\overset{d}{\to}$ means 'converges in distribution'.
Since $\exp(-t^2/2)$ is continuous everywhere and is the CF of $N(0, 1)$, conclude that
$$
  Z_n \overset{d}{\to} N(0, 1).
$$

This is a partial proof since the concept of convergence in distribution has not been formally defined. 
However, the arguments used in the proof are powerful and worth understanding.



## Exercises

Selected answers appear in Sect.\ \@ref(AnswersChapSampling).

:::{.exercise #FPCF}
When sampling *without* replacement from a finite population of size\ $N$, observations are no longer independent.
For this reason, when sampling without replacement the standard error is multiplied by the *finite population correction factor*
$$
  \sqrt{\frac{N - n}{N - 1}}
$$
where\ $n$ is the sample size.

1. What happens if the sample size is the same as the population size?
   Explain why this is a sensible result.
2. Show that, for large\ $N$, the FPCF is approximately $\sqrt{ 1 - n/N}$.
3. For what ratio $n/N$ (the *sampling fraction*)\index{Sampling fraction} do you think that using the FPCF is unnecessary in practice? 
:::


:::{.exercise}
A random sample of size\ $81$ is taken from a population with mean\ $128$ and standard deviation\ $6.3$. 

1. What is the probability that an *individual observation* will fall between\ $126.6$ and\ $129.4$?
1. What is the probability that the *sample mean* will fall between\ $126.6$ and\ $129.4$?
1. What is the probability that the *sample mean* will *not* fall between\ $126.6$ and\ $129.4$?
:::


:::{.exercise #CLTProb}
Let $Y_1$, $Y_2$, $\dots$, $Y_n$ be $n$\ independent random variables, each with PDF
$$
   f_Y(y) = \begin{cases}
               3y^2 & \text{for $0\le y \le 1$};\\
               0 & \text{otherwise}.
            \end{cases}
$$


1. Determine the probability that a *single observation* will be within one standard deviation of the population mean.
1. Determine the probability that the *sample mean* will be within one standard deviation of the population mean, using the [*Central Limit Theorem*](#CentralLimitTheorem).
:::


:::{.exercise #EggsVariance}
Suppose the weights of eggs in a carton (of twelve eggs) have a weight that is normally distributed with mean\ $59\gs$ and variance\ $0.7\gs$.

1. Find the probability that, in a sample of $20$\ cartons, the sample mean weight will exceed $59.5\gs$. 
1. Find the probability that a sample of twelve eggs will produce a sample variance of greater than\ $1$.
:::


:::{.exercise #EggsBroken}
In a carton of a twelve eggs, the number broken has a Poisson distribution with mean\ $0.2$. 

1. Find the probability that, in a sample of $20$\ cartons, the sample mean of the number of broken eggs per carton is more than one.
  (Use the *Central Limit Theorem*.)
2. Find the probability that, in any *single* carton, the probability that more than one egg is broken.
:::


:::{.exercise #ContRVM}
The random variable\ $M$ has the following probability density function
$$
    f_M(m) = 
    \begin{cases}
       3m^2 & \text{for $0 < m < 1$};\\
       0    & \text{otherwise}.
    \end{cases}
$$
A random sample of size $n = 9$ is taken from the distribution, and the sample mean\ $\overline{M}$ is computed.

1. Compute the mean of\ $M$.
2. Compute the variance of\ $M$.
3. State the approximate distribution of\ $\overline{M}$ including the values of the parameters.
4. Compute the probability that the sample mean will be within\ $0.1$ of the true mean.
:::


:::{.exercise #RandomNormalSize9}
Generate a random sample of size $n = 9$ from a $N(10, 36)$ distribution hundreds of times.
Obtain the mean and variance of\ $\sum_i X_i$ and\ $\overline{X}$ for each sample.

1. Verify that the distribution of the sample means is approximately normally distributed as expected.
1. Verify that the distribution of the sum is approximately normally distributed as expected.
1. Explain *why* these are expected.
:::



:::{.exercise #Pollution}
A manufacturing plant produces $2$\ tonnes of waste product on a given day, with a standard deviation of $0.2$\ tonnes per day. 
Find the probability that, over a $20$\ day period, the plant produces less than $25$\ tonnes of waste if daily productions can be assumed independent.
:::


:::{.exercise}
Let\ $Y$ be the *change* in depth of a river from one day to the next measured (in cms) at a specific location. 
Assume\ $Y$ is uniformly distributed for $y \in [-70, 70]$.

1. Find the probability that the mean change in depth over a period of $30$\ days will be greater than $10\cms$.
1. Use simulation to estimate the probability above.
:::


:::{.exercise #Typhoid}
The number of deaths per year due to typhoid fever is assumed to have a Poisson distribution with rate $\lambda = 4.1$ per year.

1. If deaths from year to year can be assumed to be independent, what is the distribution over a $20$\ year period?
1. Find the probability that there will be more than $100$\ deaths due to typhoid fever in period of $20$\ years.
:::


:::{.exercise #Lymphocyte}
The probability that a cell is a lymphocyte is\ $0.2$.

1. Write down an exact expression for the probability that in a sample containing $150$\ cells that at least\ $40$ are lymphocytes.
   Evaluate this expression using **R**.
1. Write down an approximate expression for this probability, and evaluate it.
:::


:::{.exercise}
Suppose the probability of a person aged $80+$ years dying after receiving influenza vaccine is\ $0.006$. 
In a sample of $200$\ persons aged $80+$\ years:

1. Write down an exact expression for the probability that more than\ $5$ will die after vaccination for influenza. 
   Evaluate this expression using **R**.
1. Write down an approximate expression for this probability and evaluate it.
1. If $4$\ persons died in a sample of\ $200$, what conclusion would you make about the probability of dying after vaccination?
   Justify your answer.
:::


:::{.exercise}
Illustrate the Central Limit Theorem for the uniform distribution on $[-1, 1]$ by simulation and repeated sampling, for various sample sizes.
:::


:::{.exercise}
Illustrate the Central Limit Theorem for the exponential distribution with mean\ $1$, by simulation and repeated sampling, for various sample sizes.
:::


:::{.exercise #SimTdist}
Consider a random variable\ $Z$ with a standard normal distribution $N(0, 1)$, and a random variable\ $V$ with a chi-squared distribution on\ $v$ degrees of freedom.

Simulate the distribution of
$$
  T = \frac{Z}{\sqrt{V/\nu}},
$$
then show this is a $t$-distribution with\ $\nu$ degrees of freedom.
*Hint*: use the **R** functions for the $t$-distribution (e.g., `dt()`) and the chi-squared distribution (e.g, `dchisq()`). 
:::


:::{.exercise #SimHurdle}
Consider a zero-modified\index{Zero-modified distribution!Gamma modified} Gamma$(2,1)$-distribution, with $p = 0.5$.

1. Derive $\operatorname{E}[\overline{Y}_n]$ and $\operatorname{var}(\overline{Y}_n)$.
1. Simulate the distribution of $\operatorname{E}[\overline{Y}_n]$ for $n \in \{10, 50, 200\}$, overlayed with the appropriate normal approximation.
   Comment.
:::


:::{.exercise #SimCPG}
Consider a compound Poisson--gamma distributions\index{Compound Poisson--gamma distribution} with $\lambda = 1.5$, $\alpha = 2$ and $\beta = 1$.

1. Derive $\operatorname{E}[\overline{Y}_n]$ and $\operatorname{var}(\overline{Y}_n)$.
1. Simulate the distribution of $\operatorname{E}[\overline{Y}_n]$ for $n \in \{10, 50, 200\}$, overlayed with the appropriate normal  approximation.
   Comment.
:::


:::{.exercise #Unbiased1}
Let $X_1, \ldots, X_n \overset{\text{iid}}{\sim} \operatorname{Exponential}(\lambda)$, where\ $\lambda$ is the rate parameter so that $\operatorname{E}[X_i] = 1/\lambda$.

1. Show that $\overline{X}$ is a biased estimator of $1/\lambda$.
2. Find an unbiased estimator of\ $\lambda$.
:::


:::{.exercise #Unbiased2}
Let $X_1, \ldots, X_n \overset{\text{iid}}{\sim} \operatorname{Uniform}(0, \theta)$. 

1. Show that $2\overline{X}$ is an unbiased estimator of\ $\theta$.
2. Let $M = \max(X_1, \ldots, X_n)$. 
   Derive the distribution of $M$ and show that\ $M$ is *not* unbiased for $\theta$.
3. Find a *function* of\ $M$ that is unbiased for\ $\theta$.
:::


:::{.exercise #UnbiasedVarianceMu}
Show that the estimator 
$$
  S^2_\mu = \frac{1}{n} \sum_{i=1}^n (X_i - \mu)^2.
$$
is an unbiased estimator of the population variance\ $\sigma^2$.
:::


:::{.exercise #UnbiasedVariance}
Show that the estimator 
$$
  S^2 = \frac{1}{n - 1} \sum_{i=1}^n (X_i - \overline{X})^2.
$$
is an unbiased estimator of the population variance\ $\sigma^2$.
:::


:::{.exercise #SumMean}
Prove Theorem\ \@ref(thm:SumMean).
:::


:::{.exercise #DistZ2}
Consider a random variable $X\sim N(0, 1)$.
Show that the distribution of $Z = X^2$ is a $\text{Gamma}(\alpha = 1/2, \beta = 2)$ distribution.
:::


:::{.exercise #SumZ2}
Consider $Z_i \overset{\text{iid}}{\sim}\sim \text{Gamma}(\alpha=1/2, \beta2)$ for $i = 1, 2, \dots, n$.

Use the MGF method (Sect.\ \@ref(TransformationMoments)) to show that the distribution of $\sum_{i=1}^n Z_i^2$ is a $\text{Gamma}(\alpha = n/2, \beta = 2)$ distribution.
:::


:::{.exercise #MeanSdIndependent}
Let $X_1, X_2,\dots, X_n$ be a random sample of size\ $n$ from $N(\mu, \sigma^2)$.
By Theorem\ \@ref(thm:\@ref(thm:MeanVarInd)), the sample mean $\overline{X}$ and the sample variance\ $S^2$ are independent.

To demonstrate using **R**, generate\ $1000$ simulations of data\ $X_i \sim N(1, 3)$ for $i = 1, 2, \dots n$ for $n = 50$.
For each of the $1\,000$\ simulations, compute the sample mean and standard deviation.

1. Plot the sample means against the sample variances.
   Comment.
2. Compute the correlation between the sample means against the sample variances.
   Comment.
3. Show that the covariance between the sample mean $\overline{X}$ and the deviations of the observations from the sample means $\overline{X} - X_i$ is zero.
:::


:::{.exercise #ChisqProb}
The variable\ $X$ has a chi-squared distribution with $12$\ df.
Determine the value of\ $X$ below which lies 90% of the distribution.
:::


:::{.exercise #ChisqProb2}
A random sample of size $n = 10$ is selected from the $N(20, 5)$ distribution.
What is the probability that the sample variance exceeds\ $10$?
:::


:::{.exercise #ExSamplingDist}
A population consists of the five values $\{1, 3, 5, 7, 9\}$.

1. Compute the population mean\ $\mu$ and population variance\ $\sigma^2$.
2. List all possible samples of size $n = 2$ (without replacement) and compute the sample mean\ $\overline{X}$ for each.
3. Find the mean and variance of the sampling distribution of\ $\overline{X}$.
4. Verify that $\operatorname{E}[\overline{X}] = \mu$.
5. How does the variance of $\overline{X}$ relate to\ $\sigma^2$ and\ $n$?
:::


:::{.exercise #ExUnbiasedMean}

Let $X_1, \ldots, X_n \overset{\text{iid}}{\sim} \operatorname{Poisson}(\lambda = 3)$.

1. State $\operatorname{E}[X_i]$ and $\operatorname{var}[X_i]$.
2. Show algebraically that $\overline{X}$ is an unbiased estimator of\ $\lambda$.
3. Using **R**, simulate $5\,000$ samples of size $n = 20$ from a $\operatorname{Poisson}(3)$ distribution. 
   Compute the sample mean for each and verify empirically that the average of the sample means is close to\ $3$.
4. Plot a histogram of the sample means. Comment on the shape.
:::


:::{.exercise #ExBiasedVariance}
Using **R**, simulate $5\,000$ samples of size $n = 5$ from a $N(\mu = 10, \sigma^2 = 4)$ distribution.

1. For each sample, compute both the biased estimator 
  $$
    S^2_n = \frac{1}{n}\sum_{i=1}^n (X_i - \overline{X})^2
  $$
  and the unbiased estimator
  $$
    S^2 = \frac{1}{n-1}\sum_{i=1}^n (X_i - \overline{X})^2.
  $$
2. Compute the average value of\ $S^2_n$ and\ $S^2$ across all simulations.
3. Compare both averages to the true population variance $\sigma^2 = 4$. 
   Which estimator is unbiased?
4. Repeat for $n = 50$. 
   Comment on how bias changes with sample size.
:::


:::{.exercise #ExSampMeanVar}
Suppose $X \sim \operatorname{Exponential}(\lambda = 0.5)$, so that $\operatorname{E}[X] = 2$ and $\operatorname{var}[X] = 4$.

1. State the theoretical mean and variance of\ $\overline{X}$ for samples of size $n = 16$.
2. Using **R**, simulate $5\,000$ sample means from samples of size $n = 16$ drawn from $\operatorname{Exponential}(0.5)$.
3. Compute the empirical mean and variance of these sample means and compare to the theoretical values.
4. Repeat for $n = 64$. 
   Comment on how sample size affects the variance of\ $\overline{X}$.
:::


:::{.exercise #ExCLTShape}
**Shape of the sampling distribution: Central Limit Theorem**

Let $X \sim \operatorname{Uniform}(0, 1)$, so that $\operatorname{E}[X] = 0.5$ and $\operatorname{var}[X] = 1/12$.

1. For sample sizes $n \in \{1, 5, 30, 100\}$, simulate $5\,000$ sample means using **R**.
2. Plot histograms of the sample means for each sample size, in a $2 \times 2$ panel.
3. Overlay each histogram with the theoretical normal distribution $N(0.5,\; 1/(12n))$.
4. Comment on how quickly the sampling distribution approaches normality as\ $n$ increases.
:::


:::{.exercise #ExCLTPoisson}
The number of customers arriving at a service desk per hour follows a $\operatorname{Poisson}(\lambda = 4)$ distribution.

1. State the mean and variance of the total number of customers over $n = 50\hs$.
2. Using the CLT, find the approximate probability that the total number of customers over $50\hs$ exceeds\ $215$.
3. Verify your answer in **R** using `ppois()` for the exact probability and compare.
:::


:::{.exercise #ExNormalSampMean}
Weights of a particular product are normally distributed with mean $500\gs$ and standard deviation $8\gs$.

1. Find the probability that a single item weighs less than $490\gs$.
2. Find the probability that the mean weight of a random sample of $n = 25$ items is less than $490\gs$.
3. Find the probability that the mean weight of a random sample of $n = 64$ items is less than $490\gs$.
4. Explain why the probabilities in Parts 1–-3 decrease as the sample size increases.
:::


:::{.exercise #ExLinearComb}
Let $X_1 \sim N(5, 9)$ and $X_2 \sim N(2, 4)$ be independent.
Define $Y = 3X_1 - 2X_2$.

1. Find $\operatorname{E}[Y]$ and $\operatorname{var}[Y]$.
2. State the distribution of\ $Y$.
3. Find $\Pr(Y > 20)$.
4. Verify using **R** simulation: generate $5\,000$ realisations of $Y$ and estimate $\Pr(Y > 20)$.
:::


:::{.exercise #ExChiSq}
Let $X \sim \chi^2_{10}$.

1. State the mean and variance of\ $X$.
2. Find $\Pr(X < 15.99)$ using **R**.
3. Find the value $x$ such that $\Pr(X > x) = 0.05$ using **R**.
4. Simulate $5\,000$ values from $\chi^2_{10}$ and plot a histogram. 
   Overlay the theoretical density curve.
:::


:::{.exercise #ExChiSqFromNormals}

1. Show by simulation that if $Z_1, Z_2, \ldots, Z_5 \overset{\text{iid}}{\sim} N(0,1)$, then $\sum_{i=1}^5 Z_i^2 \sim \chi^2_5$.
   Use $5\,000$ simulations, plot the histogram, and overlay the $\chi^2_5$ density.
2. Compute the empirical mean and variance of $\sum Z_i^2$ and compare to the theoretical values.
:::



:::{.exercise #ExSampVarChi}
Suppose $X_1, \ldots, X_{15} \overset{\text{iid}}{\sim} N(\mu = 20, \sigma^2 = 9)$.

1. State the distribution of $(n-1)S^2/\sigma^2$.
2. Find the probability that $S^2 > 14$ using **R**.
3. Find the value\ $s^2$ such that $\Pr(S^2 > s^2) = 0.10$.
4. Verify Part\ 2 by simulation: generate $5\,000$ samples of size\ $15$ and compute the proportion with $S^2 > 14$.
:::


:::{.exercise #ExTDist}
Let $T \sim t_8$.

1. State the mean and variance of\ $T$.
2. Find $\Pr(T < 1.86)$ using **R**.
3. Find the value\ $t$ such that $\Pr(|T| > t) = 0.05$ using **R**.
4. Show by simulation that $T = Z / \sqrt{V/8}$ has a $t_8$ distribution, where $Z \sim N(0,1)$ and $V \sim \chi^2_8$ independently. Use $10\,000$ simulations, plot the histogram, and overlay the $t_8$ density.
:::



:::{.exercise #ExTStatistic}
A sample of $n = 10$ observations is drawn from a normal population with unknown mean and variance.
The sample values are:
$$14.2,\ 15.8,\ 13.5,\ 16.1,\ 14.9,\ 15.3,\ 13.8,\ 16.4,\ 14.6,\ 15.0.$$

1. Compute $\overline{x}$ and $s^2$ from the data in **R**.
2. Compute the $T$-statistic assuming the population mean is $\mu_0 = 15$.
3. Find the probability of observing a $T$-statistic as extreme or more extreme than this value (two-sided) under\ $t_9$.
4. Comment on whether the data are consistent with $\mu = 15$.
:::


:::{.exercise #ExFDist}
Let $X \sim F_{4, 20}$.

1. State the mean and variance of\ $X$.
2. Find $\Pr(X < 2.87)$ using **R**.
3. Find the value $x$ such that $\Pr(X > x) = 0.01$.
4. Show by simulation that the ratio $\frac{U_1/4}{U_2/20}$, where $U_1 \sim \chi^2_4$ and $U_2 \sim \chi^2_{20}$ independently, follows an $F_{4,20}$ distribution. 
   Use $10\,000$ simulations.
:::


:::{.exercise #ExTSquaredF}

1. State the result connecting $T \sim t_n$ to the $F$-distribution.
2. Let $T \sim t_{12}$. 
   Find $\Pr(T^2 > 3)$ using the $t$-distribution in **R**.
3. Verify this answer by instead computing the equivalent probability using the $F$-distribution in **R**.
4. Confirm the equivalence by simulation: generate $10\,000$ values of $T \sim t_{12}$ and $50{,}000$ values of $W \sim F_{1,12}$ and compare histograms of\ $T^2$ and\ $W$.
:::


:::{.exercise #ExNormalApproxBinom}
In a large population, $35\%$ of people have a particular characteristic.
A random sample of $n = 120$ people is taken.

1. Let\ $Y$ be the number of people in the sample with the characteristic. 
   State the exact distribution of\ $Y$.
2. Using the CLT, state the approximate distribution of\ $Y$.
3. Use the normal approximation to find $\Pr(Y \geq 50)$.
4. Compute the exact probability using **R** and compare to the approximation.
:::


:::{.exercise #ExMeanVarIndep}
Using **R**, simulate $5\,000$ samples of size $n = 30$ from $N(0, 1)$.
For each sample, record the sample mean\ $\overline{X}$ and sample variance\ $S^2$.

1. Produce a scatterplot of $\overline{X}$ against $S^2$.
2. Compute the correlation between $\overline{X}$ and $S^2$.
3. Comment on what the plot and correlation suggest about the relationship between the sample mean and sample variance when sampling from a normal distribution.
:::


:::{.exercise #VarianceDecomposition}
Show that
$$
  \sum_{i=1}^n (X_i - \overline{X})^2
  =
  \sum_{i=1}^n (X_i - \mu)^2
  -
  n(\overline{X}-\mu)^2.
$$
Start by writing
$$
  X_i - \overline{X}
  =
  (X_i - \mu) - (\overline{X} - \mu),
$$
then expand the square and simplify the resulting sums.
:::

