# (PART) Sampling distributions {-}

# Describing samples {#ChapDescribingSamples}

::: {.objectivesBox .objectives data-latex="{iconmonstr-target-4-240.png}"}

* use appropriate techniques to determine the sampling distributions of\ $t$, $F$, and\ $\chi^2$\ distributions.
* explain how the above distributions are related to the normal distribution.
* apply the concepts of the Central Limit Theorem in appropriate circumstances.
* use the Central Limit Theorem to approximate binomial probabilities by normal probabilities in appropriate circumstances.
:::


## Populations and samples {#PopulationsSamples}

This chapter introduces the application of distribution theory to the discipline of statistics.
Distribution theory is about describing random variables in theoretical populations using probability, while statistics is about data collection and extracting information from observed data (Chap.\ \@ref(ChapterIntroduction)). 


### From populations to observations {#Sampling}

\index{Sample|(}
Until now, theoretical probability distributions describing ideal distributions of infinite populations have been our focus.
In practice, of course, ideal distributions and infinite populations do not exist, and finite (usually relatively small) samples from unknown distributions are observed.\index{Sample}


:::{.example #Heights name="Populations and models"}
Suppose our *population* of interest is Australian adult females, and we are interested in their heights.
'Height of Australian adult females' is a continuous random variable, say\ $X$.

We may decide to model this population of heights theoretically using some specified distribution, with the probability density function $f_X(x)$.
Using this model, the mean height of Australian adult females is modelled as (Def.\ 
\@ref(def:Expectation))
$$
  \mu = \operatorname{E}[X] = \int_{-\infty}^\infty x\cdot f_X(x)\,dx.
$$

If we model their heights using the normal distribution $N(\mu = 173, \sigma^2 = 9)$, the mean is modelled as\ $173\cms$ and the variance as $9\cms^2$.
:::


While height is a continuous random variable, and heights can be modelled using a continuous probability distribution, In practice height measurements of individual Australian females are *not* made with infinite precision; measured heights must be rounded as hence recorded as discrete values.
Any tool to measure actual heights must record a *rounded* version of the heights, to some minimum unit\ $\delta$.
Heights, for example, may be rounded to the nearest centimetre (i.e., $\delta = 1\cms$) or the nearest millimetre (i.e., $\delta = 0.1\cms$).
So even if it was possible to measure the height of every Australian adult female, we would in practice have a list of discrete values $\{x_1, x_2, \dots , x_N\}$, where\ $N$ is the size of the population.

So, the observed heights are actually discrete (rounded) values, and the population mean of the measured heights would be computed (from Def.\ \@ref(def:Expectation)) using
$$
  \mu 
  = \operatorname{E}[X]
  = \frac{1}{N} \sum_{i = 1}^N x_i.
$$
As $N \to \infty$ and as $\delta \to 0$, this discrete sum converges to the Riemann integral:
$$
  \mu
  =
  \lim_{\substack{\delta \to 0 \\ N \to \infty}} \frac{1}{N} \sum_{i = 1}^N x_i 
  = 
  \int_{-\infty}^\infty x\cdot f_X(x)\,dx.
$$


### Random sampling and statistics {#RandomSamples}

In practice, entire populations are very rarely studied: for example, the population are usually too large (and so studying the whole population is prohibitively expensive and time consuming), and some members of population may be inaccessible.
So rather than studying entire populations, a *sample* of observations from that population is almost always studied.

The task of the *inferential statistics*\index{Inferential statistics}\index{Statistic} is to attempt to understand the entire population, based on what is learnt from studying a small subset of that population: the *sample*. \index{Sample}
Inferential statistics requires an understanding of how samples behave.


:::{.definition #RandomSample name="Sample"}
A subset of observations from the population is called a sample.\index{Sample}
:::


This definition is very broad: *any* subset from the population is a *sample*.
Importantly, *how* the sample is drawn determines its statistical properties; the most important type of sample for statistical inference is the idea of a *random sample*.\index{Sample!random}


:::{.definition #RandomSample name="Random sample"}
The set of random variables $X_1, X_2, \dots, X_n$ from a (usually unspecified) distribution\ $D$ is a collection of random variables $X_1, X_2, \dots, X_n$ that are independently and identically distributed (iid)\index{iid} with distribution function $F_X(x)$.

We write $X_1, X_2, \dots, X_n \overset{\text{iid}}{\sim} D$, where 'iid'\index{iid} means 'independently and identically distributed'.
:::


The definition uses the distribution function and so applies to discrete, continuous and mixed random variables.
Because the elements of the random sample are assumed to be independent,\index{Independence} the results in previous chapters that are based on independence therefore apply.

Once the sample is obtained, the data can be used to compute the value of some *statistic*.
A *statistic*\index{Statistic} is simply a numerical value that is computed from a sample.


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
Since the individuals chosen to be in the sample can vary from saple to sampple, the *value* of a statistic can vary from sample to sample.
:::


:::{.definition #Statistic name="Statistic"}
Let $X_1, X_2, \ldots, X_n$ be a random sample of size\ $n$.
A *statistic* is any function $T = g(X_1, X_2, \ldots, X_n)$ of the sample.\index{Statistic}
:::


:::{.example #StatisticdsExample name="Statistic"}
For a random sample $X_1, X_2, \dots, X_n$, we could define a statistic\ $R$ such that
$$
  R = \max\{X_1, X_2, \dots, X_n\} - \min\{X_1, X_2, \dots, X_n\}.
$$
:::


An *estimator*\index{Estimator} is a *statistic* with a specific purpose: to estimate the unknown value of a population parameter.
As a statistic, an estimator is a random variable.
An *estimate*\index{Estimate} is the actual numerical value of the estimator that is computed after observing data.
We often say that an estimate is a *realisation* of an estimator.

The study of the properties of estimators---and what properties of estimators are desirable---is a branch of statistics called *statistical inference*.


:::{.example #StatisticsExampleMin name="The sample median is a statistic"}
Suppose we take a sample of $n = 12$ Australian adult females, and measure their heights to the nearest centimetre.

We could define the *statistic*\index{Statistic}
$$
  M = \text{median}(X_1, X_2, \dots, X_N).
$$
This statistic computes the median value of the sample.
If the purpose of computing the sample median is to estimate the median value in the population, then $M$ is an *estimator* of the population median.
While\ $M$ has been proposed as an estimator of the population median, we have not studied whether it is a 'good' estimator of the population median (i.e., has useful properties).

Since\ $M$ is likely to be different when computed from another sample of $12$\ Australian adult females, it is a random variable. 
If the median height in the sample is $169\cms$, then
$$
  m = 169
$$
is an *estimate* of the unknown population median.
:::
\index{Sample|)}


## Sampling distributions {#SamplingDistributions}

### Sampling distributions: a small example {#SamplingDistributionsSmallExample}

\index{Sampling distribution|(}
Since a statistic is a function of the random variables $X_1, X_2, \ldots, X_n$, *sample statistics are random variables*, and so the value of a sample statistic changes from sample to sample. 
This means that *sample statistics have a distribution* that describes *how* they vary.


:::{.definition #SamplingDistributionStatistic name="Sampling distribution of a statistic"}
The *sampling distribution of a statistic* is the theoretical probability distribution of the statistic associated with *all* possible samples of a particular size drawn from a particular population.
:::


Typically, a population will be *very* large, and listing all elements of the population is impossible.
However, for demonstration purposes, suppose the population contains only $N = 5$ discrete elements:
$$
   22\qquad 23\qquad 27\qquad 28\qquad 38.
$$
We can define the sample median as a statistic, and then the median value in the population is\ $27$.
This value is called a *parameter*\index{Parameter} since it is a number that describes the population.

The median value of any sample of size $n < 5$ drawn from the population can be found: the sample median is a *statistic*, as it is computed from the sample.
Since the population is (artificially) small in this scenario, the sampling distribution can be determined by enumerating every possible sample of any given size\ $n$.

Using **R**:

```{r}
Population <- c(22, 23, 27, 28, 38)
median_P <- median(Population)
median_P
```

Suppose we consider samples of size $n = 2$; there are $\binom{5}{2} = 10$ such samples:

```{r}
combn(Population, 2)
```

The median of each sample is easily computed also:

```{r}
median_S <- combn(Population, 2, FUN = median)
median_S
```

The *sampling distribution* is just the distribution of these sample medians (which are the sample medians from all possible samples), which can be displayed a table (or graph):

```{r}
table(median_S)
```

Since the population median is\ $27$, the sample medians are often a poor estimate of the population median.
Suppose we looked at larger sample size; say, samples of size $n = 4$:

```{r}
median_S_2 <- combn(Population, 4, FUN = median)
table(median_S_2)
```

The population median is\ $27$, so the sample median when $n = 4$ is often a very good estimate of the population median.
The larger sample size produces a better estimate, in general, and the value of the sample statistic varies less from sample to sample.


### Larger populations {#SamplingDistributionsLargerPopulations}

For larger populations, enumerating every possible is sample implausible or impossible, and obtaining the selected elements is usually impractical or impossible.
Nonetheless, the idea of a sampling distribution of a statistic can be *simulated* using\ **R**.
These sampling distributions, generated from simulated data, are called *empirical distributions*;\index{Empirical distributions} that is, they are based on sample data not theoretical ideals.


:::{.example}
Consider taking a random sample of $n = 5$ observations from a $N(\mu = 173, \sigma^2 = 9)$ distribution, and computing the median value in the sample:

```{r echo=FALSE}
set.seed(1592875)
```
```{r}
x_Mn <- 173; x_Sd <- 3        # Population (model) mean and std dev.
sample_Size <- 5              # Sample size is 5
x1 <- rnorm(sample_Size,      # Take a sample of this size...
            mean = x_Mn, sd = x_Sd)     # ... from this distribution
median(x1)                    # Compute the sample median for this sample
```

We can use\ **R** to simulate selecting many samples of size $n=5$, and compute the median for each (Fig.\ \@ref(fig:SDmedianPlots), left panel):

```{r SDmedian1, fig.show='hide'}
num_Sims <- 100000           # Number of simulations

# Generate random values
Xvals <- rnorm(num_Sims * sample_Size,
               mean = x_Mn, sd = x_Sd)

# Place these into an array, with each sample as a column 
Xvals <- array(Xvals,
               dim = c(sample_Size, num_Sims) )

# Find the median of each sample (i.e., column)
Xmed   <- apply(Xvals, 
                MARGIN = 2, 
                FUN = median) 
hist(Xmed)          # Create the histogram of sample medians
```

Suppose we take larger samples, and find the sampling distribution again (Fig.\ \@ref(fig:SDmedianPlots), right panel):

```{r SDmin2, fig.show='hide'}
sample_Size2 <- 100

# Generate random values
Xvals_2 <- rnorm(num_Sims * sample_Size2,
                 mean = x_Mn, sd = x_Sd)

# Place these into an array, with each sample as a column 
Xvals_2 <- array(Xvals_2,
                 dim = c(sample_Size2, num_Sims) )

# Find the minimum of each sample (i.e., column)
Xmed_2   <- apply(Xvals_2, 
                  MARGIN = 2,       # MARGIN = 2 means the columns
                  FUN = median)     # Find the median of each column
hist(Xmed_2)                          # Create the histogram of sample medians
```

:::

```{r SDmedianPlots, echo=FALSE, out.width='90%', fig.width=8.5, fig.height = 3.5, fig.align="center", fig.cap="Sampling distributions for the sample medians, for samples of size\ $5$ (left) and size\ $100$ (right)."}
par( mfrow = c(1, 2))

set.seed(1592875)

hist(Xmed, 
     xlim = c(165, 185),
     xlab = "Sample median",
     ylab = "",
     breaks = seq(165, 185, by = 1),
     main = "Sampling medians:\nsamples of size 5",
     axes = FALSE)
axis(side = 1)

#####

hist(Xmed_2, 
     xlim = c(165, 185),
     xlab = "Sample median",
     ylab = "",
     breaks = seq(165, 185, by = 0.5),
     main = "Sampling medians:\nsamples of size 100",
     axes = FALSE)
axis(side = 1)
```

These histograms display the sampling distributions of the sample median for samples of size $n = 5$ and $n = 100$.
The value of the sample statistic varies less from sample to sample again for the larger sample size.


## Estimating population parameters

### Estimating the population mean {#SampligDistributionMean}

The most common parameter estimated from a sample is the population mean\ $\mu$.
For a population of\ $N$ discretely-measured observed values $X_1, X_2, \dots, X_n$, the population mean is
$$
  \mu = \operatorname{E}[X] = \frac{1}{N}\sum_{i = 1}^N x_i
$$
as noted earlier.
The obvious suggestion for computing a sample mean from a sample of size\ $n$ (i.e., $x_1, x_2, \dots, x_n$) is *sample mean* $\overline{X}$, an *estimator*,\index{Estimator} defined as
$$
  \overline{X} = \frac{1}{n}\sum_{i = 1}^n X_i.
$$

We can use\ **R** to simulate the selection of many samples of size $n = 5$, and many sample of size $n = 100$, and compute the empirical mean\index{Empirical mean} for each sample, using the same samples created above (Fig.\ \@ref(fig:SDmeanPlots)):

```{r SDmean, fig.show='hide'}
# Mean: n = 5
Xmean   <- apply(Xvals, 
                 MARGIN = 2, 
                 FUN = mean) 
hist(Xmean)          # Create the histogram of sample minimums

# Mean: n = 100
Xmean_2   <- apply(Xvals_2, 
                   MARGIN = 2, 
                   FUN = mean) 
hist(Xmean_2)          # Create the histogram of sample minimums
```


```{r SDmeanPlots, echo=FALSE, out.width='90%', fig.width=8.5, fig.height = 3.5, fig.align="center", fig.cap="Sampling distributions for the sample mean for sample of size $n = 5$ (left) and $n = 100$ (right). The solid dots represent the population mean."}
par( mfrow = c(1, 2))

hist(Xmean, 
     xlim = c(165, 180),
     xlab = "Sample mean",
     ylab = "",
     breaks = seq(150, 180, by = 1),
     main = "Sampling mean:\nsamples of size 5",
     sub = paste0("(Average value: ", round(mean(Xmean), 1), ")"),
     axes = FALSE)
axis(side = 1)
points(x = x_Mn, 
       y = 0,
       pch = 19)

#####

hist(Xmean_2, 
     xlim = c(165, 180),
     xlab = "Sample mean",
     ylab = "",
     breaks = seq(150, 180, by = 0.5),
     main = "Sampling mean:\nsamples of size 100",
     sub = paste0("(Average value: ", round(mean(Xmean_2), 1), ")"),
     axes = FALSE)
axis(side = 1)
points(x = x_Mn, 
       y = 0,
       pch = 19)


```

The sample estimates of the mean appear, on average, to be good estimates of the population mean.
While some sample means are smaller than\ $\mu$, and some are larger than\ $\mu$, the estimate vary about\ $\mu$.
The larger sample produces better estimates of the population mean in general, and less variation in the values of the sample statistic.


### Unbiased estimators {#UnbiasedEstimators}

On average, the values of the sample means are centred around\ $\mu$.
It seems useful to use an estimator where, on average, the sample estimate would produce the population value.
Sometimes the sample estimate may be a little smaller than the population value, and sometimes a bit larger, but on average the value of the sample estimate is the same value of the population parameter.
Any estimator with this property is called an *unbiased estimator*.


:::{.definition #Unbiased name="Unbiased estimator"}
A statistic\ $\hat{\theta}$ is an *unbiased estimator*\index{Estimator!unbiased} of a parameter\ $\theta$ if
$$
  \operatorname{E}[\hat{\theta}] = \theta.
$$
:::


This means that an unbiased estimator from a random sample produces estimates that, *on average*, have the same value as the population value being estimated.
This seems a sensible property for a estimator.
(There are other desirable properties of estimators too, we but do not explore these here.)\index{Inferential statistics}
When we say that\ $\overline{X}$ is an unbiased estimator of\ $\mu$ then, we mean that $\operatorname{E}[\overline{X}] = \mu$,



:::{.theorem #SampleMeanUnbiased name="An unbiased estimator for $\mu$"}
The sample mean
$$
  \overline{X} = \frac{1}{n} \sum_{i = 1}^n X_i
$$
is an unbiased estimator\index{Estimator!unbiased}\index{Estimator!sample mean} of the population mean\ $\mu$.

::::{.proof}
Using the definition of an unbiased estimator and the properties of expectations:
\begin{align*}
  \operatorname{E}[\overline{X}] 
  &= \operatorname{E}\left[ \frac{1}{n}(X_1 + X_2 + \cdots X_n) \right]\\
  &= \frac{1}{n} \operatorname{E}[ X_1 + X_2 + \cdots X_n ]\\
  &= \frac{1}{n} \left( \operatorname{E}[ X_1 ] + \operatorname{E}[X_2] + \cdots \operatorname{E}[X_n] \right) \\
  &= \frac{1}{n} (n\times \mu) = \mu.
\end{align*}
That is, the expected value of\ $\overline{X}$ is equal to\ $\mu$.
The estimator\ $\overline{X}$ is an *unbiased estimator* of\ $\mu$.
::::
:::


:::{.example #UnbiasedPoisson name="A unbiased estimate"}
Let $X_1, \ldots, X_n \overset{\text{iid}}{\sim} \operatorname{Poisson}(\lambda)$, and consider the estimate $\bar{X}$.
Because the expectation is a linear operator, and because $\operatorname{E}[X_i] = \lambda$, 
$$
  \operatorname{E}[\bar{X}] 
  = 
  \operatorname{E}\!\left[\frac{1}{n}\sum_{i=1}^n X_i\right] = \frac{1}{n}\sum_{i=1}^n \operatorname{E}[X_i] = \frac{1}{n} \cdot n\lambda = \lambda.
$$
Hence $\bar{X}$ is an unbiased estimator of\ $\lambda$ for a Poisson distribution.
:::

Not all estimators\index{Estimator} are unbiased estimators; even estimators that appear to be 'obvious' estimators may not be unbiased estimators.\index{Estimator!unbiased}


:::{.example #BiasedEstimate name="A biased estimate"}
Let $X_1, \ldots, X_n \overset{\text{iid}}{\sim} \operatorname{Exponential}(\lambda)$, where $\lambda > 0$ is the rate parameter, so that $\operatorname{E}[X_i] = 1/\lambda$.
Consider the natural estimator of $\lambda$: that is, $1/\bar{X}$. 

Since $X_i \sim \operatorname{Exponential}(\lambda)$, we have $\overline{X} = \frac{1}{n}\sum_{i=1}^n X_i$, and it is known that
$$
  \overline{X} \sim \operatorname{Gamma}(\alpha = n,\, \beta = 1/(n\lambda) ),
$$
where we use the shape--scale parameterisation, so $\operatorname{E}[\overline{X}] = 1/\lambda$ and $\operatorname{Var}(\overline{X}) = 1/(n\lambda^2)$.

$\operatorname{E}[1/\overline{X}]$ can be computed directly:
$$
  \operatorname{E}\!\left[\frac{1}{\overline{X}}\right]
  = \int_0^\infty \frac{1}{x} \cdot
    \frac{(n\lambda)^n}{\Gamma(n)}\,x^{n-1} \exp(-n\lambda x)\,dx
  = \frac{(n\lambda)^n}{\Gamma(n)}
    \int_0^\infty x^{n-2} \exp(-n\lambda x)\,dx.
$$
The integral is related to the probability density function of a gamma distribution (Def.\ \@ref(def:GammaDistribution)) with shape $n - 1$ and scale $1/(n\lambda)$, and so
$$
  \int_0^\infty x^{n - 2} \exp(-n\lambda x)\,dx
  = 
  \frac{\Gamma(n - 1)}{(n\lambda)^{n - 1}},
$$
valid for $n > 1$. 
Substituting then,
$$
  \operatorname{E}\!\left[\frac{1}{\overline{X}}\right]
  = \frac{(n\lambda)^n}{\Gamma(n)} \cdot \frac{\Gamma(n - 1)}{(n\lambda)^{n - 1}}
  = n\lambda \cdot \frac{\Gamma(n - 1)}{\Gamma(n)}
  = n\lambda \cdot \frac{1}{n-1}
  = \frac{n\lambda}{n - 1},
$$
using $\Gamma(n) = (n - 1)\,\Gamma(n-1)$. 
Hence
$$
  \operatorname{E}\!\left[\frac{1}{\overline{X}}\right]
  = \frac{n}{n - 1}\,\lambda
  = \lambda + \frac{\lambda}{n - 1},
$$
and so the bias is $\lambda/(n - 1) > 0$: the estimator $1/\bar{X}$ systematically *overestimates* the value of $\lambda$.

Thus, $1/\overline{X}$ is a *biased* estimator of $\lambda$.
Note also that the bias $\lambda/(n - 1) \to 0$ as $n \to \infty$, so $1/\bar{X}$ is *asymptotically unbiased*.\index{Unbiased estimator!asymptotically}
:::


### Estimating the population variance {#EstimatePopVariance}

\index{Variance!estimation}
In the same way that a population mean can be computed from discrete, observed values, the population variance for a population of\ $N$ discretely-measured observed values $X_1, X_2, \dots, X_n$ is
$$
  \operatorname{var}[X] = \sigma^2 = \frac{1}{N}\sum_{i = 1}^N (x_i - \mu)^2.
$$
This suggests the following estimate of the population variance.


:::{.theorem #SampleVarianceUnbiased name="An unbiased estimator for $\sigma^2$"}
The sample variance
\begin{equation}
  S^2_\mu = \frac{1}{n} \sum_{i=1}^n (X_i - \mu)^2
   (\#eq:S2WithPopMean)
\end{equation}
is an unbiased estimator\index{Estimator!unbiased} of the population variance\ $\sigma^2$.

::::{.proof}
See Exercise\ \@ref(exr:UnbiasedVarianceMu)
::::
:::


Again, using the samples found earlier, the empirical estimates\ $S^2_\mu$\index{Empirical variances} can be found for samples of size $n = 5$ and $n = 100$ (Fig.\ \@ref(fig:SampleVarMu)), where $\mu = 173$:


```{r, fig.show='hide'}
s2mu <- function(x) { (sum(x - x_Mn)^2) / length(x) }  # Function to find sS2_mu
Xvar   <- apply(Xvals,                                 # Apply fn to each col
                MARGIN = 2,
                FUN = s2mu)
hist(Xvar) 
mean(Xvar)

Xvar_2   <- apply(Xvals_2,
                  MARGIN = 2,
                  FUN = s2mu)
hist(Xvar_2) 
mean(Xvar_2)
```

```{r SampleVarMu, echo=FALSE, out.width='100%', fig.width=8, fig.height=3.5, fig.cap="Sampling distributions for the sample variance for sample of size $n = 5$ (left) and $n = 100$ (right), using the unbiased estimator $S^2_\\mu$. The solid dots represent the population variance."}
par( mfrow = c(1, 2))

hist(Xvar,
     xlab = expression(italic(x)),
     main = "Variance estimates\nn = 5",
     ylab = "",
     axes = FALSE) 
axis(side = 1)

points(x = x_Sd^2,
       y = 0,
       pch = 19)

hist(Xvar_2,
     xlab = expression(italic(x)),
     main = "Variance estimates\nn = 100",
     ylab = "",
     axes = FALSE) 
axis(side = 1)
points(x = x_Sd^2,
       y = 0,
       pch = 19)
```


Again the larger sample size produces estimated with less variation compared to the estimates from the smaller sample size.
The average value of the estiamted, fromn the simulations, appear close to\ $\sigma^2 = 9$.

While $S^2_\mu$ is an unbiased estimator of\ $\sigma^2$, Eq.\ \@ref(eq:S2WithPopMean) has a practical difficulty: its computation relies on *knowing* the value of the population mean\ $\mu$.
Knowing the value of\ $\mu$ but having to estimate the population variance is highly unlikely.
If the information is available to compute the value of\ $\mu$, then\ $\sigma^2$ could also be computed (and would not need to be estimated). 
Almost always, only the estimate\ $\overline{X}$ of the population mean\ $\mu$ would be known.

This suggests using the estimator
$$
  S^2_n = \frac{1}{n} \sum_{i=1}^n (X_i - \overline{X})^2.
$$
to estimate the population variance.
To determine if this is an unbiased estimator, proceed as usual:
\begin{align*}
    \operatorname{E}[S^2_n] 
  &= \operatorname{E}\left[ \frac{1}{n} \sum_{i=1}^n (X_i - \overline{X})^2 \right]\\
  &= \frac{1}{n}\operatorname{E}\left[ \sum_{i=1}^n \left(X^2_i - 2X_i\overline{X} + \overline{X}^2\right)\right]\\
  &= \frac{1}{n}\operatorname{E}\left[ \sum_{i=1}^n X^2_i - \overline{X}\sum_{i=1}^n 2X_i + \sum_{i=1}^n \overline{X}^2\right].
\end{align*}
Since $\overline{X} = \sum_i X_i/n$, then $\sum_i X_i = n \overline{X}$; hence
\begin{align*}
     \operatorname{E}[S^2_n] 
  &= \frac{1}{n}\operatorname{E}\left[ \sum_{i=1}^n X^2_i - 2n\overline{X}^2 + n \overline{X}^2\right]\\
  &= \frac{1}{n}\operatorname{E}\left[ \sum_{i=1}^n X^2_i - n \overline{X}^2 \right]\\
  &= \frac{1}{n}\left(\operatorname{E}\left[ \sum_{i=1}^n X^2_i\right] - n\operatorname{E}\left[ \overline{X}^2 \right]\right).
\end{align*}
This expression can be simplified by noting that
$$
  \operatorname{var}[\overline{X}]
  = \operatorname{E}[\overline{X}^2] - \operatorname{E}[\overline{X}]^2 
  = \operatorname{E}[\overline{X}^2] - \mu^2, 
$$
and so $\operatorname{E}[\overline{X}^2] = \operatorname{var}(\overline{X}) + \mu^2 = \sigma^2/n + \mu^2$.
Also, 
$$
  \operatorname{var}[X] 
  = \operatorname{E}[X^2] - \operatorname{E}[X]^2 
  = \operatorname{E}[X^2] - \mu^2,
$$
so that $\operatorname{E}[X^2] = \sigma^2 + \mu^2$.
Hence,
\begin{align*}
     \operatorname{E}[S^2_n] 
  &= \frac{1}{n}\left[ \sum_{i=1}^n (\sigma^2 + \mu^2) - n\left(\frac{\sigma^2}{n} + \mu^2\right)\right]\\
  &= \frac{1}{n}\left[ n\sigma^2 + n\mu^2 - \sigma^2 - n\mu^2\right]\\
  &= \frac{n - 1}{n} \sigma^2.
\end{align*}
This shows that $\operatorname{E}[S^2_n] \ne\sigma^2$, and so the estimator is a *biased estimator* of\ $\sigma^2$.
Again, the estimator is asymptotically unbiased,\index{Unbiased estimator!asymptotically} as $\operatorname{E}[S^2_n] \to\sigma^2$as $n\to\infty$.
Since\ $\overline{X}$ estimates\ $\mu$ imprecisely (from just one of the many possible samples), an extra source of variation is introduced into the estimation of\ $\sigma^2$.

Using the sample found earlier, the empirical estimates\ $S^2_n$\index{Empirical variances} can be found for samples of size $n = 5$ and $n = 100$ (Fig.\ \@ref(fig:SDvarPlots), top panels).
The values of $\operatorname{E}[S^2_n]$ for $n = 5$ and $n = 100$ are
$$
  \operatorname{E}[S^2_n] = \frac{5 - 1}{5}\times 3^2 = 7.2
  \quad\text{and}\quad
  \operatorname{E}[S^2_n] = \frac{100 - 1}{100}\times 3^2 = 8.91
$$
respectively, similar to those means computed from the **R** simulation.
These both suggest that the estimator is unbiased.


```{r SDvarPlots, echo=FALSE, out.width='90%', fig.width=8.5, fig.height = 5.5, fig.align="center", fig.cap="Sampling distributions for the sample variance for sample of size $n = 5$ (left panels) and $n = 100$ (right panels), using the biased estimator $S^2_n$ (top panels) and the unbiased estimator $S^2$ (bottom panels). The solid dots represent the population variance."}

s2n <- function(x) { (sum( (x - mean(x))^2) ) / length(x)}

Xvarn   <- apply(Xvals, 
                MARGIN = 2, 
                FUN = s2n) 
Xvarn_2   <- apply(Xvals_2, 
                   MARGIN = 2, 
                   FUN = s2n) 


Xvar   <- apply(Xvals, 
                MARGIN = 2, 
                FUN = var) 
Xvar_2   <- apply(Xvals_2, 
                  MARGIN = 2, 
                  FUN = var) 

xRange   <- c(0, 50)
xRange_2 <- c(0, 65)

#####

par( mfrow = c(2, 2))


hist(Xvarn, 
     xlim = xRange,
     xlab = "Sample variance",
     ylab = "",
     breaks = seq(min(xRange), max(xRange), by = 2),
     main = "Sample variance:\nsamples of size 5",
     sub = paste0("(Average value: ", round(mean(Xvarn), 3), ")"),
     axes = FALSE)
axis(side = 1)
points(x = x_Sd^2, 
       y = 0,
       pch = 19)



#####

hist(Xvarn_2,
     xlim = xRange,
     xlab = "Sample variance",
     ylab = "",
     breaks = seq(min(xRange), max(xRange), by = 2),
     main = "Sample variance:\nsamples of size 100",
     sub = paste0("(Average value: ", round(mean(Xvarn_2), 3), ")"),
     axes = FALSE)
axis(side = 1)
points(x = x_Sd^2, 
       y = 0,
       pch = 19)


####
####
####


hist(Xvar, 
     xlim = xRange_2,
     xlab = "Sample variance",
     ylab = "",
     breaks = seq(min(xRange_2), max(xRange_2), by = 2),
     main = "Sample variance:\nsamples of size 5",
     sub = paste0("(Average value: ", round(mean(Xvar), 3), ")"),
     axes = FALSE)
axis(side = 1)
points(x = x_Sd^2, 
       y = 0,
       pch = 19)



#####

hist(Xvar_2, 
     xlim = xRange_2,
     xlab = "Sample variance",
     ylab = "",
     breaks = seq(min(xRange_2), max(xRange_2), by = 2),
     main = "Sample variance:\nsamples of size 100",
     sub = paste0("(Average value: ", round(mean(Xvar_2), 3), ")"),
     axes = FALSE)
axis(side = 1)
points(x = x_Sd^2, 
       y = 0,
       pch = 19)

```



To obtain an *unbiased estimator*, we could multiply $S^2_n$ by $(n - 1)/n$ (a constant), which *does* produce an unbiased estimator.


:::{.theorem #SampleVarianceUnbiasedXbar name="An unbiased estimator for $\sigma^2$ ($\mu$ unknown)"}
When the value of\ $\mu$ is unknown, an *unbiased* estimator of\ $\sigma^2$ is 
\begin{equation}
  S^2 = \frac{1}{n - 1} \sum_{i=1}^n (X_i - \overline{X})^2.
   (\#eq:S2WithoutPopMean)
\end{equation}
Notice that the divisor in front of the summation is now\ $n - 1$ rather than\ $n$. 

::::{.proof}
Easily shown by adapting the proof for $S^2_n$ above; see Exercise\ \@ref(exr:UnbiasedVariance).
::::
:::


Using the sample found earlier, the empirical estimates\ $S^2$\index{Empirical variances} can be found for samples of size $n = 5$ and $n = 100$ (Fig.\ \@ref(fig:SDvarPlots), bottom panels).


### Independence of the sample mean and sample= varance {#IndependenceMeanVar}

A result that is useful later is the perhaps-surprising result that the sample mean and the sample variance are independent.

:::{.theorem #MeanVarInd name="Sample mean and variance: Independent"}
Let $X_1, X_2,\dots, X_n$ be a random sample of size\ $n$ from $N(\mu, \sigma^2)$.
Then the sample mean
$$
   \overline{X}
   = 
   \frac1n\sum_{i=1}^n X_i
$$
and sample variance
$$
   S^2
   = \frac{1}{n - 1}\sum_{i=1}^n (X_i - \overline{X})^2
$$
are independent.

::::{.proof}
This proof is not given.
::::
:::


## The distribution of the sample mean {#SamplingDistributionsMean}

### Introduction {#SamplingDistributionsMeanIntro}

In previous sections, estimators were introduced and studied, specifically unbiased estimators for\ $mu$ and\ $\sigma^2$.
All estimators are random variables, and so they all have distributions. 
Previously, we used simulations to display the empirical sampling distribution\index{Sampling distributions!empirical}\index{Empirical distributions} of various statistics.

However, theoretical sampling distributions of various estimators can be found, which leads to numerous very powerful and practical results.


### The mean and variance of $\overline{X}$ {#SamplingDistributionMean}

Usually the sample mean\ $\overline{X}$ is of greatest interest when sampling.
Section\ \@ref(SamplingDistributionsSmallExample) shows that the sample mean, like all estimators, varies from sample to sample.
Figure\ \@ref(fig:SDmeanPlots) suggests that larger sample sizes produce sampling distributions with smaller variances.
The following theorem quantifies this idea.


:::{.theorem #SamplingDistMean name="Sampling distribution of the mean"}
If $X_1, X_2, \dots, X_n$ is a *random sample* of size\ $n$ from a population with mean\ $\operatorname{E}[X] = \mu$ and variance\ $\operatorname{var}[X] = \sigma^2$, then the sample mean\ $\overline{X}$ has a sampling distribution with mean $\mu$ and variance $\sigma^2 / n$.

::::{.proof}
To show that $\operatorname{E}[\overline{X}] = \mu$, apply Theorem\ \@ref(thm:SampleMeanUnbiased).

Similarly, using the properties of the variance, the variance of $\overline{X}$ is:
\begin{align*}
    \operatorname{var}[\overline{X}]
  &= \operatorname{var}[(X_1 + X_2 + \cdots X_n)/n]\\
  &= \frac{1}{n^2} \operatorname{var}[X_1 + X_2 + \cdots X_n]\\
  &= \frac{1}{n^2} \times n\sigma^
  = \frac{\sigma^2}{n}.
\end{align*}
Equivalently, $\operatorname{sd}[X] = \sigma/\sqrt{n}$. 
We sometimes write $\operatorname{var}[\overline{X}] = \sigma^2_{\bar{X}}$ and $\operatorname{sd}[\overline{X}] = \sigma_{\bar{X}}$. 
::::
:::


Theorem\ \@ref(thm:SamplingDistMean) applies to a random sample as defined in Sect.\ \@ref(RandomSamples), because taking a random sample\index{Sample!random} means that the observations are independent.
Theorem\ \@ref(thm:SamplingDistMean) describes the *mean* and *variance* of the sampling distribution of the mean, but not the *shape* of the sampling distribution.

Theorem\ \@ref(thm:SamplingDistMean) shows that the expected value of\ $\bar{X}$ does not depend on the size of the sample, but the variance *decreases* as the sample size *increases*: larger samples trend to produce more precise estimates.
This makes sense intuitively; after all, one reason for preferring larger samples over smaller samples is that they produce more precise estimates.

Theorem\ \@ref(thm:SamplingDistMean) contains two important pices of information.
The first is that it shows *how* variance decreases as\ $n$ gets larger: inversely proportional to\ $n$.
This means that a sample size twice as large reduces the variance by half.
The impact on the standard deviation is probably more intuitive, as the standard deviation is measured in the same units as the data (and the sample mean): a sample size twice as large reduces the standard deviation by a factor of\ $\sqrt{2}$.
In practice this means that to double the precision of\ $\bar{X}$ (i.e., to halve the size of the standard deviation), the sample size needs to be increase four-fold. 

The second important (and remarkable) feature of Theorem\ \@ref(thm:SamplingDistMean) is that the results do not rely on knowing the distribution of the population\ $X$.
The result applies for random samples drawn from *any* distribution, provided the $X_i$ are iid.
In Sect.\ \@ref(SamplingDistributionsNormal), we consider the shape of the sampling distribution when the population specifically has a normal distribution; then in Sect.\ \@ref(CentralLimitTheorem) we consider the shape of the sampling distribution when the population has *any* distribution


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
In Theorem\ \@ref(thm:SamplingDistMean), the distribution of the population from which the sample is drawn is irrelevant.
:::


:::{.example #SamlimgMnVarExp name="Sampling mean and variance from an exponential distribution"}
Suppose samples of $n = 5$ are drawn iid from an exponential distribution with $\lambda = 2$.
This means that $\operatorname{E}[X] = 1/2 = 0.5$, and $\operatorname{var}[X] = 1/\lambda^2 = 0.25$, so we would expect $\operatorname{E}[\overline{X}] = \lambda = 0.5$, and $\operatorname{var}[\overline{X}] = 0.25/n = 0.05$.

The empirical distribution of the means of the samples is shown in Fig.\ \@ref(fig:ExpSD) (left panel).
The empirical estimate (shown on the figure) are very close to the theoretical values.
The centre and right panel show sampling distributions for $n = 15$ and $n = 100$.
:::


```{r, fig.show='hide'}
set.seed(123)

ns <- c(5, 15, 100)
lambda <- 2
num_Sims <- 10000  # number of simulations

sim_means <- lapply(ns, function(n) {
  replicate(num_Sims, 
            mean(rexp(n, rate = lambda)))
})

hist(sim_means[[1]]) 
hist(sim_means[[2]])
hist(sim_means[[3]])

# The variance when n = 100 should be about twenty times larger
# than the variance for n = 5:
var(sim_means[[1]]) / var(sim_means[[3]])
```



```{r ExpSD, echo=FALSE, fig.align="center", fig.cap="Sampling distributions, for samples drawn from an exponential distribution with $\\lambda = 2$, for various sample sizes. The solid dots represent the population mean $1/\\lambda$.", out.width='100%', fig.width=7.5, fig.height=2.5}
par(mfrow = c(1, 3))

hist(sim_means[[1]], 
     axes = FALSE,
     xlim = c(0, 2),
     main = "Sample means:\nn = 5", 
     sub = bquote("(mean;" ~ .(round(mean(sim_means[[1]]), 3)) ~ "and var:" ~ .(round(var(sim_means[[1]]), 4))),
     ylab = "",
     xlab = "Sample means")
axis(side = 1)
points(x = 1/lambda,
       y = 0,
       pch = 19)

hist(sim_means[[2]], 
     axes = FALSE,
     xlim = c(0, 2),
     main = "Sample means:\nn = 15", 
     sub = bquote("(mean;" ~ .(round(mean(sim_means[[2]]), 3)) ~ "and var:" ~ .(round(var(sim_means[[2]]), 4))),
     ylab = "",
     xlab = "Sample mean")
axis(side = 1)
points(x = 1/lambda,
       y = 0,
       pch = 19)

hist(sim_means[[3]], 
     axes = FALSE,
     xlim = c(0, 2),
     breaks = 5,
     main = "Sample means:\nn = 100", 
     sub = bquote("(mean;" ~ .(round(mean(sim_means[[3]]), 3)) ~ "and var:" ~ .(round(var(sim_means[[3]]), 4))),
     ylab = "",
     xlab = "Sample mean")
axis(side = 1)
points(x = 1/lambda,
       y = 0,
       pch = 19)


```


### Sampling distribution of $\overline{X}$: when $X$ has a normal distribution {#SamplingDistributionsNormal}

The normal distribution is used to model many naturally occurring phenomena (Sect.\ \@ref(Normal)). 
Consequently, we begin by studying the sampling distributions of statistics based on data drawn from normal populations.

The following result proves useful.


:::{.theorem #LinearComb name="Linear combinations"}
Let $X_1, X_2, \dots, X_n$ be a set of independent random variables where $X_i\sim N(\mu_i,\sigma^2_i)$. 
Define the linear combination\ $Y$ as
$$
   Y = a_1 X_1 + a_2 X_2 + \dots + a_nX_n.
$$
Then $Y \sim N(\Sigma a_i\mu_i, \Sigma a^2_i\sigma^2_i)$.

::::{.proof}
From Theorem\ \@ref(thm:NormalProperties), the MGF of the random variable\ $X_i$ is
$$
   M_{X_i} (t)
   = \exp\left\{\mu_i t + \frac{1}{2} \sigma^2_i t^2\right\}\quad\text{for $i = 1, 2, \dots, n$}.
$$
So, for a constant\ $a_i$ (using Theorem\ \@ref(thm:ExpectationLinear) with $\beta = 0$, $\alpha = a_i$):
\begin{align*}
   M_{a_i X_i}(t)
   &= M_{X_i}(a_it)\\
   &= \exp\left\{\mu_i a_i t + \frac{1}{2} \sigma^2_i a^2_i t^2\right\}.
\end{align*}
Since the MGF of a sum of independent random variables is equal to the product of their MGFs,
\begin{align*}
     M_Y(t)   
     &= \prod^n_{i = 1} \exp\left( \mu_i a_i t + \frac{1}{2} \sigma^2_i a^2_i t^2 \right)\\
     &= \exp\left( t\Sigma a_i\mu_i +\frac{1}{2} t^2 \Sigma a^2_i \sigma^2_i \right).
\end{align*}
This is the MGF of a normal random variable with mean $\operatorname{E}[Y] = \Sigma a_i\mu_i$ and variance $\operatorname{var}[Y] = \Sigma a^2_i\sigma^2_i$.
::::
:::


:::{.example}
Define
$$
  X_1 \sim N(2, \sigma^2 = 5)
  \quad\text{and}\quad
  X_2 \sim N(25, \sigma^2 = 1)
$$
and $a_1 = -8$ and $a_2 = 10$.
Then, by Theorem\ \@ref(thm:LinearComb), $Y = a_1 X_1  + a_2 X_2$ has a normal distribution, with
\begin{align}
  \operatorname{E}[Y]    &= a_1\mu_1       + a_2\mu_2      = (-8\times 2) + (10\times 25) = 234;
  \text{and}\\
  \operatorname{var}[Y] &= a^2_1\sigma^2_1 + a_2\sigma^2_2 = ((-8)^2\times 5) + (10^2\times 1) = 420.
\end{align}
This can be demonstrated in **R**, as shown below.
The simulated distribution of\ $Y$ is shown in Fig.\ \@ref(fig:HistLinearComb).
:::

```{r, fig.show='hide'}
set.seed(8979704) # For reproducibility

a_1 <- -8; a_2 <- 10
num_Sims <- 10000

X_1 <- rnorm(num_Sims,
             mean = 2, 
             sd = sqrt(5) )
X_2 <- rnorm(num_Sims,
             mean = 25, 
             sd = sqrt(1) )

Y <- a_1 * X_1 + a_2 * X_2
hist(Y)

mean(Y); var(Y)

```

```{r HistLinearComb, echo=FALSE, out.width='100%', fig.width=8, fig.height=2.5, fig.align="center", fig.cap="Left: The normal distribution of $X_1$. Centre: the normal distribution of $X_2$. Right: the empirical distribution of a linear combination of the two normal random variates in a normal distribution. The solid dot represents the theoretical mean of $Y$, and the solid curve is the theoretical distribution of\\ $Y$."}
par(mfrow = c(1, 3))

curve( dnorm(x, mean=2, sd = sqrt(5)),
       from = -4, to = 9,
       lwd = 2,
       ylab = "",
       xlab = expression(italic(x)),
       main = expression(The~distribution~of~italic(X)[1]),
       axes = FALSE)
axis(side = 1)

curve( dnorm(x, mean=25, sd = sqrt(1)),
       from = 22, to = 28,
       lwd = 2,
       ylab = "",
       xlab = expression(italic(x)),
       main = expression(The~distribution~of~italic(X)[2]),
       axes = FALSE)
axis(side = 1)

hist(Y,
     axes = FALSE,
     freq = FALSE,
     ylab = "",
     main = expression(bold(atop(The~distribution~of~the,
                                 random~variable~italic(Y)))),
     xlab = expression(italic(y)))
axis(side = 1)
points(x = 234,
       y = 0,
       pch = 19)

curve(dnorm(x, mean = 234, sd = sqrt(420)),
      add = TRUE,
      lwd = 2,
      from = 150, 
      to = 350)
```

The sampling distribution of the sum and mean of a random sample from a normal population follows directly from Theorem\ \@ref(thm:LinearComb).


:::{.theorem #SumMean name="Sum and mean of a random sample"}
Let $X_1, X_2, \dots, X_n$ be a random sample of size $n$ from $N(\mu,\sigma^2)$. 
Define the sum\ $S$ and mean\ $\overline{X}$ respectively as
\begin{align*}
   S
   &= X_1 + X_2 + \dots + X_n\\
   \overline{X}
   &= (X_1+X_2 + \dots + X_n)/n.
\end{align*}
Then $S\sim N(n\mu, n\sigma^2)$ and $\overline{X}\sim N(\mu, \sigma^2/n)$.

::::{.proof}
See Exercise\ \@ref(exr:SumMean).
::::
:::


Theorem\ \@ref(thm:SumMean) relies only on the population from which the sample is drawn having a normal distribution, and on the properties of expectation; it does not depend on the sample size\ $n$.


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
For any size sample drawn from a normal distribution, the sample means have a normal distribution.
:::


:::{.example #CLTSugarSachets name="Sums of rvs"}
Sugar sachets (for sweetening hot drinks) typical are manufactured so that the weights of sugar in grams\ $W$ in each sachet are normally distributed with a mean of $4\gs$ and standard deviation $0.1\gs$.
For catering purposes, sachets are sold in boxes of\ $1000$.
What is the probability that a box contains more than\ $4020\gs$ of sugar? 

Since the weights\ $W$ follow a normal distribution, then the weight of the sugar in each box is 
$$
   B = W_1 + W_2 + \cdots + W_{1000} \sim N(1000\times 4, \sigma^2=1000 \times 0.1^2).
$$
or $N(4000, 10)$.
Then, we seek $\Pr(B > 4020)$:
$$
  \Pr(B > 4020)
  = 
  \Pr(Z > (4020 - 4000) / 10) 
  = 
  \Pr(Z > 2)
  =
  1 - \phi(2)
$$
In **R**:

```{r}
1 - pnorm(2)
```

The probability is about $2.3$\%.
:::


:::{.example #CLTweights name="CLT"}
The IQs for a large population of $10$\ year-old boys is assumed to be normally distributed, with a mean of\ $110$ and a variance of\ $144$. 
How large a sample is needed to have a probability of\ $0.9$ that the *mean* IQ of the sample would not differ from the expected value\ $110$ by more than\ $5$?

Let\ $X_i$ be the IQ of the $i$th\ boy; then $X_i \sim N(110, 144)$.
Consider a sample of size\ $n$ and let $\overline{X} = \sum^n_{i = 1}X_i/n$; then $\overline{X}\sim N(110, 144/n)$.
The question asks to find the values of\ $n$ such that
$$
   \Pr(|\overline{X} - 110|\leq 5) = 0.90.
$$
That is,
$$
   \Pr\left(\frac{|\overline{X} - 110|}{12/\sqrt{n}} \leq \frac{5}{12/\sqrt{n}}\right) = 0.90
$$
hence
$$
   \Pr(Z \leq 5\sqrt{n} /12) = 0.90. 
$$
Using **R** and being careful (see Fig.\ \@ref(fig:IQFindZ)):
```{r echo=TRUE}
z <- qnorm(0.95); z
```

Thus,
$$
  5\sqrt{n}/12 = 1.644854 \implies n = (12\times 1.644854/5)^2 = 15.58.
$$


The *smallest* size sample, then, would be a sample of $n = 16$.
:::


```{r IQFindZ, echo=FALSE, fig.align="center", out.width='90%', fig.width=8, fig.height=3, fig.cap="Left: the situation in terms of $Z$. Right: finding the required value of $Z$ =."}
par(mfrow = c(1, 2))

xstar <- 1.6

out <- plotNormal(mu = 0,
                  sd = 1,
                  showX = c(-3, -xstar, 0, xstar, 3),
                  showXlabels = c(-3, expression(-italic(z)), 0, expression(italic(z)), 3))
shadeNormal(x = out$x,
            y = out$y,
            lo = -xstar, 
            hi =  xstar,
            col = "grey")
text(x = 2.5,
     y = 0.3,
     pos = 3,
     expression(Area*":"~0.90))
arrows(x0 = 2.5,
       y0 = 0.3,
       x1 = 0.75,
       y1 = 0.1,
       lwd = 2,
       angle = 15,
       length = 0.1)

## 

out <- plotNormal(mu = 0,
                  sd = 1,
                  showX = c(-3, -2, -1, 0, xstar, 3),
                  showXlabels = c(-3, -2, -1, 0, expression(italic(z)), 3))
shadeNormal(x = out$x,
            y = out$y,
            lo = -Inf, 
            hi =  xstar,
            col = "grey")
text(x = 2.5,
     y = 0.3,
     pos = 3,
     expression(Area*":"~0.95))
arrows(x0 = 2.5,
       y0 = 0.3,
       x1 = 0.75,
       y1 = 0.1,
       lwd = 2,
       angle = 15,
       length = 0.1)
```

:::{.example #CLTPlunger name="Two random variables"}
A design involves a plunger fitting into a cylindrical tube.
The diameter of the plunger can be considered a normal random variable with mean $2.1\cms$ and standard deviation $0.1\cms$.
The inside diameter of the cylindrical tube is a normal random variable with mean $2.3\cms$ and standard deviation $0.05\cms$. 
For a plunger and tube chosen randomly from a day's production run, find the probability that the plunger will *not* fit into the cylinder.

Let\ $X$ and\ $Y$ be the diameter of the plunger and cylinder respectively.
Then $X\sim N\big(2.1, (0.1)^2\big)$ and $Y\sim N\big( 2.3, (0.05)^2\big)$, and we seek $\Pr(Y < X)$.
The distribution of $Y - X$ is $N(2.3 - 2.1, 0.0025 + 0.01)$ so that
\begin{align*}
   \Pr(Y - X < 0) 
   &= \Pr\left(Z <\frac{0 - 0.2}{\sqrt{0.0125}}\right) \quad\text{where $Z\sim N(0,1)$}\\
   &= \Pr(Z < -1.78)\\
          &= 0.0375.
\end{align*}
In **R**:
```{r}
pnorm( -1.78 )
```
:::


### The Central Limit Theorem {#CentralLimitTheorem}

Sampling distributions for various statistics of interest are well-developed when sampling is from a *normal* distribution (Sect.\ \@ref(SamplingDistributions)).
Theorem\ \@ref(thm:SumMean) applies when the population is normally distributed.
Although these results are important, what if we do not know the distribution of the population from which the random sample is drawn (as is usually the case)?

Remarkably, when the population is *not* normally distributed, the sampling distribution has an approximate normal distribution for large sample sizes (Sect.\ \@ref(CentralLimitTheorem)).

In Sect.\ \@ref(SamplingDistributionMean) general results are given describing the *mean* and *variance* of the sample mean which holds for any population distribution, but do not say anything about the *distribution* of the sample means.
The *Central Limit Theorem* (or *CLT*) describes the distribution of the sample means.


:::{.theorem #CLT name="Central Limit Theorem (CLT)"}
Let $X_1, X_2, \dots, X_n$ be a random sample from a distribution with mean\ $\mu$ and variance\ $\sigma^2$.
Then the random variable
$$
   Z_n 
   = \frac{\overline{X} - \mu}{\sigma / \sqrt{n}}
$$
converges in distribution to a standard normal variable as $n\to\infty$.

::::{.proof}
The proof is long, so is deferred to Appendix\ \@ref(CLTProof).
::::
:::


'Convergence in distribution' means that the distribution of $Z_n$ converges to a standard normal as $n\to \infty$. 
So when\ $n$ is 'large', $Z_n$ is expected to approximate a standard normal distribution. 
Transforming $Z_n$ back to the sample mean, $\overline{X}$ can be expected to approximate a $N(\mu, \sigma^2/n)$ distribution.

This is a partial proof since the concept of convergence in distribution has not been formally defined. 
However, the arguments used in the proof are powerful and worth understanding.


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
The *Central Limit Theorem* (Theorem\ \@ref(thm:CLT)) is one of the most important theorems in statistics.
:::



In practice, the distribution of\ $\overline{X}$ is sufficiently close to that of a normal distribution when\ $n$ is larger than about\ $25$ in most situations.
However, if the population distribution is severely skewed, larger samples sizes may be necessary for the approximate to be adequate in practice.


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
In simple terms, the *Central Limit Theorem* states that, for *any* random variable\ $X$ (continuous, discrete or mixed), the distribution of the *sample means* will have an:\index{Central Limit Theorem}

* approximate normal distribution,
* with mean\ $\mu = \operatorname{E}[X]$, and
* with a variance of $\sigma^2/n$, where\ $\sigma^2 = \operatorname{var}[X]$.

The approximation improves as the sample size\ $n$ gets larger.

This applies for *any* random variable\ $X$, *whatever its distribution*.
For many distributions that are not highly skewed, the approximation is reasonable for sample sizes larger than about\ $20$ to\ $30$.
If the data comes from a normal distribution, the distribution of the sample mean *always* has a normal distribution for any\ $n$ (Sect.\ \@ref(SamplingDistributionsNormal)).
:::


```{r, child = if (knitr::is_html_output())  './children/CLT/CLT_html.Rmd' }
```

```{r, child = if (knitr::is_latex_output()) './children/CLT/CLT_latex.Rmd'}
```


:::{.example #CLTPoisson name="CLT and a Poisson distribution"}
In the **R** code below, data come from a Poisson distribution (Fig.\ \@ref(fig:PoissonCLT), left panel) with $\lambda = 2$.
However, the mean of samples of size $n = 10$ (centre panel) and $n = 100$ (right panel) are approximately distributed as a normal distribution.
The variance of the sample means for the larger sample size is smaller than that for $n = 10$, and the distribution looks more like a normal distribution for the larger sample size.
:::


```{r PoissonCLT, echo=FALSE, fig.cap="The Central Limit Theorem. Left: the distribution of the individual observations that follow a Poisson distribution. Centre: the empirical sampling distribution of the sample mean for samples of size $n = 10$. Right: the empirical sampling distribution of the sample mean for samples of size $n = 100$. The thick dashed lines represent the theoretical asymptotic normal distribution.", fig.align="center", fig.width=7, out.width='100%', fig.height=3}
# The distribution of individuals has a uniform distribution:
set.seed(40516) # For reproducibility
sample_Size_1 <- 10
sample_Size_2 <- 100
sample_Means_1 <- NULL
sample_Means_2 <- NULL
lambda <-2

number_Simulations <- 10000

for (i in 1:number_Simulations){
  # Take a sample of size 5
  Xs_1 <- rpois(sample_Size_1, 
                   lambda = lambda)
  # Take a sample of size 20
  Xs_2 <- rpois(sample_Size_2, 
                   lambda = lambda)

  # Find the sample mean
  sample_Means_1 <- c(sample_Means_1, mean(Xs_1) )
  sample_Means_2 <- c(sample_Means_2, mean(Xs_2) )
}

par(mfrow = c(1, 3))

x <- 0:10
plotDiscrete( x = x,
              type = "MF",
              px = dpois(x,
                         lambda = lambda),
              las = 1,
              xlab = expression(italic(x)),
              ylab = "Prob. fn",
              main = "Distribution of the\nindividuals observations")

hist( sample_Means_1,
      las = 1,
      ylim = c(0, 0.9),
      freq = FALSE,
      axes = FALSE,
      xlab = expression(Sample~means~bar(italic(x)) ),
      main = expression( bold( atop(Distribution~of~sample~means,
                               "for"~italic(n)==10) ) ) )
curve(dnorm(x, mean=2, sd = sqrt(2) / sqrt(sample_Size_1) ),
      lwd = 2,
      lty = 2,
      add = TRUE)
axis(side = 1)

hist( sample_Means_2,
      las = 1,
      ylim = c(0, 2.75),
      freq = FALSE,
      axes = FALSE,
      xlab = expression(Sample~means~bar(italic(x)) ),
      main = expression( bold( atop(Distribution~of~sample~means,
                               "for"~italic(n)==100) ) ) )
curve(dnorm(x, mean=2, sd = sqrt(2) / sqrt(sample_Size_2) ),
      lwd = 2,
      lty = 2,
      add = TRUE)
axis(side = 1)
```


:::{.example #SoftDrinkVending name="CLT"}
When a coffee machine makes a single espresso-style shot, the volume of water in millimetres\ $W$ that is dispensed has a mean of $30\mLs$ with a standard deviation of $2.2\mLs$.

Find the probability that the *average* volume of water is a random sample of size\ $35$ shots is less than $31\mLs$.

The distribution of\ $W$ is not given, but the mean and standard deviation of\ $W$ are given.
The distribution of the sample mean\ $\overline{W}$ can be approximated by the normal distribution with mean of $\mu = 30$ and standard error of $\sigma_{\overline{W}} = \sigma/\sqrt{n} = 2.2/\sqrt{36} = 0.3718\mLs$, according to the CLT.
That is, $\overline{W} \sim N(30, 0.3718^2)$.
Now
$$
   \Pr(\overline{W} < 31) 
   \approx \Pr\left(Z < \frac{31 - 30}{0.3718}\right)
   = \Pr(Z < 2.69).
$$
In **R**:

```{r}
pnorm(2.69)
```

Hence, $\Pr(\overline{W} < 31) \approx 0.0036$, or approximately\ $99.6$%.
:::


:::{.example #FairDiceSum name="Throwing dice"}
Consider throwing a fair die\ $n$ times and observing the sum of the faces showing. 
For $n = 12$ rolls, find the probability that the sum of the faces is at least\ $52$.

Let the random variable\ $X_i$ be the number showing on the $i$th\ throw.
Then define $Y = X_1 + \dots + X_{12}$, so that we seek $\Pr(Y\geq 52)$.

To use Theorem\ \@ref(thm:CLT), see that '$Y\geq 52$' is equivalent to '$\overline{X}\geq 52/12$', where $\overline{X} = Y/12$ is the mean number showing from the $12$\ tosses.

Since the distribution of each\ $X_i$ is rectangular with $\Pr(X_i = x) = 1/6$ (for $x = 1, 2, \dots, 6$), then $\operatorname{E}[X_i] = 7/2$ and $\operatorname{var}[X_i] = 35/12$ (Sect.\ \@ref(DiscreteUniformProperties)).
Hence, $\operatorname{E}[\overline{X}] = 7/2$ and $\operatorname{var}[\overline{X}] = 35/(12^2)$, and so 
$\overline{X} \sim N(7/2, sigma^2 = 35/(12^2))$. 

Then, from Theorem\ \@ref(thm:CLT),
\begin{align*}
  \Pr(Y\geq 52) 
    & \simeq \Pr(\overline{X}\geq 52/12)\\
    &= \Pr_N \left(Z\geq \frac{52/12 - 7/2}{\sqrt{35/144}}\right)\\
    &= 1 - \Phi(1.690) = 0.0455
\end{align*}
The probability is approximately\ $4.5$%.
:::


Example\ \@ref(exm:FairDiceSum) can also be solved using a generalisation of the Central Limit Theorem.
This generalisation indicates why the normal distribution plays such an important role in statistical theory. 
It states that a large class of random variables converge in distribution to the standardized normal.


:::{.theorem #ConvergeToN name="Convergence to standard normal distribution"}
Let $X_1, X_2, \dots, X_n$ be a sequence of independent random variables with $\operatorname{E}[X_i] = \mu_i$, $\operatorname{var}[X_i] = \sigma^2_i$.
Define $Y = a_1X_1 + a_2X_2 + \dots + a_nX_n$.

Then under certain general conditions, including\ $n$ large, $Y$\ is distributed approximately as $N(\sum_i a_i\mu_i, \sum_i a^2_i\sigma^2_i)$.

::::{.proof}
Not given.
::::
:::


Theorem\ \@ref(thm:ConvergeToN) makes no assumptions about the distribution of\ $X_i$.
If the\ $X_i$ are *normally* distributed, then $Y = \sum_{i = 1}^n a_i X_i$ will have a normal distribution for any\ $n$, large or small.


SALT SACHETS:

:::{.example #NoiseVoltages name="CLT (Voltages)"}
Suppose we have a number of independent noise voltages\ $V_i$ (for $i = 1, 2, \dots, n$). 
Let\ $V$ be the sum of the voltages, and suppose each\ $V_i$ is distributed $U(0, 10)$.
For $n = 20$, find $\Pr(V > 105)$.

This is an example of Theorem\ \@ref(thm:ConvergeToN) with each $a_i = 1$. 
To find $\Pr(V > 105)$, the distribution of\ $V$ must be known.

Since $\operatorname{E}[V_i] = 5$ and $\operatorname{var}[V_i] = 100/12$, from Theorem\ \@ref(thm:ConvergeToN) $V$ has an approximate normal distribution with mean $20\times 5 = 100$ and variance $20\times 100/12$.
That is, $\displaystyle{\frac{V - 100}{10\sqrt{5/3}}}$ is distributed $N(0, 1)$ approximately. 
So
$$ 
   \Pr(V > 105) 
   \approx \Pr\left (Z > \frac{105 - 100}{12.91}\right )
   = 1 - \Phi(0.387)
   = 0.352.
$$
The probability is approximately\ $35$%.
:::



## The $\chi^2$-distribution {#ChiSquaredDistribution}

The two most important and used descriptions of a sample are the sample mean and the sample variance.
Having already discussed the sampling distribution of the sample mean, we now turn to the sampling distribution of the sample variance
\begin{equation}
  S^2 = \frac{1}{n-1} \sum_{i=1}^n (X_i - \bar{X})^2.
  (\#eq:S2)
\end{equation}
for a sample $X_1, \dots, X_n \overset{\text{iid}}{\sim} N(\mu, \sigma^2)$ (i.e., independently and identically\index{iid} distributed from a normal distribution) (Theorem\ \@ref(thm:SampleVarianceUnbiasedXbar)).
We are seeking the distribution of $S^2$.



\index{Chi-squared distribution|(}
The formula for the estimator\ $S^2$ in Eq.\ \@ref(eq:S2) involves an\ $\sum X_i^2$ term, where each $X_i$ has a normal distribution.
Hence, to find the distribution of\ $S^2$, we start with the distribution of a squared normal variate; specifically, consider
$$
  Z = \frac{X - \mu}{\sigma} \sim N(0,1)
$$
and the distribution of $Y = Z^2$.
The distribution of\ $Y$ is easy to determine using the techniques in Chap.\ \@ref(ChapterTransformations)), and yields
$$
  f_Y(y) = \frac{y^{-1/2}\exp(-y/2)}{\sqrt{2\pi}} \quad\text{for $y > 0$}.
$$
The proof is Exercise\ \@ref(exr:DistZ2).
Note that $Y \sim \text{Gamma}(\alpha = 1/2, \beta = 2)$.

Now consider $Z_1, Z_2, \dots Z_n\overset{\text{iid}}{\sim} N(0, 1)$, and the summation $\sum_{i = 1}^n Z_i^2$.
Since each\ $Z_i$ has the same gamma distribution, the distribution of the summation can be found (Exercise\ \@ref(exr:SumZ2)) as
$$
  \sum_{i = 1}^n Z_i \sim \text{Gamma}(\alpha = n/2, \beta = 2).
$$
This distribution occurs so often in practice that is has its own name: a $\chi^2$-distribution.
That is, a $\text{Gamma}(\alpha = n/2, \beta = 2)$ is defined as a $\chi^2_n$-distribution.
Thus:
\begin{equation}
  \sum_{i = 1}^n Z_i \sim \chi^2_n,
  (\#eq:SumZ2)
\end{equation}
where\ $n$ here is the *degrees of freedom*.\index{Degrees of freedom}


:::{.definition #ChiSquaredDistribution name="Chi-squared distribution"}
A continuous random variable\ $X$ with probability density function\index{Chi-squared distribution}
\begin{equation}
   f_X(x)
   = \frac{x^{(\nu/2) - 1}\exp(-x/2)}{2^{\nu/2}\,\Gamma(\nu/2)}
   \quad\text{for $x > 0$}
\end{equation}
is said to have a *chi-squared distribution* with parameter $\nu > 0$, called the *degrees of freedom*.
We write $X \sim \chi^2_\nu$.
:::


Some plots of $\chi^2$-distributions are shown in Fig.\ \@ref(fig:ChisqPlots).


::: {.linkBox .link data-latex="{iconmonstr-link-1-240.png}"}
The chi-squared distribution is a special case of the gamma distribution, with $\alpha = \nu/2$ and $\beta = 2$ (Exercise\ \@ref(exr:ChiSquaredGamma)).
This means that properties of the chi-squared distribution can be obtained directly from those for the gamma distribution (Sect.\ \@ref(GammaDistribution)).
:::


```{r ChisqPlots, echo=FALSE, out.width='100%', fig.height=2.75, fig.width=6.5, fig.cap="Some $\\chi^2\\!$-distribution probability density functions."}
par( mfrow = c(1, 3))

xx <- seq(0.01, 4, length = 100)
plot( x = xx,
      y = dchisq(xx, df = 1),
      lwd = 2,
      type = "l",
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression( bold(chi^2*"-"*distribution: df==1)))

xx <- seq(0.001, 15, length = 100)
plot( x = xx,
      y = dchisq(xx, df = 5),
      lwd = 2,
      type = "l",
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression( bold(chi^2*"-"*distribution: df==5)))

xx <- seq(0.001, 50, length = 100)
plot( x = xx,
      y = dchisq(xx, df = 20),
      lwd = 2,
      type = "l",
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression( bold(chi^2*"-"*distribution: df==20)))
```


:::{.theorem #ChisqProperties name="Properties of the chi-squared distribution"}
If $X\sim\chi^2_\nu$ then

* $\operatorname{E}(X) = \nu$.
* $\operatorname{var}(X) = 2\nu$.
* $M_X(t) = (1 - 2t)^{-\nu/2}$ for $t < 1/2$.

::::{.proof}
See Theorem\ \@ref(thm:GammaProperties).
::::
:::


:::{.example}
Some example.
:::
\index{Chi-squared distribution|)}


Having found the distribution of the sum of\ $n$ random variables, we can return to the original task: finding the sampling distribution of the sample variance\ $S^2$ (in Eq.\ \@ref(eq:S2)).
First, write
$$
  \sum_{i=1}^n (X_i - \overline{X})^2
  = 
  \sum_{i=1}^n (X_i - \mu)^2 - 
  \sum_{i=1}^n (\bar{X} - \mu)^2
$$
(see Exercise\ \@ref(Exercise3)).
Dividing this expression by\ $\sigma^2$ then gives
\begin{align*}
  \sum_{i=1}^n \frac{(X_i - \overline{X})^2}{\sigma^2}
  &= 
  \sum_{i=1}^n \frac{(X_i - \mu)^2}{\sigma^2} - 
  \sum_{i=1}^n \frac{(\bar{X} - \mu)^2}{\sigma^2}\\
  &= 
  \sum_{i=1}^n \left( \frac{X_i - \mu}{\sigma}\right)^2 - 
  \frac{n(\bar{X} - \mu)^2}{\sigma^2}.
\end{align*}
The term on the left side has a $\chi^2_n$ distribution (Eq~\@ref(\#eq:SumZ2)).
The final term has the form $Z^2$, where $Z = (\bar{X} - \mu)/(\sigma/\sqrt{n})$; thus the third term has a $\chi^2(1)$ distribution.

Now, *if those two terms are independent*, we can use the MGF method to show that
$$
  \sum_{i=1}^n \frac{(X_i - \bar{X})^2}{\sigma^2} \sim \chi^2_{n-1}
$$
Showing that the two terms are independent is not straightforward, but it is equivalent to showing that\ $S^2$ and $\bar{X}$ are independent since the\ $X_i$ are normally distributed (Exercise\ \@ref(exr:MeanSdIndependent)).
Assuming the independence holds, this means that
$$
  \frac{1}{\sigma^2} \sum_{i=1}^n (X_i - \bar{X})^2 \sim \chi^2_{n-1}
$$
and hence
$$
  \frac{(n - 1)S^2}{\sigma^2} \sim \chi^2_{n-1}
$$
after substituting Eq.\ \@ref(eq:S2) and rearranging.
This means that $S^2$ follows a scaled $\chi^2_{n-1}$ distribution.


:::{.theorem #SamplingDistMean name="Sampling distribution of the mean"}
If $X_1, X_2, \dots, X_n$ is an *iid sample* of size\ $n$ from a population with a $N(\mu, \sigma^2)$ distribution, then the sample variance\ $S^2$ has a sampling distribution distributed as  $\chi^2_{n-1}$.

::::{.proof}
The proof has been outlined in the above discussion.
::::
:::


::: {.softwareBox .software data-latex="{iconmonstr-laptop-4-240.png}"}
The four **R** functions for working with the $\chi_\nu^2$-distribution have the form `[dpqr]chisq(., df)`, where `df`${} = \nu$ is the degrees of freedom (see Appendix\ \@ref(UseRDistributions)):

* `dchisq(x, df)` computes the PDF at $X = {}$`x`;
* `pchisq(q, df)` computes the CDF at $X = {}$`q`;
* `qchisq(p, df)` computes the quantile for cumulative probability `p`; and
* `rchisq(n, df)` generates `n` random observations.
:::


:::{.example}
Consider taking samples of size $n = 5$ and $n = 20$ from a $N(5, \sigma^2 = 4)$ distribution.
The sampling distribution of
$$
  \frac{n - 1}{4} S^2 \sim \chi^2_{n - 1}.
$$
Figure\ \@ref(fig:ChiS2R) shows the empirical sampling distribution for the scaled sample standard deviation for both sample sizes.
:::

```{r fig.show='hide'}
set.seed(42)

n_a <- 5; n_b <- 20
mu <- 5; sigma2 <- 4
num_Sims   <- 10000

# Simulate nsim values of (n-1)S^2 / sigma^2
scaled_S2_a <- replicate(num_Sims, {
  x <- rnorm(n_a, 
             mean = mu, sd = sqrt(sigma2))
  (n_a - 1) * var(x) / sigma2
})
scaled_S2_b <- replicate(num_Sims, {
  x <- rnorm(n_b, 
             mean = mu, sd = sqrt(sigma2))
  (n_b - 1) * var(x) / sigma2
})

par(mfrow = c(1, 2))

# Plot histogram vs true chi^2(n-1) density
hist(scaled_S2_a, 
     freq = FALSE,
     main = "Scaled sample standard\ndeviation n = 5",
     xlab = "Value")

curve(dchisq(x, df = n_a - 1), 
      add = TRUE, 
      lwd = 2)

hist(scaled_S2_b, 
     freq = FALSE,
     main = "Scaled sample standard\ndeviation n = 20",
     xlab = "Value")

curve(dchisq(x, df = n_b - 1), 
      add = TRUE, 
      lwd = 2)

```

```{r ChiS2R, echo=FALSE, fig.align="center", fig.cap="The sampling distribution of the scaled standard deviation.", fig.width=8, fig.height=3.25, out.width='90%'}
# Plot histogram vs true chi^2(n-1) density
par(mfrow = c(1, 2))

hist(scaled_S2_a, 
     freq = FALSE,
     ylim = c(0, 0.18),
     axes = FALSE,
     main = "Scaled sample standard\ndeviation: n = 5",
     ylab = "",
     xlab = "Sample standard deviation")
axis(side = 1)

curve(dchisq(x, df = n_a - 1), 
      add = TRUE, 
      lwd = 2)

###

hist(scaled_S2_b, 
     freq = FALSE,
     axes = FALSE,
     ylab = "",
     main = "Scaled sample standard\ndeviation: n = 20",
     xlab = "Sample standard deviation")
axis(side = 1)

curve(dchisq(x, df = n_b - 1), 
      add = TRUE, 
      lwd = 2)


```


:::{.example #Chi2Eng name="Sample standard deviation"}
Quality control engineers routinely need to assess the variability of a production process. 
A process may be centred correctly yet still produce defective items if the spread is too large.

Suppose a manufacturer produces steel rods whose diameter (in mm) should follow a $N(\mu, \sigma^2)$ distribution. The target standard deviation is $\sigma = 0.05\mms$; large variances are outside tolerance.

The production line is monitored by a quality engineer by drawing a random sample of $n = 25$ rods each hour and computing the sample variance $S^2$.
In one sample of $n = 25$, the engineer finds $s^2 = 0.0031\mms^2$.
Find the probability that $S^2 > 0.0031$ when $\sigma^2 = 0.0025$.

Using the chi-squared result, 
$$
  \frac{n-1}{\sigma^2} S^2 
  \sim  \chi^2_{n-1} 
  \quad
  \text{so that}
  \quad  
  \frac{24}{0.0025} S^2 
  =
  9600\, S^2 
  \sim  \chi^2_{24} 
$$

Thus, 
$$
  \Pr(S^2 > 0.0031)
  =
  \Pr\left(9600\,S^2 > \frac{24\times 0.0031}{0.0025}\right)
$$
In\ **R**:

```{r ChiExample}
1 - pchisq(29.76, 
           df = 24)
```
There is an approximate $19$% probability of observing a sample variance this large or larger purely by chance, if the true variance is $\sigma^2 = 0.0025$.
:::


## The $T$-distribution {#TDistribution}

\index{t-distribution@$t$-distribution}
Many other useful distributions exist that are related to the normal distribution.
Because of the Central Limit Theorem (Sect.\ \@ref(CentralLimitTheorem)), these distribution prove useful in many practical situations.

In practice, a common task is to evaluate a claim or belief about population mean\ $\mu$, based on the mean of a sample\ $\bar{X}$.
For example [@lin2021sleep], is the average over-night sleep time for a sample of children equal to the recommendation of $10\hs$?

In these cases, provided the sample size is large so that the Central Limit Theorem applies, the statistic $(\bar{X} - \mu)/(\sigma/\sqrt{n})$ has an approximate normal distribution, whatever the distribution of\ $X$, provided the value of\ $\sigma^2$ is known.

But the value of\ $\sigma^2$ is almost never known, and the estimate\ $S^2$ instead is used.
However, the statistic $T = (\overline{X} - \mu)(S/\sqrt{n}\,)$ no longer has a normal distribution.

Consider, then, the statistic\ $T$ and re-express:
\begin{align*}
  T = \frac{\overline{X} - \mu}{s/\sqrt{n}}
  &=
  \frac{\overline{X} - \mu}{\sigma/\sqrt{n}} \times \frac{\sigma\sqrt{n-1}}{s\sqrt{n-1}} \\
  &=
  \frac{\overline{X} - \mu}{\sigma/\sqrt{n}} 
    \times 
  \sqrt{ \frac{\sigma^2}{s^2(n-1)} }
     \times 
  \sqrt{n-1}.
\end{align*}
This re-writing may appear strange, but the ratio now comprises two expression whose distribution is known:
\begin{align*}
  Z = \frac{\overline{X} - \mu}{\sigma/\sqrt{n}} 
  &\sim N(0, 1), \quad\text{and}\\
  V = \frac{s^2 (n-1)}{\sigma^2}
  &\sim \chi^2_{n-1}.
\end{align*}
So we can write
$$
  T = \frac{Z}{\sqrt{V/(n - 1)}}.
$$
This ratio occurs so often in statistical inference, and in practice, that it has its own name.
This defines a $T$-distribution.


:::{.definition #TDistribution name="$T$-distribution"}
Suppose $Z \sim N(0, 1)$ (i.e., a standard normal distribution) and $V \sim \chi^2_n$ (i.e., a chi-squared distribution with $n$\ degrees of freedom).
Then
\begin{equation}
  T = \frac{Z}{\sqrt{V/n}}
  (\#eq:Tdist)
\end{equation}
is defined as a $T$-distribution with $n$\ degrees of freedom.
:::

:::{.theorem #TDistributionPDF name="$T$-distribution PDF"}
A continuous random variable\ $X$ with probability density function
\begin{equation}
   f_X(x)
   = \frac{\Gamma\left((\nu + 1)/2\right)}{\sqrt{\pi \nu}\,\Gamma(\nu/2)}
     \left(1 + \frac{x^2}{\nu}\right)^{-(\nu + 1)/2}
     \quad\text{for $x\in\mathbb{R}$}
\end{equation}
is said to have a *$t$-distribution* with parameter $\nu > 0$.
The parameter\ $\nu$ is called the *degrees of freedom*. 
We write $X \sim t_\nu$.

::::{.proof}
This proof is not given.
::::
:::


The probability density function of the $T$-distribution is very similar to the standard normal distribution (Fig.\ \@ref(fig:TPlots)): bell-shaped and symmetric about zero. 
However, the variance is greater than one (Theorem\ \@ref(thm:TDistProperties)) and so is larger than the variance of the standard normal distribution, and is dependent on\ $\nu$.



::: {.softwareBox .software data-latex="{iconmonstr-laptop-4-240.png}"}
The four **R** functions for working with the $t$\ distribution have the form `[dpqr]t(., df)`:

* `dt(x, df)` computes the PDF at $X = {}$`x`;
* `pt(q, df)` computes the CDF at $X = {}$`q`;
* `qt(p, df)` computes the quantile for cumulative probability `p`; and
* `rt(n, df)` generates `n` random numbers,

where `df` is the degrees of freedom (see Appendix\ \@ref(UseRDistributions)). 
:::


```{r TPlots, echo=FALSE, out.width='100%', fig.height=2.75, fig.width=6.5, fig.cap="Some $t$-distributions (with normal distributions in dashed lines), with mean\\ $0$ and variance\\ $1$."}
par( mfrow = c(1, 3))

xx <- seq(-4, 4,
          length = 100)

plot( x = xx,
      y = dt(xx, df = 2),
      ylim = c(0, 0.4),
      lwd = 2,
      type = "l",
      axes = FALSE,
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression( bold(atop(The~italic(t)*"-"*distribution,
                                   with~2~df))) )
axis(side = 1)
lines( x = xx,
       y = dnorm(x = xx),
       lwd = 2,
       lty = 2,
       col = ColourOpaque)

plot( x = xx,
      y = dt(xx, df = 7),
      lwd = 2,
      type = "l",
      axes = FALSE,
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression( bold(atop(The~italic(t)*"-"*distribution,
                                   with~7~df))))
axis(side = 1)
lines( x = xx,
       y = dnorm(x = xx),
       lwd = 2,
       lty = 2,
       col = ColourOpaque)

plot( x = xx,
      y = dt(xx, df = 30),
      lwd = 2,
      type = "l",
      axes = FALSE,
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression( bold(atop(The~italic(t)*"-"*distribution,
                                   with~30~df))))
axis(side = 1)
lines( x = xx,
       y = dnorm(x = xx),
       lwd = 2,
       lty = 2,
       col = ColourOpaque)
```


:::{.theorem #TDistProperties name="Properties of the $T$-distribution"}
If $X\sim t_\nu$ then

* For $\nu > 1$, $\operatorname{E}[X] = 0$.
* For $\nu > 2$, $\operatorname{var}[X] = \displaystyle{\frac{\nu}{\nu - 2}}$.
* The MGF does not exist because only the first $\nu - 1$ moments exist. 

::::{.proof}
This proof is not given.
::::
:::


Although we won't prove it, as $\nu\to\infty$ the $t$\ distribution converges to the standard normal (which can be seen in Fig.\ \@ref(fig:TPlots)).
In addition, from Theorem\ \@ref(thm:TDistProperties), we see that as $\nu \to \infty$, $\operatorname{var}[X] \to 1$ as for the normal distribution.

The usefulness of the $t$-distribution derives from the following important result.


:::{.theorem #Tscores name="$T$-scores"}
Let $X_1, X_2, \dots, X_n$ be a random sample of size\ $n$ from $N(\mu, \sigma^2)$.
Then the random variable
$$
   T = \frac{\overline{X} - \mu}{S/\sqrt{n}}
$$
follows a $t_{n - 1}$-distribution where\ $\overline{X}$ is the sample mean and\ $S^2$ is the sample variance as defined in\ Eq.\ \@ref(eq:SampleVarDist).

::::{.proof}
**We give a partial proof only.**
The statistic\ $T$ can be written as
\begin{align*}
  T
    &= \frac{\overline{X} - \mu}{S/\sqrt{n}}\\
    &= \left(\frac{\overline{X} - \mu}{\sigma/\sqrt{n}}\right)   \frac{1}{\sqrt{\frac{(n - 1)S^2}{\sigma^2}/(n - 1)}}\\
    &= \frac{Z}{Y/(n - 1)},
\end{align*}
where $Z$ is a $N(0, 1)$ variable and $Y$ is a chi-squared variable with $(n - 1)$ df.
Using Def.\ \@ref(def:TDistribution) then, $T$ has a $T$-distribution.
::::
:::



Notice that\ $T$ represents a *standardised* version of the sample mean.
Because of this, the $T$\ statistic is important in statistical inference.
You may have used a $t$-test in hypothesis testing, for example.


:::{.example #CalcaltingT name="Calculating $T$"}
A random sample $21, 18, 16, 24, 16$ is drawn from a normal population with mean of\ $20$.

1. What is the value of\ $T$ for this sample?
2. In random samples from this population, what is the probability that\ $T$ is less than the value found above?

From the sample: $\overline{x} = 19.0$ and $s^2 = 12.0$. 
Therefore $t = \frac{19.0 - 20}{\sqrt{12.0/5}} = -0.645$.
(Notice lower-case symbols are used for specific values of statistics, and upper-case symbols for the random variables.)

Interest here is in $\Pr(T < -0.645)$ where $T\sim t_4$. 
From **R**, the answer is\ $0.277$:

```{r echo=TRUE}
pt(-0.645, df = 4)
```
:::


## The $F$-distribution {#FDistribution}

\index{F-distribution@$F$-distribution}
The $F$-distribution arises naturally when comparing the variability of two independent normal populations, and is also used in *analysis of variance*, a common technique for testing the equality of several means.


:::{.definition #FDistribution name="$F$-distribution"}
A continuous random variable\ $X$ with probability density function
\begin{equation}
  f_X(x)
  = \frac{\Gamma\big( (\nu_1 + \nu_2)/2 \big)
    \nu_1^{\nu_1/2} \nu_2^{\nu_2/2} x^{(\nu_1/2) - 1}}
   {\Gamma(\nu_1/2)\Gamma(\nu_2/2)(\nu_1 x + \nu_2)^{(\nu_1 + \nu_2)/2}}
  \quad\text{for $x > 0$}
\end{equation}
is said to have an *$F$-distribution* with parameters $\nu_1 > 0$ and $\nu_2 > 0$, known respectively as the numerator and denominator degrees of freedom.
We write $X \sim F_{\nu_1, \nu_2}$.
:::


Some $F$-distributions are shown in Fig.\ \@ref(fig:FPlots).


:::{.theorem #FProperties name="Properties of the $F$-distribution"}
If $X\sim F_{\nu_1, \nu_2}$ then

* For $\nu_2 > 2$,
  $\operatorname{E}[X] = \displaystyle{\frac{\nu_2}{\nu_2-2}}$.
* For $\nu_2 > 4$,
  $\operatorname{var}[X] = \displaystyle{\frac{2\nu_2^2(\nu_1 + \nu_2-2)}
  {\nu_1(\nu_2 - 2)^2(\nu_2 - 4)}}$.
* The MGF does not exist.

::::{.proof}
Not covered.
::::
:::


::: {.softwareBox .software data-latex="{iconmonstr-laptop-4-240.png}"}
The four **R** functions for working with the $F$-distribution have the form `[dpqr]f(df1, df2)`, where `df1`${} = \nu_1$ and
`df2`${} = \nu_2$ are the degrees of freedom (see Appendix\ \@ref(UseRDistributions)):

* `df(x, df1, df2)` computes the PDF at $X = {}$`x`;
* `pf(q, df1, df2)` computes the CDF at $X = {}$`q`;
* `qf(p, df1, df2)` computes the quantile for cumulative probability `p`;
  and
* `rf(n, df1, df2)` generates `n` random observations.
:::


```{r FPlots, echo=FALSE, out.width='100%', fig.height=2.5, fig.width=6.5, fig.cap="Some $F$-distributions."}
par(mfrow = c(1, 3))

xx <- seq(0.01, 4, length = 100)

plot( x = xx,
      y = df(xx, df1 = 1, df2 = 4),
      lwd = 2,
      type = "l",
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression(bold(atop(The~italic(F)*"-"*distribution,
                                  with~df=="(1, 4)"))))

plot( x = xx,
      y = df(xx, df1 = 6, df2 = 1),
      lwd = 2,
      type = "l",
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression(bold(atop(The~italic(F)*"-"*distribution,
                                  with~df=="(6, 1)"))))

plot( x = xx,
      y = df(xx, df1 = 30, df2 = 5),
      lwd = 2,
      type = "l",
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression(bold(atop(The~italic(F)*"-"*distribution,
                                  with~df=="(30, 5)"))))
```


The main result connecting the $F$-distribution to samples from normal populations is the following.


:::{.theorem #FGenerate name="Sampling distribution of the ratio of two sample variances"}
Let $X_1, X_2, \dots, X_{n_1}$ be a random sample of size\ $n_1$ from $N(\mu_1, \sigma_1^2)$ and $Y_1, Y_2,\dots, Y_{n_2}$ be an independent random sample of size\ $n_2$ from $N(\mu_2, \sigma_2^2)$.
Then the random variable
$$
   F = \frac{S_1^2/\sigma_1^2}{S_2^2/\sigma_2^2}
$$
follows an $F_{n_1 - 1,\, n_2 - 1}$-distribution.

::::{.proof}
**A partial proof only is given.**
By Theorem\ \@ref(thm:SampleVarDist),
$$
   U_1 = \frac{(n_1 - 1)S_1^2}{\sigma_1^2} \sim \chi^2_{n_1 - 1}
   \qquad\text{and}\qquad
   U_2 = \frac{(n_2 - 1)S_2^2}{\sigma_2^2} \sim \chi^2_{n_2 - 1},
$$
and $U_1$, $U_2$ are independent because the two samples are independent.
Setting $\nu_1 = n_1 - 1$ and $\nu_2 = n_2 - 1$, the statistic can be written as
$$
   F = \frac{S_1^2/\sigma_1^2}{S_2^2/\sigma_2^2}
     = \frac{U_1/\nu_1}{U_2/\nu_2}.
$$
That this ratio has an $F_{\nu_1, \nu_2}$-distribution follows directly from Def.\ \@ref(def:FDistribution); the derivation of the PDF is not given here.
::::
:::


:::{.corollary}
If $T \sim t_n$ then $T^2 \sim F_{1,\, n}$.

::::{.proof}
From Def.\ \@ref(def:TDistribution), $T = Z / \sqrt{V/n}$ where $Z \sim N(0,1)$ and $V \sim \chi^2_n$ are independent.
Then
$$
  T^2 = \frac{Z^2/1}{V/n}.
$$
Since $Z^2 \sim \chi^2_1$ (Theorem\ \@ref(thm:SumSquaredNormals)), this is a ratio of the form $\chi^2_1/1$ over $\chi^2_n/n$, which by Def.\ \@ref(def:FDistribution) has an $F_{1,n}$-distribution.
::::
:::


:::{.example #FProbs name="$F$-distribution probabilities"}
Suppose $X\sim F_{2, 10}$.
Find:

* $\Pr(X < 1)$
* $x$ such that $\Pr(X > x) = 0.01$

Using **R**:

```{r echo=TRUE}
pf(1,    df1 = 2, df2 = 10)
qf(0.99, df1 = 2, df2 = 10)
```
:::


## The normal approximation to the binomial {#NormalApproxBinomialCLT}

\index{Binomial distribution!normal approximation}
The normal approximation to the binomial distribution (Sect.\ \@ref(NormalApproxBinomial)) can be seen as an application of the [Central Limit Theorem](#thm:CLT).\index{Central Limit Theorem}
The essential observation is that a sample proportion is a sample mean. 
Consider a sequence of independent [Bernoulli trials](#def:BernoulliTrials) resulting in the random sample $X_1, X_2, \dots, X_n$ where
$$
  X_i
  =
  \begin{cases}
     0  &  \text{if failure}\\
     1  &  \text{if success}
  \end{cases}
$$
denotes whether or not the $i$th\ trial is a success.
Then the sum
$$
   Y = \sum_{i = 1}^n X_i
$$
represents the number of successes in the $n$\ trials and
$$
   \overline{X}
   = 
   \frac{1}{n}\sum_{i = 1}^n X_i
   = 
   \frac{Y}{n}
$$
is a sample mean representing the proportion or fraction of trials which are successful. 
In this context, $\overline{X}$ is usually denoted by the sample proportion\ $\widehat{p}$.

Note that $\operatorname{E}[X_i] = p$ and $\operatorname{var}[X_i] = p(1 - p)$.
Therefore
$$
   \operatorname{E}[Y] = np
   \quad\text{and}\quad
   \operatorname{var}[Y] = np(1 - p)
$$
and
$$
   \operatorname{E}[\overline{X}] = p
   \quad\text{and}\quad
   \operatorname{var}[\overline{X}] = \frac{p(1 - p)}{n}.
$$
Theorems\ \@ref(thm:CLT) and\ \@ref(thm:ConvergeToN) are applicable to\ $\overline{X}$ and\ $Y$ respectively. 
Hence
$$
   \overline{X} = \widehat{p}
   \sim N\left(p, \frac{p(1 - p)}{n}\right)\text{ approximately}
$$
and
$$
   Y = n\widehat{p}
   \sim N(np, np(1 - p))\text{ approximately}.
$$



## Appendix: Proof of the CLT {#CLTProof}

This Appendix outlines the proof of the Central Limit Theorem (Theorem\ \@ref(thm:CLT)).

Without loss of generality, centre the $X_i$ by letting $Y_i = (X_i - \mu)/\sigma$,
so that $\operatorname{E}[Y_i] = 0$, $\operatorname{var}[Y_i] = 1$, and
$$
  Z_n = \frac{1}{\sqrt{n}} \sum_{i=1}^n Y_i.
$$
Let $\varphi(t) = \operatorname{E}[\exp(itY)]$ denote the common characteristic function (CF) of the\ $Y_i$.
Since the\ $Y_i$ are independent, the CF of\ $Z_n$ is
\begin{align*}
  \varphi_{Z_n}(t)
  &= \operatorname{E}[\exp(itZ_n)]\\
  &= \operatorname{E}\left[\exp\left(\frac{it}{\sqrt{n}}\sum_{i=1}^n Y_i\right)\right]\\
  &= \prod_{i=1}^n \operatorname{E}\left[\exp(it Y_i/\sqrt{n})\right]\\
  &= \left[\varphi\!\left(t/\sqrt{n}\right)\right]^n.
\end{align*}

Since $\operatorname{E}[Y^2] = 1 < \infty$, the CF $\varphi(t)$ is twice differentiable at $t = 0$, with
$$
  \varphi'(t)\big|_{t=0} = i\operatorname{E}[Y] = 0
  \qquad\text{and}\qquad
  \varphi''(t)\big|_{t=0} = i^2 \operatorname{E}[Y^2] = -1.
$$
The Taylor expansion of $\varphi$ about $t = 0$ is therefore
\begin{align*}
  \varphi(t)
  &= 1 + \varphi'(0)\,t + \frac{1}{2}\varphi''(0)\,t^2 + o(t^2)\\
  &= 1 - \frac{1}{2}t^2 + o(t^2)
  \quad \text{as $t \to 0$},
\end{align*}
where $o(t^2)$ means $o(t^2)/t^2 \to 0$ as $t\to 0$.
Substituting to write as a function of $t/\sqrt{n}$ gives
$$
  \varphi\!\left(\frac{t}{\sqrt{n}}\right) 
  = 1 - \frac{t^2}{2n} + o\!\left(\frac{t^2}{n}\right).
$$

Taking logs (which is valid for large enough\ $n$ since $\varphi(t/\sqrt{n}\,) \to 1$) and using
$\log(1 + u) = u + o(u)$ as $u \to 0$:
$$
  \log \varphi\!\left(\frac{t}{\sqrt{n}}\right) 
  = -\frac{t^2}{2n} + o\!\left(\frac{t^2}{n}\right).
$$
Therefore
\begin{align*}
  \log \varphi_{Z_n}(t)
  &= n \log \varphi\!\left(\frac{t}{\sqrt{n}}\right)\\
  &= n\left[-\frac{t^2}{2n} + o\!\left(\frac{t^2}{n}\right)\right]\\
  &= -\frac{t^2}{2} + n\cdot o\left(\frac{t^2}{n}\right)
  \to\, -\frac{t^2}{2}
\end{align*}
as $n \to \infty$, for each fixed $t \in \mathbb{R}$. 
Taking exponents,
$$
  \varphi_{Z_n}(t) \to \exp(-t^2/2) \quad \text{for all $t \in \mathbb{R}$}.
$$

The function $\exp(-t^2/2)$ is the CF of the $N(0, 1)$ distribution.
Since $\varphi_{Z_n}(t) \to \varphi(t)$ pointwise for all\ $t$, and the limit $\varphi(t)$ is continuous at $t = 0$, then $Z_n \overset{d}{\to} Z$, where\ $Z$ has characteristic function $\varphi(t)$, and $\overset{d}{\to}$ means 'converges in distribution'.
Since $\exp(-t^2/2)$ is continuous everywhere and is the CF of $N(0, 1)$, conclude that
$$
  Z_n \overset{d}{\to} N(0, 1).
$$



## Exercises

Selected answers appear in Sect.\ \@ref(AnswersChapSampling).

:::{.exercise #FPCF}
Consider the FPCF (Def.\ \@ref(def:FPC)).

1. What happens if the sample size is the same as the population size?
   Explain why this is a sensible result.
2. Show that, for large\ $n$, the FPCF is approximately $\sqrt{ 1 - n/N}$.
:::


:::{.exercise}
A random sample of size\ $81$ is taken from a population with mean\ $128$ and standard deviation\ $6.3$. 

1. What is the probability that an individual observation will fall between\ $126.6$ and\ $129.4$?
1. What is the probability that the *sample mean* will fall between\ $126.6$ and\ $129.4$?
1. What is the probability that the *sample mean* will *not* fall between\ $126.6$ and\ $129.4$?
:::


:::{.exercise #CLTProb}
Let $Y_1$, $Y_2$, $\dots$, $Y_n$ be $n$\ independent random variables, each with PDF
$$
   f_Y(y) = \begin{cases}
               3y^2 & \text{for $0\le y \le 1$};\\
               0 & \text{otherwise}.
            \end{cases}
$$


1. Determine the probability that a single observation will be within one standard deviation of the population mean.
1. Determine the probability that the sample mean will be within one standard deviation of the population mean, using the [*Central Limit Theorem*](#CentralLimitTheorem).
:::


:::{.exercise #EggsVariance}
Suppose the weights of eggs in a dozen carton have a weight that is normally distributed with mean\ $59\gs$ and variance\ $0.7\gs$.

1. Find the probability that, in a sample of $20$\ cartons,  the sample mean weight will exceed $59.5\gs$. 
1. Find the probability that a sample of twelve eggs will produce a sample variance of greater than\ $1$.
:::


:::{.exercise #EggsBroken}
In a carton of a dozen eggs, the number broken has a Poisson distribution with mean\ $0.2$. 

1. Find the probability that, in a sample of $20$\ cartons, the sample mean of the number of broken eggs per carton is more than one.
  (Use the *Central Limit Theorem*.)
2. Find the probability that, in any *single* carton, the probability that more than one egg is broken.
:::


:::{.exercise #ContRVM}
The random variable\ $M$ has the following probability density function
$$
    f_M(m) = 
    \begin{cases}
       3m^2 & \text{for $0 < m < 1$};\\
       0    & \text{otherwise}.
    \end{cases}
$$
A random sample of size\ $9$ is taken from the distribution, and the sample mean\ $\overline{M}$ is computed.

1. Compute the mean of\ $M$.
2. Compute the variance of\ $M$.
3. State the approximate distribution of\ $\overline{M}$ including the parameters.
4. Compute the probability that the sample mean will be within\ $0.1$ of the true mean.
:::


:::{.exercise #RandomNormalSize9}
Generate a random sample of size $n = 9$ from a $N(10, 36)$ distribution hundreds of times.
Obtain the mean and variance of\ $\sum X_i$ and\ $\overline{X}$ for each sample.

1. Verify that the distribution of the sample means is approximately normally distributed as expected.
1. Explain *why* this is expected.
:::


:::{.exercise #ABW}
Consider the random variable\ $A$, defined as
$$
    A = \frac{Z}{\sqrt{W/\nu}},
$$
where\ $Z$ has a standard normal distribution independent of\ $W$, which has a $\chi^2$-distribution with $\nu$\ degrees of freedom.
Also, consider the random variable\ $B$, defined as
$$
    B = \frac{W_1/\nu_1}{W_2/\nu_2},
$$
where\ $W_1$ and\ $W_2$ are independent $\chi^2$ variables with\ $\nu_1$ and\ $\nu_2$ degrees of freedom respectively.

1. Write down (do not derive) the distribution of\ $A$, including the parameters of the distribution.
1. Write down (do not derive) the distribution of\ $B$, including the parameters of the distribution.
1. Deduce that the distribution of\ $A^2$ is a special case of the distribution of\ $B$, and state the values of the parameters for which this is true.
:::


:::{.exercise #Pollution}
A manufacturing plant produces $2$\ tonnes of waste product on a given day, with a standard deviation of $0.2$\ tonnes per day. 
Find the probability that, over a $20$\ day period, the plant produces less than $25$\ tonnes of waste if daily productions can be assumed independent.
:::


:::{.exercise}
Let\ $Y$ be the *change* in depth of a river from one day to the next measured (in cms) at a specific location. 
Assume\ $Y$ is uniformly distributed for $y \in [-70, 70]$.

1. Find the probability that the mean change in depth for a period of $30$\ days will be greater than $10\cms$.
1. Use simulation to estimate the probability above.
:::


:::{.exercise #Typhoid}
The number of deaths per year due to typhoid fever is assumed to have a Poisson distribution with rate $\lambda = 4.1$ per year.

1. If deaths from year to year can be assumed to be independent, what is the distribution over a $20$\ year period?
1. Find the probability that there will be more than $100$\ deaths due to typhoid fever in period of $20$\ years.
:::


:::{.exercise #Lymphocyte}
The probability that a cell is a lymphocyte is\ $0.2$.

1. Write down an exact expression for the probability that in a sample containing $150$\ cells that at least\ $40$ are lymphocytes.
   Evaluate this expression using **R**.
1. Write down an approximate expression for this probability, and evaluate it.
:::



:::{.exercise}
Suppose the probability of a person aged $80+$ years dying after receiving influenza vaccine is\ $0.006$. 
In a sample of $200$\ persons aged $80+$\ years:

1. Write down an exact expression for the probability that more than\ $5$ will die after vaccination for influenza. 
   Evaluate this expression using **R**.
1. Write down an approximate expression for this probability and evaluate it.
1. If $4$\ persons died in a sample of\ $200$, what conclusion would you make about the probability of dying after vaccination?
   Justify your answer.
:::


:::{.exercise}
Illustrate the Central Limit Theorem for the uniform distribution on $[-1, 1]$ by simulation and repeated sampling, for various sample sizes.
:::



:::{.exercise}
Illustrate the Central Limit Theorem for the exponential distribution with mean\ $1$, by simulation and repeated sampling, for various sample sizes.
:::



:::{.exercise}
Demonstrate the Central Limit Theorem using **R**:

1. For the uniform distribution, where $-1 < x < 1$ (a symmetric distribution).
1. For the exponential distribution with parameter\ $1$ (a skewed distribution).
:::
    


:::{.exercise #SimTdist}
Consider a random variable\ $Z$ with a standard normal distribution $N(0, 1)$, and a random variable\ $V$ with a chi-squared distribution on\ $v$ degrees of freedom.

Simulate the distribution of
$$
  T = \frac{Z}{\sqrt{V/\nu}},
$$
then show this is a $t$-distribution with\ $\nu$ degrees of freedom.
*Hint*: use the **R** functions for the $t$-distribution (e.g., `dt()`) and the chi-squared distribution (e.g, `dchisq()`). 
:::



:::{.exercise #SimTdist2}
Consider a random variable\ $Z$ with a standard normal distribution $N(0, 1)$, and a random variable\ $V_\nu$ with a chi-squared distribution on\ $\nu$ degrees of freedom.

Simulate the distribution of
$$
  T = \frac{Z}{\sqrt{V_9/9}},
$$
then show this is a $t$-distribution with\ $\nu = 9$ degrees of freedom.
*Hint*: use the **R** functions for the $t$-distribution (e.g., `dt()`) and the chi-squared distribution (e.g, `dchisq()`). 
:::


:::{.exercise #SimHurdle}
Consider a zero-modified\index{Zero-modified distribution!Gamma modified} Gamma$(2,1)$-distribution, with $p = 0.5$.

1. Derive $\operatorname{E}[\overline{Y}_n]$ and $\operatorname{var}(\overline{Y}_n)$.
1. Simulate the distribution of $\operatorname{E}[\overline{Y}_n]$ for $n \in \{10, 50, 200\}$, overlayed with the appropriate normal approximation.
   Comment.
:::


:::{.exercise #SimCPG}
Consider a compound Poisson--gamma distributions\index{Compound Poisson--gamma distribution} with $\lambda = 1.5$, $\alpha = 2$ and $\beta = 1$.

1. Derive $\operatorname{E}[\overline{Y}_n]$ and $\operatorname{var}(\overline{Y}_n)$.
1. Simulate the distribution of $\operatorname{E}[\overline{Y}_n]$ for $n \in \{10, 50, 200\}$, overlayed with the appropriate normal  approximation.
   Comment.
:::


:::{.exercise #Unbiased1}
Let $X_1, \ldots, X_n \overset{\text{iid}}{\sim} \operatorname{Exponential}(\lambda)$, where\ $\lambda$ is the rate parameter so that $\operatorname{E}[X_i] = 1/\lambda$

1. Show that $\bar{X}$ is an unbiased estimator of $1/\lambda$.
2. Find an unbiased estimator of\ $\lambda$.
:::


:::{.exercise #Unbiased2}
Let $X_1, \ldots, X_n \overset{\text{iid}}{\sim} \operatorname{Uniform}(0, \theta)$. 

1. Show that $2\bar{X}$ is an unbiased estimator of\ $\theta$.
2. Let $M = \max(X_1, \ldots, X_n)$. 
   Derive the distribution of $M$ and show that\ $M$ is not unbiased for $\theta$.
3. Find a function of\ $M$ that is unbiased for\ $\theta$.
:::


:::{.exercise #UnbiasedVarianceMu}
Show that the estimator 
$$
  S^2_\mu = \frac{1}{n} \sum_{i=1}^n (X_i - \mu)^2.
$$
is an unbiased estimator of the population variance\ $\sigma^2$.
:::



:::{.exercise #UnbiasedVariance}
Show that the estimator 
$$
  S^2 = \frac{1}{n - 1} \sum_{i=1}^n (X_i - \overline{X})^2.
$$
is an unbiased estimator of the population variance\ $\sigma^2$.
:::



:::{.exercise #SumMean}
Prove Theorem\ \@ref(thm:SumMean).
:::


:::{.exercise #DistZ2}
Consider a random variable $X\sim N(0, 1)$.
Show that the distribution of $Z = X^2$ is a $\text{Gamma}(\alpha = 1/2. \beta = 2)$ distribution.
:::


:::{.exercise #SumZ2}
Consider $Z_i \overset{\text{iid}}{\sim}\sim \text{Gamma}(\alpha=1/2. \beta2)$ for $i = 1, 2, \dots, n$.

Use the MGF method (Sect.\ \@ref(TransformationMoments)) to show that the distribution of $\sum_{i=1}^n Z_i^2$ is a $\text{Gamma}(\alpha = n/2, \beta = 2)$ distribution.
:::


:::{.exercise #MeanSdIndependent}
In this exercise, the relationship between teh samnple meand and sample variance is explored (Sect.\ \@ref(ChiSquaredDistribution)).

Using **R**, generate\ $1000$ simulations of data\ $X_i \sim N(1, 3)$ for $i = 1, 2, dots n$ for $n = 50$.
For each of the $1000$\ simulations, compute the sample mean and standard deviation.

1. Plot the sample means against the sample variances.
   Comment.
2. Compute the correlation between the sample means against the sample variances.
   Comment.
3. Show that the covariance between the sample mean $\bar{X}$ and the deviations of the observations from the sample means $\bar{X} - X_i$ is zero.
:::


:::{.exercise #ChisqProb}
The variable\ $X$ has a chi-squared distribution with $12$\ df.
Determine the value of\ $X$ below which lies 90% of the distribution.
:::


:::{.exercise #ChisqProb2}
A random sample of size $n = 10$ is selected from the $N(20, 5)$ distribution.
What is the probability that the sample variance exceeds\ $10$?
:::

