# (PART) Sampling distributions {-}

# Describing samples {#SamplingDistributions}

::: {.objectivesBox .objectives data-latex="{iconmonstr-target-4-240.png}"}

* use appropriate techniques to determine the sampling distributions of\ $t$, $F$, and\ $\chi^2$\ distributions.
* explain how the above distributions are related to the normal distribution.
* apply the concepts of the Central Limit Theorem in appropriate circumstances.
* use the Central Limit Theorem to approximate binomial probabilities by normal probabilities in appropriate circumstances.
:::




## Introduction {#ChapterSamplingIntro}

This chapter introduces the application of distribution theory to the discipline of statistics.
Distribution theory is about describing random variables in theoretical population using probability, while statistics is about data collection and extracting information from data (Chap.\ \@ref(ChapterIntroduction)). 


## Populations and samples {#PopulationsSamples}

### From populations to observations {#Sampling}

\index{Sample|(}
Until now, theoretical probability distributions describing ideal distributions of infinite populations have been our focus.
In practice, of course, infinite populations do not exist, and finite (usually relatively small) samples from unknown distributions are observed.\index{Sample}


:::{.example #Heights name="Populations and models"}
Suppose our *population* of interest is Australian females, and we are interested in their heights.
'Height of Australian females' is a continuous random variable, say\ $X$.

We may decide to model their heights theoretically using some specified distribution, with the probability density function $f_X(x)$.
Using this model, the mean height of Australian females is modelled as (Def.\ 
\@ref(def:Expectation))
$$
  \mu = \operatorname{E}[X] = \int_{-\infty}^\infty x\cdot f_X(x)\,dx.
$$

If we model their heights using the normal distribution $N(\mu = 173, \sigma^2 = 9)$, the mean is modelled as\ $173\cms$.
:::


In practice however, height measurements of individual Australian females are *not* made with infinite precision, and measured heights must be rounded as hence recorded as discrete values.
Even though the random variable 'Height' is continuous, any tool to measure actual heights must record a *rounded* version of the heights, to some minimum unit\ $\delta$.
Heights, for example, may be rounded to the nearest centimetre (i.e., $\delta = 1\cms$) or the nearest millimetre (i.e., $\delta = 0.1\cms$).
So even if it was possible to measure the height of every Australian female, we would in practice have a list of discrete values $\{x_1, x_2, \dots , x_N\}$, where\ $N$ is the size of the population.

So, these observed heights are actually discrete (rounded) values, and the mean would be computed using (Def.\ 
\@ref(def:Expectation))
$$
  \mu 
  = \operatorname{E}[X]
  = \frac{1}{N} \sum_{i = 1}^N x_i.
$$
As $N \to \infty$ and as $\delta \to 0$, this discrete sum converges to the Riemann integral:
$$
  \mu
  =
  \lim_{\substack{\delta \to 0 \\ N \to \infty}} \frac{1}{N} \sum_{i = 1}^N x_i 
  = 
  \int_{-\infty}^\infty x\cdot f_X(x)\,dx.
$$


### Random sampling and statistics {#RandomSamples}

In practice, entire populations are very rarely studied: the population may to be too large, or all members may be inaccessible.
So rather than studying entire populations, a *sample* of observations from that population are studied.

The task of the *inferential statistics*\index{Inferential statistics}\index{Statistic} is to attempt to understand the entire population, based on what is learnt from studying a small subset of that population: the *sample*. \index{Sample}


:::{.definition #RandomSample name="Sample"}
A subset of observations from the population is called a sample.\index{Sample}
:::


This definition is very broad: *any* subset from the population is a *sample*.
Importantly, *how* the sample is drawn determines its statistical properties; the most important type of sample for statistical inference is the idea of a *random sample*.\index{Sample!random}


:::{.definition #RandomSample name="Random sample"}
The set of random variables $X_1, X_2, \dots, X_n$ from a distribution\ $D$ is a collection of random variables $X_1, X_2, \dots, X_n$ that are independently and identically distributed (iid) with distribution function $F_X(x)$.

We write $X_1, X_2, \dots, X_n \sim D$.
:::


The definition uses the distribution function and so applies to discrete, continuous and mixed random variables.
Because the elements of the random sample are assumed to be independent,\index{Independence} the results in previous chapters that are based on independence therefore apply.

Once the sample is obtained, the data can be used to compute the value of some *statistic*.
A *statistic*\index{Statistic} is simply a numerical value that is computed from a sample.


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
Since the individuals in the sample can vary, the value of a statistic can vary from sample to sample.
:::


:::{.definition #Statistic name="Statistic"}
Let $X_1, X_2, \ldots, X_n$ be a random sample. 
A *statistic* is any function $T = g(X_1, X_2, \ldots, X_n)$ of the sample.\index{Statistic}
:::

:::{.example #StatisticdsExample name="Statistic"}
For a sample $X_1, X_2, \dots, X_n$, we could define a statistic\ $M$ such that
$$
  M = \left(\max\{X_1, X_2, \dots, X_n\} - \min\{X_1, X_2, \dots, X_n\}\right) / 2
$$
:::


An *estimator*\index{Estimator} is a *statistic* with a specific purpose: to estimate the unknown value of a population parameter.
As a statistic, an estimator is a random variable.
An *estimate*\index{Estimate} is the actual numerical value of the estimator that is computed after observing data.
We often say that an estimate is a *realisation* of an estimator.


:::{.example #StatisticsExampleMin name="The minimum is a statistic"}
Suppose we take a sample of $n = 12$ Australian females, and measure their heights to the nearest centimetre.

We could define the *statistic*\index{Statistic}
$$
  X_{(1)} = \min(\{X_1, X_2, \dots, X_N\}).
$$
This statistic computes the minimum value of the sample.
If the purpose of computing the sample minimum is to estimate the minimum value in the population, then $X_{(1)}$ is an *estimator* of the population minimum.
Since $X_{(1)}$ is likely to be different when computed from another sample of $12$\ Australian women, it is a random variable. 

If the smallest height in the sample is $144\cms$, then
$$
  x_{(1)} = 144
$$
is an *estimate* of the population minimum.
:::
\index{Sample|)}


## Estimators and estimates {#SamplingDistributions}

### Sampling distributions: a small example {#SamplingDistributionsSmallExample}

\index{Sampling distribution|(}
Since a statistic is a function of the random variables $X_1, X_2, \ldots, X_n$, *sample statistics are random variables*.
In other words, the value of a sample statistic changes from sample to sample, and so *sample statistics have a distribution* that describes *how* they vary.


:::{.definition #SamplingDistributionStatistic name="Sampling distribution of a statistic"}
The *sampling distribution of a statistic* is the theoretical probability distribution of the statistic associated with *all* possible samples of a particular size drawn from a particular population.
:::


Typically, a population will be *very* large, and listing all elements of the population is impossible.
However, for demonstration purposes, suppose the population contains only $N = 5$ discrete elements:
$$
   22\qquad 23\qquad 27\qquad 28\qquad 38.
$$
The mean of the population is
$$
  \mu = \frac{22 + 23 + 27 + 28 + 38}{5} = 27.6.
$$
and the minimum value in the population is\ $2$.
These two values are called *parameters*\index{Parameter} as they are numbers that describe the population.

The minimum value of any sample of size $n < 5$ drawn from the population can be found: the sample minimum is a *statistic*, as it is computed from the sample.
Since the population is (artificially) small in this scenario, the sampling distribution can be determined by enumerating every possible sample of any given size\ $n$.

Using **R**:

```{r}
Population <- c(22, 23, 27, 28, 38)
min_P <- min(Population)
min_P
```

Suppose we consider samples of size $n = 2$; there are $\binom{5}{2} = 10$ such samples:

```{r}
combn(Population, 2)
```

The minimum of each sample is easily computed also:

```{r}
min_S <- combn(Population, 2, FUN = min)
min_S
```

The sampling distribution is just the distribution of these sample minimums, which can be displayed a table (or graph):

```{r}
table(min_S)
```

This a *sampling distribution*.
The population minimum is\ $22$, so the sample minimum is often a poor estimate of the population minimum.

Suppose we looked at larger sample size; say, sample of size $n = 4$:

```{r}
min_S <- combn(Population, 4, FUN = min)
table(min_S)
```

The population minimum is\ $22$, so the sample minimum when $n = 4$ is often a very good estimate of the population minimum.
The larger sample size produces a better estimate, in general, and the value of the sample statistic varies less from sample to sample.


### Larger populations

For larger populations, enumerating every possible is sample implausible or impossible, and obtaining the selected elements is usually impractical or impossible.
Nonetheless, the idea of a sampling distribution of a statistic can be *simulated* using\ **R**.


:::{.example}
Consider taking a random samples of $n = 5$ observations from a $N(\mu = 173, \sigma^2 = 9)$ distribution, and computing the minimum value in the sample:

```{r echo=FALSE}
set.seed(1592875)
```
```{r}
x_Mn <- 173; x_Sd <- 3        # Population (model) mean, std dev.
sample_Size <- 5              # Sample size is 5
x1 <- rnorm(sample_Size,      # Take a sample of this size
            mean = x_Mn, sd = x_Sd)
min(x1)    # Compute the sample minimum for this sample
```

We can use\ **R** to simulate selecting many samples of size $n=5$, and compute the minimum for each (Fig.\ \@ref(min:SDminPlots), left panel):

```{r SDmin1, fig.show='hide'}
num_Sims <- 100000           # Number of simulations

# Generate random values
Xvals <- rnorm(num_Sims * sample_Size,
               mean = x_Mn, sd = x_Sd)

# Place these into an array, with each sample as a column 
Xvals <- array(Xvals,
               dim = c(sample_Size, num_Sims) )

# Find the minimum of each sample (i.e., column)
Xmin   <- apply(Xvals, 
                MARGIN = 2, 
                FUN = min) 
hist(Xmin)          # Create the histogram of sample minimums
```

Suppose we take larger samples, and find the sampling distribution again (Fig.\ \@ref(fig:SDminPlots), right panel):

```{r SDmin2, fig.show='hide'}
sample_Size2 <- 100

# Generate random values
Xvals <- rnorm(num_Sims * sample_Size2,
               mean = x_Mn, sd = x_Sd)

# Place these into an array, with each sample as a column 
Xvals <- array(Xvals,
               dim = c(sample_Size2, num_Sims) )

# Find the minimum of each sample (i.e., column)
Xmin   <- apply(Xvals, 
                MARGIN = 2, 
                FUN = min) 
hist(Xmin)          # Create the histogram of sample minimums
```

:::

```{r SDminPlots, echo=FALSE, out.width='90%', fig.width=8.5, fig.height = 3.5, fig.align="center", fig.cap="Sampling distributions for the sample minimums, for samples of size\ $5$ (left) and size\ $100$ (right)."}
par( mfrow = c(1, 2))

set.seed(1592875)

sample_Size <- 5

x_Mn <- 173
x_Sd <- 3
num_Sims <- 10000          # Number of simulations

#####

# Generate random values
Xvals <- rnorm(num_Sims * sample_Size,
               mean = x_Mn, sd = x_Sd)

# Place these into an array, with each sample as a column 
Xvals <- array(Xvals,
               dim = c(sample_Size, num_Sims) )

# Find the minimum of each sample (i.e., column)
Xmin   <- apply(Xvals, 
                MARGIN = 2, 
                FUN = min) 

hist(Xmin, 
     xlim = c(155, 180),
     xlab = "Sample minimum",
     ylab = "",
     breaks = seq(150, 180, by = 1),
     main = "Sampling minimums:\nsamples of size 25",
     axes = FALSE)
axis(side = 1)

#####

sample_Size2 <- 100

Xvals2 <- rnorm(num_Sims * sample_Size2,
                mean = x_Mn, sd = x_Sd)
Xvals2 <- array(Xvals2,
                dim = c(sample_Size2, num_Sims) )
Xmin2  <- apply(Xvals2, 
                MARGIN = 2, 
                FUN = "min") 

hist(Xmin2, 
     xlim = c(155, 180),
     xlab = "Sample minimum",
     ylab = "",
     breaks = seq(150, 180, by = 1),
     main = "Sampling minimums:\nsamples of size 100",
     axes = FALSE)
axis(side = 1)
```

These histograms display the sampling distributions of the sample minimums for samples of size $n = 5$ and $n = 100$.
The value of the sample statistic varies less from sample to sample again.


### Estimating the population mean {#SampligDistributionMean}

Using a sample to estimate a minimum value is not often helpful; a more useful statistic is the sample mean or the sample variance.
But how should the population mean be estimated from a sample?

For a population of\ $N$ discretely measured observed values $X_1, X_2, \dots, X_n$, the population mean is
$$
  \mu = \frac{1}{N}\sum_{i = 1}^N x_i.
$$
The obvious suggestion for computing a sample mean from a sample of size\ $n$ (i.e., $x_1, x_2, \dots, x_n$) is *sample mean* $\overline{X}$, an *estimator*,\index{Estimator} defined as
$$
  \overline{X} = \frac{1}{n}\sum_{i = 1}^n X_i.
$$

We can use\ **R** to simulate selecting many samples of size $n = 5$, and many sample of size $n = 100$, and compute the sample mean for each, using the same samples created above (Fig.\ \@ref(fig:SDmeanPlots)):

```{r SDmean, fig.show='hide'}
# Mean: n = 5
Xmean   <- apply(Xvals, 
                 MARGIN = 2, 
                 FUN = mean) 
hist(Xmean)          # Create the histogram of sample minimums

# Mean: n = 100
Xmean2   <- apply(Xvals2, 
                 MARGIN = 2, 
                 FUN = mean) 
hist(Xmean2)          # Create the histogram of sample minimums
```


```{r SDmeanPlots, echo=FALSE, out.width='90%', fig.width=8.5, fig.height = 3.5, fig.align="center", fig.cap="Sampling distributions for the sample mean for sample of size $n = 5$ (left) and $n = 100$ (right). The solid dots represent the population mean."}
par( mfrow = c(1, 2))

hist(Xmean, 
     xlim = c(165, 180),
     xlab = "Sample mean",
     ylab = "",
     breaks = seq(150, 180, by = 1),
     main = "Sampling mean:\nsamples of size 5",
     sub = paste0("(Average value: ", round(mean(Xmean), 1), ")"),
     axes = FALSE)
axis(side = 1)
points(x = x_Mn, 
       y = 0,
       pch = 19)



#####

hist(Xmean2, 
     xlim = c(165, 180),
     xlab = "Sample mean",
     ylab = "",
     #breaks = seq(150, 180, by = 1),
     main = "Sampling mean:\nsamples of size 100",
     sub = paste0("(Average value: ", round(mean(Xmean2), 1), ")"),
     axes = FALSE)
axis(side = 1)
points(x = x_Mn, 
       y = 0,
       pch = 19)


```

The sample estimates of the mean appear, on average, to be good estimates of the population mean.
The larger sample produces better estimates of the population mean in general, and less variation in the values of the sample statistic.


### Unbiased estimators {#UnbiasedEstimators}

It seems sensible that, on average, a sample estimate would produce the population value.
Sometimes the sample estimate may be a little smaller than the population value, and sometimes a bit larger, but on average the value of the sample estimate is the value of the population parameter.
Any such estimator is called *unbiased*.


:::{.definition #Unbiased name="Unbiased estimator"}
A statistic\ $\hat{\theta}$ is an *unbiased estimator*\index{Estimator!unbiased} of a parameter\ $\theta$ if
$$
  \operatorname{E}[\hat{\theta}] = \theta.
$$
:::


This means that an unbiased estimator from a *random* sample produces estimates that, *on average*, have the same value as the population value being estimated.
This seems a sensible property for a 'good' estimator.
(There are other properties of 'good' estimators too, we but do not explore these here. 
This topic is a whole branch of statistics called *inferential statistics*.)\index{Inferential statistics}
When we say that\ $\overline{X}$ is an unbiased estimator of\ $\mu$ then, we mean that $\operatorname{E}[\overline{X}] = \mu$,



:::{.theorem #SampleMeanUnbiased name="An unbiased estimator for $\mu$"}
The sample mean
$$
  \overline{X} = \frac{1}{n} \sum_{i = 1}^n X_i
$$
is an unbiased estimator\index{Estimator!unbiased}\index{Estimator!sample mean} of the population mean\ $\mu$.

::::{.proof}
Using the definition of an unbiased estimator:
\begin{align*}
  \operatorname{E}[\overline{X}] 
  &= \operatorname{E}\left[ \frac{1}{n}(X_1 + X_2 + \cdots X_n) \right]\\
  &= \frac{1}{n} \operatorname{E}[ X_1 + X_2 + \cdots X_n ]\\
  &= \frac{1}{n} \left( \operatorname{E}[ X_1 ] + \operatorname{E}[X_2] + \cdots \operatorname{E}[X_n] \right) \\
  &= \frac{1}{n} (n\times \mu) = \mu.
\end{align*}
That is, the expected value of\ $\overline{X}$ is equal to\ $\mu$.
The estimator\ $\overline{X}$ is an *unbiased estimator* of\ $\mu$.
::::
:::


### Estimating the population variance {#EstimatePopVariance}

\index{Variance!estimation}
Given the formula for the population variance, the following estimate of the population variance is suggested.


:::{.theorem #SampleVarianceUnbiased name="An unbiased estimator for $\sigma^2$"}
The sample variance
\begin{equation}
  S^2_\mu = \frac{1}{n} \sum_{i=1}^n (X_i - \mu)^2.
   (\#eq:S2WithPopMean)
\end{equation}
is an unbiased estimator\index{Estimator!unbiased} of the population variance\ $\sigma^2$.

::::{.proof}
Again starting with the definition of an unbiased estimator:
\begin{align*}
    \operatorname{E}[S^2_n] 
    &= \frac{1}{n} \operatorname{E}\left[ (X_1 - \mu)^2 + (X_2 - \mu)^2 + \cdots + (X_n - \mu)^2\right]\\
    &= \frac{1}{n} \left( \operatorname{E}[ (X_1 - \mu)^2] + \operatorname{E}[(X_2 - \mu)^2] + \cdots + \operatorname{E}[X_n - \mu]^2\right)\\
    &= \frac{1}{n} \left( \sigma^2 + \sigma^2 + \cdots + \sigma^2 \right)\\
    &= \frac{1}{n} \times n\sigma^2 = \sigma^2.
\end{align*}
Thus, $S^2_n$ is an unbiased estimator of $\sigma^2$.
::::
:::


While $S^2_\mu$ is an unbiased estimator of\ $\sigma^2$, Eq.\ \@ref(eq:S2WithPopMean) has a practical difficulty: its computation relies on *knowing* the value of the population mean\ $\mu$.
Knowing the value of\ $\mu$ is highly unlikely; knowing the value of\ $\mu$ but not the value of $\sigma^2$ almost never occurs.
Almost always, we would only have an estimate\ $\overline{X}$ of the population mean\ $\mu$.

This suggests using the estimator
$$
  S^2_n = \frac{1}{n} \sum_{i=1}^n (X_i - \overline{X})^2.
$$
to estimate the population variance.
To determine if this is an unbiased estimator, proceed as usual:
\begin{align*}
    \operatorname{E}[S^2_n] 
  &= \operatorname{E}\left[ \frac{1}{n} \sum_{i=1}^n (X_i - \overline{X})^2 \right]\\
  &= \frac{1}{n}\operatorname{E}\left[ \sum_{i=1}^n (X_i - \overline{X})^2\right]\\
  &= \frac{1}{n}\operatorname{E}\left[ \sum_{i=1}^n \left(X^2_i - 2X_i\overline{X} + \overline{X}^2\right)\right]\\
  &= \frac{1}{n}\operatorname{E}\left[ \sum_{i=1}^n X^2_i - \overline{X}\sum_{i=1}^n 2X_i + \sum_{i=1}^n \overline{X}^2\right].
\end{align*}
Now observe that since $\overline{X} = \sum_i X_i/n$, then $\sum_i X_i = n \overline{X}$; hence
\begin{align*}
     \operatorname{E}[S^2_n] 
  &= \operatorname{E}\left[ \frac{1}{n} \sum_{i=1}^n (X_i - \overline{X})^2 \right]\\
  &= \frac{1}{n}\operatorname{E}\left[ \sum_{i=1}^n X^2_i - \overline{X}^2 + n \overline{X}^2\right]\\
  &= \frac{1}{n}\operatorname{E}\left[ \sum_{i=1}^n X^2_i - n \overline{X}^2 \right]\\
  &= \frac{1}{n}\left(\operatorname{E}\left[ \sum_{i=1}^n X^2_i\right] - n\operatorname{E}\left[ \overline{X}^2 \right]\right).
\end{align*}
This expression can be simplified by noting that
$$
  \operatorname{var}[\overline{X}]
  = \operatorname{E}[\overline{X}^2] - \operatorname{E}[\overline{X}]^2 
  = \operatorname{E}[\overline{X}^2] - \mu^2, 
$$
and so $\operatorname{E}[\overline{X}^2] = \operatorname{var}(\overline{X}) + \mu^2 = \sigma^2/n + \mu^2$.
Also, 
$$
  \operatorname{var}[X] 
  = \operatorname{E}[X^2] - \operatorname{E}[X]^2 
  = \operatorname{E}[X^2] - \mu^2,
$$
so that $\operatorname{E}[X^2] = \sigma^2 + \mu^2$.
Hence,
\begin{align*}
     \operatorname{E}[S^2_n] 
  &= \operatorname{E}\left[ \frac{1}{n} \sum_{i=1}^n (X_i - \overline{X})^2 \right]\\
  &= \frac{1}{n}\left(\operatorname{E}\left[ \sum_{i=1}^n X^2_i\right] - n\operatorname{E}\left[\sum_{i=1}^n \overline{X}^2 \right]\right)\\
  &= \frac{1}{n}\left\{ \sum(\sigma^2 + \mu^2) - n\left(\frac{\sigma^2}{n} + \mu^2\right)\right\}\\
  &= \frac{1}{n}\left\{ n\sigma^2 + n\mu^2 - \sigma^2 - n\mu^2\right\}\\
  &= \frac{1}{n}\left\{ n\sigma^2  - \sigma^2\right\}\\
  &= \frac{n - 1}{n} \sigma^2.
\end{align*}
This shows that $\operatorname{E}[S^2_n] \ne\sigma^2$, and so the estimator is a *biased estimator* of\ $\sigma^2$.
Since\ $\overline{X}$ estimates\ $\mu$ imprecisely (from just one of the many possible samples), an extra source of variation is introduced into the estimation of\ $\sigma^2$.



```{r SDvarPlots, echo=FALSE, out.width='90%', fig.width=8.5, fig.height = 3.5, fig.align="center", fig.cap="Sampling distributions for the sample variance for sample of size $n = 5$ (left) and $n = 100$ (right). The solid dots represent the population variance."}
par( mfrow = c(1, 2))

Xvar   <- apply(Xvals, 
                MARGIN = 2, 
                FUN = var) 
Xvar2   <- apply(Xvals2, 
                 MARGIN = 2, 
                 FUN = var) 
xRange <- c(1, 2)

hist(Xvar, 
#     xlim = xRange,
     xlab = "Sample variance",
     ylab = "",
     #breaks = seq(150, 180, by = 1),
     main = "Sample variance:\nsamples of size 5",
     sub = paste0("(Average value: ", round(mean(Xvar), 3), ")"),
     axes = FALSE)
axis(side = 1)
points(x = x_Sd^2, 
       y = 0,
       pch = 19)



#####

hist(Xvar2, 
#     xlim = c(165, 180),
     xlab = "Sample variance",
     ylab = "",
     #breaks = seq(150, 180, by = 1),
     main = "Sample variance:\nsamples of size 100",
     sub = paste0("(Average value: ", round(mean(Xvar2), 3), ")"),
     axes = FALSE)
axis(side = 1)
points(x = x_Sd^2, 
       y = 0,
       pch = 19)

```



We could, however, multiply $S^2_n$ by $(n - 1)/n$, which *would* produce an unbiased estimator.



:::{.theorem #SampleVarianceUnbiasedXbar name="An unbiased estimator for $\sigma^2$ ($\mu$ unknown)"}
When the value of\ $\mu$ is unknown, an *unbiased* estimator of\ $\sigma^2$ is 
\begin{equation}
  S^2 = \frac{1}{n - 1} \sum_{i=1}^n (X_i - \overline{X})^2.
   (\#eq:S2WithoutPopMean)
\end{equation}
Notice that the divisor in front of the summation is now\ $n - 1$ rather than\ $n$. 

::::{.proof}
Easily shown by adapting the proof for $S^2_n$ above.
::::
:::


This shows that not all estimators\index{Estimator} are unbiased estimators; even estimators that appear to be 'obvious' estimators may not be unbiased estimators.\index{Estimator!unbiased}


:::{.example #UnbiasedPoisson name="A unbiased estimate"}
Let $X_1, \ldots, X_n \overset{\text{iid}}{\sim} \operatorname{Poisson}(\lambda)$, and consider the estimate $\bar{X}$.
Because the expectation is a linear operator, and because $\operatorname{E}[X_i] = \lambda$, 
$$
  \operatorname{E}[\bar{X}] 
  = 
  \operatorname{E}\!\left[\frac{1}{n}\sum_{i=1}^n X_i\right] = \frac{1}{n}\sum_{i=1}^n \operatorname{E}[X_i] = \frac{1}{n} \cdot n\lambda = \lambda.
$$
Hence $\bar{X}$ is an unbiased estimator of\ $\lambda$ for a Poisson distribution.
:::


:::{.example #BiasedEstimate name="A biased estimate"}
Let $X_1, \ldots, X_n \overset{\text{iid}}{\sim} \operatorname{Exponential}(\lambda)$, where $\lambda > 0$ is the rate parameter, so that $\operatorname{E}[X_i] = 1/\lambda$.
Consider the natural estimator of $\lambda$: that is, $1/\bar{X}$. 

Since $X_i \sim \operatorname{Exponential}(\lambda)$, we have $\bar{X} = \frac{1}{n}\sum_{i=1}^n X_i$, and it is known that
$$
  \bar{X} \sim \operatorname{Gamma}(n,\, n\lambda),
$$
where we use the shape--rate parameterisation, so $\operatorname{E}[\bar{X}] = 1/\lambda$ and $\operatorname{Var}(\bar{X}) = 1/(n\lambda^2)$.

$\operatorname{E}[1/\bar{X}]$ can be computed directly:
$$
  \operatorname{E}\!\left[\frac{1}{\bar{X}}\right]
  = \int_0^\infty \frac{1}{x} \cdot
    \frac{(n\lambda)^n}{\Gamma(n)}\,x^{n-1}e^{-n\lambda x}\,dx
  = \frac{(n\lambda)^n}{\Gamma(n)}
    \int_0^\infty x^{n-2}e^{-n\lambda x}\,dx.
$$
The integral is related to the probability density function of a gamma distribution (Def.\ \@ref(def:GammaDistribution)) with shape $n - 1$ and rate $n\lambda$, and so
$$
  \int_0^\infty x^{n-2}e^{-n\lambda x}\,dx
  = 
  \frac{\Gamma(n-1)}{(n\lambda)^{n-1}},
$$
valid for $n > 1$. 
Substituting then,
$$
  \operatorname{E}\!\left[\frac{1}{\bar{X}}\right]
  = \frac{(n\lambda)^n}{\Gamma(n)} \cdot \frac{\Gamma(n - 1)}{(n\lambda)^{n - 1}}
  = n\lambda \cdot \frac{\Gamma(n - 1)}{\Gamma(n)}
  = n\lambda \cdot \frac{1}{n-1}
  = \frac{n\lambda}{n - 1},
$$
using $\Gamma(n) = (n - 1)\,\Gamma(n-1)$. 
Hence
$$
  \operatorname{E}\!\left[\frac{1}{\bar{X}}\right]
  = \frac{n}{n - 1}\,\lambda
  = \lambda + \frac{\lambda}{n - 1},
$$
and so the bias is $\lambda/(n - 1) > 0$: the estimator $1/\bar{X}$ systematically *overestimates* the value of $\lambda$. 

Note also that the bias $\lambda/(n - 1) \to 0$ as $n \to \infty$, so $1/\bar{X}$ is *asymptotically unbiased*.\index{Unbiased estimator!asymptotically}
:::


## Sampling distributions {#SamplingDistributionsTheory}

### Introduction

In previous sections, estimators were introduced and studied, specifically unbiased estimators.
All estimators are random variables, and so they have distributions. 
Previously, we used simulations to display sampling distributions of various statistics.

However, theoretical sampling distributions of various estimators can be found, which leads to numerous very powerful and practical results.


### The sampling distribution of the mean {#SamplingDistributionMean}

Usually the sample mean\ $\overline{X}$ is of greatest interest when sampling.
Section\ \@ref(SamplingDistributionsSmallExample) shows that the sample mean, like all estimators, varies from sample to sample.
Figure\ \@ref(fig:SDmeanPlots) suggests that larger sample sizes produce sampling distributions with smaller variances.
The following theorem quantifies this idea.


:::{.theorem #SamplingDistMean name="Sampling distribution of the mean"}
If $X_1, X_2, \dots, X_n$ is a *random sample* of size\ $n$ from a population with mean\ $\mu$ and variance\ $\sigma^2$, then the sample mean\ $\overline{X}$ has a sampling distribution with mean $\mu$ and variance $\sigma^2 / n$.

::::{.proof}
This is a application of Corollary\ \@ref(cor:IID) with $a_i = 1/n$.
::::
:::


Theorem\ \@ref(thm:SamplingDistMean) applies to a random sample as defined in Sect.\ \@ref(RandomSamples).
The remarkable feature of Theorem\ \@ref(thm:SamplingDistMean) is that the distribution of the population\ $X$ is not important to the result.


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
In Theorem\ \@ref(thm:SamplingDistMean), the distribution of the population from which the sample is drawn is irrelevant.
:::


Theorem\ \@ref(thm:SamplingDistMean) describes the *location* and *spread* of the sampling distribution of the mean, but not the *shape* of the sampling distribution.
Remarkably, the *shape* of the sampling distribution only depends on the shape of the population distribution *when the sample is small* (Sect.\ \@ref(CentralLimitTheorem)).

Theorem\ \@ref(thm:SamplingDistMean) is remarkable in that it makes no assumption about the shape of the population distribution: the result holds whether\ $X$ is normal, Poisson, exponential, or anything else. 
However, the theorem tells us the *mean* and *variance* of the sampling distribution of\ $\overline{X}$, but says nothing about its *shape*. 

Describing the shape, and deriving the sampling distribution of $S^2$ requires additional assumptions. When the population is normally distributed, exact results are available, and these are the subject of Sect.\ \@ref(SamplingDistributionsRelatedToNormal). 
When the population is *not* normally distributed, the Central Limit Theorem (Sect.\ \@ref(CentralLimitTheorem)) provides an approximate answer for large\ $n$.


```{r, fig.show='hide'}
set.seed(123)

ns <- c(5, 15, 100)
lambda <- 4
B <- 10000  # number of simulations

sim_means <- lapply(ns, function(n) {
  replicate(B, mean(rexp(n, rate = lambda)))
})

hist(sim_means[[1]]) 
hist(sim_means[[2]])
hist(sim_means[[3]])
```


```{r ExpSD, echo=FALSE, fig.align="center", fig.cap="Sampling distributions, for samples drawn from an exponential distribution with $\\lambda = 4$, for various sample sizes.", out.width='100%', fig.width=8, fig.height=3.2}
par(mfrow = c(1, 3))

hist(sim_means[[1]], 
     axes = FALSE,
     main = "Sample means:\nn = 5", 
     ylab = "",
     xlab = "Sample means")
axis(side = 1)

hist(sim_means[[2]], 
     axes = FALSE,
     main = "Sample means:\nn = 15", 
     ylab = "",
     xlab = "Sample mean")
axis(side = 1)

hist(sim_means[[3]], 
     axes = FALSE,
     main = "Sample means:\nn = 100", 
     ylab = "",
     xlab = "Sample mean")
axis(side = 1)


```





### Sampling distributions: when $X_i$ have a normal distribution {#SamplingDistributionsRelatedToNormal}

The normal distribution is used to model many naturally occurring phenomena (Sect.\ \@ref(Normal)). 
Consequently, we begin by studying the sampling distributions of statistics of random samples that come from normal populations.

First, we present a useful result.


:::{.theorem #LinearComb name="Linear combinations"}
Let $X_1, X_2, \dots, X_n$ be a set of independent random variables where $X_i\sim N(\mu_i,\sigma^2_i)$. 
Define the linear combination\ $Y$ as
$$
   Y = a_1 X_1 + a_2 X_2 + \dots + a_nX_n.
$$
Then $Y \sim N(\Sigma a_i\mu_i, \Sigma a^2_i\sigma^2_i)$.

::::{.proof}
In Theorem\ \@ref(thm:NormalProperties), the MGF of the random variable\ $X_i$ was shown to be
$$
   M_{X_i} (t)
   = \exp\left\{\mu_i t + \frac{1}{2} \sigma^2_i t^2\right\}\quad\text{for $i = 1, 2, \dots, n$}.
$$
So, for a constant\ $a_i$ (using Theorem\ \@ref(thm:ExpectationLinear) with $\beta = 0$, $\alpha = a_i$):
\begin{align*}
   M_{a_i X_i}(t)
   &= M_{X_i}(a_it)\\
   &= \exp\left\{\mu_i a_i t + \frac{1}{2} \sigma^2_i a^2_i t^2\right\}.
\end{align*}
Since the MGF of a sum of independent random variables is equal to the product of their MGFs,
\begin{align*}
     M_Y(t)   
     &= \prod^n_{i = 1} \exp\left( \mu_i a_i t + \frac{1}{2} \sigma^2_i a^2_i t^2 \right)\\
     &= \exp\left( t\Sigma a_i\mu_i +\frac{1}{2} t^2 \Sigma a^2_i \sigma^2_i \right).
\end{align*}
This is the MGF of a normal random variable with mean $\operatorname{E}[Y] = \Sigma a_i\mu_i$ and variance $\operatorname{var}[Y] = \Sigma a^2_i\sigma^2_i$.
::::
:::


:::{.example}
Theorem\ \@ref(thm:LinearComb) can be demonstrated in **R**; see below.
:::


```{r}
set.seed(8979704) # For reproducibility
numberVars <- 1000 # That is: n

# The values of a_i, the coefficients in the linear combination
ai <- sample(x = seq(-2, 3),    # We select values of a_i between -2 and 3
             size = numberVars, # Need one for each X value
             replace = TRUE)    # Select with replacement

# The means of the individual distributions
mui <- runif(numberVars, # We take means from a uniform dist; one for each X
             min = -5,   # Min value is -5
             max = 5)    # Max value is 5
# The variances of the distributions
sigma2i <- runif(numberVars,  # Variance from a uniform distn also
                min = 1,    # Minimum values of 1
                max = 3)    # Maximum value of 3

# Compute the values directly:
Xi <- rnorm(numberVars,      # X_i with 
            mean = mui,   # means 
            sd = sqrt(sigma2i) )

# What we expect for Y = a_1 X_2 + a_2 X_2 + ...
# according to the Theorem:
Y <- ai * Xi  # The linear combination Y

# Show some:
someInfo <- cbind(ai, mui, sigma2i, Xi, Y)
someInfo[1:5, ]


# Compare:
compareTable <- array( dim = c(2, 2) )
colnames(compareTable) <- c("Means:",
                            "Variances:")
rownames(compareTable) <- c("Using theorem",
                            "Computing directly")

compareTable[, 1] <- c( mean(ai * Xi), 
                        mean(Y) )
compareTable[, 2] <- c( var(ai * Xi), 
                        var(Y) )
compareTable
```

The sampling distribution of the sum and mean of a random sample from a normal population follows directly from Theorem\ \@ref(thm:LinearComb).


:::{.corollary #SumMean name="Sum and mean of a random sample"}
Let $X_1, X_2, \dots, X_n$ be a random sample of size $n$ from $N(\mu,\sigma^2)$. 
Define the sum\ $S$ and mean\ $\overline{X}$ respectively as
\begin{align*}
   S
   &= X_1 + X_2 + \dots + X_n\\
   \overline{X}
   &= (X_1+X_2 + \dots + X_n)/n
\end{align*}
Then $S\sim N(n\mu, n\sigma^2)$ and $\overline{X}\sim N(\mu, \sigma^2/n)$.

::::{.proof}
Exercise.
::::
:::


Corollary\ \@ref(cor:SumMean) relies only on the population from which the sample is drawn having a normal distribution, and on the properties of expectation; it does not depend on the sample size\ $n$.
This means that for any size sample drawn from a normal distribution, the sample means have a normal distribution.
This is the basis for inference about the population mean of a normal distribution with known variance.


:::{.example #CLTEnvelopes name="Sums of rvs"}
Suppose envelopes are collected into packs of\ $25$ by weighing them.
The weight of an envelope is distributed normally with mean\ $3\gs$ and standard deviation\ $0.6\gs$. 
Any weighed pack of envelopes is declared as containing\ $25$ if it weighs between $70\gs$ and $80\gs$.
What is the probability that a pile of\ $25$ will *not* be counted as such?

Let random variable\ $X_i$ be the weight of the $i$th\ envelope; then $X_i \sim N(3, 0.36)$. 
Let $Y = X_1 + X_2 + \dots + X_{25}$; then $Y \sim N(75, 9)$. 
We require $1 - \Pr(70 < Y < 80)$; see Fig.\ \@ref(fig:EnvelopeDist).
Since the normal distribution is symmetric, proceed:
\begin{align*}
   1 - \Pr(70 < Y < 80)
   &= 2\times \Pr\left( Z < \frac{70 - 75}{\sqrt{9}} \right)\\
   &= 2\times \Phi(-1.667) \\
   &= 0.10.
\end{align*}
In **R**:

```{r}
2 * ( pnorm( -1.667 ) )
```
:::


```{r EnvelopeDist, echo=FALSE, out.width='90%', fig.width=8, fig.height=4, fig.align="center", fig.cap="The envelope example. Left panel: The distribution of the weight of individual envelopes; right panel: The distribution of the weight of packs of $25$\\ envelopes."}
par( mfrow = c(1, 2))

x <- seq(3 - 4 * 0.6,
         3 + 4 * 0.6,
         length = 200)
y <- dnorm(x,
           mean = 3,
           sd = 0.6)
plot( y ~ x,
      type = "l",
      lwd = 2,
      las = 1,
      axes = FALSE,
      main = "Weight of\nindividual envelopes",
      xlab = "Envelope weight (in g)",
      ylab = "Density")
axis(side = 1)
box()


x <- seq(75 - 4 * 3,
         75 + 4 * 3,
         length = 200)
y <- dnorm(x,
           mean = 75,
           sd = 3)
plot( y ~ x,
      type = "l",
      lwd = 2,
      las = 1,
      axes = FALSE,
      main = "Weight of packs of\n25 envelopes",
      xlab = "Pack weight (in g)",
      ylab = "Density")
axis(side = 1)
box()
shadeNormal(x, y,
            col = "grey",
            lo = 70,
            hi = 80)
```



:::{.example #CLTweights name="CLT"}
The IQs for a large population of $10$\ year-old boys is assumed to be normally distributed, and were determined and found to have a mean of\ $110$ and a variance of\ $144$. 
How large a sample would have to be taken in order to have a probability of\ $0.9$ that the *mean* IQ of the sample would not differ from the expected value\ $110$ by more than\ $5$?

Let\ $X_i$ be the IQ of the $i$th\ boy; then $X_i \sim N(110, 144)$.
Consider a sample of size\ $n$ and let $\overline{X} = \sum^n_{i = 1}X_i/n$; then $\overline{X}\sim N(110, 144/n)$.
Then
$$
   \Pr(|\overline{X} - 110|\leq 5) = 0.90.
$$
That is
$$
   \Pr\left(\frac{|\overline{X} - 110|\sqrt{n}}{12} \leq \frac{5\sqrt{n}}{12}\right) = 0.90
$$
hence
$$
   \Pr(Z \leq 5\sqrt{n} /12) = 0.90. 
$$
Using **R**:
```{r ech0=TRUE}
z <- qnorm(0.95)
z

n <- (z * 12 / 5) ^ 2
n
```
The *smallest* size sample, then, would be a sample of $n = 16$.
:::


:::{.example #CLTPlunger name="Two random variable"}
A certain product involves a plunger fitting into a cylindrical tube.
The diameter of the plunger can be considered a normal random variable with mean $2.1\cms$ and standard deviation $0.1\cms$.
The inside diameter of the cylindrical tube is a normal random variable with mean $2.3\cms$ and standard deviation $0.05\cms$. 
For a plunger and tube chosen randomly from a day's production run, find the probability that the plunger will not fit into the cylinder.

Let\ $X$ and\ $Y$ be the diameter of the plunger and cylinder respectively.
Then $X\sim N\big(2.1, (0.1)^2\big)$ and $Y\sim N\big( 2.3, (0.05)^2\big)$, and we seek $\Pr(Y < X)$.
The distribution of $Y - X$ is $N(2.3 - 2.1, 0.0025 + 0.01)$ so that
\begin{align*}
   \Pr(Y - X < 0) 
   &= \Pr\left(Z <\frac{0 - 0.2}{\sqrt{0.0125}}\right) \quad\text{where $Z\sim N(0,1)$}\\
   &= \Pr(Z < -1.78)\\
          &= 0.0375.
\end{align*}
In **R**:
```{r}
pnorm( -1.78 )
```
:::



## Distributions related to the normal distribution {#DistributionsRelatedToNormal}

### The $\chi^2$-distribution {#ChiSquareDistribution}

\index{Chi-squared distribution}
Having studied the sampling distribution of the *mean*, we now study the sampling distribution of the *variance* in a normal population.
We begin with the definition and properties of the chi-squared distribution, which were first encountered in Examples\ \@ref(exm:TransformationNon11) and\ \@ref(exm:TransformB).


:::{.definition #ChiSquaredDistribution name="Chi-squared distribution"}
A continuous random variable\ $X$ with probability density function\index{Chi-squared distribution}
\begin{equation}
   f_X(x)
   = \frac{x^{(\nu/2) - 1}e^{-x/2}}{2^{\nu/2}\Gamma(\nu/2)}
   \quad\text{for $x > 0$}
\end{equation}
is said to have a *chi-squared distribution* with parameter $\nu > 0$.
The parameter\ $\nu$ is called the *degrees of freedom*.
We write $X \sim \chi^2(\nu)$.
:::


Some plots of $\chi^2$-distributions are shown in Fig.\ \@ref(fig:ChisqPlots).


::: {.linkBox .link data-latex="{iconmonstr-link-1-240.png}"}
The chi-squared distribution is a special case of the gamma distribution, with $\alpha = \nu/2$ and $\beta = 2$ (Exercise\ \@ref(exr:ChiSquaredGamma)).
This means that properties of the chi-squared distribution can be obtained directly from those for the gamma distribution (Sect.\ \@ref(GammaDistribution)).
:::


```{r ChisqPlots, echo=FALSE, out.width='100%', fig.height=3, fig.width=6.5, fig.cap="Some $\\chi^2$-distributions."}
par( mfrow = c(1, 3))

xx <- seq(0.01, 4, length = 100)
plot( x = xx,
      y = dchisq(xx, df = 1),
      lwd = 2,
      type = "l",
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression( bold(chi^2*"-"*distribution: df==1)))

xx <- seq(0.001, 15, length = 100)
plot( x = xx,
      y = dchisq(xx, df = 5),
      lwd = 2,
      type = "l",
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression( bold(chi^2*"-"*distribution: df==5)))

xx <- seq(0.001, 50, length = 100)
plot( x = xx,
      y = dchisq(xx, df = 20),
      lwd = 2,
      type = "l",
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression( bold(chi^2*"-"*distribution: df==20)))
```


:::{.theorem #ChisqProperties name="Properties of the chi-squared distribution"}
If $X\sim\chi^2(\nu)$ then

* $\operatorname{E}(X) = \nu$.
* $\operatorname{var}(X) = 2\nu$.
* $M_X(t) = (1 - 2t)^{-\nu/2}$ for $t < 1/2$.

::::{.proof}
See Theorem\ \@ref(thm:GammaProperties).
::::
:::


::: {.softwareBox .software data-latex="{iconmonstr-laptop-4-240.png}"}
The four **R** functions for working with the $\chi^2$-distribution have the form `[dpqr]chisq(df)`, where `df`${} = \nu$ is the degrees of freedom (see Appendix\ \@ref(UseRDistributions)):

* `dchisq(x, df)` computes the PDF at $X = {}$`x`;
* `pchisq(q, df)` computes the CDF at $X = {}$`q`;
* `qchisq(p, df)` computes the quantile for cumulative probability `p`; and
* `rchisq(n, df)` generates `n` random observations.
:::


Two results established earlier are needed for what follows.
The first (proved in Example\ \@ref(exm:TransformationNon11)) is that if $Z\sim N(0,1)$ then $Z^2\sim\chi^2(1)$; this is
Theorem\ \@ref(thm:SumSquaredNormals).
The second extends this to a sum of squares.


:::{.theorem #SumSquaredNormals name="Sum of squared standard normals"}
If $Z_1, Z_2,\dots, Z_n$ are iid $N(0, 1)$, then $\displaystyle\sum_{i=1}^n Z_i^2 \sim \chi^2(n)$.

::::{.proof}
Since $Z_i^2 \sim \chi^2(1)$, from Theorem\ \@ref(thm:ChisqProperties) its MGF is $M_{Z_i^2}(t) = (1 - 2t)^{-1/2}$.
By independence and Theorem\ \@ref(thm:MGFIndependent),
\begin{align*}
   M_{\sum Z_i^2}(t)
   &= \prod_{i = 1}^n (1 - 2t)^{-1/2}
    = (1 - 2t)^{-n/2},
\end{align*}
which is the MGF of $\chi^2(n)$.
::::
:::


We can now state the main result connecting a random sample from a normal population to the chi-squared distribution.


:::{.theorem #ChiSquare name="Chi-squared distribution of standardised sum of squares"}
Let $X_1, X_2, \dots, X_n$ be a random sample from $N(\mu, \sigma^2)$.
Then
$$
  \sum_{i = 1}^n \left(\frac{X_i - \mu}{\sigma}\right)^2
  \sim \chi^2_n.
$$

::::{.proof}
Setting $Z_i = (X_i - \mu)/\sigma$ gives iid $N(0,1)$ variables, and the result follows immediately from Theorem\ \@ref(thm:SumSquaredNormals).
::::
:::


The sample variance is given by
\begin{equation}
   S^2 = \frac{1}{n - 1}\sum_{i = 1}^n (X_i - \overline{X})^2.
   (\#eq:SampleVarDist)
\end{equation}
Since\ $S^2$ is a statistic, it has a sampling distribution.
Theorem\ \@ref(thm:ChiSquare) almost applies, but $\overline{X}$ appears in place of the known constant\ $\mu$---a difference that
costs one degree of freedom. ??????????


:::{.theorem #SampleVarDist name="Sampling distribution of the sample variance"}
Let $X_1, X_2, \dots, X_n$ be a random sample from $N(\mu, \sigma^2)$. 
Then
$$
   \frac{(n - 1)S^2}{\sigma^2} \sim \chi^2_{n - 1}.
$$

::::{.proof}
Only a *partial proof* is provided.
From Eq.\ \@ref(eq:SampleVarDist),
$$
   \frac{(n - 1)S^2}{\sigma^2}
   = \sum_{i = 1}^n \left(\frac{X_i - \overline{X}}{\sigma}\right)^2.
$$
This resembles Theorem\ \@ref(thm:ChiSquare), except that the known constant\ $\mu$ is replaced by the random variable\ $\overline{X}$.
Writing $X_i - \overline{X} = (X_i - \mu) - (\overline{X} - \mu)$ and expanding gives
\begin{align*}
  \sum_{i = 1}^n \left(\frac{X_i - \overline{X}}{\sigma}\right)^2
    &= \sum_{i = 1}^n \left(\frac{X_i - \mu}{\sigma}\right)^2
       - n\frac{(\overline{X} - \mu)^2}{\sigma^2}.
\end{align*}
(Check!)
Define $W = \sum_{i=1}^n\bigl((X_i-\mu)/\sigma\bigr)^2$, $U = n(\overline{X}-\mu)^2/\sigma^2$, and $V = (n-1)S^2/\sigma^2$, so that $W = U + V$.
Then:

* $W \sim \chi^2_n$ by Theorem\ \@ref(thm:ChiSquare), with MGF   $M_W(t) = (1 - 2t)^{-n/2}$.
* $U \sim \chi^2_1$ because $\overline{X}\sim N(\mu, \sigma^2/n)$, so $(\overline{X}-\mu)/(\sigma/\sqrt{n})$ is standard normal and   $U$ is its square; thus $M_U(t) = (1 - 2t)^{-1/2}$.

Provided\ $U$ and\ $V$ are independent (not proved here), the MGF of\ $V$ is
$$
   M_V(t) = \frac{M_W(t)}{M_U(t)} = (1 - 2t)^{-(n-1)/2},
$$
which is the MGF of $\chi^2_{n-1}$.
::::
:::


:::{.example #ChisqProb name="Chi-squared quantile"}
The variable\ $X$ has a chi-squared distribution with $12$\ df.
Determine the value of\ $X$ below which lies 90% of the distribution.

We seek\ $c$ such that $\Pr(X \le c) = 0.90$ where $X\sim\chi^2(12)$. 
In **R**:

```{r}
qchisq(0.9, df = 12)
```

About 90% of the distribution lies below
`r round(qchisq(0.9, df = 12), 3)`.
:::


:::{.example #SampleVar name="Distribution of sample variance"}
A random sample of size $n = 10$ is selected from the $N(20, 5)$ distribution.
What is the probability that the sample variance exceeds\ $10$?

By Theorem\ \@ref(thm:SampleVarDist),
$$
   \frac{9 S^2}{5} \sim \chi^2_9,
$$
so
$$
   \Pr(S^2 > 10)
   = \Pr\!\left(\frac{9S^2}{5} > 18\right)
   = \Pr(\chi^2_9 > 18).
$$
In **R**:

```{r, echo=TRUE}
pchisq(18, df = 9, lower.tail = FALSE)
```

Hence $\Pr(S^2 > 10) \approx 0.035$.
:::


### The $t$-distribution {#TDistribution}

\index{t-distribution@$t$-distribution}
The basis of the independence assumed in the proof of Theorem\ \@ref(thm:SampleVarDist) is the (perhaps surprising) result that the mean and variance of a random sample from a normal population are independent random variables.


:::{.theorem #MeanVarInd name="Sample mean and variance: Independent"}
Let $X_1, X_2,\dots, X_n$ be a random sample of size\ $n$ from $N(\mu, \sigma^2)$.
Then the sample mean
$$
   \overline{X}
   = 
   \frac1n\sum_{i=1}^n X_i
$$
and sample variance
$$
   S^2
   = \frac{1}{n - 1}\sum_{i=1}^n (X_i - \overline{X})^2
$$
are independent.

::::{.proof}
This proof is not given.
::::
:::


Another distribution that derives from sampling a normal population is the $t$-distribution.


:::{.definition #TDistribution name="T-distribution"}
Suppose $Z \sim N(0, 1)$ (i.e., a standard normal distribution) and $V \sim \chi^2_n$ (i.e., a chi-squared distribution with $n$\ degrees of freedom).
Then
\begin{equation}
  T = \frac{Z}{\sqrt{V/n}}
  (\#eq:Tdist)
\end{equation}
is defined as a $T$-distribution with $n$\ degrees of freedom.
:::

:::{.definition #TDistributionPDF name="T-distribution"}
A continuous random variable\ $X$ with probability density function
\begin{equation}
   f_X(x)
   = \frac{\Gamma\left((\nu + 1)/2\right)}{\sqrt{\pi \nu}\,\Gamma(\nu/2)}
     \left(1 + \frac{x^2}{\nu}\right)^{-(\nu + 1)/2}
     \quad\text{for all $x$}
\end{equation}
is said to have a *$t$-distribution* with parameter $\nu > 0$.
The parameter\ $\nu$ is called the *degrees of freedom*. 
We write $X \sim t_\nu$.

::::{.proof}
This proof is not given.
::::
:::


The probability denisty function of the $T$-distribution is very similar to the standard normal distribution (Fig.\ \@ref(fig:TPlots)): bell-shaped and symmetric about zero. 

However, the variance is greater than one (Theorem\ \@ref(thm:TDistProperties)) and so is larger than the variance of the standard normal distribution, and is dependent on\ $\nu$.



::: {.softwareBox .software data-latex="{iconmonstr-laptop-4-240.png}"}
The four **R** functions for working with the $t$\ distribution have the form `[dpqr]t(df)`:

* `dt(x, df)` computes the PDF at $X = {}$`x`;
* `pt(q, df)` computes the CDF at $X = {}$`q`;
* `qt(p, df)` computes the quantile for cumulative probability `p`; and
* `rt(n, df)` generates `n` random numbers,

where `df` is the degrees of freedom (see Appendix\ \@ref(UseRDistributions)). 
:::


```{r TPlots, echo=FALSE, out.width='100%', fig.height=2.75, fig.width=6.5, fig.cap="Some $t$-distributions (with normal distributions in dashed lines), with mean\\ $0$ and variance\\ $1$."}
par( mfrow = c(1, 3))

xx <- seq(-4, 4,
          length = 100)

plot( x = xx,
      y = dt(xx, df = 2),
      ylim = c(0, 0.4),
      lwd = 2,
      type = "l",
      axes = FALSE,
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression( bold(The~italic(t)*"-"*distribution~with~2~df)))
axis(side = 1)
lines( x = xx,
       y = dnorm(x = xx),
       lwd = 2,
       lty = 2,
       col = plotColour1)

plot( x = xx,
      y = dt(xx, df = 7),
      lwd = 2,
      type = "l",
      axes = FALSE,
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression( bold(The~italic(t)*"-"*distribution~with~7~df)))
axis(side = 1)
lines( x = xx,
       y = dnorm(x = xx),
       lwd = 2,
       lty = 2,
       col = plotColour1)

plot( x = xx,
      y = dt(xx, df = 30),
      lwd = 2,
      type = "l",
      axes = FALSE,
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression( bold(The~italic(t)*"-"*distribution~with~30~df)))
axis(side = 1)
lines( x = xx,
       y = dnorm(x = xx),
       lwd = 2,
       lty = 2,
       col = plotColour1)
```


:::{.theorem #TDistProperties name="Properties of the $T$-distribution"}
If $X\sim t_\nu$ then

* For $\nu > 1$, $\operatorname{E}[X] = 0$.
* For $\nu > 2$, $\operatorname{var}[X] = \displaystyle{\frac{\nu}{\nu - 2}}$.
* The MGF does not exist because only the first $\nu - 1$ moments exist. 

::::{.proof}
This proof is not given.
::::
:::


Although we won't prove it, as $\nu\to\infty$ the $t$\ distribution converges to the standard normal (which can be seen in Fig.\ \@ref(fig:TPlots)).
In addition, from Theorem\ \@ref(thm:TDistProperties), we see that as $\nu \to \infty$, $\operatorname{var}[X] \to 1$ as for the normal distribution.

The usefulness of the $t$-distribution derives from the following important result.


:::{.theorem #Tscores name="$T$-scores"}
Let $X_1, X_2, \dots, X_n$ be a random sample of size\ $n$ from $N(\mu, \sigma^2)$.
Then the random variable
$$
   T = \frac{\overline{X} - \mu}{S/\sqrt{n}}
$$
follows a $t_{n - 1}$-distribution where\ $\overline{X}$ is the sample mean and\ $S^2$ is the sample variance as defined in\ Eq.\ \@ref(eq:SampleVarDist).

::::{.proof}
**We give a partial proof only.**
The statistic\ $T$ can be written as
\begin{align*}
  T
    &= \frac{\overline{X} - \mu}{S/\sqrt{n}}\\
    &= \left(\frac{\overline{X} - \mu}{\sigma/\sqrt{n}}\right)   \frac{1}{\sqrt{\frac{(n - 1)S^2}{\sigma^2}/(n - 1)}}\\
    &= \frac{Z}{Y/(n - 1)},
\end{align*}
where $Z$ is a $N(0, 1)$ variable and $Y$ is a chi-squared variable with $(n - 1)$ df.
Using Def.\ \@ref(def:TDistribution) then, $T$ has a $T$-distribution.
::::
:::



Notice that\ $T$ represents a *standardised* version of the sample mean.
Because of this, the $T$\ statistic is important in statistical inference.
You may have used a $t$-test in hypothesis testing, for example.


:::{.example #CalcaltingT name="Calculating $T$"}
A random sample $21, 18, 16, 24, 16$ is drawn from a normal population with mean of\ $20$.

1. What is the value of\ $T$ for this sample?
2. In random samples from this population, what is the probability that\ $T$ is less than the value found above?

From the sample: $\overline{x} = 19.0$ and $s^2 = 12.0$. 
Therefore $t = \frac{19.0 - 20}{\sqrt{12.0/5}} = -0.645$.
(Notice lower-case symbols are used for specific values of statistics, and upper-case symbols for the random variables.)

Interest here is in $\Pr(T < -0.645)$ where $T\sim t_4$. 
From **R**, the answer is\ $0.277$:

```{r echo=TRUE}
pt(-0.645, df = 4)
```
:::


### The $F$-distribution {#FDistribution}

\index{F-distribution@$F$-distribution}
The $F$-distribution arises naturally when comparing the variability of two independent normal populations, and is also used in *analysis of variance*, a common technique for testing the equality of several means.


:::{.definition #FDistribution name="$F$-distribution"}
A continuous random variable\ $X$ with probability density function
\begin{equation}
  f_X(x)
  = \frac{\Gamma\big( (\nu_1 + \nu_2)/2 \big)
    \nu_1^{\nu_1/2} \nu_2^{\nu_2/2} x^{(\nu_1/2) - 1}}
   {\Gamma(\nu_1/2)\Gamma(\nu_2/2)(\nu_1 x + \nu_2)^{(\nu_1 + \nu_2)/2}}
  \quad\text{for $x > 0$}
\end{equation}
is said to have an *$F$-distribution* with parameters $\nu_1 > 0$ and $\nu_2 > 0$, known respectively as the numerator and denominator degrees of freedom.
We write $X \sim F_{\nu_1, \nu_2}$.
:::


Some $F$-distributions are shown in Fig.\ \@ref(fig:FPlots).


:::{.theorem #FProperties name="Properties of the $F$-distribution"}
If $X\sim F_{\nu_1, \nu_2}$ then

* For $\nu_2 > 2$,
  $\operatorname{E}[X] = \displaystyle{\frac{\nu_2}{\nu_2-2}}$.
* For $\nu_2 > 4$,
  $\operatorname{var}[X] = \displaystyle{\frac{2\nu_2^2(\nu_1 + \nu_2-2)}
  {\nu_1(\nu_2 - 2)^2(\nu_2 - 4)}}$.
* The MGF does not exist.

::::{.proof}
Not covered.
::::
:::


::: {.softwareBox .software data-latex="{iconmonstr-laptop-4-240.png}"}
The four **R** functions for working with the $F$-distribution have the form `[dpqr]f(df1, df2)`, where `df1`${} = \nu_1$ and
`df2`${} = \nu_2$ are the degrees of freedom (see Appendix\ \@ref(UseRDistributions)):

* `df(x, df1, df2)` computes the PDF at $X = {}$`x`;
* `pf(q, df1, df2)` computes the CDF at $X = {}$`q`;
* `qf(p, df1, df2)` computes the quantile for cumulative probability `p`;
  and
* `rf(n, df1, df2)` generates `n` random observations.
:::


```{r FPlots, echo=FALSE, out.width='100%', fig.height=2.5, fig.width=6.5, fig.cap="Some $F$-distributions."}
par(mfrow = c(1, 3))

xx <- seq(0.01, 4, length = 100)

plot( x = xx,
      y = df(xx, df1 = 1, df2 = 4),
      lwd = 2,
      type = "l",
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression(bold(atop(The~italic(F)*"-"*distribution,
                                  with~df=="(1, 4)"))))

plot( x = xx,
      y = df(xx, df1 = 6, df2 = 1),
      lwd = 2,
      type = "l",
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression(bold(atop(The~italic(F)*"-"*distribution,
                                  with~df=="(6, 1)"))))

plot( x = xx,
      y = df(xx, df1 = 30, df2 = 5),
      lwd = 2,
      type = "l",
      xlab = expression(italic(x)),
      ylab = "Density",
      las = 1,
      main = expression(bold(atop(The~italic(F)*"-"*distribution,
                                  with~df=="(30, 5)"))))
```


The main result connecting the $F$-distribution to samples from normal populations is the following.


:::{.theorem #FGenerate name="Sampling distribution of the ratio of two sample variances"}
Let $X_1, X_2, \dots, X_{n_1}$ be a random sample of size\ $n_1$ from $N(\mu_1, \sigma_1^2)$ and $Y_1, Y_2,\dots, Y_{n_2}$ be an independent random sample of size\ $n_2$ from $N(\mu_2, \sigma_2^2)$.
Then the random variable
$$
   F = \frac{S_1^2/\sigma_1^2}{S_2^2/\sigma_2^2}
$$
follows an $F_{n_1 - 1,\, n_2 - 1}$-distribution.

::::{.proof}
**A partial proof only is given.**
By Theorem\ \@ref(thm:SampleVarDist),
$$
   U_1 = \frac{(n_1 - 1)S_1^2}{\sigma_1^2} \sim \chi^2_{n_1 - 1}
   \qquad\text{and}\qquad
   U_2 = \frac{(n_2 - 1)S_2^2}{\sigma_2^2} \sim \chi^2_{n_2 - 1},
$$
and $U_1$, $U_2$ are independent because the two samples are independent.
Setting $\nu_1 = n_1 - 1$ and $\nu_2 = n_2 - 1$, the statistic can be written as
$$
   F = \frac{S_1^2/\sigma_1^2}{S_2^2/\sigma_2^2}
     = \frac{U_1/\nu_1}{U_2/\nu_2}.
$$
That this ratio has an $F_{\nu_1, \nu_2}$-distribution follows directly from Def.\ \@ref(def:FDistribution); the derivation of the PDF is not given here.
::::
:::


:::{.corollary}
If $T \sim t_n$ then $T^2 \sim F_{1,\, n}$.

::::{.proof}
From Def.\ \@ref(def:TDistribution), $T = Z / \sqrt{V/n}$ where $Z \sim N(0,1)$ and $V \sim \chi^2_n$ are independent.
Then
$$
  T^2 = \frac{Z^2/1}{V/n}.
$$
Since $Z^2 \sim \chi^2_1$ (Theorem\ \@ref(thm:SumSquaredNormals)), this is a ratio of the form $\chi^2_1/1$ over $\chi^2_n/n$, which by Def.\ \@ref(def:FDistribution) has an $F_{1,n}$-distribution.
::::
:::


:::{.example #FProbs name="$F$-distribution probabilities"}
Suppose $X\sim F_{2, 10}$.
Find:

* $\Pr(X < 1)$
* $x$ such that $\Pr(X > x) = 0.01$

Using **R**:

```{r echo=TRUE}
pf(1,    df1 = 2, df2 = 10)
qf(0.99, df1 = 2, df2 = 10)
```
:::


## The Central Limit Theorem {#CentralLimitTheorem}

Sampling distributions for various statistics of interest are well-developed when sampling is from a *normal* distribution (Sect.\ \@ref(SamplingDistributionsRelatedToNormal)).
Although these results are important, what if we do not know the distribution of the population from which the random sample is drawn (as is usually the case)?

In Sect.\ \@ref(SamplingDistributionMean) general results are given describing the *mean* and *variance* of the sample mean which hold for any population distribution, but do not say anything about the *distribution* of the sample means.
The *Central Limit Theorem* (or *CLT*) describes the distribution of the sample means.


:::{.theorem #CLT name="Central Limit Theorem (CLT)"}
Let $X_1, X_2, \dots, X_n$ be a random sample from a distribution with mean\ $\mu$ and variance\ $\sigma^2$.
Then the random variable
$$
   Z_n 
   = \frac{\overline{X} - \mu}{\sigma / \sqrt{n}}
$$
converges in distribution to a standard normal variable as $n\to\infty$.

::::{.proof}
Without loss of generality, centre the $X_i$ by letting $Y_i = (X_i - \mu)/\sigma$,
so that $\operatorname{E}[Y_i] = 0$, $\operatorname{var}[Y_i] = 1$, and
$$
  Z_n = \frac{1}{\sqrt{n}} \sum_{i=1}^n Y_i.
$$
Let $\varphi(t) = \operatorname{E}[\exp(itY)]$ denote the common characteristic function (CF) of the\ $Y_i$.
Since the\ $Y_i$ are independent, the CF of\ $Z_n$ is
$$
  \varphi_{Z_n}(t)
  = \operatorname{E}\left[e^{itZ_n}\right]
  = \operatorname{E}\left[\exp\!\left\{\frac{it}{\sqrt{n}}\sum_{i=1}^n Y_i\right\}\right]
  = \prod_{i=1}^n \operatorname{E}\left[\exp(it Y_i/\sqrt{n})\right]
  = \left[\varphi\!\left(\frac{t}{\sqrt{n}}\right)\right]^n
$$

Since $\operatorname{E}[Y^2] = 1 < \infty$, the CF $\varphi(t)$ is twice differentiable at $t = 0$, with
$$
  \varphi'(t)\big|_{t=0} = i\operatorname{E}[Y] = 0
  \qquad\text{and}\qquad
  \varphi''(t)\big|_{t=0} = i^2 \operatorname{E}[Y^2] = -1.
$$
The Taylor expansion of $\varphi$ about $t = 0$ is therefore
$$
  \varphi(t)
  = 1 + \varphi'(0)\,t + \frac{1}{2}\varphi''(0)\,t^2 + o(t^2)
  = 1 - \frac{1}{2}t^2 + o(t^2)
  \quad \text{as $t \to 0$},
$$
where $o(t^2)$ means $o(t^2)/t^2 \to 0$ as $t\to 0$.

Substituting $t/\sqrt{n}$ gives
$$
  \varphi\!\left(\frac{t}{\sqrt{n}}\right) 
  = 1 - \frac{t^2}{2n} + o\!\left(\frac{t^2}{n}\right).
$$

Taking logs (which is valid for large enough\ $n$ since $\varphi(t/\sqrt{n}) \to 1$) and using
$\log(1 + u) = u + o(u)$ as $u \to 0$:
$$
  \log \varphi\!\left(\frac{t}{\sqrt{n}}\right) 
  = -\frac{t^2}{2n} + o\!\left(\frac{t^2}{n}\right).
$$
Therefore
$$
  \log \varphi_{Z_n}(t)
  = n \log \varphi\!\left(\frac{t}{\sqrt{n}}\right)
  = n\left[-\frac{t^2}{2n} + o\!\left(\frac{t^2}{n}\right)\right]
  = -\frac{t^2}{2} + n\cdot o\left(\frac{t^2}{n}\right)
  \to\, -\frac{t^2}{2}
$$
as $n \to \infty$, for each fixed $t \in \mathbb{R}$. 
Taking exponents,
$$
  \varphi_{Z_n}(t) \to \exp(-t^2/2) \quad \text{for all $t \in \mathbb{R}$}.
$$

The function $\exp(-t^2/2)$ is the CF of the $N(0, 1)$ distribution.
By *Lévy's continuity theorem*, if $\varphi_{Z_n}(t) \to \varphi(t)$ pointwise for all\ $t$, and the limit $\varphi(t)$ is continuous at $t = 0$, then $Z_n \overset{d}{\to} Z$, where\ $Z$ has characteristic function $\varphi(t)$.
Since $\exp(-t^2/2)$ is continuous everywhere and is the CF of $N(0, 1)$, conclude that
$$
  Z_n \overset{d}{\to} N(0, 1).
$$

::::
:::


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
The *Central Limit Theorem* is one of the most important theorems in statistics.
:::


This is a partial proof since the concept of convergence in distribution has not been defined. However, the arguments used in the proof are powerful and worth understanding.

'Convergence in distribution' means that the distribution of $Z_n$ converges to a standard normal as $n\to \infty$. 
So when\ $n$ is 'large', $Z_n$ is expected to approximate a standard normal distribution. Transforming $Z_n$ back to the sample mean, $\overline{X}$ can be expected to approximate a $N(\mu, \sigma^2/n)$ distribution.

In practice, the distribution of\ $\overline{X}$ is sufficiently close to that of a normal distribution when\ $n$ is larger than about\ $25$ in most situations.
However, if the population distribution is severely skewed, larger samples sizes may be necessary for the approximate to be adequate in practice.


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
The *Central Limit Theorem* states that, for *any* random variable\ $X$, the distribution of the *sample means* will have an:\index{Central Limit Theorem}

* approximate normal distribution,
* with mean\ $\mu$ (the mean of\ $X$), and
* with a variance of $\sigma^2/n$ (where\ $\sigma^2$ is the variance of\ $X$).

The approximation improves as the sample size\ $n$ gets larger.

This applies for *any* random variable\ $X$, *whatever its distribution*.
For many distributions that are not highly skewed, the approximation is reasonable for sample sizes larger than about\ $20$ to\ $30$.
If the data comes from a normal distribution, the distribution of the sample mean *always* has a normal distribution for any\ $n$.
:::


```{r, child = if (knitr::is_html_output())  './children/CLT/CLT_html.Rmd' }
```

```{r, child = if (knitr::is_latex_output()) './children/CLT/CLT_latex.Rmd'}
```


:::{.example #CLTUniform name="CLT and a uniform distribution"}
In the **R** code below, data come from a uniform distribution (Fig.\ \@ref(fig:UniformCLT), left panel).
However, the mean of samples of size $n = 10$ (centre panel) and $n = 20$ (right panel) are approximately distributed as a normal distribution.
The variance of the sample means for the larger sample size is smaller than that for $n = 10$, and the distribution looks more like a normal distribution for the larger sample size.
:::


```{r UniformCLT, fig.cap="The Central Limit Theorem. Left: the distribution of the individual observations. Centre: the sampling distribution of the sample mean for samples of size $n = 10$. Right: the sampling distribution of the sample mean for samples of size $n = 20$.", fig.align="center", fig.width=7, out.width='100%', fig.height=3}
# The distribution of individuals has a uniform distribution:
set.seed(4516) # For reproducibility
sampleSize1 <- 10
sampleSize2 <- 20
sampleMeans1 <- NULL
sampleMeans2 <- NULL

numberOfSimulations <- 1000
for (i in 1:numberOfSimulations){
  # Take a sample of size 10
  Xs1 <- runif(sampleSize1, 
               min = 0,
               max = 5)
  # Take a sample of size 10
  Xs2 <- runif(sampleSize2, 
               min = 0,
               max = 5)

  # Find the sample mean
  sampleMeans1 <- c(sampleMeans1, mean(Xs1) )
  sampleMeans2 <- c(sampleMeans2, mean(Xs2) )
}

par(mfrow = c(1, 3))

plot( x = c(-1, 0, 0, 5, 5, 6), 
      y = c(0, 0, 0.2, 0.2, 0, 0),
      lwd = 2,
      las = 1,
      type = "l",
      xlab = expression(italic(x)),
      ylab = "Prob. fn",
      main = "Distribution of the\nindividuals observations")

hist( sampleMeans1,
      las = 1,
      xlab = expression(Sample~means~bar(italic(x)) ),
      main = expression( bold( atop(Distribution~of~sample~means,
                               "for"~italic(n)==10) ) ) )
hist( sampleMeans2,
      las = 1,
      xlab = expression(Sample~means~bar(italic(x)) ),
      main = expression( bold( atop(Distribution~of~sample~means,
                               "for"~italic(n)==20) ) ) )
```

:::{.example #SoftDrinkVending name="CLT"}
A soft-drink vending machine is set so that the amount of drink dispensed is a random variable with a mean of $200\mLs$ and a standard deviation of $15\mLs$.
What is the probability that the *average* amount dispensed in a random sample of size\ $36$ is at least $204\mLs$?

If\ $X$ is the amount of drink dispensed in mL, then the distribution of\ $X$ is not given.
However, the mean and standard deviation of\ $X$ are given.
The distribution of the sample mean\ $\overline{X}$ can be approximated by the normal distribution with mean of $\mu = 200$ and standard error of $\sigma_{\overline{X}} = \sigma/\sqrt{n} = 15/\sqrt{36} = 15/6$, according to the CLT.
That is, $\overline{X} \sim N(200, (15/6)^2)$.
Now
$$
   \Pr(\overline{X} \ge 204) \approx \Pr_N\left(Z \ge \frac{204 - 200}{15/6}\right)
   = \Pr(Z \ge 1.6).
$$
In **R**:

```{r}
1 - pnorm(1.6)
```

Hence, $\Pr(\overline{X}\ge 204) \approx 0.0548$, or about\ $5.5$%.
(Here, $\Pr_N(A)$ is used to denote the probability of event\ $A$ involving a random variable assumed to be normal in distribution.)
:::


:::{.example #FairDiceSum name="Throwing dice"}
Consider the experiment of throwing a fair die\ $n$ times where we observe the sum of the faces showing. 
For $n = 12$, find the probability that the sum of the faces is at least\ $52$.

Let the random variable\ $X_i$ be the number showing on the $i$th\ throw.
Then, define $Y = X_1 + \dots + X_{12}$. 
We seek $\Pr(Y\geq 52)$.

In order to use Theorem\ \@ref(thm:CLT), note that the event '$Y\geq 52$' is equivalent to '$\overline{X}\geq 52/12$', where $\overline{X} = Y/12$ is the mean number showing from the $12$\ tosses.

Since the distribution of each\ $X_i$ is rectangular with $\Pr(X_i = x) = 1/6$ (for $x = 1, 2, \dots, 6$), then $\operatorname{E}[X_i] = 7/2$ and $\operatorname{var}[X_i] = 35/12$.

It follows that $\operatorname{E}[\overline{X}] = 7/2$ and $\operatorname{var}[\overline{X}] = 35/(12^2)$. 
Then from Theorem\ \@ref(thm:CLT),
\begin{align*}
  \Pr(Y\geq 52) 
    & \simeq \Pr(\overline{X}\geq 52/12)\\
    &= \Pr_N \left(Z\geq \frac{52/12 - 7/2}{\sqrt{35/144}}\right)\\
    &= 1 - \Phi(1.690)\\
    &= 0.0455
\end{align*}
The probability is *approximately*\ $4.5$%.
:::


Example\ \@ref(exm:FairDiceSum) can also be solved using a generalisation of the Central Limit Theorem.
This generalisation indicates why the normal distribution plays such an important role in statistical theory. 
It states that a large class of random variables converge in distribution to the standardized normal.


:::{.theorem #ConvergeToN name="Convergence to standard normal distribution"}
Let $X_1, X_2, \dots, X_n$ be a sequence of independent random variables with $\operatorname{E}[X_i] = \mu_i$, $\operatorname{var}[X_i] = \sigma^2_i$.
Define $Y = a_1X_1 + a_2X_2 + \dots + a_nX_n$.

Then under certain general conditions, including\ $n$ large, $Y$\ is distributed approximately as $N(\sum_i a_i\mu_i, \sum_i a^2_i\sigma^2_i)$.

::::{.proof}
Not given.
::::
:::


Theorem\ \@ref(thm:ConvergeToN) makes no assumptions about the distribution of\ $X_i$.
If the\ $X_i$ are *normally* distributed, then $Y = \sum_{i = 1}^n a_i X_i$ will have a normal distribution for any\ $n$, large or small.


:::{.example #NoiseVoltages name="CLT (Voltages)"}
Suppose we have a number of independent noise voltages\ $V_i$ (for $i = 1, 2, \dots, n$). 
Let\ $V$ be the sum of the voltages, and suppose each\ $V_i$ is distributed $U(0, 10)$.
For $n = 20$, find $\Pr(V > 105)$.

This is an example of Theorem\ \@ref(thm:ConvergeToN) with each $a_i = 1$. 
To find $\Pr(V > 105)$, the distribution of\ $V$ must be known.

Since $\operatorname{E}[V_i] = 5$ and $\operatorname{var}[V_i] = 100/12$, from Theorem\ \@ref(thm:ConvergeToN) $V$ has an approximate normal distribution with mean $20\times 5 = 100$ and variance $20\times 100/12$.
That is, $\displaystyle{\frac{V - 100}{10\sqrt{5/3}}}$ is distributed $N(0, 1)$ approximately. 
So
$$ 
   \Pr(V > 105) 
   \approx \Pr_N\left (Z > \frac{105 - 100}{12.91}\right )
   = 1 - \Phi(0.387)
   = 0.352.
$$
The probability is approximately\ $35$%.
:::



## The normal approximation to the binomial {#NormalApproxBinomialCLT}

\index{Binomial distribution!normal approximation}
The normal approximation to the binomial distribution (Sect.\ \@ref(NormalApproxBinomial) can be seen as an application of the [Central Limit Theorem](#thm:CLT).\index{Central Limit Theorem}
The essential observation is that a sample proportion is a sample mean. 
Consider a sequence of independent [Bernoulli trials](#def:BernoulliTrials) resulting in the random sample $X_1, X_2, \dots, X_n$ where
$$
  X_i
  =
  \begin{cases}
     0  &  \text{if failure}\\
     1  &  \text{if success}
  \end{cases}
$$
denotes whether or not the $i$th\ trial is a success.
Then the sum
$$
   Y = \sum_{i = 1}^n X_i
$$
represents the number of successes in the $n$\ trials and
$$
   \overline{X}
   = 
   \frac{1}{n}\sum_{i = 1}^n X_i
   = 
   \frac{Y}{n}
$$
is a sample mean representing the proportion or fraction of trials which are successful. 
In this context, $\overline{X}$ is usually denoted by the sample proportion\ $\widehat{p}$.

Note that $\operatorname{E}[X_i] = p$ and $\operatorname{var}[X_i] = p(1 - p)$.
Therefore
$$
   \operatorname{E}[Y] = np
   \quad\text{and}\quad
   \operatorname{var}[Y] = np(1 - p)
$$
and
$$
   \operatorname{E}[\overline{X}] = p
   \quad\text{and}\quad
   \operatorname{var}[\overline{X}] = \frac{p(1 - p)}{n}.
$$
Theorems\ \@ref(thm:CLT) and\ \@ref(thm:ConvergeToN) are applicable to\ $\overline{X}$ and\ $Y$ respectively. 
Hence
$$
   \overline{X} = \widehat{p}
   \sim N\left(p, \frac{p(1 - p)}{n}\right)\text{ approximately}
$$
and
$$
   Y = n\widehat{p}
   \sim N(np, np(1 - p))\text{ approximately}.
$$



## Exercises

Selected answers appear in Sect.\ \@ref(AnswersChapSampling).

:::{.exercise #FPCF}
Consider the FPCF (Def.\ \@ref(def:FPC)).

1. What happens if the sample size is the same as the population size?
   Explain why this is a sensible result.
2. Show that, for large\ $n$, the FPCF is approximately $\sqrt{ 1 - n/N}$.
:::


:::{.exercise}
A random sample of size\ $81$ is taken from a population with mean\ $128$ and standard deviation\ $6.3$. 

1. What is the probability that an individual observation will fall between\ $126.6$ and\ $129.4$?
1. What is the probability that the *sample mean* will fall between\ $126.6$ and\ $129.4$?
1. What is the probability that the *sample mean* will *not* fall between\ $126.6$ and\ $129.4$?
:::


:::{.exercise #CLTProb}
Let $Y_1$, $Y_2$, $\dots$, $Y_n$ be $n$\ independent random variables, each with PDF
$$
   f_Y(y) = \begin{cases}
               3y^2 & \text{for $0\le y \le 1$};\\
               0 & \text{otherwise}.
            \end{cases}
$$


1. Determine the probability that a single observation will be within one standard deviation of the population mean.
1. Determine the probability that the sample mean will be within one standard deviation of the population mean, using the [*Central Limit Theorem*](#CentralLimitTheorem).
:::


:::{.exercise #EggsVariance}
Suppose the weights of eggs in a dozen carton have a weight that is normally distributed with mean\ $59\gs$ and variance\ $0.7\gs$.

1. Find the probability that, in a sample of $20$\ cartons,  the sample mean weight will exceed $59.5\gs$. 
1. Find the probability that a sample of twelve eggs will produce a sample variance of greater than\ $1$.
:::


:::{.exercise #EggsBroken}
In a carton of a dozen eggs, the number broken has a Poisson distribution with mean\ $0.2$. 

1. Find the probability that, in a sample of $20$\ cartons, the sample mean of the number of broken eggs per carton is more than one.
  (Use the *Central Limit Theorem*.)
2. Find the probability that, in any *single* carton, the probability that more than one egg is broken.
:::


:::{.exercise #ContRVM}
The random variable\ $M$ has the following probability density function
$$
    f_M(m) = 
    \begin{cases}
       3m^2 & \text{for $0 < m < 1$};\\
       0    & \text{otherwise}.
    \end{cases}
$$
A random sample of size\ $9$ is taken from the distribution, and the sample mean\ $\overline{M}$ is computed.

1. Compute the mean of\ $M$.
2. Compute the variance of\ $M$.
3. State the approximate distribution of\ $\overline{M}$ including the parameters.
4. Compute the probability that the sample mean will be within\ $0.1$ of the true mean.
:::


:::{.exercise #RandomNormalSize9}
Generate a random sample of size $n = 9$ from a $N(10, 36)$ distribution hundreds of times.
Obtain the mean and variance of\ $\sum X_i$ and\ $\overline{X}$ for each sample.

1. Verify that the distribution of the sample means is approximately normally distributed as expected.
1. Explain *why* this is expected.
:::


:::{.exercise #ABW}
Consider the random variable\ $A$, defined as
$$
    A = \frac{Z}{\sqrt{W/\nu}},
$$
where\ $Z$ has a standard normal distribution independent of\ $W$, which has a $\chi^2$-distribution with $\nu$\ degrees of freedom.
Also, consider the random variable\ $B$, defined as
$$
    B = \frac{W_1/\nu_1}{W_2/\nu_2},
$$
where\ $W_1$ and\ $W_2$ are independent $\chi^2$ variables with\ $\nu_1$ and\ $\nu_2$ degrees of freedom respectively.

1. Write down (do not derive) the distribution of\ $A$, including the parameters of the distribution.
1. Write down (do not derive) the distribution of\ $B$, including the parameters of the distribution.
1. Deduce that the distribution of\ $A^2$ is a special case of the distribution of\ $B$, and state the values of the parameters for which this is true.
:::


:::{.exercise #Pollution}
A manufacturing plant produces $2$\ tonnes of waste product on a given day, with a standard deviation of $0.2$\ tonnes per day. 
Find the probability that, over a $20$\ day period, the plant produces less than $25$\ tonnes of waste if daily productions can be assumed independent.
:::


:::{.exercise}
Let\ $Y$ be the *change* in depth of a river from one day to the next measured (in cms) at a specific location. 
Assume\ $Y$ is uniformly distributed for $y \in [-70, 70]$.

1. Find the probability that the mean change in depth for a period of $30$\ days will be greater than $10\cms$.
1. Use simulation to estimate the probability above.
:::


:::{.exercise #Typhoid}
The number of deaths per year due to typhoid fever is assumed to have a Poisson distribution with rate $\lambda = 4.1$ per year.

1. If deaths from year to year can be assumed to be independent, what is the distribution over a $20$\ year period?
1. Find the probability that there will be more than $100$\ deaths due to typhoid fever in period of $20$\ years.
:::


:::{.exercise #Lymphocyte}
The probability that a cell is a lymphocyte is\ $0.2$.

1. Write down an exact expression for the probability that in a sample containing $150$\ cells that at least\ $40$ are lymphocytes.
   Evaluate this expression using **R**.
1. Write down an approximate expression for this probability, and evaluate it.
:::



:::{.exercise}
Suppose the probability of a person aged $80+$ years dying after receiving influenza vaccine is\ $0.006$. 
In a sample of $200$\ persons aged $80+$\ years:

1. Write down an exact expression for the probability that more than\ $5$ will die after vaccination for influenza. 
   Evaluate this expression using **R**.
1. Write down an approximate expression for this probability and evaluate it.
1. If $4$\ persons died in a sample of\ $200$, what conclusion would you make about the probability of dying after vaccination?
   Justify your answer.
:::


:::{.exercise}
Illustrate the Central Limit Theorem for the uniform distribution on $[-1, 1]$ by simulation and repeated sampling, for various sample sizes.
:::



:::{.exercise}
Illustrate the Central Limit Theorem for the exponential distribution with mean\ $1$, by simulation and repeated sampling, for various sample sizes.
:::



:::{.exercise}
Demonstrate the Central Limit Theorem using **R**:

1. For the uniform distribution, where $-1 < x < 1$ (a symmetric distribution).
1. For the exponential distribution with parameter\ $1$ (a skewed distribution).
:::
    


:::{.exercise #SimTdist}
Consider a random variable\ $Z$ with a standard normal distribution $N(0, 1)$, and a random variable\ $V$ with a chi-squared distribution on\ $v$ degrees of freedom.

Simulate the distribution of
$$
  T = \frac{Z}{\sqrt{V/\nu}},
$$
then show this is a $t$-distribution with\ $\nu$ degrees of freedom.
*Hint*: use the **R** functions for the $t$-distribution (e.g., `dt()`) and the chi-squared distribution (e.g, `dchisq()`). 
:::



:::{.exercise #SimTdist2}
Consider a random variable\ $Z$ with a standard normal distribution $N(0, 1)$, and a random variable\ $V_\nu$ with a chi-squared distribution on\ $\nu$ degrees of freedom.

Simulate the distribution of
$$
  T = \frac{Z}{\sqrt{V_9/9}},
$$
then show this is a $t$-distribution with\ $\nu = 9$ degrees of freedom.
*Hint*: use the **R** functions for the $t$-distribution (e.g., `dt()`) and the chi-squared distribution (e.g, `dchisq()`). 
:::


:::{.exercise #SimHurdle}
Consider a zero-modified\index{Zero-modified distribution!Gamma modified} Gamma$(2,1)$-distribution, with $p = 0.5$.

1. Derive $\operatorname{E}[\overline{Y}_n]$ and $\operatorname{var}(\overline{Y}_n)$.
1. Simulate the distribution of $\operatorname{E}[\overline{Y}_n]$ for $n \in \{10, 50, 200\}$, overlayed with the appropriate normal approximation.
   Comment.
:::


:::{.exercise #SimCPG}
Consider a compound Poisson--gamma distributions\index{Compound Poisson--gamma distribution} with $\lambda = 1.5$, $\alpha = 2$ and $\beta = 1$.

1. Derive $\operatorname{E}[\overline{Y}_n]$ and $\operatorname{var}(\overline{Y}_n)$.
1. Simulate the distribution of $\operatorname{E}[\overline{Y}_n]$ for $n \in \{10, 50, 200\}$, overlayed with the appropriate normal  approximation.
   Comment.
:::


:::{.exercise #Unbiased1}
Let $X_1, \ldots, X_n \overset{\text{iid}}{\sim} \operatorname{Exponential}(\lambda)$, where\ $\lambda$ is the rate parameter so that $\operatorname{E}[X_i] = 1/\lambda$

1. Show that $\bar{X}$ is an unbiased estimator of $1/\lambda$.
2. Find an unbiased estimator of\ $\lambda$.
:::


:::{.exercise #Unbiased2}
Let $X_1, \ldots, X_n \overset{\text{iid}}{\sim} \operatorname{Uniform}(0, \theta)$. 

1. Show that $2\bar{X}$ is an unbiased estimator of\ $\theta$.
2. Let $M = \max(X_1, \ldots, X_n)$. 
   Derive the distribution of $M$ and show that\ $M$ is not unbiased for $\theta$.
3. Find a function of\ $M$ that is unbiased for\ $\theta$.

:::
