\part{Appendices}

# (APPENDIX) Appendices {-}





# Symbols used {#SymbolsUsed}




## Symbols and abbreviations used {#Symbols}

(ref:Sets) Chap.\ \@ref(ChapterSetTheory)

(ref:Probability) Chap.\ \@ref(ChapterProbability)

(ref:RandomVariables) Chap.\ \@ref(DistributionRandomVariables)

(ref:Expectation) Chap.\ \@ref(ChapExpectation)

(ref:ContinuousDistributions) Chap.\ \@ref(ContinuousDistributions)




```{r OtherSymbols, echo=FALSE}
SymbolsTab <- array( dim = c(38, 3))

colnames(SymbolsTab) <- c("Symbol", 
                          "Meaning", 
                          "Reference")

SymbolsTab[1, ] <- c("$\\in$", 
                     "Is an element of", 
                     "(ref:Sets)")
SymbolsTab[2, ] <- c("$\\subset$", 
                     "Proper subset", 
                     "(ref:Sets)")
SymbolsTab[3, ] <- c("$\\subseteq$", 
                     "Subset", 
                     "(ref:Sets)")
SymbolsTab[4, ] <- c("$\\cup$", 
                     "Set union; or", 
                     "(ref:Sets)")
SymbolsTab[5, ] <- c("$\\cap$", 
                     "Set intersection; and", 
                     "(ref:Sets)")
SymbolsTab[6, ] <- c("$A\\setminus B$", 
                     "Set difference", 
                     "(ref:Sets)")
SymbolsTab[7, ] <- c("Superscript $c$", 
                     "Set complement", 
                     "(ref:Sets)")
SymbolsTab[8, ] <- c("$A\\Delta B$", 
                     "Symmetric difference", 
                     "(ref:Sets)")
SymbolsTab[9, ] <- c("$|A|$", 
                     "Cardinality of set\\ $A$", 
                     "(ref:Sets)")
SymbolsTab[10, ] <- c("$\\aleph_0$", 
                     "The smallest infinite cardinality", 
                     "(ref:Sets)")
SymbolsTab[11, ] <- c("$\\mathfrak{c}$", 
                     "Cardinality of the continuum", 
                     "(ref:Sets)")
SymbolsTab[12, ] <- c("$\\mathbb{N}$", 
                     "Set of natural (counting) numbers", 
                     "(ref:Sets)")
SymbolsTab[13, ] <- c("$\\mathbb{Z}$", 
                     "Set of integers", 
                     "(ref:Sets)")
SymbolsTab[14, ] <- c("$\\mathbb{Q}$", 
                     "Set rational numbers" , 
                     "(ref:Sets)")
SymbolsTab[15, ] <- c("$\\mathbb{R}$", 
                     "Set of real numbers", 
                     "(ref:Sets)")
SymbolsTab[16, ] <- c("$\\mathbb{R}^{+}_0$", 
                     "Set of non-negative real numbers", 
                     "(ref:Sets)")
SymbolsTab[17, ] <- c("$\\mathbb{R}_{+}$", 
                     "Set of positive real numbers", 
                     "(ref:Sets)")
SymbolsTab[18, ] <- c("$\\mathbb{R}_{-}$", 
                     "Set of negative real numbers", 
                     "(ref:Sets)")
SymbolsTab[19, ] <- c("$\\mathbb{C}$", 
                     "Set of complex numbers", 
                     "(ref:Sets)")


SymbolsTab[20, ] <- c("$P^n_r$", 
                     "Permutation of\\ $r$ elements from\\ $n$", 
                     "(ref:Probability)")
SymbolsTab[21, ] <- c("$C^n_r$; $\\binom{n}{r}$", 
                     "Combination of\\ $r$ elements from\\ $n$", 
                     "(ref:Probability)")

SymbolsTab[22, ] <- c("$p_X(x)$", 
                     "Probability mass function of\\ $X$", 
                     "(ref:Probability)")
SymbolsTab[23, ] <- c("$f_X(x)$", 
                     "Probability density function of\\ $X$", 
                     "(ref:Probability)")
SymbolsTab[24, ] <- c("$F_X(x)$", 
                     "Distribution function for\\ $X$", 
                     "(ref:Probability)")

SymbolsTab[25, ] <- c("$\\operatorname{E}[X]$; $\\mu$", 
                     "Expected value, or mean, of\\ $X$", 
                     "(ref:RandomVariables)")
SymbolsTab[26, ] <- c("$\\operatorname{var}[X]$; $\\sigma^2$", 
                     "Variance of\\ $X$", 
                     "(ref:RandomVariables)")
SymbolsTab[27, ] <- c("$\\gamma_1$", 
                     "Skewness", 
                     "(ref:RandomVariables)")

SymbolsTab[28, ] <- c("$\\mathcal{R}_X$", 
                     "Range for rv $X$", 
                     "(ref:Expectation)")
SymbolsTab[29, ] <- c("$\\gamma_2$", 
                     "Excess kurtosis", 
                     "(ref:Expectation)")
SymbolsTab[30, ] <- c("$M_X(t)$", 
                     "Moment generating function (MGF)", 
                     "(ref:Expectation)")
SymbolsTab[31, ] <- c("$\\mu'_r$", 
                     "$r$th raw moment", 
                     "(ref:Expectation)")
SymbolsTab[32, ] <- c("$\\mu_r$", 
                     "$r$th central moment", 
                     "(ref:Expectation)")
SymbolsTab[33, ] <- c("$K_X(t)$", 
                     "Cumulant generating function (CGF)", 
                     "(ref:Expectation)")
SymbolsTab[34, ] <- c("$\\varphi_X(t)$", 
                     "Characteristic function", 
                     "(ref:Expectation)")
SymbolsTab[35, ] <- c("$\\kappa_r$", 
                     "$r$th cumulant", 
                     "(ref:Expectation)")


SymbolsTab[36, ] <- c("$\\phi(\\cdot)$", 
                     "Standard normal density function", 
                     "(ref:ContinuousDistributions)")
SymbolsTab[37, ] <- c("$\\Phi(\\cdot)$", 
                     "Cumulative standard normal distribution function", 
                     "(ref:ContinuousDistributions)")
SymbolsTab[38, ] <- c("$\\operatorname{erf}(x)$", 
                     "Error function", 
                     "(ref:ContinuousDistributions)")


if( knitr::is_latex_output() ) {
  kable( SymbolsTab,
         format = "latex",
         longtable = TRUE,
         booktabs = TRUE,
         escape = FALSE,
         align = c("c", "l", "c") ) %>%
#         linesep = c("\\addlinespace", "", "\\addlinespace", "",  "\\addlinespace", "\\addlinespace", "", "\\addlinespace", "", "")) %>%  # Else adds a space every 5 lines...
        # caption = "Some symbols used") %>%
	row_spec(0, bold = TRUE) %>%
  column_spec(column = 1, width = "25mm") %>% 
  kable_styling(font_size = 9) %>%
  kable_styling(latex_options = "repeat_header")
}
if( knitr::is_html_output() ) {
  kable( SymbolsTab,
                format = "html",
                booktabs = TRUE,
                align = c("c", "l", "c")) %>%
                #caption = "Some symbols used")
  row_spec(0, bold = TRUE)
}
```







# Some useful series {#UsefulSeries}

For convenience and reference, some results are given here involving series that occur frequently in probability and statistical theory.


## Finite series {#FiniteSeries}

* Sum of natural numbers:\index{Finite series}
\begin{equation}
   1 + 2 + 3 + \ldots + n 
   = \frac{n(n + 1)}{2}.
   (\#eq:SumNaturalNumbers)
\end{equation}

* Sum of squares of natural numbers:
\begin{equation}
   1^2 + 2^2 + 3^2 + \ldots + n^2
   = \frac{n}{6}(n + 1)(2n + 1).
   (\#eq:SumSquaredNaturalNumbers)
 \end{equation}
 
* Geometric series:
\begin{equation}
   a + ar + ar^2 + \ldots + ar^{n - 1} 
   = \frac{a(1 - r^n)}{1 - r}
   (\#eq:SumGeometricFinite).
\end{equation}
See the next section for the case where $n\to\infty$.

* Binomial theorem:\index{Binomial theorem}
\begin{equation}
   (a + b)^n = b^n + \binom{n}{1}ab^{n - 1} + \ldots + \binom{n}{r}a^r b^{n - r} + \ldots a^n.
   (\#eq:BinomialSeries)
\end{equation}



## Infinite series and limits {#InfiniteSeriesLimits}


* Geometric series:\index{Infinite series}
\begin{equation}
     \sum_{n = 0}^\infty a r^n
   = a + ar + ar^2 + \ldots
   \rightarrow \frac{a}{1 - r} \text{ as $n \rightarrow \infty$} \quad\text{for $|r|<1$.}
  (\#eq:SumGeometricInfinite)
\end{equation}
* Geometric-like series:
\begin{align}
     \sum_{n = 0}^\infty a n r^n
   &= ar + 2ar^2 + 3ar^3 + \ldots
   \rightarrow \frac{ar}{(1 - r)^2} \quad\text{for $|r|<1$.}
  (\#eq:SumGeometricInfinite2)\\
     \sum_{n=0}^\infty a n^2 r^n
   &= ar + 4ar^2 + 9ar^3 \ldots 
   \rightarrow \frac{ar(r + 1)}{(1 - r)^3} \quad\text{for $|r|<1$.}
  (\#eq:SumGeometricInfinite3)
\end{align}


* Exponential functions:  
\begin{align}
   \exp z
   &= \sum_{i = 1}^n \frac{z^i}{i!}
    = 1 + z + \frac{z^2}{2!} + \frac{z^3}{3!} + \ldots \quad\text{for all values of $z$}.
   (\#eq:Exponential)\\
   \exp(-z)
   &= \sum_{i = 1}^n \frac{(-z)^i}{i!}
    = 1 - z + \frac{z^2}{2!} - \frac{z^3}{3!} + \ldots \quad\text{for all values of $z$}.
   (\#eq:ExponentialNegative) \\
   e
   &= \lim_{z\to\infty} \left(1 + \frac{1}{z}\right)^z
   (\#eq:ExponentialLimit)\\
   1/e
   &= \lim_{z\to\infty} \left(1 - \frac{1}{z}\right)^z
   (\#eq:ExponentialNegLimit)
\end{align}

* Logarithmic series:  
\begin{align}
   \log(1 + z)
   &= z - \frac{z^2}{2} + \frac{z^3}{3} - \ldots \quad\text{for $|z|<1$}.
   (\#eq:Logarithmic)\\
   \log(1 - z)
   &= -z - \frac{z^2}{2} - \frac{z^3}{3} - \ldots \quad\text{for $|z|<1$}.
   (\#eq:LogarithmicNegative)
\end{align}

* Others:  
\begin{align}
   (1 - z)^{-1}
   &= 1 + z + z^2 + z^3 + \ldots \quad\text{for $|z|<1$}.
   (\#eq:Other1)\\
   (1 - z)^{-r}
   &= 1 + rz + \frac{r(r + 1)z^2}{2!} + \frac{r(r + 1)(r + 2)z^3}{3!} + \ldots \quad\text{for $|z|<1$}.
   (\#eq:Other2)
\end{align}

Note that the expressions on the RHS above do not converge to the LHS for all values of $z$.




<!-- ------------------------------------------------------------------------ -->

# Some useful integrals {#UsefulIntegrals}

\index{Integrals}

For convenience and reference, integrals that arise frequently in probability and statistical distribution theory are collected here.
Throughout, it is assumed that the parameters are such that the integrals converge (conditions are noted where they are non-obvious).
The results are stated without proof; derivations typically use integration by parts, substitution, or contour integration.

## Gamma and related integrals {#GammaIntegrals}

The *gamma function* is defined by\index{Gamma function}
\begin{equation}
  \Gamma(\alpha) = \int_0^{\infty} x^{\alpha - 1} \exp(-x)\, dx \quad\text{for $\alpha > 0$}.
  (\#eq:GammaFunction)
\end{equation}
The gamma function satisfies the recursion $\Gamma(\alpha) = (\alpha - 1)\,\Gamma(\alpha - 1)$, with $\Gamma(1) = 1$ and $\Gamma\!\left(\frac{1}{2}\right) = \sqrt{\pi}$.
For positive integers, $\Gamma(n) = (n - 1)!$.

A direct consequence is that for $\alpha > 0$ and $\beta > 0$:
\begin{equation}
  \int_0^{\infty} x^{\alpha - 1} \exp(-\beta x)\, dx
  = \frac{\Gamma(\alpha)}{\beta^{\alpha}}.
  (\#eq:GammaScaled)
\end{equation}
This is perhaps the single most-used integral in distribution theory; it underlies the gamma, exponential, chi-squared, and related distributions.

Setting $\alpha = n + 1$ for non-negative integer $n$:
\begin{equation}
  \int_0^{\infty} x^{n} \exp(-\beta x)\, dx
  = \frac{n!}{\beta^{n + 1}} \quad\text{for $n = 0, 1, 2, \ldots$ and $\beta > 0$}.
  (\#eq:GammaInteger)
\end{equation}

The *incomplete gamma integral*\index{Incomplete gamma integral} appears in CDFs of the gamma and chi-squared distributions:
\begin{equation}
  \int_0^{x} t^{\alpha - 1} \exp(-t)\, dt = \gamma(\alpha,\, x),
  (\#eq:IncompleteGamma)
\end{equation}
where $\gamma(\alpha, x)$ is the lower incomplete gamma function.\index{Lower incomplete gamma integral}
The corresponding survival integral is $\Gamma(\alpha) - \gamma(\alpha, x)$.


## Beta and related integrals {#BetaIntegrals}

The *beta function*\index{Beta function} is defined by
\begin{equation}
  B(\alpha,\, \beta)
  = \int_0^{1} x^{\alpha - 1}(1 - x)^{\beta - 1}\, dx
  = \frac{\Gamma(\alpha)\,\Gamma(\beta)}{\Gamma(\alpha + \beta)},
    \quad\text{for $\alpha > 0$ and $\beta > 0$}.
  (\#eq:BetaFunctionAppendix)
\end{equation}
This is the normalising constant of the beta distribution, and arises in order-statistic distributions and Bayesian conjugate analysis.

The substitution $x = u/(1 + u)$ converts Eq.\ \@ref(eq:BetaFunction) to
\begin{equation}
  \int_0^{\infty} \frac{u^{\alpha - 1}}{(1 + u)^{\alpha + \beta}}\, du
  = B(\alpha,\, \beta),
  \qquad \alpha > 0,\; \beta > 0,
  (\#eq:BetaF)
\end{equation}
which appears in the derivation of the $F$-distribution.

More generally, for $a > 0$ and $b > 0$:
\begin{equation}
  \int_0^{\infty} \frac{x^{\alpha - 1}}{(a + bx)^{\alpha + \beta}}\, dx
  = \frac{B(\alpha,\,\beta)}{a^{\beta}\, b^{\alpha}}.
  (\#eq:BetaGeneralised)
\end{equation}


## Gaussian integrals {#GaussianIntegrals}

The fundamental Gaussian integral is\index{Gaussian integral}
\begin{equation}
  \int_{-\infty}^{\infty} \exp(-x^2)\, dx = \sqrt{\pi}.
  (\#eq:GaussianFundamental)
\end{equation}
Rescaling gives, for $\sigma > 0$:
\begin{equation}
  \int_{-\infty}^{\infty} \exp(-x^2 / (2\sigma^2))\, dx = \sigma\sqrt{2\pi}.
  (\#eq:GaussianScaled)
\end{equation}
This is the normalising constant of the normal distribution.

For the moments of the standard normal, using symmetry and integration by parts:
\begin{align}
  \int_{-\infty}^{\infty} x^{2k}\, \exp(-x^2/2)\, dx
  &= \sqrt{2\pi}\,(2k - 1)!! \quad\text{for $k = 1, 2, 3, \ldots$}
  (\#eq:GaussianEvenMoment)\\
  \int_{-\infty}^{\infty} x^{2k+1}\, \exp(-x^2/2)\, dx
  &= 0,
  (\#eq:GaussianOddMoment)
\end{align}
where $(2k - 1)!! = 1 \cdot 3 \cdot 5 \cdots (2k - 1)$ is the double factorial.\index{Double factorial}

The *moment-generating integral* of the normal is\index{Normal distribution!MGF}
\begin{equation}
  \int_{-\infty}^{\infty}
    \exp\!\left(tx - \frac{x^2}{2\sigma^2}\right) dx
  = \sigma\sqrt{2\pi}\,\exp\!\left(\frac{\sigma^2 t^2}{2}\right),
  (\#eq:GaussianMGF)
\end{equation}
obtained by completing the square in the exponent.
This result is used in deriving the MGF of the normal distribution.


## Exponential and Laplace integrals {#ExponentialIntegrals}

For $\lambda > 0$ and $n = 0, 1, 2, \ldots$:\index{Exponential integral}
\begin{equation}
  \int_0^{\infty} x^n \exp(-\lambda x)\, dx 
  = \frac{n!}{\lambda^{n + 1}}.
  (\#eq:ExponentialMoment)
\end{equation}
(This is Eq.\ \@ref(eq:GammaInteger) restated for emphasis.)

The bilateral (two-sided) exponential integral, useful for the Laplace distribution, is
\begin{equation}
  \int_{-\infty}^{\infty} |x|^n \exp(-\lambda |x|)\, dx
  = \frac{2\, n!}{\lambda^{n + 1}} \quad\text{where $n = 0, 1, 2, \ldots$}.
  (\#eq:LaplaceIntegral)
\end{equation}

The *convolution integral*\index{Convolution integral} for two exponentials with possibly different rates ($\lambda_1 \neq \lambda_2$, both positive) is
\begin{equation}
  \int_0^x \exp(-\lambda_1 t)\, \exp(-\lambda_2(x - t))\, dt
  = \frac{\exp(-\lambda_2 x) - \exp(-\lambda_1 x)}{\lambda_1 - \lambda_2}.
  (\#eq:ExponentialConvolution)
\end{equation}
This arises in the distribution of the sum of two independent exponential random variables with different rates.



<!-- 

## Integration by parts?????????????? standard forms {#IBPForms}

Integration by parts ($\int u\, dv = uv - \int v\, du$) generates several
results that recur in moment calculations.

For $n \geq 1$ and $\lambda > 0$, repeated integration by parts gives:
\begin{equation}
  \int x^n e^{-\lambda x}\, dx
  = -e^{-\lambda x} \sum_{k=0}^{n} \frac{n!}{(n-k)!\,\lambda^{k+1}}\, x^{n-k} + C.
  (\#eq:IBPGeneral)
\end{equation}
When evaluated on $(0, \infty)$ the boundary terms vanish (assuming $\lambda > 0$)
and Eq.\ \@ref(eq:GammaInteger) is recovered.

The **LOTUS identity** for integration by parts with a survival function:
\begin{equation}
  \int_0^{\infty} \bar{F}(x)\, dx = \mathrm{E}[X],
  (\#eq:SurvivalMean)
\end{equation}
valid for non-negative random variables with $\bar{F}(x) = P(X > x)$.
More generally,
\begin{equation}
  \int_0^{\infty} x^{n-1} \bar{F}(x)\, dx = \frac{\mathrm{E}[X^n]}{n},
  \qquad n \geq 1.
  (\#eq:SurvivalMoment)
\end{equation}
-->


## Characteristic function integrals {#CFIntegrals}

The *inversion formula* for recovering the PDF from the characteristic function $\varphi_X(t) = \operatorname{E}[\exp(itX)]$\index{Characteristic function} is\index{Characteristic function!inversion}
\begin{equation}
  f_X(x)
  = \frac{1}{2\pi} \int_{-\infty}^{\infty} \varphi_X(t)\exp(-itx)\, dt.
  (\#eq:CFInversion)
\end{equation}
(See Sect.\ \@ref(DistributionFromMGF) and the surrounding discussion.)

Two integrals that arise in evaluating specific characteristic functions via contour integration are
\begin{align}
  \int_{-\infty}^{\infty} \frac{\cos(tx)}{1 + t^2}\, dt 
  &= \pi \exp(-|x|),
   (\#eq:CauchyCF1)\\
  \int_{-\infty}^{\infty} \frac{t\sin(tx)}{1 + t^2}\, dt 
  &= \pi\, \mathrm{sgn}(x)\, \exp(-|x|),
  (\#eq:CauchyCF2)
\end{align}
and together give
\begin{equation}
  \frac{1}{2\pi}\int_{-\infty}^{\infty}
    \frac{\cos(tx) + t\sin(tx)}{1 + t^2}\, dt
  = \frac{1}{2}\bigl(1 + \mathrm{sgn}(x)\bigr)\, \exp(-|x|)
  = \exp(-x)\,\mathbf{1}(x \geq 0),
  (\#eq:ExponentialCFInversion)
\end{equation}
which is the result used in Example\ \@ref(exm:DistFromMGFContinuous) to invert the exponential characteristic function.


## Dirichlet and order-statistic integrals {#DirichletIntegrals}

The *Dirichlet integral*,\index{Dirichlet integral} a multivariate generalisation of the beta function, is
\begin{equation}
  \int_0^1 \cdots \int_0^1
    x_1^{\alpha_1 - 1} \cdots x_k^{\alpha_k - 1}
    (1 - x_1 - \cdots - x_k)^{\alpha_{k+1} - 1}\, dx_1 \cdots dx_k
  = \frac{\prod_{i=1}^{k+1} \Gamma(\alpha_i)}{\Gamma\!\left(\sum_{i=1}^{k+1}\alpha_i\right)},
  (\#eq:DirichletIntegral)
\end{equation}
where the integration is over the $k$-dimensional simplex $\{x_i > 0,\; \sum x_i < 1\}$.
This is the normalising constant of the Dirichlet distribution.

The *order-statistic integral*\index{Order-statistic integral!inversion} gives the joint density of the $r$-th and $s$-th order statistics ($r < s$) from a sample of size $n$:
\begin{equation}
  \int_x^y (F(t) - F(x))^{s - r - 1}\, f(t)\, dt
  = \frac{(F(y) - F(x))^{s - r}}{s - r},
  (\#eq:OrderStatIntegral)
\end{equation}
a simple power substitution that reduces the marginalisation integrals for order statistics.



<!-- -------------------------------------------------------------------------------------------- -->
# Short R introduction {#ShortRIntro}

## General ideas

\index{R@\textbf{R}}
When you start **R**, you are presented with a *prompt* asking you enter an instruction:

```
>
```

We do not show this prompt with the **R** code in this book, to enable any code that we show do be cut-and-paste directly into **R** and evaluated.
The `>` is how **R** prompts you to enter something for it to do.\index{pi@\texttt{pi}}\index{R!prompt@\textbf{R}}

**R** can be used a fancy calculator:

```{r}
(-5 + 7) * 2
2 * pi * 3
exp(-1)
sqrt( 9 ) * sin(3 * pi / 2)
```

The output begins with `[1]` to indicate the first item of output; sometimes, numerous values appear in the output (see Sect.\ \@ref(RVector) for examples).

Spaces are not important to **R**, so these are all the same:

```{r}
1 +                2
1          +  2
1+2
1 + 2
```

The last is better for human readability, so is recommended.

**R** also allows *variables* to be used, and values can be assigned to variables using `=`\index{=@\texttt{=}  (assignment)} or the more traditional `<-`\index{<-@\texttt{<-} (assignment)} combination of characters. 
Variables cannot start with a digit.\index{pi@\texttt{pi}}\index{sqrt()@\texttt{sqrt()}}\index{\^@\texttt{\^}}

```{r}
Radius = 2
Area = pi * Radius^2
Area

side1 <- 2
side2 <- 4
hypot <- sqrt( side1^2 + side2^2 )
hypot
```

The *value* of a variable is displayed by typing the variable name.

Comments can be added using `#` followed by text; any text following a `#` is ignored by **R**:\index{R@\textbf{R}!comments}\index{\#@\texttt{\#} (comment character)}


```{r}
hourly_Wage <- 20  # Hourly wage, in $ per hr
hours_Worked <- 16 # Number of hours worked

# This all assumes no overtime rates!

weekly_Income <- hourly_Wage * hours_Worked # Income in dollar for the week
weekly_Income
```

In complete lines of instructions cause **R** to wait for the instruction to be completed.\index{R@\textbf{R}!line continuation}

For instance, suppose you type:

```{r eval=FALSE}
sqrt( 4  # Notice: no closing of parentheses
```

**R** will change the prompt from a `>` to `+`, indicating that you need to *add* more information to complete the command.\index{R!prompt@\textbf{R}}
Once the command is complete, **R** evaluates the instruction:\index{sqrt()@\texttt{sqrt()}}

```{r}
sqrt(4
)
```

## Using functions in R {#FunctionsInR}

**R** has thousands of functions to perform specific tasks.\index{R@\textbf{R}!standard functions}
We have already seen two being used above:

* `sqrt()` takes the **sq**uare **r**oo**t** of a given value.\index{sqrt()@\texttt{sqrt()}}
* `exp()` find the value of $\exp(x) = \exp x$ for some given value of\ $x$.\index{exp()@\texttt{exp()}}
* `log()` finds the **log**arithm of a function.\index{log()@\texttt{log()}}

Functions, in general, can take more than one input.
Consider the function `log()`.
We can type

```{r}
log(100)
```

If we look at the help for the `log()` function (by typing `?log`), we see the usage is described like this:\index{?@\texttt{?} (\textbf{R} help)}

```
Usage
log(x, base = exp(1))
```

That is, the function can take *two* inputs: one called `x`, and one called `base`.
The help information is starting that the default value of `base`, if it is not told otherwise, is assumed to be `exp(1)` (which is $\exp(1)= \exp 1 = 2.71828\dots$).
In other words, the default logarithm is the *natural logarithm*.

Since the command above (i.e., `log(100)`) only gave one input to the `log()` function, the value of the second input (i.e., `base`) uses the default value of `exp(1)`.
That is, $\log_e 100 = \ln 100 = 4.605\dots$.

To specify a different base, we can set the value of `base` explicitly:

```{r}
log(100, 10)
```

The inputs are assumed by **R** to have been given in the order shown in the help; that is `x = 100` and `base = 10`. 
It is clearer, though, to *name* the value of the second input:

```{r}
log(100, base = 10)
```

The value of the first input can be named also:

```{r}
log(x = 100, base = 10)
```
When inputs are *named*, they can appear in any order:

```{r}
log(base = 10, x = 100)
```

Some functions in **R** have *many* inputs.
Some functions are collected into _**R** packages_ that provide additional functionality. 


## General functions {#OtherUsefulRGeneral}

\index{R@\textbf{R}!standard functions}

* `seq()` produces a **seq**uence of integers:\index{seq()@\texttt{seq()}}
  * `seq(1, 4)` produces the list: $1$, $2$, $3$, $4$.
  * `seq(0, 10, by = 2)` produces a list going up **by** two each time: $0$, $2$, $4$, $6$, $8$, $10$.
  * `seq(0, 10, length = 3)` produces a list of **length** three: $0$, $5$, $10$.
  * A colon can also be used in special cases: `3 : 7` produces: $3$, $4$, $5$, $6$, $7$.\index{:@\texttt{:}}
* `c()` is used to **c**oncatenate (join together) a series of values:\index{c()@\texttt{c()}}
  * `c("fred", "martha")` creates a vector with two text elements.
  * `c(1, 8, 3.14, -2)`create a vector with four numerical elements.
* `cat()` is often used to print information.\index{cat()@\texttt{cat()}}
* `matrix()` is used to produces matrices (of $2$-dimensions):\index{matrix()@\texttt{matrix()}}
  * `matrix(data = c(1, 2, 3, 4, 5, 6), byrow = TRUE, ncol = 3)` produces the $2\times 3$ array (i.e., `ncol = 3` means the number of columns is\ $3$):
$$
  \left[
  \begin{array}{ccc}
    1 & 2 & 3\\
    4 & 5 & 6\\
  \end{array}
  \right].
$$

```{r}
matrix(data = c(1, 2, 3, 4, 5, 6), 
       byrow = TRUE, 
       ncol = 3)
```
  * `matrix(data = c(1, 2, 3, 4, 5, 6), bycol = TRUE, nrow = 2` produces the $2\times 3$ array (i.e., `nrow = 2` means the number of rows is\ $2$):
$$
  \left[
  \begin{array}{ccc}
    1 & 3 & 5\\
    2 & 4 & 6\\
  \end{array}
  \right].
$$
* `numeric()` creates a numeric vector.\index{numeric()@\texttt{numeric()}}
  * `numeric(4)` produces an empty vector of length\ $4$.
* `array()` produces matrix-like structure, but they can have many dimensions (not just two, like `matrix()`)\index{array()@\texttt{array()}}
  * `array( dim = c(2, 4))` produces an empty $2\times 4$ array.
  * `array( data = c(1, 2, 3, 4, 5, 6, 7, 8), dim = c(2, 2, 2))` produces a $2\times 2\times 2$ array:
```{r}
array( data = c(1, 2, 3, 4, 5, 6, 7, 8), dim = c(2, 2, 2))
```
* `t()` **t**ransposes a matrix or array:\index{t()@\texttt{t()}}

```{r}
A <- matrix( c(1, -2, -20, 3), ncol = 2)
A
t(A)
```

Logical comparisons in **R** are possible (using of `==` rather than `=`):\index{R@\textbf{R}!logic operators}\index{==@\texttt{==} (logical EQUALS)}\index{>@\texttt{>}}\index{>=@\texttt{>=}}\index{\textbar@\texttt{\textbar} (logical OR)}\index{\&@\texttt{\&} (logical AND)}\index{R|indexing()@\textbf{R}!indexing}\index{\textbackslash n@\texttt{\textbackslash n} (new line)}

```{r}
Days_Of_Week <- c("Mon", "Tues", "Wed", "Thurs", "Fri", "Sat", "Sun")
Hours_Sleep <- c(8, 8, 8, 8, 6, 6, 10)

Hours_Sleep == 8
Hours_Sleep > 8
Hours_Sleep >= 8

Hours_Sleep == 6 | Hours_Sleep == 10 # | means "OR"

Hours_Sleep > 6 & Hours_Sleep < 10 # & means "AND"

which(Hours_Sleep == 8)
which(Hours_Sleep > 8)
which(Hours_Sleep >= 8)

which(Hours_Sleep == 6 | Hours_Sleep == 10 ) # | means "OR"

which(Hours_Sleep > 6 & Hours_Sleep < 10 ) # & means "AND"

Weekend <- Days_Of_Week == "Sat" | Days_Of_Week == "Sun"

Week_Day <- !Weekend # ! means "NOT"

Hours_Sleep[Week_Day]

Hours_Sleep[Weekend]


cat("These are the hours of sleep on weekdays:", Hours_Sleep[Week_Day], "\n")
   # \n  is used to create a new line.
```

Specific elements of a vector can be accessed using square brackets:\index{R|indexing()@\textbf{R}!indexing}

```{r}
Hours_Sleep[1]
Hours_Sleep[3:5]

Hours_Sleep[Weekend]
```



## Vector operations {#RVector}

**R** is a *vectorised* system; that is, operations work on *all* elements of a vector:\index{R@\textbf{R}!vector operations}\index{cos()@\texttt{cos()}}

```{r}
x <- 0 : 8   # x  is a vector
x
x + 3
2 * x

sqrt(x) # The square root function
cos(x)  # The cosine function
exp( -x ) # The exponential function
```


## Statistical functions {#OtherUsefulRStatistical}

Because **R** is primarily a statistical package and environment, all the basic statistical functions are available, including:\index{R@\textbf{R}!statistical functions}


* `mean()`: Find the **mean** of a *sample* of values.\index{mean()@\texttt{mean()}}
   Example: `mean( c(1, 2, 3, 4))`.
* `median()`: Find the **median** of a *sample* of values.\index{median()@\texttt{median()}}
   Example: `median( c(1, 2, 3, 4))`.
* `sd()`: Finds the **s**tandard **d**eviation of a *sample* of values.\index{sd()@\texttt{sqrt()}}
   Example: `sd( c(1, 2, 3, 4))`.
* `var()`: Finds the **var**iance of a *sample* of values.\index{var()@\texttt{var()}}
   Example: `var( c(1, 2, 3, 4))`.
* `lm()`: To fit a **l**inear regression **m**odel:\index{lm()@\texttt{lm()}}

```{r}
x <- c(1, 2, 3, 4, 5)
y <- c(12, 10, 9, 7, 4)

out <- lm(y ~ x) # Read: 'y as a function of x'
coef(out)
summary(out)
```


## Probability and counting functions {#RProbCounting}

\index{R@\textbf{R}!counting functions}

* the *number* of *combinations* of `n` elements taken `k` at a time is found using `choose(n, k)`.\index{choose()@\texttt{choose()}}
* $n!$ is given by `factorial(n)`.\index{factorial()@\texttt{factorial()}}
* the *number* of *permutations* of `n` elements taken `k` at a time is found using `choose(n, k) * factorial(k)`.
* a list of all combinations of `n` elements, `m` at a time, is given by `combn(x, m)`.\index{combn()@\texttt{combn()}}
* $\Gamma(x)$ is given by `gamma(x)`.\index{gamma()@\texttt{gamma()}}




## Plotting functions {#OtherUsefulRPlotting}

Three systems exists for plotting in **R**, believe it or not.
Here, we discuss the base system.\index{R@\textbf{R}!plotting functions}


`plot()` is the basic function for plotting, with *many* options:\index{plot()@\texttt{plot()}}

* `plot(x, y)` and `plot(y ~ x)` both produce a scatterplot, with `y` on the vertical axis and `x` on the horizontal axis.
* `plot(..., type = "l")` plots with **l**ines rather than points.
* `plot(..., lwd = "3")` plots with a **l**ine **w**i**d**th thrice as thick.
* `plot(..., xlab = "text", ylab = "Info")` adds an **lab**el to the **x** and **y** axes respectively.
* `plot(..., main = "Title")` adds a **main** title to the plot.
* `plot(..., col = "green")` plots in green rather than the default (black).

`plot()` starts a new canvas every time it is called.
`lines()` and `points()` add **lines** and **points** respectively to an existing plot.\index{lines()@\texttt{lines()}}\index{points()@\texttt{points()}}
`legend()` adds a legend to an existing plot.\index{legend()@\texttt{legend()}}


```{r ExamplePlotExponential, fig.align='center', fig.cap="An example plot: the probability function for two exponential distributions", out.width='60%', fig.width=5.5}
x <- seq(0, 4, length = 100)
y1 <- dexp(x, rate = 2) # Probability function for an exponential distn

plot(y1 ~ x,
     type = "l", # Use lines, not points
     lwd = 2, # Make lines a bit thicker
     xlab = "Values of x",
     ylab = "Probability function",
     main = "Probability function for\nexponential distribution",
      # NOTE: using  \n  adds a line break, and # is used for comments
     col = "red", 
     las = 1, # Makes the axis labels all horizontal
     xlim = c(0, 4.5), # Changes the displayed limits of the x-axis
     ylim = c(0, 3) ) # Changes the displayed limits on the y-axis

y2 <- dexp(x, rate = 3)
lines( y2 ~ x, # ADD a line to existing point
       lwd = 2,
       col = "blue")

legend("topright", # The location
       lwd = 2,
       col = c("red", "blue"),
       legend = c("Rate = 2",
                  "Rate = 3")
       )

```


# Using R with distributions {#UseRDistributions}

In **R**, standard distributions have four associated functions:\index{R@\textbf{R}!standard distributions}


* Computing the probability function: All begin with `p` (such as `pnorm()` for the normal distribution).
* Computing the distribution function: All begin with `d`(such as `dnorm()`).
* Computing the quantile function: All begin with `q` (such as `qnorm()`).
* Generating random numbers: All begin with `r` (such as `rnorm()`).

For example, for the normal distribution (denoted `norm` in **R**) the four functions above are called:

* `dnorm()`
* `pnorm()`
* `qnorm()`
* `rnorm()`





## Univariate distributions

Common parameters:

* `x`: Values where to evaluate the probability function
* `q`: Quantiles
* `p`: Probabilities
* `n`: The number of random observations to generate

Most distributions have other parameters also; check the help (e.g., `?punif`).

```{r RDistributionsFunctions, echo=FALSE}
#library(knitr)
#library(kableExtra)

# Function to put R function in tt mode
code_fn <- function(x) {
  if (knitr::is_html_output()) {
    sprintf("<code>%s</code>", x)
  } else {
    sprintf("\\texttt{%s}", x)
  }
}




# Merge some celss in some rows: For parameters
span_cols <- function(df, row, start_col = 2, num_cols = 5, 
                      text, align = "center", bold = TRUE) {
  # 1. Detect the current knitr output format
  # Defaults to HTML if run interactively in the console
  fmt <- knitr::opts_current$get("rmarkdown.pandoc.to")
  if (is.null(fmt)) fmt <- "html" 
  
  # 2. Format the text (apply bold if requested)
  formatted_text <- text
  
  # 3. Generate format-specific syntax
  if (grepl("html", fmt)) {
    style <- paste0("text-align:", align, ";")
    if (bold) style <- paste0(style, " font-weight:bold;")
    
    df[row, start_col] <- paste0("<td colspan='", num_cols, "' style='", style, "'>", formatted_text, "</td>")
    
  } else if (grepl("latex|pdf", fmt)) {
    lat_align <- switch(align, "left" = "l", "right" = "r", "c")
    if (bold) formatted_text <- paste0("\\textbf{", formatted_text, "}")
    
    df[row, start_col] <- paste0("\\multicolumn{", num_cols, "}{", lat_align, "}{", formatted_text, "}")
  }
  
  # 4. Clear out the subsequent columns so they don't break table alignment
  end_col <- start_col + num_cols - 1
  if (num_cols > 1) {
    df[row, (start_col + 1):end_col] <- ""
  }
  
  return(df)
}

# 
# 
# 
# library(flextable)
# 
# # 1. Your original data frame
# my_data <- data.frame(
#   A = c("R1C1", "R2C1", "R3C1"),
#   B = c("R1C2", "R2C2", "R3C2"),
#   C = c("R1C3", "R2C3", "R3C3"),
#   D = c("R1C4", "R2C4", "R3C4"),
#   stringsAsFactors = FALSE
# )
# 
# # 2. Assign text to the main cell
# my_data[2, "B"] <- "This spans 3 columns!"
# 
# # 3. Build and merge using flextable
# ft <- flextable(my_data) %>%
#   # Merge row 2, from column 2 (B) to column 4 (D)
#   merge_at(i = 2, j = 2:4, part = "body") %>%
#   # Apply text formatting properties
#   align(i = 2, j = 2, align = "center", part = "body") %>%
#   bold(i = 2, j = 2, part = "body") %>%
#   # Match the kable styling look
#   theme_booktabs() %>%
#   autofit()
# 
# # 4. Print table (flextable automatically renders to HTML or PDF based on output)
# ft
# 
# 
# 


# Parameters: $n$ (`size`) and so on

###


dist_table <- data.frame(
  Distribution = c(
    "Discrete uniform",
    "Binomial",
    "Poisson",
    "Geometric",
    "Negative binomial",
    "Continuous uniform",
    "Normal",
    "Exponential",
    "Gamma",
    "Log-normal",
    "$t$",
    "$\\chi$-squared",
    "$F$"
  ),
  Prob = c(
    "",
    code_fn("dbinom()"),
    code_fn("dpois()"),
    code_fn("dgeom()"),
    code_fn("dnbinom()"),
    code_fn("dunif()"),
    code_fn("dnorm()"),
    code_fn("dexp()"),
    code_fn("dgamma()"),
    code_fn("dlnorm()"),
    code_fn("dt()"),
    code_fn("dchisq()"),
    code_fn("df()")
  ),
  Dist = c(
    "",
    code_fn("pbinom()"),
    code_fn("ppois()"),
    code_fn("pgeom()"),
    code_fn("pnbinom()"),
    code_fn("punif()"),
    code_fn("pnorm()"),
    code_fn("pexp()"),
    code_fn("pgamma()"),
    code_fn("plnorm()"),
    code_fn("pt()"),
    code_fn("pchisq()"),
    code_fn("pf()")
  ),
  Quant = c(
    "",
    code_fn("qbinom()"),
    code_fn("qpois()"),
    code_fn("qgeom()"),
    code_fn("qnbinom()"),
    code_fn("qunif()"),
    code_fn("qnorm()"),
    code_fn("qexp()"),
    code_fn("qgamma()"),
    code_fn("qlnorm()"),
    code_fn("qt()"),
    code_fn("qchisq()"),
    code_fn("qf()")
  ),
  Random = c(
    code_fn("sample()"),
    code_fn("rbinom(n, )"),
    code_fn("rpois(n, )"),
    code_fn("rgeom(n, )"),
    code_fn("rnbinom(n, )"),
    code_fn("runif(n, )"),
    code_fn("rnorm(n, )"),
    code_fn("rexp(n, )"),
    code_fn("rgamma(n, )"),
    code_fn("rlnorm(n, )"),
    code_fn("rt(n, )"),
    code_fn("rchisq(n, )"),
    code_fn("rf(n, )")
  ),
  check.names = FALSE
)



if (knitr::is_html_output()) {

  kable(
    dist_table,
    format = "html",
    escape = FALSE,
    booktabs = TRUE,
    col.names = c(
      "Distribution",
      "function",
      "function",
      "function",
      "numbers"
    ),
    align = c("l", "c", "c", "c", "c")
  ) |>
    add_header_above(
      c(
        " " = 1,
        "Probability" = 1,
        "Distribution" = 1,
        "Quantile" = 1,
        "Random" = 1
      ),
      bold = TRUE,
      line = FALSE
    ) |>
    kable_styling(full_width = FALSE) |>
    row_spec(0, bold = TRUE)

} else {

  kable(
    dist_table,
    format = "latex",
    escape = FALSE,
    booktabs = TRUE,
    col.names = c(
      "Distribution",
      "function",
      "function",
      "function",
      "numbers"
    ),
    align = c("l", "c", "c", "c", "c")
  ) |>
    add_header_above(
      c(
        " " = 1,
        "Probability" = 1,
        "Distribution" = 1,
        "Quantile" = 1,
        "Random" = 1
      ),
      bold = TRUE,
      line = FALSE
    ) |>
    row_spec(0, bold = TRUE)
}
```

\index{dbinom()@\texttt{dbinom()}}\index{pbinom()@\texttt{pbinom()}}\index{qbinom()@\texttt{qbinom()}}\index{rbinom()@\texttt{rbinom()}}
\index{dpois()@\texttt{dpois()}}\index{ppois()@\texttt{ppois()}}\index{qpois()@\texttt{qpois()}}\index{rpois()@\texttt{rpois()}}
\index{dgeom()@\texttt{dgeom()}}\index{pgeom()@\texttt{pgeom()}}\index{qgeom()@\texttt{qgeom()}}\index{rgeom()@\texttt{rgeom()}}
\index{dnbinom()@\texttt{dnbinom()}}\index{pnbinom()@\texttt{pnbinom()}}\index{qnbinom()@\texttt{qnbinom()}}\index{rnbinom()@\texttt{rnbinom()}}
\index{dunif()@\texttt{dunif()}}\index{punif()@\texttt{punif()}}\index{qunif()@\texttt{qunif()}}\index{runif()@\texttt{runif()}}
\index{dnorm()@\texttt{dnorm()}}\index{pnorm()@\texttt{pnorm()}}\index{qnorm()@\texttt{qnorm()}}\index{rnorm()@\texttt{rnorm()}}
\index{dgamma()@\texttt{dgamma()}}\index{pgamma()@\texttt{pgamma()}}\index{qgamma()@\texttt{qgamma()}}\index{rgamma()@\texttt{rgamma()}}
\index{dexp()@\texttt{dexp()}}\index{pexp()@\texttt{pexp()}}\index{qexp()@\texttt{qexp()}}\index{rexp()@\texttt{rexp()}}
\index{dlnorm()@\texttt{dlnorm()}}\index{plnorm()@\texttt{plnorm()}}\index{qlnorm()@\texttt{qlnorm()}}\index{rlnorm()@\texttt{rlnorm()}}
\index{dt()@\texttt{dt()}}\index{pt()@\texttt{pt()}}\index{qt()@\texttt{qt()}}\index{rt()@\texttt{rt()}}
\index{dchisq()@\texttt{dchisq()}}\index{pchisq()@\texttt{pchisq()}}\index{qchisq()@\texttt{qchisq()}}\index{rchisq()@\texttt{rchisq()}}
\index{df()@\texttt{df()}}\index{pf()@\texttt{pf()}}\index{qf()@\texttt{qf()}}\index{rf()@\texttt{rf()}}





## Multivariate distributions 

The **multinomial** distribution: \index{dmultinom()@\texttt{dmultinom()}}\index{rmultinom()@\texttt{rmultinom()}}

* Probability function: `dmultinom()`
* Random number generation: `rmultinom()`

The **multivariate normal** distribution:\index{dmnorm()@\texttt{dmnorm()}}\index{pmnorm()@\texttt{pmnorm()}}\index{rmnorm()@\texttt{rmnorm()}}
  
* Probability function: `dmnorm()`
* Distribution function: `pmnorm()`
* Random number generation: `rmnorm()`


