\part{Appendices}

# (APPENDIX) Appendices {-}



# Some useful series {#UsefulSeries}

For convenience and reference we state some results here involving series that occur quite frequently in probability and statistical theory.


## Finite series

* Sum of natural numbers:
\begin{equation}
   1 + 2 + 3 + \ldots + n 
   = \frac{n(n + 1)}{2}.
   (\#eq:SumNaturalNumbers)
\end{equation}

* Sum of squares of natural numbers:
\begin{equation}
   1^2 + 2^2 + 3^2 + \ldots + n^2
   = \frac{n}{6}(n + 1)(2n + 1)
   (\#eq:SumSquaredNaturalNumbers)
 \end{equation}
 
* Geometric series
\begin{align}
   a + ar + ar^2 + \ldots + ar^{n - 1}
   &= \frac{a(1 - r^n)}{1 - r}
   (\#eq:SumGeometricFinite)\\
   &\rightarrow \frac{a}{1 - r} \text{ as $n  \rightarrow  \infty$}
  (\#eq:SumGeometricInfinite)
\end{align}
where the infinite series only converges if $|r| < 1$.

* Binomial expansion:
\begin{equation}
   (a + b)^n = b^n + {n \choose 1}ab^{n - 1} + \ldots + {n \choose r}a^r b^{n - r} + \ldots a^n.
   (\#eq:BinomialSeries)
\end{equation}



## Power series

A power series is a series of the form $\sum_{n = 0}^{\infty}a_n z^n$.

* Exponential function:
\begin{align}
   e^z
   &= 1 + z + \frac{z^2}{2!} + \frac{z^3}{3!} + \ldots
   (\#eq:Exponential)\\
   e^{-z}
   &= 1 - z + \frac{z^2}{2!} - \frac{z^3}{3!} + \ldots
   (\#eq:ExponentialNegative)
\end{align}

* Logarithmic series:
\begin{align}
   \log(1 + z)
   &= z - \frac{z^2}{2} + \frac{z^3}{3} - \ldots
   (\#eq:Logarithmic)\\
   \log(1 - z)
   &= -z - \frac{z^2}{2} - \frac{z^3}{3} - \ldots
   (\#eq:LogarithmicNegative)
\end{align}

* Others:
\begin{align}
   (1 - z)^{-1}
   &= 1 + z + z^2 + z^3 + \ldots
   (\#eq:Other1)\\
   (1 - z)^{-r}
   &= 1 + rz + \frac{r(r + 1)z^2}{2!} + \frac{r(r + 1)(r + 2)z^3}{3!} + \ldots
   (\#eq:Other2)
\end{align}
Note that the expressions on the RHS above do not converge to the LHS for all values of $z$.

DOES THIS COMMENT APPLY TO ALL????



# Using R with univariate distributions {#UseRDistributions}

In **R**, standard distributions have four associated function:

* Computing the [probability function](#ProbabilityFunction): All begin with `p` (such as `pnorm()` for the normal distribution).
* Computing the [distribution function](DistributionFunction): All begin with `d`(such as `dnorm()`).
* Compiting the quantile function (XREF): All begin with `q` (such as `qnorm()`).
* Generating random numbers (XREF): All begin with `r` (such as `rnorm()`).

ADD PARAMETERS TOO... if enough room.

In **R**, four function are available for most distributions

* Computing the [probability function](#ProbabilityFunction): Functions start with `d`
* Computing the [distribution function](#DistributionFunction): Functions start with `p`
* Computing the quantiles: Functions start with `q`
* Generatimg random numbers: Functions start with `r`.

For example, the fours function above are called, for the normal distribution (denoted `norm` in **R**):

* `dnorm()`
* `pnorm()`
* `qnorm()`
* `rnorm()`

ADD a chapter on how to generate random numbers and quantiles and so on.



## Discrete distributions

Distribution       | **R** function          | Common parameters
-------------------|-------------------------|-----------------------
Discrete uniform   | `sample()`              | ~                     
Binomial           | `dbinom(x, size, prob)` | `x`:  Values where to evaluate the probability function
.                  | `pbinom(q, size, prob)` | `q`: Quantiles
.                  | `qbinom(p, size, prob)` | `p`: Probabilities
.                  | `rbinom(n, size, prob)` | `n`: The number of random observations to generate
.                  |  .                      | `size`: The number of trials
.                  |  .                      | `prob`: The probability of 'success' in each trial
Poisson            | `dpois(x, lambda)`      | `x`: Values where to evaluate the probability function  
.                  | `ppois(q, lambda)`      | `q`: Quantiles
.                  | `qpois(p, lambda)`      | `p`: Probabilities
.                  | `rpois(n, lambda)`      | `n`: The number of random observations to generate
.                  |  .                      | `lambda`: The Poisson mean
Geometric          | `dgeom(x, prob)`        | `x`: Values where to evaluate the probability function  
.                  | `pgeom(q, prob)`        | `q`: Quantiles
.                  | `qgeom(p. prob)`        | `p`: Probabilities   
.                  | `rgeom(n, prob)`        | `n`: The number of random observations to generate
.                  |  .                      | `prob`: The probability of 'success' in each trial
Negative binomial  | `dnbinom(x, size, prob, mu)` | `x`: Values where to evaluate the probability function          
.                  | `pnbinom(q, size, prob, mu)` | `q`: Quantiles
.                  | `qnbinom(p, size, prob, mu)` | `p`: Probabilities   
.                  | `rnbinom(p, size, prob, mu)` | `n`: The number of random observations to generate
.                  |  .                           | `size`: The number of trials
.                  |  .                           | `prob`: The probability of 'success' in each trial
.                  |  .                           | `mu`: Useful for alternative parameterisation


## Continuous distributions


Distribution       | Probability function | Distribution function | Quantile function     | Random numbers
-------------------|----------------------|-----------------------|-----------------------|-----------------
Discrete uniform   | `sample()`           | ~                     |   ~                   | ~
Binomial           | `dunif()`            | `punif()`             | `qunif()`             | `runif()`
Poisson            | `dunif()`            | `punif()`             | `qunif()`             | `runif()`
Geometric          | `dgeom()`            | `pgeom()`             | `qunif()`             | `runif()`
Negative binomial  | `dunif()`            | `punif()`             | `qunif()`             | `runif()`
Continuous uniform | `dunif()`            | `punif()`             | `qunif()`             | `runif()`
Normal|            | `dunif()`            | `punif()`             | `qunif()`             | `runif()`
Gamma|             | `dunif()`            | `punif()`             | `qunif()`             | `runif()`
$t$                | `dunif()`            | `punif()`             | `qunif()`             | `runif()`
$\chi$-squared     | `dunif()`            | `punif()`             | `qunif()`             | `runif()`
$F$                | `dunif()`            | `punif()`             | `qunif()`             | `runif()`



## Multivariate distributions 

ADD??




Distribution       | Probability function | Distribution function | Quantile function     | Random numbers
-------------------|----------------------|-----------------------|-----------------------|-----------------
Discrete uniform   | `sample()`           | ~                     |   ~                   | ~
Binomial           | `dunif()`            | `punif()`             | `qunif()`             | `runif()`
Poisson            | `dunif()`            | `punif()`             | `qunif()`             | `runif()`
Geometric          | `dgeom()`            | `pgeom()`             | `qunif()`             | `runif()`
Negative binomial  | `dunif()`            | `punif()`             | `qunif()`             | `runif()`
Continuous uniform | `dunif()`            | `punif()`             | `qunif()`             | `runif()`
Normal|            | `dunif()`            | `punif()`             | `qunif()`             | `runif()`
Gamma|             | `dunif()`            | `punif()`             | `qunif()`             | `runif()`
$t$                | `dunif()`            | `punif()`             | `qunif()`             | `runif()`
$\chi$-squared     | `dunif()`            | `punif()`             | `qunif()`             | `runif()`
$F$                | `dunif()`            | `punif()`             | `qunif()`             | `runif()`



Distribution       | **R**  functions (with common parameter)
-------------------|-----------------------------------------------------------------------------
Discrete uniform   | Random sample: `sample()`  
Binomial           | `dbinom(x, size, prob)`   
.                   | `pbinom(q, size, prob)`  
.                   | `qbinom(p, size, prob)`  
.                   | `rbinom(n, size, prob)`
Poisson            | `dpois(x, lambda)`   
.                   | `ppois(q, lambda)`   
.                   | `qpois(p, lambda)`   
.                   | `rpois(n, lambda)`
Geometric          | `dgeom(x, prob)`   
.                   | `pgeom(q, prob)`   
.                   | `qgeom(p, prob)`   
.                   | `rgeom(n, prob)`
Negative binomial  | `dnbinom(x, size, prob, mu)` 
.                   | `pnbinom(q, size, prob, mu)` 
.                   | `qdbinom(p, size, prob, mu)` 
.                   | `rnbinom(n, size, prob, mu)`
                   

## Standard univariate continuous distributions
                 
                   
Distribution       | **R**  functions (with common parameter)
-------------------|-----------------------------------------------------------------------------
Continuous uniform | `dunif(x, min = 0, max = 1)`   
.                   | `punif(q, min = 0, max = 1)`   
.                   | `qunif(p, min = 0, max = 1)`   
.                   | `runif(r, min = 0, max = 1)`
Normal             | `dnorm(x, mean = 0, sd = 1)`   
.                   | `pnorm(q, mean = 0, sd = 1)`   
.                   | `qnorm(p, mean = 0, sd = 1)`   
.                   | `rnorm(n, mean = 0, sd = 1)`
Gamma              | `dgamma(x, shape, rate = 1, scale = 1 / rate)`  
.                   | `pgamma(q, shape, rate = 1, scale = 1 / rate)`  
.                   | `qgamma(p, shape, rate = 1, scale = 1 / rate)`  
.                   | `rgamma(n, shape, rate = 1, scale = 1 / rate)`
$t$                | `dt(x, df)`      
.                   | `pt(q, df)`      
.                   | `qt(p, df)`      
.                   | `rt(n, df)`
$\chi$-squared     | `dchisq(x, df)`  
.                   | `pchisq(q, df)`  
.                   | `qchisq(p, df)`  
.                   | `rchisq(n, df)`
$F$                | `df(x, df1, df2)`      
.                   | `pf(q, df1, df2)`      
.                   | `qf(p, df1, df2)`      
.                   | `rf(n, df1, df2)`



## Standard bivariate distributions

Add `mvtnorm::dmvnor()`???



# Solutions

## Answers for Chap. 1

:::{.answer}
Answer to Exercise \@ref(exr:BasicProbs)

1. $0.11$
2. $0.24$
3. $0.32$
4. $0.11/0.35$
5. $\Pr(A) \times \Pr(B) = 0.44\times 0.35 = 0.154 \ne \Pr(A\cap B) = 0.11$; not independent.
:::





:::{.answer}
**Answer to Exercise \@ref(exr:DrawNumbers).**

1. $\displaystyle{\frac{188}{3201}}$
1. $\displaystyle{\frac{1225}{3201}}$
1. $\Pr(\text{2 odd before 1st even}) = \Pr(\text{odd, odd, even, odd or even}) + \Pr(\text{odd, odd, odd, even)} =\displaystyle{\frac{3625}{19206}}$.
1. $\Pr(\text{sum odd}) = \Pr(\text{1 odd number drawn, OR 3 odd numbers drawn}) = \displaystyle{\frac{1600}{3201}}$
:::




:::{.answer}
Answer to Exercise \@ref(exr:Quadratic).

* $S = \{ (a,b,c) \mid a\in(-\infty, 0)\cap (0, \infty), -\infty < b < \infty, -\infty < c < \infty\}$.
* The solutions to a quadratic equation are given by
\[
   x = \frac{-b \pm \sqrt{b^2 - 4ac} }{2a}.
\]
For two equal roots, $b^2 - 4ac = 0$, so event $A$ is defined as $A = \{ (a,b,c) \mid b^2 - 4ac = 0\}$.
:::


:::{.answer}
Answer to Exercise \@ref(exr:HatData).

Tree diagram: Fig, \@ref(fig:HatTree); table: Table \@ref(tab:HatTable); Venn diagram: Fig. \@ref(fig:HatVenn).
:::


```{r HatTree, echo=FALSE, fig.align="center", fig.cap="Tree diagram for the hat-wearing example", fig.width=4, fig.height=4, out.width='70%'}

plot( x = c(0, 1),
      y = c(0, 1),
      type = "n",
      main = "Tree diagram for the hat data",
      xlab = "",
      ylab = "",
      axes = FALSE)
#STEP 1
lines( x = c(0.2, 0.40),
       y = c(0.5, 0.75),
       lwd = 2)
lines( x = c(0.2, 0.40),
       y = c(0.5, 0.25),
       lwd = 2)
# STEP 2
lines( x = c(0.60, 0.80),
       y = c(0.75, 0.90),
       lwd = 2)
lines( x = c(0.60, 0.80),
       y = c(0.75, 0.60),
       lwd = 2)

lines( x = c(0.60, 0.80),
       y = c(0.25, 0.40),
       lwd = 2)
lines( x = c(0.60, 0.80),
       y = c(0.25, 0.10),
       lwd = 2)

# TEXT
text(x = 0.50,
     y = 0.75,
     labels = "Males")
text(x = 0.50,
     y = 0.25,
     labels = "Females")

text(x = 0.90,
     y = 0.90,
     labels = "Hat")
text(x = 0.90,
     y = 0.60,
     labels = "No hat")

text(x = 0.90,
     y = 0.40,
     labels = "Hat")
text(x = 0.90,
     y = 0.10,
     labels = "No hat")

# ADD PROBS
text(x = 0.38,
     y = 0.63,
     cex = 0.9,
     labels = "0.513")
text(x = 0.38,
     y = 0.37,
     cex = 0.9,
     labels = "0.487")

text(x = 0.65,
     y = 0.84,
     cex = 0.9,
     labels = "0.205")
text(x = 0.65,
     y = 0.66,
     cex = 0.9,
     labels = "0.795")

text(x = 0.65,
     y = 0.34,
     cex = 0.9,
     labels = "0.060")
text(x = 0.65,
     y = 0.16,
     cex = 0.9,
     labels = "0.940")

## STEPS
text(x = 0.50,
     y = 0.01,
     label = expression(bold("Step 1"))
     )
text(x = 0.90,
     y = 0.01,
     label = expression(bold("Step 2"))
     )
```


```{r HatTable, echo=FALSE}
BRIS <- read.csv("Data/sunglasses.csv")
BRIS$Gender <- factor(BRIS$Gender,
                      levels = c(1, 2),
                      labels = c("Males", 
                                 "Females"))
BRIS$Hat <- factor(BRIS$Hat,
                   levels = c(0, 1),
                   labels = c("Not Hat",
                              "Hat"))

HatTable <- xtabs( Count ~ Gender + Hat,
                   data = BRIS)
HatTable <- cbind( HatTable,
                   "Total" = rowSums(HatTable) )
knitr::kable(HatTable,
             caption = "The numbers of males and females wearing a hat in the middle of the day in Brisbane")
```



```{r HatVenn, echo=FALSE, fig.width=7, fig.height=4, out.width='70%', fig.align="center", fig.cap="The hat-wearing data as a Venn diagram"}
par( mfrow = c(2, 2))
colourMale <- rgb(0, 0, 255, 
                  max = 255, 
                  alpha = 125)
colourHat  <- rgb(0, 255, 0, 
                  max = 255, 
                  alpha = 125)

plot( x = c(0, 1),
      y = c(0, 1),
      type = "n",
      ylab = "",
      xlab = "",
      main = expression(paste("Events ", italic(M), " and ", bar(italic(M)))),
      axes = FALSE)

polygon( x = c(0, 0, 1, 1),
         y = c(0, 1, 1, 0))
plotrix::draw.circle(x = 0.4, 
                     y = 0.5,
                     radius = 0.2,
                     col = colourMale)
text(x = 0.4,
     y = 0.5,
     labels = expression(paste(italic(M), ": 307")))
text(x = 0.8,
     y = 0.5,
     labels = expression(paste(bar(italic(M)), ": 366")))
mtext(text = expression(italic(S)),
      side = 1,
      adj = 0.25)


###############


plot( x = c(0, 1),
      y = c(0, 1),
      type = "n",
      ylab = "",
      xlab = "",
      main = expression(paste("Events ", italic(H), " and ", bar(italic(H)))) ,
      axes = FALSE)


polygon( x = c(0, 0, 1, 1),
         y = c(0, 1, 1, 0))

plotrix::draw.circle(x = 0.6,
                     y = 0.5,
                     radius = 0.2,
                     col = colourHat)

text(x = 0.6,
     y = 0.5,
     labels = expression(paste(italic(H), ": 101")))
text(x = 0.2,
     y = 0.5,
     labels = expression(paste( bar(italic(H)), ": 651")))

mtext(text = expression(italic(S)),
      side = 1,
      adj = 0.25)


####################


plot( x = c(0, 1),
      y = c(0, 1),
      type = "n",
      ylab = "",
      xlab = "",
      main = expression(paste("Events ", italic(M), " and ", italic(H))) ,
      axes = FALSE)


polygon( x = c(0, 0, 1, 1),
         y = c(0, 1, 1, 0))
plotrix::draw.circle(x = 0.4, 
                     y = 0.5,
                     radius = 0.2,
                     col = colourMale)
plotrix::draw.circle(x = 0.6,
                     y = 0.5,
                     radius = 0.2,
                     col = colourHat)
text(x = 0.3,
     y = 0.5,
     labels = expression(italic(M)))
text(x = 0.7,
     y = 0.5,
     labels = expression(italic(H)))
mtext(text = expression(italic(S)),
      side = 1,
      adj = 0.25)


#################



plot( x = c(0, 1),
      y = c(0, 1),
      type = "n",
      ylab = "",
      xlab = "",
      main = "The numbers in each section",
      axes = FALSE)


polygon( x = c(0, 0, 1, 1),
         y = c(0, 1, 1, 0))
plotrix::draw.circle(x = 0.4, 
                     y = 0.5,
                     radius = 0.2,
                     col = colourMale)
plotrix::draw.circle(x = 0.6,
                     y = 0.5,
                     radius = 0.2,
                     col = colourHat)
text(x = 0.3,
     y = 0.5,
     labels = "307")
text(x = 0.7,
     y = 0.5,
     labels = "22")
text(x = 0.7,
     y = 0.85,
     labels = expression(bar(italic(M)) ~ intersect() ~ bar(italic(H)) ~ ": 344") )
mtext(text = expression(italic(S)),
      side = 1,
      adj = 0.25)

# Intersection
text(x = 0.5, 
     y = 0.5, 
     label = "79")

```



:::{.answer}
**Answer to Exercise \@ref(exr:MatchingBrackets).**

Many document processors help users match brackets.
Bracket matching is an interesting mathematical problem!
For instance, the string `(())` is syntactically valid, whereas `())(` is not, even though both contain two opening and two closing brackets.


1. `()()` and `(())`. 
   There are two ways.
1. `()()()` and `(())()` and `()(())` and `((()))` and `(()())`. 
   There are five ways.
1. $\displaystyle \frac{1}{n + 1} {2n \choose n} = \frac{1}{n + 1}\frac{(2n)!}{n!n!}  = \frac{1}{(n + 1)!} \frac{(2n)!}{n!}$ as to be shown.
1. ANSWER: Show that another expression for $C_n$ is $\displaystyle {2n \choose n} - {2n \choose n + 1}$ for $n\geq 0$.
1. The first nine Catalan numbers, for $n = 0, \dots 8$, are $1, 1, 2, 5, 14, 42, 132, 429, 1430$
:::



:::{.answer}
**Answer to Exercise \@ref(exr:CompareMeansStress).**

????
:::



:::{.answer}
**Answer to Exercise \@ref(exr:GreenLights).**

1. The length of time (in seconds) between green lights at the intersection, say $G$.
2. $S = \{G \mid 15 \le G \le 150\}$.
3. No---no equally likely events are defined
4. The probability can be *approximated*---observe the lights many times, and count how often there is less than 90 seconds between green
lights.
:::




:::{.answer}
**Answer to Exercise \@ref(exr:WorldCup).**

1. $C^7_5 \times C5_4 \times C^2_1 \times C^1_1$; $11^2 + (2\times 22) = 1 65$.
2. ???
:::





:::{.answer}
**Answer to Exercise \@ref(exr:Stirling).**

```{r}
# Define a function to compute Stirling numbers:
stirling <- function(n){ 
  sqrt(2 * pi *( n)) * (n/exp(1))^n
}

n <- 1:10
Actual <- factorial(n) 
Approx <- stirling(n)
RelError <- (Actual - Approx)/Actual * 100

cbind(Actual, Approx, RelError)

plot(RelError ~ n,
     ylim = c(0, 8),
     las = 1, 
     type = "b",
     pch = 19,
     lwd = 2,
     xlab = expression(italic(n)),
     ylab = "Relative error (%)",
     main = "Relative error of\nStirling's approximation")
```
:::


## Answers for Chap. 2



## Answers for Chap. 3



:::{.answer}
Answer to Exercise \@ref(exr:C3DiscreteA).

1. $\alpha = 1/2$.
1. $\text{E}(M) = 13/6$, $\text{var}(N) \approx 0.8056$.
1. $M_M(t) = \exp(t)/3 + \exp(2t)/6 + \exp(3t)/2$.
:::


:::{.answer}
Answer to Exercise \@ref(exr:C3DiscreteB).

1. $M_Z'(t) = 0.6\exp(t)[0.3\exp(t) + 0.7]$ so $\text{E}(Z) = 0.6$.
   Also, $M''_Z(t) = 0.18\exp(2t) + 0.6\exp(t)[0.3\exp(t) + 0.7]$ so $\text{E}(Z^2) = 0.78$, hence so $\text{var}(Z) = 0.42$ (be careful with the derivatives here!); expand the quadratic and find $\Pr(Z = 0) = 0.49$, $\Pr(Z = 1) = 0.42$,
$\Pr(Z = 2) = 0.09$.
:::


:::{.answer}
Answer to Exercise \@ref(exr:C3MGFA).

1. $M'_G(t) = \alpha\beta(1 - \beta t)^{-\alpha - 1}$ so $\text{E}(G) = \alpha\beta$.
   $M''_G(t) = \alpha\beta^2(\alpha + 1)(1 - \beta t)^{-\alpha - 2}$ so $\text{E}(G^2) = \alpha\beta^2(\alpha + 1)$ and $\text{var}(G) = \alpha\beta^2$.
:::


:::{.answer}
Answer to Exercise \@ref(exr:C3Moments).

1. $\mu'_r = \text{E}(X^r) = \int_{x = 0}^1 x^r 2(1 - x)\, dx = 2\left[ \left(\frac{x^{r + 1}}{r + 1} - \frac{x^{r + 2}}{r + 2}\right)\Big|_{0}^{1}\right] = 2\left[ \frac{1}{r + 1} - \frac{1}{r + 2}\right]$.
1. Expanding, $\text{E}((X + 3)^2) = \text{E}(X^2) + 6\text{E}(X) + 9$.
   Now, $\text{E}(X) = \mu'_1 = 1/3$ from above, and $\text{E}(X^2) = \mu'_2 = 1/6$ from above.
   Hence $\text{E}((X + 3)^2) = 67/6$.
1. $\text{var}(X) = \text{E}(X^2) - \text{E}(X)^2 = 1/18$.
:::






:::{.answer}
Answer to Exercise \@ref(exr:C3Cauchy).

1. See Fig. \@ref(fig:Cauchy) (left panel).
1. Evaluate $F_X(t) = \displaystyle \int_{-\infty}^t \frac{1}{\pi(1 + x^2)}\,dx = \frac{1}{\pi}\arctan(x) + \frac{1}{2}$.
   See Fig. \@ref(fig:Cauchy) (right panel).
1. $\displaystyle \text{E}(X) =  \int_{-\infty}^t \frac{x}{\pi(1 + x^2)}\,dx = ...$, which is not defined.
:::

```{r Cauchy, fig.align="center", fig.cap="The Cauchy distribution", fig.height=4, fig.width=7,out.width='70%'}
par( mfrow = c(1, 2))
x <- seq(-5, 5,
         length = 100)
y <- 1 / (pi * (1 + x^2))
plot(y ~ x,
     lwd = 2,
     las = 1,
     type = "l",
     main = "Density function for\nthe Cauchy distribution",
     xlab = "X",
     ylab = "Density function")

Y <- atan(x)/pi + 0.5
plot(Y ~ x,
     lwd = 2,
     las = 1,
     ylim = c(0, 1),
     type = "l",
     main = "Distribution function for\nthe Cauchy distribution",
     xlab = "X",
     ylab = "Distribution function")
```



:::{.answer}
Answer to Exercise \@ref(exr:C3MGFSymmetric).

Begin with Definition \@ref(def:MGF) for $M_X(t)$ and use fact that if a distribution is symmetric about 0 then $f_X(x) = f_X(-x)$ using symmetry. 
Transform the resulting integral.
:::






## Answers for Chap. 4



## Answers for Chap. 5



## Answers for Chap. 6



## Answers for Chap. 7



## Answers for Chap. 8

