## Answers for Chap. 1

:::{.answer}
**Answer to Exercise \@ref(exr:BasicProbs).**

1. Probably the Venn diagram is best.
1. $\Pr(A\cup B) = \Pr(A) + \Pr(B) = \Pr(A\cap B) = 0.66$.
1. $0.13$.
1. $0.89$.
1. $0.11/0.24 \approx 0.4583...$
1. $\Pr(A) \times \Pr(B) = 0.53\times 0.24 = 0.1272 \ne \Pr(A\cap B) = 0.11$; not independent... but close!
:::





:::{.answer}
**Answer to Exercise \@ref(exr:DrawNumbers).**

1. $(50/100)\times (49/99)\times (48/98)\times (47/97) = ^{50}C_4 / ^{100}C_4\approx 0.0587$.
1. $^{50}C_2\times ^{50}C_2 / ^{100}C_4  = 1225/3201 \approx 0.3826$.
1. $\Pr(\text{at least 2 odd before 1st even}) = \Pr(\text{odd, odd, even, either}) + \Pr(\text{odd, odd, odd, even)} = (50/100)\times(49/99)\times(50/98) + (50/100)\times (49/99)\times(48/98)\times(50/97) \approx 0.1887$.
1. $\Pr(\text{sum odd}) = \Pr(\text{exactly 1 odd number drawn, OR exactly 3 odd numbers drawn}) = \displaystyle{\frac{1600}{3201}}$.
:::


:::{.answer}
**Answer to Exercise \@ref(exr:Quadratic).**

1. $S = \{ (a,b,c) \mid a\in(-\infty, 0)\cap (0, \infty), -\infty < b < \infty, -\infty < c < \infty\}$, which can be written $X \in \{(a, b, c) | (a, b, c) \in \mathbb{R}, a \ne 0\}$ where $\mathbb{R}$ represents the real numbers.
1. The solutions to a quadratic equation are given by  
\[
   x = \frac{-b \pm \sqrt{b^2 - 4ac} }{2a}.
\]
For two equal roots, $b^2 - 4ac = 0$, so event $R$ is defined as $R = \{ (a,b,c) \mid (a, b, c)\in \mathbb{R}, a\ne 0, b^2 - 4ac = 0\}$.
1. For no real roots, $b^2 - 4ac < 0$, so event $Z$ is defined as $Z = \{ (a,b,c) \mid (a, b, c)\in \mathbb{R}, a\ne 0, b^2 - 4ac < 0\}$.
:::



<!-- :::{.answer} -->
<!-- **Answer to Exercise \@ref(exr:GreenLights).** -->

<!-- 1. The length of time (in seconds) between green lights at the intersection, say $G$. -->
<!-- 2. $S = \{G \mid 15 \le G \le 150\}$. -->
<!-- 3. No---no equally likely events are defined. -->
<!-- 4. The probability can be *approximated*---observe the lights many times, and count how often there is less than 90 seconds between green -->
<!-- lights. -->
<!-- ::: -->



:::{.answer}
**Answer to Exercise \@ref(exr:Cricket).**

1. $C^7_5 \times C^5_4 \times C^2_1 \times C^1_1 = 210$.
2. $11^2 + (2\times 22) = 165$.
:::




<!-- :::{.answer} -->
<!-- **Answer to Exercise \@ref(exr:HatData).** -->

<!-- Tree diagram: Fig, \@ref(fig:HatTree); table: Table \@ref(tab:HatTable); Venn diagram: Fig. \@ref(fig:HatVenn). -->
<!-- ::: -->


<!-- ```{r HatTree, echo=FALSE, fig.align="center", fig.cap="Tree diagram for the hat-wearing example", fig.width=6.5, fig.height=4, out.width='70%'} -->

<!-- plot( x = c(0, 1), -->
<!--       y = c(0, 1), -->
<!--       type = "n", -->
<!--       main = "Tree diagram for the hat data", -->
<!--       xlab = "", -->
<!--       ylab = "", -->
<!--       axes = FALSE) -->
<!-- #STEP 1 -->
<!-- lines( x = c(0.2, 0.40), -->
<!--        y = c(0.5, 0.75), -->
<!--        lwd = 2) -->
<!-- lines( x = c(0.2, 0.40), -->
<!--        y = c(0.5, 0.25), -->
<!--        lwd = 2) -->
<!-- # STEP 2 -->
<!-- lines( x = c(0.60, 0.80), -->
<!--        y = c(0.75, 0.90), -->
<!--        lwd = 2) -->
<!-- lines( x = c(0.60, 0.80), -->
<!--        y = c(0.75, 0.60), -->
<!--        lwd = 2) -->

<!-- lines( x = c(0.60, 0.80), -->
<!--        y = c(0.25, 0.40), -->
<!--        lwd = 2) -->
<!-- lines( x = c(0.60, 0.80), -->
<!--        y = c(0.25, 0.10), -->
<!--        lwd = 2) -->

<!-- # TEXT -->
<!-- text(x = 0.50, -->
<!--      y = 0.75, -->
<!--      labels = "Males") -->
<!-- text(x = 0.50, -->
<!--      y = 0.25, -->
<!--      labels = "Females") -->

<!-- text(x = 0.90, -->
<!--      y = 0.90, -->
<!--      labels = "Hat") -->
<!-- text(x = 0.90, -->
<!--      y = 0.60, -->
<!--      labels = "No hat") -->

<!-- text(x = 0.90, -->
<!--      y = 0.40, -->
<!--      labels = "Hat") -->
<!-- text(x = 0.90, -->
<!--      y = 0.10, -->
<!--      labels = "No hat") -->

<!-- # ADD PROBS -->
<!-- text(x = 0.38, -->
<!--      y = 0.63, -->
<!--      cex = 0.9, -->
<!--      labels = "0.513") -->
<!-- text(x = 0.38, -->
<!--      y = 0.37, -->
<!--      cex = 0.9, -->
<!--      labels = "0.487") -->

<!-- text(x = 0.65, -->
<!--      y = 0.85, -->
<!--      cex = 0.9, -->
<!--      labels = "0.205") -->
<!-- text(x = 0.65, -->
<!--      y = 0.65, -->
<!--      cex = 0.9, -->
<!--      labels = "0.795") -->

<!-- text(x = 0.65, -->
<!--      y = 0.345, -->
<!--      cex = 0.9, -->
<!--      labels = "0.060") -->
<!-- text(x = 0.65, -->
<!--      y = 0.15, -->
<!--      cex = 0.9, -->
<!--      labels = "0.940") -->

<!-- ## STEPS -->
<!-- text(x = 0.50, -->
<!--      y = 0.01, -->
<!--      label = expression(bold("Step 1")) -->
<!--      ) -->
<!-- text(x = 0.90, -->
<!--      y = 0.01, -->
<!--      label = expression(bold("Step 2")) -->
<!--      ) -->
<!-- ``` -->


<!-- ```{r HatTable, echo=FALSE} -->
<!-- BRIS <- read.csv("Data/sunglasses.csv") -->
<!-- BRIS$Gender <- factor(BRIS$Gender, -->
<!--                       levels = c(1, 2), -->
<!--                       labels = c("Males",  -->
<!--                                  "Females")) -->
<!-- BRIS$Hat <- factor(BRIS$Hat, -->
<!--                    levels = c(0, 1), -->
<!--                    labels = c("No Hat", -->
<!--                               "Hat")) -->

<!-- HatTable <- xtabs( Count ~ Gender + Hat, -->
<!--                    data = BRIS) -->
<!-- HatTable <- cbind( HatTable, -->
<!--                    "Total" = rowSums(HatTable) ) -->

<!-- if( knitr::is_latex_output() ) { -->
<!--   knitr::kable(HatTable, -->
<!--                format = "latex", -->
<!--                booktabs = TRUE, -->
<!--                align = "r", -->
<!--                longtable = FALSE,                 -->
<!--                escape = FALSE, -->
<!--                caption = "The numbers of males and females wearing a hat in the middle of the day in Brisbane") %>% -->
<!--     kable_styling(font_size = 10) %>% -->
<!--     row_spec(0, bold = TRUE)  -->
<!-- } -->
<!-- if( knitr::is_html_output() ) { -->
<!--   knitr::kable(HatTable, -->
<!--                escape = TRUE, -->
<!--                format = "html", -->
<!--                booktabs = TRUE, -->
<!--                align = "r", -->
<!--                longtable = FALSE,                 -->
<!--                caption = "The numbers of males and females wearing a hat in the middle of the day in Brisbane") %>% -->
<!--     row_spec(1, bold = TRUE)  -->
<!-- } -->
<!-- ``` -->



<!-- ```{r HatVenn, echo=FALSE, fig.width=7, fig.height=7, out.width='70%', fig.align="center", fig.cap="The hat-wearing data as a Venn diagram"} -->
<!-- par( mfrow = c(2, 2)) -->
<!-- colourMale <- rgb(0, 0, 255,  -->
<!--                   max = 255,  -->
<!--                   alpha = 125) -->
<!-- colourHat  <- rgb(0, 255, 0,  -->
<!--                   max = 255,  -->
<!--                   alpha = 125) -->

<!-- plot( x = c(0, 1), -->
<!--       y = c(0, 1), -->
<!--       type = "n", -->
<!--       ylab = "", -->
<!--       xlab = "", -->
<!--       main = expression(paste("Events ", italic(M), " and ", bar(italic(M)))), -->
<!--       axes = FALSE) -->

<!-- polygon( x = c(0, 0, 1, 1), -->
<!--          y = c(0, 1, 1, 0)) -->
<!-- plotrix::draw.circle(x = 0.4,  -->
<!--                      y = 0.5, -->
<!--                      radius = 0.2, -->
<!--                      col = colourMale) -->
<!-- text(x = 0.4, -->
<!--      y = 0.5, -->
<!--      labels = expression(paste(italic(M), ": 307"))) -->
<!-- text(x = 0.8, -->
<!--      y = 0.5, -->
<!--      labels = expression(paste(bar(italic(M)), ": 366"))) -->
<!-- mtext(text = expression(italic(S)), -->
<!--       side = 1, -->
<!--       adj = 0.25) -->


<!-- ############### -->


<!-- plot( x = c(0, 1), -->
<!--       y = c(0, 1), -->
<!--       type = "n", -->
<!--       ylab = "", -->
<!--       xlab = "", -->
<!--       main = expression(paste("Events ", italic(H), " and ", bar(italic(H)))) , -->
<!--       axes = FALSE) -->


<!-- polygon( x = c(0, 0, 1, 1), -->
<!--          y = c(0, 1, 1, 0)) -->

<!-- plotrix::draw.circle(x = 0.6, -->
<!--                      y = 0.5, -->
<!--                      radius = 0.2, -->
<!--                      col = colourHat) -->

<!-- text(x = 0.6, -->
<!--      y = 0.5, -->
<!--      labels = expression(paste(italic(H), ": 101"))) -->
<!-- text(x = 0.2, -->
<!--      y = 0.5, -->
<!--      labels = expression(paste( bar(italic(H)), ": 651"))) -->

<!-- mtext(text = expression(italic(S)), -->
<!--       side = 1, -->
<!--       adj = 0.25) -->


<!-- #################### -->


<!-- plot( x = c(0, 1), -->
<!--       y = c(0, 1), -->
<!--       type = "n", -->
<!--       ylab = "", -->
<!--       xlab = "", -->
<!--       main = expression(paste("Events ", italic(M), " and ", italic(H))) , -->
<!--       axes = FALSE) -->


<!-- polygon( x = c(0, 0, 1, 1), -->
<!--          y = c(0, 1, 1, 0)) -->
<!-- plotrix::draw.circle(x = 0.4,  -->
<!--                      y = 0.5, -->
<!--                      radius = 0.2, -->
<!--                      col = colourMale) -->
<!-- plotrix::draw.circle(x = 0.6, -->
<!--                      y = 0.5, -->
<!--                      radius = 0.2, -->
<!--                      col = colourHat) -->
<!-- text(x = 0.3, -->
<!--      y = 0.5, -->
<!--      labels = expression(italic(M))) -->
<!-- text(x = 0.7, -->
<!--      y = 0.5, -->
<!--      labels = expression(italic(H))) -->
<!-- mtext(text = expression(italic(S)), -->
<!--       side = 1, -->
<!--       adj = 0.25) -->


<!-- ################# -->



<!-- plot( x = c(0, 1), -->
<!--       y = c(0, 1), -->
<!--       type = "n", -->
<!--       ylab = "", -->
<!--       xlab = "", -->
<!--       main = "The numbers in each section", -->
<!--       axes = FALSE) -->


<!-- polygon( x = c(0, 0, 1, 1), -->
<!--          y = c(0, 1, 1, 0)) -->
<!-- plotrix::draw.circle(x = 0.4,  -->
<!--                      y = 0.5, -->
<!--                      radius = 0.2, -->
<!--                      col = colourMale) -->
<!-- plotrix::draw.circle(x = 0.6, -->
<!--                      y = 0.5, -->
<!--                      radius = 0.2, -->
<!--                      col = colourHat) -->
<!-- text(x = 0.3, -->
<!--      y = 0.5, -->
<!--      labels = "307") -->
<!-- text(x = 0.7, -->
<!--      y = 0.5, -->
<!--      labels = "22") -->
<!-- text(x = 0.7, -->
<!--      y = 0.85, -->
<!--      labels = expression(bar(italic(M)) ~ intersect() ~ bar(italic(H)) ~ ": 344") ) -->
<!-- mtext(text = expression(italic(S)), -->
<!--       side = 1, -->
<!--       adj = 0.25) -->

<!-- # Intersection -->
<!-- text(x = 0.5,  -->
<!--      y = 0.5,  -->
<!--      label = "79") -->

<!-- ``` -->



<!-- :::{.answer} -->
<!-- **Answer to Exercise \@ref(exr:CarPassengers).** -->
<!-- 1. Choose one driver, and the other seven passengers can sit anywhere: -->
<!--    $2!\times 7! = 10080$. -->
<!-- 1. Choose one driver, and the other seven passengers can sit anywhere: -->
<!--    $3!\times 7! = 30240$. -->
<!-- 1. Choose one driver, choose who sits in what car sets, and the other five passengers can sit anywhere: -->
<!--    $2!\times 2! \times 5! = 480$. -->
<!-- ::: -->



<!-- :::{.answer} -->
<!-- **Answer to Exercise \@ref(exr:MatchingBrackets).** -->

<!-- 1. `()()` and `(())`.  -->
<!--    There are two ways. -->
<!-- 1. `()()()` and `(())()` and `()(())` and `((()))` and `(()())`.  -->
<!--    There are five ways. -->
<!-- 1. $\displaystyle \frac{1}{n + 1} \binom{2n}{n} = \frac{1}{n + 1}\frac{(2n)!}{n!n!}  = \frac{1}{(n + 1)!} \frac{(2n)!}{n!}$ as to be shown. -->
<!-- 1. Write as $\displaystyle \frac{(2n)!}{n!\, n!} - \frac{(2n)!}{(n + 1)! (2n - n - 1)!}$. -->
<!--    Simplifying:  -->

<!-- \begin{align*} -->
<!--      \frac{(2n)!}{n!\, n!} - \frac{(2n)!}{(n + 1)! (n - 1)!} -->
<!--    &= \frac{(2n)!}{n!\, n!} - \left( \frac{n}{n + 1}\right) \frac{(2n)!}{n!\,n!}\\ -->
<!--    &= \binom{2n}{n}\left(1 - \frac{n}{n + 1}\right) \\ -->
<!--    &= \frac{1}{n + 1} \binom{2n}{n} -->
<!-- \end{align*} -->
<!-- 1. The first nine Catalan numbers, for $n = 0, \dots 8$, are $1, 1, 2, 5, 14, 42, 132, 429, 1430$ -->
<!-- ::: -->




:::{.answer}
**Answer to Exercise \@ref(exr:Stirling).**
See Fig.\ \@ref(fig:Stirling).

```{r Stirling, fig.cap="Relative error of Stirling's approximation", fig.align="center", out.width="70%"}
# Define a function to compute Stirling numbers:
stirling <- function(n){ 
  sqrt(2 * pi *n) * (n/exp(1))^n
}

n <- 1:10
Actual <- factorial(n) 
Approx <- stirling(n)
RelError <- (Actual - Approx)/Actual * 100

cbind(Actual, Approx, RelError)

plot(RelError ~ n,
     ylim = c(0, 8),
     las = 1, 
     type = "b",
     pch = 19,
     lwd = 2,
     xlab = expression(italic(n)),
     ylab = "Relative error (%)",
     main = "Relative error of\nStirling's approximation")
```
:::




:::{.answer}
**Answer to Exercise \@ref(exr:DiceGame).**

1. $\Pr(\text{Player throwing first wins})$ means $\Pr(\text{First six on throw 1 or 3 or 5 or ...})$.
   So: $\Pr(\text{First six on throw 1}) + \Pr(\text{First six on throw 3}) + \cdots$.
   This produces a geometric progression that can be summed obtained (see App.\ \@ref(UsefulSeries)).
2. Use Theorem\ \@ref(thm:TotalProb).
   Define the events $A = \text{Player 1 wins}$, $B_1 = \text{Player 1 throws first}$, and $B_2 = \text{Player 1 throws second}$.
:::


<!-- :::{.answer} -->
<!-- **Answer to Exercise \@ref(exr:RandomisedResponse).** -->

<!-- Write $y = \Pr(\text{Answers `yes'})$. -->

<!-- 1. Then, from a tree diagram (Step 1: Used drugs/did not use drugs; Step 2: Card says "Used drugs"/Card says "Did not use drugs"): -->

<!-- \begin{align*} -->
<!--    y  -->
<!--    &= \Pr(\text{Never takes drugs and say so}) + \Pr(\text{Takes drugs and says so})\\ -->
<!--    &= (1 - p) \times\left(\frac{N - m}{N}\right) + p\times\frac{m}{N}. -->
<!-- \end{align*} -->
<!-- Solving for the unknown $p$: -->

<!-- \[ -->
<!--   p = \frac{yN - N + m}{2m - N}. -->
<!-- \] -->
<!-- 2. When $m = 0$, $\Pr(Y) = p$: This mean every card says 'I *have* used an illegal drug in the past twelve months', so the proportion that have used an illegal drug is just the same as the proportion responding with 'Yes' (and there is no anonymity). -->
<!--    When $m = 0$, $\Pr(Y) = 1 - p$: This mean every card says 'I *have not* used an illegal drug in the past twelve months', so the proportion that have used an illegal drug is just the same as the proportion responding with 'No' (and there is no anonymity). -->
<!--    When $m = N/2$, $\Pr(Y) = 0.5$; we have learnt nothing: The probability is 50--50.   -->
<!-- 3. $\displaystyle d = \frac{yN - N + m}{2m - N}$, so plugging in $N = 100$, $m = 25$ and $\Pr(Y) = 175/400 = 0.4375$ gives $d = 0.625$. -->
<!-- ::: -->



:::{.answer}
**Answer to Exercise \@ref(exr:MCExam).**

Use Theorem \@ref(thm:TotalProb) to find $\Pr(C)$ where $C = \text{select correct answer}$, $K = \text{student knows answer}$. 
Then, $\Pr(C)  =\displaystyle {\frac{mp + q}{m}}$
:::



:::{.answer}
**Answer to Exercise \@ref(exr:EPL).**
???
:::



<!-- :::{.answer} -->
<!-- **Answer to Exercise \@ref(exr:TwoPointsInSquare).** -->
<!-- Write $d$ as the distance; then $S = \{d: 0 \le d \le \sqrt{2}\}$. -->
<!-- For the grid, let's use **R** to find what values are possible: -->

<!-- ```{r} -->
<!-- set.seed(8757) # For reproducibility -->
<!-- numSims <- 5000 -->
<!-- distance <- array( dim = numSims) -->

<!-- for (i in 1:numSims) { -->
<!--   xA <- sample( seq(0, 1, by = 0.25), -->
<!--                 size = 1)  -->
<!--   xB <- sample( seq(0, 1, by = 0.25), -->
<!--                 size = 1)  -->
<!--   yA <- sample( seq(0, 1, by = 0.25), -->
<!--                 size = 1)  -->
<!--   yB <- sample( seq(0, 1, by = 0.25), -->
<!--                 size = 1)  -->

<!--   distance[i] <- sqrt( (xB - xA)^2 + (yA - yB) ^ 2) -->
<!-- } -->
<!-- levels( factor( distance)) -->
<!-- ``` -->
<!-- It seems there are 15 possible values for the distance: -->

<!-- * $0$, $0.25$, $0.50$, $0.75$ and $1$ along grid lines; -->
<!-- * $\sqrt{2}/4$, $\sqrt{5}/4$, $\sqrt{10}/4$ and $\sqrt{17}/4$ when the line is moved one grid-square right; -->
<!-- * $\sqrt{13}/4$ and $\sqrt{20}/4$ when the line is moved two grid-squares right; -->
<!-- * $\sqrt{25}/4$ when the line is moved three grid-squares right; -->

<!-- and so on. -->
<!-- ::: -->




:::{.answer}
**Answer to Exercise \@ref(exr:LocalNewsletter).**

1. Anywhere between $0$% and $8$%.
1. $0.06/0.30 = 0.20$.
1. $0.06/0.08 = 0.75$.
:::



:::{.answer}
**Answer to Exercise \@ref(exr:SchoolKids).**
The total number of children: $69\ 279$.
Define $N$ as 'a first-nations student', $F$ as 'a female student', and $G$ as 'attends a government school'.

1. $(2540 + 2734 + 391 + 362) / 69,279 \approx 0.0870$.
1. $49,067/69,279 \approx 0.708$.
1. Many ways to approach this. 
   For **government schools** only: 
   $\Pr(F\mid N) = 2540/(2540 + 2734) \approx 0.4816$.
   $\Pr(\bar{F} \mid N)  = 2734/(2540 + 2734) \approx 0.5184$.
   These are---*very* roughly--about the same, so *approximately* independent.
1. Many ways to approach this. 
   For **non-government schools** only: 
   $\Pr(F\mid N) = 391/(391 + 9496)  \approx 0.0395$.
   $\Pr(\bar{F} \mid N)  = 362/(391 + 9496) \approx 0.0366$.
   These are---*very* roughly--about the same, so *approximately* independent.
1. Many ways to approach this.
   $\Pr(N) =  6027/69,279 \approx 0.0870$.
   $\Pr(N | G) = 5274/49,067 \approx 0.1075$,
   These are not very close, so not really independent.
1. For *female* students:
   $\Pr(N) =  (2540 + 391)/33,646 \approx 0.0871$.
   $\Pr(N | G) = 2540/23,759 \approx 0.1069$.
   These are not very close, so not really independent.
1. For *male* students:
   $\Pr(N) =  (2734 + 362)/35,633 \approx 0.0767$.
   $\Pr(N | G) = 2734/25,308 \approx 0.1080$.
   These are not very close, so not really independent.
1. Based on the above, what can you conclude from the data?
:::




:::{.answer}
**Answer to Exercise \@ref(exr:CardsAces).**
1. The probability depends on what happens with the first card:
   \begin{align*}
     \Pr(\text{Ace second}) 
     &= \Pr(\text{Ace, then Ace}) + |Pr(\text{Non-Ace, then Ace})\\
     &= \left(\frac{4}{52}\times \frac{3}{51}\right) + \left(\frac{48}{52}\times \frac{4}{51}\right) \\
     &= \frac{204}{52\times 51} \approx 0.07843.
   \end{align*}
   You can use a tree diagram, for example.
2. Be careful:

\begin{align*}
  &\Pr(\text{1st card lower rank than second card})\\
  &= \Pr(\text{2nd card a K}) \times \Pr(\text{1st card from Q to Ace}) + {}\\
  &\qquad   \Pr(\text{2nd card a Q}) \times \Pr(\text{1st card from J to Ace}) +{}\\
  &\qquad   \Pr(\text{2nd card a J}) \times \Pr(\text{1st card from 10 to Ace}) + \dots + {}\\
  &\qquad   \Pr(\text{2nd card a 2}) \times \Pr(\text{1st card an Ace}) \\
  &= \frac{4}{51} \times \frac{12\times 4}{52} + {}\\
  &\qquad   \frac{4}{51} \times \frac{11\times 4}{52} +\\
  &\qquad   \frac{4}{51} \times \frac{10\times 4}{52} + \dots +\\
  &\qquad   \frac{4}{51} \times \frac{1\times 4}{52} \\
  &= \frac{4}{51}\frac{4}{52}\left[ 12 + 11 + 10 + \cdots + 1\right]\\
  &= \frac{4}{51}\frac{4}{52} \frac{13\times 12}{2} \approx 0.4705882.
\end{align*}
3. We can select any of the 52 cards to begin.
   Then, there are four cards higher and four lower, so a total of 16 options for the second card, a total of $52\times 16 = 832$ ways it can happen.
   The number of ways of getting two cards is $52\times 51 = 2652$, so the probability is $832/2652 \approx 0.30 = 137$.
:::




:::{.answer}
**Answer to Exercise \@ref(exr:Octaves).**

:::





<!-- :::{.answer} -->
<!-- **Answer to Exercise \@ref(exr:SomeProb).** -->
<!-- ::: -->



<!-- :::{.answer} -->
<!-- **Answer to Exercise \@ref(exr:SolvePermk).** -->

<!-- \begin{align*} -->
<!--    12\times {}^7P_k &= 7\times {}^6P_{k + 1} \\ -->
<!--    12\times \frac{7!}{(7 - k)!} &= 7\times\frac{6!}{(5 - k)!}\\ -->
<!--    \frac{12}{(7 - k)!} &= \frac{1}{(5 - k)!}\\ -->
<!--    \frac{12}{(7 - k)\times (6 - k)\times (5 - k)!} &= \frac{1}{(5 - k)!}\\ -->
<!--    12 &= (7 - k)(6 - k)\\ -->
<!--    k^2 - 13k +30 &= 0\\ -->
<!--    (k - 10)(k - 3) &= 0 -->
<!-- \end{align*} -->
<!-- and so $k = 10$ or $k = 3$. -->
<!-- But of $k = 10$, we get silly things like ${}^7P_{10}$. -->
<!-- So the solution must be $k = 3$. -->


<!-- ```{r} -->
<!-- for (k in (1:5)){ # Answer must e less than 6 -->
<!--   cat("FOR k = ",k, ":") -->
<!--   cat("LHS =", 12 * factorial(7) / factorial(7 - k), "; ") -->
<!--   cat("RHS =", 7 * factorial(6) / factorial(5 - k), "\n") -->
<!-- } -->
<!-- ``` -->
<!-- ::: -->




<!-- :::{.answer} -->
<!-- **Answer to Exercise \@ref(exr:SolvePermk).** -->

<!-- \begin{align*} -->
<!--    \frac{ {}^7 P_{r + 1}}{ {}^{7}C_r} -->
<!--    &= \frac{7!}{(7 - (r + 1))!} \times \frac{(7 - r)!\, r!}{7!}\\ -->
<!--    &= \frac{7!}{(6 - r)!} \times \frac{(7 - r)!\, r!}{7!}\\ -->
<!--    &= \frac{(7 - r)!\, r!}{(6 - r)!}\\ -->
<!--    &= \frac{(7 - r)\times (6 - r)!\, r!}{(6 - r)!}\\ -->
<!--    &= (7 - r) r!\\ -->
<!--    &= 10. -->
<!-- \end{align*} -->
<!-- Rewrite: $r! = 10/(7 - r)$. -->
<!-- Since $r!$ is a positive integer, $r$ must be either $r = 2$ or $r = 5$. -->
<!-- Trying both, clearly $r = 2$. -->
<!-- ::: -->




:::{.answer}
**Answer to Exercise \@ref(exr:BirthdayProblem).**

1. Proceed:  
\begin{align*}
  \Pr(\text{at least two have same birthday}) 
  &= 1 - \Pr(\text{no two birthdays the same}) \\
  &= 1 - \Pr(\text{every birthday is different}) \\
  &= 1 - \left(\frac{365}{365}\right) \times \left(\frac{364}{365}\right) \times \left(\frac{363}{365}\right) \times \cdots \times \left(\frac{365 - n + 1}{365}\right)\\
  &= 1 - \left(\frac{1}{365}\right)^{n} \times (365\times 364 \times\cdots (365 - n + 1) )
\end{align*}

2. Graph the relationship for various values of $N$ (from $2$ to $60$), using the above form to compute the probability.
3. How many people are needed to have at least a $50$% chance that at least two people share a birthday?
4. Birthdays are independent and randomly occur through the year (i.e., each day is equally likely).

```{r, fig.width = 5, fig.height=3.5, fig.align="center", fig.cap="Question 1", out.width='50%'}
N <- 2:80
probs <- array( dim = length(N) )

for (i in 1:length(N)){
  
  probs[i] <- 1 - prod( (365 - (1:N[i]) + 1)/365 )

}
plot( probs  ~ N,
      type = "l",
      lwd = 2,
      ylab = "Prob. at least two share birthday",
      xlab = expression(Group~size~italic(N)),
      las = 1)
```
:::





<!-- :::{.answer} -->
<!-- **Answer to Exercise \@ref(exr:PairsPoker).** -->

<!-- There are $\binom{52}{5} = 2,598,960$ ways to get five cards from 52 (without replacement). -->

<!-- 1. Pick the denomination: there are 13 to select from. -->
<!--    We need two of those 4 cards: $13\times \binom{4}{2}$. -->
<!--    Now the other three cards are drawn from the other 12 denominations: $\binom{12}{3}$. -->
<!--    There are also 4 ways to choose the suit of *each* of those 3 cards.   -->
<!--    All up then: the number of ways is $13\times\binom{4}{2}\times\binom{12}{3}\times 4^3 = 1,098,240$.   -->
<!--    The probability is therefore $1,098,240/2,598,960 = 0.422569$. -->
<!--    TAKE THE TREE OF A KIND AND FOUR OF A KIND!! -->
<!-- 1. There are $4$ suits each with $4$ picture cards, so $16$ picture cards in total. -->
<!--    We want to select five picture cards from $52$ cards, without replacement: -->
<!--    There are $16\times 15\times 14\times 13\times 12 = 524\ 160$ ways to do this. -->
<!--    So the probability is $524\ 160/^{52}C_5 = 0.2016807$. -->

<!-- Using the above logic again: -->
<!--    There are 13 denomination, and we select three of those four cards: $13\times \binom{4}{3}$. -->
<!--    The remaining two cards must be one of the other 12 denominations: $\binom{12}{2}\times 4^2$.   -->
<!--     All up then: the number of ways is $13\times\binom{4}{3}\times\binom{12}{2}\times 4^2 = 1,098,240$.   -->
<!--    The probability is therefore $54912/2,598,960 = 0.02112845$. -->
<!-- ::: -->


:::{.answer}
**Answer to Exercise \@ref(exr:NoTechP).**
$\binom{25}{8}/\binom{26}{6} = (25!\, 6!\, 19!)/(8!\,17!\,25!) = 171/28$.
:::



  
:::{.answer}
**Answer to Exercise \@ref(exr:Cards).**
$P = \{ (s, d) | s \in \{\text{`r if (knitr::is_latex_output()) {
   '\\Largedice{2}'
} else {
   '<span class="larger-die">&#9857;</span>'
}`,
`r if (knitr::is_latex_output()) {
   '\\Largedice{4}'
} else {
   '<span class="larger-die">&#9859;</span>'
}`,
`r if (knitr::is_latex_output()) {
   '\\Largedice{6}'
} else {
   '<span class="larger-die">&#9861;</span>'
}`}\} \text{and} D \in \{ 2, 3, \dots, 10, \text{Jack}, \text{Queen}, \text{King}, \text{Ace}\} \}$.
:::


:::{.answer}
**Answer to Exercise \@ref(exr:ProbHarder).**
$\mathbb{R} \in \mathbb{C}$ but $\mathbb{C} \notin \mathbb{R}$ 
:::


  