## Answers for Chap. 2

:::{.answer}
**Answer to Exercise \@ref(exr:SSpaceRV).**

1. $R_X = \{X \mid x \in (0, 1, 2) \}$; discrete.
2. $R_X = \{X \mid x \in (1, 2, 3\dots) \}$; discrete.
3. $R_X = \{X \mid x \in (0, \infty) \}$; continuous.
4. $R_X = \{X \mid x \in (0, \infty) \}$; continuous.
5. $R_X = \{X \mid x \in (0, 1, 2, \dots) \}$; discrete.
6. $R_X = \{X \mid x \in (0, 1, 2, \dots) \}$; discrete.
7. $R_X = \{X \mid x \in (0, \infty) \}$; continuous.
:::


:::{.answer}
**Answer to Exercise \@ref(exr:DFdiscrete).**

1. $\displaystyle F_X(x) = 
   \begin{cases}
      0   & \text{for $x < 10$};\\
      0.3 & \text{for $10 \le x < 15$};\\
      0.5 & \text{for $15 \le x < 20$};\\
      1   & \text{for $x \ge 20$}.
    \end{cases}$
1. $\Pr(X > 13) = 1 - F_X(13) = 0.7$.
1. $\Pr(X \le 10 \mid X\le 15) = \Pr(X \le 10) / \Pr(X \le 15) = F_X(10)/F_X(15) = 0.3/0.5 = 0.6$.
:::



:::{.answer}
**Answer to Exercise \@ref(exr:DFcontinuous).**

1. $\displaystyle F_Z(z) = 
    \begin{cases}
        0                     &  \text{for $z \le -1$};\\
        6z/15 - z^2/15 + 7/15 & \text{for $-1 < z < 2$};\\
        1                     & \text{for $z \ge 2$}.
    \end{cases}$
1. $\Pr(Z < 0) = F_Z(0) = 7/15 \approx 0.4666...$.
:::

```{r DFZcontinuous, fig.align="center", fig.cap="The DF for $f(z)$"}
F1 <- function(x){
  F <- array( dim = length(x))
  F[ x < -1] <- 0
  F[ x > 2] <- 1
  Interesting <- (x > -1) & (x < 2)
  F[ Interesting ] <- 6*z[Interesting]/15 - (z[Interesting]^2)/15 + 7/15
  F
}

z <- seq(-2, 3, 
         length = 200)
plot( F1(z) ~ z,
      las = 1,
      type = "l",
      xlab = expression(italic(x)),
      ylab = "Dist. fn.",
      lwd = 2)
```



:::{.answer}
**Answer to Exercise \@ref(exr:DFdiscreteBackwards).**

1. $\displaystyle
   p_W(w) = \begin{cases}
               0.3 & \text{for $w = 10$};\\
               0.4 & \text{for $w = 20$};\\
               0.2 & \text{for $w = 30$};\\
               0.1 & \text{for $w = 40$};\\
               0   & \text{elsewhere}.
            \end{cases}$
1. $\Pr(W < 25) = 0.7$.
:::


:::{.answer}
**Answer to Exercise \@ref(exr:DFcontinuousBackwards).**

1. $\displaystyle
   f_Y(y) = \begin{cases}
               (4/3) - y^2 & \text{for $0 < y < 1$};\\
               0 & \text{elsewhere}.
            \end{cases}$
1. $\Pr(Y < 0.5) = 0.625$.
:::

```{r, fig.align="center", fig.cap="A pdf"}
fy <- function(y){
  pdfy <- array(0, dim = length(y))
  pdfy <- ifelse( (y > 0) & (y < 1),
                  (4/3) - y^2,
                  0)
  pdfy
}
y <- seq(-1, 2,
         length = 500)
plot(fy(y) ~ y,
     lwd = 2,
     type = "l",
     xlab = expression(italic(y)),
     ylab = "PDF",
     las = 1)
```


<!-- :::{.answer} -->
<!-- **Answer to Exercise \@ref(exr:Chlorine).** -->

<!-- 1. Easiest to draw, and see that this represents a *triangular distribution*, and find the area of the said triangle. -->
<!--    Integration can be used though. -->
<!-- 2. PDF: -->
<!-- \begin{align*} -->
<!--    f_X(x)  -->
<!--    &=  -->
<!--    \begin{cases} -->
<!--       \frac{2}{8.8\times 4.4} (x - 0.6) & \text{for $0.6 < x < 5$};\\ -->
<!--       \frac{2}{8.8\times 4.4} (9.4 - x) & \text{for $5 < x < 9.4$} -->
<!--    \end{cases}\\ -->
<!--    &=  -->
<!--    \begin{cases} -->
<!--       \frac{2}{38.72} (x - 0.6) & \text{for $0.6 < x < 5$};\\ -->
<!--       \frac{2}{38.72} (9.4 - x) & \text{for $5 < x < 9.4$}. -->
<!--    \end{cases} -->
<!-- \end{align*} -->
<!-- 3.  -->
<!-- \[ -->
<!--    F_X(x)  -->
<!--    =  -->
<!--    \begin{cases} -->
<!--       0                                   & \text{for $x < 0.6$}\\ -->
<!--       [(0.6 - x)^2 ]/ 38.72               & \text{for $0.6 \le x < 5$};\\ -->
<!--       [-69 + 18.8x - x^2] / 38.72         & \text{for $5 \le x < 9.4$}.\\ -->
<!--       1                                   & \text{for $x \ge 9.4$} -->
<!--    \end{cases} -->
<!-- \] -->

<!-- 4. See below. -->
<!-- 5. $\Pr(X > 3) = 1 - \Pr(X < 3) =  1 - 0.1487603 =  0.8512397$ (or use areas of  triangles) -->
<!-- 6. $\Pr(X > 3 \mid X > 1) = \Pr(X > 3) / \Pr(X > 1) = 0.8512397 / 0.9958678 = 0.8547718$. -->
<!-- 7. One is just if $X$ exceeds 3; the other if $X$ exceeds 3 if we already know $X$ has exceeded 1. -->
<!-- ::: -->

<!-- ```{r, Triangular, out.width='90%', fig.align="center", fig.cap="A pdf and cdf"} -->
<!-- f <- function(x){ -->
<!--   f <- array( dim = length(x)) -->
<!--   f[x < 0.6] <- 0 -->
<!--   f[x > 9.4] <- 0 -->
<!--   xSmall <- (x >= 0.6) & ( x <= 5) -->
<!--   xLarge <- (x > 5) & (x <= 9.4) -->
<!--   f[xSmall] <- 2 / (8.8 * 4.4) * (x[xSmall] - 0.6) -->
<!--   f[xLarge] <- 2 / (8.8 * 4.4) * (9.4 - x[xLarge] ) -->
<!--   f -->
<!-- } -->

<!-- F <- function(x){ -->
<!--   FF <- array( dim = length(x)) -->

<!--   for (i in (1:length(x))){ -->
<!--     FF[i] <- integrate(f, lower = 0.6, upper = x[i])$value -->
<!--   } -->
<!--   FF -->
<!-- } -->


<!-- par(mfrow = c(1, 2)) -->
<!-- x <- seq(0, 10, -->
<!--          length = 200) -->
<!-- plot( f(x) ~ x, -->
<!--       type = "l", -->
<!--       lwd = 2, -->
<!--       las = 1, -->
<!--       main = "PROB fn", -->
<!--       xlab = expression(italic(x)), -->
<!--       ylab = "Prob. fn.") -->



<!-- x <- seq(0, 10, -->
<!--          length = 200) -->
<!-- plot( F(x) ~ x, -->
<!--       type = "l", -->
<!--       lwd = 2, -->
<!--       las = 1, -->
<!--       main = "DIST fn", -->
<!--       xlab = expression(italic(x)), -->
<!--       ylab = "Dist fn.") -->
<!-- ``` -->


<!-- ```{r} -->
<!-- 1 - F(3) -->
<!-- (1 - F(3)) / (1 - F(1))  -->
<!-- ``` -->



:::{.answer}
**Answer to Exercise \@ref(exr:TriangularHospital).**

Using the area of triangles, the 'height' is $2/27 \approx 0.07407407$ as shown below.
Then, after some algebra:

\[
   f_S(s) = 
   \begin{cases}
      \frac{4}{81}s - \frac{14}{81} & \text{for $3.5 < s < 5$};\\
      -\frac{4}{1377}s + \frac{122}{1377} & \text{for $5 < s < 30.5$}.
   \end{cases}
\]

Also,
\[
   F_S(s) = 
   \begin{cases}
      0                                    & \text{for $s < 3.5$};\\
      \frac{(7 - 2s)^2}{162}               & \text{for $3.5 < s < 5$};\\
      \frac{2}{1377} (s^2 - 61s + 280)     & \text{for $5 < s < 30.5$};\\
      1                                    & \text{for $s > 30.5$}.
   \end{cases}
\]

Then, 
$\Pr(S > 20 \mid S > 10) = \Pr( (S > 20) \cap (S > 10) )/\Pr(S > 10) = \Pr(S > 20)/\Pr(S > 10) = 0.1601311 / 0.6103854$.
This is about 0.2623.
:::


```{r, TriangularHospital, echo=FALSE, out.width='90%', fig.align="center", fig.cap="A pdf and cdf"}
par( mfrow = c(1, 2))

f <- function(x){
  f <- array( dim = length(x))
  f[x < 3.5] <- 0
  f[x > 30.5] <- 0
  xLO <- (x >= 3.5) & (x <= 5)
  xHI <- (x > 5) & (x <= 30.5)
  f[xLO] <- 4 * x[xLO] / 81 - (14 / 81)
  f[xHI] <- -4 * x[xHI]/1377 + 122/1377
  f
}

x <- seq(0, 35, 
         length = 500)

plot( f(x) ~ x,
      type = "l",
      lwd = 2,
      las = 1,
      main = "PROBABILITY function",
      xlab = expression(italic(s)),
      ylab = "Prob. fn.")
points(x = 5, 
       y = f(5), 
       pch = 19)
lines( x = c(0, 5),
       y = c(f(5), f(5)),
       lty = 2,
       lwd = 2)
axis(side = 2,
     at = f(5),
     labels = "2/27")




Fx <- function(x){
  F <- array( dim = length(x))
  
  for (i in (1:length(x))) {
    F[i] <- integrate(f, lower = 3.5, upper = x[i])$value
  }
  F
}

x <- seq(0, 35, 
         length = 500)

plot( Fx(x) ~ x,
      type = "l",
      lwd = 2,
      las = 1,
      ylim = c(0, 1),
      main = "DISTRIBUTION function",
      xlab = expression(italic(s)),
      ylab = "Dist. fn.")
```

```{r}
(1 - Fx(20) ) / ( 1 - Fx(10))
```

:::{.answer}
**Answer to Exercise \@ref(exr:Exceedance).**

1. Producers usually need that they will receive *at least* a certain amount of rainfall.
2. A *very poor* graph below; I really have to fix that.
3. Six months have recorded over 60mm; so 6/81. 
   But taking half of the previous '40 to under 60' category, we'd get 12/81.
   So somewhere around there.
4. In June, there are 81 observations, so the median is the 41st: The median rainfall is between 0 to under 20mm.  
   In December, there are 80 observations, so the median is the 40.5th: The median rainfall is between 40 to under 60mm.
5. Median; *very* skewed to the right.
6. ...
:::

```{r ExceedanceCharts, echo=FALSE, fig.align="center", fig.cap="Exceedance charts"}
Rain12 <- read.csv("Data/IDCJAC0001_044021/IDCJAC0001_044021_Data12.csv", 
                   na.strings = "null")
RainDec <- Rain12$Dec
RainJun <- Rain12$Jun

breaks <- c( 0, 0.01, 20,
             seq(40, 220, by = 20) )
outDec <- hist(RainDec,
               breaks = breaks,
               plot = FALSE)
outJun <- hist(RainJun, 
               breaks = breaks,
               plot = FALSE)

Buckets <- c("Zero", 
           "0 < R < 20",
           "20 <= R < 40",
           "40 <= R < 60",
           "60 <= R < 80",
           "80 <= R < 100",
           "100 <= R < 120",
           "120 <= R < 140",
           "140 <= R < 160",
           "160 <= R < 180",
           "180 <= R < 200",
           "200 <= R < 220")

cbind("Buckets" = Buckets,
      "Jun" = outJun$counts,
      "Dec" = outDec$counts)

exceedanceJun <-  ( sum(outJun$counts) - cumsum(outJun$counts) ) / sum(outJun$counts)
exceedanceDec <-  ( sum(outDec$counts) - cumsum(outDec$counts) ) / sum(outDec$counts)

plot( exceedanceJun ~ breaks[1:12], 
      type = "b",
      xlab = "Rainfall (in mm)",
      ylab = "Exceedance probability",
      main = "Charleville",
      ylim = c(0, 1.1),
      lwd = 2,
      lty = 1,
      pch = 19,
      las = 1)
lines( exceedanceDec ~ breaks[1:12], 
       type = "b",
      pch = 1,
      lty = 2,
      las = 1)
legend("topright",
       lwd = 2,
       pch = c(19, 1),
       lty = c(1, 2),
       legend = c("June",
                  "December"))
```



<!-- :::{answer} -->
<!-- **Answer for Exercise \@ref(exr:YouFriendInLine).** -->

<!-- Suppose I stand at Position 1; my friend can be at Positions 2 to 5, with distances 1, 2, 3 4. -->
<!-- Similar if I stand at Position 5. -->

<!-- Suppose I stand at Position 2; my friend can be at Positions 1, 3 to 5, with distances 1, 1, 2, 3. -->
<!-- Similar if I stand at Position 4. -->

<!-- Suppose I stand at Position 3; my friend can be at Positions 1, 2, 4, 5, with distances 1, 2, 2, 1. -->

<!-- So counting, the pdf for the number of people between us, say $Y$, is: -->

<!-- \[ -->
<!--    f_Y(y) =  -->
<!--    \begin{cases} -->
<!--       4/10 & \text{for $y = 0$};\\ -->
<!--       3/10 & \text{for $y = 1$};\\ -->
<!--       2/10 & \text{for $y = 2$};\\ -->
<!--       1/10 & \text{for $y = 3$};\\ -->
<!--          0 & \text{elsewhere}. -->
<!--    \end{cases} -->
<!-- \] -->
<!-- or $f_Y(y) = (4 - y)/10$ for $y\in\{0, 1, 2, 3\}$. -->
<!-- ::: -->


:::{answer}
**Answer for Exercise \@ref(exr:Continuous1).**

1. $a\int_0^1 (1 - y)^2\,dy = \left. (1 - y)^3/3\right|_0^1 = 1/3$.
1. $\Pr(|Y - 1/2| > 1/4) = \Pr(Y > 3/4) + \Pr(Y < 1/4) = 1 - \Pr(1/4 < Y < 3/4) = 13/32$.
1. See Fig. \@ref(fig:PDFy).

```{r PDFy, fig.cap="The pdf of $y$", fig.align="center"}
y <- seq(0, 1, length = 500)
fy <- ( (1 - y)^2 ) * 3

plot(fy ~ y,
     lwd = 2,
     xlab = "y",
     ylab = "PDF",
     type = "l")
```

:::




:::{answer}
**Answer for Exercise \@ref(exr:Discrete1).**

1. Write as $p(x) = \log_{10}(1 + x) - \log_{10}x$; then the sum is
\begin{align*}
    (\log_{10}(2) - \log_{10} 1) 
    + (\log_{10} 3 - \log_{10} 2)
    + (\log_{10} 4 - \log_{10} 3) 
    + \dots\\
    + (\log_{10} 9 - \log_{10} 8)
    + (\log_{10} 10 - \log_{10} 9)
\end{align*}
   and things cancel, leaving $\log_{10}10 = 1$.
1. DF:
:::





:::{answer}
**Answer for Exercise \@ref(exr:Discrete1).**
:::


:::{answer}
**Answer for Exercise \@ref(exr:Quartic).**
$\Pr(60 < Y < 70) = 154\,360\,000k/3\approx 51\,453\,333k$ and $\Pr(Y > 70) = 54\,360\,000k$.
Using this model, the larger probability is dying over 70.
:::



<!-- :::{answer} -->
<!-- **Answer for Exercise \@ref(exr:PooledTest).** -->
<!-- 1. See that $X = 1$ (i.e., one pooled test, and all are negative; no further testing needed) or $X = n + 1$ (the pooled test is positive, so $n$ individual tests are needed in addition to the pooled test). -->
<!--    So the sample space is $\{1, n + 1\}$. -->
<!-- 2. $X = 1$ only occurs if the test is negative; that is, $\Pr(X = 1) = (1 - p)^n$ -->
<!-- So:   -->
<!-- \[ -->
<!--   p_X(x) =  -->
<!--   \begin{cases} -->
<!--      (1 - p)^n     & \text{for $x = 1$ (i.e., the pooled test is negative)};\\ -->
<!--      1 - (1 - p)^n & \text{for $x = n + 1$} -->
<!--   \end{cases} -->
<!-- \] -->
<!-- and zero elsewhere. -->
<!-- ::: -->





<!-- :::{answer} -->
<!-- **Answer for Exercise \@ref(exr:CardDistance).** -->
<!-- So we have cards like:   -->
<!-- \[ -->
<!--   2, 3, 4, \dots, 8, 9, 10, 10 (J), 10 (Q), 10 (K), 10 (A). -->
<!-- \] -->
<!-- For a 'distance' say $D$, of $8$, we need either $(2, 10)$ or $(10, 2)$, with prob.   -->
<!-- \[ -->
<!--   \Pr(D = 8) = \frac{4 \times 20}{52 \times 51} = 0.03016591. -->
<!-- \] -->
<!-- For a 'distance' of $7$, we need either $(3, 10)$ or $(10, 3)$, or $(2, 9)$ or $(9, 2)$, with prob.   -->
<!-- \[ -->
<!--   \Pr(D = 7) = \frac{4 \times 20}{52 \times 51} + \frac{4 \times 4}{52\times 51} \times 2 = 0.04223228. -->
<!-- \] -->
<!-- For a 'distance' of $6$, we need either $(4, 10)$ or $(10, 4)$, or $(3, 9)$ or $(9, 3)$, or $(2, 8)$ or $(8, 2)$, with prob.   -->
<!-- \[ -->
<!--   \Pr(D = 6) = \frac{4 \times 20}{52 \times 51} + \frac{4 \times 4}{52\times 51} \times 4 = 0.05429864. -->
<!-- \] -->
<!-- So in general, for $D = 1, 2, \dots 8$:   -->
<!-- \[ -->
<!--   \Pr(D = d) = \frac{4 \times 20}{52 \times 51} + \frac{4 \times 4}{52\times 51} \times 2(8 - d). -->
<!-- \] -->

<!-- The case $D = 0$ is different. -->
<!-- A 'distance' of zero can occur if we get 1 'ten' and another 'ten', **or** two of the other (non-'ten') cards:   -->
<!-- \[ -->
<!--   \Pr(D = 0) = \frac{4 \times 20}{52 \times 51} \times 2 + \frac{4\times 4}{52\times 51}\times 8 = 0.1085973. -->
<!-- \] -->
<!-- See the pmf in Fig.\ \@ref(fig:CardDistance). -->

<!-- ::: -->


<!-- ```{r CardDistance, fig.cap="The pdf for the 'distance' between cards", fig.align="center"} -->
<!-- d <- 1:8 -->
<!-- pd <- (4 * 20 * 2)/(52 * 51) + (4 * 4)/(52 * 51) * 2 * (8 - d) -->

<!-- d <- c(0, d) -->
<!-- pd <- c( (4 * 3 * 8)/(52 * 51) + (20 * 19)/(52 * 51),  -->
<!--          pd ) -->
<!-- plot(pd ~ d,  -->
<!--      type = "p", -->
<!--      las = 1, -->
<!--      pch = 19, -->
<!--      xlab = expression(italic(d)), -->
<!--      ylab = "pmf") -->
<!-- ``` -->



:::{answer}
**Answer to Exercise \@ref(exr:BenfordsLaw).**

The df is
\begin{align*}
   F_D(d) 
   &= \sum_{i = 1}^d \log[ (1 + d)/ \log(d) ] \\
   &= \sum_{i = 1}^d \log(1 + d) - \log d\\
   &= (\log 2 - \log 1) + (\log 3 - \log 2) + \log 4 - \log 3) + \cdots\\
   &  \quad {} + (\log d - \log(d - 1) ) + (\log (1 + d) - \log d)\\
   &= \log (1 + d).
\end{align*}
:::

```{r, BenfordPlots, fig.align="center", fig.cap="A pdf and cdf"}
fd <- function(x){
  log10( 1 + 1/d)
}

Fd <- function(x){
  log10(1 + d)
}

d <- 1:9

par(mfrow = c(1, 2))

plot( fd(d) ~ d,
      type = "p",
      pch = 19,
      las = 1,
      main = "Prob. fn",
      xlab = expression(italic(d)),
      ylab = "Prob. fn.")
points( fd(d) ~ d,
        type = "h",
        lty = 2,
        col = "grey",
        lwd = 2)

plot( Fd(d) ~ d,
      type = "p",
      pch = 19,
      las = 1,
      main = "Distribution fn",
      ylim = c(0, 1),
      xlab = expression(italic(d)),
      ylab = "Prob. fn.")
points( Fd(d) ~ d,
        type = "h",
        lty = 2,
        col = "grey",
        lwd = 2)
```


<!-- :::{answer} -->
<!-- **Answer for Exercise \@ref(exr:Arcsine).** -->

<!-- We find $k = 1/\pi$. -->
<!-- $F(y) = \int_0^x \frac{1}{\pi\sqrt{t(1 - t)}}\, dt = \frac{1}{\pi}\arcsin(2x - 1) + \frac{1}{2}$ for $0 < x < 1$. -->
<!-- ::: -->

<!-- ```{r, ArcsinePlots, fig.align="center", fig.cap="A pdf and cdf"} -->
<!-- y <- seq(-0.5, 1.5,  -->
<!--          length = 100) -->
<!-- fy <- ifelse( (y > 0) & (y < 1), -->
<!--               1/( pi * sqrt( y * (1 - y))), -->
<!--               0) -->

<!-- Fy <- array( 0,  -->
<!--              dim = length(y)) -->
<!-- Fy[ y < 0 ] <- 0 -->
<!-- Fy[ y > 1 ] <- 1 -->
<!-- Fy[ (y > 0) & (y < 1) ] <- 1/pi * asin(2*y[(y > 0) & (y < 1)] - 1) + 0.5 -->

<!-- par(mfrow = c(1, 2)) -->

<!-- plot( fy ~ y, -->
<!--       type = "l", -->
<!--       lwd = 2, -->
<!--       xlab = expression(italic(x)), -->
<!--       ylab = "Prob. fn", -->
<!--       las = 1) -->
<!-- plot( Fy ~ y, -->
<!--       type = "l", -->
<!--       lwd = 2, -->
<!--       ylim = c(0, 1), -->
<!--       xlab = expression(italic(x)), -->
<!--       ylab = "Dist. fn", -->
<!--       las = 1) -->
<!-- ``` -->




:::{answer}
**Answer for Exercise \@ref(exr:DFExpInverse).**

The pdf is
\[
   f_X(x) 
   = \frac{d}{dx} \exp(-1/x) 
   = \frac{\exp(-1/x)}{x^2}
\]
for $x > 0$.
See Fig. \@ref(fig:PlotDFExpInverse).
:::

```{r PlotDFExpInverse, echo=FALSE, fig.cap="A pdf", fig.align="center", out.width='80%'}
par( mfrow = c(1, 2))

f1 <- function(x){
  fy <- array(0, dim = length(x))
  fy <- ifelse(x > 0,
               exp(-1/x) / x^2,
               0)
  fy
}

x <- seq(-0.5, 5, length = 100)

plot( f1(x) ~ x,
      type = "l",
      lwd = 2,
      xlab = expression(italic(x)),
      ylab = expression(italic(f)(italic(x))),
      las = 1)



F1 <- function(x){
  fy <- array(0, dim = length(x))
  fy <- ifelse(x > 0,
               exp(-1/x),
               0)
  fy
}

x <- seq(-0.5, 8, length = 100)

plot( F1(x) ~ x,
      type = "l",
      lwd = 2,
      ylim = c(0, 1),
      xlab = expression(italic(x)),
      ylab = expression(italic(F)(italic(x))),
      las = 1)


```

:::{answer}
**Answer for Exercise \@ref(exr:Cards4).**

* 1 suit: Select 4 cards from the 13 of that suit, and there are four suits to select.
* 2 suits: There are two scenarios here: 
  - Three from one suit, and one from another: 
    Choose a suit, and select three cards from it: $\binom{4}{3}\binom{13}{3}$.
    Then we need another suit (three choices remain) and one card from (any of the 13).
  - Chose two suits, and two cards from each of two suits:
    $\binom{4}{2}\binom{13}{2}\binom{13}{2}$
* 3 suits: Umm...
* 4 suits: Choose one from each of the four suits.

One way to get 3 suits is to realise that the total probablity must add to one...

```{r echo=TRUE}
### 1 SUIT
suits1 <- 4 * choose(13, 4) / choose(52, 4)

### 2 SUITS
suits2 <- choose(4, 3) * choose(13, 3) * choose(3, 1) * choose(13, 1) + 
          choose(4, 2) * choose(13, 2) * choose(13, 2)
suits2 <- suits2 / choose(52, 4)

### 4 SUITS:
suits4 <- choose(13, 1) * choose(13, 1) * choose(13, 1) * choose(13, 1)
suits4 <- suits4  / choose(52, 4)

suits3 <- 1 - suits1 - suits2 - suits4

round( c(suits1, suits2, suits3, suits4), 3)
```
:::
  