## Answers for Chap. 7

:::{.answer}
**Answer to Exercise \@ref(exr:X2).**

For $0 < x < 2$, the transformation is one-to-one.
The inverse transform is $X = Y^{1/3}$, and so $0 < y < 8$.

1. $F_Y(y) = \Pr(Y\le y)
   = \Pr(X^3 \le y) = \Pr(X\le y^{1/3}) = F_X( y^{1/3})
   = \int_{u = 0}^{y^{1/3}} u/2\,du = u^2/4\Big|_{u = 0}^{u = y^{1/3}} = y^{2/3}/4.$
Differentiate to find the pdf:
$\frac{d}{dy} y^{2/3}/4 = y^{-1/3}/6$.
The pdf of $Y$ is

\[
   f_Y(y) = \begin{cases}
               y^{-1/3}/6 & \text{for $0 < y < 8$};\\
               0 & \text{otherwise}.
            \end{cases}
\]

2. Since $w(y) = y^{1/3}$, then $w'(y) = y^{-2/3}$.
Then:
\begin{align*}
   f_Y(y)
   &= f_X(y) |J|\\
   &= y^{1/3}/2 \times \overbrace{y^{-2/3}/3}^{|J|} \\
   &= y^{-1/3}/6.
\end{align*}
So the pdf of $Y$ is as above.
:::


:::{.answer}
**Answer to Exercise \@ref(exr:JointDiscrete).**

First:

* For $X_1 = 0$, $X_2= 0$ (with prob: $0$):   $Y_1 = 0$; $Y_2 = 0$;
* For $X_1 = 0$, $X_2= 1$ (with prob. $1/6$): $Y_1 = 1$; $Y_2 = 1$;
* For $X_1 = 1$, $X_2= 0$ (with prob. $2/6$): $Y_1 = 1$; $Y_2 = 0$;
* For $X_1 = 1$, $X_2= 1$ (with prob. $3/6$): $Y_1 = 2$; $Y_2 = 1$.

Effectively, the first line can be ignored (since the probability is zero), so $Y_1 \in \{1, 2\}$ and $Y_2\in\{0, 1\}$.

1. From the above, the joint pf is shown in Table \@ref(tab:JDTable).
1. Hence, from the table:

\[
   f_{Y_1}(y_1) =
   \begin{cases}
      1/2 & \text{if $y_1 = 1$};\\
      1/2 & \text{if $y_1 = 2$};\\
      0    & \text{otherwise}.
    \end{cases}
\]

```{r JDTable, echo=FALSE}
JDTable <- array( dim = c(2, 3))

colnames(JDTable) <- c("",
                       "$Y_1 = 1$",
                       "$Y_1 = 2$")
JDTable[1, ] <- c("$Y_2 = 0$",
                  "$2/6$",
                  "$0$")
JDTable[2, ] <- c("$Y_2 = 1$",
                  "$1/6$",
                  "$3/6$")
if (knitr::is_latex_output()) {
  kable(JDTable,
    format = "latex",
    escape = FALSE,
    caption = "The joint distribution of $Y_1$ and $Y_2$",
    booktabs = TRUE) %>%
  kable_styling(font_size = 10) %>%
  row_spec(0, bold = TRUE)
}
if (knitr::is_html_output()){
  kable(JDTable,
    caption = "The joint distribution of $Y_1$ and $Y_2$",
    format = "html",
    booktabs = TRUE) %>%
  kable_styling(font_size = 10) %>%
  row_spec(0, bold = TRUE)
}
```
:::


:::{.answer}
**Answer to Exercise \@ref(exr:GammaSum).**

$Y \sim\text{Gam}(\sum\alpha, \beta)$.
:::




<!-- :::{.answer} -->
<!-- **Answer to Exercise \@ref(exr:CauchySquared).** -->

<!-- Transformation is not 1-1; and $Y > 0$. -->

<!-- Then:   -->
<!-- \begin{align*} -->
<!--    F_Y(y)  -->
<!--    &= \Pr(Y < y)\\ -->
<!--    &= \Pr( X^1 < y)\\ -->
<!--    &= \Pr( -\sqrt{y} < X < \sqrt{y} ) \quad\text{(draw a diagram!)}\\ -->
<!--    &= \int_{-\sqrt{y}}^{\sqrt{y}} \frac{1}{\pi(1 + x^2)}\, dx\\ -->
<!--    &= \frac{2}{\pi} \tan^{-1}(\sqrt{y}). -->
<!-- \end{align*} -->
<!-- and so   -->
<!-- \begin{align*} -->
<!--    f_Y(y) -->
<!--    &= \frac{d}{dy} \frac{2}{\pi} \tan^{-1}(\sqrt{y})\\ -->
<!--    &= \frac{1}{\pi(y + 1)\sqrt{y}}\quad\text{for $y > 0$}. -->
<!-- \end{align*} -->

<!-- ::: -->


:::{.answer}
**Answer to Exercise \@ref(exr:ContTransform).**

1. Differentiating:  
   \[
     f_X(x) = 
     \begin{cases}
       1 & \text{for $-1/2 < x < /1/2$};\\
       0 & \text{elsewhere}.
     \end{cases}
   \]
2. First: this transformation is not 1:1 (Fig. \@ref(fig:TransformSquared)).
   See that $X = \pm\sqrt{4 - Y}$, and that $3.75 < Y < 4$.
   So,   
   \begin{align*}
     F_Y(y) 
     &= \Pr(Y < y) \\
     &= 1 - \Pr( -\sqrt{4 - y} < X < \sqrt{4 - y} )\\
     &= 1 - \int_{-\sqrt{4 - y}}^{\sqrt{4 - y}} 1\, dx\\
     &= 1 - 2\sqrt{4 - y},
   \end{align*}
   and so, differentiating:
   \[
     f_Y(y) = \frac{1}{\sqrt{4 - y}} \quad\text{for $3.75 < Y < 4$}
   \]
:::

```{r TransformSquared, out.width='100%', fig.width=9, fig.height=3,fig.cap="The transformation", fig.align="center"}
par( mfrow = c(1, 3))

xwide <- seq(-0.75, 0.75,
             length = 100)
x <- seq( -0.5, 0.5,
          length = 100)
y <- dunif(xwide, 
           min = -0.5,
           max = 0.5)
plot( y ~ xwide,
      lwd = 2,
      las = 1,
      type = "l",
      main = expression(The~PDF~"of"~italic(X)),
      xlab = expression(italic(X)),
      ylab = expression(PDF~"of"~italic(X)) )


###

Y <- (4 - x^2)
Ywide <- (4 - xwide^2)

plot(Ywide ~ xwide,
     main = "The transformation",
     las = 1,
     lwd = 2,
     type = "l",
     col = "grey",
     xlab = expression(italic(X)),
     ylab = expression(italic(Y)))
lines(Y ~ x,
      lwd = 4 ,
      col = "black")

abline(v = 0)

yht <- 3.85
x1 <- -sqrt(4 - yht)
x2 <- sqrt(4 - yht)
abline( h = yht,
        col = "grey",
        lty = 2)
abline( v = x1,
        col = "grey",
        lty = 2)
abline( v = x2,
        col = "grey",
        lty = 2)

text(x = x1,
     y = yht,
     pos = 1,
     labels = expression(italic(x)[1]) )
text(x = x2,
     pos = 1,
     y = yht,
     labels = expression(italic(x)[2]) )
mtext(expression(italic(y)),
      at = yht,
      line = 1,
      side = 2,
      las = 1)
##
y <- seq(3.75, 4,
         length = 100)
PDFy <- 1 / sqrt(4 - y)

plot( PDFy ~ y,
      lwd = 2,
      las = 1,
      type = "l",
      ylim = c(0, 10),
      xlim = c(3.65, 4.15),
      main = expression(The~PDF~"of"~italic(Y)),
      xlab = expression(italic(Y)),
      ylab = expression(PDF~"of"~italic(Y)) )
lines( x = c(3.5, 3.75),
       y = c(0, 0),
       lwd = 2)
lines( x = c(4, 4.15),
       y = c(0, 0),
       lwd = 2)
abline( v = c(3.75, 4),
        lty = 2,
        col = "grey")
```




<!-- :::{.answer} -->
<!-- **Answer to Exercise \@ref(exr:Projectile).** -->

<!-- Given that $f(\theta) = 4/\pi$ for $0 < \theta< \pi/4$, and hence $0 < D < (v^2/g)$. -->
<!-- Then:   -->
<!-- \begin{align*} -->
<!--   F_D(d) = \Pr(D < d)  -->
<!--   &= \Pr( v^2/g\sin2\theta < d) \\ -->
<!--   &= \Pr(\theta < \frac{1}{2} \sin^{-1}(d g/v^2 ) \\ -->
<!--   &= \int_0^{\sin^{-1}(d g/v^2 )/2} \frac{4}{\pi}\, d\theta \\ -->
<!--   &= \frac{2}{\pi} \sin^{-1} \left( \frac{dg}{v^2} \right). -->
<!-- \end{align*} -->
<!-- So    -->
<!-- \[ -->
<!--   f_D(d) = \frac{2g}{\pi}\frac{1}{\sqrt{v^4 - d^2 g^2}}\quad\text{for $0 < D < v^2 /g$}. -->
<!-- \] -->

<!-- Alternatively: see that $D = v^2\sin 2\theta/g \Rightarrow sin 2\theta = Dg/v^2$, and so   -->
<!-- \[ -->
<!--   \frac{dd}{d\theta} = \frac{2v^2}{g}\cos 2\theta. -->
<!-- \] -->
<!-- Then,   -->
<!-- \begin{align*} -->
<!--   f_D(d)  -->
<!--   &= \frac{4}{\pi} \left| \frac{g}{2 v^2\cos 2\theta} \right|\\ -->
<!--   &= \frac{2g}{\pi v^2 \cos2\theta}\\ -->
<!--   &= \frac{2g}{\pi \sqrt{v^4 - d^2 g^2}} -->
<!-- \end{align*} -->
<!-- after (e.g.) drawing the right-angled triangle to re-write $\sin 2\theta = Dg/v^2$.  -->
<!-- See Fig. \@ref(fig:Projectile). -->
<!-- ::: -->

<!-- ```{r Projectile, fig.align="center", fig.cap="The distance travelled by a projectile"} -->
<!-- g <- 9.8 -->
<!-- v <- 12 -->
<!-- dvec <- seq(0, v^2/g, -->
<!--             length = 100) -->
<!-- pdf <- ( 2 * g / pi ) / sqrt(v^4 - dvec^2 * g^2) -->

<!-- plot(pdf ~ dvec, -->
<!--      type = "l", -->
<!--      las = 1, -->
<!--      lwd = 2, -->
<!--      xlab = expression(Distance~italic(d)), -->
<!--      ylab = "Prob. fn.") -->
<!-- abline(v = v^2/g, -->
<!--        col = "grey", -->
<!--        lty = 2) -->
<!-- ``` -->


<!-- :::{.answer} -->
<!-- **Answer to Exercise \@ref(exr:RandomExp).** -->
<!-- Proceed:   -->
<!-- \begin{align*} -->
<!--   F_Y(y)  -->
<!--   &= \Pr(Y \le y)\\ -->
<!--   &= \Pr(X \ge \exp(-y/\alpha)) \qquad\text{note the change of direction!}\\ -->
<!--   &= \int_{\exp(-y/\alpha)}^1 1\, dx -->
<!-- \end{align*} -->
<!-- so that $f_Y(y) = \frac{1}{\alpha} \exp(-y/\alpha)$ for $y > 0$, which is the exponential distribution.  -->
<!-- ::: -->



<!-- :::{.answer} -->
<!-- **Answer to Exercise \@ref(exr:RandomW).** -->

<!-- 2. $\text{E}[W] = -1/3$; $\text{var}[W] = 17/9\approx 1.8889\dots$. -->
<!-- 3. $\Pr(V = 0) = 1/2$; $\Pr(V = 4) = 1/2$. -->
<!-- 4. Find:   -->
<!-- \[ -->
<!--   F_W(w) =  -->
<!--   \begin{cases} -->
<!--     0 & \text{for $w < -2$};\\ -->
<!--     1/3 & \text{for $-2 \le w < 0$};\\ -->
<!--     5/6 & \text{for $0 \le w < 2$};\\ -->
<!--     1 & \text{for $w \ge 2$}. -->
<!-- \] -->
<!-- ::: -->



:::{.answer}
**Answer to Exercise \@ref(exr:LogNormal)**.

1. First see that $Y = \log X$.
   Then:  
   \begin{align*}
     F_Y(y)
     &= \Pr(Y < y) \\
     &= \Pr( \exp X < y)\\
     &= \Pr( X < \log y)\\
     &= \Phi\big((\log y - \mu)/\sigma\big) 
   \end{align*}
   by the definition of $\Phi(\cdot)$.
2. Proceed:
   \begin{align*}
     f_Y(y) 
     &= \frac{d}{dy}  \Phi\big((\log y - \mu)/\sigma\big) \\
     &= \frac{1}{y} \Phi\big((\log y - \mu)/\sigma\big) \\
     &= \frac{1}{y\sqrt{2\pi\sigma^2}}
        \exp\left[\left(
           -\frac{ (\log y - \mu)^2}{\sigma}\right)^2\right]
   \end{align*}
   since the derivative of $\Phi(\cdot)$ (the df of a standard normal distribution) is $\phi(\cdot)$ (the pdf of a standard normal distribution).
3. See Fig. \@ref(fig:LogNormalPlots).
4. See below: About 0.883.
:::


```{r LogNormalPlots, fig.align="center", fig.cap="Log-normal distributions", out.width='100%', fig.width=9.5, fig.height = 7}
par( mfrow = c(2, 2))

x <- seq(0, 8,
         length = 500)

plot( dlnorm(x, meanlog = log(1), sdlog = 1) ~ x,
      xlab = expression(italic(y)),
      ylab = "PDF",
      type = "l",
      main = expression(Log~normal*":"~mu==1~and~sigma==1),
      lwd = 2,
      las = 1)
plot( dlnorm(x, meanlog = log(3), sdlog = 1) ~ x,
      xlab = expression(italic(y)),
      ylab = "PDF",
      type = "l",
      main = expression(Log~normal*":"~mu==3~and~sigma==1),
      lwd = 2,
      las = 1)
plot( dlnorm(x, meanlog = log(1), sdlog = 2) ~ x,
      xlab = expression(italic(y)),
      ylab = "PDF",
      main = expression(Log~normal*":"~mu==1~and~sigma==2),
      type = "l",
      lwd = 2,
      las = 1)
plot( dlnorm(x, meanlog = log(3), sdlog = 2) ~ x,
      xlab = expression(italic(y)),
      ylab = "PDF",
      main = expression(Log~normal*":"~mu==3~and~sigma==2),
      type = "l",
      lwd = 2,
      las = 1)
```

```{r}
(1 - plnorm(2, meanlog=2, sdlog=2)) / 
  (1 - plnorm(1, meanlog=2, sdlog=2))
```





:::{.answer}
**Answer to Exercise \@ref(exr:BinomXform).**
See that $Y\in\{0, 1, \sqrt{2}, \sqrt{3}, 2\}$ and so  
\[
   \Pr(Y = y) = \binom{4}{y^2} (0.2)^{y^2} (0.8)^{4 - y^2} \quad \text{for $y = 0, 1, \sqrt{2}, \sqrt{3}, 2$}.
\]
:::



<!-- :::{.answer} -->
<!-- **Answer to Exercise \@ref(exr:Xminus3).** -->

<!-- 1. First, see the relationships:   -->
<!-- \begin{align*} -->
<!--   X = 1 &\to Y = (X - 3)^2 = 4;\\ -->
<!--   X = 2 &\to Y = (X - 3)^2 = 1;\\ -->
<!--   X = 3 &\to Y = (X - 3)^2 = 0;\\ -->
<!--   X = 4 &\to Y = (X - 3)^2 = 1. -->
<!-- \end{align*} -->
<!--   So adding probabilities as appropriate:   -->
<!-- \[ -->
<!--   p_Y(y) =  -->
<!--   \begin{cases} -->
<!--      9/30  & \text{for $y = 0$}\\ -->
<!--      20/30 & \text{for $y = 1$}\\ -->
<!--      1/30  & \text{for $y = 4$}\\ -->
<!--      0     & \text{elsewhere}. -->
<!-- \end{cases} -->
<!-- \] -->
<!-- ::: -->




:::{.answer}
**Answer to Exercise \@ref(exr:BetaHospital2).**

For the given beta distribution, $\text{E}(V) = 0.287/(0.287 + 0.926) = 0.2366...$ and $\text{Var}(V) = 0.08161874$.

1. $\text{E}(S) = \text{E}(4.5 + 11V) = 4.5 + 11\text{E}(V) = 7.10$ minutes.
   $\text{var}(S) = 11^2\times\text{Var}(V) = 9.875$ minutes^2^.
1. $V\in (4.5, 15.5)$.
   See Fig. \@ref(fig:ServiceTimes).
1. This corresponds to $V = 10.5/11 = 0.9545455$, so $\Pr(S > 15) = \Pr(V > 0.9545455) = 0.01745087$.
1. With $V$, the largest 20% correspond to $V = 0.004080076$, so that $S = 4.544881$; the *quickest* 20% are within 4.54 minutes.
:::


```{r ServiceTimes, fig.cap="Service times", fig.align="center"}
v <- seq(0, 1, length = 500)
stime <- 4.5 + 11 * v
  
PDFv <- dbeta(v, shape1 = 0.287, shape2 = 0.926)

plot( PDFv ~ stime,
      type = "l",
      las = 1,
      xlim = c(0, 15.5),
      ylim = c(0, 5),
      xlab = expression(italic(S)),
      ylab = "PDF",
      main = "Distribution of S",
      lwd = 2)
abline(v = 4.5,
       lty = 2,
       col = "grey",
       lwd = 2)
## Part 3
1 - pbeta( 0.9545455, shape1 = 0.287, shape2 = 0.926)

## Part 4
qbeta(0.2, shape1 = 0.287, shape2 = 0.926) + 4.5
```



