## Censored model {#CensoredModels}
\index{Censored models|(}

### Definitions {#CensoredModelsDefinitions}

Censored model assume a distribution for\ $Y$, such as a normal (or other continuous) distribution, but the value cannot always be observed exactly.
Commonly, the value of the random variable is only observed when it exceeds some threshold (typically zero). 
Below that threshold, the threshold vaklue is recorded.
For example, daily rainfall in Australia less than $0.2\mms$ is recorded as zero rainfall; rainfall amount less than this cannot be measured.

Censoring\index{Censoring} can occur at the lower end of the scale (left censoring, where values below a threshold are unobserved) or the upper end (right censoring, where values above a threshold are unobserved). 
Right censoring\index{Censoring!right} is common in survival analysis, when individuals live beyond the end if the study; left censoring\index{Censoring!left} arises naturally when instruments cannot record below a detection limit.


:::{.definition #Censoring name="Censoring"}
  A random variable\ $Y$ is censored if its exact value is only observed when it falls within some interval $[L, U]$; otherwise, only the bound that was exceeded is recorded.\index{Censoring}
  
  *Left censoring*\index{Censoring!left} occurs when there is a finite lower threshold\ $L$ and no upper censoring: values below\ $L$ are unobserved and recorded as\ $L$.
  For left censoring at\ $L$:
    $$
    Y^\text{obs} 
  = 
    \begin{cases}
  L & \text{if $Y \le L$};\\
  Y & \text{if $L > Y$}.
  \end{cases}
  $$
    
    
    *Right censoring*\index{Censoring!right} occurs when there is a finite upper threshold\ $U$ and no lower censoring: values above\ $U$ are unobserved and recorded as\ $U$.
  For right censoring at\ $L$:
    $$
    Y^\text{obs} 
  = 
    \begin{cases}
  Y & \text{if $Y < U$};\\
  U & \text{if $U \ge Y$}.
  \end{cases}
  $$
    :::
    
    
    One specific censored model is a *Tobit model*\index{Tobit model}.
  A Tobit model assumes normal distribution,\index{Nornal distribution} left-censored\index{Censoring!left} at zero (i.e., $L = 0$).
  We will focus on the left-censored case where $L = 0$, but the underlying distribution is not necessarily a normal distribution (CHECK!).
  
  
  :::{.example #TobitAlcoholSpend name="Latent variables and censoring"}
    FIX and change (Fig.\ \@ref(fig:AlcoholSpend)).
    :::
      
      ```{r AlcoholSpend, echo=FALSE, fig.align="center", fig.cap="Left: the latent variable\\ $Y$, the desire to buy alcohol. Right: the observed variable\\ $X$, the average weekly spend on alcohol. The solid dot is the sum of all probabilities for which $Z \\le 0$.", fig.width=8, fig.height=4, out.width="100%"}
    par(mfrow = c(1, 2))
    
    ############### Latent
    zz <- seq(-3, 3.5, 
              length = 1000) + 1 # 1 is the ncp
    z <- zz * 15
    
    fz <- dt(zz, 
             df = 4, 
             ncp = 1)
    
    plot(fz ~ z,
         type = "l",
         xlim = range(z),
         ylim = c(0, max(fz)),
         lwd = 3,
         col = plotColour1,
         axes = FALSE,
         main = expression(atop(Latent~variable~italic(Y)*":",
                                Desire~to~buy~alcohol)),
         xlab = expression(Latent~variable~italic(Y)),
         ylab = expression(Probability~fn.)
    )
    lines(fz[z<0] ~ z[z<0],
          lwd = 3,
          col = "grey")
    axis(side = 1)
    box()
    abline(h = 0,
           col = "grey")
    abline(v = 0,
           lty = 2,
           col = "grey",
           lwd = 2)
    
    text(x = -18,
         y = max(fz) * 0.75, 
         label = "Aversion\nto buying\nalcohol",
         cex = 0.9)
    text(x = 45,
         y = max(fz) * 0.75, 
         label = "Propensity\nto buying\nalcohol",
         cex = 0.9)
    
    
    ############### Observed
    y0 <- pt( 0, 
              df = 4,
              ncp = 1)
    
    plot(fz[z >= 0] ~ z[z >= 0],
         xlim = range(z),
         ylim = c(0, max(fz)),
         type = "l",
         lwd = 3,
         col = plotColour1,
         axes = FALSE,
         main = expression(atop(Observed~variable~italic(X)*":",
                                Average~weekly~spend~on~alcohol)),
         xlab = expression(Observed~variable~italic(X)),
         ylab = expression(Probability~fn.)
    )
    abline(h = 0,
           col = "grey")
    axis(side = 1)
    axis(side = 2,
         at = round(y0, 2),
         las = 2)
    box()
    
    # Grey where latent variable is less than zero
    lines(x = z[z < 0],
          y = fz[z < 0],
          col = "grey",
          lwd = 2,
          lty = 3)
    
    abline(v = 0,
           lty = 2,
           col = "grey",
           lwd = 2)
    
    # Points for Y = 0
    points(x = 0,
           y = y0,
           pch = 19)
    
    #points(x = Z_to_Y(0),
    #       y = dt(0, 
    #              df = 4,
    #              ncp = 1),
    #       pch = 1)
    text(x = 28,
         y = max(fz)/5,
         label = "Prob. that no\nalcohol purchased",
         pos = 1,
         cex = 0.9)
    arrows(x0 = 28,
           y0 = max(fz)/5,
           x1 = 0,
           y1 = pt(0, 
                   df = 4,
                   ncp = 1),
           angle = 15,
           length = 0.15)
    
    ```
    
    
    In this context, the value of the observed variable\ $X$ is said to be *censored*.
    
    
    :::{.definition #Censoring name="Censoring"}
      A variable is called *censored*\index{Censoring} if its value is only observed when its above or below a certain threshold.
      
      *Left censoring*\index{Censoring!left} occurs when the true (latent) value is less than or equal to a known threshold, but the exact value is unknown.
      We only know that the true (latent) value is *below* a certain threshold.
      
      *Right censoring*\index{Censoring!right} occurs when the true (latent) value is greater than or equal to a known threshold, but the exact value is unknown.
      We only know that the true (latent) value is *above* a certain threshold.
      :::
        
        
        :::{.example #AlcoholLeftCensoring name="Left censoring"}
          The latent variable 'desire to purchase alcohol' used in Example\ \@ref(exm:TobitAlcoholSpend) is *left censored*.\index{Censoring!left}
          :::
            
            
            :::{.example #insectLeftCensoring ame="Right censoring"}
              In survival analysis,\index{Survival analysis} the time it takes insects to die (say\ $X$) may be studied.
              If the study ends at time\ $t^*$, the time of death for any insects still alive is some value larger than $t^*$; that is, $X\ge t^*$ but the exact time of death remains unknown.
              The time to death is *right censored*.\index{Censoring!right}
              :::
                
                
                
                ::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
              *Censoring*\index{Censoring} is not the same as *truncation*.\index{Truncation}
              
              *Censoring* retains observations, but the exact value remains unknown.
              *Truncation* removes observations whose value falls above or below a threshold.
              :::
                
                
                
                Usually $y^* = 0$ (as in Example\ \@ref(exm:TobitAlcoholSpend)), so that any value of\ $Y \le 0$ is recorded as\ $y = 0$.
              That is, we observe the random variable\ $X$ such that
              $$
                X = \max(Y, 0).
              $$
                This means the probability function of\ $X$ is
              $$
                f_X(x) = 
                \begin{cases}
              0       & \text{if $X < 0$}\\
              F_Y(0)  & \text{if $X = 0$}\\
              f_Y(y)  & \text{if $X > 0$}
              \end{cases}
              $$
                where $F_Y(y)$ is the distribution function of\ $Y$.
              The probability function for $X\mid X > 0$, the continuous part of the distribution only, is
              \begin{equation}
              f_{X\mid X > 0}(x) 
              = \frac{f_X(x)}{\Pr(X > 0)} 
              = \frac{f_Y(y)}{1 - F_Y(0)}
              (\#eq:PDFpositiveConditional)
                \end{equation}
                where again $F_Y(y)$ is the distribution function of\ $Y$.
                The distribution function of\ $X$ is
                $$
                  F_{X\mid X>0}(x) =
                  \begin{cases}
                0             & \text{for $x < 0$}\\
                F_Y(0)        & \text{for $x = 0$}\\ 
                F_Y(x)        & \text{for $x > 0$}
                \end{cases}
                $$
                  
                  
                  If the latent distribution is a normal distribution,\index{Normal distribution} Equation\ \@ref(eq:PDFpositiveConditional) can be written in terms of the standard normal density function $\phi(\cdot)$ and the standard normal distribution function $\Phi(\cdot)$ (see Sect.\ \@ref(StandardNormal)).
                In this case only, notice that 
                \begin{align*}
                \Pr(X > 0) 
                &= \Pr(Z > (0 - \mu)/\sigma)\quad\text{(where $Z$ is a standard normal variate)} \\ 
                &= \Pr(Z > -\mu/\sigma) \\
                &= 1 - \Phi( -\mu/\sigma) \\
                &= \Phi(\mu/\sigma) \quad\text{(using the symmetry of the normal distribution)},
                \end{align*}
                and hence Equation\ \@ref(eq:PDFpositiveConditional) can be written as
                \begin{equation}
                f_{X\mid X>0}(x) 
                = \frac{\phi\big((x-\mu)/\sigma\big)}{\Phi(\mu/\sigma)}.
                (\#eq:PDFpositiveConditional2)
                  \end{equation}
                  If we write $t = \mu/\sigma$ and define $\lambda(t) = \phi(t)/\Phi(t)$, this expression becomes
                  $$
                    f_{X\mid X>0}(x) = ?????????????
                    $$
                    LOST:
                    $$
                    f_{X\mid X>0}(x) 
                  = \frac{f_Z(x)}{\Pr(Z > 0)} 
                  = \frac{f_Z(x)}{1 - F_Z(0)}
                  $$
                    
                    
                    
                    
                    :::{.example #TobitExamplesLatent2 name="Latent variables and censoring"}
                      Consider the number of hours worked (which can be observed), which must take a non-negative value.
                      We could assume an underlying, unobserved latent variable such as 'willingness to work'.
                      Negative values for 'willingness to work', no matter how small or how large, all result in zero observed hours of work.
                      :::
                        
                        
                        ### Properties {#CensoredModelsProperties}
                        
                        The expected value of the censored random variable\ $X$ can be found first separating the discrete and continuous parts of the distribution:
                        \begin{align*}
                      \operatorname{E}[X] 
                      &= \operatorname{E}[X\mid X = 0] + \operatorname{E}[X\mid X > 0]\\
                      &= \Pr(X = 0)\times 0 + \operatorname{E}[X\mid X > 0]\\
                      &= \operatorname{E}[X\mid X > 0]\\
                      &= \int_0^\infty x\cdot f_{X\mid X > 0}(x)\,dx,\\
                      &= \frac{1}{\Phi()} \int_0^\infty x\cdot f_{X}(x)\,dx,
                      \end{align*}
                      where $f_{X\mid X > 0}(x)$ is given in Equation\ \@ref(eq:PDFpositiveConditional2)).
                
                When the censoring happens at $x^* = 0$, the standardised score is $(x^* - \mu)/\sigma = -t$, where defining $t = \mu/\sigma$ proves useful.
                Then $1 - \Phi()$
                  Thus,
                \begin{align*}
                \operatorname{E}[X] 
                &= \int_0^\infty x\cdot f_{X\mid X > 0}(x)\,dx,
                &= \int_0^\infty x\cdot f_{X\mid X > 0}(x)\,dx,
                \end{align*}
                
                Similarly,
                $$
                  \operatorname{E}[X^2] 
                = \operatorname{E}[X^2 \mid X > 0]
                = \int_0^\infty x^2 \cdot f_{X\mid X>0}(x)\, dx,
                $$
                  from which the variance can be obtained.
                The same approach also gives the MGF as
                $$
                  M_X(t) = M_{X\mid X > 0}(t).
                $$
                  Clearly, further progress with these expressions requires knowing the distribution of\ $X$ and hence the distribution of the latent variable\ $Z$.
                It is common, but by no means universal, for the latent variable to be described by a normal distribution (when the models are also called *Tobit models*).\index{Tobit models}
                
                
                :::{.example #TobitExample0 name="Censored model"}
                  Consider a latent variable\ $Y$, and an observed variable\ $X = \text{max}(0, Z)$.
                  Suppose that\ $Z$ has the normal distribution\index{Normal distribution} (Fig.\ \@ref(fig:TobitPic), left panel))
              $$
                Z \sim N(\mu = 0, \sigma^2 = 1),
              $$
                so that $\operatorname{E}[Z] = 0$ and $\operatorname{var}[Z] = 1$.
              That is, $Z$ has a standard normal distribution (Sect\ \@ref(StandardNormal)), and so
              $$
                \Pr(Z < 0) = \Phi(0) = 0.5.
              $$
                Recall the notation: $\Phi(\cdot)$ is the distribution function for a standard normal variate, and $\phi(\cdot)$ is the density function for a standard normal variate.
              
              Then, define the random variable\ $X$ as
              $$
                f_X(x) = 
                \begin{cases}
              0          & \text{if $Z < 0$};\\
              \Phi(0)    & \text{if $Z = 0$};\\
              \phi(x)    & \text{if $Z > 0$},
              \end{cases}
              $$
                as shown in Fig.\ \@ref(fig:TobitPic) (right panel).
              
              From Equation\ \@ref(eq:PDFpositiveConditional),
              $$
                f_{X\mid X > 0}(x) 
              = \frac{\phi(x)}{1 - \Phi(0)}
              = 2 \phi(x)
              $$
                for $X > 0$.
              From this expression, 
              $$
                \operatorname{E}[X] 
              = \operatorname{E}[X \mid X > 0] 
              = 2 \int_0^\infty x\cdot \phi(x)\, dx 
              = \frac{2}{\sqrt{2\pi}} 
              \approx 0.7979,
              $$
                integrating the expression for $\phi(x)$ (Equation\ \@ref(eq:StandardNormalPMF)).
              Similarly, using integration by parts,
              $$
                \operatorname{E}[X^2] 
              = \operatorname{E}[X^2 \mid X > 0] 
              = 2 \int_0^\infty x^2\cdot \phi(x)\, dx 
              = 1/2.
              $$
                Combining, this means that 
              $$
                \operatorname{var}[X] 
              = \frac{1}{2} - \left(\frac{2}{\sqrt{2\pi}}\right)^2 
              = \frac{1}{2} - \frac{1}{\pi} 
              \approx 0.182.
              $$
                :::
                
                
                
                ```{r TobitPic, echo=FALSE, fig.align="center", fig.cap="A censored model, using a normal distribution for the continuous component, with the threshold value at $Z = 0$.", fig.width=8, fig.height=4, out.width="90%"}
              par( mfrow = c(1, 2))
              
              x.mean <- 0
              x.sd <- 1
              p <- pnorm(0,
                         mean = x.mean,
                         sd = x.sd)
              
              x <- seq(from = x.mean - (3.5 * x.sd),
                       to =   x.mean + (3.5 * x.sd),
                       length = 1000)
              
              y <-  dnorm(x,
                          mean = x.mean,
                          sd = x.sd)
              
              plot(y ~ x,
                   xlim = range(x),
                   ylim = c(0, 0.5),
                   type = "n",
                   axes = FALSE,
                   xlab = expression(Latent~variable~italic(Z)),
                   ylab = expression(Prob.~"function"),
                   main = expression( atop(A~censored~model*":",
                                           the~latent~variable~italic(Z))),
                   lwd = 3)
              axis(side = 1)
              abline(h = 0,
                     col = "grey")
              
              # Shade for X < 0
              xNeg <- x[ x <= 0]
              yNeg <- y[ x <= 0]
              
              polygon( x = c(xNeg, rev(xNeg)),
                       y = c(yNeg, rep(0, length(yNeg))),
                       col = "lightgrey",
                       border = NA) # Omit border
              
              # Different colours for x >= 0 and x < 0
              lines( y[x > 0] ~ x[x > 0],
                     lwd = 2,
                     col = "black")
              lines( y[x <= 0] ~ x[x <= 0],
                     lwd = 3,
                     col = "darkgrey")
              
              # x = 0
              abline(v = 0,
                     col = "grey",
                     lty = 2)
              
              box()
              
              
              # Explanatory labels
              arrows(x0 = -2,
                     x1 = -1,
                     y0 = 0.2,
                     y1 = 0.08,
                     length = 0.15,
                     angle = 15)
              text(x = -2,
                   y = 0.2,
                   pos = 3,
                   cex = 0.9,
                   label = expression(atop(Where,
                                           italic(Z)<0) ) )
              
              ################################################################################
              
              plot(y[x > 0] ~ x[x > 0],
                   xlim = range(x),
                   ylim = c(0, 0.5),
                   type = "l",
                   axes = FALSE,
                   xlab = expression(Observed~variable~italic(X)),
                   ylab = expression(Prob.~"function"),
                   main = expression( atop(A~censored~model*":",
                                           the~censored~variable~italic(X))),
                   
                   lwd = 3)
              abline(h = 0,
                     col = "grey")
              axis(side = 1)
              axis(side = 2,
                   las = 1,
                   at = p)
              box()
              
              
              points(x = 0,
                     y = p,
                     pch = 19)
              
              
              # Explanatory labels
              arrows(x0 = 2.25,
                     x1 = 0,
                     y0 = 0.4,
                     y1 = p,
                     length = 0.15,
                     angle = 15)
              text(x = 2.25,
                   y = 0.4,
                   pos = 1,
                   cex = 0.9,
                   label = expression( atop(Discrete~prob.,
                                            at~italic(X)==0) ) )
              
              
              # arrows(x0 = 4,
              #        x1 = 3,
              #        y0 = 0.3,
              #        y1 = dnorm(3, mean = x.mean, sd = x.sd),
              #        length = 0.15,
              #        angle = 15)
              # text(x = 4,
              #      y = 0.3,
              #      pos = 3,
              #      cex = 0.9,
              #      label = expression( atop(Normal~distribution)) ) 
              
              # X < 0 part
              lines( y[x <= 0] ~ x[x <= 0],
                     lwd = 1,
                     lty = 3,
                     col = "darkgrey")
              
              # x = 0
              abline(v = 0,
                     col = "grey",
                     lty = 2)
              
              
              ```
              
              
              
              
              
              \index{Censored models|)}
              
              
              
              
              

```{r CensoredSim, echo=FALSE, fig.align="center", fig.cap="FIX!!!!! The solid dot at $X = 0$ represents the discrete probability.", fig.width=8, fig.height=4, out.width="90%"}
set.seed(76103) # For reproducibility
num_Sims <- 1000

Y <- rnorm(num_Sims,
           mean = 1,
           sd = 1.5)
Yneg <- Y < 0  ### OR LESS THAN OR EQUAL TO????????????????????????????/
X <- Y
X[Yneg] <- 0
p0 <- sum(Yneg)/num_Sims

par(mfrow = c(1, 2))
latent_Breaks <- seq(-5, 8,
                     by = 1)
hist(Y,
     freq = FALSE,
     las = 1,
     breaks = latent_Breaks,
     col = c( rep( "grey", length(latent_Breaks[latent_Breaks<0])),
              rep(plotColour1, length(latent_Breaks[latent_Breaks>=0])) ),  
     main = expression(Latent~variable),
     ylab = "Probabiity fn",
     xlab = expression(Observed~variable~italic(X)) )

abline(v = 0,
       lwd = 2,
       col = "grey",
       lty = 2)

hist(X[Y > 0],
     freq = FALSE,
     las = 1,
     breaks = latent_Breaks[latent_Breaks>=0],
     main = expression(Observed~variable),
     ylab = "Probabiity fn",
     xlab = expression(Observed~variable~italic(X)) )
points(x = 0,
       y = p0,
       pch = 19)
```

...





First consider a censored model.
The latent variable could represent how favourable the conditions are for rainfall' say $Y\sim N(\mu_R = 50, \sigma_R^2)$, with $\Phi(0) = \Pr(Y < 0) = 0.30$, for some variance $\sigma_R^2$.
First, we can find the value of the standard normal variate\ $Z$ such that $\Phi(z) = 0.30 = \Pr(Y < 0)$:


```{r}
p_Zero_Rainfall <- 0.30
qnorm(p_Zero_Rainfall) # This find z such that Phi(z) = 0.30
```

This value of\ $z$ is the value such that
$$
  z = -0.5244 = \frac{x^* - \mu_R}{\sigma^2_R} = \frac{0 - 50}{\sigma_R^2},
$$
from which we find $\sigma_R = 95.35$.
Thus, the monthly rainfall can be simulated:




```{r, fig.align="center", fig.width=7, fig.height=4,out.width="80%", fig.cap="The latent variable left panel), and the observed rainfall (right panel)."}
set.seed(76103) # For reproducibility
num_Sims <- 1000

mean_R <- 50
sd_R <- -mean_R / qnorm(p_Zero_Rainfall) 

Y <- rnorm(num_Sims,
           mean = mean_R,
           sd = sd_R)
p0 <- 0.30

# To get sd_R, note: z = (0 - mu)/sigma = \Phi(p0) => -mu/sigma = qnorm(p0)
threshold_R <- mean_R  - qnorm(p_Zero_Rainfall) * sd_R # SHOULD BE AT ZERO

###

par(mfrow = c(1, 2))
latent_Breaks <- seq(-250, 450,
                     by = 50)

hist(Y,
     freq = FALSE,
     las = 1,
     breaks = latent_Breaks,
     col = c( rep( "grey", length(latent_Breaks[latent_Breaks < 0])),
              rep(plotColour1, length(latent_Breaks[latent_Breaks >= 0])) ),  
     main = expression(Latent~variable),
     ylab = "Probabiity fn",
     xlab = expression(Latent~variable~italic(X)) )

abline(v = 0,
       lwd = 2,
       col = "grey",
       lty = 2)

hist(Y[Y>0],
     freq = FALSE, 
     las = 1,
     breaks = latent_Breaks[latent_Breaks>=0],
     main = expression(Observed~variable),
     ylab = "Probabiity fn",
     xlab = expression(Observed~variable~italic(X)) )
points(x = 0,
       y = p0,
       pch = 19)
```


```{r}
num_Sims <- 1000

#Simulate

# Compute max monthly rainfall

```


---




If we use the variance from the censored model (i.e., $\sigma^2 = 95.35$), we can find the variance of the gamma distribution:
\begin{align*}
  \operatorname{var}[X] 
  &= p(\sigma^2)^+ + p(1 - p)(\mu^+)^2\\
  95.35
  &= 0.7(\sigma^2)^+ + 0.7\times 0.3(71.43)^2;
\end{align*}
solving shows that we should use $(\sigma^2)^ = ??$ for the variance of the gamma distribution (when $X > 0$).


We could suppose a gamma distribution for the rainfall amounts when rain is recorded; thus $\alpha\beta = 71.43$. 
If we use the variance from the censored model (i.e., $\sigma^2 = 95.35$), we can find the variance of the gamma distribution:
\begin{align*}
  \operatorname{var}[X] 
  &= p(\sigma^2)^+ + p(1 - p)(\mu^+)^2\\
  95.35
  &= 0.7(\sigma^2)^+ + 0.7\times 0.3(71.43)^2;
\end{align*}
solving shows that we should use $(\sigma^2)^ = ??$ for the variance of the gamma distribution (when $X > 0$).

```{r}
( 95.35 - 0.7*0.3*(71.43)^2 )/0.7
```

OOPS!!
